### 2019-01-10 paper/main.tex
initial commit

\documentclass[a4paper,oribibl,envcountsame]{llncs}
\pagestyle{plain}
\usepackage[T1]{fontenc}
\usepackage[misc]{ifsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{bussproofs}
\usepackage{cite}
\usepackage{color}
\usepackage{enumitem}
%\let\vec=\undefined
%\usepackage{mdsymbol}  %%% indirectly needs texlive-generic-extra
\usepackage{stmaryrd}
\usepackage{subfig}
\usepackage{textcomp}
\usepackage{url}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{todonotes}
\usepackage{enumitem}

\usepackage{cprotect}
\usepackage{listings}
\usepackage{lstautogobble}

\usepackage{array}
\newcolumntype{x}[1]{>{\centering\arraybackslash\hspace{0pt}}b{#1}}
\usepackage{diagbox}
\usepackage{siunitx}

\usepackage{tikz}
\usetikzlibrary{positioning,shadows,decorations, arrows, shapes, decorations.markings,decorations.pathmorphing}

\usepackage[final,tracking=true,kerning=true,spacing=true,stretch=10,shrink=10]{microtype}

%% BEGIN Times font
\usepackage{mathptmx}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
%% END Times font

\newcommand\rulename[1]{\ensuremath{\mathsf{#1}}}

\usepackage[
   a4paper,
   pdftex,
   pdftitle={Isabelle LLVM},
   pdfauthor={Peter Lammich},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}

\hyphenation{Isa-belle Nieuw-en-huis meta-logic multi-set multi-sets}

\include{lstisabelle}

\lstset{autogobble}

\begin{document}

\title{Isabelle LLVM}
\titlerunning{Isabelle LLVM}

\author{Peter Lammich}
\authorrunning {P. Lammich}
\institute{
  Technische Universit\"at M\"unchen, Munich, Germany \\
  \email{lammich@in.tum.de}
}

\maketitle

\setcounter{footnote}{0}

\begin{abstract}
We formalize the semantics of a fragment of the LLVM intermediate representation in Isabelle/HOL.
On top of this semantics, we have developed
\begin{enumerate}
\item a code generator, to translate from Isabelle/HOL to actual LLVM text; 
\item a verification condition generator, to reason about LLVM programs;
\item a backend for the Isabelle Refinement Framework, to enable a high-level refinement based approach to developing verified LLVM programs.
\end{enumerate}
As an example, we adopt an existing formalization of the Knuth-Morris-Pratt string search algorithm to our new framework.
With only slight changes to the original formalization, we obtain a verified LLVM implementation, which runs faster and uses less memory 
than the originally verified Standard ML implementation.

The trusted code base of our approach is the formalization of the semantics, which we have engineered for conciseness and readability, 
and the code generator that translates from Isabelle to LLVM text, which is merely a pretty printer. 
The other tools that we developed cannot compromise the soundness of the approach, as they only generate theorems that are proved via Isabelle's kernel.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
The Isabelle Refinement Framework~\cite{LaTu12,La13} features a stepwise refinement approach to verified algorithms, using the Isabelle/HOL theorem prover~\cite{NPW02,NiKl14}.
It has been successfully applied to verify many algorithms and software systems, 
among them LTL and timed automata model checkers\cite{ELNN13,BrLa18,WiLa18}, network flow algorithms~\cite{LaSe17,LaSe16}, a verified SAT-solver certification tool~\cite{La17_CADE,La17_SAT}, and even a verified SAT solver~\cite{FBL18}.
Using Isabelle/HOL's code generator~\cite{HKKN13}, the verified algorithms can be extracted to functional languages like Haskell or Standard ML.
However, the code generator only provides partial correctness guarantees, i.e., termination of the generated code cannot be proved.
Moreover, the generated code is typically much slower than the same algorithms implemented in C or Java.

The original Refinement Framework could only generate purely functional code. 
The first remedy to the performance problem was to introduce array data structures that behave like a functional 
list on the surface, but are implemented by a destructively updated array behind the scenes, similar to Haskell's now deprecated DiffArray.
While this gained some performance, the array-based implementation itself was not verified, such that we had to trust its correctness. 
Moreover, an array access still required a significant amount of overhead compared to a simple pointer dereference in C.

The next step was the Sepref tool~\cite{La15}. It generates code for Imperative HOL~\cite{BKHEM08}, which provides a heap monad inside Isabelle/HOL, 
and a code generator extension to generate code that uses the stateful arrays provided by ML, or the heap monad of Haskell.
The Sepref tool performs automatic data refinement from abstract data types like maps or sets to concrete implementations like hash tables, which 
can be placed on the heap and destructively updated. 
Moreover, it provides tools~\cite{La16} to assist in the definition of new data structures, exploiting `free theorems'~\cite{Wad89} that it 
obtains from parametricity properties of the abstract data types.
%
Using Imperative/HOL as backend, we gained some additional performance: For example, the GRAT tool~\cite{La17_SAT} provides a 
verified checker for UNSAT certificates in the DRAT format~\cite{WHH14}. It is faster than the unverified state-of-the art 
checker \textsc{drat-trim}~\cite{WHH14}, which is written in C. However, the GRAT tool spends most of its run time in an unverified 
certificate preprocessor. Nevertheless, optimizing the verified part of the code is important: The very same technique was also implemented in Coq,
using purely functional data structures~\cite{CMS17,CHHKS17}. There, the verified code was actually the 
bottleneck\footnote{Later, the checker was rewritten in ACL2, also using imperative data structures~\cite{CHHKS17,HHKW17}}.

This paper presents a next step towards efficient verified software: A refinement framework to generate verified code in LLVM intermediate representation~\cite{LLVM-manual} 
with total correctness guarantees. LLVM is an imperative intermediate language with a powerful and well-tested optimizing compiler.
We first formalize the semantics of a fragment\footnote{We support integer instructions, pointers, a simplified memory model, and simplified control flow. We expect this fragment to be sufficient for most algorithms that have already been verified with the refinement framework.} of LLVM in Isabelle/HOL, and provide a code generator to actual LLVM text (\S\ref{sec:semantics}). 
On top of this semantics, we build a separation logic and a verification condition generator, which allows convenient reasoning about LLVM programs (\S\ref{sec:vcg}).
Finally, we modify the Sepref tool to target our LLVM semantics instead of Imperative/HOL (\S\ref{sec:auto_ref}). 
This way, we connect the Refinement Framework to our LLVM code generator. This only affects the last refinement step, 
such that most parts of existing verifications can be reused. As a case study, we apply only minor modifications to an existing formalization~\cite{HeLa17} of
the Knuth-Morris-Pratt string search algorithm~\cite{KMP77}, and obtain a verified 
LLVM implementation that is an order of magnitude faster and uses less memory than the originally verified implementation in Standard ML.
The Isabelle theories described in this paper are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm}.

\section{LLVM Semantics}\label{sec:semantics}
\subsection{State Monad}
The basis of our LLVM semantics is a state-error monad, which we use to conveniently model the preconditions of instructions, their effect 
on memory, as well as arbitrary recursive programs. We define the algebraic data types
\begin{lstlisting}
('a,'s) M = M (run: "'s => ('a,'s) mres")              ('a,'s) mres = NTERM | FAIL | SUCC 'a 's
\end{lstlisting}
An entity of type \q{\is{('a,'s) M}} contains a function \q{\is{run}} that maps a start state of type \q{\is{'s}} to 
a \emph{monad result} that indicates either nontermination, a failure, or a successful execution with a result of type \q{\is{'a}} and a new state.
We define the standard monad combinators:
\begin{lstlisting}
  "return x  = M (%s. SUCC x s)"                                           "get   = M (%s. SUCC s s)"
  "fail         = M (%_. FAIL)"                                                  "set s = M (%_. SUCC () s)"
  "bind m f   = M (%s. case run m s of SUCC x s => run (f x) s | r => r)"
  "assert \<Phi> = if \<Phi> then return () else fail"
\end{lstlisting}
That is, \q{\is{return x}} returns result \q{\is{x}} without changing the state, \q{\is{fail}} aborts the computation,
\q{\is{get}} returns the current state, and \q{\is{set s}} updates the current state.
Finally, \q{\is{bind m f}} first executes \q{\is{m}}, and then \q{\is{f}} with the result of \q{\is{m}}.
If \q{\is{m}} fails or does not terminate, the whole bind fails or does not terminate.  
The derived \q{\is{assert \<Phi>}} combinator can be conveniently used to abort the computation 
if some precondition is violated, e.g., on division by zero.

We use do-notation, i.e.\ \q{\is|do { x<-m; f x }|} is short for \q{\is{bind m (\<lambda>x. f x)}}.
Moreover, we define a flat CCPO\cite{Mark76} on \q{\is{mres}}, with \q{\is{\<bot> := NTERM}}.
We define \q{\is{REC F}} to be the least fixed point of a monotonic function \q{\is{F :: ('a => ('b,'s) M) => 'a => ('b,'s) M}}.
As functions defined using the monad combinators are monotonic by construction~\cite{Kr10}, 
we can define arbitrary recursive computations. The partial function package~\cite{Kr10} supports automation for monotonicity 
proofs and for defining simple recursive functions. Mutual recursion still requires some manual effort, though it could be automated, too.

% \subsubsection{Lenses}
% Lenses \cite{FGMPS07} are a convenient tool to express positions in structured data, which can be read from and written to.
% We define a notion of lenses that allows for the position to be invalid, depending on the data. 
% For example, the position ``nth element'' may be valid or not, depending on the element of the list. 
% A lens \q{\is{L}} is a pair of functions get and put, such that 
% \begin{lstlisting}
% get_put: "put L x s = Some s' ==> get L s' = Some x"
% put_get: "get L s = Some x ==> put L x s = Some s"
% put_put: "[|put L x s = Some s'; put L y s = Some s''|] ==> put L x s'' = Some s'"
% 
% put_indep: "put L y s != None <--> get L s != None"
% \end{lstlisting}
% 
% The type of a lens is \q{\is{'s ==> 'b}}, where \q{\is{'s}} (small) is the type at the specified position, and \q{\is{'b}} is the type of the whole.
% 
% The first three laws formalize the intuition of a position which we can get a value from, and put a value to:
% The value we get is the last one that we have put (\q{\is{get_put}}); if we put the value that is already there, nothing changes (\q{\is{put_get}}); 
% and if we first put one value, and then another one, the first value is overwritten by the second one (\q{\is{put_put}}).
% Finally, validity of a position does not depend on the value stored at this position (\q{\is{put_indep}}). 
% 
% Lenses can be composed. The lens \q{\is{L_1 \<bullet> L_2}} first focuses on a position using \q{\is{L_1}}, and then further focuses using \q{\is{L_2}}.
% \begin{example}
%   \q{\is{fstL :: 'a ==> 'a \x 'b}} is the lens that focuses on the first element of a pair, and \q{\is{idxL i :: 'a ==> 'a list}} is the lens that 
%   focuses on the $i$th position of a list.
%   That is, \q{\is{fstL\<bullet>idxL i :: 'a ==> 'a list \x 'b}} focuses on the ith position of the first element of a pair, e.g., we have
%   \q{\is{put (fstL\<bullet>idxL 2) 42 ([0,1,2,3], y) = Some ([0,1,42,3],y)}} and \q{\is{get (idxL 2) [] = None}}.
%   In the first example, we replace element 2 (indexing starts at 0) by 42, in the first element of the pair. 
%   In the second example, the position is invalid.
% \end{example}
% 
% We combine lenses and state monads by defining a \q{\is{zoom :: ('s ==> 'b) => ('a,'s,'f) M => ('a,'b,'f) M}} combinator.
% \q{\is{zoom L m}} executes \q{\is{m}} on the position focused by \q{\is{L}} of the state. 
% 
% Moreover, we use shortcut notations for applying lenses to the current state: 
% \q{\is{L:=x}} puts the value \q{\is{x}} into the part of the state focused by \q{\is{L}}, and \q{\is{use L}} gets the part of the state focused by \q{\is{L}}.
% 
%

\subsection{Memory Model}
We use a high-level memory model that does not directly expose the bit-level representation of values, and assumes an infinite supply of memory. 
The memory is modeled as a list of blocks. Each block is either deallocated, or it is a list of values.
A value is a pair of values, a pointer, or an integer. We model memory by the following data types\footnote{We have simplified the presentation a bit. 
The actual implementation defines the concepts memory, block, and value independently, in order to ease future extensions, to, e.g., arbitrary structs instead of pairs.
% . The memory is parameterized over the block type, 
% the block type is parameterized over the value type, and the value type is parameterized over the primitive value type. This allows to 
% exchange parts of the memory model (e.g.\ include arbitrary structures instead of pairs) without breaking the rest of the memory model. 
% What we present here is the instance of our generic memory model that we use for this project.
}:
\begin{lstlisting}
memory = MEMORY (block option list)             block     = val list 
val = PAIR val val | PRIM primval                     primval = PV_INT lint | PV_PTR rptr
\end{lstlisting}
%
Here, the type \q{\is{lint}} is a fixed bit width integer type with a two's complement semantics, as used by LLVM. 
The type \q{\is{rptr}} is either null, or an address. An address is a path through the memory structure to a value:
\begin{lstlisting}
rptr  = NULL | ADDR nat nat (va_dir list)                                 va_dir = PFST | PSND
\end{lstlisting}
An address consists of a \emph{block index}, a \emph{value index}, and a \emph{value address}, which is a list of directions 
to either descend into the first or the second value of a pair.

For the rest of this paper, we will use the state monad with a memory as state. Thus, we define the type \q{\is{'a llM = ('a,memory) M}}.
It is straightforward to define functions \q{\is{load :: rptr => val llM}} and \q{\is{put :: val => rptr => unit llM}} to read/write a value from/to a pointer, 
or fail if the pointer is invalid.
For the actual store function, we check that the structure of the value does not change, i.e. pairs remain pairs, pointers remain pointers, and integers of width $w$ remain integers of width $w$.
\begin{lstlisting}
  store x p = do { y <- load p; assert (vstruct x = vstruct y); put x p }
\end{lstlisting}
%
% 
% 
% Given a pointer \q{\is{p :: rptr}}, it is straightforward to define a lens \q{\is{ptr_L p :: val ==> memory}} that focuses on the pointed to value, 
% or fails if the pointer is invalid (null pointer, pointer to deallocated block, index out of bounds).
% Using this lens, we define a load and store function:
% \begin{lstlisting}
%   load p = use (ptr_L p)
%   store x p = do { y <- use (ptr_L p); assert (vstruct x = vstruct y); ptr_L p := x }
% \end{lstlisting}
% The load function simply returns the pointed-to value. The store function additionally checks that the structure of the value does not change, i.e.,
% pairs remain pairs, pointers remain pointers, and integers of width $w$ remain integers of width $w$. 
% This check is important in order to map our memory model to the actual LLVM memory model, which allocates an amount of memory matching the type of the value.
%
Similarly, we define an allocate and a free function:\\[-2ex]
\begin{minipage}[t]{.46\textwidth}
\begin{lstlisting}
  allocn v n = do {
    blocks <- get; 
    set (blocks@[Some (replicate n v)]);
    return (ADDR |blocks| 0 []) }
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{.52\textwidth}
\begin{lstlisting}
  free (ADDR bi 0 []) = do {
    blocks <- get; 
    assert (bi < |blocks| \and blocks!bi ~= None);
    set (blocks[bi:=None]) }
  free _ = fail
\end{lstlisting}
\end{minipage}
Here, \q{\is{l_1@l_2}} concatenates two lists, \q{\is{|l|}} is the length of list \q{\is{l}}, \q{\is{l!i}} is the $i$th 
element of \q{\is{l}}, and \q{\is{l[i:=x]}} replaces the $i$th element of \q{\is{l}} by \q{\is{x}}.
The allocate function takes an initial value and a block size, appends a new block to the memory, and returns a pointer to the start of the new block (value index $0$, and value address $[]$).
The free function expects a pointer to the start of a block,
checks that this block is not already deallocated, and then deallocates the block by setting it to \q{\is{None}}.

\subsection{Towards a Shallow Embedding}
While we explicitly model values in memory by the type \q{\is{val}}, we model values in registers in a more shallow fashion:
We will identify LLVM registers with Isabelle variables that have a type of the shape \q{\is{T = T \x T | n word | T ptr}}. 
Here, \q{\is{\x}} is Isabelle's product type, \q{\is{n word}} is the $n$ bit word type from Isabelle's word 
library\cprotect\footnote{For convenient notation, we use the type \q{\is{n word}} as if it were a type depending on a variable $n$. 
Isabelle/HOL is not dependently typed. Instead, $n$ is actually a type variable with type-class \q{\is{len}}, which provides a 
function \q{\is{len_of :: 'a::len itself => nat}} to extract the length as a term.}, 
and \q{\is{'a ptr}} is a pointer with an attached phantom type for the value pointed to (\q{\is{'a ptr = PTR rptr}}).
%
For each type \q{\is{'a}} of shape \q{\is{T}}, we define the functions:
\begin{lstlisting}
to_val      :: "'a => val"             struct_of :: "'a itself => vstruct"
from_val :: "val => 'a"             init          :: 'a
\end{lstlisting}
such that
\begin{lstlisting}
"from_val o to_val = id"                "vstruct (to_val x) = (struct_of TYPE('a))"
"to_val init             = zero_initializer (struct_of TYPE('a))"
\end{lstlisting}
Here, \q{\is{TYPE('a) :: 'a itself}} reflects type \q{\is{'a}} into a term.
The functions \q{\is{to_val}} and \q{\is{from_val}} inject a T-shaped type \q{\is{'a}} into a value with structure \q{\is{struct_of TYPE('a)}}.
Moreover, \q{\is{init::'a}} corresponds to the all-zeroes value, i.e., the value where all pointers are null pointers, and all integers are $0$.

\subsection{LLVM Instructions}
In a next step, we define the LLVM instructions. Each LLVM instruction is identified with an Isabelle function. For example, the load instruction
is modeled by:
\begin{lstlisting}
  ll_load :: "'a ptr => 'a llM"
  ll_load (PTR p) = do {
    v <- load p;
    assert (vstruct v = struct_of TYPE('a));
    return (from_val v) }
\end{lstlisting}
It loads a value from the specified pointer, checks that its structure matches the expected type \q{\is{'a}}, and then 
converts the value to \q{\is{'a}}.

For allocation and deallocation, we provide the instructions:
\begin{lstlisting}
  ll_malloc :: 'a itself => n word => 'a ptr llM            ll_free :: "'a ptr => unit llM"
\end{lstlisting}
Note that LLVM does not contain a heap manager. 
Instead, we assume that the generated code will be linked with the C standard library, 
and produce calls to \q{\is{calloc}} and \q{\is{free}}.

We also define
instructions to extract/insert the first/second element of a pair, to offset a pointer, and to advance a pointer to the first/second element.
The code generator maps these instructions to the corresponding LLVM instructions \q{\is{getelementptr}}, \q{\is{insertvalue}}, and \q{\is{extractvalue}}.

The non-memory-related instructions are defined directly on the shallow embedding. For example, we define:
\begin{lstlisting}
  ll_udiv :: "n word => n word => n word"
  ll_udiv a b = do { assert (a ~= 0); return (a div b) }
\end{lstlisting}
where \q{\is{div}} is the unsigned division from the word library.
Note that we use assertions to model undefined behavior in LLVM by failure in our semantics.
Similarly, we define the other integer instructions.

\subsection{Modeling Control Flow}\label{sec:modeling_ctrl_flow}
Next, we put together instructions to form procedure bodies. Instead of arbitrary control flow graphs, 
we only allow structured control flow via if-then-else, procedure calls, and sequential composition: The body of
a procedure is modeled by an Isabelle term of type \q{\is{'a llM}} and shape \q{\is{block}}, where
\begin{lstlisting}
block = do { var <- cmd; block } | return var
cmd = ll_<opcode> arg^* | proc_name arg^* | llc_if arg block block
arg = var | number | null | init
\end{lstlisting}
and \q{\is{llc_if :: 1 word => 'a llM => 'a llM => 'a llM}} is defined as
\begin{lstlisting}
llc_if b t e = if b=1 then t else e
\end{lstlisting}
That is, a block is a list of commands whose results are bound to variables, terminated by a return instruction. 
A command is either an instruction, a procedure call, or an \q{\is{llc_if}}. 
The arguments of instructions and procedure calls, as well as the condition of an \q{\is{llc_if}}, must be variables or constants (numbers, the null pointer, or a zero-initialized value).
%
A program is represented by a set of (monomorphic) theorems of the shape
\q{\is{proc\_i x_1 ... x\_n = cmd}},
where the \q{\is{proc\_i}} are Isabelle functions, the \q{\is{x\_i}} are variables, and all free variables on the right hand side are among the \q{\is{x\_i}}.
As our monad allows arbitrary recursive definitions, we can define arbitrary recursive programs.

\begin{figure}
\centering
\begin{minipage}[t]{.39\textwidth}
  \begin{lstlisting}
    fib:: 64 word => 64 word llM
    fib n = do {
      t <- ll_icmp_ule n 1;
      llc_if t (return n) (do {
        n_1 <- ll_sub n 1;
        a   <- fib n_1;
        n_2 <- ll_sub n 2;
        b   <- fib n_2;
        c   <- ll_add a b;
        return c
      })
    }    
  \end{lstlisting}
  \caption{Program in Isabelle}\label{fig:fib_isabelle}
\end{minipage}
\hspace*{.09\textwidth}
\begin{minipage}[t]{.5\textwidth}
  \begin{lstlisting}[language=LLVM, literate={}]
  define i64 @fib(i64 %x) {
    start:
      %t = icmp ule i64 %x, 1
      br i1 %t, label %then, label %else
    then:
      br label %ctd_if
    else:
      %n_1 = sub i64 %x, 1
      %a     = call i64 @fib (i64 %n_1)
      %n_2 = sub i64 %x, 2
      %b     = call i64 @fib (i64 %n_2)
      %c     = add i64 %a, %b
      br label %ctd_if
    ctd_if:
      %x1a = phi i64 [ %x, %then ], [ %c, %else ]
      ret i64 %x1a
  }  
  \end{lstlisting}
  \caption{Generated LLVM text}\label{fig:fib_llvm}
\end{minipage}
\end{figure}

\begin{example}\label{ex:fib_isabelle}
  Figure~\ref{fig:fib_isabelle} shows the Isabelle specification of a procedure named \q{\is{fib}}, 
  which takes a 64 bit integer argument, and returns a 64 bit integer. 
  Our semantics can be directly executed inside Isabelle. The following Isabelle command evaluates our function on the first few natural numbers, and an empty memory:
  \begin{lstlisting}
    value \<open>map (%n. run (fib n) (MEMORY [])) [0,1,2,3]\<close>
    (* output: [SUCC 0 (MEMORY []), SUCC 1 ..., SUCC 1 ..., SUCC 2 ...] *)
  \end{lstlisting}
\end{example}

\subsection{Code Generation}
The LLVM intermediate representation \cite{LLVM-manual} is a strongly typed control flow graph (CFG) based intermediate language 
that uses single static assignment (SSA) form~\cite{CFRWZ91}.
A procedure is a list of basic blocks, the first block in the list being the entry point of the procedure. 
A basic block is a list of instructions, finished by a terminator instruction that determines the next basic block to execute (or to return from the current procedure).
Each non-void instruction defines a fresh register containing its result. A register can only be accessed in the part of the CFG that is dominated by its definition.
To transfer values from registers to other parts of the CFG, $\phi$-instructions have to be used. 
A $\phi$-instruction must be located at the start of a basic block. 
It lists, for each possible predecessor block, an accessible register in this predecessor block. 
The $\phi$-instruction then evaluates to the value of the register from those predecessor block from which execution was actually transferred. 
As for all instructions, the result of the $\phi$-instruction is bound to a fresh register, which can then be accessed from the current basic block.

It is straightforward to map a program defined in Isabelle to an actual LLVM program.
Each equation of the form \q{\is{proc x_1 .. x\_n = block}} is mapped to an LLVM procedure named \q{\is{proc}}. 
A block is mapped to a control flow graph. Instructions and procedure calls are directly mapped to LLVM instructions and calls.
An \q{\is{x <- llc_if b t e}} is translated to conditional branching, using a $\phi$-instruction to define the result register \q{\is{x}} when joining the control flow.

% 
% A command is mapped to a set of basic blocks and a register that holds the return value. 
% To concisely implement the translation, we use a builder pattern, that keeps track of the current basic block it is emitting instructions to.  
% The function \q{\is{build dst cmd}} translates a command, putting the result into the register named \q{\is{dst}}.
% The function \q{\is{build_block block}} translates a block, and returns the register that holds the result value of the block.
% For example, an \q{\is{llc_if}} command is translated as follows\footnote{The code displayed here is simplified, the actual code additionally 
% maintains the state of the builder and a mapping from Isabelle's variables to actual registers in LLVM. 
% We also omitted the case where the if-command returns unit, such that we do not create a $\phi$-node.}
% \begin{lstlisting}[language=ML]
%   fun build_if dst (llc_if b blk_t blk_e) = 
%   (* Omitted: Obtain fresh labels l_then, l_else, l_ctd_if *)
%   val _ = Builder.mk_cbr b l_then l_else
%   
%   val _ = Builder.open_bb l_then 
%   val r_then = build_block blk_t
%   val l_then' = Builder.mk_br l_ctd_if
%   
%   val _ = Builder.open_bb l_else 
%   val r_else = build_block blk_e
%   val l_else' = Builder.mk_br l_ctd_if
% 
%   val _ = Builder.open_bb b l_ctd_if
%   val _ = Builder.mk_phi dst [(r_then,l_then')] [(r_else,l_else')]
% \end{lstlisting}
% 
% The function takes a register \q{\is{dst}} where it should put the result to, and the Isabelle representation of the if command.
% We first obtain three fresh label names. Then, we close the current basic block with a conditional branch to the \q{\is{then}} and \q{\is{else}} label (\q{\is{mk_cbr}}).
% We then emit the code for the then and else part into the respective basic blocks, and close these basic blocks with unconditional branches to the \q{\is{ctd_if}} label.
% Finally, we open the \q{\is{ctd_if}} basic block, and emit a $\phi$-node to bind the result of the then and else part to the register \q{\is{dst}}.

\begin{example}
Figure~\ref{fig:fib_llvm} displays the output of our code generator for the \q{\is{fib}} function displayed in Figure~\ref{fig:fib_isabelle}.
\end{example}

\subsubsection{Mapping the Memory Model}
Some of the translations are slightly more involved.
For example, recall the \q{\is{ll_malloc :: 'a itself => n word => 'a ptr llM}} instruction. It has to be mapped to the function \q{\is{void* calloc(size_t, size_t)}} from
the C standard library. For this, we have to parameterize the code generator with the architecture dependent size of the \q{\is{size_t}} type. 
Next, we have to obtain the size of type \q{\is{'a}} and cast the \q{\is{n word}} parameter to \q{\is{size_t}}. Here, our code generator will refuse downcast, as this might result in bits being dropped. Finally, we have to cast the returned \q{\is{void*}} to the correct return type. 
Moreover, the \q{\is{calloc}} function returns \q{\is{null}} if not enough memory is available. In contrast, our semantics always returns a new block of memory.
We insert code to terminate the program in a defined way if it runs out of memory. The relation between our semantics and the actual LLVM program then 
becomes: Either the program terminates with an out-of-memory condition, or it behaves as modeled by the semantics. 
Our current implementation simply prints an error message and terminates the process with exit code $1$ if it runs out of memory.
A similar issue arises when comparing pointers: LLVM does not have instructions for pointer comparison. Instead, pointers have to be cast to integers, which can then be compared. However, this requires to know the bit-width of a pointer, which we cannot model in our semantics.
Instead, we model the instructions \q{\is{ll_ptrcmp_eq}} and \q{\is{ll_ptrcmp_ne}}, and let the code generator generate the cast to integers and the integer comparison.

\subsection{Preprocessing}
In the previous sections we have described the semantics of a fragment of LLVM, shallowly embedded in Isabelle/HOL. 
However, programs have to stick to a very restrictive shape (cf.~\S\ref{sec:modeling_ctrl_flow}), 
which makes them easy to map to actual LLVM code, but tedious to write directly.
In this section, we describe the preprocessing techniques that we implemented to make programs simpler to write.
Note that these techniques work by \emph{proving} the equations in restricted shape from the user specified ones. 
Thus, errors in the preprocessor cannot affect soundness of the generated code: Either, it fails to prove the equations, or it produces ill-shaped equations, 
which the code generator will reject.

Our preprocessor includes two main functionalities, monomorphization and inlining.
The goal is to produce well-shaped equations from a given set of Isabelle functions.
We assume that the functions we start with are instantiated to a monomorphic type, i.e., do not contain any type variables.

\subsubsection{Monomorphization}
We first collect the defining equations of the specified functions, and, recursively, the equations of the functions used in them.
During the collection process we perform monomorphization, i.e., 
we instantiate polymorphic definitions to the types they are actually used with. 
After this step, we have a set of monomorphic equations that define all functions that will occur in the final program.
However, the right hand sides of the equations may still be ill-shaped.

\subsubsection{Inlining}
We first apply user defined rewrite rules, and then flatten nested expressions, converting function calls to the shape 
\q{\is{r <- f x_1 ... x\_n}} or \q{\is{r <- return (f x_1 ... x\_n)}}, where the $x_i$ are either constants, variables, or \emph{monadic} arguments
of type \q{\is{... => _ llM}}. Subterms of type \q{\is{_ llM}} are recursively flattened.
We iterate the rewriting and flattening steps until a fixed point is reached. 
While this preprocessing is highly incomplete, i.e., it cannot convert every term to a well-shaped one, it works well in practice, allowing for concise specifications.

\begin{example}
Consider the following definition of the function \q{\is{fib'}}:
\begin{lstlisting}
fib' :: "m word => m word llM"
fib' n = if n <= 1 then return n 
         else do { n_1 <- fib' (n - 1); n_2 <- fib' (n - 2); return (n_1 + n_2) }
\end{lstlisting}
When started with \q{\is{fib' :: 64 word => 64 word llM}}, the preprocessor automatically translates this equation 
to the equation displayed in Figure~\ref{fig:fib_isabelle}. During the translation, it uses the following inlining rules:
\begin{lstlisting}
  if b then c else t = llc_if (from_bool b) c t                 return (a + b) = ll_add a b
  return (from_bool (a<=b)) = ll_icmp_ule a b               return (a - b) = ll_sub a b
\end{lstlisting}
Our default setup contains similar rules for the other operations, as well as rules to map tuples and case-distinctions over tuples to \q{\is{insertvalue}} and \q{\is{extractvalue}} instructions.
\end{example}

\subsubsection{Generation of Recursive Equations}
A third functionality of the preprocessor is to generate recursive functions from fixed-point combinators. In general,
it examines the right hand side of an equation for patterns \q{\is{p}} for which it has an unfold rule of the form \q{\is{p = F p}}.
It then defines a new function \q{\is{f x_1 ... x\_n = F (f x_1 ... x\_n)}}, where the \q{\is{x\_i}} are the free variables in the pattern \q{\is{p}}.
Finally, it replaces \q{\is{p}} by \q{\is{f x_1 ... x\_n}} in the equation. This way, specifications with fixed point combinators are automatically 
transformed to a set of recursive equations, as required by the code generator.

A standard example is the while loop combinator, which we define by the equation
\begin{lstlisting}
  llc_while b c s = do {ctd <- b s; llc_if ctd (do {s <- c s; llc_while b c s}) (return s)}
\end{lstlisting}
We use this equation as unfold rule, such that while loops are automatically converted to tail calls. Note that this 
feature allows for the concise specifications of loops, without adding additional trusted code. 
As LLVM has tail-call optimization, the compiled programs will be efficient.

\begin{example}\label{ex:euclid}
Consider the following program:
\begin{lstlisting}
euclid :: 64 word => 64 word => 64 word
euclid a b = do {
  (a,b) <- llc_while 
    (%(a,b) => ll_cmp (a ~= b))
    (%(a,b) => if (a<=b) then return (a,b-a) else return (a-b,b))
    (a,b);
  return a }
\end{lstlisting}
From this, the preprocessor proves the following two equations (before inlining):
\begin{lstlisting}
euclid a b = do {
    (a, b) <- euclid_0 (a, b);
    return a }
euclid_0 s = do {
    ctd <- case s of (a, b) => ll_cmp (a ~= b);
    llc_if ctd (do {
      s <- case s of (a, b) => if a <= b then return (a, b - a) else return (a - b, b);
      euclid_0 s
    }) (return s) }
\end{lstlisting}
That is, it defined a new function \q{\is{euclid_0}} to replace the while loop by tail recursion.
\end{example}

\section{Verification Condition Generator}\label{sec:vcg}
The next step towards generating verified LLVM programs is to establish a reasoning infrastructure. 
In this section, we describe our separation logic~\cite{Rey02} based verification condition generator.
Note that, while applying complex operations on the proof state, at the end, our VCG conducts a proof 
that goes through Isabelle's inference kernel. Thus, bugs in the VCG cannot cause unsoundness.

\subsection{Separation Algebra}
The first step to obtain a separation logic is to define a separation algebra on a suitable abstraction of the memory.
A separation algebra~\cite{CHY07} is a structure with a zero, a disjointness predicate $a\#b$, 
and a disjoint union $a+b$.
Intuitively, elements describe parts of the memory. 
Zero describes the empty memory, $a\#b$ means that $a$ and $b$ describe disjoint parts of the memory,
and $a+b$ describes the memory described by the union of $a$ and $b$. 
For the exact definition of a separation algebra, we refer to \cite{CHY07,KKB12}. We note that 
separation algebras naturally extend over functions, pairs, and option types.

We abstract a value by a partial function from value addresses (\q{\is{va_dir list}}) to primitive values, 
such that the addresses in the domain of the function are independent, i.e., no address is the prefix of another address:
\begin{lstlisting}
  typedef aval = "{ m :: vaddr => 'a option. \<forall>va,va'\<in>dom m. va~=va' --> indep va va' }"
  val_\<alpha> :: val => aval
  val_\<alpha> (PRIM x) = [[] |-> x]
  val_\<alpha> (PAIR x y) = PFST \<cdot> val_\<alpha> x + PSND \<cdot> val_\<alpha> y
\end{lstlisting}
here, \q{\is{[k |->v]}} is the partial function that maps $k$ to $v$, and \q{\is{i \<cdot> a}} 
prepends the item $i$ to all addresses in the domain of $a$. 
It is straightforward (though technically involved) to show that abstract values form a separation algebra, 
where the empty map is zero, maps are disjoint iff their domains are pairwise independent, and union merges two maps.

A natural abstraction of a block (\q{\is{val list}})
would be a function from indexes to abstract values, mapping invalid indexes to $0$. 
However, this abstraction does not contain enough information to reason about deallocation. 
In order to deallocate a block, we have to own the whole block. However, from the abstraction, we cannot infer the size of the block, and thus we cannot
specify an assertion that ensures that we own the whole block. A remedy (which the author has seen in \cite{App14}) is to additionally 
abstract a block to its size. Thus, abstract blocks have the type \q{\is{ablock = (nat => aval) \x nat option}}. 
The option type is required to make the second elements of the tuples a separation algebra. We use the trivial separation algebra here, where
two elements are only disjoint if at least one of them is \q{\is{None}}. 
Finally, we define \q{\is{amemory = nat => ablock}}, and a function \q{\is{\<alpha> :: memory => amemory}} that abstracts memory
by a function from block indexes to abstract blocks, mapping deallocated or invalid indexes to zero.

\subsection{Basic Reasoning Infrastructure}
Predicates of type \q{\is{assn = memory => bool}} are called \emph{assertions}.
The \emph{weakest precondition} of a program \q{\is{c :: 'a llM}}, a \emph{postcondition} \q{\is{Q :: 'a => assn}}, and a memory \q{\is{s}} is a predicate defined as:
\begin{lstlisting}
wp c Q s = (\<exists>r s'. run c s = SUCC r s' \and Q r (\<alpha> s'))
\end{lstlisting}
Intuitively, \q{\is{wp c Q s}} states that program \q{\is{c}}, if run on memory \q{\is{s}}, terminates successfully and the result \q{\is{r}} and the abstraction of the new 
state \q{\is{s'}} satisfy \q{\is{Q}}.

For assertions \q{\is{P}} and \q{\is{Q}}, the \emph{separating conjunction} \q{\is{P*Q}} describes a memory that can be split into two disjoint parts described by \q{\is{P}} and \q{\is{Q}}, respectively:
\begin{lstlisting}
(P * Q) s = \<exists>s_1 s_2. s_1 # s_2 \and s = s_1 + s_2 \and P s_1 \and Q s_2
\end{lstlisting}
%
Validity of a \emph{Hoare triple} \q{\is|{P} c {Q}|} is defined as follows:
\begin{lstlisting}
|= {P} c {Q} = \<forall>F s. (P*F) (\<alpha> s) --> wp c (%r s'. (Q r * F) s') s
\end{lstlisting}
That is, if the memory can be split into a part described by the \emph{precondition} \q{\is{P}}, and a \emph{frame} described by \q{\is{F}},
then command \q{\is{c}} will succeed, and the new memory consists of a part described by the postcondition \q{\is{Q}} and the unchanged frame. 
Our Hoare triples satisfy the frame rule: \q{\is$|= {P} c {Q} ==> |= {P * F} c {\<lambda>r. Q r * F}$} for all \q{\is{F}}.

\subsection{Basic Rules}
Once we have set up the separation algebra and the abstraction function, we can prove Hoare triples for the basic operations of our memory model.
For example, we prove the following rules for \q{\is{allocn}} and \q{\is{free}}:
\begin{lstlisting}
|= {\<box>} allocn v n {%p. malloc_tag n p * range {0..<n} (%_. v) p}
|= {malloc_tag n p * range {0..<n} blk p} free p {%_. \<box>}
\end{lstlisting}
where \q{\is{\<box> = \<lambda>s. s=0}} describes the empty memory, 
\q{\is{malloc_tag n p}} asserts that \q{\is{p}} points to the beginning of a block, and the size field of this block's abstraction is \q{\is{n}}, 
and \q{\is{range I f p}} describes that for all \q{$i\in I$}, \q{$p+i$} points to value \q{$f~i$}.
Intuitively, \q{\is{allocn}} creates a block of size \q{$n$}, initialized with values \q{$v$}, and a tag. 
If one possesses both, the whole block and the tag, it can be deallocated by free. 
% Note that we use the size encoded in the tag to ensure that we possess the whole block. 
For the actual LLVM memory instructions, we obtain the following rules:
\begin{lstlisting}
|= {n~=0} ll_malloc TYPE('a) n {%p. range {0..<n} (%_. init) p * malloc_tag n p}
|= {range {0..<n} blk p * malloc_tag n p} ll_free p {%_. \<box>}
|= {pto x p} ll_load p {%r. r=x * pto x p}
|= {pto xx p} ll_store x p {%_. pto x p}
\end{lstlisting}
Here, \q{\is{pto x p}} describes that $p$ points to value $x$. We prove similar rules for the other instructions.

\subsection{Automating the VCG}
In order to efficiently prove Hoare triples, some automation is required.
We provide a verification condition generator with a frame inference heuristics. 
The first step to prove a Hoare triple is to convert it to a proposition on weakest preconditions:
\begin{lstlisting}
  [|!!F s. STATE (P*F) s ==> wp c (%r s'. (Q r * F) s') s|] ==> |= {P} c {Q}
\end{lstlisting}
where \q{\is{STATE P s = P (\<alpha> s)}}.
In general, the VCG operates on subgoals of the form \q{\is{STATE P s ==> wp c Q s}}. 
It then iteratively performs one of the following steps\footnote{This is a simplified presentation. 
The actual VCG is an instantiation of a generic VCG framework that can be configured with various solvers, rules, and heuristics.}:
\begin{description}
  \item[simplification] Apply a rewrite rule to transform \q{\is{wp c Q s}} into some equivalent proposition. For example,
    binding is resolved by the rule:
    \begin{lstlisting}
      wp (do {x<-m; f x}) Q s = wp m (%x. wp (f x) Q) s
    \end{lstlisting}
  \item[rule] If there is a Hoare triple of the form \q{\is$|= {P'} c {Q'}$}, the VCG tries to infer a frame \q{\is{F}} 
  such that \q{\is{P |- P'*F}}, and replaces the goal by \q{\is{STATE (Q'*F) s' ==> Q s'}} for a fresh \q{\is{s'}}. 
  Here, \q{\is{P|-Q = \<forall>s. P s ==> Q s}} denotes entailment.
  \item[final] If the goal has the form \q{\is{STATE P s ==> Q s}} such that \q{\is{Q}} is not of the form \q{\is{wp _ _ _}}, a heuristics is used to prove 
  \q{\is{P |- Q}}.
\end{description}
The actual verification conditions are generated during frame inference and the final proof heuristics.
For example, the rule for \q{\is{ll_malloc}} requires to prove that the size operand is not zero. 
The VCG will try to prove these goals by a default tactic, and leave them to the user if this tactic fails.

\begin{example}\label{ex:euclid-proof}
Recall the function \q{\is{euclid :: 64 word => 64 word => 64 word llM}} from Example~\ref{ex:euclid}.
We prove the following Hoare triple:
\begin{lstlisting}
  |= {uint a a\impl * uint b b\impl * 0<a * 0<b} euclid a\impl b\impl {%r\impl. uint (gcd a b) r\impl}
\end{lstlisting}
Here, \q{\is{uint a a\impl}} states that \q{\is{a\impl::32 word}} is an unsigned integer with value \q{\is{a::int}}, 
where \q{\is{int}} is the type of (mathematical) integers in Isabelle, and \q{\is{gcd}} is Isabelle's greatest common divisor function. 
After annotating a suitable loop invariant, the VCG generates the following two verification conditions:
\begin{lstlisting}
  [| gcd x y = gcd a b; x ~= y; x <= y; ... |] ==> gcd x (y - x) = gcd a b
  [| gcd x y = gcd a b; \<not> x <= y; ... |] ==> gcd (x - y) y = gcd a b
\end{lstlisting}
These are straightforward to prove in Isabelle, e.g., using sledgehammer\cite{BBP13}.
\end{example}

\subsection{Data Structures and Basic Refinement}
Recall Example~\ref{ex:euclid-proof}. The Hoare triple that is proved there first maps the 64 bit word arguments and results to mathematical integers,
and then phrases the correctness statement in terms of mathematical integers. 
This approach is often more feasible than stating correctness on the concrete implementation directly. 
In our case, we would have to define the concept of greatest common divisor for 64 bit words. In general, an algorithm often 
computes some function on abstract mathematical concepts like integers or sets, but has to implement these by concrete data structures like 64 bit words or hash-tables.
Thus, a concise way to specify the correctness statement is to first map the implementations back to the abstract concepts, and then state 
the actual correctness abstractly.

In separation logic based reasoning, a data structure provides a \emph{refinement assertion} \q{\is{A x x\impl :: assn}}, which describes 
that the abstract value \q{\is{x}} is implemented by the concrete value \q{\is{x\impl}}. 
We define refinement assertions to implement integers and natural numbers by $n$ bit words, and to implement lists by blocks of memory. 
Note that new data structures and refinement assertions can easily be added. In general, an implementation does not completely implement an abstract mathematical concept.
For example, $n$ bit words can only represent the integers \q{\is|sints n = $\{-2^{\text{n}-1}..<2^{\text{n}-1}\}$|}, and hash-tables can only represent finite sets. Thus, the rules for the operations generally come with additional preconditions. 
For example, the rule to implement subtraction on integers by subtraction on $n$ bit words is the following:
\begin{lstlisting}
|= {sint a a\impl * sint b b\impl * a-b \<in> sints n} ll_sub a\impl b\impl {%r\impl. sint (a-b) r\impl}
  for a\impl b\impl :: n word and a b :: int
\end{lstlisting}
Note that the postcondition does not mention the operands \q{\is{a,b}} again, though they are still valid after the operation. 
As \q{\is{sint}} is \emph{pure}, i.e., does not use the memory, our VCG can automatically add the corresponding assertions to the postcondition.

\section{Automatic Refinement}\label{sec:auto_ref}
Our basic VCG infrastructure can be used to verify simple algorithms, like \q{\is{euclid}} from Example~\ref{ex:euclid-proof}.
However, many complex algorithms have already been verified using the Isabelle Refinement Framework~\cite{LaTu12}.
It features a non-deterministic programming language with a refinement calculus and a VCG. 
It allows to express an algorithm on the abstract mathematical concepts, and then refine it in multiple steps towards an efficient implementation.
The last step of a refinement is typically performed by the Sepref tool~\cite{La15}, which translates a program from the non-deterministic monad of the Refinement Framework 
into the deterministic heap monad of Imperative HOL~\cite{BKHEM08}, replacing abstract data types by concrete implementations. 
% It comes with some powerful automation and support for easily defining new data structures~\cite{La16}.
%
We have modified the Sepref tool to translate to our LLVM monad instead.
We only had to modify the translation phase. The preprocessing phases, which only work on the abstract program, remained unchanged.

The translation phase works by symbolically executing the abstract program, thereby synthesizing a structurally similar concrete program.
During the symbolic execution, the relation between the abstract and concrete variables is modeled by refinement assertions. 
The predicate \q{\is{hnr \<Gamma> m\impl \<Gamma>' R m}} means that concrete program \q{\is{m\impl}} implements abstract program \q{\is{m}},
where \q{\is{\<Gamma>}} contains the refinements for the variables before the execution, \q{\is{\<Gamma>'}} contains the 
refinements after the execution, and \q{\is{R}} is the refinement assertion for the result of \q{\is{m}}. For example, a \q{\is{bind}} is 
processed by the following rule:
\begin{lstlisting}[numbers=left]
[| hnr \<Gamma> m\impl \<Gamma>' R\_x m;
  !!x x\impl. hnr (R\_x x x\impl * \<Gamma>') (f\impl x\impl) ($R_x'$ x x\impl * \<Gamma>'') R\_y (f x);
  MK_FREE $R_x'$ free;
|] ==> hnr \<Gamma> (do {x\impl<-m\impl;r\impl<-f\impl x\impl; free x\impl; return r\impl}) \<Gamma>'' R\_y (do {x<-m; f x})
\end{lstlisting}
To refine \q{\is{x<-m; f x}}, we first execute \q{\is{m}}, synthesizing the concrete program \q{\is{m\impl}} (line~1).
The state after \q{\is{m}} is \q{\is{R\_x x x\impl * \<Gamma>'}}, where \q{\is{x}} is the result created by \q{\is{m}}.
From this state, we execute \q{\is{f x}} (line~2). The new state is \q{\is{$R_x'$ x x\impl * \<Gamma>'' * R\_y y y\impl}}, where \q{\is{y}} is the result of \q{\is{f x}}.
% , which is only implicit in the above rule.
Now, the variable \q{\is{x}} goes out of scope, such that it has to be deallocated. 
The predicate \q{\is&MK_FREE $R_x'$ free = \<forall>x x\impl. |= {$R_x'$ x x\impl} free x\impl {\<lambda>_. \<box>}&} (line~3) states that 
\q{\is{free}} is a deallocator for data structures implemented by refinement assertion \q{\is{$R_x'$}}. 
Note that the refinement for variable \q{\is{x}} may change: If \q{\is{f\impl x\impl}} overwrites \q{\is{x\impl}},
the refinement assertion for \q{\is{x}} will be changed to the special assertion \q{\is{invalid}}. 
The deallocator for \q{\is{invalid}} is simply a no-op.
%
Adding support for deallocators was the most substantial change we applied to the Sepref tool. Its original target 
language, Imperative HOL, is garbage collected, such that there is no need for explicit deallocation.

\subsection{Data Structure Library}
Once the basic Sepref tool is adapted, we can define data structures. 
Reusing the basic data structures from the original Sepref tool is not possible, as Imperative HOL 
uses arbitrary precision integers and algebraic data types, while we have only fixed width integers and pairs.
Up to now, we have added the implementation of integers and natural numbers 
by $n$ bit words, the implementation of lists by arrays, and, on top of this, the implementation of a list 
by an array and a length field. However, we could reuse the existing infrastructure of the Sepref tool: For example,
there is support to automatically generate rules that also support refinement of the elements of a data structure, 
exploiting `free theorems'~\cite{Wad89} which stem from parametricity properties of the abstract types.

\subsection{Example: Knuth-Morris-Pratt String Search}
Finally, we conducted a case study: 
We took an existing formalization~\cite{HeLa17} of the Knuth-Morris-Pratt string search~\cite{KMP77} in the Refinement Framework.
It uses Sepref to generate Imperative HOL code, which is then translated to Standard ML by Isabelle's code generator.
We only had to slightly change the abstract part of the proof, in order to insert a few in-bounds assertions. 
For example, when incrementing an index, we have to explicitly assert that it won't overflow. 
As the invariants used in the original proof already contain this knowledge, adapting the proofs was straightforward.
We then used the modified Sepref tool to synthesize LLVM instead of Imperative HOL. 
For this, we fixed the lengths and indexes to 64 bit integers, and the characters to 8 bit integers.
For the search string and text we use a pair of an array and a length field.
The result of the automatic synthesis is an LLVM specification \q{\is{kmp_impl}}, and the theorem:
\begin{lstlisting}
(kmp_impl, kmp_SPEC) : [%s t. |s| + |t| < $2^{63}$] larray^k \x larray^k -> snat_option
\end{lstlisting}
Here, \q{\is{kmp_SPEC}} is the abstract specification of a string search algorithm, \q{\is{larray}} is 
the refinement of a list by an array and a length, and \q{\is{snat_option}} implements the type \q{\is{nat option}} 
by signed 64 bit integers, mapping \q{\is{None}} to $-1$. 
The notation \q{\is{[\<Phi>] $A_1^{k|d}$ \x ... \x $A_n^{k|d}$ -> R}} specifies a refinement with precondition \q{\is{\Phi}}, such that the 
arguments are refined by \q{\is{A_1 ... A\_n}} and the result is refined by \q{\is{R}}. 
The $\cdot^{k|d}$ annotations specify whether the argument is overwritten ($k$ for keep, $d$ for destroy).
While we use this notation a lot in the Refinement Framework, it is straightforward to prove a standard Hoare triple from it. By unfolding some definitions we get:
\begin{lstlisting}
|= {larray s s\impl * larray t t\impl * |s| + |t| < $2^{63}$}
   kmp_impl s\impl t\impl
     {%i. larray s s\impl * larray t t\impl 
      * if i=-1 then \<nexists>i. sublist_at s t i 
          else i>=0 \and sublist_at s t (unat i) \and (\<forall>j<unat i. \<not>sublist_at s t j)}
\end{lstlisting}
That is, our algorithm requires the length of the search string $s$ plus the length of the text $t$ to be less than $2^{63}$. 
Then, it will return the index of the first occurrence of $s$ in $t$, or $-1$ if $s$ does not occur in $t$. 
Moreover, the algorithm will neither change $s$ nor $t$, as expressed by the respective assertions in the postcondition.

The code generator produces actual LLVM text from \q{\is{kmp_impl}}, which we compile and link with a simple command line interface written in C, 
using \textsc{clang}~\cite{Clang}.
Similarly, we generate Standard ML code from the original formalization, which we compile using \textsc{MLton}~\cite{MLton}.
We compare the programs by running them on a large text that does not contain the search string: 
The program compiled by LLVM is an order of magnitude faster than the one compiled by MLton, and uses a third of the memory.
We leave more systematic testing to future work.

\section{Conclusions}
We have formalized a semantics and a code generator for a fragment of LLVM. On top of the semantics, we have built a verification 
infrastructure, and re-targeted the Sepref tool to connect the Refinement Framework to LLVM. As a case-study, we have 
only slightly changed an existing formalization of the Knuth-Morris-Pratt string search algorithm to obtain a verified LLVM implementation,
which is an order of magnitude faster than the Standard-ML implementation from the original formalization.

Our LLVM semantics makes a few simplifying assumptions: It assumes an infinite supply of memory, and thus cannot assign a bit-size to pointers. 
This assumption helps us to retain a deterministic semantics, which allows us to evaluate the semantics inside the theorem prover (cf.~Example~\ref{ex:fib_isabelle}). 
We plan to use this feature for systematic testing of our semantics against the actual LLVM compiler. 
A similar assumption is implicitly made for the stack, as our semantics permits arbitrarily deep recursive procedure calls.
We remedy this mismatch between semantics and reality by terminating the program in a defined way if it runs out of heap.
To protect against stack overflows, LLVM provides mechanisms like stack probing or split stack, which, however, require some effort to enable. 
We leave that to future work, and note that our generated code allocates no large blocks of memory on the stack. 
Thus, stack overflows are likely to hit the guard pages inserted by most operating systems, which will cause termination of the process.

This project would not have been possible without several independent Isabelle developments:
We use the Separation Algebra library~\cite{KKB12_afp,KKB12} as basis for our separation logic. We substantially extended this library by a 
frame inference heuristics, and formalized the extension of separation algebras over functions, products, and options. 
Moreover, we use Isabelle's machine word library~\cite{Word_Lib-AFP} to model the 2's complement arithmetic of LLVM. We slightly extended this library by adding a few lemmas. Finally, the Eisbach language~\cite{MMW16} was a great help for prototyping the verification condition generator, although most of the 
final VCG is now implemented directly in the more low-level Isabelle/ML.

Related work includes the Vellvm project \cite{ZNMZ12,ZNMZ13}, which formalizes a substantial fragment of LLVM in Coq.
While we focus on generating verified programs, they focus on proving correctness of program analyses and transformations. 
Moreover, we drew some of the ideas for our separation logic from the Verifiable C project~\cite{App14}, a Coq formalization of a separation logic on top of the CompCert C semantics~\cite{BL09}.

Future work will include the verification of more basic data structures, which will enable us to 
port more existing verification projects to LLVM, hopefully increasing their performance. 
One challenge will be the support of sum types: Unfortunately, union types have been discontinued in LLVM, such that we have to resort 
to low-level bit-fiddling to support them. However, this will increase the complexity of the code generator. An alternative would be a more sophisticated memory model.
The current shallow embedding of the control flow semantics also has the disadvantage that we cannot easily model more complex control flow like breaking from loops or exceptions, without significantly increasing the complexity of the code generator. To this end, we are already working on formalizing the semantics of 
arbitrary control flow graphs, such that we can verify the translation of our control flow constructs to CFGs.

{\footnotesize
\paragraph{\footnotesize Acknowledgement}
We thank Maximilian P.\ L.\ Haslbeck and Simon Wimmer for proofreading and useful suggestions.
We received funding from DFG grant LA 3292/1 "Verifizierte Model Checker".
}

\clearpage

% 
% 
% Discussion:
%   assumption on infinite supply of memory assumption, deterministic semantics
% 
% Future work:
%   handle stack overflow
%   systematically test semantics: execute in Isabelle and LLVM, compare results
%   start to develop data structures, port more algorithms to use LLVM as backend.
% 
% Mention used 3rd party libraries:
%   Word, Sep-Algebra, Eisbach
% 
% Related work
%   Vellvm, Compcert and its verification infrastructure
%

% {\footnotesize
% 
% \paragraph{\footnotesize Acknowledgement}
% Max Haslbeck, Anders Schlichtkrull, and Mark Summerfield suggested textual
% improvements.
% %
% The work has received funding from the European Research Council
% under the European Union's Horizon 2020 research and innovation program
% (grant agreement No.\ 713999, Matryoshka).
% 
% }

\bibliographystyle{splncs03}
\bibliography{bib}

\end{document}

### 2019-01-10 make_artifact.sh

@@ -1,0 +1,13 @@
+#!/bin/sh
+
+all_of () {
+  find "./$1" -path "*/obsolete" -prune -o -not -name "*~" -and -not -path "*/shallow/code2/*" \
+    -not -name "lorem.txt" -not -name "kmp" \
+    -type f -print
+}
+
+
+( all_of lib; all_of shallow; all_of paper/origKMPtest; echo ./README.md ) | sort > DIST_FILES
+
+tar -czf isabelle_llvm.tgz -T DIST_FILES --transform 's|^.|isabelle_llvm|'
+

### 2019-01-10 mkhtml.sh

@@ -1,0 +1,61 @@
+#!/bin/bash
+
+BASEDIR=$PWD
+
+REBUILD=false
+UPLOAD=false
+
+while [[ $# -gt 0 ]]
+do
+key="$1"
+
+case $key in
+    -r|--rebuild)
+    REBUILD=true
+    ;;
+    -u|--upload)
+    UPLOAD=true
+    ;;
+    *)
+    echo "Unknown command line argument $1"        # unknown option
+    exit 1
+    ;;
+esac
+shift # past argument or value
+done
+
+
+
+if $REBUILD; then
+./make_artifact.sh
+fi
+
+
+rm -rf html
+mkdir -p html
+
+
+cp README.md html/
+cp isabelle_llvm.tgz html/
+cp LICENSE html/
+
+
+pandoc -V pagetitle="Isabelle LLVM" -s README.md > html/index.html
+
+
+if $UPLOAD; then
+  LOCAL_DEST=~/devel/www21-lammich/isabelle_llvm
+  rm -rf $LOCAL_DEST
+  cp -a html $LOCAL_DEST
+  cd $LOCAL_DEST
+  echo ADD
+  hg add .
+  echo COMMIT
+  hg commit -m "Automatic update of Isabelle-LLVM" .
+  echo PUSH
+  hg push
+  echo DONE
+  cd $BASEDIR
+fi
+
+

### 2019-01-10 paper/make.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+latexmk --pdf main
+rm -f *.cpt

### 2019-01-10 shallow/code2/arraycpy_cli.c

@@ -1,0 +1,36 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+
+
+void arraycpy(char *dst, char *src, int64_t n);
+
+void arraycp2(char *dst, char *src, int64_t n) {
+  int64_t i=0;
+  
+  while (i<n) {
+    dst[i] = src[i];
+    ++i;
+  }
+}
+
+
+int main (int argc, char** argv) {
+  
+  int64_t size = 1000 * 1024*1024;
+  
+  char *x = calloc(1, size);
+  char *y = calloc(1, size);
+  
+  memset(y,27,size);
+  arraycpy(x,y,size);
+  
+  int j=0;
+  for (int i=0;i<size;++i) j+=x[i];
+  
+  return j;
+}
+
+
+

### 2019-01-10 shallow/code2/cr_big_al_cli.c

@@ -1,0 +1,27 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+
+
+int64_t cr_big_al(int64_t);
+
+
+int32_t euclid2(int32_t a, int32_t b) {
+  while (a!=b) {
+    if (a<=b) b=b-a; else a=a-b;
+  }
+  return a;
+}
+
+
+int main (int argc, char** argv) {
+  
+  if (argc != 2) {printf("Expecting one argument\n"); exit(1);}
+  
+  int64_t n = strtoul (argv[1],NULL,10);
+  
+  printf ("sum %ld = %ld\n", n, cr_big_al(n));
+}
+
+
+

### 2019-01-10 shallow/code2/cr_big_vector.cpp

@@ -1,0 +1,24 @@
+#include <vector>
+#include <iostream>   // std::cout
+#include <string>     // std::string, std::stoi
+
+
+int main(int argc, char** argv) {
+  int64_t n = std::stol(argv[1]);
+  
+  auto v = std::vector<int64_t>();
+  
+  for (int64_t i=n;i>0;--i) v.push_back(i);
+  
+  int64_t s=0;
+  
+  while (n>0) {
+    --n;
+    s+=v[n];
+  }
+  
+  std::cout<<"result = "<<s<<std::endl;
+  return 0;
+}  
+
+

### 2019-01-10 shallow/code2/euclid_cli.c

@@ -1,0 +1,28 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+
+
+int32_t euclid(int32_t, int32_t);
+
+
+int32_t euclid2(int32_t a, int32_t b) {
+  while (a!=b) {
+    if (a<=b) b=b-a; else a=a-b;
+  }
+  return a;
+}
+
+
+int main (int argc, char** argv) {
+  
+  if (argc != 3) {printf("Expecting two arguments\n"); exit(1);}
+  
+  int32_t a = strtoul (argv[1],NULL,10);
+  int32_t b = strtoul (argv[2],NULL,10);
+  
+  printf ("gcd(%d,%d) = %d\n", a, b, euclid(a,b));
+}
+
+
+

### 2019-01-10 shallow/code2/qs_quicksort_cli.c

@@ -1,0 +1,32 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+
+
+int64_t qs_quicksort(int64_t*, int64_t, int64_t);
+
+
+int main (int argc, char** argv) {
+  
+  if (argc != 2) {printf("Expecting one argument\n"); exit(1);}
+
+  int64_t n = strtoul(argv[1],NULL,10);
+
+  int64_t *A = malloc (n * sizeof(int64_t));
+  
+  srand(192);
+  
+  for (int64_t i=0; i<n; ++i) A[i] = rand() % (2*n);
+
+  for (int64_t i=0; i<n; ++i) printf ("%ld ",A[i]);
+  printf("\n");
+    
+  qs_quicksort(A,0,n-1);
+  
+  for (int64_t i=0; i<n; ++i) printf ("%ld ",A[i]);
+  printf("\n");
+  
+}
+
+
+

### 2019-01-10 shallow/code2/string_cli.c

@@ -1,0 +1,20 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+
+
+int64_t my_strlen(char *);
+char streq(char *, char *);
+
+
+int main (int argc, char** argv) {
+  
+  if (argc != 3) {printf("Expecting two arguments\n"); exit(1);}
+
+  if (streq(argv[1],argv[2])) printf("Equal\n");
+  else if (my_strlen(argv[1]) == my_strlen(argv[2])) printf("Not equal (but eq len) %ld\n", my_strlen(argv[1]));
+  else printf("Not equal, len not equal %ld != %ld\n",my_strlen(argv[1]),my_strlen(argv[2]));
+}
+
+
+

### 2019-01-10 shallow/sepref/Examples/kmpcli.c

@@ -1,0 +1,79 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+
+
+typedef struct {
+  int64_t len;
+  char *str;
+} string;
+
+int64_t kmp(string,string);
+
+
+string string_of_str(char *s) { string r = {strlen(s),s}; return r;}
+
+void error(char * msg) {
+  printf("Error: %s\n",msg);
+  exit(1);
+}
+
+
+string string_of_file(char *fname) {
+  // code snippet found on fundza.com
+  /* declare a file pointer */
+  FILE    *infile;
+  char    *buffer;
+  int64_t numbytes;
+  
+  /* open an existing file for reading */
+  infile = fopen(fname, "r");
+  
+  /* quit if the file does not exist */
+  if(infile == NULL) error("No such file");
+  
+  /* Get the number of bytes */
+  fseek(infile, 0L, SEEK_END);
+  numbytes = ftell(infile);
+  
+  /* reset the file position indicator to 
+  the beginning of the file */
+  fseek(infile, 0L, SEEK_SET);	
+  
+  /* grab sufficient memory for the 
+  buffer to hold the text */
+  buffer = (char*)calloc(numbytes, sizeof(char));	
+  
+  /* memory error */
+  if(buffer == NULL) error("Out of memory");
+  
+  /* copy all the text into the buffer */
+  fread(buffer, sizeof(char), numbytes, infile);
+  fclose(infile);
+  
+  string res = {numbytes,buffer};
+  return res;
+}
+
+void free_string(string s) { free (s.str); }
+
+
+
+int main (int argc, char** argv) {
+  
+  if (argc != 4) {printf("usage: kmp <num-iterations> <search-string> <textfile>\n"); exit(1);}
+  
+  int n = strtol(argv[1],NULL,10);
+  string s1 = string_of_str(argv[2]);
+  string s2 = string_of_file(argv[3]);
+  
+  for (int i=1;i<n;++i) kmp(s1,s2);
+  int64_t res = kmp(s1,s2);
+  
+  printf("result (%d iterations) = %ld\n", n, res);
+  
+  free_string(s2);
+  
+  return 0;
+}

### 2019-01-10 shallow/sepref/Examples/lib_isabelle_llvm.c

@@ -1,0 +1,23 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+void isabelle_llvm_abort() {
+  printf("%s\n","Isabelle-LLVM (abort): Dynamic check failed.");
+  abort();
+}
+
+void isabelle_llvm_abort_msg(char *msg) {
+  printf("Isabelle-LLVM (abort): %s\n",msg);
+  abort();
+}
+
+char* isabelle_llvm_calloc(size_t n, size_t m) {
+  char *res = calloc(n,m);
+  if (!res) isabelle_llvm_abort_msg("Out of memory");
+  return res;
+}
+
+void isabelle_llvm_free(char *p) {
+  free(p);
+}
+

### 2019-06-07 paper/lipics_ex.tex
Major restructuring, added thys-2019

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"

\q{\is{l}}
%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Dummy title} %TODO Please add

\titlerunning{Dummy short title}%optional, please use if title is longer than one line

\author{John Q. Public}{Dummy University}{}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional

% \author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{J.\,Q. Public and J.\,R. Public}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{John Q. Public and Joan R. Public}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{General and reference~General literature}
\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword}%TODO mandatory; please add comma-separated list of keywords

\category{}%optional, e.g. invited paper

\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. 
\end{abstract}

\section{Typesetting instructions -- Summary}
\label{sec:typesetting-summary}

LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl. 
In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others. The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.

Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume. 
\paragraph*{Minimum requirements}

\begin{itemize}
\item Use pdflatex and an up-to-date \LaTeX{} system.
\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
\item Use a spellchecker to correct typos.
\end{itemize}

\paragraph*{Mandatory metadata macros}
Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
\begin{itemize}
\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
\item \verb+\keywords+ (a comma-separated list of keywords).
\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
\item \verb+\begin{abstract}...\end{abstract}+ .
\end{itemize}

\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
Generally speaking, please do not override the \texttt{lipics-v2019}-style defaults. To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2019} defaults}. In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
\begin{itemize}
\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
\item Do not alter the spacing of the \texttt{lipics-v2019.cls} style file.
\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
 \verb+\begin{enumerate}[(a)]+ or the like.)
\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions. 
\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
\item Do not change the bibliography style. In particular, do not use author-year citations. (The
\texttt{natbib} package is not supported.)
\end{itemize}

\enlargethispage{\baselineskip}

This is only a summary containing the most relevant details. Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2019} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
\href{http://drops.dagstuhl.de/styles/lipics-v2019/lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2019/\newline lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}}

\section{Lorem ipsum dolor sit amet}

Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. Donec eget odio et magna ullamcorper vehicula ut vitae libero. Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis. Pellentesque id ante erat. In hac habitasse platea dictumst. Curabitur a scelerisque odio. Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam. Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.

\begin{lemma}[Lorem ipsum]
\label{lemma:lorem}
Vestibulum sodales dolor et dui cursus iaculis. Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper. Proin facilisis gravida rutrum. Etiam sed sollicitudin lorem. Proin pellentesque risus at elit hendrerit pharetra. Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
\end{lemma}

\begin{proof}
Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.

\begin{claim}
content...
\end{claim}
\begin{claimproof}
content...
\end{claimproof}

\end{proof}

\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
\label{lemma:curabitur}
Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
\end{corollary}

\begin{proposition}\label{prop1}
This is a proposition
\end{proposition}

\autoref{prop1} and \cref{prop1} \ldots

\subsection{Curabitur dictum felis id sapien}

Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat. Curabitur sed velit diam. Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu. Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet. Duis rutrum dapibus ornare. Nulla vehicula vulputate iaculis. Proin a consequat neque. Donec ut rutrum urna. Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Aenean nec faucibus leo. Cras ut nisl odio, non tincidunt lorem. Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat. Fusce risus libero, convallis at fermentum in, dignissim sed sem. Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}. Donec non suscipit lorem. Nam sit amet enim vitae nisl accumsan pretium.

\begin{lstlisting}[caption={Useless code},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
for i:=maxint to 0 do 
begin 
    j:=square(root(i));
end;
\end{lstlisting}

\subsection{Proin ac fermentum augue}

Proin ac fermentum augue. Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis. Nulla facilisi. Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac. Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna. Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui. Praesent a pretium quam. Mauris non mauris sit amet eros rutrum aliquam id ut sapien. Nulla aliquet fringilla sagittis. Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor. Nulla cursus aliquet enim. Cras sapien risus, accumsan eu cursus ut, commodo vel velit. Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel. Integer faucibus faucibus felis.

\begin{itemize}
\item Ut vitae diam augue. 
\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem. 
\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit. 
\end{itemize}

\begin{remark}
content...
\end{remark}

\section{Pellentesque quis tortor}

Nec urna malesuada sollicitudin. Nulla facilisi. Vivamus aliquam tempus ligula eget ornare. Praesent eget magna ut turpis mattis cursus. Aliquam vel condimentum orci. Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi. Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.

\begin{lemma} [Quisque blandit tempus nunc]
Sed interdum nisl pretium non. Mauris sodales consequat risus vel consectetur. Aliquam erat volutpat. Nunc sed sapien ligula. Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus. Nunc vestibulum nunc ac massa pretium pharetra. Nulla facilisis turpis id augue venenatis blandit. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
\end{lemma}

Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.

\appendix
\section{Morbi eros magna}

Morbi eros magna, vestibulum non posuere non, porta eu quam. Maecenas vitae orci risus, eget imperdiet mauris. Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis. Sed condimentum convallis dolor, a dignissim est ultrices eu. Donec consectetur volutpat eros, et ornare dui ultricies id. Vivamus eu augue eget dolor euismod ultrices et sit amet nisi. Vivamus malesuada leo ac leo ullamcorper tempor. Donec justo mi, tempor vitae aliquet non, faucibus eu lacus. Donec dictum gravida neque, non porta turpis imperdiet eget. Curabitur quis euismod ligula.

%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{lipics-v2019-sample-article}

\appendix

\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

List of different predefined enumeration styles:

\begin{itemize}
\item \verb|\begin{itemize}...\end{itemize}|
\item \dots
\item \dots
%\item \dots
\end{itemize}

\begin{enumerate}
\item \verb|\begin{enumerate}...\end{enumerate}|
\item \dots
\item \dots
%\item \dots
\end{enumerate}

\begin{alphaenumerate}
\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
\item \dots
\item \dots
%\item \dots
\end{alphaenumerate}

\begin{romanenumerate}
\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
\item \dots
\item \dots
%\item \dots
\end{romanenumerate}

\begin{bracketenumerate}
\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
\item \dots
\item \dots
%\item \dots
\end{bracketenumerate}

\begin{description}
\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
\item[Description 2] Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\item[Description 3]  \dots
%\item \dots
\end{description}

\section{Theorem-like environments}\label{sec:theorem-environments}

List of different predefined enumeration styles:

\begin{theorem}\label{testenv-theorem}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{theorem}

\begin{lemma}\label{testenv-lemma}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{lemma}

\begin{corollary}\label{testenv-corollary}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{corollary}

\begin{proposition}\label{testenv-proposition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proposition}

\begin{exercise}\label{testenv-exercise}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{exercise}

\begin{definition}\label{testenv-definition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{definition}

\begin{example}\label{testenv-example}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{example}

\begin{note}\label{testenv-note}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note}

\begin{note*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note*}

\begin{remark}\label{testenv-remark}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark}

\begin{remark*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark*}

\begin{claim}\label{testenv-claim}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim}

\begin{claim*}\label{testenv-claim2}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim*}

\begin{proof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proof}

\begin{claimproof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claimproof}

\end{document}

### 2019-06-07 paper/main.tex
Major restructuring, added thys-2019

@@
-\documentclass[a4paper,oribibl,envcountsame]{llncs}
-\pagestyle{plain}
-\usepackage[T1]{fontenc}
-\usepackage[misc]{ifsym}
+\documentclass[a4paper,USenglish,cleveref, autoref]{lipics-v2019}
+% \documentclass[a4paper,oribibl,envcountsame]{llncs}
+% \pagestyle{plain}
+% \usepackage[T1]{fontenc}
+% \usepackage[misc]{ifsym}
  \usepackage{amsmath}
  \usepackage{amssymb}
-\usepackage{booktabs}
-\usepackage{bussproofs}
-\usepackage{cite}
-\usepackage{color}
-\usepackage{enumitem}
+% \usepackage{booktabs}
+% \usepackage{bussproofs}
+% \usepackage{cite}
+% \usepackage{color}
+% \usepackage{enumitem}
 %\let\vec=\undefined
@@
  \usepackage{stmaryrd}
-\usepackage{subfig}
-\usepackage{textcomp}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{mathtools}
-\usepackage{todonotes}
-\usepackage{enumitem}
+% \usepackage{subfig}
+% \usepackage{textcomp}
+% \usepackage{url}
+% \usepackage{xspace}
+% \usepackage{mathtools}
+% \usepackage{todonotes}
+% \usepackage{enumitem}
 
@@
 
-\usepackage{array}
-\newcolumntype{x}[1]{>{\centering\arraybackslash\hspace{0pt}}b{#1}}
-\usepackage{diagbox}
-\usepackage{siunitx}
+\usepackage{relsize}
+
+% \usepackage{array}
+% \newcolumntype{x}[1]{>{\centering\arraybackslash\hspace{0pt}}b{#1}}
+% \usepackage{diagbox}
+% \usepackage{siunitx}
 
@@
 
-\usepackage[final,tracking=true,kerning=true,spacing=true,stretch=10,shrink=10]{microtype}
+% \usepackage[final,tracking=true,kerning=true,spacing=true,stretch=10,shrink=10]{microtype}
 
 %% BEGIN Times font
-\usepackage{mathptmx}
-\usepackage[scaled=.82]{beramono}
-\usepackage[scaled=.86]{helvet}
-\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
+% \usepackage{mathptmx}
+% \usepackage[scaled=.82]{beramono}
+% \usepackage[scaled=.86]{helvet}
+% \DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
 %% END Times font
@@
 
-\usepackage[
-   a4paper,
-   pdftex,
-   pdftitle={Isabelle LLVM},
-   pdfauthor={Peter Lammich},
-   pdfkeywords={},
-   pdfborder={0 0 0},
-   draft=false,
-   bookmarksnumbered,
-   bookmarks,
-   bookmarksdepth=2,
-   bookmarksopenlevel=2,
-   bookmarksopen]{hyperref}
-
-\hyphenation{Isa-belle Nieuw-en-huis meta-logic multi-set multi-sets}
+% \usepackage[
+%    a4paper,
+%    pdftex,
+%    pdftitle={Isabelle LLVM},
+%    pdfauthor={Peter Lammich},
+%    pdfkeywords={},
+%    pdfborder={0 0 0},
+%    draft=false,
+%    bookmarksnumbered,
+%    bookmarks,
+%    bookmarksdepth=2,
+%    bookmarksopenlevel=2,
+%    bookmarksopen]{hyperref}
+
+% \hyphenation{Isa-belle Nieuw-en-huis meta-logic multi-set multi-sets}
+
+\bibliographystyle{plainurl}% the mandatory bibstyle
 
@@
 
+\ccsdesc[500]{Theory of computation~Program verification}
+\ccsdesc[300]{Theory of computation~Logic and verification}
+\ccsdesc[300]{Theory of computation~Separation logic}
+% \ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\keywords{Isabelle/HOL,LLVM,Separation Logic,Verification Condition Generator,Code Generation}%TODO mandatory; please add comma-separated list of keywords
+
 \begin{document}
 
-\title{Isabelle LLVM}
-\titlerunning{Isabelle LLVM}
-
-\author{Peter Lammich}
+\title{Generating Verified LLVM from Isabelle/HOL}
+% \titlerunning{Isabelle LLVM}
+
+\author{Peter Lammich}{The University of Manchester}{peter.lammich@manchester.ac.uk}{}{}
 \authorrunning {P. Lammich}
-\institute{
-  Technische Universit\"at M\"unchen, Munich, Germany \\
-  \email{lammich@in.tum.de}
-}
+
+% \author{Peter Lammich}
+% \authorrunning {P. Lammich}
+% \institute{
+%   The University of Manchester, England
+%   \email{peter.lammich@manchester.ac.uk}
+% %   Technische Universit\"at M\"unchen, Munich, Germany \\
+% %   \email{lammich@in.tum.de}
+% }
 
@@
 \begin{abstract}
-We formalize the semantics of a fragment of the LLVM intermediate representation in Isabelle/HOL.
-On top of this semantics, we have developed
-\begin{enumerate}
-\item a code generator, to translate from Isabelle/HOL to actual LLVM text; 
-\item a verification condition generator, to reason about LLVM programs;
-\item a backend for the Isabelle Refinement Framework, to enable a high-level refinement based approach to developing verified LLVM programs.
-\end{enumerate}
-As an example, we adopt an existing formalization of the Knuth-Morris-Pratt string search algorithm to our new framework.
-With only slight changes to the original formalization, we obtain a verified LLVM implementation, which runs faster and uses less memory 
-than the originally verified Standard ML implementation.
-
-The trusted code base of our approach is the formalization of the semantics, which we have engineered for conciseness and readability, 
-and the code generator that translates from Isabelle to LLVM text, which is merely a pretty printer. 
-The other tools that we developed cannot compromise the soundness of the approach, as they only generate theorems that are proved via Isabelle's kernel.
+We present a framework to generate verified LLVM programs from Isabelle/HOL.
+It consists of a code generator that generates LLVM text from 
+a simplified fragment of LLVM, shallowly embedded into Isabelle/HOL.
+On top, we have developed a separation logic, a verification condition 
+generator, and an LLVM backend to the Isabelle Refinement Framework.
+
+As case studies, we have produced verified LLVM implementations of binary search and the 
+Knuth-Morris-Pratt string search algorithm. These are one order of magnitude faster than 
+the Standard-ML implementations produced with the original Refinement Framework, and on par 
+with unverified C implementations. Adoption of the original correctness proofs to the new LLVM backend was straightforward.
+
+The trusted code base of our approach is the shallow embedding of the LLVM fragment 
+and the code generator, which is a pretty printer combined with some straightforward compilation steps.
+
+% We formalize the semantics of Isabelle-LLVM, an imperative language shallowly 
+% embedded into Isabelle/HOL, designed to be easily translated to actual LLVM text.
+% On top of Isabelle-LLVM, we have developed
+% \begin{enumerate}
+% \item a code generator, to translate from Isabelle-LLVM to actual LLVM text; 
+% \item a verification condition generator, to reason about Isabelle-LLVM programs;
+% \item a backend for the Isabelle Refinement Framework, to enable a high-level refinement based approach to developing verified LLVM programs.
+% \end{enumerate}
+% 
+% As an example, we adopt an existing formalization of the Knuth-Morris-Pratt string search algorithm to our new framework.
+% With only slight changes to the original formalization, we obtain a verified LLVM implementation, which runs faster and uses less memory 
+% than the originally verified Standard ML implementation.
+% 
+% The trusted code base of our approach is the Isabelle-LLVM semantics, which we have engineered for conciseness and readability, 
+% and the code generator that translates from Isabelle to LLVM text, which is merely a pretty printer. 
+% The other tools that we developed cannot compromise the soundness of the approach, as they only generate theorems that are proved via Isabelle's kernel.
 \end{abstract}
 
-\section{Introduction}
-\label{sec:introduction}
-The Isabelle Refinement Framework~\cite{LaTu12,La13} features a stepwise refinement approach to verified algorithms, using the Isabelle/HOL theorem prover~\cite{NPW02,NiKl14}.
+\section{Introduction}\label{sec:introduction}
+
+% Current code generator technology in Isabelle and other interactive theorem provers is limited to emit high-level functional target 
+% languages, which are inherently less efficient than low-level imperative languages such as C or LLVM~\cite{LLVM-manual}.
+% This work presents a framework to generate verified LLVM programs from Isabelle/HOL~\cite{NPW02,NiKl14}.
+% 
+% We developed the Isabelle-LLVM language, an imperative language shallowly embedded into Isabelle/HOL.
+% Isabelle-LLVM is designed to be easily translated into actual LLVM text, which is done by our code generator.
+% On top of Isabelle/LLVM we build a separation logic and a verification condition generator, which enables us to verify
+% basic data structures and algorithms. To simplify verification of more complex algorithms, we provide a stepwise refinement 
+% approach, based on the Isabelle Refinement Framework~\cite{LaTu12,La13}.
+
+The Isabelle Refinement Framework~\cite{LaTu12,La13,La15} features a stepwise refinement approach to verified algorithms, using the Isabelle/HOL theorem prover~\cite{NPW02,NiKl14}.
 It has been successfully applied to verify many algorithms and software systems, 
-among them LTL and timed automata model checkers\cite{ELNN13,BrLa18,WiLa18}, network flow algorithms~\cite{LaSe17,LaSe16}, a verified SAT-solver certification tool~\cite{La17_CADE,La17_SAT}, and even a verified SAT solver~\cite{FBL18}.
+among them LTL and timed automata model checkers~\cite{ELNN13,BrLa18,WiLa18}, network flow algorithms~\cite{LaSe17,LaSe16}, a SAT-solver certification tool~\cite{La17_CADE,La17_SAT}, and even a SAT solver~\cite{FBL18}.
 Using Isabelle/HOL's code generator~\cite{HKKN13}, the verified algorithms can be extracted to functional languages like Haskell or Standard ML.
 However, the code generator only provides partial correctness guarantees, i.e., termination of the generated code cannot be proved.
-Moreover, the generated code is typically much slower than the same algorithms implemented in C or Java.
-
-The original Refinement Framework could only generate purely functional code. 
-The first remedy to the performance problem was to introduce array data structures that behave like a functional 
-list on the surface, but are implemented by a destructively updated array behind the scenes, similar to Haskell's now deprecated DiffArray.
-While this gained some performance, the array-based implementation itself was not verified, such that we had to trust its correctness. 
+Moreover, the generated code is typically slower than the same algorithms implemented in C or Java.
+
+The original Refinement Framework~\cite{LaTu12,La13} could only generate purely functional code. 
+The first remedy to the performance problem was to introduce array data structures that behave like functional 
+lists on the surface, but are implemented by destructively updated arrays behind the scenes, similar to Haskell's now deprecated DiffArray.
+While this gained some performance, the array implementation itself was not verified, such that we had to trust its correctness. 
 Moreover, an array access still required a significant amount of overhead compared to a simple pointer dereference in C.
 
-The next step was the Sepref tool~\cite{La15}. It generates code for Imperative HOL~\cite{BKHEM08}, which provides a heap monad inside Isabelle/HOL, 
+The next step towards more efficient verified implementations was the Sepref tool~\cite{La15}. 
+It generates code for Imperative HOL~\cite{BKHEM08}, which provides a heap monad inside Isabelle/HOL, 
 and a code generator extension to generate code that uses the stateful arrays provided by ML, or the heap monad of Haskell.
@@
 %
-Using Imperative/HOL as backend, we gained some additional performance: For example, the GRAT tool~\cite{La17_SAT} provides a 
+Using Imperative HOL as backend, we gained some additional performance: For example, the GRAT tool~\cite{La17_CADE,La17_SAT} provides a 
 verified checker for UNSAT certificates in the DRAT format~\cite{WHH14}. It is faster than the unverified state-of-the art 
@@
 
-This paper presents a next step towards efficient verified software: A refinement framework to generate verified code in LLVM intermediate representation~\cite{LLVM-manual} 
+This paper presents a next step towards efficient verified algorithms: A refinement framework to generate verified code in LLVM intermediate representation~\cite{LLVM-manual} 
 with total correctness guarantees. LLVM is an imperative intermediate language with a powerful and well-tested optimizing compiler.
-We first formalize the semantics of a fragment\footnote{We support integer instructions, pointers, a simplified memory model, and simplified control flow.
-We expect this fragment to be sufficient for most algorithms that have already been verified with the refinement framework.} of LLVM in Isabelle/HOL, and provide a code generator to actual LLVM text (\S\ref{sec:semantics}).
-On top of this semantics, we build a separation logic and a verification condition generator, which allows convenient reasoning about LLVM programs (\S\ref{sec:vcg}).
-Finally, we modify the Sepref tool to target our LLVM semantics instead of Imperative/HOL (\S\ref{sec:auto_ref}). 
-This way, we connect the Refinement Framework to our LLVM code generator. This only affects the last refinement step, 
-such that most parts of existing verifications can be reused. As a case study, we apply only minor modifications to an existing formalization~\cite{HeLa17} of
-the Knuth-Morris-Pratt string search algorithm~\cite{KMP77}, and obtain a verified 
-LLVM implementation that is an order of magnitude faster and uses less memory than the originally verified implementation in Standard ML.
+We first formalize the semantics of Isabelle-LLVM, a simple imperative language shallowly embedded into Isabelle/HOL, 
+and designed to be easily translated to actual LLVM text (\S\ref{sec:semantics}). 
+On top of Isabelle-LLVM, we build a separation logic and a verification condition generator, which allows convenient reasoning about Isabelle-LLVM programs (\S\ref{sec:vcg}).
+Finally, we modify the Sepref tool to target Isabelle-LLVM instead of Imperative/HOL (\S\ref{sec:auto_ref}), connecting 
+the Refinement Framework to our LLVM code generator. This only affects the last refinement step, 
+such that most parts of existing verifications can be reused. 
+As case studies (\S\ref{sec:casestudies}), we verify a binary search algorithm and adopt an existing formalization~\cite{HeLa17} of
+the Knuth-Morris-Pratt string search algorithm~\cite{KMP77}. The resulting LLVM code is significantly
+faster than the corresponding Standard-ML code and on par with unverified C implementations.
+%
 The Isabelle theories described in this paper are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm}.
 
-\section{LLVM Semantics}\label{sec:semantics}
+\section{Isabelle-LLVM}\label{sec:semantics}
 \subsection{State Monad}
-The basis of our LLVM semantics is a state-error monad, which we use to conveniently model the preconditions of instructions, their effect 
-on memory, as well as arbitrary recursive programs. We define the algebraic data types
+The basis of Isabelle-LLVM is a state-error monad, which we use to conveniently model the preconditions of instructions, their effect 
+on memory, as well as arbitrary recursive programs. We define the algebraic data types:
 \begin{lstlisting}
@@
 \begin{lstlisting}
-  "return x  = M (%s. SUCC x s)"                                           "get   = M (%s. SUCC s s)"
-  "fail         = M (%_. FAIL)"                                                  "set s = M (%_. SUCC () s)"
+  "return x  = M (%s. SUCC x s)"                                get  = M (%s. SUCC s s)
+  "fail       = M (%_. FAIL)"                                     set s = M (%_. SUCC () s)
   "bind m f  = M (%s. case run m s of SUCC x s => run (f x) s | r => r)"
@@
 We use do-notation, i.e.\ \q{\is|do { x<-m; f x }|} is short for \q{\is{bind m (\<lambda>x. f x)}}.
-Moreover, we define a flat CCPO\cite{Mark76} on \q{\is{mres}}, with \q{\is{\<bot> := NTERM}}.
-We define \q{\is{REC F}} to be the least fixed point of a monotonic function \q{\is{F :: ('a => ('b,'s) M) => 'a => ('b,'s) M}}.
+Moreover, we define a flat chain complete partial order~\cite{Mark76} on \q{\is{mres}}, with \q{\is{\<bot> := NTERM}}.
+For a monotonic function \q{\is{F :: ('a => ('b,'s) M) => 'a => ('b,'s) M}}, \q{\is{REC F}} is the least fixed point.
 As functions defined using the monad combinators are monotonic by construction~\cite{Kr10}, 
@@
 \subsection{Memory Model}
-We use a high-level memory model that does not directly expose the bit-level representation of values, and assumes an infinite supply of memory. 
+We use a high-level memory model that does not directly expose the bit-level representation of values and assumes an infinite supply of memory. 
 The memory is modeled as a list of blocks. Each block is either deallocated, or it is a list of values.
-A value is a pair of values, a pointer, or an integer. We model memory by the following data types\footnote{We have simplified the presentation a bit. 
-The actual implementation defines the concepts memory, block, and value independently, in order to ease future extensions, to, e.g., arbitrary structs instead of pairs.
+A value is a pair of values, a pointer, or an integer. We model memory by the following data types\footnote{We have slightly simplified the presentation. 
+The actual implementation defines the concepts memory, block, and value in a modular fashion, in order to ease future extensions.
 % . The memory is parameterized over the block type, 
@@
 \begin{lstlisting}
-memory = MEMORY (block option list)             block     = val list 
+memory = MEMORY (block list)                 block     = val list option
 val = PAIR val val | PRIM primval              primval = PV_INT lint | PV_PTR rptr
@@
 %
-Here, the type \q{\is{lint}} is a fixed bit width integer type with a two's complement semantics, as used by LLVM. 
-The type \q{\is{rptr}} is either null, or an address. An address is a path through the memory structure to a value:
+Here, the type \q{\is{lint}} is a fixed bit width word type with a two's complement semantics, as used by LLVM. 
+The type \q{\is{rptr}} is either null or an address. An address is a path through the memory structure to a value:
 \begin{lstlisting}
@@
 or fail if the pointer is invalid.
-For the actual store function, we check that the structure of the value does not change, i.e. pairs remain pairs, pointers remain pointers, and integers of width $w$ remain integers of width $w$.
+For the actual store function, we check that the structure of the value does not change, i.e. pairs remain pairs, pointers remain pointers, and words of width $w$ remain words of width $w$:
 \begin{lstlisting}
   store x p = do { y <- load p; assert (vstruct x = vstruct y); put x p }
-\end{lstlisting}
+  $\text{where}$
+  vstruct (PAIR a b) = VS_PAIR (vstruct a) (vstruct b)
+  vstruct (PRIM (PV_PTR _)) = VS_PTR
+  vstruct (PRIM (PV_INT w)) = VS_INT (width w) 
+\end{lstlisting}
+
 %
@@
 %
-Similarly, we define an allocate and a free function:\\[-2ex]
+Similarly, we define an allocate and a free function:\\
 \begin{minipage}[t]{.46\textwidth}
@@
 While we explicitly model values in memory by the type \q{\is{val}}, we model values in registers in a more shallow fashion:
-We will identify LLVM registers with Isabelle variables that have a type of the shape \q{\is{T = T \x T | n word | T ptr}}. 
+We identify LLVM registers with Isabelle variables that have a type of the shape \q{\is{T = T \x T | n word | T ptr}}. 
 Here, \q{\is{\x}} is Isabelle's product type, \q{\is{n word}} is the $n$ bit word type from Isabelle's word 
@@
 
-\subsection{LLVM Instructions}
-In a next step, we define the LLVM instructions. Each LLVM instruction is identified with an Isabelle function. For example, the load instruction
+\subsection{Instructions}
+In a next step, we define the instructions of Isabelle-LLVM. Each instruction is identified with an Isabelle constant. For example, the load instruction
 is modeled by:
@@
 Instead, we assume that the generated code will be linked with the C standard library, 
-and produce calls to \q{\is{calloc}} and \q{\is{free}}.
-
-We also define
-instructions to extract/insert the first/second element of a pair, to offset a pointer, and to advance a pointer to the first/second element.
+and let the code generator produce calls to \q{\is{calloc}} and \q{\is{free}}.
+%
+We also define instructions to access the elements of a pair, to offset a pointer, and to advance a pointer into a pair.
 The code generator maps these instructions to the corresponding LLVM instructions \q{\is{getelementptr}}, \q{\is{insertvalue}}, and \q{\is{extractvalue}}.
 
-The non-memory-related instructions are defined directly on the shallow embedding. For example, we define:
+Integer instructions are defined on the \q{\is{n word}} type. For example, we define:
 \begin{lstlisting}
   ll_udiv :: "n word => n word => n word"
-  ll_udiv a b = do { assert (a ~= 0); return (a div b) }
-\end{lstlisting}
-where \q{\is{div}} is the unsigned division from the word library.
-Note that we use assertions to model undefined behavior in LLVM by failure in our semantics.
-Similarly, we define the other integer instructions.
+  ll_udiv a b = do { assert (b ~= 0); return (a div b) }
+\end{lstlisting}
+where \q{\is{div}} is the unsigned division from Isabelle's word library.
+Note the use of assertions to exclude undefined behavior, e.g., division by zero.
 
 \subsection{Modeling Control Flow}\label{sec:modeling_ctrl_flow}
-Next, we put together instructions to form procedure bodies. Instead of arbitrary control flow graphs, 
-we only allow structured control flow via if-then-else, procedure calls, and sequential composition: The body of
+Next, we put together instructions to form procedure bodies. We only allow structured control flow 
+via if-then-else, while, procedure calls, and sequential composition: The body of
 a procedure is modeled by an Isabelle term of type \q{\is{'a llM}} and shape \q{\is{block}}, where
@@
 block = do { var <- cmd; block } | return var
-cmd = ll_<opcode> arg^* | proc_name arg^* | llc_if arg block block
+cmd = ll_<opcode> arg^* | proc_name arg^* | llc_if arg block block | llc_while block block
 arg = var | number | null | init
 \end{lstlisting}
-and \q{\is{llc_if :: 1 word => 'a llM => 'a llM => 'a llM}} is defined as
-\begin{lstlisting}
+% and \q{\is{llc_if :: 1 word => 'a llM => 'a llM => 'a llM}} is defined as
+with
+\begin{lstlisting}
+llc_if :: 1 word => 'a llM => 'a llM => 'a llM
 llc_if b t e = if b=1 then t else e
+
+llc_while :: ('a => 1 word llM) => ('a => 'a llM) => 'a => 'a llM
+llc_while b c s = do {ctd <- b s; llc_if ctd (do {s <- c s; llc_while b c s}) (return s)}
 \end{lstlisting}
 That is, a block is a list of commands whose results are bound to variables, terminated by a return instruction. 
-A command is either an instruction, a procedure call, or an \q{\is{llc_if}}. 
-The arguments of instructions and procedure calls, as well as the condition of an \q{\is{llc_if}}, must be variables or constants (numbers, the null pointer, or a zero-initialized value).
+A command is either an instruction, a procedure call, or an if-then-else or while statement.
+The arguments of instructions and procedure calls, as well as the condition of an if-then-else statement, must be variables or 
+constants (i.e., numbers, the null pointer, or a zero-initialized value). 
+The condition of a while statement is modeled as a block returning a \q{\is{1 word}}, such that it can be re-evaluated prior to each loop iteration.
 %
@@
 where the \q{\is{proc\_i}} are Isabelle functions, the \q{\is{x\_i}} are variables, and all free variables on the right hand side are among the \q{\is{x\_i}}.
-As our monad allows arbitrary recursive definitions, we can define arbitrary recursive programs.
+% As our monad allows arbitrary recursive definitions, we can define arbitrary recursive programs.
 
@@
 \centering
-\begin{minipage}[t]{.39\textwidth}
+\begin{minipage}[t]{.35\textwidth}
   \begin{lstlisting}
@@
         return c
-      })
-    }    
-  \end{lstlisting}
-  \caption{Program in Isabelle}\label{fig:fib_isabelle}
+      }) }    
+  \end{lstlisting}
+  \caption{Isabelle-LLVM program}\label{fig:fib_isabelle}
 \end{minipage}
-\hspace*{.09\textwidth}
-\begin{minipage}[t]{.5\textwidth}
+% \hspace*{.04\textwidth}
+\hfill
+\begin{minipage}[t]{.57\textwidth}
   \begin{lstlisting}[language=LLVM, literate={}]
@@
       %x1a = phi i64 [ %x, %then ], [ %c, %else ]
-      ret i64 %x1a
-  }  
+      ret i64 %x1a }  
   \end{lstlisting}
@@
   Figure~\ref{fig:fib_isabelle} shows the Isabelle specification of a procedure named \q{\is{fib}}, 
-  which takes a 64 bit integer argument, and returns a 64 bit integer. 
-  Our semantics can be directly executed inside Isabelle. The following Isabelle command evaluates our function on the first few natural numbers, and an empty memory:
+  which takes a 64 bit word argument, and returns a 64 bit word. 
+  Our semantics can be directly executed inside Isabelle. The following Isabelle command evaluates \q{\is{fib}} on the first few natural numbers, and an empty memory:
   \begin{lstlisting}
@@
 \subsection{Code Generation}
-The LLVM intermediate representation \cite{LLVM-manual} is a strongly typed control flow graph (CFG) based intermediate language 
+The LLVM intermediate representation~\cite{LLVM-manual} is a strongly typed control flow graph (CFG) based intermediate language 
 that uses single static assignment (SSA) form~\cite{CFRWZ91}.
@@
 Each non-void instruction defines a fresh register containing its result. A register can only be accessed in the part of the CFG that is dominated by its definition.
-To transfer values from registers to other parts of the CFG, $\phi$-instructions have to be used. 
+To transfer values from registers to other parts of the CFG, $\phi$-instructions are used. 
 A $\phi$-instruction must be located at the start of a basic block. 
 It lists, for each possible predecessor block, an accessible register in this predecessor block. 
-The $\phi$-instruction then evaluates to the value of the register from those predecessor block from which execution was actually transferred. 
-As for all instructions, the result of the $\phi$-instruction is bound to a fresh register, which can then be accessed from the current basic block.
-
-It is straightforward to map a program defined in Isabelle to an actual LLVM program.
+The $\phi$-instruction evaluates to the value of the register from those predecessor block from which execution was actually transferred. 
+The result of the $\phi$-instruction is bound to a fresh register, which can then be accessed from the current basic block.
+
+It is straightforward to map an Isabelle-LLVM program to an actual LLVM program.
 Each equation of the form \q{\is{proc x_1 .. x\_n = block}} is mapped to an LLVM procedure named \q{\is{proc}}. 
@@
 An \q{\is{x <- llc_if b t e}} is translated to conditional branching, using a $\phi$-instruction to define the result register \q{\is{x}} when joining the control flow.
+An \q{\is{x <- llc_while b c s}} is translated similarly.
 
@@
 \begin{example}
-Figure~\ref{fig:fib_llvm} displays the output of our code generator for the \q{\is{fib}} function displayed in Figure~\ref{fig:fib_isabelle}.
+Figure~\ref{fig:fib_llvm} displays the output of our code generator for the \q{\is{fib}} constant displayed in Figure~\ref{fig:fib_isabelle}.
 \end{example}
@@
 \subsubsection{Mapping the Memory Model}
-Some of the translations are slightly more involved.
+Mapping the abstract memory model of Isabelle-LLVM to actual LLVM is slightly more involved.
+% Some translations are slightly more involved.
 For example, recall the \q{\is{ll_malloc :: 'a itself => n word => 'a ptr llM}} instruction. It has to be mapped to the function \q{\is{void* calloc(size_t, size_t)}} from
@@
 becomes: Either the program terminates with an out-of-memory condition, or it behaves as modeled by the semantics. 
-Our current implementation simply prints an error message and terminates the process with exit code $1$ if it runs out of memory.
+Our current implementation prints an error message and terminates the process with exit code $1$ if it runs out of memory.
+
 A similar issue arises when comparing pointers: LLVM does not have instructions for pointer comparison.
 Instead, pointers have to be cast to integers, which can then be compared.
-However, this requires to know the bit-width of a pointer, which we cannot model in our semantics.
+However, this requires to know the bit-width of a pointer, which we cannot model in our semantics that admits unboundedly many different pointers.
 Instead, we model the instructions \q{\is{ll_ptrcmp_eq}} and \q{\is{ll_ptrcmp_ne}}, and let the code generator generate the cast to integers and the integer comparison.
@@
 \subsection{Preprocessing}
-In the previous sections we have described the semantics of a fragment of LLVM, shallowly embedded in Isabelle/HOL. 
-However, programs have to stick to a very restrictive shape (cf.~\S\ref{sec:modeling_ctrl_flow}), 
+In the previous sections we have described the semantics of Isabelle-LLVM and its translation to actual LLVM.
+However, Isabelle-LLVM programs have to adhere to a very restrictive shape (cf.~\S\ref{sec:modeling_ctrl_flow}), 
 which makes them easy to map to actual LLVM code, but tedious to write directly.
-In this section, we describe the preprocessing techniques that we implemented to make programs simpler to write.
-Note that these techniques work by \emph{proving} the equations in restricted shape from the user specified ones. 
-Thus, errors in the preprocessor cannot affect soundness of the generated code: Either, it fails to prove the equations, or it produces ill-shaped equations, 
+Thus, we implement a preprocessor that tries to automatically transform user-specified equations to 
+valid Isabelle-LLVM. While the preprocessing is highly incomplete, i.e., it cannot convert every equation to a well-shaped one, 
+it works well in practice, allowing for concise specifications.
+Note that the preprocessor \emph{proves} the new equations from the original ones.
+Thus, errors in the preprocessor cannot affect soundness: Either, it fails to prove the equations, or it produces ill-shaped equations, 
 which the code generator will reject.
 
-Our preprocessor includes two main functionalities, monomorphization and inlining.
-The goal is to produce well-shaped equations from a given set of Isabelle functions.
-We assume that the functions we start with are instantiated to a monomorphic type, i.e., do not contain any type variables.
-
-\subsubsection{Monomorphization}
-We first collect the defining equations of the specified functions, and, recursively, the equations of the functions used in them.
-During the collection process we perform monomorphization, i.e., 
-we instantiate polymorphic definitions to the types they are actually used with. 
-After this step, we have a set of monomorphic equations that define all functions that will occur in the final program.
-However, the right hand sides of the equations may still be ill-shaped.
+The user specifies an initial set of constants, which must be instantiated to monomorphic types, i.e., must not contain any type variables.
+For each constant, the preprocessor then gathers the defining equation, 
+instantiates it to the actual monomorphic type of the constant, transforms it by inlining and fixed point unfolding,
+and then repeats the process for any new constant occurring on the right-hand side of the transformed equation.
+Note that a constant is identified by its name and type, such that a constant with the same name can occur multiple 
+times in the final Isabelle-LLVM program. The code generator will disambiguate the names.
+At the end, we have a set of monomorphic equations that define all constants that occur in the final program, and can be passed to the actual code generator.
+We now describe the inlining and fixed point unfolding transformations.
+
+% 
+% 
+% In this section, we describe the preprocessing techniques that we implemented to make programs simpler to write.
+% The preprocessor starts with a set of user-specified constants. It then gathers the defining equations of these 
+% constants, and, recursively, of all used constants, and transforms them to the shape expected by the code generator.. 
+% Note that the preprocessor \emph{proves} the new equations from the original ones.
+% Thus, errors in the preprocessor cannot affect soundness of the generated code: Either, it fails to prove the equations, or it produces ill-shaped equations, 
+% which the code generator will reject.
+% Our preprocessor includes three main functionalities: monomorphization, inlining, and fixed-point unfolding.
+
+% \subsubsection{Monomorphization}
+% We assume that the initial constants are instantiated to monomorphic types, i.e., do not contain any type variables.
+% For each constant, the preprocessor then gathers the defining equation, 
+% instantiates it to the actual monomorphic type of the constant, performs inlining and fixed point unfolding,
+% and then repeats the process for the constants occurring on the right-hand side of the transformed equation.
+% At the end, we have a set of monomorphic equations that define all constants that occur in the final program.
+% We now describe the inlining and fixed point unfolding in more detail.
 
 \subsubsection{Inlining}
-We first apply user defined rewrite rules, and then flatten nested expressions, converting function calls to the shape 
+Inlining first applies user defined rewrite rules and then flattens nested expressions, converting function calls to the shape 
 \q{\is{r <- f x_1 ... x\_n}} or \q{\is{r <- return (f x_1 ... x\_n)}}, where the $x_i$ are either constants, variables, or \emph{monadic} arguments
@@
 We iterate the rewriting and flattening steps until a fixed point is reached.
-While this preprocessing is highly incomplete, i.e., it cannot convert every term to a well-shaped one, it works well in practice, allowing for concise specifications.
 
 \begin{example}
-Consider the following definition of the function \q{\is{fib'}}:
+Consider the following definition of the constant \q{\is{fib'}}:
 \begin{lstlisting}
@@
 
-\subsubsection{Generation of Recursive Equations}
-A third functionality of the preprocessor is to generate recursive functions from fixed-point combinators. In general,
-it examines the right hand side of an equation for patterns \q{\is{p}} for which it has an unfold rule of the form \q{\is{p = F p}}.
-It then defines a new function \q{\is{f x_1 ... x\_n = F (f x_1 ... x\_n)}}, where the \q{\is{x\_i}} are the free variables in the pattern \q{\is{p}}.
+\subsubsection{Fixed-Point Unfolding}
+The preprocessor generates recursive functions from fixed-point combinators. 
+It examines the right hand side of an equation for patterns \q{\is{p}} for which it has an unfold rule of the form \q{\is{p = F p}}.
+It then defines a new constant \q{\is{f x_1 ... x\_n = F (f x_1 ... x\_n)}}, where the \q{\is{x\_i}} are the free variables in the pattern \q{\is{p}}.
 Finally, it replaces \q{\is{p}} by \q{\is{f x_1 ... x\_n}} in the equation. This way, specifications with fixed point combinators are automatically 
@@
 
-A standard example is the while loop combinator, which we define by the equation
-\begin{lstlisting}
-  llc_while b c s = do {ctd <- b s; llc_if ctd (do {s <- c s; llc_while b c s}) (return s)}
-\end{lstlisting}
-We use this equation as unfold rule, such that while loops are automatically converted to tail calls. Note that this 
-feature allows for the concise specifications of loops, without adding additional trusted code. 
-As LLVM has tail-call optimization, the compiled programs will be efficient.
+For example, the \q{\is{llc_while}} combinator is defined as a fixed point (cf.~\S\ref{sec:modeling_ctrl_flow}). 
+Using its definition as an unfold rule, the preprocessor will automatically convert while loops into tail calls.
+This allows for using while-loops without trusting their translation in the code generator. 
+% However, to be efficient, we have to enable LLVM's tail call optimization, which will translate the function back into a loop.
+A configuration option in our tool lets the user choose between direct while-loop translation or unfolding to a tail call.
+
+% A standard example is the while loop combinator, which we define by the equation
+% \begin{lstlisting}
+%   llc_while b c s = do {ctd <- b s; llc_if ctd (do {s <- c s; llc_while b c s}) (return s)}
+% \end{lstlisting}
+% We use this equation as unfold rule, such that while loops are automatically converted to tail calls. Note that this 
+% feature allows for the concise specifications of loops, without adding additional trusted code. 
+% As LLVM has tail-call optimization, the compiled programs will be efficient.
 
@@
 \end{lstlisting}
-That is, it defined a new function \q{\is{euclid_0}} to replace the while loop by tail recursion.
+That is, it defined a new constant \q{\is{euclid_0}} to replace the while loop by tail recursion.
 \end{example}
@@
 and $a+b$ describes the memory described by the union of $a$ and $b$. 
-For the exact definition of a separation algebra, we refer to \cite{CHY07,KKB12}. We note that 
+For the exact definition of a separation algebra, we refer to~\cite{CHY07,KKB12}. We note that 
 separation algebras naturally extend over functions, pairs, and option types.
@@
 \end{lstlisting}
-here, \q{\is{[k |->v]}} is the partial function that maps $k$ to $v$, and \q{\is{i \<cdot> a}} 
-prepends the item $i$ to all addresses in the domain of $a$. 
+Here, \q{\is{[k |->v]}} is the partial function that maps \q{\is{k}} to \q{\is{v}}, and \q{\is{i \<cdot> a}} 
+prepends the item \q{\is{i}} to all addresses in the domain of \q{\is{a}}. 
 It is straightforward (though technically involved) to show that abstract values form a separation algebra, 
@@
 In order to deallocate a block, we have to own the whole block. However, from the abstraction, we cannot infer the size of the block, and thus we cannot
-specify an assertion that ensures that we own the whole block. A remedy (which the author has seen in \cite{App14}) is to additionally 
+specify an assertion that ensures that we own the whole block. A remedy (which the author has seen in~\cite{App14}) is to additionally 
 abstract a block to its size. Thus, abstract blocks have the type \q{\is{ablock = (nat => aval) \x nat option}}. 
@@
 Predicates of type \q{\is{assn = memory => bool}} are called \emph{assertions}.
-The \emph{weakest precondition} of a program \q{\is{c :: 'a llM}}, a \emph{postcondition} \q{\is{Q :: 'a => assn}}, and a memory \q{\is{s}} is a predicate defined as:
+The \emph{weakest precondition} of a program \q{\is{c :: 'a llM}}, a \emph{postcondition} \q{\is{Q :: 'a => assn}}, and a memory \q{\is{s}} is defined as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Intuitively, \q{\is{wp c Q s}} states that program \q{\is{c}}, if run on memory \q{\is{s}}, terminates successfully and the result \q{\is{r}} and the abstraction of the new 
-state \q{\is{s'}} satisfy \q{\is{Q}}.
+Intuitively, \q{\is{wp c Q s}} states that program \q{\is{c}}, if run on memory \q{\is{s}}, terminates successfully with the result \q{\is{r}}, and the abstraction of the new 
+state \q{\is{s'}} satisfies \q{\is{Q}}.
 
@@
 % Note that we use the size encoded in the tag to ensure that we possess the whole block. 
-For the actual LLVM memory instructions, we obtain the following rules:
+For the Isabelle-LLVM memory instructions, we obtain the following rules:
 \begin{lstlisting}
@@
   such that \q{\is{P |- P'*F}}, and replaces the goal by \q{\is{STATE (Q'*F) s' ==> Q s'}} for a fresh \q{\is{s'}}. 
-  Here, \q{\is{P|-Q = \<forall>s. P s ==> Q s}} denotes entailment.
+  Here, \q{\is{P |- Q = \<forall>s. P s ==> Q s}} denotes entailment.
   \item[final] If the goal has the form \q{\is{STATE P s ==> Q s}} such that \q{\is{Q}} is not of the form \q{\is{wp _ _ _}}, a heuristics is used to prove 
@@
 \begin{lstlisting}
-  |= {uint a a\impl * uint b b\impl * 0<a * 0<b} euclid a\impl b\impl {%r\impl. uint (gcd a b) r\impl}
-\end{lstlisting}
-Here, \q{\is{uint a a\impl}} states that \q{\is{a\impl::32 word}} is an unsigned integer with value \q{\is{a::int}}, 
+  |= {uint$_{64}$ a a\impl * uint$_{64}$ b b\impl * 0<a * 0<b} euclid a\impl b\impl {%r\impl. uint$_{64}$ (gcd a b) r\impl}
+\end{lstlisting}
+Here, \q{\is{uint$_{64}$ a a\impl}} states that \q{\is{a\impl::64 word}} is an unsigned integer with value \q{\is{a::int}}, 
 where \q{\is{int}} is the type of (mathematical) integers in Isabelle, and \q{\is{gcd}} is Isabelle's greatest common divisor function. 
@@
 \end{lstlisting}
-These are straightforward to prove in Isabelle, e.g., using sledgehammer\cite{BBP13}.
+These are straightforward to prove in Isabelle, e.g., using sledgehammer~\cite{BBP13}.
 \end{example}
@@
 \begin{lstlisting}
-|= {sint a a\impl * sint b b\impl * a-b \<in> sints n} ll_sub a\impl b\impl {%r\impl. sint (a-b) r\impl}
+|= {sint$_n$ a a\impl * sint$_n$ b b\impl * a-b \<in> sints n} ll_sub a\impl b\impl {%r\impl. sint$_n$ (a-b) r\impl}
   for a\impl b\impl :: n word and a b :: int
 \end{lstlisting}
+Here, \q{\is{sint$_n$}} implements mathematical integers by $n$-bit words.
 Note that the postcondition does not mention the operands \q{\is{a,b}} again, though they are still valid after the operation. 
-As \q{\is{sint}} is \emph{pure}, i.e., does not use the memory, our VCG can automatically add the corresponding assertions to the postcondition.
+As \q{\is{sint$_n$}} is \emph{pure}, i.e., does not use the memory, our VCG will automatically add the corresponding assertions to the postcondition.
 
 \section{Automatic Refinement}\label{sec:auto_ref}
-Our basic VCG infrastructure can be used to verify simple algorithms, like \q{\is{euclid}} from Example~\ref{ex:euclid-proof}.
+Our basic VCG infrastructure can be used to verify simple algorithms like \q{\is{euclid}} from Example~\ref{ex:euclid-proof}.
 However, many complex algorithms have already been verified using the Isabelle Refinement Framework~\cite{LaTu12}.
@@
 %
-We have modified the Sepref tool to translate to our LLVM monad instead.
+We have modified the Sepref tool to translate to Isabelle-LLVM's monad instead.
 We only had to modify the translation phase. The preprocessing phases, which only work on the abstract program, remained unchanged.
@@
 Reusing the basic data structures from the original Sepref tool is not possible, as Imperative HOL 
-uses arbitrary precision integers and algebraic data types, while we have only fixed width integers and pairs.
+uses arbitrary precision integers and algebraic data types, while we have only fixed width words and pairs.
 Up to now, we have added the implementation of integers and natural numbers 
-by $n$ bit words, the implementation of lists by arrays, and, on top of this, the implementation of a list 
-by an array and a length field. However, we could reuse the existing infrastructure of the Sepref tool: For example,
+by $n$ bit words and the implementation of lists by both fixed size and resizeable arrays.
+However, we could reuse the existing infrastructure of the Sepref tool: For example,
 there is support to automatically generate rules that also support refinement of the elements of a data structure, 
@@
 
-\subsection{Example: Knuth-Morris-Pratt String Search}
-Finally, we conducted a case study: 
-We took an existing formalization~\cite{HeLa17} of the Knuth-Morris-Pratt string search~\cite{KMP77} in the Refinement Framework.
-It uses Sepref to generate Imperative HOL code, which is then translated to Standard ML by Isabelle's code generator.
-We only had to slightly change the abstract part of the proof, in order to insert a few in-bounds assertions. 
-For example, when incrementing an index, we have to explicitly assert that it won't overflow. 
-As the invariants used in the original proof already contain this knowledge, adapting the proofs was straightforward.
-We then used the modified Sepref tool to synthesize LLVM instead of Imperative HOL. 
-For this, we fixed the lengths and indexes to 64 bit integers, and the characters to 8 bit integers.
-For the search string and text we use a pair of an array and a length field.
-The result of the automatic synthesis is an LLVM specification \q{\is{kmp_impl}}, and the theorem:
-\begin{lstlisting}
-(kmp_impl, kmp_SPEC) : [%s t. |s| + |t| < $2^{63}$] larray^k \x larray^k -> snat_option
-\end{lstlisting}
-Here, \q{\is{kmp_SPEC}} is the abstract specification of a string search algorithm, \q{\is{larray}} is 
-the refinement of a list by an array and a length, and \q{\is{snat_option}} implements the type \q{\is{nat option}} 
-by signed 64 bit integers, mapping \q{\is{None}} to $-1$. 
+\section{Case Studies}\label{sec:casestudies}
+To assess the usability of our approach, we have verified a binary search algorithm and 
+the Knuth-Morris-Pratt string search~\cite{KMP77} algorithm. 
+Both algorithms have also been verified with the original Sepref tool, such that we can compare the two approaches.
+
+\subsection{Binary Search}
+Binary search is a simple algorithm to find a value in a sorted array. 
+Despite its simplicity, it has a history of flawed implementations\footnote{A buggy implementation in the Java Standard Library has gone undetected for nearly a decade~\cite{bs_flaw_blogpost}.}, making it a natural example for formal verification.
+
+We start with a high-level specification: For a list \q{\is{xs}} and a value \q{\is{x}}, find the index of the first element greater or equal to \q{\is{x}}.
+We define the following constant:
+\begin{lstlisting}
+  fi_spec xs x = spec i. find_index (%y. x<=y) xs
+\end{lstlisting}
+where \q{\is{find_index P xs}} is a standard list function that returns the index of the 
+first element in \q{\is{xs}} that satisfies \q{\is{P}}, or \q{\is{length xs}} if there is no such element.
+
+Next, we phrase the binary search algorithm in the Isabelle Refinement Framework:
+\begin{lstlisting}
+  bin_search xs x == do {
+    (l,h) <- while
+      (%(l,h). l<h) 
+      (%(l,h). do {
+        assert (l<length xs \and h<=length xs \and l<=h);
+        let m = l + (h-l) div 2;
+        if xs!m < x then return (m+1,h) else return (l,m)
+      }) 
+      (0,length xs);
+    return l }"
+\end{lstlisting}
+It is a standard exercise to prove that the algorithm adheres to its specification:
+\begin{lstlisting}
+  bs_correct: sorted xs ==> bin_search xs x <= fi_spec xs x
+\end{lstlisting}
+Finally, we invoke our adapted Sepref tool: 
+\begin{lstlisting}
+  sepref_definition bs_impl [llvm_code] is bin_search
+    :: "(larray${}_{64}$ sint${}_{64}$)$^k$ -> sint${}_{64}^k$ -> snat$_{64}$"
+    unfolding bin_search_def [...] by sepref
+  export_llvm bs_impl file "bin_search.ll"
+  lemmas bs_impl_correct = bs_impl.refine[FCOMP bs_correct]
+\end{lstlisting}
+This produces an Isabelle-LLVM program \q{\is{bs_impl}}, exports it to actual LLVM text, and proves the refinement theorem
+\begin{lstlisting}
+  bs_impl_correct: (bs_impl, fi_spec) : [%(xs, _). sorted xs] (larray$_{64}$ sint$_{64}$)${}^k$ \x sint$_{64}^k$ -> snat$_{64}$
+\end{lstlisting}
+Here, \q{\is{snat${}_{w}$}} implements natural numbers by signed $w$-bit words\footnote{As LLVM's index operations are on signed words, 
+  it's convenient to always implement sizes and indexes by signed types, even if they are natural numbers.}.
+Moreover, \q{\is{larray${}_{w}$ A}} refines a list to an array and a $w$-bit length field, the elements of the list being refined by assertion \q{\is{A}}.
 The notation \q{\is{[\<Phi>] $A_1^{k|d}$ \x ... \x $A_n^{k|d}$ -> R}} specifies a refinement with precondition \q{\is{\Phi}}, such that the 
 arguments are refined by \q{\is{A_1 ... A\_n}} and the result is refined by \q{\is{R}}. 
-The $\cdot^{k|d}$ annotations specify whether the argument is overwritten ($k$ for keep, $d$ for destroy).
+The $\cdot^{k|d}$ annotations specify whether an argument is overwritten ($k$ for keep, $d$ for destroy).
 While we use this notation a lot in the Refinement Framework, it is straightforward to prove a standard Hoare triple from it. By unfolding some definitions we get:
 \begin{lstlisting}
-|= {larray s s\impl * larray t t\impl * |s| + |t| < $2^{63}$}
-   kmp_impl s\impl t\impl
-     {%i. larray s s\impl * larray t t\impl 
-      * if i=-1 then \<nexists>i. sublist_at s t i 
-          else i>=0 \and sublist_at s t (unat i) \and (\<forall>j<unat i. \<not>sublist_at s t j)}
-\end{lstlisting}
-That is, our algorithm requires the length of the search string $s$ plus the length of the text $t$ to be less than $2^{63}$. 
-Then, it will return the index of the first occurrence of $s$ in $t$, or $-1$ if $s$ does not occur in $t$. 
-Moreover, the algorithm will neither change $s$ nor $t$, as expressed by the respective assertions in the postcondition.
-
-The code generator produces actual LLVM text from \q{\is{kmp_impl}}, which we compile and link with a simple command line interface written in C, 
-using \textsc{clang}~\cite{Clang}.
-Similarly, we generate Standard ML code from the original formalization, which we compile using \textsc{MLton}~\cite{MLton}.
-We compare the programs by running them on a large text that does not contain the search string: 
-The program compiled by LLVM is an order of magnitude faster than the one compiled by MLton, and uses a third of the memory.
-We leave more systematic testing to future work.
+  |= {larray$_{64}$ sint$_{64}$ xs xs\impl ** sint$_{64}$ x x\impl ** sorted xs }
+      (bs_impl xs\impl x\impl)
+      (%i\impl. \<exists>i. larray$_{64}$ sint$_{64}$ xs xs\impl ** snat$_{64}$ i i\impl ** i=find_index (%y. x<=y) xs)"
+\end{lstlisting}
+% TODO: Perhaps just annotate the bit-width!
+That is, if we start with an array \q{\is{xs\impl}} representing the sorted list \q{\is{xs}}, and 
+a 64-bit word \q{\is{x\impl}} representing the integer \q{\is{x}},
+then the array still represents \q{\is{xs\impl}}, and the result \q{\is{i\impl}} represents a natural number \q{\is{i}}, which is equal to the correct index.
+
+The Sepref tool implements mathematical integers by 64-bit words, proving absence of overflows.
+This is only possible because the assertion in \q{\is{bin_search}} explicitly states that the indexes are in bounds. 
+Moreover, note the expression \q{\is{l + (h-l) div 2}} that we used to compute the midpoint index. 
+On mathematical integers, it is equal to \q{\is{(l+h) div 2}}. However, on fixed-width words, the latter may overflow, while the former does not\footnote{Exactly this 
+overflowing midpoint computation caused the infamous bug in the Java Standard Library~\cite{bs_flaw_blogpost}.}.
+
+\subsection{Knuth-Morris-Pratt String Search}
+Next, we regard the Knuth-Morris-Pratt (KMP) string search algorithm~\cite{KMP77}, a well-known 
+linear time algorithm to find the index of the first occurrence of a string $s$ in a string $t$:
+\begin{lstlisting}
+"ss_spec s t = spec
+  None => \<nexists>i. sublist_at s t i |
+  Some i => sublist_at s t i \<and> (\<forall>ii<i. \<not>sublist_at s t ii))"
+\end{lstlisting}
+where \q{\is{sublist_at s t i}} specifies that list \q{\is{s}} occurs in list \q{\is{t}} at index \q{\is{i}}:
+\begin{lstlisting}
+  "sublist_at s t i = \<exists>ps ss. t = ps@s@ss \<and> i = length ps"
+\end{lstlisting}
+
+We have recently formalized KMP with the original Sepref tool~\cite{HeLa17}.
+The adaption of the existing formalization was straightforward: 
+In the abstract part, we had to explicitly add a few in-bounds assertions. Most of them were already contained implicitly in the original proof.
+For the synthesis step, we only had to add setup for the fixed-width word types.
+The result of the automatic synthesis is an Isabelle-LLVM program \q{\is{kmp_impl}}, and the theorem:
+\begin{lstlisting}
+(kmp_impl, ss_spec)
+  : [%s t. |s| + |t| < $2^{63}$] (larray$_{64}$ sint$_{64}$)^k \x (larray$_{64}$ sint$_{64}$)^k -> snat_option$_{64}$
+\end{lstlisting}
+Here \q{\is{snat_option$_{64}$}} implements the type \q{\is{nat option}} by signed 64-bit words, mapping \q{\is{None}} to $-1$.
+
+\subsection{Runtime}
+We have compared our verified LLVM implementations to unverified C/C++ implementations of the same algorithms, 
+as well as to the Standard ML (SML) implementations generated by the original Sepref tool. 
+While we have implemented binary search in C ourselves, we used a publicly available code snippet~\cite{kmp_cpp} for 
+KMP\footnote{One easily finds many C implementations of KMP, mainly differing in the loop structure. We tried to choose one that is close to our implementation.}.
+The programs were compiled with MLton-$2018$~\cite{MLton} and clang-$6.0$~\cite{clang}, and run on a standard laptop machine (2.8GHz Quadcore i7 with 16MiB RAM).
+%
+Tables~\ref{tab:bs-runtimes} and \ref{tab:kmp-runtimes} display the 
+results: The verified LLVM implementations are on par with the unverified C/C++ implementations, and an order of magnitude faster 
+than the SML implementations.
+
+Isabelle's code generator uses arbitrary precision integers, which tend to be significantly slower than fixed-width integers.
+The SML${}^*$ column shows the results when we manually replace the arbitrary precision integers by 64-bit integers in the generated code.
+While this is unsound in general, it gives us a lower bound of what would be possible in SML with 
+more elaborate code generator configurations\footnote{Fleury et al.~\cite{FBL18} have successfully experimented with such code generator tuning.}.
+SML${}^*$ is significantly faster than the original SML, but still $1.5$ times slower than LLVM.
+
+\begin{table}
+\begin{minipage}{.49\textwidth}
+  \centering  
+  \begin{tabular}{c|cccc}
+    $n/10^6$ & C    & LLVM & SML    & SML${}^*$ \\
+    \hline
+    $1$      & 121  & 100  & 1999   & 139      \\
+    $2$      & 251  & 204  & 4209   & 289      \\
+    $3$      & 379  & 304  & 6516   & 440      \\
+    $4$      & 513  & 412  & 8843   & 600      \\
+    $5$      & 635  & 514  & 11494  & 756      \\
+    $6$      & 767  & 617  & 13646  & 917      \\
+    $7$      & 908  & 726  & 16032  & 1076     \\
+    $8$      & 1038 & 854  & 18421  & 1250     \\
+    $9$      & 1162 & 945  & 20957  & 1409     \\
+    $10$     & 1293 & 1045 & 23409  & 1564     \\
+  \end{tabular}
+  \caption{Time (ms) to search for the values $0,2,\ldots<5n$ in an array $[0,5,\ldots<5n]$.}\label{tab:bs-runtimes}
+\end{minipage}
+\hfill
+\begin{minipage}{.49\textwidth}
+  \centering  
+  \begin{tabular}{c|cccc}
+    $a$-$l$  & C++  & LLVM & SML  & SML${}^*$\\
+    \hline    
+    16-8     & 499 & 597  & 4616 & 918 \\
+    16-64    & 511 & 598  & 4621 & 926 \\
+    16-512   & 513 & 590  & 4573 & 909 \\
+    32-8     & 453 & 551  & 4471 & 850 \\
+    32-64    & 465 & 552  & 4523 & 857 \\
+    32-512   & 463 & 544  & 4456 & 840 \\
+    64-8     & 418 & 530  & 4433 & 803 \\
+    64-64    & 420 & 531  & 4514 & 809 \\
+    64-512   & 416 & 523  & 4411 & 799 \\
+  \end{tabular}
+  \caption{Time (ms) to run the $a$-$l$ benchmark suite from StringBench~\cite{stringbench}. 
+    Here $a$ is the alphabet size, and $l$ the pattern size. The sample size is $3\cdot2^{20}$ characters. 
+    The algorithm stops after finding the first match.}\label{tab:kmp-runtimes}
+\end{minipage}
+\end{table}
+
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% The same abstract algorithm can also be refined to imperative Standard ML code, using the original Sepref tool. 
+% Moreover, we have implemented the algorithm directly in C. 
+% While Isabelle's code generator uses arbitrary precision integers by default, there are extensions to support 
+% machine integers~\cite{Native_Word-AFP}, but require some effort to be used with the Refinement Framework. 
+% As a safe lower bound for the possible efficiency of these approaches, we have manually replaced the arbitrary precision integers by 64bit integers in the generated code\footnote{This is, of course, unsound in general.}.
+% 
+% Table~\ref{tab:bs-runtimes} shows the result of a benchmark ran on standard laptop hardware, using MLton~\cite{MLton} and Clang~\cite{Clang} as compilers.
+% The unverified C implementation and the verified LLVM implementation are equally fast, while the SML code is an order of magnitude slower.
+% When manually replacing arbitrary precision by 64bit integers (SML${}^*$), the resulting program is still slower by a factor of roughly $1.5$.
+% 
+% 
+% 
+% Table~\ref{tab:kmp-runtimes} shows the results of comparing the runtimes of
+% the generated LLVM code against the original implementation (SML), the original implementation with manual replacements of arbitrary precision by fixed-width integers (SML${}^*$), and a C++ implementation taken from~\cite{kmp_cpp}\footnote{There are several implementations of KMP floating around the internet, implementing different variants of the algorithm.
+We tried to choose one that is similar to our variant.}.
+% Again, LLVM is on par with C++, and one order of magnitude faster than SML, and roughly 1.5 times faster than SML${}^*$. 
+%
+
+% 
+% The code generator produces actual LLVM text from \q{\is{kmp_impl}}, which we compile and link with a simple command line interface written in C, 
+% using \textsc{clang}~\cite{Clang}.
+% Similarly, we generate Standard ML code from the original formalization, which we compile using \textsc{MLton}~\cite{MLton}.
+% We compare the programs by running them on a large text that does not contain the search string: 
+% The program compiled by LLVM is an order of magnitude faster than the one compiled by MLton, and uses a third of the memory.
+% We leave more systematic testing to future work.
+% 
+% Table~\ref{tab:kmp-runtimes} shows the comparisons of the run-times.
+
+% 
+% 
+% We only had to slightly change the abstract part of the proof, in order to insert a few in-bounds assertions. 
+% For example, when incrementing an index, we have to explicitly assert that it won't overflow. 
+% As the invariants used in the original proof already contain this knowledge, adapting the proofs was straightforward.
+% We then used the adapted Sepref tool to generate LLVM code for 64 bit indexes and 8 bit characters.
+% 
+% 
+% We report on a case study on a slightly more complicated algorithm.
+% We took an existing formalization~\cite{HeLa17} of the Knuth-Morris-Pratt string search~\cite{KMP77} in the Refinement Framework.
+% It uses Sepref to generate Imperative HOL code, which is then translated to Standard ML by Isabelle's code generator.
+% We only had to slightly change the abstract part of the proof, in order to insert a few in-bounds assertions. 
+% For example, when incrementing an index, we have to explicitly assert that it won't overflow. 
+% As the invariants used in the original proof already contain this knowledge, adapting the proofs was straightforward.
+% We then used the modified Sepref tool to synthesize Isabelle-LLVM instead of Imperative HOL. 
+% For this, we fixed the lengths and indexes to 64 bit integers, and the characters to 8 bit integers.
+% For the search string and text we use a pair of an array and a length field.
+% The result of the automatic synthesis is an Isabelle-LLVM program \q{\is{kmp_impl}}, and the theorem:
+% \begin{lstlisting}
+% (kmp_impl, kmp_SPEC) 
+%   : [%s t. |s| + |t| < $2^{63}$] (larray sint)^k \x (larray sint)^k -> snat_option
+% \end{lstlisting}
+% Here, \q{\is{kmp_SPEC}} is the abstract specification of a string search algorithm, \q{\is{larray}} is 
+% the refinement of a list by an array and a length, and \q{\is{snat_option}} implements the type \q{\is{nat option}} 
+% by signed 64 bit integers, mapping \q{\is{None}} to $-1$. 
+% The notation \q{\is{[\<Phi>] $A_1^{k|d}$ \x ... \x $A_n^{k|d}$ -> R}} specifies a refinement with precondition \q{\is{\Phi}}, such that the 
+% arguments are refined by \q{\is{A_1 ... A\_n}} and the result is refined by \q{\is{R}}. 
+% The $\cdot^{k|d}$ annotations specify whether the argument is overwritten ($k$ for keep, $d$ for destroy).
+% While we use this notation a lot in the Refinement Framework, it is straightforward to prove a standard Hoare triple from it. By unfolding some definitions we get:
+% \begin{lstlisting}
+% |= {larray s s\impl * larray t t\impl * |s| + |t| < $2^{63}$}
+%    kmp_impl s\impl t\impl
+%      {%i. larray s s\impl * larray t t\impl 
+%       * if i=-1 then \<nexists>i. sublist_at s t i 
+%           else i>=0 \and sublist_at s t (unat i) \and (\<forall>j<unat i. \<not>sublist_at s t j)}
+% \end{lstlisting}
+% That is, our algorithm requires the length of the search string $s$ plus the length of the text $t$ to be less than $2^{63}$. 
+% Then, it will return the index of the first occurrence of $s$ in $t$, or $-1$ if $s$ does not occur in $t$. 
+% Moreover, the algorithm will neither change $s$ nor $t$, as expressed by the respective assertions in the postcondition.
+
+% The code generator produces actual LLVM text from \q{\is{kmp_impl}}, which we compile and link with a simple command line interface written in C, 
+% using \textsc{clang}~\cite{Clang}.
+% Similarly, we generate Standard ML code from the original formalization, which we compile using \textsc{MLton}~\cite{MLton}.
+% We compare the programs by running them on a large text that does not contain the search string: 
+% The program compiled by LLVM is an order of magnitude faster than the one compiled by MLton, and uses a third of the memory.
+% We leave more systematic testing to future work.
+%TODO Add exact numbers.
 
 \section{Conclusions}
-We have formalized a semantics and a code generator for a fragment of LLVM. On top of the semantics, we have built a verification 
-infrastructure, and re-targeted the Sepref tool to connect the Refinement Framework to LLVM. As a case-study, we have 
-only slightly changed an existing formalization of the Knuth-Morris-Pratt string search algorithm to obtain a verified LLVM implementation,
-which is an order of magnitude faster than the Standard-ML implementation from the original formalization.
-
-Our LLVM semantics makes a few simplifying assumptions: It assumes an infinite supply of memory, and thus cannot assign a bit-size to pointers. 
-This assumption helps us to retain a deterministic semantics, which allows us to evaluate the semantics inside the theorem prover (cf.~Example~\ref{ex:fib_isabelle}). 
-We plan to use this feature for systematic testing of our semantics against the actual LLVM compiler. 
+We have developed Isabelle-LLVM, a shallowly embedded imperative language designed to be easily translated to actual LLVM text. 
+On top of this, we have built a verification 
+infrastructure, and re-targeted the Sepref tool to connect the Refinement Framework to LLVM. 
+As case studies, we have generated verified LLVM code for a binary search algorithm and the Knuth-Morris-Pratt string search algorithm.
+Both implementations are an order of magnitude faster than the ones generated with the original Sepref tool, and on par with unverified C implementations.
+The additional effort required to refine to LLVM instead of Standard ML was quite low.
+
+While our case studies only cover medium complex algorithms, we expect that our approach will scale to more complex algorithms, e.g.
+model checkers~\cite{WiLa18,FBL18} and SAT solvers~\cite{FBL18}, which have already been formalized with the original refinement framework. 
+While these formalizations use a combination of functional and imperative data structures, the LLVM backend only supports imperative data structures. 
+We expect the necessary changes to be manageable, but non-trivial.
+
+Compared to actual LLVM, Isabelle-LLVM makes a few simplifying assumptions: It assumes an infinite supply of memory, and thus cannot assign a bit-size to pointers. 
+This assumption helps us to retain a deterministic semantics, which is executable inside the theorem prover (cf.~Example~\ref{ex:fib_isabelle}). 
+We plan to use this feature for systematic testing of our code generator against the actual LLVM compiler. 
 A similar assumption is implicitly made for the stack, as our semantics permits arbitrarily deep recursive procedure calls.
@@
 
-Related work includes the Vellvm project \cite{ZNMZ12,ZNMZ13}, which formalizes a substantial fragment of LLVM in Coq.
-While we focus on generating verified programs, they focus on proving correctness of program analyses and transformations. 
-Moreover, we drew some of the ideas for our separation logic from the Verifiable C project~\cite{App14}, a Coq formalization of a separation logic on top of the CompCert C semantics~\cite{BL09}.
-
-Future work will include the verification of more basic data structures, which will enable us to 
-port more existing verification projects to LLVM, hopefully increasing their performance. 
-One challenge will be the support of sum types: Unfortunately, union types have been discontinued in LLVM, such that we have to resort 
-to low-level bit-fiddling to support them. However, this will increase the complexity of the code generator. An alternative would be a more sophisticated memory model.
-The current shallow embedding of the control flow semantics also has the disadvantage that we cannot easily model more complex control flow like breaking from loops or exceptions, without significantly increasing the complexity of the code generator.
+Related work includes the Vellvm project~\cite{ZNMZ12,ZNMZ13} that verifies LLVM program transformations in Coq.
+To be useful, e.g. as backend for clang, they have to formalize a substantial fragment of LLVM. 
+On the other hand, we can afford to formalize a simplified and abstract semantics that 
+is just powerful enough to cover what Sepref generates.
+We drew some of the ideas for our separation logic from the Verifiable C project~\cite{App14}, 
+a Coq formalization of a separation logic on top of the CompCert C semantics~\cite{BL09}.
+
+There exists various formalizations of low-level imperative languages, eg~\cite{Loch12,Watt18}. 
+These are focused on specifying the semantics, and we are not aware of any complex 
+algorithm verifications using these formalizations.
+% CakeML~\cite{KTMK19} is a verified compiler from StandardML down to assembly code. 
+% While it contains various low-level semantics, we are not aware direct correctness proofs
+The DeepSpec project~\cite{deepspec} aims at a completely verified computation environment, 
+down to machine code, including the operating system. 
+This is much more ambitious than the work presented here, which stops at a (simplified) LLVM semantics.
+For proving correct imperative programs, they have a separation logic based VCG for a fragment of C~\cite{App14,CBGD18},
+which they apply to several small C programs, mainly for cryptographic algorithms.
+
+Future work includes porting more existing verifications and data structures to Isabelle-LLVM, hopefully increasing their performance. 
+One challenge will be the limited support of nested container data structures by Sepref: The nested containers must either be 
+purely functional, or laborious ad-hoc proofs on the separation logic level are required. 
+As the former is not an option for (imperative) LLVM, we plan to add more support for nested imperative data structures to Sepref. 
+Other challenges will be the translation of sum-types (which are not supported by LLVM), and support for more 
+complex control flow statements.
+
+% 
+% 
+% Future work will include the verification of more basic data structures, which will enable us to 
+% port more existing verification projects to LLVM, hopefully increasing their performance. 
+% One challenge will be the support of sum types: Unfortunately, union types have been discontinued in LLVM, such that we have to resort 
+% to low-level bit-fiddling to support them. However, this will increase the complexity of the code generator. An alternative would be a more sophisticated memory model.
+% The current shallow embedding of the control flow semantics also has the disadvantage that we cannot easily model more complex control flow like breaking from loops or exceptions, without significantly increasing the complexity of the code generator.
 To this end, we are already working on formalizing the semantics of
-arbitrary control flow graphs, such that we can verify the translation of our control flow constructs to CFGs.
+% arbitrary control flow graphs, such that we can verify the translation of our control flow constructs to CFGs.
+% Finally, Sepref does not support nested container data structures on the heap. 
+% Currently, these have to be verified in an ad-hoc fashion using the underlying separation logic.
+% For functional code, a common workaround is to use purely functional inner data structures. This is obviously not an option for LLVM. 
+% We plan to investigate ways to add support for nested imperative data structures to Sepref.
 
 {\footnotesize
-\paragraph{\footnotesize Acknowledgement}
+\paragraph*{\footnotesize Acknowledgement}
 We thank Maximilian P.\ L.\ Haslbeck and Simon Wimmer for proofreading and useful suggestions.
-We received funding from DFG grant LA 3292/1 "Verifizierte Model Checker".
+We received funding from DFG grant LA 3292/1 "Verifizierte Model Checker" and VeTSS grant "Formal Verification of Information Flow Security for Relational Databases".
 }
@@
 
-\bibliographystyle{splncs03}
+% \bibliographystyle{splncs03}
 \bibliography{bib}

### 2019-06-07 benchmarks/_run_kmp_bench.sh

@@ -1,0 +1,28 @@
+#!/bin/bash
+
+# Usage: _run_kmp_bench name kmp_binary iterations
+
+NAME=$1
+KMP=$2
+ITERS=$3
+
+
+
+for als in 16 32 64; do
+  SAM="kmp_data/sample-$als.sample"
+  for ss in 8 64 512; do
+    PAT="kmp_data/pattern-$als-$ss.result"
+  
+    LOG="kmp-$NAME-$als-$ss.log"
+    date >$LOG
+  
+    sed 's/:.*//' "$PAT" | while read s; do
+      # echo "'$s'"
+      $KMP $ITERS "$s" "$SAM" | tee -a $LOG
+    done  
+  
+  done
+done
+
+
+

### 2019-06-07 benchmarks/bench_bs.sh

@@ -1,0 +1,57 @@
+#!/bin/bash
+
+
+function run() {
+  DIR=$1
+  NAME=$2
+
+  make -C "$DIR" bs
+
+  LOG="bs-$NAME.log"
+  
+  date >$LOG
+
+  for ((n=1000000;n<=10000000;n+=1000000)) do
+    $DIR/bs $n | tee -a "$LOG"
+  done
+}
+
+function eval() {
+  NAME=$1
+  LOG="bs-$NAME.log"
+  
+  egrep "^@" $LOG | sed -re 's/^@//'
+}
+
+
+RUN=false
+
+while [[ $# -gt 0 ]]
+do
+key="$1"
+
+case $key in
+    -r|--run)
+    RUN=true
+    ;;
+    *)
+    echo "Unknown command line argument $1"        # unknown option
+    exit 1
+    ;;
+esac
+shift # past argument or value
+done
+
+if $RUN; then
+  run ../etc/binsearch_c CPP
+  run ../thys/examples/code LLVM
+  run ../etc/binsearch_impholx SMLX
+  run ../etc/binsearch_imphol SML
+fi  
+  
+eval CPP
+eval LLVM
+eval SMLX
+eval SML
+
+

### 2019-06-07 benchmarks/bench_kmp.sh

@@ -1,0 +1,53 @@
+#!/bin/bash
+
+function run() {
+  DIR=$1
+  NAME=$2
+  ITER=$3
+
+  make -C "$DIR" kmp
+  
+  ./_run_kmp_bench.sh $NAME "$DIR/kmp" $ITER
+}
+
+function eval() {
+  NAME=$1
+  
+  for i in kmp-$NAME-*-*.log; do
+    echo -n "$i: "
+    grep '^@' "$i" | sed 's/.*://' | gawk '{x=x+$1} END {print x/1000}'
+  done
+}
+
+
+
+
+RUN=false
+
+while [[ $# -gt 0 ]]
+do
+key="$1"
+
+case $key in
+    -r|--run)
+    RUN=true
+    ;;
+    *)
+    echo "Unknown command line argument $1"        # unknown option
+    exit 1
+    ;;
+esac
+shift # past argument or value
+done
+
+if $RUN; then
+  run ../etc/kmp_c CPP 50
+  run ../thys/examples/code LLVM 50
+  run ../etc/kmp_impholx SMLX 30
+  run ../etc/kmp_imphol SML 10
+fi  
+  
+eval CPP
+eval LLVM
+eval SMLX
+eval SML

### 2019-06-07 etc/binsearch_c/bs.c

@@ -1,0 +1,54 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <time.h>
+
+typedef uint64_t elem_t;
+
+typedef struct {
+  int64_t len;
+  elem_t *data;
+} larray_t;
+
+int64_t bin_search (larray_t xs, elem_t x) {
+  int64_t l = 0;
+  int64_t h = xs.len;
+  
+  while (l<h) {
+    int64_t m = l + (h-l) / 2;
+    if (xs.data[m] < x) l=m+1;
+    else h=m;
+  }
+  
+  return l;
+}
+
+
+int main(int argc, char **argv) {
+
+  if (argc != 2) {printf("usage: bs <size>\n"); exit(1);}
+  
+  int64_t len = strtol(argv[1],NULL,10);
+
+  elem_t* ptr = (elem_t*)calloc(len,sizeof(elem_t));
+  
+  for (int64_t i=0;i<len;++i) ptr[i] = ((elem_t)i) * 5;
+  
+  larray_t arr = { len, ptr };
+  
+  clock_t time = clock();
+  
+  int64_t i = 0;
+  
+  for (elem_t x=0;x<5*len;x+=2)
+    i += bin_search(arr,x);
+  
+  time = clock() - time;  
+    
+  printf("Time: %fs, (idx checksum: %ld)\n",((double)time)/CLOCKS_PER_SEC,i);
+  
+  uint64_t time_ms = ((double)time)/CLOCKS_PER_SEC * 1000;
+  printf("@ %ld %ld\n",len,time_ms);
+  
+}

### 2019-06-07 etc/kmp_c/kmp.cpp

@@ -1,0 +1,119 @@
+#include <iostream>
+#include <string>
+#include <vector>
+#include <fstream>
+#include <streambuf>
+#include <cerrno>
+#include <ctime>
+
+
+using namespace std;
+
+// From https://gist.github.com/gongzhitaao/5e9d8f80aaba60e14a2c
+int64_t kmp(const string &T, const string &P) {
+  if (P.empty()) return 0;
+
+  vector<int64_t> pi(P.size(), 0);
+  for (int64_t i = 1, k = 0; i < P.size(); ++i) {
+    while (k && P[k] != P[i]) k = pi[k - 1];
+    if (P[k] == P[i]) ++k;
+    pi[i] = k;
+  }
+
+  for (int64_t i = 0, k = 0; i < T.size(); ++i) {
+    while (k && P[k] != T[i]) k = pi[k - 1];
+    if (P[k] == T[i]) ++k;
+    if (k == P.size()) return i - k + 1;
+  }
+
+  return -1;
+}
+
+
+/// from https://gist.github.com/yongpitt/5704216
+int* getPrefixFunction(string str)
+{
+  if(str=="")
+		return NULL;
+	int n=str.size();
+	int k=0;
+	int *A = new int[n];
+	A[0] = 0;
+	for(int i=1; i<n; i++){
+		while(k&&str[k]!=str[i])  //this loop is to find the k-1
+			k = A[k-1];          //The condition str[k]!=str[i] makes sure it finds 
+		if(str[i]==str[k])
+			k = k+1;
+		A[i] = k;
+	}
+	return A;
+}
+
+bool kmp2(string text, string pattern)
+{
+	if(pattern=="") return true;
+	if(text=="") return false;
+	int n = text.size();
+	int p = pattern.size();
+	int *sf = getPrefixFunction(pattern);
+	int localMatch = 0;
+	for(int i=0; i<=n-p; i++){
+		for(int j=localMatch, k=0; j<p; j++,k++){
+			if(text[i+k]==pattern[j])
+				localMatch++;
+			else{
+				localMatch = sf[localMatch];
+				i=i+(localMatch-sf[localMatch])+sf[localMatch]; //I previous put this before the above statement.
+				break;
+			}
+		}
+		if(localMatch==p)
+			return true;
+	}
+	return false;
+}
+
+std::string get_file_contents(const char *filename)
+{
[…]

### 2019-06-07 etc/llvm_tests/struct_cli.c

@@ -1,0 +1,49 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <time.h>
+
+
+typedef struct {
+  int64_t l;
+  int64_t c;
+  int64_t d;
+  //int64_t *data;
+} test;
+
+
+test create_test(void);
+
+int64_t read_test(test a) {
+  return a.c;
+}
+
+
+test create_test2() {
+  test a;
+  a.l=0;
+  a.c=0;
+  a.d=0;
+  return a;
+}
+
+
+int main(int argc, char **argv) {
+  test a;
+  
+  a.l=-1;
+  a.c=-1;
+  a.d=-1;
+//  a.data=(void*)0xFEFEFEFEFEFEFEFE;
+
+  printf("%ld,%ld,%ld\n",a.l,a.c,a.d);
+  
+  
+  a = create_test();
+
+  printf("%ld,%ld,%ld\n",a.l,a.c,a.d);
+  
+
+  return 0;
+}

### 2019-06-07 make_artifact.sh

@@ -1,13 +1,0 @@
-#!/bin/sh
-
-all_of () {
-  find "./$1" -path "*/obsolete" -prune -o -not -name "*~" -and -not -path "*/shallow/code2/*" \
-    -not -name "lorem.txt" -not -name "kmp" \
-    -type f -print
-}
-
-
-( all_of lib; all_of shallow; all_of paper/origKMPtest; echo ./README.md ) | sort > DIST_FILES
-
-tar -czf isabelle_llvm.tgz -T DIST_FILES --transform 's|^.|isabelle_llvm|'
-

### 2019-06-07 mkdist.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+
+DIR=${PWD##*/}
+
+cd ..
+tar -czf "$DIR/dist.tgz" --exclude-ignore-recursive=.distignore --transform "s|^$DIR|isabelle-llvm|" "$DIR"
+cd - >/dev/null 

### 2019-06-07 mkhtml.sh

@@ -27,17 +27,28 @@ done
 
 
 if $REBUILD; then
-./make_artifact.sh
+  cd paper
+  ./make.sh
+  cd ..
+  
+  cd thys
+  isabelle build -v -d '$AFP' -D .
+  cd ..
+  
+  ./mkdist.sh
 fi
 
 
 rm -rf html
 mkdir -p html
 
+ISABELLE_BROWSER_INFO=$(isabelle getenv ISABELLE_BROWSER_INFO | sed -re 's/.*=//')
 
+cp -a $ISABELLE_BROWSER_INFO/Unsorted/Isabelle_LLVM html/
 cp README.md html/
-cp isabelle_llvm.tgz html/
+cp dist.tgz html/
 cp LICENSE html/
+cp paper/main.pdf html/
 
 
 pandoc -V pagetitle="Isabelle LLVM" -s README.md > html/index.html

### 2019-06-07 shallow/code2/arraycpy_cli.c

@@ -1,36 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-
-
-void arraycpy(char *dst, char *src, int64_t n);
-
-void arraycp2(char *dst, char *src, int64_t n) {
-  int64_t i=0;
-  
-  while (i<n) {
-    dst[i] = src[i];
-    ++i;
-  }
-}
-
-
-int main (int argc, char** argv) {
-  
-  int64_t size = 1000 * 1024*1024;
-  
-  char *x = calloc(1, size);
-  char *y = calloc(1, size);
-  
-  memset(y,27,size);
-  arraycpy(x,y,size);
-  
-  int j=0;
-  for (int i=0;i<size;++i) j+=x[i];
-  
-  return j;
-}
-
-
-

### 2019-06-07 shallow/code2/cr_big_al_cli.c

@@ -1,27 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-
-
-int64_t cr_big_al(int64_t);
-
-
-int32_t euclid2(int32_t a, int32_t b) {
-  while (a!=b) {
-    if (a<=b) b=b-a; else a=a-b;
-  }
-  return a;
-}
-
-
-int main (int argc, char** argv) {
-  
-  if (argc != 2) {printf("Expecting one argument\n"); exit(1);}
-  
-  int64_t n = strtoul (argv[1],NULL,10);
-  
-  printf ("sum %ld = %ld\n", n, cr_big_al(n));
-}
-
-
-

### 2019-06-07 shallow/code2/cr_big_vector.cpp

@@ -1,24 +1,0 @@
-#include <vector>
-#include <iostream>   // std::cout
-#include <string>     // std::string, std::stoi
-
-
-int main(int argc, char** argv) {
-  int64_t n = std::stol(argv[1]);
-  
-  auto v = std::vector<int64_t>();
-  
-  for (int64_t i=n;i>0;--i) v.push_back(i);
-  
-  int64_t s=0;
-  
-  while (n>0) {
-    --n;
-    s+=v[n];
-  }
-  
-  std::cout<<"result = "<<s<<std::endl;
-  return 0;
-}  
-
-

### 2019-06-07 shallow/code2/euclid_cli.c

@@ -1,28 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-
-
-int32_t euclid(int32_t, int32_t);
-
-
-int32_t euclid2(int32_t a, int32_t b) {
-  while (a!=b) {
-    if (a<=b) b=b-a; else a=a-b;
-  }
-  return a;
-}
-
-
-int main (int argc, char** argv) {
-  
-  if (argc != 3) {printf("Expecting two arguments\n"); exit(1);}
-  
-  int32_t a = strtoul (argv[1],NULL,10);
-  int32_t b = strtoul (argv[2],NULL,10);
-  
-  printf ("gcd(%d,%d) = %d\n", a, b, euclid(a,b));
-}
-
-
-

### 2019-06-07 shallow/code2/lib_isabelle_llvm.c

@@ -1,23 +1,0 @@
-#include <stdlib.h>
-#include <stdio.h>
-
-void isabelle_llvm_abort() {
-  printf("%s\n","Isabelle-LLVM (abort): Dynamic check failed.");
-  abort();
-}
-
-void isabelle_llvm_abort_msg(char *msg) {
-  printf("Isabelle-LLVM (abort): %s\n",msg);
-  abort();
-}
-
-char* isabelle_llvm_calloc(size_t n, size_t m) {
-  char *res = calloc(n,m);
-  if (!res) isabelle_llvm_abort_msg("Out of memory");
-  return res;
-}
-
-void isabelle_llvm_free(char *p) {
-  free(p);
-}
-

### 2019-06-07 shallow/code2/qs_quicksort_cli.c

@@ -1,32 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-
-
-int64_t qs_quicksort(int64_t*, int64_t, int64_t);
-
-
-int main (int argc, char** argv) {
-  
-  if (argc != 2) {printf("Expecting one argument\n"); exit(1);}
-
-  int64_t n = strtoul(argv[1],NULL,10);
-
-  int64_t *A = malloc (n * sizeof(int64_t));
-  
-  srand(192);
-  
-  for (int64_t i=0; i<n; ++i) A[i] = rand() % (2*n);
-
-  for (int64_t i=0; i<n; ++i) printf ("%ld ",A[i]);
-  printf("\n");
-    
-  qs_quicksort(A,0,n-1);
-  
-  for (int64_t i=0; i<n; ++i) printf ("%ld ",A[i]);
-  printf("\n");
-  
-}
-
-
-

### 2019-06-07 shallow/code2/string_cli.c

@@ -1,20 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-
-
-int64_t my_strlen(char *);
-char streq(char *, char *);
-
-
-int main (int argc, char** argv) {
-  
-  if (argc != 3) {printf("Expecting two arguments\n"); exit(1);}
-
-  if (streq(argv[1],argv[2])) printf("Equal\n");
-  else if (my_strlen(argv[1]) == my_strlen(argv[2])) printf("Not equal (but eq len) %ld\n", my_strlen(argv[1]));
-  else printf("Not equal, len not equal %ld != %ld\n",my_strlen(argv[1]),my_strlen(argv[2]));
-}
-
-
-

### 2019-06-07 shallow/sepref/Examples/kmpcli.c

@@ -1,79 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-
-
-typedef struct {
-  int64_t len;
-  char *str;
-} string;
-
-int64_t kmp(string,string);
-
-
-string string_of_str(char *s) { string r = {strlen(s),s}; return r;}
-
-void error(char * msg) {
-  printf("Error: %s\n",msg);
-  exit(1);
-}
-
-
-string string_of_file(char *fname) {
-  // code snippet found on fundza.com
-  /* declare a file pointer */
-  FILE    *infile;
-  char    *buffer;
-  int64_t numbytes;
-  
-  /* open an existing file for reading */
-  infile = fopen(fname, "r");
-  
-  /* quit if the file does not exist */
-  if(infile == NULL) error("No such file");
-  
-  /* Get the number of bytes */
-  fseek(infile, 0L, SEEK_END);
-  numbytes = ftell(infile);
-  
-  /* reset the file position indicator to 
-  the beginning of the file */
-  fseek(infile, 0L, SEEK_SET);	
-  
-  /* grab sufficient memory for the 
-  buffer to hold the text */
-  buffer = (char*)calloc(numbytes, sizeof(char));	
-  
-  /* memory error */
-  if(buffer == NULL) error("Out of memory");
-  
-  /* copy all the text into the buffer */
-  fread(buffer, sizeof(char), numbytes, infile);
-  fclose(infile);
-  
-  string res = {numbytes,buffer};
-  return res;
-}
-
-void free_string(string s) { free (s.str); }
-
-
-
-int main (int argc, char** argv) {
-  
-  if (argc != 4) {printf("usage: kmp <num-iterations> <search-string> <textfile>\n"); exit(1);}
-  
-  int n = strtol(argv[1],NULL,10);
-  string s1 = string_of_str(argv[2]);
-  string s2 = string_of_file(argv[3]);
-  
-  for (int i=1;i<n;++i) kmp(s1,s2);
-  int64_t res = kmp(s1,s2);
-  
-  printf("result (%d iterations) = %ld\n", n, res);
-  
-  free_string(s2);
-  
-  return 0;
-}

### 2019-06-07 src/lib_isabelle_llvm.c

@@ -1,0 +1,24 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+void isabelle_llvm_abort() {
+  printf("%s\n","Isabelle-LLVM (abort): Dynamic check failed.");
+  abort();
+}
+
+void isabelle_llvm_abort_msg(char *msg) {
+  printf("Isabelle-LLVM (abort): %s\n",msg);
+  abort();
+}
+
+char* isabelle_llvm_calloc(size_t n, size_t m) {
+  printf("calloc(%ld,%ld)",n,m);
+  char *res = calloc(n,m);
+  if (!res) isabelle_llvm_abort_msg("Out of memory");
+  return res;
+}
+
+void isabelle_llvm_free(char *p) {
+  free(p);
+}
+

### 2019-06-07 thys-2019/examples/code/kmpcli.c

@@ -1,0 +1,87 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <time.h>
+
+typedef struct {
+  int64_t len;
+  char *str;
+} string;
+
+int64_t kmp(string,string);
+
+
+string string_of_str(char *s) { string r = {strlen(s),s}; return r;}
+
+void error(char * msg) {
+  printf("Error: %s\n",msg);
+  exit(1);
+}
+
+
+string string_of_file(char *fname) {
+  // code snippet found on fundza.com
+  /* declare a file pointer */
+  FILE    *infile;
+  char    *buffer;
+  int64_t numbytes;
+  
+  /* open an existing file for reading */
+  infile = fopen(fname, "r");
+  
+  /* quit if the file does not exist */
+  if(infile == NULL) error("No such file");
+  
+  /* Get the number of bytes */
+  fseek(infile, 0L, SEEK_END);
+  numbytes = ftell(infile);
+  
+  /* reset the file position indicator to 
+  the beginning of the file */
+  fseek(infile, 0L, SEEK_SET);	
+  
+  /* grab sufficient memory for the 
+  buffer to hold the text */
+  buffer = (char*)calloc(numbytes, sizeof(char));	
+  
+  /* memory error */
+  if(buffer == NULL) error("Out of memory");
+  
+  /* copy all the text into the buffer */
+  fread(buffer, sizeof(char), numbytes, infile);
+  fclose(infile);
+  
+  string res = {numbytes,buffer};
+  return res;
+}
+
+void free_string(string s) { free (s.str); }
+
+
+
+int main (int argc, char** argv) {
+  
+  if (argc != 4) {printf("usage: kmp <num-iterations> <search-string> <textfile>\n"); exit(1);}
+  
+  int n = strtol(argv[1],NULL,10);
+  string s1 = string_of_str(argv[2]);
+  string s2 = string_of_file(argv[3]);
+  
+  clock_t time = clock();
+  
+  for (int i=1;i<n;++i) kmp(s1,s2);
+  int64_t res = kmp(s1,s2);
+
+  time = clock() - time;
+  double dtime = ((double)time)/CLOCKS_PER_SEC*1000000/n;
+    
+  printf("Time: %fus\n",dtime);
+  printf("result (%d iterations) = %ld\n", n, res);
+  
+  printf("@(%d) '%s': %d\n",n, s1.str, (int)(dtime));
+  
+  free_string(s2);
+  
+  return 0;
+}

### 2019-06-07 thys-2019/sepref/IICF/Impl/Heaps/sort_cli.c

@@ -1,0 +1,35 @@
+#include <stdint.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <string.h>
+#include <time.h>
+
+
+typedef struct {
+  int64_t l;
+  int64_t c;
+  int64_t *data;
+} array_list;
+
+extern array_list sort(array_list);
+extern array_list arl_new();
+extern array_list arl_push_back(array_list,int64_t);
+
+
+int main(int argc, char **argv) {
+  array_list a;
+  a.l=0;
+  a.c=0;
+  a.data=0;
+  
+  a = arl_new();
+  
+  printf("%ld, %ld\n",a.l,a.c);
+  
+  arl_push_back(a,42);
+  
+  // for (int i=0;i<5;i++) a=arl_push_back(a,i);
+  
+  return 0;
+}
+

### 2019-06-24 paper/main.tex
paper tuned

@@
 We present a framework to generate verified LLVM programs from Isabelle/HOL.
-It consists of a code generator that generates LLVM text from 
+It is based on a code generator that generates LLVM text from 
 a simplified fragment of LLVM, shallowly embedded into Isabelle/HOL.
@@
 It has been successfully applied to verify many algorithms and software systems, 
-among them LTL and timed automata model checkers~\cite{ELNN13,BrLa18,WiLa18}, network flow algorithms~\cite{LaSe17,LaSe16}, a SAT-solver certification tool~\cite{La17_CADE,La17_SAT}, and even a SAT solver~\cite{FBL18}.
+among them LTL and timed automata model checkers~\cite{ELNN13,BrLa18,WiLa18}, network flow algorithms~\cite{LaSe19,LaSe16}, a SAT-solver certification tool~\cite{La17_CADE,La17_SAT}, and even a SAT solver~\cite{FBL18}.
 Using Isabelle/HOL's code generator~\cite{HKKN13}, the verified algorithms can be extracted to functional languages like Haskell or Standard ML.
@@
 faster than the corresponding Standard-ML code and on par with unverified C implementations.
+The paper ends with the discussion of future work (\S\ref{sec:future_work}) and related work (\S\ref{sec:related_work}).
 %
@@
 %
-Here, the type \q{\is{lint}} is a fixed bit width word type with a two's complement semantics, as used by LLVM. 
+Here, the type \q{\is{lint}} is a fixed bit width word type with a two's complement semantics, as used by LLVM, 
+and pair corresponds to a 2-element structure in LLVM. 
 The type \q{\is{rptr}} is either null or an address. An address is a path through the memory structure to a value:
@@
 It is straightforward to map an Isabelle-LLVM program to an actual LLVM program.
-Each equation of the form \q{\is{proc x_1 .. x\_n = block}} is mapped to an LLVM procedure named \q{\is{proc}}. 
+Each equation of the form \q{\is{proc x_1 .. x\_n = block}} is mapped to an LLVM function named \q{\is{proc}}. 
 A block is mapped to a control flow graph. Instructions and procedure calls are directly mapped to LLVM instructions and calls.
@@
 \subsection{Basic Reasoning Infrastructure}
-Predicates of type \q{\is{assn = memory => bool}} are called \emph{assertions}.
+Predicates of type \q{\is{assn = amemory => bool}} are called \emph{assertions}.
 The \emph{weakest precondition} of a program \q{\is{c :: 'a llM}}, a \emph{postcondition} \q{\is{Q :: 'a => assn}}, and a memory \q{\is{s}} is defined as:
@@
 |= {\<box>} allocn v n {%p. malloc_tag n p * range {0..<n} (%_. v) p}
-|= {malloc_tag n p * range {0..<n} blk p} free p {%_. \<box>}
+|= {malloc_tag n p * \<exists>blk. range {0..<n} blk p} free p {%_. \<box>}
 \end{lstlisting}
@@
 \end{lstlisting}
-Here, \q{\is{pto x p}} describes that $p$ points to value $x$. We prove similar rules for the other instructions.
+Here, \q{\is{pto x p}} describes that $p$ points to value $x$, and we write predicates as if they were assertions 
+on the empty memory, e.g., \q{\is{r=x}} instead of \q{\is{\<lambda>s. s=0 \<and> r=x}}.
+We prove similar rules for the other instructions.
 
@@
 It features a non-deterministic programming language with a refinement calculus and a VCG. 
-It allows to express an algorithm on the abstract mathematical concepts, and then refine it in multiple steps towards an efficient implementation.
+It allows to express an algorithm using abstract mathematical concepts, and then refine it in multiple steps towards an efficient implementation.
 The last step of a refinement is typically performed by the Sepref tool~\cite{La15}, which translates a program from the non-deterministic monad of the Refinement Framework 
@@
 Up to now, we have added the implementation of integers and natural numbers 
-by $n$ bit words and the implementation of lists by both fixed size and resizeable arrays.
-However, we could reuse the existing infrastructure of the Sepref tool: For example,
+by $n$ bit words and some basic container data structures like dynamic arrays, bit-vectors, and min-heaps.
+Thereby, we could reuse the existing infrastructure of the Sepref tool: For example,
 there is support to automatically generate rules that also support refinement of the elements of a data structure, 
@@
 \end{lstlisting}
-This produces an Isabelle-LLVM program \q{\is{bs_impl}}, exports it to actual LLVM text, and proves the refinement theorem
-\begin{lstlisting}
-  bs_impl_correct: (bs_impl, fi_spec) : [%(xs, _). sorted xs] (larray$_{64}$ sint$_{64}$)${}^k$ \x sint$_{64}^k$ -> snat$_{64}$
+This produces an Isabelle-LLVM program \q{\is{bs_impl}}, exports it to actual LLVM text, and proves the refinement theorem \q{\is{bs_impl_correct}}:
+\begin{lstlisting}
+  (bs_impl, fi_spec) : [%(xs, _). sorted xs] (larray$_{64}$ sint$_{64}$)${}^k$ \x sint$_{64}^k$ -> snat$_{64}$
 \end{lstlisting}
@@
 \begin{lstlisting}
-  |= {larray$_{64}$ sint$_{64}$ xs xs\impl ** sint$_{64}$ x x\impl ** sorted xs }
-      (bs_impl xs\impl x\impl)
-      (%i\impl. \<exists>i. larray$_{64}$ sint$_{64}$ xs xs\impl ** snat$_{64}$ i i\impl ** i=find_index (%y. x<=y) xs)"
+  |= {larray$_{64}$ sint$_{64}$ xs xs\impl * sint$_{64}$ x x\impl * sorted xs }
+     bs_impl xs\impl x\impl
+     {%i\impl. \<exists>i. larray$_{64}$ sint$_{64}$ xs xs\impl * snat$_{64}$ i i\impl * i=find_index (%y. x<=y) xs}"
 \end{lstlisting}
@@
 
-\section{Conclusions}
-We have developed Isabelle-LLVM, a shallowly embedded imperative language designed to be easily translated to actual LLVM text. 
-On top of this, we have built a verification 
-infrastructure, and re-targeted the Sepref tool to connect the Refinement Framework to LLVM. 
-As case studies, we have generated verified LLVM code for a binary search algorithm and the Knuth-Morris-Pratt string search algorithm.
-Both implementations are an order of magnitude faster than the ones generated with the original Sepref tool, and on par with unverified C implementations.
-The additional effort required to refine to LLVM instead of Standard ML was quite low.
-
+\section{Future Work}\label{sec:future_work}
 While our case studies only cover medium complex algorithms, we expect that our approach will scale to more complex algorithms, e.g.
@@
 We expect the necessary changes to be manageable, but non-trivial. 
+In particular, the current Sepref tool only supports pure data structures to be nested in containers. In the Imperative HOL setting, we simply 
+use functional data structures inside containers.
+For LLVM, nested container data structures currently require ad-hoc proofs on the separation logic level.
+We leave the lifting of Sepref to support nested imperative data structures to future work.
+
+Moreover, the refinement from arbitrary precision integers to fixed size integers was quite straightforward for our case studies, 
+and we expect these refinements to be more complex in general.
+We leave it to future work to explore this issue more systematically, and to provide semi-automated tools,
+e.g.~along the lines of AutoCorres~\cite{GLAK14}.
+
+Our code generator, as well as most standard code generators in theorem provers, 
+translate from logic to target language code, implicitly identifying logical 
+concepts with programming language concepts. This approach is simple, however, 
+the translation algorithm and its implementation become part of the trusted code base.
+More recently, code generators that translate into a deeply embedded semantics of the 
+target language have been developed~\cite{MyOw14,HuNi18}. 
+We leave a translation to a deep embedding of LLVM to future work, and note that a 
+deep embedding will also enable more advanced control flow constructs like exceptions and breaking from loops, 
+without significantly increasing the trusted code base.
 
@@
 
+Currently, we interface our generated LLVM code from C programs compiled by clang. However, the ABIs of C and LLVM only partially match, 
+and some LLVM constructs cannot be expressed in C at all.
+Currently, it is the user's responsibility to implement a correct header file. 
+We plan to automatically generate header files and adapter functions to make the exported code accessible from C.
+
+\section{Related Work}\label{sec:related_work}
 This project would not have been possible without several independent Isabelle developments:
@@
 
-Related work includes the Vellvm project~\cite{ZNMZ12,ZNMZ13} that verifies LLVM program transformations in Coq.
+The Vellvm project~\cite{ZNMZ12,ZNMZ13} verifies LLVM program transformations in Coq.
 To be useful, e.g. as backend for clang, they have to formalize a substantial fragment of LLVM. 
@@
 is just powerful enough to cover what Sepref generates.
+
 We drew some of the ideas for our separation logic from the Verifiable C project~\cite{App14}, 
@@
 % While it contains various low-level semantics, we are not aware direct correctness proofs
+
 The DeepSpec project~\cite{deepspec} aims at a completely verified computation environment, 
@@
 
-Future work includes porting more existing verifications and data structures to Isabelle-LLVM, hopefully increasing their performance. 
-One challenge will be the limited support of nested container data structures by Sepref: The nested containers must either be 
-purely functional, or laborious ad-hoc proofs on the separation logic level are required. 
-As the former is not an option for (imperative) LLVM, we plan to add more support for nested imperative data structures to Sepref. 
-Other challenges will be the translation of sum-types (which are not supported by LLVM), and support for more 
-complex control flow statements.
+\section{Conclusions}
+We have developed Isabelle-LLVM, a shallowly embedded imperative language designed to be easily translated to actual LLVM text. 
+On top of this, we have built a verification 
+infrastructure, and re-targeted the Sepref tool to connect the Refinement Framework to LLVM. 
+As case studies, we have generated verified LLVM code for a binary search algorithm and the Knuth-Morris-Pratt string search algorithm.
+Both implementations are an order of magnitude faster than the ones generated with the original Sepref tool, and on par with unverified C implementations.
+The additional effort required to refine to LLVM instead of Standard ML was quite low.
+
+% 
+% Future work includes porting more existing verifications and data structures to Isabelle-LLVM, hopefully increasing their performance. 
+% One challenge will be the limited support of nested container data structures by Sepref: The nested containers must either be 
+% purely functional, or laborious ad-hoc proofs on the separation logic level are required. 
+% As the former is not an option for (imperative) LLVM, we plan to add more support for nested imperative data structures to Sepref. 
+% Other challenges will be the translation of sum-types (which are not supported by LLVM), and support for more 
+% complex control flow statements.
 

### 2019-07-02 paper/main.tex
bounds checker

@@
 \authorrunning {P. Lammich}
+\Copyright{Peter Lammich}
 
@@
 \setcounter{footnote}{0}
+
+\supplement{\url{http://www21.in.tum.de/~lammich/isabelle_llvm}}
 
@@
 %
-The Isabelle theories described in this paper are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm}.
+The Isabelle theories described in this paper are available as supplement material (URL displayed in paper header).
 
@@
 As functions defined using the monad combinators are monotonic by construction~\cite{Kr10}, 
-we can define arbitrary recursive computations. The partial function package~\cite{Kr10} supports automation for monotonicity 
+we can define arbitrary recursive computations. The partial function package~\cite{Kr10} provides automation for monotonicity 
 proofs and for defining simple recursive functions. Mutual recursion still requires some manual effort, though it could be automated, too.
@@
 While we explicitly model values in memory by the type \q{\is{val}}, we model values in registers in a more shallow fashion:
-We identify LLVM registers with Isabelle variables that have a type of the shape \q{\is{T = T \x T | n word | T ptr}}. 
+We identify LLVM registers with Isabelle variables that have a type of shape \q{\is{T = T \x T | n word | T ptr}}. 
 Here, \q{\is{\x}} is Isabelle's product type, \q{\is{n word}} is the $n$ bit word type from Isabelle's word 
@@
 % However, to be efficient, we have to enable LLVM's tail call optimization, which will translate the function back into a loop.
-A configuration option in our tool lets the user choose between direct while-loop translation or unfolding to a tail call.
+A configuration option in our tool lets the user choose between direct while-loop translation or unfolding into a tail call.
 
@@
 Here, \q{\is{pto x p}} describes that $p$ points to value $x$, and we write predicates as if they were assertions 
-on the empty memory, e.g., \q{\is{r=x}} instead of \q{\is{\<lambda>s. s=0 \<and> r=x}}.
+on the empty memory, e.g., \q{\is{n~=0}} instead of \q{\is{\<lambda>s. s=0 \<and> n~=0}}.
 We prove similar rules for the other instructions.
@@
 We define refinement assertions to implement integers and natural numbers by $n$ bit words, and to implement lists by blocks of memory.
-Note that new data structures and refinement assertions can easily be added. In general, an implementation does not completely implement an abstract mathematical concept.
+On top of that, we define more complex data structures like dynamic arrays.
+Note that new data structures can easily be added. In general, an implementation does not completely implement an abstract mathematical concept.
 For example, $n$ bit words can only represent the integers \q{\is|sints n = $\{-2^{\text{n}-1}..<2^{\text{n}-1}\}$|}, and hash-tables can only represent finite sets.
@@
 \begin{lstlisting}
-  fi_spec xs x = spec i. find_index (%y. x<=y) xs
+  fi_spec xs x = spec i. i = find_index (%y. x<=y) xs
 \end{lstlisting}
@@
 On mathematical integers, it is equal to \q{\is{(l+h) div 2}}. However, on fixed-width words, the latter may overflow, while the former does not\footnote{Exactly this 
-overflowing midpoint computation caused the infamous bug in the Java Standard Library~\cite{bs_flaw_blogpost}.}.
+overflow caused the infamous bug in the Java Standard Library~\cite{bs_flaw_blogpost}.}.
 
@@
 Our code generator, as well as most standard code generators in theorem provers, 
-translate from logic to target language code, implicitly identifying logical 
+translates from logic to target language code, implicitly identifying logical 
 concepts with programming language concepts. This approach is simple, however, 
@@
 
-Compared to actual LLVM, Isabelle-LLVM makes a few simplifying assumptions: It assumes an infinite supply of memory, and thus cannot assign a bit-size to pointers. 
+Compared to actual LLVM, Isabelle-LLVM makes a few simplifying assumptions: 
+We do not support floating point arithmetic, though this could be added, e.g.~based on Lei Yu's floating point formalization \cite{Yu13}.
+Moreover, we only support two-element structures (pairs). This nicely fits Isabelle HOL's product datatype, and the nested structures 
+resulting from longer tuples should not be a problem for LLVM's optimizer.
+Also, we do not support concepts that are handy for program optimization, but not required for code generation, like poison values.
+Isabelle-LLVM assumes an infinite supply of memory, and thus cannot assign a bit-size to pointers. 
 This assumption helps us to retain a deterministic semantics, which is executable inside the theorem prover (cf.~Example~\ref{ex:fib_isabelle}). 
@@
 We leave that to future work, and note that our generated code allocates no large blocks of memory on the stack. 
-Thus, stack overflows are likely to hit the guard pages inserted by most operating systems, which will cause termination of the process.
+Thus, stack overflows are likely to hit the guard pages inserted by most operating systems, which will cause defined termination of the process.
 
@@
 The Vellvm project~\cite{ZNMZ12,ZNMZ13} verifies LLVM program transformations in Coq.
-To be useful, e.g. as backend for clang, they have to formalize a substantial fragment of LLVM. 
+To be useful, e.g.~as backend for clang, they have to formalize a substantial fragment of LLVM. 
 On the other hand, we can afford to formalize a simplified and abstract semantics that 

### 2019-09-03 paper/main.tex
bugfixes

@@
 \begin{lstlisting}
-  ll_udiv :: "n word => n word => n word"
+  ll_udiv :: "n word => n word => n word llM"
   ll_udiv a b = do { assert (b ~= 0); return (a div b) }
@@
 A program is represented by a set of (monomorphic) theorems of the shape
-\q{\is{proc\_i x_1 ... x\_n = cmd}},
+\q{\is{proc\_i x_1 ... x\_n = block}},
 where the \q{\is{proc\_i}} are Isabelle functions, the \q{\is{x\_i}} are variables, and all free variables on the right hand side are among the \q{\is{x\_i}}.

### 2019-11-22 papers/sorting/main.tex
sorting stuff

\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{balance}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
\usepackage{mathpartir}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{cprotect}

\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

\usepackage{multirow}
\usepackage{siunitx}

\usepackage{tikz}
\usetikzlibrary{positioning,fit,shadows,decorations, arrows, shapes, decorations.markings,decorations.pathmorphing,calc}

\tikzset{
  -|-/.style={
    to path={
      (\tikztostart) -| ($(\tikztostart)!#1!(\tikztotarget)$) |- (\tikztotarget)
      \tikztonodes
    }
  },
  -|-/.default=0.5,
  |-|/.style={
    to path={
      (\tikztostart) |- ($(\tikztostart)!#1!(\tikztotarget)$) -| (\tikztotarget)
      \tikztonodes
    }
  },
  |-|/.default=0.5,
}

\clubpenalty = 10000
\widowpenalty = 10000
\displaywidowpenalty = 10000

\input{lstisabelle}
\lstset{captionpos=b}
\lstset{numberbychapter=false}
\lstset{autogobble}

\newcolumntype{Y}{>{\RaggedRight\arraybackslash}X} % Multiline column, automatic width

\theoremstyle{definition}
\newtheorem{example}{Example}[section]

% \overfullrule=8pt

\begin{document}

\title{Efficient Verified Implementation of Introsort}

\author{Peter Lammich}
\email{peter.lammich@machester.ac.uk}
\affiliation{%
  \institution{The University of Manchester, UK}
}
%
% \authorinfo{Peter Lammich}
%            {The University of Manchester, UK}
%            {peter.lammich@machester.ac.uk}

\begin{abstract}
  Sorting algorithm are an important part of most standard libraries, and
  both, their correctness and efficiency is crucial for many applications.
%   There has been extensive research on improving sorting algorithms for
%   both, the general case of sorting arbitrary objects wrt.\ a weak ordering,
%   or special cases like sorting integers or strings.

  As generic sorting algorithm, the GNU C++ Standard Library implements the Introsort algorithm,
  a combination of quicksort, heapsort, and insertion sort.

  We verify this algorithm in the Isabelle LLVM verification framework,
  including most of the optimizations from the GNU implementation.
  On an extensive benchmark set, our verified algorithm performs on par with the original.
\end{abstract}

\maketitle

% \category{F.3.1}{Logics and Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}

% general terms are not compulsory anymore,
% you may leave them out
% \terms Verification
% term1, term2

% \keywords
% Sorting, Interactive Theorem Proving, Isabelle HOL, Stepwise Refinement, Separation Logic, LLVM

\section{Introduction}
  The goal of this paper is to provide a case study to show that the verification of
  state-of-the art implementations of medium-complex algorithms is feasible.
  We chose Musser's introspective sorting algorithm (Introsort)~\cite{Muss97}, and its
  implementation in the GNU C++ Library (libstdc++)~\cite{libstdc++}.
  Using the Isabelle-LLVM verification framework~\cite{La19-llvm}, we verify the same algorithm,
  including most of the optimizations from libstdc++. The result is a verified sorting algorithm, which performs
  on-par with the original implementation on an extensive set of benchmarks.

  Sorting algorithms are an important part of any standard library. The Introsort algorithm,
  being a combination of quicksort, heapsort, and insertion sort, is far from being trivial, but still
  has a manageable complexity. Finally, libstdc++ provides a state-of-the art implementation with several
  non-trivial but crucial optimizations. This makes Introsort a good candidate for our case study.

  Moreover, sorting algorithms in standard libraries have not always been correct.
  The Timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
  Also, many real-world mergesort implementations suffered from an overflow bug~\cite{bs_flaw_blogpost}.
  Finally, LLVM's libc++~\cite{libc++} implements a different quicksort based sorting algorithm. While it may be functionally correct,
  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by October 2019.}.

  To the best of our knowledge, this paper provides the first verification of the Introsort algorithm.
  This algorithm and two Java sorting algorithms~\cite{GRBBH15,BSSU17} are the only verified "real-world" implementations
  of sorting algorithms we are aware of. On the technical side, this paper provides a case study of using stepwise refinement
  for the design of verified algorithms.
  Moreover, we report on two extensions to the Isabelle-LLVM framework, to handle nested container data structures and to automatically
  generate C-header files to interface the generated code. These were required for our benchmark programs, which sort arrays of arrays of characters,
  and interface the generated LLVM code from C++.

  The complete Isabelle/HOL formalization and the benchmarks are submitted as artefact accompanying this paper.

%
%
%   Mention implementation in libc++, but not standard-conformant! $O(n^2)$ worst case
%
%   The Isabelle LLVM verification framework
%     stepwise refinement, code export to LLVM
%
%   Contributions
%     D First verified introsort? (that is on-par with state-of-the art implementation!)
%     D Case-Study: Verification of state-of-the art implementations of medium-complex algorithms possible!
%     ! Prototype for handling nested container data structures.

\section{The Introsort Algorithm}
  The Introsort algorithm by Musser~\cite{Muss97} is a sorting algorithm
  that combines the good average-case runtime of quicksort~\cite{Hoare61} with the worst-case complexity of heapsort~\cite{Will64}.
  The basic idea is to use quicksort as main sorting algorithm, but switch to heapsort when the recursion
  depth exceeds a given limit, usually $2\lfloor\log_2 n\rfloor$ for $n$ elements.

  Introsort is the default generic sorting algorithm in the GNU C++ Library~\cite{libstdc++}. Alg.~\ref{alg:introsort} displays its pseudo-code.
  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{partial\_introsort}{$a,l,h,d$}
      \If{$h-l>\textrm{threshold}$}
        \If{$d=0$}
          \State{\Call{heapsort}{$a,l,h$}}
        \Else
          \State{$m\gets\Call{partition\_pivot}{a,l,h}$}
          \State{\Call{partial\_introsort}{$a,l,m,d-1$}}
          \State{\Call{partial\_introsort}{$a,m,h,d-1$}}
        \EndIf
      \EndIf
    \EndProcedure

    \Procedure{introsort}{$a,l,h$}
      \If{$h-l>1$}
        \State{\Call{partial\_introsort}{$a,l,h,2\lfloor\log_2(h-l)\rfloor$}}
        \State{\Call{final\_insort}{$a,l,h$}}
      \EndIf
    \EndProcedure
  \end{algorithmic}
  \caption{Introsort}\label{alg:introsort}
  \end{algorithm}
  The \textsc{introsort} function sorts the slice $[l{..}{<}h]$ of an array $a$. For this, it first
  calls \textsc{partial\_introsort} to partially sort the array, i.e., all elements are no more than $\textrm{threshold}$ from their final position,
  and then finishes the sorting by invoking insertion sort. The \textsc{partial\_introsort} procedure implements a standard quicksort, that switches
  to heapsort when the recursion depth gets greater than $d$, and stops when the partition size becomes less than or equal to $\textrm{threshold}$.
  Note that the actual implementation in the GNU C++ Library contains a manual tail-call optimization,
  replacing the second recursive call to \textsc{partial\_introsort} by looping. For our formalization, we chose to not implement this
  optimization, in particular due to the fact that the LLVM optimizer recognizes and eliminates this tail call automatically.

  \subsection{Partitioning}
  The \textsc{partition\_pivot($a,l,h$)} algorithm must re-arrange the elements in $a$ and return a \emph{pivot} index $m\in[l{<}{..}{<}h]$, such
  that all elements $a[l{..}{<}m]$ are less than or equal to $a[m]$, and all elements $a[m{..}{<}h]$ are greater than or equal to $a[m]$.
  Note that the performance of quicksort crucially depends on pivot selection: Ideally, the pivot index should be close to $(h-l)/2$,
  i.e., the pivot element should be close to the median. The algorithm implemented by the GNU C++ Library is sketched in Alg.~\ref{alg:partition}.
  It selects the median of the three elements $a[l+1]$, $a[\lfloor (h-l)/2 \rfloor]$, and $a[h-1]$ as pivot (l.~\ref{line:partition:pivot_selection}).
  It then swaps the selected median element to the front $a[l]$, and invokes the standard Hoare partitioning algorithm,
  which searches for pairs of mismatched elements simultaneously from the start and end of the array, and swaps them.
  The chosen pivot element remains at position $a[l]$.
  \begin{algorithm}
  \begin{algorithmic}[1]
    \Function{partition}{$a,l,h,p$}
      \While{True}
        \While{$a[l]<a[p]$} ${+}{+}l$ \EndWhile \label{partition:nextl_loop}
        \State{${-}{-}h$}
        \While{$a[p]<a[h]$} ${-}{-}h$ \EndWhile \label{partition:nexth_loop}
        \If{$\neg l<h$} \Return{$l$}\EndIf
        \State{\Call{swap}{$a[l],a[h]$}}
        \State{${+}{+}l$}
      \EndWhile
    \EndFunction

    \Function{partition\_pivot}{$a,l,h$}
      \State{$p\gets \Call{median\_of\_three}{a,l+1,\lfloor h-l\rfloor/2,h-1}$} \label{line:partition:pivot_selection}
      \State{\Call{swap}{$a[l],a[p]$}}
      \State{\Return{\Call{partition}{a,l+1,h,l}}}
    \EndFunction
  \end{algorithmic}
  \caption{Partitioning}\label{alg:partition}
  \end{algorithm}

  Note that this function contains a subtle optimization: The loops in lines~\ref{partition:nextl_loop} and \ref{partition:nexth_loop}
  will iterate until they find a mismatched element. This iteration will not exceed the array bounds, only because the median selection
  placed \emph{stopper elements} at the beginning and end of the array. In particular when the element comparison function is
  cheap (e.g.\ for integer arrays), omitting an extra bounds check in the very inner loop of the sorting procedure can make a difference in practice.

  \subsection{Unguarded Insertion Sort}
  Insertion sort is a well-known simple sorting algorithm, which works by inserting the elements of the
  array, one after the other, into the sorted prefix of the array. Its worst case complexity is quadratic in general,
  but linear for partially sorted arrays with a constant threshold. For small thresholds, it performs better than quicksort,
  such that it is effective to stop the quicksort procedure early, and do the final sorting with insertion sort.
  Algorithm~\ref{alg:insort} shows the algorithm we implemented\footnote{The guarded
  version ($G=\textrm{true}$) slightly differs from the implementation in the GNU C++ Library. However, as it is only called for a small
  number of elements, we don't expect this to have any significant effect.}.

  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{insert}{$G,a,l,i$}
      \State{$t\gets a[i]$}
      \While{$ (\neg G \vee l<i) \wedge t<a[i-1]$} \label{alg:insert:loophead}
        \State{$a[i]\gets a[i-1]$}
        \State{${-}{-}i$}
      \EndWhile
      \State{$a[i]\gets t$}
    \EndProcedure

    \Procedure{gen\_insort}{$G,a,l,i,h$}
      \While{$i<h$}
        \State{\Call{insert}{$G,a,l,i$}}
        \State{${+}{+}i$}
      \EndWhile
    \EndProcedure

    \Procedure{final\_insort}{$a,l,h$}
      \If{$h-l\le\textrm{threshold}$}
        \State{\Call{gen\_insort}{$\textrm{true},a,l,l+1,h$}}
      \Else
        \State{\Call{gen\_insort}{$\textrm{true},a,l,l+1,l+\textrm{threshold}$}}
        \State{\Call{gen\_insort}{$\textrm{false},a,l,l+\textrm{threshold},h$}}
      \EndIf
    \EndProcedure
  \end{algorithmic}
  \caption{Final Insertion Sort}\label{alg:insort}
  \end{algorithm}

  This algorithm contains a similar optimization as the partitioning algorithm: As the array is already partially sorted,
  we know that elements beyond the threshold index won't be inserted at the very beginning of the array.
  This is exploited to get rid of the bounds check in line~\ref{alg:insert:loophead}.
  Only the first $\textrm{threshold}$ elements of the array are sorted with activated bounds check. In the displayed code, we control
  the bounds check by the Boolean parameter $G$, and assume that the algorithm will be specialized for both values of $G$.

%   Combination of quicksort, heapsort, and insertion sort
%   Worst-case $O(n\log n)$, average case efficiency of quicksort.

\section{Notation}
  Throughout this paper, we use some specific notation. We write function application in curried form, i.e., \is{f x_1 ... x$_n$}.
  Definitions of types and functions are denoted by \is{==}. Indexes start at zero. The \is$i$th element of a list is \is$xs[i]$,
  the slice of a list from index \is$i$ inclusive to \is$j$ exclusive is \is$xs[i..<j]$. We also use the notations \is{xs[i<..<j]}, \is{xs[i<..j]},
  and \is{xs[i..j]} to specify exclusiveness of bounds. Moreover, an omitted lower bound denotes index $0$, and an omitted upper bound denotes
  the length of the list, e.g., \is$xs[..<n]$ is the list of the first \is$n$ elements of $xs$, and \is$xs[n..]$ is the list of the remaining elements.
  List concatenation is written as \is{xs_1@xs_2}, and \is$xs[i:=x]$ is the list \is$xs$, with index \is$i$ updated to element \is$x$.
  The length of list \is$xs$ is denoted by \is{|xs|}.

\section{The Isabelle Refinement Framework}
  The Isabelle Refinement Framework~\cite{LaTu12} provides tools for
  verified program development by stepwise refinement, using the Isabelle HOL proof assistant~\cite{NPW02}.
  It is based on a nondeterminism-error monad, defined as follows:
  \begin{lstlisting}
    \<alpha> nres == FAIL | RES (\<alpha> set)
    return x == RES {x}
    bind FAIL f == FAIL | bind (RES X) f = |_|x\<in>X. f x
  \end{lstlisting}
  Intuitively, \is{FAIL} represents a program that may fail (e.g.~not terminate or fail explicitly by violating an assertion),
  and \is{RES X} represents a program that, nondeterministically, returns a value \is{x\<in>X}.
  By extending the subset ordering, with \is{FAIL} as greatest element, \is{\<alpha> nres} becomes a complete lattice.
  The program \is{return x} returns exactly the value \is{x}.
  The program \is{bind m f} sequentially composes \is{m} and \is{f}: it nondeterministically chooses a
  return value \is{x} of \is{m}, and then behaves like \is{f x}. If either \is{m}, or one of the \is{f x} may fail,
  then also \is{bind m f} may fail.

  We use sequential notation for bind, i.e., \is$x<-m; f x$ stands for \is$bind m (\<lambda>x. f x)$,
  and \is$m_1; m_2$ stands for \is$bind m_1 (\<lambda>_. m_2)$, i.e., the result of \is$m_1$ is ignored.
  We also use the standard HOL functions \is{if-then-else} and \is{let}, and define the following notations:
  \begin{lstlisting}
    assert \Phi == if \Phi then return () else FAIL
    spec x. \Phi x == RES {x | \Phi x}
  \end{lstlisting}
  An assertion will fail if its argument is false, and \is{spec x. \Phi x} describes all results that satisfy \is{\Phi}.

  Recursion is defined by the greatest fixed point over a monotonic functor \is{F}{\hspace{1pt}}\footnote{Monadic
    programs are monotonic by construction~\cite{Kr10}. Moreover, on monadic programs, the greatest fixed point over the \is{nres}-ordering
    coincides with the fixed point over a flat CCPO, which is the standard construction for recursion. In the actual
    formalization, monotonicity wrt.\ both orderings is asserted. }
  , and while loops are defined via recursion:
  \begin{lstlisting}
    rec F x == assert (mono F); fixp F x
    while b c s == rec (\<lambda>W s.
      if b s then s' <- c s; W s' else return s) s
  \end{lstlisting}
  From now on, we will omit the \is{rec} syntax, and define recursive monadic functions by their recursion equation, e.g.:
  \begin{lstlisting}
    while b c s ==
      if b s then s' <- c s; while b c s' else return s
  \end{lstlisting}

  The ordering on \is{nres} can be interpreted as refinement: \is{m_1 <= m_2} means
  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
  The statement
  \begin{lstlisting}
    f x <= { assert (P x); spec r. Q r}
  \end{lstlisting}
  means that function $f$ satisfies the specification with precondition \is{P} and postcondition \is{Q}.

\section{Verifying Introsort}
  The first step to verify a sorting algorithm is to specify the desired result.
  We specify a sorting algorithm as follows:
  \begin{lstlisting}
    sort_spec xs == spec xs'. sorted xs' \<and> mset xs' = mset xs
    $\textrm{where}$
    sorted xs == \<forall>i<j<|xs|. \<not>(xs[i] > xs[j])
  \end{lstlisting}
  That is, the result will be sorted and a permutation of the input (\is{mset xs} denotes the multiset of elements of list $xs$).

  Note that \is{<} is a \emph{strict weak ordering}, i.e., a relation with:
  \begin{lstlisting}
    a<b ==> \<not>b<a      and      a<c ==> a<b \or b<c
  \end{lstlisting}
  Weak orderings are used by C++ for sorting~\cite{Josu12}.
  They are a natural model for objects that are sorted by linearly ordered keys:
  An equivalent characterization can be given by an equivalence relation on the elements (e.g.~objects with same keys),
  and a linear ordering on the equivalence classes.
  In particular, any linear ordering is a weak ordering, with the equivalence relation being equality, and
  $\neg (a<b)$ meaning $b\le a$. While our formalization is over arbitrary weak orderings, we assume a linear ordering for the rest of
  this paper, to simplify the presentation.

  For a partial sorting algorithm wrt.\ threshold \is{n}, we replace \is{sorted} by:
  \begin{lstlisting}
    part_sorted xs == \<exists>ss. is_slicing xs ss \<and> sorted$_\preceq$ ss"
    $\textrm{where}$
    is_slicing xs ss == xs = concat ss \<and> (\<forall>s\<in>ss. |s| <= n)"
    xs $\preceq$ ys == \<forall>x\<in>xs. \<forall>y\<in>ys. x <= y
  \end{lstlisting}
  That is, a list is partially sorted if we can split it into slices that are not longer than
  the threshold $n$, and each element in a slice is less than or equal to any element in a subsequent slice.

  The next step to verify quicksort like algorithms is to specify a partitioning function:
  \begin{lstlisting}
    partition_spec xs == assert (|xs| >= 4);
      spec (\<lambda>(xs_1,xs_2). xs_1~=[] \and xs_2~=[]
             \and  mset xs = mset xs_1 + mset xs_2 \and xs_1 $\preceq$ xs_2)
  \end{lstlisting}
  Given a list \is{xs} of sufficient length, two non-empty lists \is{xs_1,xs_2} are returned,
  which together contain the same elements as $xs$, and each element of the first list is less than or equal to any element of the second list.
  We assume lists of at least four elements, because this is the size where the pivot selection makes sense (cf.~Alg.~\ref{alg:partition}).
  In practice, the lists will always be longer than the threshold, which is typically 16.

%   The final sorting step has to completely sort an already sorted list:
%   \begin{lstlisting}
%     final_sort_spec xs == assert (part_sorted xs); sort_spec xs
%   \end{lstlisting}
%
  Finally, we can specify the partial Introsort algorithm in the Refinement Framework:
  \begin{lstlisting}
    partial_introsort xs d ==
        if |xs| > threshold then {
            if d=0 then sort_spec xs
            else {
                (xs_1,xs_2) <- partition_spec xs;
                xs_1 <- partial_introsort xs_1 (d-1);
                xs_2 <- partial_introsort xs_2 (d-1);
                return (xs_1@xs_2) }}
        else return xs
  \end{lstlisting}
%
  A straightforward Isabelle proof yields:
  \begin{theorem}\label{thm:partial_introsort_correct}
  \begin{lstlisting}
    partial_introsort xs d <= part_sort_spec xs
  \end{lstlisting}
  \end{theorem}

\subsection{Refinement}\label{sec:refinement}
  The partial introsort algorithm that we have specified clearly captures the idea of the algorithm.
  However, it splits the input list and later concatenates it, which makes it unsuitable for an in-place
  implementation on arrays. In this section we will \emph{refine} this algorithm into one that is better suited
  for an implementation on arrays. Note that refinement preserves correctness, such that the new algorithm will
  automatically be correct.

  Given a relation \is{R} that relates \emph{concrete} with \emph{abstract} values, and a program \is{m} that returns abstract values,
  the program \is{\<Down>R m} is the biggest program that returns only concrete values with corresponding abstract values in \is{m}.
  Moreover, \is{\<Down>R m} may fail if and only if \is{m} may fail. Formally:
  \begin{lstlisting}
    \<Down> R FAIL == FAIL
    \<Down> R (RES A) == RES { c. \<exists>a\<in>A. (c,a)\<in>R }
  \end{lstlisting}
  Note that for the identity relation \is{I}, we have \is{\<Down> I m = m}.

  \begin{example}
    As a simple but constructed example, consider the specification of a program that returns some set of numbers divisible by \is$n$:
    \begin{lstlisting}
      dvd_set_spec n == assert (n>0); SPEC S. \<forall>m\<in>S. n dvd m
    \end{lstlisting}

    A possible implementation of this program is
    \begin{lstlisting}
      dvd_set_1 n == return {n,2*n}
    \end{lstlisting}
    and we have \is{dvd_set_1 n <= dvd_set_spec n}. This restricts the number of possible results.

    Now consider the relation \is{list_set_rel} that relates distinct lists to the set of their elements:
    \begin{lstlisting}
      list_set_rel == { (l, set l) | distinct l }
    \end{lstlisting}

    We have \is{return [n,2*n] <=\<Down> list_set_rel (dvd_set_1 n)}. This changes the
    representation from sets to lists. Note that, by transitivity and monotonicity of \is{\<Down>} we also have:
    \begin{lstlisting}
      return [n,2*n] <=\<Down> list_set_rel (dvd_set_spec n)
    \end{lstlisting}
    Intuitively, this limits the number of possible results and changes their representation at the same time.
  \end{example}

  We now refine the specifications and the introsort algorithm to work on a \emph{slice} of a list rather than on the whole list.
  The relation \is{slice_rel xsr l h} relates a list with its slice from \is{l} to \is{h}. Outside the slice, the list must
  be equal to \is{xsr}:
  \begin{lstlisting}
    slice_rel xsr l h == {(xs, xs[l..<h]) | l <= h \and h <= |xs|
      \and |xs|=|xsr| \and xs[..<l] = xsr[..<l] \and xs[h..] = xsr[h..]
    }
  \end{lstlisting}
  We then define:
  \begin{lstlisting}
    slice_sort_spec xs l h == \<Down> (slice_rel xs l h) (
      assert (l <= h \and h <= |xs|);
      sort_spec (xs[l..<h]) )
  \end{lstlisting}
  That is, for valid ranges \is{l,h}, we return a list where the range is sorted, and the rest is unchanged.
  Partial sorting of a slice is specified analogously.

  Similarly, we refine the specification for partitioning:
  \begin{lstlisting}
    slice_partition_spec xs l h == \<Down>({((xs',m),(xs_1,xs_2)).
      (xs',xs_1@xs_2)\<in>slice_rel xs l h \and m = l + |xs_1| })
      (assert (l<=h \and h<=|xs|); partition_spec (xs[l..<h])})
  \end{lstlisting}
  This returns a pair of a list \is{xs'} and an index \is{m}.
  This list \is{xs'} contains the two partitions \is{xs_1@xs_2} at range \is{l}, \is{h},
  and is equal to the input list \is{xs} outside this range. Moreover, the index \is{m}
  is the index where the second partition starts in \is{xs'}.

  Having defined the concrete specifications, we define a more concrete version of the Introsort algorithm:
  \begin{lstlisting}
    slice_partial_introsort xs l h d ==
        assert (l<=h);
        if h-l > threshold then {
            if d=0 then slice_sort_spec xs l h
            else {
                (xs,m) <- slice_partition_spec xs l h;
                xs <- slice_partial_introsort xs l m (d-1);
                xs <- slice_partial_introsort xs m h (d-1);
                return xs }}
        else return xs
  \end{lstlisting}
  To show that this algorithm refines \is{slice_sort_spec}, we first show that it refines \is{partial_introsort}:
  \begin{theorem}\label{thm:slice_partial_introsort_refines}
  \begin{lstlisting}
    l<=h \and h<=|xs| ==>
      slice_partial_introsort xs l h d
      <= \<Down> (slice_rel xs l h) (partial_introsort (xs[l..<h]) d)
  \end{lstlisting}
  \end{theorem}
  The proof is straightforward: First, we use the verification condition generator of the
  Refinement Framework\footnote{we had to do a bit of manual hinting to get the recursion refinement right.}.
  This reduces the proof to a set of proof obligations on lists. With some help from the sledgehammer
  tool~\cite{BBP13}, these could be solved easily. For example, a crucial proof obligation relates the first abstract and recursively
  sorted partition with the respective slice of the concrete result:
  \begin{lstlisting}
    lemma (xs', xs_1 @ xs_2) \<in> slice_rel xs l h
        ==> xs_1 = xs'[l..<l+|xs_1|${}$]
        apply (clarsimp simp: slice_rel_def)
        by (metis slice_def add_diff_cancel_left' append.assoc ...)
  \end{lstlisting}
  The first line of the proof unfolds the definition of \is{slice_rel} and simplifies the goal.
  These simplification steps are standard, and usually don't require too much thinking.
  The second line of the proof was then automatically found by sledgehammer.

  Finally, we combine the refinement statement (Thm~\ref{thm:slice_partial_introsort_refines}) with the
  correctness result for \is{partial_introsort} (Thm~\ref{thm:partial_introsort_correct}).
  By transitivity and monotonicity of \is{\<Down>} we get correctness of \is{slice_partial_introsort}:
  \begin{theorem}\label{thm:slice_partial_introsort_correct}
  {\ }\vspace*{-.5em}

  \begin{lstlisting}
    slice_partial_introsort xs l h d <= slice_part_sort_spec xs l h
  \end{lstlisting}
  \end{theorem}
  Note how refinement allowed us to split this correctness proof into two independent parts:
  We first established the correctness of Introsort on a whole list, using a partition function that returns two separate lists.
  Although this \emph{abstract} representation is not suitable for an in-place implementation with arrays, the correctness proof is much simpler than a
  direct correctness proof on slices. The second part of the proof is independent from the abstract correctness proof.
  It just shows how to implement the abstract algorithm on the \emph{concrete} data representation.

  The main Introsort algorithm is simple enough to define it on slices directly:
  The additional overhead introduced by refinement would outweigh the
  simplification of the proof.

  We first specify the final sorting step to completely sort an already partially sorted slice\footnote{
    We will check for an empty or singleton slice, which is trivially sorted, in the main algorithm.
    Thus, the additional precondition \is{h-l > 1} allows us to omit this check in the final sorting algorithm.
  }:
  \begin{lstlisting}
    slice_final_sort_spec xs l h ==
      assert (h-l > 1 \and part_sorted (xs[l..<h]));
      slice_sort_spec xs l h
  \end{lstlisting}
  The Introsort algorithm then simply combines partial sorting and final sorting:
  \begin{lstlisting}
    introsort xs l h ==
        assert (l <= h);
        if h-l>1 then {
            xs <- slice_part_sort_spec xs l h;
            xs <- final_sort_spec xs l h;
            return xs
        } else return xs
  \end{lstlisting}
  Its correctness proof is straightforward, and we get:
  \begin{theorem}\label{thm:introsort_correct}
    \begin{lstlisting}
    introsort xs l h <= slice_sort_spec xs l h
    \end{lstlisting}
  \end{theorem}

\section{Assembling the Implementation}
  Note that our \is{introsort} algorithm itself is independent from both,
  the partial and the final sorting algorithms: It only uses
  abstract specifications of them.
  However, in order to arrive at a concrete implementation, one has
  to \emph{plug in} concrete implementations for these specifications.
  Similarly, our \is{slice_partial_introsort} algorithm depends on abstract
  specifications of partitioning and sorting (when the recursion depth limit is hit).
  Figure~\ref{fig:modular} shows the complete architecture of our Introsort implementation.
  \begin{figure}
  \begin{tikzpicture}[
    node distance=2mm,
%     algoname/.style={font=\fontsize{6}{6}\color{black!50}\ttfamily},
%     depname/.style={rectangle, draw=black!50, font=\scriptsize\ttfamily, anchor=west}
    algoname/.style={font=\scriptsize\color{blue}\ttfamily},
    depname/.style={font=\fontsize{6}{6}\color{red}\ttfamily, anchor=west}
  ]
%     \node (introsort) [algoname] { introsort };
%     \node (introsort-ps) [below=of introsort.west, depname, xshift=2mm] { part\_sort };
%     \node (introsort-fs) [below=of introsort-ps.west, depname] { final\_sort };
%     \node (introsort-box) [draw=black!50, fit={(introsort) (introsort-ps) (introsort-fs)}] {};
%
%     \node (partintrosort) [right=6mm of introsort-box.north east, yshift=-1.15mm, anchor=north west, algoname] { partial\_introsort };
%     \node (partintrosort-s) [below=of partintrosort.west, depname, xshift=2mm] { sort };
%     \node (partintrosort-p) [below=of partintrosort-s.west, depname] { partition };
%     \node (partintrosort-box) [draw=black!50, fit={(partintrosort) (partintrosort-s) (partintrosort-p)}] {};
%
%     \node (insort) [below=2mm of partintrosort-box.south west, anchor=north west, algoname,draw=black!50] { final\_insertion\_sort };
%
%     \node (heapsort) [right=6mm of partintrosort-box.north east, anchor=north west, yshift=-1mm, algoname,draw=black!50] { heapsort };
%     \node (ungrdpart) [below=2mm of heapsort.south west, anchor=north west, algoname,draw=black!50] { unguarded\_partition };
%
%     \draw[->] (heapsort.west) to[-|-=.17] (partintrosort-s.east);
%     \draw[->] (ungrdpart.west) to[-|-=.22] (partintrosort-p.east);
%
%     \draw[->] (insort.west) to[-|-] (introsort-fs.east);
%     \draw[->] (partintrosort-box.west) to[-|-] (introsort-ps.east);

    \node (introsort) [algoname] { introsort };
    \node (introsort-ps) [below=of introsort.west, depname, xshift=2mm] { part\_sort };
    \node (introsort-fs) [below=of introsort-ps.west, depname] { final\_sort };
    \node (introsort-box) [draw=black!50, fit={(introsort) (introsort-ps) (introsort-fs)}] {};

    \node (partintrosort) [right=6mm of introsort-box.north east, yshift=-1.15mm, anchor=north west, algoname] { partial\_introsort };
    \node (partintrosort-s) [below=of partintrosort.west, depname, xshift=2mm] { sort };
    \node (partintrosort-p) [below=of partintrosort-s.west, depname] { partition };
    \node (partintrosort-box) [draw=black!50, fit={(partintrosort) (partintrosort-s) (partintrosort-p)}] {};

    \node (insort) [below=3mm of partintrosort-box.south west, xshift=-1mm, anchor=north west, algoname] { final\_insertion\_sort };
    \node (insort-s) [below=of insort.west, depname, xshift=2mm] { sort\_one\_more };
    \node (insort-box) [draw=black!50, fit={(insort) (insort-s)}] {};

    \node (heapsort) [right=6mm of partintrosort-box.north east, anchor=north west, yshift=-1mm, algoname,draw=black!50] { heapsort };
    \node (ungrdpart) [below=2mm of heapsort.south west, anchor=north west, algoname,draw=black!50] { unguarded\_partition };
    \node (insert) [below=3.85mm of ungrdpart.south west, anchor=north west, algoname,draw=black!50] { insert };

    \draw[->] (heapsort.west) to[-|-=.17] (partintrosort-s.east);
    \draw[->] (ungrdpart.west) to[-|-=.22] (partintrosort-p.east);
    \draw[->] (insert.west) to[-|-=.22] (insort-s.east);

    \draw[->] (insort-box.west) to[-|-] (introsort-fs.east);
    \draw[->] (partintrosort-box.west) to[-|-] (introsort-ps.east);
  \end{tikzpicture}
  \caption{Modular verification of Introsort. The boxes show the different
    algorithms and the specifications they depend on. Arrows point from an algorithm to the specification that it implements.}\label{fig:modular}
  \end{figure}

  % TODO: Despite this explicit modularization, we also do implicit modularization, by inserting concrete algorithms
  %  early, but only using their spec for the correctness proof

\subsection{Unguarded Insertion Sort}\label{sec:unguarded_insort}
  We also use refinement for the implementation of insertion sort.
  As suggested in the pseudocode (Alg.~\ref{alg:insort}), we use a Boolean variable to discriminate
  between the guarded and unguarded versions. Only in the last refinement step we will specialize the algorithm for both cases.

  For the correctness proof, we first specify insertion as sorting one more element of an already sorted list:
  \begin{lstlisting}
    sort_one_more_spec G xs i ==
      assert (i<|xs| \and sorted (xs[0..<i]));
      assert (\<not>G --> 0<i \and \<not>(xs[i] < xs[0]));
      spec xs'. inres (slice_sort_spec xs 0 (i+1)) xs'
                \and (\<not>G --> xs'[0] = xs[0]))
  \end{lstlisting}
  Assuming the initial segment of the list up to \is{i} is already sorted,
  afterwards, the segment up to \is{i+1} will be sorted, and the rest of the list
  will be unchanged. We reuse \is{slice_sort_spec} for this (\is{inres m x} states that \is{x} is a possible result of \is{m}).
  In the unguarded case (\is{\<not>G}), we assume that the first element of the list to be sorted is not greater than
  the element to be inserted, which guarantees that insertion will stop latest at this element. Moreover, we explicitly guarantee
  that the first element of the list will not change\footnote{This is stronger than required, as in the general case of weak orderings,
  it would be fine to exchange the first element by an equivalent one. However, it holds for our implementation and slightly simplifies the proofs.}.

  Using this specification, and its canonical refinement to slices, it is straightforward to prove correct the final
  insertion sorting algorithm (cf.~Alg.~\ref{alg:insort}). We now focus on the actual insertion procedure.
  In a first step, we refine the specification to a version that describes what the insertion procedure actually does,
  rather than focusing on the sorting:
  \begin{lstlisting}
    insert_spec G xs i ==
      assert (i<|xs| \and (\<not>G --> 0<i \and \<not>(xs[i] < xs[0])));
      spec xs'. \<exists>i'<=i.
        xs'=xs[0..<i'] @ xs[i] @ xs[i'..<i] @ xs[i<${}$..]
      \and \<forall>j\<in>{i'<${}$..i}. xs[i] < xs'[j]
      \and i'>0 --> \<not>(xs[i] < xs'[i'-1])
      "
  \end{lstlisting}
  This precisely describes how the element at index \is{i} is moved backwards over greater elements.
  Its new index is \is{i'}, and if this is not the beginning of the list, the element at \is{i'-1} was not greater,
  such that the moving could be stopped.

  This specification refines \is{sort_one_more_spec}:
  \begin{theorem}
  \begin{lstlisting}
    insert_spec xs i <= sort_one_more_spec xs i
  \end{lstlisting}
  \end{theorem}
  The proof requires some basic arguments about relating equalities on list indices to equalities on lists and sets.
  Although straightforward in principle, it required a few auxiliary lemmas.
  Again, refinement allows us to do this reasoning in isolation, and then focus on how to implement the moving
  of the element.

  We implement the insert procedure on whole lists first, and then refine it to slices beginning at index \is$l$.
  We only display the version on slices here:
  \begin{lstlisting}
    slice_insert G xs l i ==
        x <- xs[i];
        (xs,i) <- while$_I$ (\<lambda>(xs,i). (\<not>G \or i>l) \and x<xs[i-1]) (\<lambda>(xs,i).
            xs <- xs[i:=xs[i-1]]
            return (xs,i-1) ) (xs,i)
        xs <- xs[i:=x]
        return xs
  \end{lstlisting}
%
%   In a second refinement step, the algorithm is refined to work on slices of lists
%
%
%
%
%   The insertion algorithm resembles the one displayed in Alg.~\ref{alg:insort}:
%   \begin{lstlisting}
%     insert G xs i ==
%         x <- xs[i];
%         (xs,i) <- while$_I$ (\<lambda>(xs,i). (\<not>G \or i>0) \and x<xs[i-1]) (\<lambda>(xs,i).
%             xs <- xs[i:=xs[i-1]]
%             return (xs,i-1) ) (xs,i)
%         xs <- xs[i:=x]
%         return xs
%   \end{lstlisting}
  As this is a functional program, the variables \is{xs} and \is{i} have to be threaded through the loop explicitly.
  Moreover, instead of updating a single index, we update the whole list and rebind the updated list to the same name as
  the old list (\is{xs <- xs[i:=x]}). Finally, the loop is annotated with an invariant $I$. It is completely standard
  and not displayed here.

  We easily prove that \is{slice_insert} refines \is{slice_insert_spec}, and, by transitivity,
  it also refines \is{slice_sort_one_more_spec}. Thus, we can plug it into the insertion sort algorithm, which, in turn, is plugged into the main Introsort algorithm (cf.\ Fig.~\ref{fig:modular}).

\subsection{Heapsort and Partition}
  \newcommand{\leftc}{\textrm{left}}
  \newcommand{\rightc}{\textrm{right}}

  The proofs of heapsort and partitioning follow a similar plot, and are not displayed here in full.
  We used some existing Isabelle proofs as guideline \cite{LaWi19-IMP2,Gri19-AFP,La16}.
  However, we point out one more interesting application of refinement: recall the sift-down function on heaps,
  that restores the heap property by \emph{floating down} an element\footnote{see, e.g., \cite[Ch.~6]{CLRS09} or \cite[Ch.~2.4]{SeWa11} for a description of heapsort.}. Usually, this function is presented by swapping the element with one of its children, until the heap property is restored (Alg.~\ref{alg:sift_down_swap}).
  \begin{algorithm}
    \begin{algorithmic}
    \Procedure{sift\_down}{$a,i$}
      \While{has right child $i$}
        \If{$a[\leftc~i] < a[\rightc~i]$}
          \If{$a[i] < a[\rightc~i]$}
            \State{\Call{swap}{$a[i], a[\rightc~i]$}}
            \State{$i\gets \rightc~i$}
          \Else~\Return{}
          \EndIf
        \ElsIf{$a[i] < a[\leftc~i]$}
          \State{\Call{swap}{$a[i], a[\leftc~i]$}}
          \State{$i\gets \leftc~i$}
        \Else~\Return{}
        \EndIf
      \EndWhile
      \If{has left child $i$ and $a[i] < a[\leftc~i]$}
        \State{\Call{swap}{$a[i], a[\leftc~i]$}}
      \EndIf
    \EndProcedure
    \end{algorithmic}
  \caption{Pseudocode of sift-down function with swap}\label{alg:sift_down_swap}
  \end{algorithm}
  \begin{algorithm}
    \begin{algorithmic}
    \Procedure{sift\_down\_opt}{$a',i$}
      \State{$t\gets a'[i]$}
      \While{has right child $i$}
        \If{$a'[\leftc~i] < a'[\rightc~i]$}
          \If{$t < a'[\rightc~i]$}
            \State{$a'[i]\gets a'[\rightc~i]$}
            \State{$i\gets \rightc~i$}
          \Else~\Return{}
          \EndIf
        \ElsIf{$t < a[\leftc~i]$}
          \State{$a'[i]\gets a'[\leftc~i]$}
          \State{$i\gets \leftc~i$}
        \Else~\Return{}
        \EndIf
      \EndWhile
      \If{has left child $i$ and $t < a'[\leftc~i]$}
        \State{$a'[i]\gets a'[\leftc~i]$}
        \State{$i\gets \leftc~i$}
      \EndIf
      \State{$a'[i]\gets t$}
    \EndProcedure
    \end{algorithmic}
  \caption{Pseudocode of optimized sift-down function}\label{alg:sift_down_opt}
  \end{algorithm}
  However, the element that is written to $a[\rightc~i]$ or $a[\leftc~i]$ by the swap will get overwritten in the next loop iteration.
  Thus, a common optimization to save half of the writes is to store the element to be moved down in a temporary variable, and only
  assign it to its final position after the loop (Alg.~\ref{alg:sift_down_opt}). Note that the insertion procedure of insertion sort does a similar optimization.
  However, for the insertion procedure, it was feasible to prove the optimization together with the actual algorithm.
  For the slightly more complicated sift-down procedure, we first prove correct the simpler algorithm with swaps,
  and then refine it to the optimized version. Inside the loop, the refinement relation between the concrete array \is{a'} and the
  abstract array \is{a} states that
  we obtain \is{a} as \is{a'[i:=t]}. The proof that the optimized version refines the version with swaps requires only about 20
  lines of straightforward Isabelle script.

\section{Imperative Implementation}
  In the previous sections, we have presented a refinement based approach to verify the Introsort algorithm,
  including most optimizations we found in its GNU C++ Library implementation. However, the algorithm is still expressed
  as a nondeterministic monadic program that modifies functional lists. In this section, we discuss how the Isabelle LLVM framework~\cite{La19-llvm} is
  used to (semi-)automatically transfer the algorithm into an LLVM program on arrays.

  \subsection{The Sepref Tool}
  The idea of the Sepref tool~\cite{La15,La19-llvm} is to symbolically execute an abstract program in the \is{nres}-monad,
  keeping track of refinements of every abstract variable to a concrete representation that may use pointers to dynamically allocated memory.
  During this symbolic execution, the tool can synthesize an imperative program in the Isabelle-LLVM semantics, together with a refinement proof.
  The tool works automatically, but usually requires some program-specific setup and boilerplate. For a detailed discussion of the Sepref tool and the Isabelle-LLVM framework, we refer the reader to~\cite{La15,La19-llvm}.

  Isabelle-LLVM comes with standard setup to refine lists to arrays. List update operations are refined to destructive array element updates,
  as long as the old version of the abstract list is not used any more after the update. Note that, by rebinding the updated list to the same
  variable name as the original list\footnote{as e.g.\ in \is{xs <- xs[i:=x]}, cf.\ Sec.~\ref{sec:unguarded_insort}}, this property is syntactically ensured. Moreover, Isabelle-LLVM provides setup to refine the unbounded integers of Isabelle/HOL to bounded integers.
  The resulting proof obligations to show that the integers are in bounds are discharged automatically, but typically require some hinting from the user.
  A common technique to provide such hints is to insert additional assertions into the abstract program.
  Usually, these can be proved easily. For example, in the definition of \is{slice_partial_introsort} (Sec.~\ref{sec:refinement}),
  we have inserted the assertion \is{l<=h} at the very beginning. This is required by the Sepref tool to prove that the operation \is{h-l}
  in the next line cannot underflow. This assertion adds a proof obligation to the proof that \is{slice_partial_introsort} refines \is{slice_part_sort_spec} (Thm.~\ref{thm:slice_partial_introsort_correct}), which is, however, trivial, as the specification
  explicitly states \is{l<=h} as precondition. On the other hand, when refining \is{slice_partial_introsort} to an implementation with bounded
  integers, one can assume \is{l<=h}, thus discharging the non-underflow proof obligation for the \is{h-l} operation.
  Thus, assertions provide a convenient tool to pass properties down the refinement chain.

  Using the Sepref tool, it is straightforward to refine the Introsort algorithm to an Isabelle-LLVM program that sorts arrays of integers.
  For example, Figure~\ref{fig:ug_insort_isa_llvm} shows the Isabelle-LLVM code that is generated for the \is{insert} procedure with \is{G=false} (cf.~\ref{sec:unguarded_insort}).
  Moreover, the Sepref tool will generate a theorem that the generated program actually implements the abstract one:
  \begin{theorem}
    \begin{lstlisting}
      (ug_insert_impl, slice_insert False)
        : arr_u64$^d$ \x snat64$^k$ \x snat64$^k$ -> arr_u64
    \end{lstlisting}
  \end{theorem}
  This specifies the relations for the parameters and the result, where \is{arr_u64} relates arrays of unsigned 64-bit integers with lists of integers,
  and \is{snat64} relates non-negative 64-bit integers with natural numbers.
  The $\cdot^d$ annotation means that the parameter will be \emph{destroyed} by the function call,
  while $\cdot^k$ means that the parameter is \emph{kept}.
  Here, the insertion is done in place, such that the input list has no corresponding refinement after the function call.

  The final correctness statement for our Introsort implementation (for integers) is
  \begin{theorem}
    \begin{lstlisting}
      (introsort_impl, slice_sort_spec)
        : arr_u64$^d$ \x snat64$^k$ \x snat64$^k$ -> arr_u64
    \end{lstlisting}
  \end{theorem}
  Here, \is$introsort_impl$ is the Isabelle-LLVM program generated by Sepref from \is$introsort$ (cf.\ Sec.\ \ref{sec:refinement}).
  This theorem combines the refinement theorem generated by Sepref, and Theorem~\ref{thm:introsort_correct}.

  \begin{figure}
    \begin{lstlisting}
      ug_insert_impl == %xs l i. doM {
          x <- array_nth xs i;
          (xs, i) <- llc_while (%(xs, i). doM {
              bi <- ll_sub i 1;
              t <- array_nth xs bi;
              ll_icmp_ult x t
          }) (%(xs, i). doM {
              i' <- ll_sub i 1;
              t <- array_nth xs i';
              xs <- array_upd xs i t;
              i <- ll_sub i 1;
              return (xs, i)
          }) (xs, i);
          array_upd xs i x
        }
    \end{lstlisting}
    \caption{Implementation for \is{insert} procedure with \is{G=false} generated by the Sepref tool, for 64bit signed integer elements.
      This definition is within the executable fragment of Isabelle-LLVM, i.e.,
      the Isabelle LLVM code generator can, after preprocessing, translate it to LLVM-IR. Moreover, note that the function does not
      depend on the lower bound parameter \is$l$ any more, as this was only required in the guarded version.
    }\label{fig:ug_insort_isa_llvm}
  \end{figure}

  \begin{figure}
    \begin{lstlisting}[language=llvm,literate={}]
      define i64* @ug_insert_impl(i64* %xs, i64 %l, i64 %i) {
      start:
        %x3 = getelementptr i64, i64* %xs, i64 %i
        %r = load i64, i64* %x3, align 8
        br label %while_start

      while_start:
        %bib.sink = phi i64 [ %bib, %while_body ], [ %i, %start ]
        %bib = add i64 %bib.sink, -1
        %xda = getelementptr i64, i64* %xs, i64 %bib
        %ra = load i64, i64* %xda, align 8
        %xg = icmp ult i64 %r, %ra
        %p1 = getelementptr i64, i64* %xs, i64 %bib.sink
        br i1 %xg, label %while_body, label %while_end

      while_body:
        store i64 %ra, i64* %p1, align 8
        br label %while_start

      while_end:
        store i64 %r, i64* %p1, align 8
        ret i64* %xs
      }
    \end{lstlisting}
    \caption{LLVM-IR generated by Isabelle-LLVM for the definition from Figure~\ref{fig:ug_insort_isa_llvm}.
    We display the code after an LLVM optimization pass that eliminates some redundant operations generated by Isabelle-LLVM,
    and thus makes the code more readable. Note that we kept the unused parameter \%l. LLVM will eliminate it by inlining the function.
    }\label{fig:ug_insort_llvm}
  \end{figure}

  \subsection{The Isabelle-LLVM Code Generator}
  Isabelle-LLVM comes with a code generator that exports this program to actual LLVM text, which is subsequently compiled using the LLVM toolchain.
  Fig.~\ref{fig:ug_insort_llvm} shows the code generated for the \is{slice_insert} procedure.

  To make the generated programs usable, one has to link them to C wrappers, that handle parsing of command line options and printing of results.
  However, the original Isabelle-LLVM framework provides no support for this: one has to manually write a C header file, that hopefully matches the object file generated by the LLVM compiler. If it doesn't, the program has undefined behaviour\footnote{In practice, this means it will probably SEGFAULT. However, it also might return wrong results, or be prone to various kinds of exploits.}. What makes the situation even worse is, that the C compiler sometimes changes function signatures: For example, a function with a non-primitive return value (e.g. a structure) is changed to a void function with an extra
  reference parameter for the return value. LLVM, on the other hand, does not perform this change, which makes the function generated by LLVM incompatible with the function call generated by the C compiler.

  To this end, we extended Isabelle-LLVM to also generate a header file for the exported functions, refusing to export functions whose signatures
  would be altered by the C compiler. Figure~\ref{fig:export} shows the extended Isabelle LLVM export command and the generated header file for our Introsort algorithm on strings.
  \begin{figure}
    \begin{flushleft}\bfseries\color{black!50} Isabelle command\end{flushleft}
    \begin{lstlisting}
      export_llvm
        str_sort_introsort_impl
          is llstring* str_introsort(llstring*, int64_t, int64_t)
        defines \<open>
          typedef struct {
            int64_t size; struct {int64_t capacity; char *data;};
          } llstring; \<close>
    \end{lstlisting}
    \medskip

    \begin{flushleft}\bfseries\color{black!50} Generated C header\end{flushleft}
    \begin{lstlisting}[language=C]
      typedef struct {
        int64_t size;
        struct {
          int64_t capacity;
          char*data;
        };
      } llstring;

      llstring* str_introsort(llstring*, int64_t, int64_t);
    \end{lstlisting}

    \caption{An example Isabelle command to export LLVM code for a string sorting function, and the generated header file.
    Strings are represented as dynamic arrays of characters.
    Isabelle will refuse to export functions whose signatures
    would be changed by C. Moreover, structure fields in LLVM are indexed by consecutive numbers, rather than named by meaningful names.
    Our extension allows to define properly named structures for the header file, and checks that these definitions are actually compatible with
    the LLVM signatures.
    }\label{fig:export}
  \end{figure}

  \subsection{Separation Logic and Ownership}
  Internally, the Sepref tool represents the symbolic state, which contains all abstract variables and their refinements to concrete variables,
  as an assertion in separation logic~\cite{Rey02,CHY07}. For example, the assertion
  \begin{lstlisting}
    array xs_1 p_1 * array xs_2 p_2
  \end{lstlisting}
  describes a symbolic state that refines the two abstract list variables \is{xs_1} and \is{xs_2} by
  two arrays at addresses \is{p_1} and \is{p_2}. Separation logic guarantees that the arrays at \is{p_1} and \is{p_2}
  share no common memory, i.e.,
  updates to the first array cannot change the second array. As long as the array elements are
  primitive values like integers, this works as expected, and a list of integers can be refined
  to an array of integers. An array lookup operation will simply return a copy of the integer
  value at the specified index. It can be specified by the following Hoare-triple, which can
  be used by Sepref to synthesize LLVM code for the operation \is{r <- xs[i]}.
  \begin{lstlisting}
    { array xs p * i<|xs| } r = p[i] { array xs p * r=xs[i] }
  \end{lstlisting}
  Note that we use C like notation for the program, and abstracted away the fact that the \is{i}
  in the program is a bounded integer register from LLVM, while the \is{i} in the assertions refers to
  its content as an unbounded Isabelle integer.

  Now, let's assume we have a list of strings, where a string is itself a list of characters.
  The following assertion refines both, the outer and the inner lists to arrays:
  \begin{lstlisting}
    sarray xs p == \<exists>ys.
      |ys|=|xs| * array ys p * array xs_1 ys_1 * ... array xs$_n$ ys$_n$
  \end{lstlisting}
  Here, \is{ys} represents the list of string pointers, which point to their own separate array each.
  The canonical Hoare triple for array lookup would be:
  \begin{lstlisting}
    { sarray xs p * i<|xs| } r = p[i] { sarray xs p * array (xs[i]) r }
  \end{lstlisting}
  However, this does not hold, as the inner array at position \is{i} is now pointed to twice:
  by \is{r} and by the \is{ys[i]} from within the \is{sarray} assertion. The only way to make a
  Hoare triple of this form valid is to make a copy of the inner array, which would come with an
  unacceptable performance penalty.

  The Rust Programming Language~\cite{rustlang} has to solve a similar problem.
  There, they use the concept of borrowing: the inner array is \emph{borrowed} from the outer one,
  i.e., ownership is transferred to the variable \is{r}. Only at the end of the lifetime of \is{r},
  ownership will return to the outer array\footnote{Actually, borrowing in Rust does not work for array elements,
  as the type-checking would become undecidable. Fortunately, in an interactive theorem prover, we can burden the
  user with parts of this type checking.}. We realize a similar concept in Sepref, by refining abstract lists of
  optional values to arrays, that only own those elements which are present (not None) in the abstract list\footnote{For simplicity, the presentation sticks to the array of strings example. Actually, our implementation is parameterized over the refinement for the inner type.
  We believe that a further generalization to include outer types other than lists (e.g.\ maps) is possible, but leave it for future work.}:
  \begin{lstlisting}
    oarray xs p == \<exists>ys. array ys p * A' xs_1 ys_1 * ... * A' xs$_n$ ys$_n$
    $\textrm{where }$
    A' None _ = \<box>  |  A' (Some x) y = array x y
  \end{lstlisting}
  We define abstract operations to borrow and return (insert) elements, as well as to convert between a
  standard list and a list with explicit ownership:
  \begin{lstlisting}
    borrow xs i == assert (i<|xs| \and xs[i]~=None);
                  return (the (xs[i]), xs[i:=None])
    ins xs i x == assert (i<|xs| \and xs[i]=None);
                  return xs[i:=Some x]
    to_olist xs = return (map Some xs)
    to_slist xs = assert (None \<notin> xs); return (map the xs)
  \end{lstlisting}
  Here, \is{the (Some x) = x} is the selector function of the option data type.
  The \is{borrow} function assumes that the element is still present in the list, and then returns the element
  and a new list where the element is gone. Symmetrically, the \is{ins} function assumes that there is no element
  present at the specified index. The \is{to_olist} and \is{to_slist} functions convert between lists of plain values and lists
  of optional values, where \is{to_slist} assumes that all values in the list are actually present.

  This approach allows us to explicitly model ownership transfer of array elements in the \is{nres} monad, and let Sepref
  generate efficient code, sparing explicit copies of elements.

  The necessary conversion operations can usually be inserted by a straightforward refinement step.
  For example, we refine the \is{insert} function (Sec.~\ref{sec:unguarded_insort}) as follows:
  \begin{lstlisting}
    is_insert_o G xs' i ==
      xs' <- to_olist xs';
      (x,xs') <- borrow xs' i;
      (xs',i) <- while
        (%(xs',i). (\<not>G \or i>0) \and xs' < xs'[i-1]) (%(xs',i).
          (t,xs') <- borrow xs' (i-1);
          xs' <- ins xs' i t;
          return (xs', i-1)
        ) (xs',i);

      xs' <- ins xs' i x;
      return (to_slist xs')
    }
  \end{lstlisting}
  First, this function converts the input list \is{xs'} to one with explicit ownership.
  It then borrows the element \is{i} from this list. In each loop iteration, it moves an element
  from \is{i-1} to \is{i}, by borrowing and immediately re-inserting it.
  Finally, the element borrowed at the beginning is re-inserted, and the list is converted back to implicit ownership.
  The refinement proof is, again, straightforward. The refinement relation for the loop states that the abstract and concrete array
  contain the same values, except at index \is{i}, where the concrete value is not present:
  \begin{lstlisting}
    |xs'| = |xs|
    \and (\<forall>j<|xs|. j~=i --> xs'[j] = Some (xs[j])) \and xs'[i]=None}
  \end{lstlisting}

  Using this explicit ownership extension to Sepref, we can also synthesize the Introsort algorithm for arrays of arrays of characters,
  where the inner arrays are compared lexicographically.

%   xxx, ctd here: Define slice_partition_spec the same way! Then show refinement of algo!
%     lemma "partition3_spec xs l h = ⇓(slice_rel xs l h ×⇩r Id) (doN { ASSERT (h-l≥4 ∧ h≤length xs); (xs⇩1,xs⇩2) ← partition1_spec (slice l h xs); RETURN (xs⇩1@xs⇩2, l+length xs⇩1) })"
%
%
%
%   xxx, ctd here: slicep_rel and slice_rel. Do we need two steps?
%
%

\section{Benchmarks}
  The goal of this paper was to verify
  an existing real-world implementation of a medium-complex algorithm.
  We chose the Introsort implementation from the GNU C++ library. As we did not directly verify the existing C++ code,
  but re-implemented the algorithm using Isabelle-LLVM, we have to test how close we came to the performance of the original.

  Our benchmark set is based on the Boost C++ Library's~\cite{boost} sorting algorithm benchmarks, extended with further benchmarks indicated in~\cite{Bent93}.
  Apart from sorting a random list of elements that are mostly different (random), we also sort lists of length $n$ that contain only $n/10$ different elements (random-dup-10), and random lists of only two different elements (random-boolean), as well as lists where all elements are equal (equal).
  We also consider already sorted sequences (sorted, rev-sorted), as well as a sequence of $n/2$ elements in ascending order, followed by the same elements in descending order (organ-pipe). We also consider sorted sequences where we applied $pn/100$ random swap operations (almost-sorted-$p$).
  Finally, we consider sorted sequences with $pn/100$ random elements inserted at the end or in the middle ([rev-]sorted-end-$p$, [rev-]sorted-middle-$p$).

  We sorted integer arrays with $n=10^8$ elements, and string arrays with $n=10^7$ elements.
  To represent strings, we used dynamic arrays of characters, and used the same compare function for our algorithm and std::sort,
  in order to eliminate differences in the compare function implementation\footnote{Initially, we used std::string elements for std::sort,
  for which the compare function seems to be significantly slower than our compare function on dynamic arrays.}.

  We compile both, our verified algorithm and the std::sort benchmark with clang-6.0.0.
  The benchmarks were run on a laptop with an Intel(R) Core(TM) i7-8665U CPU and 32GiB of RAM, as well as on a server machine with
  24 AMD Opteron 6176 cores and 128GiB of RAM. Ideally, the sorting algorithm should take exactly the same time when run on the same data and machine.
  However, in practice, we encountered some noise. We encountered both, \emph{high frequency} noise, that slowed down single sorting
  runs, and \emph{low frequency} noise, that slowed down the machine for several minutes. To counter the high-frequency noise,
  we ran each experiment 10 times in a row, dropped the 6 extremal results, and report the average of the remaining 4 results.
  If the smallest remaining result was more than 2\% smaller than the largest remaining result, we repeated the experiment.
  Countering the low-frequency noise is more time consuming. For all benchmarks, we repeated the slower experiment when the
  difference between our algorithm and std::sort was more than 15\%\footnote{The outlier for sorted-end-1 on the server machine seems to be genuine. We have no explanation for that.}. Moreover, for the random and organ-pipe integer benchmarks,
  we ran all the experiments in a loop for several hours, and report the intervals of the averages of the median 4 values.
  The results are displayed in Tables~\ref{tab:benchmarks} and ~\ref{tab:bench_longrun}.

  \begin{table}{\footnotesize
    \begin{tabular}{|l|S|S||S|S|}
      \hline
      \multirow{2}{*}{Data Set}&\multicolumn{2}{c||}{uint64}&\multicolumn{2}{c|}{llstring}\\
      & Lap & Srv & Lap & Srv \\
      \hline
      random & -0.5 & 2.4&-14.5 & -2.1\\
      random-dup-10 & -0.2 & 3.2&-1.2 & -1.7\\
      random-boolean & -0.1 & -0.7&4.7 & 5.3\\
      equal & -13.3 & 0.9&6.9 & 4.0\\
      sorted & 6.6 & 3.0&2.2 & 4.2\\
      rev-sorted & 1.9 & 3.8&2.1 & 2.0\\
      organ-pipe & -8.1 & 12.8&7.0 & 4.9\\

      almost-sorted-10 & 3.8 & 1.3&-0.8 & -2.0\\
      almost-sorted-.1 & 3.2 & 1.7&2.3 & -0.7\\
      almost-sorted-1 & -8.7 & 2.2&-11.1 & -1.1\\
      almost-sorted-50 & -0.0 & 2.5&-12.2 & -1.8\\

      sorted-end-10 & -4.3 & 1.3&-1.0 & -1.4\\
      sorted-end-.1 & -4.0 & 8.3&-4.7 & 11.9\\
      sorted-end-1 & -5.2 & 19.9&10.4 & 5.6\\
      sorted-middle-10 & -1.1 & 3.2&-0.7 & -2.7\\
      sorted-middle-1 & 12.5 & 4.5&5.4 & 2.1\\
      sorted-middle-.1 & -2.8 & 4.4&6.1 & 6.6\\

      rev-sorted-end-10 & 1.1 & 3.1&-1.5 & -2.4\\
      rev-sorted-end-.1 & 5.6 & 8.7&0.9 & 5.4\\
      rev-sorted-end-1 & 9.7 & 3.9&-12.6 & 0.6\\
      rev-sorted-middle-10 & 1.2 & 6.2&-14.3 & -0.6\\
      rev-sorted-middle-1 & -2.3 & 3.6&-12.0 & 0.9\\
      rev-sorted-middle-.1 & -5.7 & 4.2&3.4 & 5.5\\
      \hline
    \end{tabular}}
    \caption{Benchmark results, as the ratio $(i/c - 1)*100$, where $i$ and $c$ are the average wall-times of verified Introsort and std::sort.
    That is, negative values indicate that Introsort was faster, and positive values indicate that std::sort was faster.
    }\label{tab:benchmarks}
  \end{table}
  \begin{table}\footnotesize
    \begin{tabular}{|l|SSS|SSS|}
      \hline
      \multirow{2}{*}{Algorithm} & \multicolumn{3}{c|}{random} & \multicolumn{3}{c|}{organ-pipe} \\
      & {min} & {max} & {r} & {min} & {max} & {r} \\
      \hline
      isabelle::sort &7.35 & 8.58 & 16.7 & 8.43 & 9.86 & 17.0 \\
      std::sort &7.38 & 8.63 & 17.0 & 8.97 & 10.49 & 16.9 \\
      \hline
    \end{tabular}

    \caption{Intervals of the results (wall time in seconds) for repeating each experiment 74 times,
      in an interleaved fashion. The $r$ fields display $(\max/\min-1)*100$, i.e., the value that
      would appear in Table~\ref{tab:benchmarks} when comparing the slowest and fastest run of the same experiment.
      }\label{tab:bench_longrun}
  \end{table}

  The results clearly show that our verified sorting algorithm performs on-par with the GNU C++ Library's implementation,
  within the expected noise of up to $17\%$. Table~\ref{tab:bench_longrun} even indicates that our algorithm performs
  slightly better on our laptop machine. We leave it to future work to do more precise measurements on different machines.

%   \subsection{Proof Architecture of Introsort}
%     Start with introsort-scheme. Motivate stepwise refinement "in the small" to get
%     from list-concatenation to slicing.
%
%     Then illustrate big-picture architecture. Mention that we use similar "in the small" techniques for other modules.
%
%     Finally show how final algorithm is assembled (nres-level)
%
%     THEN introduce sepref tool, and illustrate how final algorithm is refined, module by module.
%     mention parameterization over sort function.
%
%     Can we motivate the array-ownership stuff?
%
%
%
%
% \section{Isabelle-LLVM}
%   Quick intro to Isabelle-LLVM
%
%   \subsection{Nested Containers and Ownership}
%     Our prototype solution
%     Rust's ownership model
%     Outlook: Fractional separation logic for read-only borrowing
%     Still unsolved: Abstractly immutable, but concrete mutations
%       (E.g. union-find with path compression, amortized search-trees?) Invalid in concurrent setting, but perfectly OK in sequential setting!
%
%
%
%   \subsection{Isabelle-LLVM Code Generator}
%   Small TCB.
%
%
% \section{Verification of Introsort}
%
%   Layout general architecture (Quicksort-Scheme)
%   Main proof steps and concepts:
%     Partially sorted
%     Guards
%     Plugging together the implementation
%
%
%   Observations
%     sledgehammer tool is really useful. But prone to not properly abstract/modularize the proofs,
%       which will only work until sledgehammer can't solve a goal. At this point, usually
%       refactoring of the whole proof is required to make it manageable again.
%
%
%
%
% \section{Benchmarks}
%   64 bit integers,
%   character arrays (we don't have a std::string implementation in Isabelle-LLVM yet)
%
%   benchmark suite taken from BOOST. Added some more test cases from Timsort(?) paper.
%
%
%   Benchmark the heapsort-algorithm alone!
%
%
%   Add Test:
%     Organpipe  123..nn.. 321
%     Random 0-1 arrays

\section{Conclusions}
In this paper we have verified the state-of-the art Introsort sorting algorithm using Isabelle-LLVM.
We included most optimizations found in the GNU C++ Library's (libstdc++) implementation.
On an extensive set of benchmarks, our verified implementation performs on par with the libstdc++ implementation.

The verification took us about 100 person hours distributed over 4 weeks. After two weeks and roughly 60 person hours,
we had arrived at an algorithm that was only about 10\% slower than std::sort. However, it only worked for integers, and always
used guarded insertion sort. The remaining time was spent on implementing explicit ownership for Sepref, such
that we could sort strings and other objects, and on verifying the unguarded insertion sort optimization.

% Verification Effort: ca 100 person-hours, including benchmarking and extension of Isabelle-LLVM to support nested containers.
% ca 60 person-hours to arrive at first integer sorting algorithm that was competitive [this time, however, includes heapsort implementation, which is not strictly necessary for competitive algorithm, but was done quite early in the project].
% XXX lines of proof.

This shows that Isabelle-LLVM can be used to generate competitive verified implementations of medium-complex algorithms.
The Refinement Framework has been successfully used for much larger projects, e.g., model-checkers~\cite{ELNN13,BrLa18,WiLa18}.
However, these developments produce implementations that use arbitrary precision integers and target Standard ML,
which means a principal performance deficit compared to more low-level implementations. Using Isabelle-LLVM as backend to the Refinement Framework,
these limitations can now be overcome. Thanks to the modular structure of the Refinement Framework, we can even hope to gradually
transform existing verification projects to Isabelle-LLVM, e.g., replacing invocations of functional mergesort on lists by
our Introsort algorithm on arrays in a first step.

While we can easily instantiate our sorting algorithm to arbitrary weak orderings on arbitrary element types \emph{within} the Refinement Framework,
we cannot so easily replace C++'s std::sort. The reason is that std::sort is a template which is instantiated by the C++ compiler,
i.e., before LLVM intermediate code is even generated. Verifying the std::sort template itself would require a formal semantics of C++,
including its template mechanism and the relevant traits from the standard template library (orderings, iterators, etc.). To the best of our knowledge, such a semantics has not yet been formalized, let alone being used to verify non-trivial algorithms.
However, we can \emph{specialize} the std::sort template to use our Introsort algorithm for any fixed combination of element type and ordering, on containers backed by an array (e.g.\ std::vector).

Finally, the trusted code base of Isabelle-LLVM is relatively small, even compared to other common program verification methods.
Following the LCF tradition~\cite{gordon-et-al-1979}, the Isabelle prover is designed around a small inference kernel, which is the only code
whose bugs can affect the correctness of proved theorems.
The Isabelle-LLVM framework contains a relatively straightforward formalization of the LLVM semantics, with a straightforward code
generator~\cite{La19-llvm} to LLVM intermediate language.
Finally, the LLVM compiler is an industrial strength, well-tested, and widely used compiler.
In contrast, the default code generators of most theorem provers target high level languages, which includes the
compilers and the high-level language semantics into the trusted code base.
Deductive verification tools like KeY~\cite{BHS07} depend on the correct axiomatization of the highly complex Java semantics,
as well as on several automatic theorem provers, which, themselves, are highly complex and optimized C programs.

To reduce the trusted code base even further, one can include a compiler correctness proof in the theorem prover:
The Verifiable C project~\cite{App14} provides a verification infrastructure on top of the verified CompCert compiler~\cite{BL09}.
However, to the best of our knowledge, this has only been applied to verify rather simple programs~\cite{CBGD18}.
The CakeML project \cite{KTMK19,MyOw14} has developed a code generator for HOL4 and a verified ML compiler.
However, the performance of functional high-level languages seems to be inherently limited wrt.\ low-level imperative languages,
and we are not aware of any high-performance algorithm that has been verified using CakeML.

\subsection{More Related Work}
Sorting algorithms are a standard benchmark for program verification tools, such that we cannot give an exhaustive overview.
Nevertheless, we discuss a few notable examples: The arguably first formal proof of quicksort was given
by Foley and Hoare himself~\cite{FoHoa71}, though, due to the lack of powerful enough theorem provers at these times,
it was only done on paper.

One of the first mechanical verifications of imperative sorting algorithms is by
Filli{\^a}tre and Magaut~\cite{FiMa99}, who prove correct imperative versions of quicksort, heapsort, and insertion sort in Coq.
However, they do not report on code generation or benchmarking, nor do they combine their separate algorithms to Introsort.

The Timsort algorithm, which was used in the Java standard library, has been verified
with the KeY tool~\cite{GRBBH15}. A bug was found and fixed during the verification.
Subsequently, KeY has been used to also verify the dual-pivot quicksort algorithm from the
Java standard library~\cite{BSSU17}. This time, no bugs were found.

\subsection{Future Work}
Actually, the C++ standard requires the std::sort algorithm to not only be functional correct,
but to also have a worst-case complexity of $O(n\log n)$. Recently, the Refinement Framework has been extended to reason
about time complexity~\cite{HaLa19}, and it has been proposed that Isabelle-LLVM could be used as a backend.
This would enable us to also formally prove the worst-case complexity of Introsort.

Another obvious direction is to formalize other state-of-the-art implementations of sorting algorithms,
for example std::stable\_sort, or Boost's pattern defeating quicksort~\cite{boost} with branch prediction optimization~\cite{EdWe16}.

Finally, we proposed an explicit ownership model for nested lists.
However, we only support write-ownership. A next step would be to include read-ownership, which allows aliasing on read-only objects.
Formally, this could be realized with fractional permissions~\cite{BCOP05}.

%
% Future work:
%   Proof complexity (Use refinement-with-time)
%   More optimizations (eg PD-Qsort)
%   Specialized algorithms for integers and strings (radix sort, burstsort, etc ... -> Boost library)

\paragraph{Acknowledgements}Hidden for blind review.
%
% We received funding from DFG grant LA 3292/1 "Verifizierte Model Checker" and VeTSS grant "Formal Verification of Information Flow Security for Relational Databases".

%
%
% The Compcert project~\cite{compcert} has formalized a correct compiler for C to machine code
%
% This has been done in Coq for a fragment of the C language
%
%
%
% Most other algorithms that are verified with interactive theorem provers such as Isabelle, Coq~\cite{XX}, or ACL2 \cite{XX}
%
%
% xxx: CompCert, CakeML --> actually smaller code-base
%
%
%
%
% Apart from the obvious things such as the machine and operating system,
% and the Theorem Provers kernel
%
%
%
%
%
%
%
%
%
%
%   As our verification targets LLVM, we cannot directly replace std::sort, which is a C++ template.
%   However, we could provide template specializations for std::sort that direct to our verified
%   sorting algorithm for certain data types. A direct verification of std::sort would require a formal verification framework
%   for C++ including templates, which, to the best of our knowledge, does not exist.
%
%   Inside the Isabelle-LLVM framework, however, our algorithm can easily be instantiated to any data structure that implements
%   random access, and to any element type that implements a weak ordering.
%
%
%
%
%
%
%
% makes the achievable performance significantly worse than what can be done
%
%
% only went down to either functional code, or the imperative fragment of Standard ML,
%
%
%
%
% --> mention goal, achieved!
%
% --> mention that Isabelle provides means to use this as building block for
%   even more complex algorithms, mention CAVA, Timed-Automata MC
%
%
% Our verified implementation contains most optimizations that we
%
%
%
% Verified state-of-the art sorting algorithm implementation. On par with original from libstdc++.
%
% top-down vs. bottom up. Verification of C++ would be different game.
% TCB! ours is smaller, as it excludes C++ compiler.
% Drawback: Not as universally usable from C++, b\c not integrated into STL.
%
%
%
%
% Verification Effort: ca 100 person-hours, including benchmarking and extension of Isabelle-LLVM to support nested containers.
% ca 60 person-hours to arrive at first integer sorting algorithm that was competitive [this time, however, includes heapsort implementation, which is not strictly necessary for competitive algorithm, but was done quite early in the project].
% XXX lines of proof.
%
% Extended Isabelle-LLVM:
%   Explicit ownership on nres-level
%   Header file generation
%
%
% Usability:
%   As our verification targets LLVM, we cannot directly replace std::sort, which is a C++ template.
%   However, we could provide template specializations for std::sort that direct to our verified
%   sorting algorithm for certain data types. A direct verification of std::sort would require a formal verification framework
%   for C++ including templates, which, to the best of our knowledge, does not exist.
%
%   Inside the Isabelle-LLVM framework, however, our algorithm can easily be instantiated to any data structure that implements
%   random access, and to any element type that implements a weak ordering.
%
%
%
% Related Work:
%   Filliatre COQ 2012: Insertion sort, quicksort, heapsort
%     no combination to introsort
%     not so many optimizations
%     imperative. Executable? (They do not report on benchmarks)
%
%   Timsort verification in KeY
%     verifies the Java implementation (after slight modifications, mainly stripping generics)
%       This is a bottom-up approach, while ours is a top-down approach.
%       The trusted code-base of KeY is much larger, in particular, it includes SMT solvers.
%
%
%
%     There's also a follow-up on dual pivot quicksort??
%
%       Proving JDK’s Dual Pivot Quicksort Correct
%       Bernhard Beckert, Jonas Schiffl, Peter H. Schmitt, and Mattias Ulbrich
%
%         Bottom up. Refactored the code to make it more prover-friendly.
%
%
%
%
%     12. de Gouw S, de Boer FS, Rot J. Proof pearl: the KeY to correct and stable sorting. J. Autom. Reason. 2014;53(2):129–139. doi: 10.1007/s10817-013-9300-y. [CrossRef] [Google Scholar]
%       radix sort, counting sort
%
%
%   Why3? Verified some simple sorting algorithm. No evidence of implementations or benchmarks.
%
%   Functional quicksort: Common knowledge. Few lines of Isabelle!
%
%   Heapsort-AFP
%     not executable
%
%   Bockeneck et al.
%     Insertion sort, quicksort partitioning (Lomutu?)
%     Not executable
%
%
%
%
%
% Future work:
%   Proof complexity (Use refinement-with-time)
%   More optimizations (eg PD-Qsort)
%   Specialized algorithms for integers and strings (radix sort, burstsort, etc ... -> Boost library)

\bibliographystyle{ACM-Reference-Format}
\bibliography{bib}

\end{document}

### 2019-11-22 benchmarks/sorting/BlockQuicksort/driver.cpp

@@ -1,0 +1,650 @@
+/******************************************************************************
+* driver.cpp
+*
+* Driver to benchmark sorting algorithms. Comparison and move measurements do NOT work for all algorithms.
+* Modified version of sort-driver.c++ from http://www.diku.dk/~jyrki/Myris/Kat2014S.html
+******************************************************************************
+* Modified (more test cases etc.) by
+* Copyright (C) 2016 Stefan Edelkamp <edelkamp@tzi.de>
+* Copyright (C) 2016 Armin Weiß <armin.weiss@fmi.uni-stuttgart.de>
+******************************************************************************
+*
+* This program is free software: you can redistribute it and/or modify it
+* under the terms of the GNU General Public License as published by the Free
+* Software Foundation, either version 3 of the License, or (at your option)
+* any later version.
+*
+* This program is distributed in the hope that it will be useful, but WITHOUT
+* ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+* FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License for
+* more details.
+*
+* You should have received a copy of the GNU General Public License along with
+* this program.  If not, see <http://www.gnu.org/licenses/>.
+*******************************************************************************
+*
+* Copyright (C) 2014 Amr Elmasry
+* Copyright (C) 2014 Jyrki Katajainen
+* Copyright (C) 2014 Max Stenmark
+*
+*****************************************************************************/
+
+#if ! defined(MAXSIZE)
+#define MAXSIZE (1024 * 1024 * 1024)
+#endif
+
+
+//Maximal number of bytes to sort
+#ifndef BYTESTOSORT
+#define BYTESTOSORT (128 * 1024 * 1024)
+#endif
+
+#include <algorithm>  	
+#include <functional> 	
+#include <iostream>   	
+#include <iterator>   	
+#include <string> 
+#include <cmath> 
+#include <random>
+#include <string>
+#include <chrono>
+#include <typeinfo>
+
+
+
+//to print types
+#define QUOTE(name) #name
+#define STR(macro) QUOTE(macro)
+#define ALG_NAME STR(NAME)
+
+
[…]

### 2019-11-22 benchmarks/sorting/all_benchmarks.sh

@@ -1,0 +1,78 @@
+#!/bin/bash
+
+# make do_benchmark
+
+########################## BEGIN CONFIG SECTION ##################
+
+# Defaults
+
+export NITER=10   # Number of iterations
+export NIELEM=100000000 # Number of integer elements
+export NSELEM=10000000 # Number of string elements
+
+export INT_DATA="random random-dup-10 random-boolean organ-pipe sorted equal almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
+
+export STR_DATA="random random-dup-10 random-boolean organ-pipe sorted equal   almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
+
+# CUSTOM Ad-HOC Overrides
+
+export INT_DATA="random organ-pipe"
+# export STR_DATA=""
+# export NIELEM=1000 # Number of integer elements
+# export NSELEM=1000 # Number of string elements
+
+
+########################## END CONFIG SECTION ##################
+
+
+export LOGFILE="log/sortbench-$(date -Iseconds).log"
+
+echo "Writing log to $LOGFILE"
+
+echo "# Sorting benchmark" | tee "$LOGFILE"
+
+( echo -n "# "; date ) | tee -a "$LOGFILE"
+
+( echo -n "# "; uname -a ) | tee -a "$LOGFILE"
+
+function run() {
+  ( ./do_benchmark $@ || echo "\n***ERROR $?" ) 2>&1 | tee -a "$LOGFILE"
+  echo | tee -a "$LOGFILE"
+}
+
+function run_int_std() {
+  run uint64 std::sort $1 $NIELEM $NITER
+}
+
+function run_int_isa() {
[…]

### 2019-11-22 benchmarks/sorting/benchmark_numbers.cpp

@@ -30,13 +30,16 @@
 
 #include <boost/sort/sort.hpp>
 
+#include "BlockQuicksort/blocked_double_pivot_check_mosqrt.h++"
+
+
 
 extern "C" {
   #include "introsort.h"
 }
 
-#define NELEM 1000000
-#define NITER 100
+#define NELEM 10000000
+#define NITER 10
 
 using namespace std;
 namespace bsort = boost::sort;
@@ -52,6 +55,7 @@ using bsort::spinsort;
 using bsort::flat_stable_sort;
 using bsort::spreadsort::spreadsort;
 using bsort::pdqsort;
+using bsort::pdqsort_branchless;
 
 
 template<typename A> double time_op(A op) {
@@ -76,6 +80,18 @@ template<typename A> void time_op(std::string name, A op) {
 namespace annot {
 
 
+  template<class Iter, class Compare>
+  inline void pdqsort(Iter first, Iter last, Compare comp) {
+      if (first == last) return;
+      boost::sort::pdqsort_detail::pdqsort_loop<Iter, Compare,
+          false
+          >(
+          first, last, comp, boost::sort::pdqsort_detail::log2(last - first));
+  }
+
+
+
+
   double insort_duration=0;
 
   void ll_introsort(uint64_t *a, int64_t l, int64_t h) {
@@ -97,6 +113,72 @@ namespace annot {
     std::cout<<"partitionings: "<<cnt_partitionings<<std::endl;
     std::cout<<"partitions: "<<cnt_partitions<<std::endl;
   }
+
+
+
+
+  /// This is a helper function for the sort routine.
+  template<typename _RandomAccessIterator, typename _Compare>
+    void
+    __guarded_linear_insert(_RandomAccessIterator __first, _RandomAccessIterator __last,
+			      _Compare __comp)
+    {
+      typename iterator_traits<_RandomAccessIterator>::value_type
+	__val = _GLIBCXX_MOVE(*__last);
+      _RandomAccessIterator __i = __last;
+      while (__i != __first && __comp(__val, __i-1))
+	{
+	  *__i = _GLIBCXX_MOVE(*(__i-1));
+	  --__i;
+	}
+      *__i = _GLIBCXX_MOVE(__val);
+    }
+
+  /// This is a helper function for the sort routine.
+  template<typename _RandomAccessIterator, typename _Compare>
+    void
+    __insertion_sort(_RandomAccessIterator __first,
[…]

### 2019-11-22 benchmarks/sorting/benchmark_strings.cpp

@@ -1,0 +1,524 @@
+//----------------------------------------------------------------------------
+/// @file benchmark_strings.cpp
+/// @brief Benchmark of several sort methods with strings
+///
+/// @author Copyright (c) 2017 Francisco José Tapia (fjtapia@gmail.com )\n
+///         Distributed under the Boost Software License, Version 1.0.\n
+///         ( See accompanying file LICENSE_1_0.txt or copy at
+///           http://www.boost.org/LICENSE_1_0.txt )
+///
+///         This program use for comparison purposes, the Threading Building
+///         Blocks which license is the GNU General Public License, version 2
+///         as  published  by  the  Free Software Foundation.
+///
+/// @version 0.1
+///
+/// @remarks
+//-----------------------------------------------------------------------------
+#include <algorithm>
+#include <iostream>
+#include <iomanip>
+#include <random>
+#include <stdlib.h>
+#include <vector>
+
+#include <boost/sort/common/time_measure.hpp>
+#include <boost/sort/common/file_vector.hpp>
+#include "boost/sort/common/int_array.hpp"
+
+#include <boost/sort/sort.hpp>
+
+extern "C" {
+  #include "introsort.h"
+}
+
+
+#define NMAXSTRING 1000000
+
+using namespace std;
+namespace bsort = boost::sort;
+namespace bsc = boost::sort::common;
+
+using bsc::time_point;
+using bsc::now;
+using bsc::subtract_time;
+using bsc::fill_vector_uint64;
+using bsc::write_file_uint64;
+
+using bsort::spinsort;
+using bsort::flat_stable_sort;
+using bsort::spreadsort::spreadsort;
+using bsort::pdqsort;
+
+void Generator_random (void);
+void Generator_sorted(void);
+void Generator_sorted_end(size_t n_last);
+void Generator_sorted_middle (size_t n_last);
+void Generator_reverse_sorted(void);
+void Generator_reverse_sorted_end(size_t n_last);
+void Generator_reverse_sorted_middle(size_t n_last);
+
+template <class IA, class compare>
+int Test(std::vector<IA> &B, compare comp = compare());
+
[…]

### 2019-11-22 benchmarks/sorting/do_benchmark.cpp

@@ -1,0 +1,761 @@
+// Re-used parts from BOOST sorting benchmark suite
+
+#include <algorithm>
+#include <iostream>
+#include <iomanip>
+#include <random>
+#include <stdlib.h>
+#include <vector>
+
+#include <boost/sort/common/time_measure.hpp>
+#include <boost/sort/common/file_vector.hpp>
+#include "boost/sort/common/int_array.hpp"
+
+#include <boost/sort/sort.hpp>
+
+#ifndef NO_LLVM
+extern "C" {
+  #include "introsort.h"
+}
+#endif
+
+using namespace std;
+namespace bsort = boost::sort;
+namespace bsc = boost::sort::common;
+
+using bsc::time_point;
+using bsc::now;
+using bsc::subtract_time;
+using bsc::fill_vector_uint64;
+using bsc::write_file_uint64;
+
+
+template<typename A> double time_op(A op) {
+  auto start = now ();
+  op();
+  auto finish = now ();
+  return subtract_time (finish, start);
+}
+
+
+template<typename A> void time_op(std::string name, A op) {
+  auto start = now ();
+  op();
+  auto finish = now ();
+  auto duration = subtract_time (finish, start);
+
+  std::cout<<"  "<<name<<"."<<"time: "<<duration<<std::endl;
+}
+
+template<class IA, class compare>
+void test_sorted(std::vector<IA> &A, compare comp)
+{
+  if (!std::is_sorted(A.begin (), A.end (), comp)) {
+    std::cout<<"*** NOT SORTED *** "<<std::endl;
+  }
+}
+
+template<class IA, class compare, class Op>
+void sort_test(std::vector<IA> &A, std::vector<IA> &B, size_t NITER, compare comp, Op op, bool verify=true) {
+//   std::cout<<name<<std::endl;
+
+  vector<double> durations;
+
+  for (size_t i=0;i<NITER;++i) {
+    A=B;
+
+    durations.push_back(time_op(op) * 1000.0);
+    if (verify) test_sorted(A,comp);
+  }
+
+  std::sort(durations.begin(),durations.end());
+
+  auto N=durations.size();
+//   auto l=durations.begin()+N/10;
+//   auto h=durations.end()-N/10;
+  auto l=durations.begin();
+  auto h=durations.end();
+
+  N=h-l;
+
[…]

### 2019-11-22 benchmarks/sorting/repeat_tests.sh

@@ -1,0 +1,4 @@
+#!/bin/bash
+./do_benchmark uint64 isabelle::sort almost-sorted-50 100000000 10
+./do_benchmark uint64 std::sort sorted-middle-10 100000000 10
+./do_benchmark llstring isabelle::sort random-boolean 10000000 10

### 2019-11-22 benchmarks/sorting/stest.cpp

@@ -14,7 +14,7 @@ extern "C" {
 }
 
 
-size_t NELEM = 100;
+size_t NELEM = 32;
 
 namespace bsc = boost::sort::common;
 
@@ -45,6 +45,25 @@ template <typename A> void print_range(A begin, A end) {
   std::cout<<std::endl;
 }
 
+void prn(int x) {
+  if (x<10) std::cout<<" "; std::cout<<x;
+}
+
+template <typename A> void print_range2(A l0, A begin, A end) {
+
+  size_t n = end-begin;
+  size_t base = begin-l0;
+
+  for (size_t i=0;i<n;++i) {prn(base+i); std::cout<<" ";}
+  std::cout<<std::endl;
+  for (size_t i=0;i<n;++i) {prn(*(begin+i)); std::cout<<" ";}
+  std::cout<<std::endl;
+}
+
+
+
+
+
 /*
 template <typename A> void print_range(A begin, A end) {
   for (auto i=begin;i<end;++i) std::cout<<*i<<" ";
@@ -63,6 +82,14 @@ extern "C" {
 
 }
 
+
+template<typename RandomAccessIterator>
+  inline void
+  std_heapsort(RandomAccessIterator first, RandomAccessIterator last)
+  {
+    std::make_heap(first,last);
+    std::sort_heap(first,last);
+  }
 
 
 
@@ -160,7 +187,119 @@ namespace annot {
 
 
 
+namespace test2 {
 
+
+  std::vector<uint64_t>::iterator l0;
+
+  /// This is a helper function...
+  template<typename _RandomAccessIterator, typename _Compare>
+    _RandomAccessIterator
+    __unguarded_partition(_RandomAccessIterator __first,
+			  _RandomAccessIterator __last,
+			  _RandomAccessIterator __pivot, _Compare __comp)
+    {
+      while (true)
+	{
+	  while (__first!=__last && __comp(__first, __pivot))
+	    ++__first;
+	  --__last;
+	  while (__first!=__last && __comp(__pivot, __last))
+	    --__last;
+	  if (!(__first < __last))
+	    return __first;
+	  std::iter_swap(__first, __last);
+	  ++__first;
+	}
+    }
+
+  /// This is a helper function...
+  template<typename _RandomAccessIterator, typename _Compare>
+    inline _RandomAccessIterator
+    __unguarded_partition_pivot(_RandomAccessIterator __first,
+				_RandomAccessIterator __last, _Compare __comp)
+    {
[…]

### 2019-11-22 mkhtml.sh

@@ -51,8 +51,9 @@ cp -a $ISABELLE_BROWSER_INFO/Unsorted/Isabelle_LLVM html/
 cp index.md html/
 cp dist.tgz html/
 cp LICENSE html/
-cp paper/main.pdf html/paper.pdf
-cp paper/talk/pres.pdf html/slides.pdf
+cp etc/logo/logo_200.png html/
+cp papers/ITP2019/main.pdf html/paper.pdf
+cp papers/ITP2019/talk/pres.pdf html/slides.pdf
 
 
 pandoc -V pagetitle="Isabelle LLVM" -s index.md > html/index.html

### 2019-11-22 src/lib_isabelle_llvm.c

@@ -6,14 +6,14 @@ void isabelle_llvm_abort() {
   abort();
 }
 
-void isabelle_llvm_abort_msg(char *msg) {
+void isabelle_llvm_abort_msg(char const *msg) {
   printf("Isabelle-LLVM (abort): %s\n",msg);
   abort();
 }
 
 char* isabelle_llvm_calloc(size_t n, size_t m) {
   // printf("calloc(%ld,%ld)",n,m);
-  char *res = calloc(n,m);
+  char *res = (char*)(calloc(n,m));
   if (!res) isabelle_llvm_abort_msg("Out of memory");
   return res;
 }

### 2019-11-22 src/lib_isabelle_llvm.cpp

@@ -1,0 +1,3 @@
+extern "C" {
+  #include "lib_isabelle_llvm.c"
+}

### 2019-11-22 thys/examples/code/bin_search.h

@@ -9,6 +9,6 @@ typedef struct {
   elem_t*data;
 } larray_t;
 
-int64_t bin_search(larray_t,elem_t);
+int64_t bin_search(larray_t, elem_t);
 
 #endif

### 2019-11-22 thys/examples/code/euclid.h

@@ -5,6 +5,6 @@
 
 
 
-uint64_t euclid(uint64_t,uint64_t);
+uint64_t euclid(uint64_t, uint64_t);
 
 #endif

### 2019-11-22 thys/examples/code/introsort.h

@@ -3,11 +3,21 @@
 #define _INTROSORT_H 1
 
 
+typedef struct {
+  int64_t size;
+  struct {
+    int64_t capacity;
+    char*data;
+  };
+} llstring;
 
-
+char llstrcmp(llstring*, llstring*);
+void str_init(llstring*);
+void str_append(llstring*, char);
+llstring* str_introsort(llstring*, int64_t, int64_t);
 uint64_t* heapsort(uint64_t*, int64_t, int64_t);
 uint64_t* introsort(uint64_t*, int64_t, int64_t);
-uint64_t* insertion_sort(uint64_t*, int64_t, int64_t);
 uint64_t* introsort_aux(uint64_t*, int64_t, int64_t, int64_t);
+uint64_t* insertion_sort(uint64_t*, int64_t, int64_t);
 
 #endif

### 2019-11-22 thys/examples/code/kmp.h

@@ -8,6 +8,6 @@ typedef struct {
   char*str;
 } string;
 
-int64_t kmp(string,string);
+int64_t kmp(string, string);
 
 #endif

### 2020-02-13 benchmarks/sorting/all_benchmarks.sh
Pdqsort

@@ -10,13 +10,16 @@ export NITER=10   # Number of iterations
 export NIELEM=100000000 # Number of integer elements
 export NSELEM=10000000 # Number of string elements
 
+# export NIELEM=10000000 # Number of integer elements
+# export NSELEM=1000000 # Number of string elements
+
 export INT_DATA="random random-dup-10 random-boolean organ-pipe sorted equal almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
 
 export STR_DATA="random random-dup-10 random-boolean organ-pipe sorted equal   almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
 
 # CUSTOM Ad-HOC Overrides
 
-export INT_DATA="random organ-pipe"
+# export INT_DATA="random organ-pipe"
 # export STR_DATA=""
 # export NIELEM=1000 # Number of integer elements
 # export NSELEM=1000 # Number of string elements
@@ -56,23 +59,42 @@ function run_str_isa() {
   run llstring isabelle::sort $1 $NSELEM $NITER
 }
 
+function run_pdq_int_std() {
+  run uint64 boost::pdqsort $1 $NIELEM $NITER
+}
+
+function run_pdq_int_isa() {
+  run uint64 isabelle::pdqsort $1 $NIELEM $NITER
+}
+
+function run_pdq_str_std() {
+  run llstring boost::pdqsort $1 $NSELEM $NITER
+}
+
+function run_pdq_str_isa() {
+  run llstring isabelle::pdqsort $1 $NSELEM $NITER
+}
+
+
+
 function run_int_gcc() {
[…]

### 2020-02-13 benchmarks/sorting/do_benchmark.cpp

@@ -681,15 +681,32 @@ struct less<llstring>
 
 #endif
 
+
+// Disable branch-prediction optimization for all iterators, to make it comparable with our impl!
+template<class Iter, class Compare>
+inline void boost_pdqsort(Iter first, Iter last, Compare comp) {
+    if (first == last) return;
+    boost::sort::pdqsort_detail::pdqsort_loop<Iter, Compare, false >( first, last, comp, boost::sort::pdqsort_detail::log2(last - first));
+}
+
+template<class Iter>
+inline void boost_pdqsort(Iter first, Iter last) {
+    typedef typename std::iterator_traits<Iter>::value_type T;
+    boost_pdqsort(first, last, std::less<T>());
+}
+
+
 void test_uint(string name, size_t NITER, vector<uint64_t> B) {
   std::vector<uint64_t> A (B);
 
   auto comp = std::less<uint64_t>();
 #ifndef NO_LLVM
   if (name=="isabelle::sort") sort_test(A,B,NITER,comp,[&]{ introsort(A.data (), 0, A.size());});
+  else if (name=="isabelle::pdqsort") sort_test(A,B,NITER,comp,[&]{ pdqsort(A.data (), 0, A.size());});
   else
 #endif
   if (name=="std::sort") sort_test(A,B,NITER,comp,[&]{ std::sort(A.begin (), A.end (), comp);});
+  else if (name=="boost::pdqsort") sort_test(A,B,NITER,comp,[&]{ boost_pdqsort(A.begin (), A.end (), comp);});
   else {
     cout<<"No such sorting algorithm "<<name<<endl;
     exit(1);
@@ -703,6 +720,8 @@ void test_llstring(string name, size_t NITER, vector<llstring> &B) {
   auto comp = std::less<llstring>();
 
   if (name=="isabelle::sort") sort_test(A,B,NITER,comp,[&]{ str_introsort(A.data (), 0, A.size());});
+  else if (name=="isabelle::pdqsort") sort_test(A,B,NITER,comp,[&]{ str_pdqsort(A.data (), 0, A.size());});
+  else if (name=="boost::pdqsort") sort_test(A,B,NITER,comp,[&]{ boost_pdqsort(A.begin (), A.end (), comp);});
   else if (name=="std::sort") sort_test(A,B,NITER,comp,[&]{ std::sort(A.begin (), A.end (), comp);});
   else {
     cout<<"No such sorting algorithm "<<name<<endl;

### 2020-02-13 benchmarks/sorting/int_generator.h

@@ -1,0 +1,286 @@
+#ifndef _INT_GENERATOR_H
+#define _INT_GENERATOR_H
+
+#include <algorithm>
+#include <iostream>
+#include <iomanip>
+#include <random>
+#include <stdlib.h>
+#include <vector>
+
+#include <boost/sort/common/time_measure.hpp>
+#include <boost/sort/common/file_vector.hpp>
+#include "boost/sort/common/int_array.hpp"
+
+#include <boost/sort/sort.hpp>
+
+using namespace std;
+namespace bsort = boost::sort;
+namespace bsc = boost::sort::common;
+
+using bsc::time_point;
+using bsc::now;
+using bsc::subtract_time;
+using bsc::fill_vector_uint64;
+using bsc::write_file_uint64;
+
+class Int_Generator {
+
+public:
+
+  static vector<uint64_t> random(size_t NELEM) {
+    vector<uint64_t> A;
+    A.reserve (NELEM);
+    A.clear ();
+    if (fill_vector_uint64 ("input.bin", A, NELEM) != 0)
+    {
+        std::cout << "Error in the input file\n";
+        exit(1);
+    };
+    return A;
+  }
+
+  static vector<uint64_t> random_dup (size_t NELEM, size_t nvals)
+  {
+    vector<uint64_t> A;
+    A.reserve (NELEM);
+    A.clear ();
+    if (fill_vector_uint64 ("input.bin", A, NELEM) != 0)
+    {
+        std::cout << "Error in the input file\n";
+        exit(1);
+    };
+
+    for (size_t i=0;i<A.size();++i) A[i]%=nvals;
+
+    return A;
+  }
+
+
+
+
+  static vector<uint64_t> sorted (size_t NELEM)
+  {
+    vector<uint64_t> A;
+
+    A.reserve (NELEM);
+    A.clear ();
+    for (size_t i = 0; i < NELEM; ++i)
+        A.push_back (i);
+
+    return A;
+  }
+
+  static vector<uint64_t> organ_pipe (size_t NELEM)
+  {
+    vector<uint64_t> A;
+
+    A.reserve (NELEM);
+    A.clear ();
+
+    size_t n = NELEM/2;
+    uint64_t v=0;
+
+    for (size_t i=0;i<n;++i) A.push_back(v++);
+    for (size_t i=n;i<2*n;++i) A.push_back(--v);
+    return A;
+  }
+
+
+  static vector<uint64_t> almost_sorted (size_t NELEM, size_t nperm)
+  {
[…]

### 2020-02-13 benchmarks/sorting/repeat_laptop.sh

@@ -1,0 +1,43 @@
+#!/bin/bash 
+echo -n '# '; date 
+echo -n '# '; uname -a 
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::pdqsort rev-sorted-end-10 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring isabelle::pdqsort rev-sorted-end-10 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::pdqsort sorted-end-10 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::pdqsort rev-sorted-middle-.1 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::pdqsort rev-sorted-middle-10 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::pdqsort almost-sorted-.1 10000000 10
+
+echo '# Repeated b/c difference > 15\% (154,201)'
+./do_benchmark uint64 boost::pdqsort sorted 100000000 10
+./do_benchmark uint64 isabelle::pdqsort sorted 100000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::pdqsort rev-sorted-middle-1 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring isabelle::pdqsort rev-sorted-middle-1 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::pdqsort sorted-middle-10 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring isabelle::pdqsort sorted-middle-10 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring isabelle::pdqsort sorted-end-1 10000000 10
+
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring isabelle::pdqsort random-dup-10 10000000 10
+

### 2020-02-13 benchmarks/sorting/repeat_server.sh

@@ -1,0 +1,14 @@
+#!/bin/bash 
+echo -n '# '; date 
+echo -n '# '; uname -a 
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark uint64 isabelle::pdqsort sorted-end-.1 100000000 10
+
+echo '# Repeated b/c difference > 15\% (8142,6584)'
+./do_benchmark llstring boost::pdqsort rev-sorted-middle-10 10000000 10
+./do_benchmark llstring isabelle::pdqsort rev-sorted-middle-10 10000000 10
+
+echo '# Repeated b/c difference > 15\% (3899,3371)'
+./do_benchmark uint64 boost::pdqsort almost-sorted-1 100000000 10
+./do_benchmark uint64 isabelle::pdqsort almost-sorted-1 100000000 10
+

### 2020-02-13 benchmarks/sorting/repeat_tests.sh

@@ -1,4 +1,0 @@
-#!/bin/bash
-./do_benchmark uint64 isabelle::sort almost-sorted-50 100000000 10
-./do_benchmark uint64 std::sort sorted-middle-10 100000000 10
-./do_benchmark llstring isabelle::sort random-boolean 10000000 10

### 2020-02-13 benchmarks/sorting/sexplore.cpp

@@ -1,0 +1,1199 @@
+
+#include <algorithm>
+#include <iostream>
+#include <iomanip>
+#include <random>
+#include <stdlib.h>
+#include <vector>
+
+#include <boost/sort/common/time_measure.hpp>
+#include <boost/sort/common/file_vector.hpp>
+#include "boost/sort/common/int_array.hpp"
+
+#include <boost/sort/sort.hpp>
+
+#include "int_generator.h"
+
+
+const size_t N = 100000000;
+
+// const size_t N = 1000;
+
+using namespace std;
+namespace bsort = boost::sort;
+namespace bsc = boost::sort::common;
+
+#define BOOST_PDQSORT_PREFER_MOVE(x) std::move(x)
+
+#define STAT_INCR(x) ++x
+// #define STAT_INCR(x)
+
+class sort_stat {
+public:
+  size_t bad = 0;
+  size_t heapsort = 0;
+  size_t leftp = 0;
+  size_t ap = 0;
+  size_t ap_sorted = 0;
+  size_t small=0;
+
+  void reset() {
+    bad = 0;
+    heapsort = 0;
+    leftp = 0;
+    ap = 0;
+    ap_sorted = 0;
+    small=0;
+  }
+
+  void print() {
+    cout<<"bad: "<<bad<<endl;
+    cout<<"heapsort: "<<heapsort<<endl;
+    cout<<"leftp: "<<leftp<<endl;
+    cout<<"ap: "<<ap<<endl;
+    cout<<"ap_sorted: "<<ap_sorted<<endl;
+    cout<<"small: "<<small<<endl;
+  }
+};
+
+sort_stat mystat;
+sort_stat pdqstat;
+
+
+namespace pdqcopy {
+  using namespace boost;
+  using namespace boost::sort;
+
+    enum {
+        // Partitions below this size are sorted using insertion sort.
+        insertion_sort_threshold = 24,
+
+        // Partitions above this size use Tukey's ninther to select the pivot.
+        ninther_threshold = 128,
+
+        // When we detect an already sorted partition, attempt an insertion sort that allows this
+        // amount of element moves before giving up.
+        partial_insertion_sort_limit = 8,
+
+        // Must be multiple of 8 due to loop unrolling, and < 256 to fit in unsigned char.
+        block_size = 64,
+
+        // Cacheline size, assumes power of two.
+        cacheline_size = 64
+    };
+
+    template<class T> struct is_default_compare : boost::false_type { };
[…]

### 2020-02-13 thys/examples/code/introsort.h

@@ -14,6 +14,8 @@ typedef struct {
 char llstrcmp(llstring*, llstring*);
 void str_init(llstring*);
 void str_append(llstring*, char);
+llstring* str_pdqsort(llstring*, int64_t, int64_t);
+uint64_t* pdqsort(uint64_t*, int64_t, int64_t);
 llstring* str_introsort(llstring*, int64_t, int64_t);
 uint64_t* heapsort(uint64_t*, int64_t, int64_t);
 uint64_t* introsort(uint64_t*, int64_t, int64_t);

### 2020-02-13 papers/sorting/main.tex
pdqsort paper

@@
-\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-
+% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[runningheads]{llncs}
+
+\usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{amssymb}
-\usepackage{amsthm}
+% \usepackage{amsthm}
 
@@
 \usepackage{cprotect}
+
+\usepackage[numbers]{natbib}
 
@@
 \usetikzlibrary{positioning,fit,shadows,decorations, arrows, shapes, decorations.markings,decorations.pathmorphing,calc}
+
+\usepackage{pgfplots}
 
@@
 
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+  \newcommand{\false}{\textrm{false}}
+  \newcommand{\true}{\textrm{true}}
+
+% \theoremstyle{definition}
+% \newtheorem{example}{Example}[section]
 
@@
 
-\title{Efficient Verified Implementation of Introsort}
+\title{Efficient Verified Implementation of Introsort and Pdqsort}
 
 \author{Peter Lammich}
-\email{peter.lammich@machester.ac.uk}
-\affiliation{%
-  \institution{The University of Manchester, UK}
-}
+\institute{The University of Manchester, UK}
+% \email{peter.lammich@machester.ac.uk}
 %
@@
 
+\maketitle
+
 \begin{abstract}
-  Sorting algorithm are an important part of most standard libraries, and
+  Sorting algorithms are an important part of most standard libraries, and
   both, their correctness and efficiency is crucial for many applications.
-%   There has been extensive research on improving sorting algorithms for
-%   both, the general case of sorting arbitrary objects wrt.\ a weak ordering,
-%   or special cases like sorting integers or strings.
 
   As generic sorting algorithm, the GNU C++ Standard Library implements the Introsort algorithm,
-  a combination of quicksort, heapsort, and insertion sort.
-
-  We verify this algorithm in the Isabelle LLVM verification framework,
-  including most of the optimizations from the GNU implementation.
-  On an extensive benchmark set, our verified algorithm performs on par with the original.
+  a combination of quicksort, heapsort, and insertion sort. The Boost C++ Libraries implement Pdqsort,
+  an extension of Introsort that achieves linear runtime on inputs with certain patterns.
+
+  We verify Introsort and Pdqsort in the Isabelle LLVM verification framework,
+  closely following the state-of-the-art implementations from GNU and Boost.
+  On an extensive benchmark set, our verified implementations perform on par with the originals.
 \end{abstract}
-
-\maketitle
 
@@
 \section{Introduction}
-  The goal of this paper is to provide a case study to show that the verification of
-  state-of-the art implementations of medium-complex algorithms is feasible.
-  We chose Musser's introspective sorting algorithm (Introsort)~\cite{Muss97}, and its
-  implementation in the GNU C++ Library (libstdc++)~\cite{libstdc++}.
-  Using the Isabelle-LLVM verification framework~\cite{La19-llvm}, we verify the same algorithm,
-  including most of the optimizations from libstdc++. The result is a verified sorting algorithm, which performs
-  on-par with the original implementation on an extensive set of benchmarks.
-
-  Sorting algorithms are an important part of any standard library. The Introsort algorithm,
-  being a combination of quicksort, heapsort, and insertion sort, is far from being trivial, but still
-  has a manageable complexity. Finally, libstdc++ provides a state-of-the art implementation with several
-  non-trivial but crucial optimizations. This makes Introsort a good candidate for our case study.
-
-  Moreover, sorting algorithms in standard libraries have not always been correct.
+%   In this paper, we present verified and efficient implementations of
+%   Musser's introspective sorting algorithm (Introsort)~\cite{Muss97} and its
+%   extension \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}.
+%   Using the Isabelle Refinement Framework with its LLVM backend~\cite{La19-llvm},
+%   we closely follow the implementations of Introsort and Pdqsort from
+%   the GNU C++ Library (libstdc++)~\cite{libstdc++} and the Boost C++ Libraries~\cite{boost}.
+%   The resulting sorting algorithms
+%   are verified down to their representation in the LLVM intermediate language~\cite{LLVM-manual},
+%   and perform on-par with the unverified original implementations on an extensive set of benchmarks.
+
+  Sorting algorithms are an important part of any standard library.
+  The GNU C++ Library (libstdc++)~\cite{libstdc++} implements Musser's introspective sorting algorithm (Introsort)~\cite{Muss97}.
+  It is a combination of quicksort, heapsort, and insertion sort,
+  which has the fast average case runtime of quicksort and the optimal $O(n\log(n))$
+  worst-case runtime of heapsort.
+  The Boost C++ Libraries~\cite{boost} provide a state-of-the-art implementation of
+  \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}, an extension of Introsort to
+  achieve better performance on inputs that contain certain patterns like already sorted sequences.
+  Verification of these algorithms and their state-of-the-art implementations is far from trivial,
+  but turns out to be manageable when handled with adequate tools.
+
+  Sorting algorithms in standard libraries have not always been correct.
   The Timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
@@
   Finally, LLVM's libc++~\cite{libc++} implements a different quicksort based sorting algorithm. While it may be functionally correct,
-  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by October 2019.}.
-
-  To the best of our knowledge, this paper provides the first verification of the Introsort algorithm.
-  This algorithm and two Java sorting algorithms~\cite{GRBBH15,BSSU17} are the only verified "real-world" implementations
-  of sorting algorithms we are aware of. On the technical side, this paper provides a case study of using stepwise refinement
-  for the design of verified algorithms.
-  Moreover, we report on two extensions to the Isabelle-LLVM framework, to handle nested container data structures and to automatically
-  generate C-header files to interface the generated code. These were required for our benchmark programs, which sort arrays of arrays of characters,
-  and interface the generated LLVM code from C++.
-
-  The complete Isabelle/HOL formalization and the benchmarks are submitted as artefact accompanying this paper.
+  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by January 2020.}.
+
+  In this paper, we present efficient implementations of Introsort and Pdqsort that are verified down to their
+  LLVM intermediate representation~\cite{LLVM-manual}. The verification uses the
+  Isabelle Refinement Framework~\cite{LaTu12}, and its recent Isabelle-LLVM backend~\cite{La19-llvm}.
+  We also report on two extensions of Isabelle-LLVM, to handle nested container data structures and to automatically
+  generate C-header files to interface the generated code. Thanks to the modularity of the Isabelle Refinement Framework,
+  our verified algorithms can easily be reused in larger verification projects.
+
+  While sorting algorithms are a standard benchmark for theorem provers and program verification tools,
+  verified real-world implementations seem to be rare: apart from our work, we are only aware of
+  two verified sorting algorithms~\cite{GRBBH15,BSSU17} from the Java standard library.
+
+  The complete Isabelle/HOL formalization and the benchmarks are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm/}.
 
@@
 
-\section{The Introsort Algorithm}
-  The Introsort algorithm by Musser~\cite{Muss97} is a sorting algorithm
-  that combines the good average-case runtime of quicksort~\cite{Hoare61} with the worst-case complexity of heapsort~\cite{Will64}.
-  The basic idea is to use quicksort as main sorting algorithm, but switch to heapsort when the recursion
+\section{The Introsort and Pdqsort Algorithms}
+  The Introsort algorithm by Musser~\cite{Muss97} is a generic unstable sorting algorithm
+  that combines the good average-case runtime of quicksort~\cite{Hoare61} with the optimal $O(n\log(n))$ worst-case complexity of heapsort~\cite{Will64}.
+  The basic idea is to use quicksort as main sorting algorithm, insertion sort for small partitions, and heapsort when the recursion
   depth exceeds a given limit, usually $2\lfloor\log_2 n\rfloor$ for $n$ elements.
 
-  Introsort is the default generic sorting algorithm in the GNU C++ Library~\cite{libstdc++}. Alg.~\ref{alg:introsort} displays its pseudo-code.
   \begin{algorithm}
   \begin{algorithmic}[1]
-    \Procedure{partial\_introsort}{$a,l,h,d$}
-      \If{$h-l>\textrm{threshold}$}
-        \If{$d=0$}
-          \State{\Call{heapsort}{$a,l,h$}}
+    \Procedure{introsort}{$xs,l,h$}
+      \If{$h-l>1$} \label{l:intrs:trivial}
+        \State{\Call{introsort\_aux}{$xs,l,h,2\lfloor\log_2(h-l)\rfloor$}} \label{l:intrs:aux}
+        \State{\Call{final\_insort}{$xs,l,h$}} \label{l:intrs:finalis}
+      \EndIf
+    \EndProcedure
+
+    \Procedure{introsort\_aux}{$xs,l,h,d$}
+      \If{$h-l>\textrm{threshold}$} \label{l:intrs:threshold}
+        \If{$d=0$}~\Call{heapsort}{$xs,l,h$} \label{l:intrs:heapsort}
         \Else
-          \State{$m\gets\Call{partition\_pivot}{a,l,h}$}
-          \State{\Call{partial\_introsort}{$a,l,m,d-1$}}
-          \State{\Call{partial\_introsort}{$a,m,h,d-1$}}
-        \EndIf
-      \EndIf
-    \EndProcedure
-
-    \Procedure{introsort}{$a,l,h$}
-      \If{$h-l>1$}
-        \State{\Call{partial\_introsort}{$a,l,h,2\lfloor\log_2(h-l)\rfloor$}}
-        \State{\Call{final\_insort}{$a,l,h$}}
+          \State{$m\gets\Call{partition\_pivot}{xs,l,h}$} \label{l:intrs:part}
+          \State{\Call{introsort\_aux}{$xs,l,m,d-1$}}  \label{l:intrs:rec1}
+          \State{\Call{introsort\_aux}{$xs,m,h,d-1$}}  \label{l:intrs:rec2}
+        \EndIf
       \EndIf
@@
   \end{algorithm}
-  The \textsc{introsort} function sorts the slice $[l{..}{<}h]$ of an array $a$. For this, it first
-  calls \textsc{partial\_introsort} to partially sort the array, i.e., all elements are no more than $\textrm{threshold}$ from their final position,
-  and then finishes the sorting by invoking insertion sort. The \textsc{partial\_introsort} procedure implements a standard quicksort, that switches
-  to heapsort when the recursion depth gets greater than $d$, and stops when the partition size becomes less than or equal to $\textrm{threshold}$.
-  Note that the actual implementation in the GNU C++ Library contains a manual tail-call optimization,
-  replacing the second recursive call to \textsc{partial\_introsort} by looping. For our formalization, we chose to not implement this
-  optimization, in particular due to the fact that the LLVM optimizer recognizes and eliminates this tail call automatically.
-
-  \subsection{Partitioning}
-  The \textsc{partition\_pivot($a,l,h$)} algorithm must re-arrange the elements in $a$ and return a \emph{pivot} index $m\in[l{<}{..}{<}h]$, such
-  that all elements $a[l{..}{<}m]$ are less than or equal to $a[m]$, and all elements $a[m{..}{<}h]$ are greater than or equal to $a[m]$.
-  Note that the performance of quicksort crucially depends on pivot selection: Ideally, the pivot index should be close to $(h-l)/2$,
-  i.e., the pivot element should be close to the median. The algorithm implemented by the GNU C++ Library is sketched in Alg.~\ref{alg:partition}.
-  It selects the median of the three elements $a[l+1]$, $a[\lfloor (h-l)/2 \rfloor]$, and $a[h-1]$ as pivot (l.~\ref{line:partition:pivot_selection}).
-  It then swaps the selected median element to the front $a[l]$, and invokes the standard Hoare partitioning algorithm,
-  which searches for pairs of mismatched elements simultaneously from the start and end of the array, and swaps them.
-  The chosen pivot element remains at position $a[l]$.
+
+  Algorithm~\ref{alg:introsort} shows our implementation of Introsort, which closely follows the implementation in libstdc++~\cite{libstdc++}.
+  The function \textsc{introsort} sorts the slice \is$l..<h$ of the list\footnote{%
+    Our formalization initially uses lists to represent the sequence of elements to be sorted, and refines them to arrays later (cf.\ Sec.\ \ref{sec:imp_impl}).
+  } $xs$.
+  If there is more than one element (line~\ref{l:intrs:trivial}), it initializes a depth
+  counter and calls the function \textsc{introsort\_aux} (line~\ref{l:intrs:aux}), which partially sorts the list such that every element is no more than
+  threshold positions away from its final position in the sorted list. The remaining sorting is then done by insertion sort (line~\ref{l:intrs:finalis}).
+  The function \textsc{introsort\_aux} implements a recursive quicksort scheme: recursion stops if the slice becomes smaller than the the threshold (line~\ref{l:intrs:threshold}).
+  If the maximum recursion depth is exhausted, heapsort is used to sort the slice (line~\ref{l:intrs:heapsort}).
+  Otherwise, the slice is partitioned (line~\ref{l:intrs:part}), and the procedure is
+  recursively invoked for the two partitions (line~\ref{l:intrs:rec1}--\ref{l:intrs:rec2}).
+
+  Note that we do not try to invent our own implementation, but closely follow the existing (and hopefully well-thought) libstdc++ implementation.
+  This includes the slightly idiosyncratic partitioning scheme, which leaves the pivot-element as first element of the left partition.
+  Moreover, the libstdc++ implementation contains a manual tail-call optimization, replacing the recursive call in line~\ref{l:intrs:rec2} by a loop.
+  While we could easily add this optimization in an additional refinement step, it turned out to be unnecessary, as LLVM recognizes and
+  eliminates this tail call automatically.
+
   \begin{algorithm}
   \begin{algorithmic}[1]
-    \Function{partition}{$a,l,h,p$}
-      \While{True}
-        \While{$a[l]<a[p]$} ${+}{+}l$ \EndWhile \label{partition:nextl_loop}
-        \State{${-}{-}h$}
-        \While{$a[p]<a[h]$} ${-}{-}h$ \EndWhile \label{partition:nexth_loop}
-        \If{$\neg l<h$} \Return{$l$}\EndIf
-        \State{\Call{swap}{$a[l],a[h]$}}
-        \State{${+}{+}l$}
-      \EndWhile
-    \EndFunction
-
-    \Function{partition\_pivot}{$a,l,h$}
-      \State{$p\gets \Call{median\_of\_three}{a,l+1,\lfloor h-l\rfloor/2,h-1}$} \label{line:partition:pivot_selection}
-      \State{\Call{swap}{$a[l],a[p]$}}
-      \State{\Return{\Call{partition}{a,l+1,h,l}}}
-    \EndFunction
+    \Procedure{pdqsort}{$lm,xs,l,h,d$}
+      \If{$h-l>1$}~\Call{pdqsort\_aux}{$\true,xs,l,h,\log(h-l)$}\EndIf \label{l:pdq:aux}
+    \EndProcedure
+
+    \Procedure{pdqsort\_aux}{$lm,xs,l,h,d$}
+      \If{$h-l<\textrm{threshold}$}~\Call{insort}{$lm,xs,l,h$} \label{l:pdq:threshold}
+      \Else
+        \State{\Call{pivot\_to\_front}{$xs,l,h$}} \label{l:pdq:pivotfront}
+        \If{$\neg lm \wedge xs[l-1] \not< xs[l]$} \label{l:pdq:eqopt}
+          \State{$m\gets \Call{partition\_left}{xs,l,h}$} \label{l:pdq:partleft}
+          \State{\textbf{assert} $m+1\le h$} \label{l:pdq:assert}
+          \State{\Call{pdqsort\_aux}{$\false,xs,m+1,h,d$}} \label{l:pdq:recright}
+        \Else
+          \State{$(m,ap)\gets\Call{partition\_right}{xs,l,h}$} \label{l:pdq:part}
+          \If{$m-l < \lfloor(h-l)/8\rfloor \vee h-m-1 < \lfloor(h-l)/8\rfloor$} \label{l:pdq:unbal}
+            \If{${-}{-}d = 0$}~\Call{heapsort}{xs,l,h};~\Return\EndIf \label{l:pdq:heapsort}
+            \State{\Call{shuffle}{xs,l,h,m}} \label{l:pdq:shuffle}
+          \ElsIf{$ap \wedge \Call{maybe\_sort}{xs,l,m} \wedge \Call{maybe\_sort}{xs,m+1,h}$}\State{\Return} \label{l:pdq:maybe}
+          \EndIf
+          \State{\Call{pdqsort\_aux}{$lm,xs,l,m,d$}}        \label{l:pdq:rec1}
+          \State{\Call{pdqsort\_aux}{$\false,xs,m+1,h,d$}}  \label{l:pdq:rec2}
+        \EndIf
+      \EndIf
+    \EndProcedure
   \end{algorithmic}
-  \caption{Partitioning}\label{alg:partition}
+  \caption{Pdqsort}\label{alg:pdqsort}
   \end{algorithm}
 
-  Note that this function contains a subtle optimization: The loops in lines~\ref{partition:nextl_loop} and \ref{partition:nexth_loop}
-  will iterate until they find a mismatched element. This iteration will not exceed the array bounds, only because the median selection
-  placed \emph{stopper elements} at the beginning and end of the array. In particular when the element comparison function is
-  cheap (e.g.\ for integer arrays), omitting an extra bounds check in the very inner loop of the sorting procedure can make a difference in practice.
-
-  \subsection{Unguarded Insertion Sort}
-  Insertion sort is a well-known simple sorting algorithm, which works by inserting the elements of the
-  array, one after the other, into the sorted prefix of the array. Its worst case complexity is quadratic in general,
-  but linear for partially sorted arrays with a constant threshold. For small thresholds, it performs better than quicksort,
-  such that it is effective to stop the quicksort procedure early, and do the final sorting with insertion sort.
-  Algorithm~\ref{alg:insort} shows the algorithm we implemented\footnote{The guarded
-  version ($G=\textrm{true}$) slightly differs from the implementation in the GNU C++ Library. However, as it is only called for a small
-  number of elements, we don't expect this to have any significant effect.}.
+  Algorithm~\ref{alg:pdqsort} shows our implementation of Pdqsort.
+  As for Introsort, the wrapper \textsc{pdqsort} just initializes a depth counter, and then calls the function \textsc{pdqsort\_aux}
+  (line~\ref{l:pdq:aux}),
+  which, in contrast to Introsort, completely sorts the list, such that no final insertion sort is necessary.
+  Again, the \textsc{pdqsort\_aux} function implements a recursive quicksort scheme, however, with a few additional optimizations.
+  Slices smaller than the threshold are sorted with insertion sort (line~\ref{l:pdq:threshold}). If the current slice is not the leftmost one
+  of the list, as indicated by the parameter $lm$,
+  the element before the start of the slice is guaranteed to be smaller than any element of the slice itself. This can be exploited to omit a comparison
+  in the inner loop of insertion sort (cf.~Sec.~\ref{sec:insort}).
+  If the slice is not smaller than the threshold, a pivot element is selected and moved to the front of the slice (line~\ref{l:pdq:pivotfront}).
+  If the pivot is equal to the element before the current slice (line~\ref{l:pdq:eqopt}), this indicates a lot of equal elements.
+  The \textsc{partition\_left} function (line~\ref{l:pdq:partleft}) will put them in the left partition, and then only the right partition needs
+  to be sorted recursively (line~\ref{l:pdq:recright}).
+  Otherwise, \textsc{partition\_right} (line~\ref{l:pdq:part}) places elements equal to the pivot in the right partition.
+  Additionally, it returns a flag $ap$ that indicates that the slice was already partitioned.
+  Next, we check for a highly unbalanced partitioning (line~\ref{l:pdq:unbal}), i.e., if one partition is less than 1/8th of the overall size.
+  After encountering a certain number of highly unbalanced partitionings, Pdqsort switches to heapsort (line~\ref{l:pdq:heapsort}).
+  Otherwise, it will shuffle some elements in both partitions, trying to break up patterns in the input (line~\ref{l:pdq:shuffle}).
+  If the input was already partitioned wrt.\ the selected pivot (indicated by the flag $ap$),
+  Pdqsort will optimistically try to sort both partitions with insertion sort (line~\ref{l:pdq:maybe}).
+  However, these insertion sorts abort if they cannot sort the list with a small number of swaps, limiting the penalty for being too optimistic.
+  Finally, the two partitions are recursively sorted (lines~\ref{l:pdq:rec1}--\ref{l:pdq:rec2}).
+
+  Our implementation of Pdqsort closely follows the implementation we found in Boost~\cite{boost}.
+  Again, we omitted a manual tail call optimization that LLVM does automatically.
+  Moreover, for certain comparison functions, Boost's Pdqsort uses a special branch-aware partitioning algorithm~\cite{EdWe16}.
+  We leave its verification to future work, but note that it will easily integrate in our existing formalization.
+
+  While Introsort and Pdqsort are based on the same idea, this presentation focuses on the more complex Pdqsort:
+  apart from the more involved \textsc{pdqsort\_aux} function,
+  \textsc{pivot\_to\_front} uses Tukey's 'ninther' pivot selection~\cite{Bent93}, while Introsort uses the simpler median-of-three scheme.
+  It has two partitioning algorithms used in different situations, and the \textsc{partition\_right} algorithm also
+  checks for already partitioned slices. Finally, with \textsc{insort} and \textsc{maybe\_sort}, it uses two different
+  versions of insertion sort.
+
+  \section{Verification}
+  We use the Isabelle Refinement Framework~\cite{LaTu12,La19-llvm} to formally verify our algorithms.
+  It provides tools to develop algorithms by stepwise refinement, and generates code in the LLVM intermediate representation~\cite{LLVM-manual}.
+
+  A program returns an element of the following datatype:
+  \begin{lstlisting}
+    \<alpha> nres == fail | spec (\<alpha> => bool)
+  \end{lstlisting}
+  Here \is{fail} represents possible non-termination or assertion violation, and \is{spec P}
+  a result nondeterministically chosen to satisfy predicate $P$. Note that we use \is{==} to indicate defining equations.
+  We define a \emph{refinement ordering} on \is{nres} by
+  \begin{lstlisting}
+    spec P <= spec Q == \<forall>x. P x ==> Q x         fail $\not\le$ spec Q          m <= fail
+  \end{lstlisting}
+  Intuitively, \is{m_1 <= m_2} means
+  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
+  Note that \is$<=$ is a complete lattice, with top element \is$fail$.
+  The \emph{monad combinators} are then defined as
+  \begin{lstlisting}
+    return x == spec y. y=x
+    bind (spec P) f == |_|{f x | P x}         bind fail f == fail
+  \end{lstlisting}
+  Here, \is{return x} deterministically returns \is{x}, and \is{bind m f} chooses a result of \is$m$ and then applies
+  \is$f$ to it. If \is$m$ may fail, then the bind may also fail.
+  We write \is{ x<-m; f x} for \is{bind m (\<lambda>x. f x)}, and \is{m_1; m_2} for \is{bind m_1 (\<lambda>_. m_2)}.
+
+%   Programs are expressed in the following nondeterminism-error monad.
+%   \begin{lstlisting}
+%     \<alpha> nres == fail | spec (\<alpha> => bool)                return x == spec y. y=x
+%     bind fail f == fail | bind (spec P) f = |_|{f x | P x}
+%   \end{lstlisting}
+%   Here \is{fail} represents possible non-termination or assertion violation, and \is{spec P}
+%   a result nondeterministically chosen to satisfy predicate $P$.
+%   We define a \emph{refinement ordering} on \is{nres} by
+%   \begin{lstlisting}
+%     spec P <= spec Q == \<forall>x. P x --> Q x         fail $\not\le$ spec Q          m <= fail
+%   \end{lstlisting}
+%   Intuitively, \is{m_1 <= m_2} means
+%   that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
+%   Note that \is$<=$ is a complete lattice, with top element \is$fail$.
+%   The program \is{return x} returns exactly the value \is{x}.
+%   The program \is{bind m f} chooses a result of \is$m$ and then applies \is$f$ to it. If \is$m$ may fail, then the bind may also fail.
+%   We will write \is{ x<-m; f x} for \is{bind m (\<lambda>x. f x)}, and \is{m_1; m_2} for \is{bind m_1 (\<lambda>_. m_2)}.
+
+  Arbitrary recursive programs can be defined via a fixed-point construction~\cite{Kr10}.
+  An \emph{assertion} fails if its condition is not met, otherwise it returns the unit value:
+  \begin{lstlisting}
+    assert P == if P then return () else fail;
+  \end{lstlisting}
+  Assertions are used to express that a program \is{m} satisfies the Hoare triple with
+  precondition \is$P$ and postcondition \is$Q$:
+  \begin{lstlisting}
+    m <= assert P; spec x. Q x
+  \end{lstlisting}
+  If the precondition is false, the right hand side is \is{fail}, and the statement trivially holds.
+  Otherwise, $m$ cannot fail, and every possible result $x$ of $m$ must satisfy $Q$.
+
+  While the Isabelle Refinement Framework provides some syntax to express programs, for better
+  readability, we use the slightly more informal syntax that we have already used in Algorithms~\ref{alg:introsort} and \ref{alg:pdqsort}.
+  In particular, we treat lists as if they were updated in place, while our actual formalization is purely functional, i.e., generates a new
+  version of the list on each update, which is explicitly threaded through the program. Destructively updated arrays will only be introduced in a
+  later refinement step (cf.~Sec.~\ref{sec:imp_impl}).
+
+  \subsection{Specification of Sorting Algorithms}\label{sec:sortspec}
+  The first step to verify a sorting algorithm is to specify the desired result.
+  We specify a sorting algorithm as follows:
+  \begin{lstlisting}
+    sort_spec xs l h
+      == assert l<=h \<and> h<=|xs|; spec xs'. xs =$_{l,h}$ xs' \<and> sorted (xs'[l..<h])
+  \end{lstlisting}
+%     xs =$_{l,h}$ xs'
+%       == xs[0..<l]=xs'[0..<l] \<and> mset (xs[l..<h]) = mset (xs'[l..<h]) \<and> xs[h..]=xs'[h..]
+  here \is$|xs|$ is the length of the list \is{xs} and
+  \is{xs[I]} is the slice of the list \is{xs} for indexes in the interval \is{I}.
+%   and \is{xs[l..]} is short for \is{xs[l..<|xs|${}$]}.
+  The equivalence relation \is{xs =$_{l,h}$ xs'} relates lists $xs$ and $xs'$ iff they are equal outside the slice \is$l..<h$
+  and $xs$ is a permutation of $xs'$.
+  To simplify the presentation, we assume a linear ordering on the elements.
+  Note that both C++ and our actual formalization support arbitrary weak orderings~\cite{Josu12}.
+
+  \subsection{Quicksort Scheme}
+  We split a call of \textsc{pdqsort\_aux} into phases, described by the following predicates:
+  \begin{lstlisting}
+    pvt xs == a_0 =$_{l,h}$ xs \<and> (\<exists>i \<in> l<..<h. xs[i] <= xs[l]) \<and> (\<exists>i \<in> l<..<h. xs[i] >= xs[l])
+    part m xs == a_0 =$_{l,h}$ xs \<and> l<=m \<and> m<h
+                  \<and> (\<forall>i \<in> l..<m. xs[i]<=xs[m]) \<and> (\<forall>i \<in> m<..<h. xs[m]<=xs[i])
+    sortl m xs == part m xs \<and> sorted (xs[l..<m])
+    sortr m xs == sortl m xs \<and> sorted (xs[m<..<h])
+  \end{lstlisting}
+  Let \is$a_0$ denote the original list. First, a pivot element is selected and moved to the beginning of the slice (phase \is{pvt}).
+  The pivot is selected in a way such there is at least one smaller (\is$<=$) and one greater (\is$>=$) element, e.g., by a median-of-three selection.
+  This knowledge can later be exploited to optimize the inner loops of the partitioning algorithm.
+  After the partitioning (phase \is$part m$), $m$ points to the pivot element, and all elements before $m$ are smaller, and all elements after $m$ are greater.
+  Then, first the left (phase \is$sortl m$), and then the right (phase \is$sortr m$) partition gets sorted, while the list remains partitioned around $m$.
+
+  This approach allows us to prove correct the algorithm, without assuming too many details of the underlying subroutines.
+  The following is all we need to know about the subroutines:
+  \begin{lstlisting}
+    (a) lm \<or> notleft xs l h ==> insort lm xs l h <= sort_spec xs l h
+    (b) l+4<h ==> pivot_to_front xs l h <= spec xs'. pvt xs'
+    (c) pvt xs ==> partition_right xs l h <= spec (xs',m,_). part m xs'
+                     \<and> partition_left xs l h   <= spec (xs',m,_). part m xs'
+    (d) heapsort xs l h <= sort_spec xs l h
+    (e) part m xs ==> shuffle xs l h m <= spec xs'. part m xs'
+    (f) i<=j \<and> j<=|xs|
+      ==> maybe_sort xs i j <= spec (b,xs'). xs=$_{i,j}$xs' \<and> (\<not>b \<or> sorted xs'[i..<j])
+  \end{lstlisting}
+  where \is$notleft xs l h == 0<l \<and> \<forall>i \<in> l..<h. xs[l-1] <= xs[i]$
+  states that the element \is{xs[l-1]} before the slice is smaller than any element of the slice.
+  Note that we explicitly mention the changed list $xs'$ in these specifications, while we left the list changes implicit in the algorithm description.
+
+  Intuitively, (a,d,f) state correctness of the sorting subroutines, (b) states that pivot selection goes to phase \is{pvt},
+  (c) states that partitioning transitions from phase \is{pvt} to phase \is{part}, and (e) states that shuffling preserves phase \is{part}.
+  From the above, we easily prove the following lemmas:
+  \begin{lstlisting}
+    (g) part m xs \<and> notleft xs l h \<and> xs[m] <= xs[l-1] ==> sorted xs[l..<m]
+    (h) part m xs \<and> ( xs=$_{l,m}$xs' \<or> xs=$_{m+1,h}$xs') ==> part m xs'
+    (i) sortl m xs \<and> xs=$_{m+1,h}$xs' ==> sortl m xs'
+    (j) part m xs ==> sort_spec xs l m <= spec xs'. sortl m xs'
+    (k) sortl m xs ==> sort_spec xs (m+1) h <= spec xs'. sortr m xs'
+    (l) sortr m xs ==> sorted xs[l..<h]
+  \end{lstlisting}
+  The correctness statement for \textsc{pdqsort\_aux} is:
+  \begin{lstlisting}
+  lm \<or> notleft xs l h ==> pdqsort_aux lm xs l h d <= sort_spec xs l h
+  \end{lstlisting}
+  The proof is done by using the Refinement Framework's verification condition generator, and then discharging the generated VCs using the above lemmas.
+  The line numbers in the following brief sketch refer to Algorithm~\ref{alg:pdqsort}.
+  As termination measure for the recursion, we use the size $h-l$ of the slice to be sorted.
+  If we switch to insertion sort in line~\ref{l:pdq:threshold}, (a) implies that the slice gets sorted, and we are done.
+  Otherwise, we select a pivot in line~\ref{l:pdq:pivotfront}, going to phase \is{pvt} (b).
+  When the equals optimization is triggered in line~\ref{l:pdq:eqopt}, we transition to phase \is{part} (c),
+  and the left partition is already sorted\footnote{actually all elements in the left partition are equal to the pivot.} (g), such that we can transition to phase \is{sortl} (j), and, via a recursive call in line~\ref{l:pdq:recright}
+  to phase \is{sortr} (k). This implies that the slice is sorted (l), and we are done.
+  When the equals optimization is not used, (c) shows that we transition to phase \is{part} in line~\ref{l:pdq:part}.
+  If the partition is unbalanced, we either use heapsort (line~\ref{l:pdq:heapsort}) to directly sort the slice (d),
+  or shuffle the elements (line~\ref{l:pdq:shuffle}) and stay in phase \is{part} (e).
+  In line~\ref{l:pdq:maybe}, the algorithm may attempt to sort the slice.
+  If this succeeds, we are done (f). Otherwise, we stay in phase \is{part} (h),
+  and the recursive calls in lines~\ref{l:pdq:rec1} and \ref{l:pdq:rec2} will take us to
+  phase \is{sortr} (j,k), which implies sortedness of the slice (l).
+
+  Using the above statement, and an analogous statement for Introsort, we can prove the main correctness theorem:
+  \begin{theorem}\label{thm:sorting_correct}
+  \begin{lstlisting}
+    pdqsort xs l h <= sort_spec xs l h
+              $\textbf{and}$  introsort xs l h <= sort_spec xs l h
+  \end{lstlisting}
+  \end{theorem}
+
+  Note that we could prove the correctness of our algorithm with only minimal assumptions about the used subroutines.
+%   Actually, in our formalization we use the specifications (a-f) directly, and replace them by their
+%   implementations in a later refinement step.
+  This decoupling of the algorithm from its subroutines simplifies the proof, as it is not obfuscated with unnecessary details.
+  For example, correctness of the algorithm does not depend on the exact partitioning scheme being used, as long as it implements
+  a transition from the \is$pvt$ to the \is$part$ phase. It also simplifies changing the subroutines later, e.g.,
+  adding further optimizations such as branch-aware partitioning~\cite{EdWe16}.
+
+  Breaking down an algorithm into small and decoupled modules is often the key to its successful verification.
+  Note that the original implementation in Boost is more coarse grained, inlining much of the functionality into the main algorithm.
+  After having proved correct an algorithm, we can always do the inlining in a later refinement step, or rely on the LLVM optimizer to do
+  the inlining for us. In our formalization, we use the inlining feature of Isabelle-LLVM's preprocessor.
+%   Thus, splitting up an algorithm into small parts does not come with a performance penalty by procedure calling overhead.
 
   \begin{algorithm}
+  \begin{minipage}{.41\textwidth}
   \begin{algorithmic}[1]
-    \Procedure{insert}{$G,a,l,i$}
-      \State{$t\gets a[i]$}
-      \While{$ (\neg G \vee l<i) \wedge t<a[i-1]$} \label{alg:insert:loophead}
-        \State{$a[i]\gets a[i-1]$}
-        \State{${-}{-}i$}
-      \EndWhile
-      \State{$a[i]\gets t$}
-    \EndProcedure
-
-    \Procedure{gen\_insort}{$G,a,l,i,h$}
+    \Procedure{insort}{$G,xs,l,h$}
+      \If{$l=h$}~\Return\EndIf
+      \State{$i\gets l+1$}
       \While{$i<h$}
-        \State{\Call{insert}{$G,a,l,i$}}
+        \State{\Call{insert}{$G,xs,l,i$}}
         \State{${+}{+}i$}
@@
     \EndProcedure
-
-    \Procedure{final\_insort}{$a,l,h$}
-      \If{$h-l\le\textrm{threshold}$}
-        \State{\Call{gen\_insort}{$\textrm{true},a,l,l+1,h$}}
-      \Else
-        \State{\Call{gen\_insort}{$\textrm{true},a,l,l+1,l+\textrm{threshold}$}}
-        \State{\Call{gen\_insort}{$\textrm{false},a,l,l+\textrm{threshold},h$}}
-      \EndIf
+    \algstore{algs:insort-ctd}
+  \end{algorithmic}
+  \end{minipage}\hfill
+  \begin{minipage}{.52\textwidth}
+  \begin{algorithmic}[1]
+    \algrestore{algs:insort-ctd}
+    \Procedure{insert}{$G,xs,l,i$}
+      \State{$t\gets xs[i]$}
+      \While{$ (\neg G \vee l<i) \wedge t<xs[i-1]$} \label{alg:insert:loophead}
+        \State{$xs[i]\gets xs[i-1]$}
+        \State{${-}{-}i$}
+      \EndWhile
+      \State{$xs[i]\gets t$}
     \EndProcedure
   \end{algorithmic}
-  \caption{Final Insertion Sort}\label{alg:insort}
+  \end{minipage}
+  \caption{Insertion Sort}\label{alg:insort}
   \end{algorithm}
-
-  This algorithm contains a similar optimization as the partitioning algorithm: As the array is already partially sorted,
-  we know that elements beyond the threshold index won't be inserted at the very beginning of the array.
-  This is exploited to get rid of the bounds check in line~\ref{alg:insert:loophead}.
-  Only the first $\textrm{threshold}$ elements of the array are sorted with activated bounds check. In the displayed code, we control
-  the bounds check by the Boolean parameter $G$, and assume that the algorithm will be specialized for both values of $G$.
-
-%   Combination of quicksort, heapsort, and insertion sort
-%   Worst-case $O(n\log n)$, average case efficiency of quicksort.
-
-\section{Notation}
-  Throughout this paper, we use some specific notation. We write function application in curried form, i.e., \is{f x_1 ... x$_n$}.
-  Definitions of types and functions are denoted by \is{==}. Indexes start at zero. The \is$i$th element of a list is \is$xs[i]$,
-  the slice of a list from index \is$i$ inclusive to \is$j$ exclusive is \is$xs[i..<j]$. We also use the notations \is{xs[i<..<j]}, \is{xs[i<..j]},
-  and \is{xs[i..j]} to specify exclusiveness of bounds. Moreover, an omitted lower bound denotes index $0$, and an omitted upper bound denotes
-  the length of the list, e.g., \is$xs[..<n]$ is the list of the first \is$n$ elements of $xs$, and \is$xs[n..]$ is the list of the remaining elements.
-  List concatenation is written as \is{xs_1@xs_2}, and \is$xs[i:=x]$ is the list \is$xs$, with index \is$i$ updated to element \is$x$.
-  The length of list \is$xs$ is denoted by \is{|xs|}.
-
-\section{The Isabelle Refinement Framework}
-  The Isabelle Refinement Framework~\cite{LaTu12} provides tools for
-  verified program development by stepwise refinement, using the Isabelle HOL proof assistant~\cite{NPW02}.
-  It is based on a nondeterminism-error monad, defined as follows:
-  \begin{lstlisting}
-    \<alpha> nres == FAIL | RES (\<alpha> set)
-    return x == RES {x}
-    bind FAIL f == FAIL | bind (RES X) f = |_|x\<in>X. f x
-  \end{lstlisting}
-  Intuitively, \is{FAIL} represents a program that may fail (e.g.~not terminate or fail explicitly by violating an assertion),
-  and \is{RES X} represents a program that, nondeterministically, returns a value \is{x\<in>X}.
-  By extending the subset ordering, with \is{FAIL} as greatest element, \is{\<alpha> nres} becomes a complete lattice.
-  The program \is{return x} returns exactly the value \is{x}.
-  The program \is{bind m f} sequentially composes \is{m} and \is{f}: it nondeterministically chooses a
-  return value \is{x} of \is{m}, and then behaves like \is{f x}. If either \is{m}, or one of the \is{f x} may fail,
-  then also \is{bind m f} may fail.
-
-  We use sequential notation for bind, i.e., \is$x<-m; f x$ stands for \is$bind m (\<lambda>x. f x)$,
-  and \is$m_1; m_2$ stands for \is$bind m_1 (\<lambda>_. m_2)$, i.e., the result of \is$m_1$ is ignored.
-  We also use the standard HOL functions \is{if-then-else} and \is{let}, and define the following notations:
-  \begin{lstlisting}
-    assert \Phi == if \Phi then return () else FAIL
-    spec x. \Phi x == RES {x | \Phi x}
-  \end{lstlisting}
-  An assertion will fail if its argument is false, and \is{spec x. \Phi x} describes all results that satisfy \is{\Phi}.
-
-  Recursion is defined by the greatest fixed point over a monotonic functor \is{F}{\hspace{1pt}}\footnote{Monadic
-    programs are monotonic by construction~\cite{Kr10}. Moreover, on monadic programs, the greatest fixed point over the \is{nres}-ordering
-    coincides with the fixed point over a flat CCPO, which is the standard construction for recursion. In the actual
-    formalization, monotonicity wrt.\ both orderings is asserted. }
-  , and while loops are defined via recursion:
-  \begin{lstlisting}
-    rec F x == assert (mono F); fixp F x
-    while b c s == rec (\<lambda>W s.
-      if b s then s' <- c s; W s' else return s) s
-  \end{lstlisting}
-  From now on, we will omit the \is{rec} syntax, and define recursive monadic functions by their recursion equation, e.g.:
-  \begin{lstlisting}
-    while b c s ==
-      if b s then s' <- c s; while b c s' else return s
-  \end{lstlisting}
-
-  The ordering on \is{nres} can be interpreted as refinement: \is{m_1 <= m_2} means
-  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
-  The statement
-  \begin{lstlisting}
-    f x <= { assert (P x); spec r. Q r}
-  \end{lstlisting}
-  means that function $f$ satisfies the specification with precondition \is{P} and postcondition \is{Q}.
-
-\section{Verifying Introsort}
-  The first step to verify a sorting algorithm is to specify the desired result.
-  We specify a sorting algorithm as follows:
-  \begin{lstlisting}
-    sort_spec xs == spec xs'. sorted xs' \<and> mset xs' = mset xs
-    $\textrm{where}$
-    sorted xs == \<forall>i<j<|xs|. \<not>(xs[i] > xs[j])
-  \end{lstlisting}
-  That is, the result will be sorted and a permutation of the input (\is{mset xs} denotes the multiset of elements of list $xs$).
-
-  Note that \is{<} is a \emph{strict weak ordering}, i.e., a relation with:
-  \begin{lstlisting}
-    a<b ==> \<not>b<a      and      a<c ==> a<b \or b<c
-  \end{lstlisting}
-  Weak orderings are used by C++ for sorting~\cite{Josu12}.
-  They are a natural model for objects that are sorted by linearly ordered keys:
-  An equivalent characterization can be given by an equivalence relation on the elements (e.g.~objects with same keys),
-  and a linear ordering on the equivalence classes.
-  In particular, any linear ordering is a weak ordering, with the equivalence relation being equality, and
-  $\neg (a<b)$ meaning $b\le a$. While our formalization is over arbitrary weak orderings, we assume a linear ordering for the rest of
-  this paper, to simplify the presentation.
-
-  For a partial sorting algorithm wrt.\ threshold \is{n}, we replace \is{sorted} by:
-  \begin{lstlisting}
-    part_sorted xs == \<exists>ss. is_slicing xs ss \<and> sorted$_\preceq$ ss"
-    $\textrm{where}$
-    is_slicing xs ss == xs = concat ss \<and> (\<forall>s\<in>ss. |s| <= n)"
-    xs $\preceq$ ys == \<forall>x\<in>xs. \<forall>y\<in>ys. x <= y
-  \end{lstlisting}
-  That is, a list is partially sorted if we can split it into slices that are not longer than
-  the threshold $n$, and each element in a slice is less than or equal to any element in a subsequent slice.
-
-  The next step to verify quicksort like algorithms is to specify a partitioning function:
-  \begin{lstlisting}
-    partition_spec xs == assert (|xs| >= 4);
-      spec (\<lambda>(xs_1,xs_2). xs_1~=[] \and xs_2~=[]
-             \and  mset xs = mset xs_1 + mset xs_2 \and xs_1 $\preceq$ xs_2)
-  \end{lstlisting}
-  Given a list \is{xs} of sufficient length, two non-empty lists \is{xs_1,xs_2} are returned,
-  which together contain the same elements as $xs$, and each element of the first list is less than or equal to any element of the second list.
-  We assume lists of at least four elements, because this is the size where the pivot selection makes sense (cf.~Alg.~\ref{alg:partition}).
-  In practice, the lists will always be longer than the threshold, which is typically 16.
-
-%   The final sorting step has to completely sort an already sorted list:
-%   \begin{lstlisting}
-%     final_sort_spec xs == assert (part_sorted xs); sort_spec xs
-%   \end{lstlisting}
-%
-  Finally, we can specify the partial Introsort algorithm in the Refinement Framework:
-  \begin{lstlisting}
-    partial_introsort xs d ==
-        if |xs| > threshold then {
-            if d=0 then sort_spec xs
-            else {
-                (xs_1,xs_2) <- partition_spec xs;
-                xs_1 <- partial_introsort xs_1 (d-1);
-                xs_2 <- partial_introsort xs_2 (d-1);
-                return (xs_1@xs_2) }}
-        else return xs
-  \end{lstlisting}
-%
-  A straightforward Isabelle proof yields:
-  \begin{theorem}\label{thm:partial_introsort_correct}
-  \begin{lstlisting}
-    partial_introsort xs d <= part_sort_spec xs
-  \end{lstlisting}
-  \end{theorem}
-
-\subsection{Refinement}\label{sec:refinement}
-  The partial introsort algorithm that we have specified clearly captures the idea of the algorithm.
-  However, it splits the input list and later concatenates it, which makes it unsuitable for an in-place
-  implementation on arrays. In this section we will \emph{refine} this algorithm into one that is better suited
-  for an implementation on arrays. Note that refinement preserves correctness, such that the new algorithm will
-  automatically be correct.
-
-  Given a relation \is{R} that relates \emph{concrete} with \emph{abstract} values, and a program \is{m} that returns abstract values,
-  the program \is{\<Down>R m} is the biggest program that returns only concrete values with corresponding abstract values in \is{m}.
-  Moreover, \is{\<Down>R m} may fail if and only if \is{m} may fail. Formally:
-  \begin{lstlisting}
-    \<Down> R FAIL == FAIL
-    \<Down> R (RES A) == RES { c. \<exists>a\<in>A. (c,a)\<in>R }
-  \end{lstlisting}
-  Note that for the identity relation \is{I}, we have \is{\<Down> I m = m}.
-
-  \begin{example}
-    As a simple but constructed example, consider the specification of a program that returns some set of numbers divisible by \is$n$:
-    \begin{lstlisting}
-      dvd_set_spec n == assert (n>0); SPEC S. \<forall>m\<in>S. n dvd m
-    \end{lstlisting}
-
-    A possible implementation of this program is
-    \begin{lstlisting}
-      dvd_set_1 n == return {n,2*n}
-    \end{lstlisting}
-    and we have \is{dvd_set_1 n <= dvd_set_spec n}. This restricts the number of possible results.
-
-    Now consider the relation \is{list_set_rel} that relates distinct lists to the set of their elements:
-    \begin{lstlisting}
-      list_set_rel == { (l, set l) | distinct l }
-    \end{lstlisting}
-
-    We have \is{return [n,2*n] <=\<Down> list_set_rel (dvd_set_1 n)}. This changes the
-    representation from sets to lists. Note that, by transitivity and monotonicity of \is{\<Down>} we also have:
-    \begin{lstlisting}
-      return [n,2*n] <=\<Down> list_set_rel (dvd_set_spec n)
-    \end{lstlisting}
-    Intuitively, this limits the number of possible results and changes their representation at the same time.
-  \end{example}
-
-  We now refine the specifications and the introsort algorithm to work on a \emph{slice} of a list rather than on the whole list.
-  The relation \is{slice_rel xsr l h} relates a list with its slice from \is{l} to \is{h}. Outside the slice, the list must
-  be equal to \is{xsr}:
-  \begin{lstlisting}
-    slice_rel xsr l h == {(xs, xs[l..<h]) | l <= h \and h <= |xs|
-      \and |xs|=|xsr| \and xs[..<l] = xsr[..<l] \and xs[h..] = xsr[h..]
-    }
-  \end{lstlisting}
-  We then define:
-  \begin{lstlisting}
-    slice_sort_spec xs l h == \<Down> (slice_rel xs l h) (
-      assert (l <= h \and h <= |xs|);
-      sort_spec (xs[l..<h]) )
-  \end{lstlisting}
-  That is, for valid ranges \is{l,h}, we return a list where the range is sorted, and the rest is unchanged.
-  Partial sorting of a slice is specified analogously.
-
-  Similarly, we refine the specification for partitioning:
-  \begin{lstlisting}
-    slice_partition_spec xs l h == \<Down>({((xs',m),(xs_1,xs_2)).
-      (xs',xs_1@xs_2)\<in>slice_rel xs l h \and m = l + |xs_1| })
-      (assert (l<=h \and h<=|xs|); partition_spec (xs[l..<h])})
-  \end{lstlisting}
-  This returns a pair of a list \is{xs'} and an index \is{m}.
-  This list \is{xs'} contains the two partitions \is{xs_1@xs_2} at range \is{l}, \is{h},
-  and is equal to the input list \is{xs} outside this range. Moreover, the index \is{m}
-  is the index where the second partition starts in \is{xs'}.
-
-  Having defined the concrete specifications, we define a more concrete version of the Introsort algorithm:
-  \begin{lstlisting}
-    slice_partial_introsort xs l h d ==
-        assert (l<=h);
-        if h-l > threshold then {
-            if d=0 then slice_sort_spec xs l h
-            else {
-                (xs,m) <- slice_partition_spec xs l h;
-                xs <- slice_partial_introsort xs l m (d-1);
-                xs <- slice_partial_introsort xs m h (d-1);
-                return xs }}
-        else return xs
-  \end{lstlisting}
-  To show that this algorithm refines \is{slice_sort_spec}, we first show that it refines \is{partial_introsort}:
-  \begin{theorem}\label{thm:slice_partial_introsort_refines}
-  \begin{lstlisting}
-    l<=h \and h<=|xs| ==>
-      slice_partial_introsort xs l h d
-      <= \<Down> (slice_rel xs l h) (partial_introsort (xs[l..<h]) d)
-  \end{lstlisting}
-  \end{theorem}
-  The proof is straightforward: First, we use the verification condition generator of the
-  Refinement Framework\footnote{we had to do a bit of manual hinting to get the recursion refinement right.}.
-  This reduces the proof to a set of proof obligations on lists. With some help from the sledgehammer
-  tool~\cite{BBP13}, these could be solved easily. For example, a crucial proof obligation relates the first abstract and recursively
-  sorted partition with the respective slice of the concrete result:
-  \begin{lstlisting}
-    lemma (xs', xs_1 @ xs_2) \<in> slice_rel xs l h
-        ==> xs_1 = xs'[l..<l+|xs_1|${}$]
-        apply (clarsimp simp: slice_rel_def)
-        by (metis slice_def add_diff_cancel_left' append.assoc ...)
-  \end{lstlisting}
-  The first line of the proof unfolds the definition of \is{slice_rel} and simplifies the goal.
-  These simplification steps are standard, and usually don't require too much thinking.
-  The second line of the proof was then automatically found by sledgehammer.
-
-  Finally, we combine the refinement statement (Thm~\ref{thm:slice_partial_introsort_refines}) with the
-  correctness result for \is{partial_introsort} (Thm~\ref{thm:partial_introsort_correct}).
-  By transitivity and monotonicity of \is{\<Down>} we get correctness of \is{slice_partial_introsort}:
-  \begin{theorem}\label{thm:slice_partial_introsort_correct}
-  {\ }\vspace*{-.5em}
-
-  \begin{lstlisting}
-    slice_partial_introsort xs l h d <= slice_part_sort_spec xs l h
-  \end{lstlisting}
-  \end{theorem}
-  Note how refinement allowed us to split this correctness proof into two independent parts:
-  We first established the correctness of Introsort on a whole list, using a partition function that returns two separate lists.
-  Although this \emph{abstract} representation is not suitable for an in-place implementation with arrays, the correctness proof is much simpler than a
-  direct correctness proof on slices. The second part of the proof is independent from the abstract correctness proof.
-  It just shows how to implement the abstract algorithm on the \emph{concrete} data representation.
-
-  The main Introsort algorithm is simple enough to define it on slices directly:
-  The additional overhead introduced by refinement would outweigh the
-  simplification of the proof.
-
-  We first specify the final sorting step to completely sort an already partially sorted slice\footnote{
-    We will check for an empty or singleton slice, which is trivially sorted, in the main algorithm.
-    Thus, the additional precondition \is{h-l > 1} allows us to omit this check in the final sorting algorithm.
-  }:
-  \begin{lstlisting}
-    slice_final_sort_spec xs l h ==
-      assert (h-l > 1 \and part_sorted (xs[l..<h]));
-      slice_sort_spec xs l h
-  \end{lstlisting}
-  The Introsort algorithm then simply combines partial sorting and final sorting:
-  \begin{lstlisting}
-    introsort xs l h ==
-        assert (l <= h);
-        if h-l>1 then {
-            xs <- slice_part_sort_spec xs l h;
-            xs <- final_sort_spec xs l h;
-            return xs
-        } else return xs
-  \end{lstlisting}
-  Its correctness proof is straightforward, and we get:
-  \begin{theorem}\label{thm:introsort_correct}
-    \begin{lstlisting}
-    introsort xs l h <= slice_sort_spec xs l h
-    \end{lstlisting}
-  \end{theorem}
-
-\section{Assembling the Implementation}
-  Note that our \is{introsort} algorithm itself is independent from both,
-  the partial and the final sorting algorithms: It only uses
-  abstract specifications of them.
-  However, in order to arrive at a concrete implementation, one has
-  to \emph{plug in} concrete implementations for these specifications.
-  Similarly, our \is{slice_partial_introsort} algorithm depends on abstract
-  specifications of partitioning and sorting (when the recursion depth limit is hit).
-  Figure~\ref{fig:modular} shows the complete architecture of our Introsort implementation.
-  \begin{figure}
-  \begin{tikzpicture}[
-    node distance=2mm,
-%     algoname/.style={font=\fontsize{6}{6}\color{black!50}\ttfamily},
-%     depname/.style={rectangle, draw=black!50, font=\scriptsize\ttfamily, anchor=west}
-    algoname/.style={font=\scriptsize\color{blue}\ttfamily},
-    depname/.style={font=\fontsize{6}{6}\color{red}\ttfamily, anchor=west}
-  ]
-%     \node (introsort) [algoname] { introsort };
-%     \node (introsort-ps) [below=of introsort.west, depname, xshift=2mm] { part\_sort };
-%     \node (introsort-fs) [below=of introsort-ps.west, depname] { final\_sort };
-%     \node (introsort-box) [draw=black!50, fit={(introsort) (introsort-ps) (introsort-fs)}] {};
-%
-%     \node (partintrosort) [right=6mm of introsort-box.north east, yshift=-1.15mm, anchor=north west, algoname] { partial\_introsort };
-%     \node (partintrosort-s) [below=of partintrosort.west, depname, xshift=2mm] { sort };
-%     \node (partintrosort-p) [below=of partintrosort-s.west, depname] { partition };
-%     \node (partintrosort-box) [draw=black!50, fit={(partintrosort) (partintrosort-s) (partintrosort-p)}] {};
-%
-%     \node (insort) [below=2mm of partintrosort-box.south west, anchor=north west, algoname,draw=black!50] { final\_insertion\_sort };
-%
-%     \node (heapsort) [right=6mm of partintrosort-box.north east, anchor=north west, yshift=-1mm, algoname,draw=black!50] { heapsort };
-%     \node (ungrdpart) [below=2mm of heapsort.south west, anchor=north west, algoname,draw=black!50] { unguarded\_partition };
-%
-%     \draw[->] (heapsort.west) to[-|-=.17] (partintrosort-s.east);
-%     \draw[->] (ungrdpart.west) to[-|-=.22] (partintrosort-p.east);
-%
-%     \draw[->] (insort.west) to[-|-] (introsort-fs.east);
-%     \draw[->] (partintrosort-box.west) to[-|-] (introsort-ps.east);
-
-    \node (introsort) [algoname] { introsort };
-    \node (introsort-ps) [below=of introsort.west, depname, xshift=2mm] { part\_sort };
-    \node (introsort-fs) [below=of introsort-ps.west, depname] { final\_sort };
-    \node (introsort-box) [draw=black!50, fit={(introsort) (introsort-ps) (introsort-fs)}] {};
-
-    \node (partintrosort) [right=6mm of introsort-box.north east, yshift=-1.15mm, anchor=north west, algoname] { partial\_introsort };
-    \node (partintrosort-s) [below=of partintrosort.west, depname, xshift=2mm] { sort };
-    \node (partintrosort-p) [below=of partintrosort-s.west, depname] { partition };
-    \node (partintrosort-box) [draw=black!50, fit={(partintrosort) (partintrosort-s) (partintrosort-p)}] {};
-
-    \node (insort) [below=3mm of partintrosort-box.south west, xshift=-1mm, anchor=north west, algoname] { final\_insertion\_sort };
-    \node (insort-s) [below=of insort.west, depname, xshift=2mm] { sort\_one\_more };
-    \node (insort-box) [draw=black!50, fit={(insort) (insort-s)}] {};
-
-    \node (heapsort) [right=6mm of partintrosort-box.north east, anchor=north west, yshift=-1mm, algoname,draw=black!50] { heapsort };
-    \node (ungrdpart) [below=2mm of heapsort.south west, anchor=north west, algoname,draw=black!50] { unguarded\_partition };
-    \node (insert) [below=3.85mm of ungrdpart.south west, anchor=north west, algoname,draw=black!50] { insert };
-
-    \draw[->] (heapsort.west) to[-|-=.17] (partintrosort-s.east);
-    \draw[->] (ungrdpart.west) to[-|-=.22] (partintrosort-p.east);
-    \draw[->] (insert.west) to[-|-=.22] (insort-s.east);
-
-    \draw[->] (insort-box.west) to[-|-] (introsort-fs.east);
-    \draw[->] (partintrosort-box.west) to[-|-] (introsort-ps.east);
-  \end{tikzpicture}
-  \caption{Modular verification of Introsort. The boxes show the different
-    algorithms and the specifications they depend on. Arrows point from an algorithm to the specification that it implements.}\label{fig:modular}
-  \end{figure}
-
-  % TODO: Despite this explicit modularization, we also do implicit modularization, by inserting concrete algorithms
-  %  early, but only using their spec for the correctness proof
-
-\subsection{Unguarded Insertion Sort}\label{sec:unguarded_insort}
-  We also use refinement for the implementation of insertion sort.
-  As suggested in the pseudocode (Alg.~\ref{alg:insort}), we use a Boolean variable to discriminate
-  between the guarded and unguarded versions. Only in the last refinement step we will specialize the algorithm for both cases.
-
-  For the correctness proof, we first specify insertion as sorting one more element of an already sorted list:
-  \begin{lstlisting}
-    sort_one_more_spec G xs i ==
-      assert (i<|xs| \and sorted (xs[0..<i]));
-      assert (\<not>G --> 0<i \and \<not>(xs[i] < xs[0]));
-      spec xs'. inres (slice_sort_spec xs 0 (i+1)) xs'
-                \and (\<not>G --> xs'[0] = xs[0]))
-  \end{lstlisting}
-  Assuming the initial segment of the list up to \is{i} is already sorted,
-  afterwards, the segment up to \is{i+1} will be sorted, and the rest of the list
-  will be unchanged. We reuse \is{slice_sort_spec} for this (\is{inres m x} states that \is{x} is a possible result of \is{m}).
-  In the unguarded case (\is{\<not>G}), we assume that the first element of the list to be sorted is not greater than
-  the element to be inserted, which guarantees that insertion will stop latest at this element. Moreover, we explicitly guarantee
-  that the first element of the list will not change\footnote{This is stronger than required, as in the general case of weak orderings,
-  it would be fine to exchange the first element by an equivalent one. However, it holds for our implementation and slightly simplifies the proofs.}.
-
-  Using this specification, and its canonical refinement to slices, it is straightforward to prove correct the final
-  insertion sorting algorithm (cf.~Alg.~\ref{alg:insort}). We now focus on the actual insertion procedure.
-  In a first step, we refine the specification to a version that describes what the insertion procedure actually does,
-  rather than focusing on the sorting:
-  \begin{lstlisting}
-    insert_spec G xs i ==
-      assert (i<|xs| \and (\<not>G --> 0<i \and \<not>(xs[i] < xs[0])));
-      spec xs'. \<exists>i'<=i.
-        xs'=xs[0..<i'] @ xs[i] @ xs[i'..<i] @ xs[i<${}$..]
-      \and \<forall>j\<in>{i'<${}$..i}. xs[i] < xs'[j]
-      \and i'>0 --> \<not>(xs[i] < xs'[i'-1])
-      "
-  \end{lstlisting}
-  This precisely describes how the element at index \is{i} is moved backwards over greater elements.
-  Its new index is \is{i'}, and if this is not the beginning of the list, the element at \is{i'-1} was not greater,
-  such that the moving could be stopped.
-
-  This specification refines \is{sort_one_more_spec}:
-  \begin{theorem}
-  \begin{lstlisting}
-    insert_spec xs i <= sort_one_more_spec xs i
-  \end{lstlisting}
-  \end{theorem}
-  The proof requires some basic arguments about relating equalities on list indices to equalities on lists and sets.
-  Although straightforward in principle, it required a few auxiliary lemmas.
-  Again, refinement allows us to do this reasoning in isolation, and then focus on how to implement the moving
-  of the element.
-
-  We implement the insert procedure on whole lists first, and then refine it to slices beginning at index \is$l$.
-  We only display the version on slices here:
-  \begin{lstlisting}
-    slice_insert G xs l i ==
-        x <- xs[i];
-        (xs,i) <- while$_I$ (\<lambda>(xs,i). (\<not>G \or i>l) \and x<xs[i-1]) (\<lambda>(xs,i).
-            xs <- xs[i:=xs[i-1]]
-            return (xs,i-1) ) (xs,i)
-        xs <- xs[i:=x]
-        return xs
-  \end{lstlisting}
-%
-%   In a second refinement step, the algorithm is refined to work on slices of lists
-%
-%
-%
-%
-%   The insertion algorithm resembles the one displayed in Alg.~\ref{alg:insort}:
-%   \begin{lstlisting}
-%     insert G xs i ==
-%         x <- xs[i];
-%         (xs,i) <- while$_I$ (\<lambda>(xs,i). (\<not>G \or i>0) \and x<xs[i-1]) (\<lambda>(xs,i).
-%             xs <- xs[i:=xs[i-1]]
-%             return (xs,i-1) ) (xs,i)
-%         xs <- xs[i:=x]
-%         return xs
-%   \end{lstlisting}
-  As this is a functional program, the variables \is{xs} and \is{i} have to be threaded through the loop explicitly.
-  Moreover, instead of updating a single index, we update the whole list and rebind the updated list to the same name as
-  the old list (\is{xs <- xs[i:=x]}). Finally, the loop is annotated with an invariant $I$. It is completely standard
-  and not displayed here.
-
-  We easily prove that \is{slice_insert} refines \is{slice_insert_spec}, and, by transitivity,
-  it also refines \is{slice_sort_one_more_spec}. Thus, we can plug it into the insertion sort algorithm, which, in turn, is plugged into the main Introsort algorithm (cf.\ Fig.~\ref{fig:modular}).
-
-\subsection{Heapsort and Partition}
+%
+  \subsection{Insertion Sort}\label{sec:insort}
+  Algorithm~\ref{alg:insort} shows our implementation of insertion sort.
+  The \textsc{insort} procedure repeatedly calls \textsc{insert} to add elements to a sorted prefix of the list.
+  The flag $G$ controls the \emph{unguarded} optimization: if it is false, we assume that \textsc{insert} will hit a smaller
+  element before underflowing the list index $i$, and thus omit the comparison $l<i$ (line~\ref{alg:insert:loophead}) in the inner loop.
+  We later specialize the \textsc{insort} algorithm for the two cases of $G$, and simplify the loop conditions accordingly.
+
+%   It contains the \emph{unguarded}
+%   optimization: if parameter $G$ is false, we assume that we will hit a smaller element before underflowing the list, and thus can omit
+%   the comparison $l<i$ in line~\ref{alg:insert:loophead}. In particular when element comparisons are cheap, i.e.~have cost similar to index comparisons,
+%   omitting a comparison from the very inner loop can have a significant effect. Again, we show the version for Pdqsort here, where the unguarded optimization
+%   is used for non-leftmost partitions. The final insertion sort of Introsort uses a similar optimization, where we switch to unguarded insertion once
+%   the first threshold elements have been sorted.
+
+  Again, we split the insertion sort algorithm into two smaller parts, which are proved separately via the following
+  specification for \textsc{insert}:
+  \begin{lstlisting}
+    ASSERT (sorted xs[l..<i] \<and> l<=i \<and> i<|xs| \<and> (G \<or> xs[l-1] <= xs[i]));
+    SPEC (%xs'. xs=$_{l,i+1}$xs' \<and> sorted xs[l..<i+1])
+  \end{lstlisting}
+  This captures the intuition that \textsc{insert} goes from a slice that is sorted up to index $i$ to one that is sorted up to index $i+1$.
+
+\subsection{The Remaining Subroutines}
   \newcommand{\leftc}{\textrm{left}}
   \newcommand{\rightc}{\textrm{right}}
-
-  The proofs of heapsort and partitioning follow a similar plot, and are not displayed here in full.
-  We used some existing Isabelle proofs as guideline \cite{LaWi19-IMP2,Gri19-AFP,La16}.
-  However, we point out one more interesting application of refinement: recall the sift-down function on heaps,
-  that restores the heap property by \emph{floating down} an element\footnote{see, e.g., \cite[Ch.~6]{CLRS09} or \cite[Ch.~2.4]{SeWa11} for a description of heapsort.}.
-Usually, this function is presented by swapping the element with one of its children, until the heap property is restored (Alg.~\ref{alg:sift_down_swap}).
+  \newcommand{\hasleftc}{\textrm{has\_left}}
+  \newcommand{\hasrightc}{\textrm{has\_right}}
+
+  The proofs of the remaining subroutines follow a similar plot, and are not displayed here in full.
+  Most of them were straightforward, and we could use existing Isabelle proofs as guideline \cite{LaWi19-IMP2,Gri19-AFP,La16}.
+  For the \textsc{shuffle} and \textsc{pivot\_to\_front} procedures, which contain a large number of indexing and update operations,
+  we ran into a scalability problem: the many partially redundant in-bound statements for
+  the indexes overwhelmed the linear arithmetic solver that is hard-wired into the simplifier.
+  We worked around this problem by introducing auxiliary definitions, which hide the in-bound statements from the simplifier,
+  and allow us to precisely control when it sees them.
+
+  Finally, we point out another interesting application of refinement: the \textsc{sift\_down} function of heapsort
+  restores the heap property by \emph{floating down} an element\footnote{see, e.g., \cite[Ch.~6]{CLRS09} or \cite[Ch.~2.4]{SeWa11} for a description of heapsort.}.
+  A straightforward implementation swaps the element with one of its children, until the heap property is restored (Alg.~\ref{alg:sift_down} (left)).
   \begin{algorithm}
+    \begin{minipage}[t]{.49\textwidth}
     \begin{algorithmic}
-    \Procedure{sift\_down}{$a,i$}
-      \While{has right child $i$}
-        \If{$a[\leftc~i] < a[\rightc~i]$}
-          \If{$a[i] < a[\rightc~i]$}
-            \State{\Call{swap}{$a[i], a[\rightc~i]$}}
-            \State{$i\gets \rightc~i$}
+    \Procedure{sift\_down}{$xs,i$}
+      \State{}
+      \While{$\hasrightc(i)$}
+        \If{$xs[\leftc(i)] < xs[\rightc(i)]$}
+          \If{$xs[i] < xs[\rightc(i)]$}
+            \State{\Call{swap}{$xs[i], xs[\rightc(i)]$}}
+            \State{$i\gets \rightc(i)$}
           \Else~\Return{}
           \EndIf
-        \ElsIf{$a[i] < a[\leftc~i]$}
-          \State{\Call{swap}{$a[i], a[\leftc~i]$}}
-          \State{$i\gets \leftc~i$}
+        \ElsIf{$xs[i] < xs[\leftc(i)]$}
+          \State{\Call{swap}{$xs[i], xs[\leftc(i)]$}}
+          \State{$i\gets \leftc(i)$}
         \Else~\Return{}
@@
       \EndWhile
-      \If{has left child $i$ and $a[i] < a[\leftc~i]$}
-        \State{\Call{swap}{$a[i], a[\leftc~i]$}}
+      \If{$\hasleftc(i) \wedge xs[i] < xs[\leftc(i)]$}
+        \State{\Call{swap}{$xs[i], xs[\leftc(i)]$}}
       \EndIf
@@
     \end{algorithmic}
-  \caption{Pseudocode of sift-down function with swap}\label{alg:sift_down_swap}
-  \end{algorithm}
-  \begin{algorithm}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[t]{.49\textwidth}
     \begin{algorithmic}
-    \Procedure{sift\_down\_opt}{$a',i$}
-      \State{$t\gets a'[i]$}
-      \While{has right child $i$}
-        \If{$a'[\leftc~i] < a'[\rightc~i]$}
-          \If{$t < a'[\rightc~i]$}
-            \State{$a'[i]\gets a'[\rightc~i]$}
-            \State{$i\gets \rightc~i$}
+    \Procedure{sift\_down\_opt}{$xs',i$}
+      \State{$t\gets xs'[i]$}
+      \While{$\hasrightc(i)$}
+        \If{$xs'[\leftc(i)] < xs'[\rightc(i)]$}
+          \If{$t < xs'[\rightc(i)]$}
+            \State{$xs'[i]\gets xs'[\rightc(i)]$}
+            \State{$i\gets \rightc(i)$}
           \Else~\Return{}
           \EndIf
-        \ElsIf{$t < a[\leftc~i]$}
-          \State{$a'[i]\gets a'[\leftc~i]$}
-          \State{$i\gets \leftc~i$}
+        \ElsIf{$t < xs[\leftc(i)]$}
+          \State{$xs'[i]\gets xs'[\leftc(i)]$}
+          \State{$i\gets \leftc(i)$}
         \Else~\Return{}
@@
       \EndWhile
-      \If{has left child $i$ and $t < a'[\leftc~i]$}
-        \State{$a'[i]\gets a'[\leftc~i]$}
-        \State{$i\gets \leftc~i$}
-      \EndIf
-      \State{$a'[i]\gets t$}
+      \If{$\hasleftc(i) \wedge t < xs'[\leftc(i)]$}
+        \State{$xs'[i]\gets xs'[\leftc(i)]$}
+        \State{$i\gets \leftc(i)$}
+      \EndIf
+      \State{$xs'[i]\gets t$}
     \EndProcedure
     \end{algorithmic}
-  \caption{Pseudocode of optimized sift-down function}\label{alg:sift_down_opt}
+    \end{minipage}
+    \caption{The standard (left) and optimized (right) sift-down function.}\label{alg:sift_down}
   \end{algorithm}
-  However, the element that is written to $a[\rightc~i]$ or $a[\leftc~i]$ by the swap will get overwritten in the next loop iteration.
-  Thus, a common optimization to save half of the writes is to store the element to be moved down in a temporary variable, and only
-  assign it to its final position after the loop (Alg.~\ref{alg:sift_down_opt}). Note that the insertion procedure of insertion sort does a similar optimization.
-  However, for the insertion procedure, it was feasible to prove the optimization together with the actual algorithm.
+  However, the element that is written to $xs[\rightc(i)]$ or $xs[\leftc(i)]$ by the swap will get overwritten in the next iteration.
+  A common optimization to save half of the writes is to store the element to be moved down in a temporary variable, and only
+  assign it to its final position after the loop (Alg.~\ref{alg:sift_down} (right)). Note that the \textsc{insert} procedure of insertion sort
+  (cf.~Alg.~\ref{alg:insort}) does a similar optimization.
+  However, for \textsc{insert}, it was feasible to prove the optimization together with the actual algorithm.
   For the slightly more complicated sift-down procedure, we first prove correct the simpler algorithm with swaps,
-  and then refine it to the optimized version. Inside the loop, the refinement relation between the concrete array \is{a'} and the
-  abstract array \is{a} states that
-  we obtain \is{a} as \is{a'[i:=t]}. The proof that the optimized version refines the version with swaps requires only about 20
-  lines of straightforward Isabelle script.
-
-\section{Imperative Implementation}
-  In the previous sections, we have presented a refinement based approach to verify the Introsort algorithm,
-  including most optimizations we found in its GNU C++ Library implementation. However, the algorithm is still expressed
-  as a nondeterministic monadic program that modifies functional lists. In this section, we discuss how the Isabelle LLVM framework~\cite{La19-llvm} is
-  used to (semi-)automatically transfer the algorithm into an LLVM program on arrays.
+  and then refine it to the optimized version. Inside the loop, the refinement relation
+  between the abstract list \is{xs} and the concrete list \is{xs'} is \is{xs = xs'[i:=t]}.
+  Using the tool support of the Isabelle Refinement Framework, the proof that
+  the optimized version refines the version with swaps requires only about 20 lines of straightforward Isabelle script.
+
+\section{Imperative Implementation}\label{sec:imp_impl}
+  We have presented a refinement based approach to verify the Introsort and Pdqsort algorithms,
+  including most optimizations we found in their libstdc++ and Boost implementations. However, the algorithms are still expressed
+  as nondetermistic programs on functional lists and unbounded natural numbers.
+  In this section, we use the Isabelle-LLVM framework~\cite{La19-llvm} to (semi-)automatically
+  refine them to LLVM programs on arrays and 64 bit integers.
 
   \subsection{The Sepref Tool}
-  The idea of the Sepref tool~\cite{La15,La19-llvm} is to symbolically execute an abstract program in the \is{nres}-monad,
-  keeping track of refinements of every abstract variable to a concrete representation that may use pointers to dynamically allocated memory.
-  During this symbolic execution, the tool can synthesize an imperative program in the Isabelle-LLVM semantics, together with a refinement proof.
-  The tool works automatically, but usually requires some program-specific setup and boilerplate.
-For a detailed discussion of the Sepref tool and the Isabelle-LLVM framework, we refer the reader to~\cite{La15,La19-llvm}.
-
-  Isabelle-LLVM comes with standard setup to refine lists to arrays. List update operations are refined to destructive array element updates,
-  as long as the old version of the abstract list is not used any more after the update. Note that, by rebinding the updated list to the same
-  variable name as the original list\footnote{as e.g.\ in \is{xs <- xs[i:=x]}, cf.\ Sec.~\ref{sec:unguarded_insort}}, this property is syntactically ensured.
-Moreover, Isabelle-LLVM provides setup to refine the unbounded integers of Isabelle/HOL to bounded integers.
-  The resulting proof obligations to show that the integers are in bounds are discharged automatically, but typically require some hinting from the user.
+  The the Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{nres}-monad,
+  keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
+  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
+  The synthesis is automatic, but usually requires some program-specific setup and boilerplate. For a detailed discussion of Sepref and Isabelle-LLVM,
+  we refer the reader to~\cite{La15,La19-llvm}.
+
+  Sepref comes with standard setup to refine lists to arrays. List updates are refined to destructive array updates,
+  as long as the old version of the list is not used after the update.
+  It also provides setup to refine unbounded natural numbers to bounded integers.
+  It tries to discharge the resulting in-bounds proof obligations automatically.
+  If this is not possible, it relies on hints from the user.
+%   Note that, by rebinding the updated list to the same
+%   variable name as the original list\footnote{as e.g.\ in \is{xs <- xs[i:=x]}, cf.\ Sec.~\ref{sec:unguarded_insort}}, this property is syntactically ensured.
+%   Moreover, Isabelle-LLVM provides setup to refine the unbounded integers of Isabelle/HOL to bounded integers.
+%   Sepref tries to discharge the resulting in-bounds proof obligations automatically. Where this is not possible, it relies on hints from the user.
+
   A common technique to provide such hints is to insert additional assertions into the abstract program.
-  Usually, these can be proved easily. For example, in the definition of \is{slice_partial_introsort} (Sec.~\ref{sec:refinement}),
-  we have inserted the assertion \is{l<=h} at the very beginning. This is required by the Sepref tool to prove that the operation \is{h-l}
-  in the next line cannot underflow.
-This assertion adds a proof obligation to the proof that \is{slice_partial_introsort} refines \is{slice_part_sort_spec} (Thm.~\ref{thm:slice_partial_introsort_correct}), which is, however, trivial, as the specification
-  explicitly states \is{l<=h} as precondition. On the other hand, when refining \is{slice_partial_introsort} to an implementation with bounded
-  integers, one can assume \is{l<=h}, thus discharging the non-underflow proof obligation for the \is{h-l} operation.
+  Usually, these can be proved easily. For example, in the \textsc{pdqsort\_aux} algorithm (Alg.~\ref{alg:pdqsort}, line~\ref{l:pdq:assert}), the
+  assertion \is{m+1<=h} ensures that the addition \is$m+1$ in the next line cannot overflow.
+  This assertion adds a proof obligation to the correctness proof of \textsc{pdqsort\_aux},
+  which is easily discharged (we are in phase \is{part}, which guarantees \is$m<h$).
+  When refining \textsc{pdqsort\_aux} to an implementation with bounded
+  integers, one can assume \is{m+1<=h} to discharge the non-overflow proof obligation.
+  Note that re-proving \is{m+1<=h} when doing the refinement would require duplicating large parts of the correctness proof.
   Thus, assertions provide a convenient tool to pass properties down the refinement chain.
-
-  Using the Sepref tool, it is straightforward to refine the Introsort algorithm to an Isabelle-LLVM program that sorts arrays of integers.
-  For example, Figure~\ref{fig:ug_insort_isa_llvm} shows the Isabelle-LLVM code that is generated for the \is{insert} procedure with \is{G=false} (cf.~\ref{sec:unguarded_insort}).
-  Moreover, the Sepref tool will generate a theorem that the generated program actually implements the abstract one:
-  \begin{theorem}
-    \begin{lstlisting}
-      (ug_insert_impl, slice_insert False)
-        : arr_u64$^d$ \x snat64$^k$ \x snat64$^k$ -> arr_u64
-    \end{lstlisting}
-  \end{theorem}
-  This specifies the relations for the parameters and the result, where \is{arr_u64} relates arrays of unsigned 64-bit integers with lists of integers,
-  and \is{snat64} relates non-negative 64-bit integers with natural numbers.
+  Our actual formalization contains multiple such assertions, which we have omitted in this presentation for the sake of readability.
+
+%   This assertion adds a proof obligation to the correctness proof of \textsc{pdqsort\_aux}, which is trivial at this point, as $l\le h$ is a precondition of \is{sort_spec} (cf.~Sec.~\ref{sec:sortspec}).
+%   On the other hand, when refining \textsc{pdqsort\_aux} to an implementation with bounded
+%   integers, one can assume \is{l<=h}, thus discharging the non-underflow proof obligation for the \is{h-l} operation.
+%   Thus, assertions provide a convenient tool to pass properties down the refinement chain.
+%   Note that our actual formalization contains multiple such assertions, which we have omitted in this presentation for the sake of readability.
+
+  \begin{figure}
+    \begin{minipage}{.5\textwidth}
+    \begin{lstlisting}
+      ug_insert_impl == %a l i. doM {
+          x <- array_nth a i;
+          (a, i) <- llc_while (%(a, i). doM {
+              bi <- ll_sub i 1;
+              t <- array_nth a bi;
+              ll_icmp_ult x t
+          }) (%(a, i). doM {
+              i' <- ll_sub i 1;
+              t <- array_nth a i';
+              a <- array_upd a i t;
+              i <- ll_sub i 1;
+              return (a, i)
+          }) (a, i);
+          array_upd a i x
+        }
+    \end{lstlisting}
+    \end{minipage}
+    \begin{minipage}{.49\textwidth}
+    \caption{Implementation of the \textsc{insert} procedure for 64bit unsigned integer elements and \is{G=false}, which
+      is generated by the Sepref tool.
+      This definition lies within the executable fragment of Isabelle-LLVM, i.e.,
+      the Isabelle LLVM code generator can translate it to LLVM intermediate representation.
+      Note that the function does not
+      depend on the lower bound parameter \is$l$ any more, as this was only required in the guarded version.
+      Inlining will remove this bogus parameter.
+    }\label{fig:ug_insort_isa_llvm}
+    \end{minipage}
+  \end{figure}
+  Using the Sepref tool, it is straightforward to refine the sorting algorithms and their subroutines to an Isabelle-LLVM program.
+  For example, Figure~\ref{fig:ug_insort_isa_llvm} shows the Isabelle-LLVM code that is generated for the \textsc{insert} procedure
+  for unsigned 64 bit integer elements and \is{G=false} (cf.~\ref{sec:insort}).
+  Moreover, the Sepref tool proves that the generated program actually implements the abstract one:
+  \begin{lstlisting}
+    (ug_insert_impl, insert False) : arr$^d$ \x nat$_{64}^k$ \x nat$_{64}^k$ -> arr
+  \end{lstlisting}
+  This specifies the refinement relations for the parameters and the result, where \is{arr} relates arrays with lists,
+  and \is{nat$_{64}$} relates 64-bit integers with natural numbers.
   The $\cdot^d$ annotation means that the parameter will be \emph{destroyed} by the function call,
   while $\cdot^k$ means that the parameter is \emph{kept}.
-  Here, the insertion is done in place, such that the input list has no corresponding refinement after the function call.
-
-  The final correctness statement for our Introsort implementation (for integers) is
+  Here, the insertion is done in place, such that the original array is destroyed.
+
+  The final correctness statement for our implementations is:
   \begin{theorem}
     \begin{lstlisting}
-      (introsort_impl, slice_sort_spec)
-        : arr_u64$^d$ \x snat64$^k$ \x snat64$^k$ -> arr_u64
+      (introsort_impl, sort_spec) : arr$^d$ \x nat$_{64}^k$ \x nat$_{64}^k$ -> arr
+                 $\textbf{and}$  (pdqsort_impl, sort_spec)  : arr$^d$ \x nat$_{64}^k$ \x nat$_{64}^k$ -> arr
     \end{lstlisting}
   \end{theorem}
-  Here, \is$introsort_impl$ is the Isabelle-LLVM program generated by Sepref from \is$introsort$ (cf.\ Sec.\ \ref{sec:refinement}).
-  This theorem combines the refinement theorem generated by Sepref, and Theorem~\ref{thm:introsort_correct}.
-
-  \begin{figure}
-    \begin{lstlisting}
-      ug_insert_impl == %xs l i. doM {
-          x <- array_nth xs i;
-          (xs, i) <- llc_while (%(xs, i). doM {
-              bi <- ll_sub i 1;
-              t <- array_nth xs bi;
-              ll_icmp_ult x t
-          }) (%(xs, i). doM {
-              i' <- ll_sub i 1;
-              t <- array_nth xs i';
-              xs <- array_upd xs i t;
-              i <- ll_sub i 1;
-              return (xs, i)
-          }) (xs, i);
-          array_upd xs i x
-        }
-    \end{lstlisting}
-    \caption{Implementation for \is{insert} procedure with \is{G=false} generated by the Sepref tool, for 64bit signed integer elements.
-      This definition is within the executable fragment of Isabelle-LLVM, i.e.,
-      the Isabelle LLVM code generator can, after preprocessing, translate it to LLVM-IR. Moreover, note that the function does not
-      depend on the lower bound parameter \is$l$ any more, as this was only required in the guarded version.
-    }\label{fig:ug_insort_isa_llvm}
-  \end{figure}
-
-  \begin{figure}
+  Here, \is$introsort_impl$ and \is$pdqsort_impl$ are the Isabelle-LLVM programs generated by Sepref from \textsc{introsort} and \textsc{pdqsort} (Algs.\ \ref{alg:introsort} and \ref{alg:pdqsort}).
+  The theorem is easily proved by combining Theorem~\ref{thm:sorting_correct} with the theorems generated by Sepref.
+
+  \subsection{Separation Logic and Ownership}
+  Internally, the Sepref tool represents the symbolic state that contains all abstract variables and their refinements to concrete variables
+  as an assertion in separation logic~\cite{Rey02,CHY07}. Thus, two variables can never reference the same memory.
+  This is a problem for nested container data structures like arrays of strings: when indexing the array, both the array element and the result of
+  the indexing operation would point to the same string.
+  In the original Sepref tool~\cite{La15}, which targeted Standard ML, we worked around this problem by always using
+  functional data types (e.g. lists) to represent the inner type of a nested container. This workaround is no longer applicable for the purely imperative LLVM, such that
+  we could not use Sepref for nested container data structures\footnote{We could still reason about such structures on a lower level.}.
+
+  We now describe an approach towards solving this problem for Sepref.
+  Abstractly, we model an array by the type \is{\<alpha> option list}\footnote{Here, \is$\\<alpha> option = None | Some \\<alpha>$ is Isabelle's option datatype,
+  and \is{the (Some x) == x} is the corresponding selector function.},
+  where \is{None} means that the array does currently not own the respective element.
+  The abstract indexing operation then moves the
+  element from the list to the result:
+  \begin{lstlisting}
+    move xs i == assert (i<|xs| \and xs[i]~=None); return (the (xs[i]), xs[i:=None])
+  \end{lstlisting}
+%   where \is{the (Some x) = x} is the selector function of the option data type.
+  As no memory is shared between the result and the array, we can show the following refinement:
+  \begin{lstlisting}
+    (%a i. (a[i],a), %xs i. move xs i) : oarr$^d$ \x nat$_{64}^k$ -> A \x oarr
+  \end{lstlisting}
+  where \is{oarr} is the relation between an array and an \is{\<alpha> option list}, and \is$A$ is the relation for the array elements.
+  Note that this operation does not change the concrete array \is{a}. The movement of ownership is a purely abstract concept, which results
+  in no implementation overhead.
+
+  The transition from \is{\<alpha> list} to \is{\<alpha> option list} can typically be done in an additional refinement step,
+  and thus does not obfuscate the actual correctness proofs, which are still done on plain \is{\<alpha> list}.
+  Moreover, the \is{\<alpha> option list} representation is only required for subroutines where extracted array elements are actually visible.
+  For example, we define an operation to compare two array elements:
+  \begin{lstlisting}
+    cmp_idxs xs i j == assert (i<|xs| \<and> j<|xs|); return xs[i] < xs[j]
+  \end{lstlisting}
+  Inside this operation, we have to temporarily extract the elements $i$ and $j$ from the array,
+  requiring an intermediate refinement step to \is{\<alpha> option list}.
+  However, at the start and end of this operation, the array owns all its elements. For the whole operation, we thus get
+  a refinement on plain arrays:
+  \begin{lstlisting}
+    (cmp_impl, cmp_idxs) : arr$^k$ \x nat$_{64}^k$ \x nat$_{64}^k$ -> bool
+  \end{lstlisting}
+  In our case, we only have to explicitly refine \textsc{insert} and \textsc{sift\_down}.
+  The other subroutines use \is{cmp_idxs} and \is{swap} operations on plain lists.
+
+  \subsection{The Isabelle-LLVM Code Generator}
+  \newcommand{\lis}{\lstinline[language=llvm,literate={}]}
+
+  The programs that are generated by Sepref (cf.\ Fig.~\ref{fig:ug_insort_isa_llvm}) lie in the fragment for which Isabelle-LLVM~\cite{La19-llvm}
+  can generate LLVM text.
+  For example, the Pdqsort algorithm for strings yields an LLVM function with the signature:
+%   Isabelle-LLVM comes with a code generator that exports this program to actual LLVM text, which is subsequently compiled using the LLVM toolchain.
+%   For example, the Pdqsort algorithms for strings yields an LLVM function with the signature:
   \begin{lstlisting}[language=llvm,literate={}]
-      define i64* @ug_insert_impl(i64* %xs, i64 %l, i64 %i) {
-      start:
-        %x3 = getelementptr i64, i64* %xs, i64 %i
-        %r = load i64, i64* %x3, align 8
-        br label %while_start
-
-      while_start:
-        %bib.sink = phi i64 [ %bib, %while_body ], [ %i, %start ]
-        %bib = add i64 %bib.sink, -1
-        %xda = getelementptr i64, i64* %xs, i64 %bib
-        %ra = load i64, i64* %xda, align 8
-        %xg = icmp ult i64 %r, %ra
-        %p1 = getelementptr i64, i64* %xs, i64 %bib.sink
-        br i1 %xg, label %while_body, label %while_end
-
-      while_body:
-        store i64 %ra, i64* %p1, align 8
-        br label %while_start
-
-      while_end:
-        store i64 %r, i64* %p1, align 8
-        ret i64* %xs
-      }
-    \end{lstlisting}
-    \caption{LLVM-IR generated by Isabelle-LLVM for the definition from Figure~\ref{fig:ug_insort_isa_llvm}.
-    We display the code after an LLVM optimization pass that eliminates some redundant operations generated by Isabelle-LLVM,
-    and thus makes the code more readable. Note that we kept the unused parameter \%l. LLVM will eliminate it by inlining the function.
-    }\label{fig:ug_insort_llvm}
-  \end{figure}
-
-  \subsection{The Isabelle-LLVM Code Generator}
-  Isabelle-LLVM comes with a code generator that exports this program to actual LLVM text, which is subsequently compiled using the LLVM toolchain.
-  Fig.~\ref{fig:ug_insort_llvm} shows the code generated for the \is{slice_insert} procedure.
-
-  To make the generated programs usable, one has to link them to C wrappers, that handle parsing of command line options and printing of results.
-  However, the original Isabelle-LLVM framework provides no support for this: one has to manually write a C header file, that hopefully matches the object file generated by the LLVM compiler.
+  { i64, { i64, i8* } }* @str_pdqsort({ i64, { i64, i8* } }*, i64, i64)
+  \end{lstlisting}
+  Here, the type \lis${ i64, { i64, i8* } }$ represents dynamic arrays of characters\footnote{%
+    For strings, we use the verified dynamic array implementation provided by Isabelle-LLVM.
+    Note that C++ uses a similar representation, with an additional optimization for small strings.
+  }, represented by length, capacity, and a data pointer.
+
+%   Fig.~\ref{fig:ug_insort_llvm} shows the code generated for the \is{slice_insert} procedure.
+%
+%   \begin{figure}
+%     \begin{minipage}{.7\textwidth}
+%     \begin{lstlisting}[language=llvm,literate={}]
+%       define i64* @ug_insert_impl(i64* %xs, i64 %l, i64 %i) {
+%       start:
+%         %x3 = getelementptr i64, i64* %xs, i64 %i
+%         %r = load i64, i64* %x3, align 8
+%         br label %while_start
+%
+%       while_start:
+%         %bib.sink = phi i64 [ %bib, %while_body ], [ %i, %start ]
+%         %bib = add i64 %bib.sink, -1
+%         %xda = getelementptr i64, i64* %xs, i64 %bib
+%         %ra = load i64, i64* %xda, align 8
+%         %xg = icmp ult i64 %r, %ra
+%         %p1 = getelementptr i64, i64* %xs, i64 %bib.sink
+%         br i1 %xg, label %while_body, label %while_end
+%
+%       while_body:
+%         store i64 %ra, i64* %p1, align 8
+%         br label %while_start
+%
+%       while_end:
+%         store i64 %r, i64* %p1, align 8
+%         ret i64* %xs
+%       }
+%     \end{lstlisting}
+%     \end{minipage}
+%     \begin{minipage}{.29\textwidth}
+%     \caption{LLVM-IR generated by Isabelle-LLVM for the definition from Figure~\ref{fig:ug_insort_isa_llvm}.
+%     We display the code after an LLVM optimization pass that eliminates some redundant operations generated by Isabelle-LLVM,
+%     and thus makes the code more readable. Note that we kept the unused parameter \%l. LLVM will eliminate it by inlining the function.
+%     }\label{fig:ug_insort_llvm}
+%     \end{minipage}
+%   \end{figure}
+
+  The generated LLVM text is then compiled to machine code using the LLVM toolchain.
+  To make the generated program usable, one has to link it to a C wrapper, which handles parsing of command line options and printing of results.
+  However, the original Isabelle-LLVM framework provides no support for interfacing the generated code from C: one has to manually write a C header file, which hopefully matches the object file generated by the LLVM compiler.
 If it doesn't, the program has undefined behaviour\footnote{In practice, this means it will probably SEGFAULT.
 However, it also might return wrong results, or be prone to various kinds of exploits.}.
-What makes the situation even worse is, that the C compiler sometimes changes function signatures: For example, a function with a non-primitive return value (e.g.
+
+% > Could not reproduce this behaviour in an isolated test, so commenting out for now!
+%   What makes the situation even worse is, that the C compiler sometimes changes function signatures: For example, a function with a non-primitive return value (e.g.
 a structure) is changed to a void function with an extra
-  reference parameter for the return value.
+%   reference parameter for the return value.
 LLVM, on the other hand, does not perform this change, which makes the function generated by LLVM incompatible with the function call generated by the C compiler.
 
-  To this end, we extended Isabelle-LLVM to also generate a header file for the exported functions, refusing to export functions whose signatures
-  would be altered by the C compiler. Figure~\ref{fig:export} shows the extended Isabelle LLVM export command and the generated header file for our Introsort algorithm on strings.
-  \begin{figure}
-    \begin{flushleft}\bfseries\color{black!50} Isabelle command\end{flushleft}
-    \begin{lstlisting}
-      export_llvm
-        str_sort_introsort_impl
+  To this end, we extended Isabelle-LLVM to also generate a header file for the exported functions. For example, the Isabelle command
+  {\small\begin{lstlisting}
+    export_llvm str_sort_introsort_impl
       is llstring* str_introsort(llstring*, int64_t, int64_t)
-        defines \<open>
-          typedef struct {
+      defines \<open> type${}$def struct {
           int64_t size; struct {int64_t capacity; char *data;};
         } llstring; \<close>
-    \end{lstlisting}
-    \medskip
-
-    \begin{flushleft}\bfseries\color{black!50} Generated C header\end{flushleft}
-    \begin{lstlisting}[language=C]
-      typedef struct {
-        int64_t size;
-        struct {
-          int64_t capacity;
-          char*data;
-        };
-      } llstring;
-
+  \end{lstlisting}}
+  \noindent will check that the specified signature actually matches the Isabelle definition, and generate the following
+  C declarations:
+  {\small\begin{lstlisting}[language=C]
+    typedef struct { int64_t size; struct { int64_t capacity; char *data; }; } llstring;
     llstring* str_introsort(llstring*, int64_t, int64_t);
-    \end{lstlisting}
-
-    \caption{An example Isabelle command to export LLVM code for a string sorting function, and the generated header file.
-    Strings are represented as dynamic arrays of characters.
-    Isabelle will refuse to export functions whose signatures
-    would be changed by C. Moreover, structure fields in LLVM are indexed by consecutive numbers, rather than named by meaningful names.
-    Our extension allows to define properly named structures for the header file, and checks that these definitions are actually compatible with
-    the LLVM signatures.
-    }\label{fig:export}
+  \end{lstlisting}}
+
+  \newcommand{\plotintrosort}{
+  \begin{tikzpicture}
+    \begin{axis}[
+      xlabel near ticks,
+      legend style = {
+%         legend pos= north east,
+        at = {(.77,.97)},
+        cells={anchor=west},
+        font=\scriptsize
+      },
+      title={\large Introsort},
+      title style={at={(0.15,.85)}},
+      ybar=0pt,
+      width=\textwidth,
+      height=.4\textheight,
+      bar width=1.8pt,
+      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,}
+      ,      xtick=data,
+      x tick label style={rotate=45,anchor=east,font=\tiny},
+    ]
+
+      \input{data_introsort}
+
+    \end{axis}
+  \end{tikzpicture}
+  }
+
+  \newcommand{\plotpdqsort}{
+  \begin{tikzpicture}
+    \begin{axis}[
+      xlabel near ticks,
+      legend style = {
+        legend pos= north east,
+        cells={anchor=west},
+        font=\scriptsize
+      },
+      title={\large Pdqsort},
+      title style={at={(0.15,.85)}},
+      ybar=0pt,
+      width=\textwidth,
+      height=.4\textheight,
+      bar width=1.8pt,
+      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,}
+      ,      xtick=data,
+      x tick label style={rotate=45,anchor=east,font=\tiny},
+    ]
+
+      \input{data_pdqsort}
+
+    \end{axis}
+  \end{tikzpicture}
+
+  }
+
+\section{Benchmarks}
+%   We have verified the libstdc++ implementation of Introsort, and the Boost implementation of Pdqsort.
+%   We benchmarked our verified implementations against the unverified originals.
+%   As we did not verify the C++ code directly, re-implemented the algorithm using Isabelle-LLVM, we have to
+%   test how close we came to the performance of the original.
+
+  The Boost library comes with a sorting algorithm benchmark suite, which we extended with further benchmarks
+  indicated in~\cite{Bent93}: apart from sorting random lists of elements that are mostly different (random),
+  we also sort lists of length $n$ that contain only $n/10$ different elements (random-dup-10),
+  and lists of only two different elements (random-boolean), as well as lists where all elements are equal (equal).
+  We also consider already sorted sequences (sorted, rev-sorted), as well as a sequence of $n/2$ elements in ascending order, followed by the
+  same elements in descending order (organ-pipe). We also consider sorted sequences where we applied $pn/100$ random swap operations (almost-sorted-$p$).
+  Finally, we consider sorted sequences with $pn/100$ random elements inserted at the end or in the middle ([rev-]sorted-end-$p$, [rev-]sorted-middle-$p$).
+
+  We sorted integer arrays with $n=10^8$ elements, and string arrays with $n=10^7$ elements.
+  For strings, all implementations use the same data structure and compare function.
+  For integers, we disable Pdqsorts branch-aware partitioning, which we have not yet verified.
+  For strings, it does not apply anyway.
+
+  We compile both, the verified and unverified algorithms with clang-6.0.0, and run them on a laptop with an Intel(R) Core(TM) i7-8665U CPU and 32GiB of RAM,
+  as well as on a server machine with 24 AMD Opteron 6176 cores and 128GiB of RAM.
+  Ideally, the same algorithm should take exactly the same time when repeatedly run on the same data and machine. However, in practice,
+  we encountered some noise up to 17\%. Thus, we have repeated each experiment at least ten times, and more often to confirm outliers
+  where the verified and unverified algorithms' run times differ significantly. Assuming that the noise only slows down an algorithm,
+  we take the fastest time measured over all repetitions.
+  The results are displayed in Figure~\ref{fig:benchres}.
+%   The charts below compare the run-times for the verified and unverified algorithm.
+  \begin{figure}
+  \plotpdqsort
+  \plotintrosort
+  \caption{Benchmarking our verified implementations against the unverified originals.
+    For each element type, machine, and distribution, the value $(t_1/t_2-1)*s$ is shown,
+    where $t_1$ is the slower time, $t_2$ is the faster time, and $s = 100$ if the unverified algorithm is faster,
+    and $s=-100$ if the verified algorithm is faster.
+  }\label{fig:benchres}
   \end{figure}
 
-  \subsection{Separation Logic and Ownership}
-  Internally, the Sepref tool represents the symbolic state, which contains all abstract variables and their refinements to concrete variables,
-  as an assertion in separation logic~\cite{Rey02,CHY07}. For example, the assertion
-  \begin{lstlisting}
-    array xs_1 p_1 * array xs_2 p_2
-  \end{lstlisting}
-  describes a symbolic state that refines the two abstract list variables \is{xs_1} and \is{xs_2} by
-  two arrays at addresses \is{p_1} and \is{p_2}. Separation logic guarantees that the arrays at \is{p_1} and \is{p_2}
-  share no common memory, i.e.,
-  updates to the first array cannot change the second array. As long as the array elements are
-  primitive values like integers, this works as expected, and a list of integers can be refined
-  to an array of integers. An array lookup operation will simply return a copy of the integer
-  value at the specified index. It can be specified by the following Hoare-triple, which can
-  be used by Sepref to synthesize LLVM code for the operation \is{r <- xs[i]}.
-  \begin{lstlisting}
-    { array xs p * i<|xs| } r = p[i] { array xs p * r=xs[i] }
-  \end{lstlisting}
-  Note that we use C like notation for the program, and abstracted away the fact that the \is{i}
-  in the program is a bounded integer register from LLVM, while the \is{i} in the assertions refers to
-  its content as an unbounded Isabelle integer.
-
-  Now, let's assume we have a list of strings, where a string is itself a list of characters.
-  The following assertion refines both, the outer and the inner lists to arrays:
-  \begin{lstlisting}
-    sarray xs p == \<exists>ys.
-      |ys|=|xs| * array ys p * array xs_1 ys_1 * ... array xs$_n$ ys$_n$
-  \end{lstlisting}
-  Here, \is{ys} represents the list of string pointers, which point to their own separate array each.
-  The canonical Hoare triple for array lookup would be:
-  \begin{lstlisting}
-    { sarray xs p * i<|xs| } r = p[i] { sarray xs p * array (xs[i]) r }
-  \end{lstlisting}
-  However, this does not hold, as the inner array at position \is{i} is now pointed to twice:
-  by \is{r} and by the \is{ys[i]} from within the \is{sarray} assertion. The only way to make a
-  Hoare triple of this form valid is to make a copy of the inner array, which would come with an
-  unacceptable performance penalty.
-
-  The Rust Programming Language~\cite{rustlang} has to solve a similar problem.
-  There, they use the concept of borrowing: the inner array is \emph{borrowed} from the outer one,
-  i.e., ownership is transferred to the variable \is{r}. Only at the end of the lifetime of \is{r},
-  ownership will return to the outer array\footnote{Actually, borrowing in Rust does not work for array elements,
-  as the type-checking would become undecidable. Fortunately, in an interactive theorem prover, we can burden the
-  user with parts of this type checking.}. We realize a similar concept in Sepref, by refining abstract lists of
-  optional values to arrays, that only own those elements which are present (not None) in the abstract list\footnote{For simplicity, the presentation sticks to the array of strings example.
-Actually, our implementation is parameterized over the refinement for the inner type.
-  We believe that a further generalization to include outer types other than lists (e.g.\ maps) is possible, but leave it for future work.}:
-  \begin{lstlisting}
-    oarray xs p == \<exists>ys. array ys p * A' xs_1 ys_1 * ... * A' xs$_n$ ys$_n$
-    $\textrm{where }$
-    A' None _ = \<box>  |  A' (Some x) y = array x y
-  \end{lstlisting}
-  We define abstract operations to borrow and return (insert) elements, as well as to convert between a
-  standard list and a list with explicit ownership:
-  \begin{lstlisting}
-    borrow xs i == assert (i<|xs| \and xs[i]~=None);
-                  return (the (xs[i]), xs[i:=None])
-    ins xs i x == assert (i<|xs| \and xs[i]=None);
-                  return xs[i:=Some x]
-    to_olist xs = return (map Some xs)
-    to_slist xs = assert (None \<notin> xs); return (map the xs)
-  \end{lstlisting}
-  Here, \is{the (Some x) = x} is the selector function of the option data type.
-  The \is{borrow} function assumes that the element is still present in the list, and then returns the element
-  and a new list where the element is gone. Symmetrically, the \is{ins} function assumes that there is no element
-  present at the specified index. The \is{to_olist} and \is{to_slist} functions convert between lists of plain values and lists
-  of optional values, where \is{to_slist} assumes that all values in the list are actually present.
-
-  This approach allows us to explicitly model ownership transfer of array elements in the \is{nres} monad, and let Sepref
-  generate efficient code, sparing explicit copies of elements.
-
-  The necessary conversion operations can usually be inserted by a straightforward refinement step.
-  For example, we refine the \is{insert} function (Sec.~\ref{sec:unguarded_insort}) as follows:
-  \begin{lstlisting}
-    is_insert_o G xs' i ==
-      xs' <- to_olist xs';
-      (x,xs') <- borrow xs' i;
-      (xs',i) <- while
-        (%(xs',i). (\<not>G \or i>0) \and xs' < xs'[i-1]) (%(xs',i).
-          (t,xs') <- borrow xs' (i-1);
-          xs' <- ins xs' i t;
-          return (xs', i-1)
-        ) (xs',i);
-
-      xs' <- ins xs' i x;
-      return (to_slist xs')
-    }
-  \end{lstlisting}
-  First, this function converts the input list \is{xs'} to one with explicit ownership.
-  It then borrows the element \is{i} from this list. In each loop iteration, it moves an element
-  from \is{i-1} to \is{i}, by borrowing and immediately re-inserting it.
-  Finally, the element borrowed at the beginning is re-inserted, and the list is converted back to implicit ownership.
-  The refinement proof is, again, straightforward. The refinement relation for the loop states that the abstract and concrete array
-  contain the same values, except at index \is{i}, where the concrete value is not present:
-  \begin{lstlisting}
-    |xs'| = |xs|
-    \and (\<forall>j<|xs|. j~=i --> xs'[j] = Some (xs[j])) \and xs'[i]=None}
-  \end{lstlisting}
-
-  Using this explicit ownership extension to Sepref, we can also synthesize the Introsort algorithm for arrays of arrays of characters,
-  where the inner arrays are compared lexicographically.
-
-%   xxx, ctd here: Define slice_partition_spec the same way! Then show refinement of algo!
-%     lemma "partition3_spec xs l h = ⇓(slice_rel xs l h ×⇩r Id) (doN { ASSERT (h-l≥4 ∧ h≤length xs); (xs⇩1,xs⇩2) ← partition1_spec (slice l h xs); RETURN (xs⇩1@xs⇩2, l+length xs⇩1) })"
-%
-%
-%
-%   xxx, ctd here: slicep_rel and slice_rel. Do we need two steps?
-%
-%
-
-\section{Benchmarks}
-  The goal of this paper was to verify
-  an existing real-world implementation of a medium-complex algorithm.
-  We chose the Introsort implementation from the GNU C++ library. As we did not directly verify the existing C++ code,
-  but re-implemented the algorithm using Isabelle-LLVM, we have to test how close we came to the performance of the original.
-
-  Our benchmark set is based on the Boost C++ Library's~\cite{boost} sorting algorithm benchmarks, extended with further benchmarks indicated in~\cite{Bent93}.
-  Apart from sorting a random list of elements that are mostly different (random), we also sort lists of length $n$ that contain only $n/10$ different elements (random-dup-10), and random lists of only two different elements (random-boolean), as well as lists where all elements are equal (equal).
-  We also consider already sorted sequences (sorted, rev-sorted), as well as a sequence of $n/2$ elements in ascending order, followed by the same elements in descending order (organ-pipe).
-We also consider sorted sequences where we applied $pn/100$ random swap operations (almost-sorted-$p$).
-  Finally, we consider sorted sequences with $pn/100$ random elements inserted at the end or in the middle ([rev-]sorted-end-$p$, [rev-]sorted-middle-$p$).
-
-  We sorted integer arrays with $n=10^8$ elements, and string arrays with $n=10^7$ elements.
-  To represent strings, we used dynamic arrays of characters, and used the same compare function for our algorithm and std::sort,
-  in order to eliminate differences in the compare function implementation\footnote{Initially, we used std::string elements for std::sort,
-  for which the compare function seems to be significantly slower than our compare function on dynamic arrays.}.
-
-  We compile both, our verified algorithm and the std::sort benchmark with clang-6.0.0.
-  The benchmarks were run on a laptop with an Intel(R) Core(TM) i7-8665U CPU and 32GiB of RAM, as well as on a server machine with
-  24 AMD Opteron 6176 cores and 128GiB of RAM. Ideally, the sorting algorithm should take exactly the same time when run on the same data and machine.
-  However, in practice, we encountered some noise. We encountered both, \emph{high frequency} noise, that slowed down single sorting
-  runs, and \emph{low frequency} noise, that slowed down the machine for several minutes. To counter the high-frequency noise,
-  we ran each experiment 10 times in a row, dropped the 6 extremal results, and report the average of the remaining 4 results.
-  If the smallest remaining result was more than 2\% smaller than the largest remaining result, we repeated the experiment.
-  Countering the low-frequency noise is more time consuming. For all benchmarks, we repeated the slower experiment when the
-  difference between our algorithm and std::sort was more than 15\%\footnote{The outlier for sorted-end-1 on the server machine seems to be genuine.
-We have no explanation for that.}.
-Moreover, for the random and organ-pipe integer benchmarks,
-  we ran all the experiments in a loop for several hours, and report the intervals of the averages of the median 4 values.
-  The results are displayed in Tables~\ref{tab:benchmarks} and ~\ref{tab:bench_longrun}.
-
-  \begin{table}{\footnotesize
-    \begin{tabular}{|l|S|S||S|S|}
-      \hline
-      \multirow{2}{*}{Data Set}&\multicolumn{2}{c||}{uint64}&\multicolumn{2}{c|}{llstring}\\
-      & Lap & Srv & Lap & Srv \\
-      \hline
-      random & -0.5 & 2.4&-14.5 & -2.1\\
-      random-dup-10 & -0.2 & 3.2&-1.2 & -1.7\\
-      random-boolean & -0.1 & -0.7&4.7 & 5.3\\
-      equal & -13.3 & 0.9&6.9 & 4.0\\
-      sorted & 6.6 & 3.0&2.2 & 4.2\\
-      rev-sorted & 1.9 & 3.8&2.1 & 2.0\\
-      organ-pipe & -8.1 & 12.8&7.0 & 4.9\\
-
-      almost-sorted-10 & 3.8 & 1.3&-0.8 & -2.0\\
-      almost-sorted-.1 & 3.2 & 1.7&2.3 & -0.7\\
-      almost-sorted-1 & -8.7 & 2.2&-11.1 & -1.1\\
-      almost-sorted-50 & -0.0 & 2.5&-12.2 & -1.8\\
-
-      sorted-end-10 & -4.3 & 1.3&-1.0 & -1.4\\
-      sorted-end-.1 & -4.0 & 8.3&-4.7 & 11.9\\
-      sorted-end-1 & -5.2 & 19.9&10.4 & 5.6\\
-      sorted-middle-10 & -1.1 & 3.2&-0.7 & -2.7\\
-      sorted-middle-1 & 12.5 & 4.5&5.4 & 2.1\\
-      sorted-middle-.1 & -2.8 & 4.4&6.1 & 6.6\\
-
-      rev-sorted-end-10 & 1.1 & 3.1&-1.5 & -2.4\\
-      rev-sorted-end-.1 & 5.6 & 8.7&0.9 & 5.4\\
-      rev-sorted-end-1 & 9.7 & 3.9&-12.6 & 0.6\\
-      rev-sorted-middle-10 & 1.2 & 6.2&-14.3 & -0.6\\
-      rev-sorted-middle-1 & -2.3 & 3.6&-12.0 & 0.9\\
-      rev-sorted-middle-.1 & -5.7 & 4.2&3.4 & 5.5\\
-      \hline
-    \end{tabular}}
-    \caption{Benchmark results, as the ratio $(i/c - 1)*100$, where $i$ and $c$ are the average wall-times of verified Introsort and std::sort.
-    That is, negative values indicate that Introsort was faster, and positive values indicate that std::sort was faster.
-    }\label{tab:benchmarks}
-  \end{table}
-  \begin{table}\footnotesize
-    \begin{tabular}{|l|SSS|SSS|}
-      \hline
-      \multirow{2}{*}{Algorithm} & \multicolumn{3}{c|}{random} & \multicolumn{3}{c|}{organ-pipe} \\
-      & {min} & {max} & {r} & {min} & {max} & {r} \\
-      \hline
-      isabelle::sort &7.35 & 8.58 & 16.7 & 8.43 & 9.86 & 17.0 \\
-      std::sort &7.38 & 8.63 & 17.0 & 8.97 & 10.49 & 16.9 \\
-      \hline
-    \end{tabular}
-
-    \caption{Intervals of the results (wall time in seconds) for repeating each experiment 74 times,
-      in an interleaved fashion. The $r$ fields display $(\max/\min-1)*100$, i.e., the value that
-      would appear in Table~\ref{tab:benchmarks} when comparing the slowest and fastest run of the same experiment.
-      }\label{tab:bench_longrun}
-  \end{table}
-
-  The results clearly show that our verified sorting algorithm performs on-par with the GNU C++ Library's implementation,
-  within the expected noise of up to $17\%$. Table~\ref{tab:bench_longrun} even indicates that our algorithm performs
-  slightly better on our laptop machine. We leave it to future work to do more precise measurements on different machines.
-
-%   \subsection{Proof Architecture of Introsort}
-%     Start with introsort-scheme. Motivate stepwise refinement "in the small" to get
-%     from list-concatenation to slicing.
-%
-%     Then illustrate big-picture architecture. Mention that we use similar "in the small" techniques for other modules.
-%
-%     Finally show how final algorithm is assembled (nres-level)
-%
-%     THEN introduce sepref tool, and illustrate how final algorithm is refined, module by module.
-%     mention parameterization over sort function.
-%
-%     Can we motivate the array-ownership stuff?
-%
-%
-%
-%
-% \section{Isabelle-LLVM}
-%   Quick intro to Isabelle-LLVM
-%
-%   \subsection{Nested Containers and Ownership}
-%     Our prototype solution
-%     Rust's ownership model
-%     Outlook: Fractional separation logic for read-only borrowing
-%     Still unsolved: Abstractly immutable, but concrete mutations
-%       (E.g. union-find with path compression, amortized search-trees?) Invalid in concurrent setting, but perfectly OK in sequential setting!
-%
-%
-%
-%   \subsection{Isabelle-LLVM Code Generator}
-%   Small TCB.
-%
-%
-% \section{Verification of Introsort}
-%
-%   Layout general architecture (Quicksort-Scheme)
-%   Main proof steps and concepts:
-%     Partially sorted
-%     Guards
-%     Plugging together the implementation
-%
-%
-%   Observations
-%     sledgehammer tool is really useful. But prone to not properly abstract/modularize the proofs,
-%       which will only work until sledgehammer can't solve a goal. At this point, usually
-%       refactoring of the whole proof is required to make it manageable again.
-%
-%
-%
-%
-% \section{Benchmarks}
-%   64 bit integers,
-%   character arrays (we don't have a std::string implementation in Isabelle-LLVM yet)
-%
-%   benchmark suite taken from BOOST. Added some more test cases from Timsort(?) paper.
-%
-%
-%   Benchmark the heapsort-algorithm alone!
-%
-%
-%   Add Test:
-%     Organpipe  123..nn.. 321
-%     Random 0-1 arrays
+%   \begin{figure}
+%   \plotpdqsort
+%   \end{figure}
+%
+%   \begin{figure}
+%   \plotintrosort
+%   \end{figure}
+  They indicate that both our Pdqsort and Introsort implementations are competitive.
+  There is one outlier for Pdqsort for already sorted integer arrays on the laptop. We have not yet understood its exact reason.
+  The remaining cases differ by less than 20\%, and in many cases our verified algorithm is actually faster.
+
+%   Except for already sorted integer sequences on the laptop machine, the algorithm
+%   For Pdqsort, we have
+%   some outliers $>20\%$ for equal, rev-sorted, and sorted. We have not yet understood their exact reason, but they may hint at
+%   a missed (or inefficiently implemented) optimization.
+%   The remaining cases differ by less than 20\%, and in many cases our verified algorithm is actually faster.
 
 \section{Conclusions}
-In this paper we have verified the state-of-the art Introsort sorting algorithm using Isabelle-LLVM.
-We included most optimizations found in the GNU C++ Library's (libstdc++) implementation.
-On an extensive set of benchmarks, our verified implementation performs on par with the libstdc++ implementation.
-
-The verification took us about 100 person hours distributed over 4 weeks. After two weeks and roughly 60 person hours,
-we had arrived at an algorithm that was only about 10\% slower than std::sort. However, it only worked for integers, and always
-used guarded insertion sort. The remaining time was spent on implementing explicit ownership for Sepref, such
-that we could sort strings and other objects, and on verifying the unguarded insertion sort optimization.
-
-% Verification Effort: ca 100 person-hours, including benchmarking and extension of Isabelle-LLVM to support nested containers.
-% ca 60 person-hours to arrive at first integer sorting algorithm that was competitive [this time, however, includes heapsort implementation, which is not strictly necessary for competitive algorithm, but was done quite early in the project].
-% XXX lines of proof.
-
-This shows that Isabelle-LLVM can be used to generate competitive verified implementations of medium-complex algorithms.
-The Refinement Framework has been successfully used for much larger projects, e.g., model-checkers~\cite{ELNN13,BrLa18,WiLa18}.
-However, these developments produce implementations that use arbitrary precision integers and target Standard ML,
-which means a principal performance deficit compared to more low-level implementations. Using Isabelle-LLVM as backend to the Refinement Framework,
-these limitations can now be overcome. Thanks to the modular structure of the Refinement Framework, we can even hope to gradually
-transform existing verification projects to Isabelle-LLVM, e.g., replacing invocations of functional mergesort on lists by
-our Introsort algorithm on arrays in a first step.
-
-While we can easily instantiate our sorting algorithm to arbitrary weak orderings on arbitrary element types \emph{within} the Refinement Framework,
-we cannot so easily replace C++'s std::sort. The reason is that std::sort is a template which is instantiated by the C++ compiler,
-i.e., before LLVM intermediate code is even generated. Verifying the std::sort template itself would require a formal semantics of C++,
-including its template mechanism and the relevant traits from the standard template library (orderings, iterators, etc.).
-To the best of our knowledge, such a semantics has not yet been formalized, let alone being used to verify non-trivial algorithms.
-However, we can \emph{specialize} the std::sort template to use our Introsort algorithm for any fixed combination of element type and ordering, on containers backed by an array (e.g.\ std::vector).
-
-Finally, the trusted code base of Isabelle-LLVM is relatively small, even compared to other common program verification methods.
-Following the LCF tradition~\cite{gordon-et-al-1979}, the Isabelle prover is designed around a small inference kernel, which is the only code
-whose bugs can affect the correctness of proved theorems.
-The Isabelle-LLVM framework contains a relatively straightforward formalization of the LLVM semantics, with a straightforward code
-generator~\cite{La19-llvm} to LLVM intermediate language.
-Finally, the LLVM compiler is an industrial strength, well-tested, and widely used compiler.
-In contrast, the default code generators of most theorem provers target high level languages, which includes the
-compilers and the high-level language semantics into the trusted code base.
-Deductive verification tools like KeY~\cite{BHS07} depend on the correct axiomatization of the highly complex Java semantics,
+We have presented the first verification of the Introsort and Pdqsort algorithms.
+We verified state-of-the-art implementations, down to LLVM intermediate representation.
+On an extensive set of benchmarks, our verified implementations perform on par with their
+unverified counterparts from the GNU C++ and Boost C++ libraries.
+Apart from our work, the only other verified real-world implementations of sorting algorithms
+are Java implementations of Timsort and dual-pivot quicksort that have been verified with KeY~\cite{GRBBH15,BSSU17}.
+
+% To the best of our knowledge, our work and two sorting algorithms from the Java
+% standard library that have been verified with KeY~\cite{GRBBH15,BSSU17} are the only
+% verified real-world sorting algorithms.
+
+Compared to other program verification methods, the trusted code base of our approach is small:
+apart from the well-tested and widely used LLVM compiler, it only includes Isabelle's logical inference kernel, and the
+relatively straightforward Isabelle-LLVM semantics and code generation~\cite{La19-llvm}.
+In contrast, deductive verification tools like KeY~\cite{BHS07} depend on the correct axiomatization of the highly complex Java semantics,
 as well as on several automatic theorem provers, which, themselves, are highly complex and optimized C programs.
 
-To reduce the trusted code base even further, one can include a compiler correctness proof in the theorem prover:
-The Verifiable C project~\cite{App14} provides a verification infrastructure on top of the verified CompCert compiler~\cite{BL09}.
-However, to the best of our knowledge, this has only been applied to verify rather simple programs~\cite{CBGD18}.
-The CakeML project \cite{KTMK19,MyOw14} has developed a code generator for HOL4 and a verified ML compiler.
-However, the performance of functional high-level languages seems to be inherently limited wrt.\ low-level imperative languages,
-and we are not aware of any high-performance algorithm that has been verified using CakeML.
-
-\subsection{More Related Work}
-Sorting algorithms are a standard benchmark for program verification tools, such that we cannot give an exhaustive overview.
-Nevertheless, we discuss a few notable examples: The arguably first formal proof of quicksort was given
+Our verified algorithms can readily be used in
+larger verification projects, and we have already replaced a naive quicksort implementation that caused a stack overflow in
+an ongoing SAT-solver verification project~\cite{FBL18}.
+For fixed element types and containers based on arrays (e.g.\ std::vector), we can use our verified algorithms as
+a drop-in replacement for C++'s std::sort. A direct verification of the C++ code of std::sort, however,
+would require a formal semantics of C++, including templates and the relevant concepts from the standard template library (orderings, iterators, etc.).
+To the best of our knowledge, such a semantics has not yet been formalized, let alone been used to verify non-trivial algorithms.
+
+The verification of Introsort took us about 100 person hours.
+After we had set up most of the infrastructure for Introsort, we could verify the more
+complex Pdqsort in about 30h.
+
+\subsection{Related Work}
+Sorting algorithms are a standard benchmark for program verification tools, such that we cannot give an exhaustive overview here.
+Nevertheless, we discuss a few notable examples: the arguably first formal proof of quicksort was given
 by Foley and Hoare himself~\cite{FoHoa71}, though, due to the lack of powerful enough theorem provers at these times,
@@
 One of the first mechanical verifications of imperative sorting algorithms is by
-Filli{\^a}tre and Magaut~\cite{FiMa99}, who prove correct imperative versions of quicksort, heapsort, and insertion sort in Coq.
-However, they do not report on code generation or benchmarking, nor do they combine their separate algorithms to Introsort.
+Filli{\^a}tre and Magaud~\cite{FiMa99}, who prove correct imperative versions of quicksort, heapsort, and insertion sort in Coq.
+However, they use a simplistic partitioning scheme, do not report on code generation or benchmarking,
+nor do they combine their separate algorithms to get Introsort.
 
@@
 \subsection{Future Work}
-Actually, the C++ standard requires the std::sort algorithm to not only be functional correct,
-but to also have a worst-case complexity of $O(n\log n)$. Recently, the Refinement Framework has been extended to reason
-about time complexity~\cite{HaLa19}, and it has been proposed that Isabelle-LLVM could be used as a backend.
-This would enable us to also formally prove the worst-case complexity of Introsort.
-
-Another obvious direction is to formalize other state-of-the-art implementations of sorting algorithms,
-for example std::stable\_sort, or Boost's pattern defeating quicksort~\cite{boost} with branch prediction optimization~\cite{EdWe16}.
-
-Finally, we proposed an explicit ownership model for nested lists.
-However, we only support write-ownership. A next step would be to include read-ownership, which allows aliasing on read-only objects.
-Formally, this could be realized with fractional permissions~\cite{BCOP05}.
+An obvious next step is to verify a branch-aware partitioning algorithm~\cite{EdWe16}.
+Thanks to our modular approach, this will easily integrate with our existing formalization.
+We also plan to extend our work to stable sorting algorithms.
+Recently, we have extended the Refinement Framework to support reasoning about algorithmic complexity~\cite{HaLa19}.
+Once this work has been integrated with Isabelle-LLVM, we can also prove that our implementations
+have a worst-case complexity of $O(n\log(n))$, as required by the C++ standard.
+Finally, we proposed an explicit ownership model for nested lists. We plan to extend this
+to more advanced concepts like read-only shared ownership, inspired by Rust's~\cite{rustlang} ownership system.
+Formally, this could be realized with fractional permission separation logic~\cite{BCOP05}.
 
@@
 
-\paragraph{Acknowledgements}Hidden for blind review.
-%
-% We received funding from DFG grant LA 3292/1 "Verifizierte Model Checker" and VeTSS grant "Formal Verification of Information Flow Security for Relational Databases".
-
-%
-%
-% The Compcert project~\cite{compcert} has formalized a correct compiler for C to machine code
-%
-% This has been done in Coq for a fragment of the C language
-%
-%
-%
-% Most other algorithms that are verified with interactive theorem provers such as Isabelle, Coq~\cite{XX}, or ACL2 \cite{XX}
-%
-%
-% xxx: CompCert, CakeML --> actually smaller code-base
-%
-%
-%
-%
-% Apart from the obvious things such as the machine and operating system,
-% and the Theorem Provers kernel
-%
-%
-%
-%
-%
-%
-%
-%
-%
-%
-%   As our verification targets LLVM, we cannot directly replace std::sort, which is a C++ template.
-%   However, we could provide template specializations for std::sort that direct to our verified
-%   sorting algorithm for certain data types. A direct verification of std::sort would require a formal verification framework
-%   for C++ including templates, which, to the best of our knowledge, does not exist.
-%
-%   Inside the Isabelle-LLVM framework, however, our algorithm can easily be instantiated to any data structure that implements
-%   random access, and to any element type that implements a weak ordering.
-%
-%
-%
-%
-%
-%
-%
-% makes the achievable performance significantly worse than what can be done
-%
-%
-% only went down to either functional code, or the imperative fragment of Standard ML,
-%
-%
-%
-%
-% --> mention goal, achieved!
-%
-% --> mention that Isabelle provides means to use this as building block for
-%   even more complex algorithms, mention CAVA, Timed-Automata MC
-%
-%
-% Our verified implementation contains most optimizations that we
-%
-%
-%
-% Verified state-of-the art sorting algorithm implementation. On par with original from libstdc++.
-%
-% top-down vs. bottom up. Verification of C++ would be different game.
-% TCB! ours is smaller, as it excludes C++ compiler.
-% Drawback: Not as universally usable from C++, b\c not integrated into STL.
-%
-%
-%
-%
-% Verification Effort: ca 100 person-hours, including benchmarking and extension of Isabelle-LLVM to support nested containers.
-% ca 60 person-hours to arrive at first integer sorting algorithm that was competitive [this time, however, includes heapsort implementation, which is not strictly necessary for competitive algorithm, but was done quite early in the project].
-% XXX lines of proof.
-%
-% Extended Isabelle-LLVM:
-%   Explicit ownership on nres-level
-%   Header file generation
-%
-%
-% Usability:
-%   As our verification targets LLVM, we cannot directly replace std::sort, which is a C++ template.
-%   However, we could provide template specializations for std::sort that direct to our verified
-%   sorting algorithm for certain data types. A direct verification of std::sort would require a formal verification framework
-%   for C++ including templates, which, to the best of our knowledge, does not exist.
-%
-%   Inside the Isabelle-LLVM framework, however, our algorithm can easily be instantiated to any data structure that implements
-%   random access, and to any element type that implements a weak ordering.
-%
-%
-%
-% Related Work:
-%   Filliatre COQ 2012: Insertion sort, quicksort, heapsort
-%     no combination to introsort
-%     not so many optimizations
-%     imperative. Executable? (They do not report on benchmarks)
-%
-%   Timsort verification in KeY
-%     verifies the Java implementation (after slight modifications, mainly stripping generics)
-%       This is a bottom-up approach, while ours is a top-down approach.
-%       The trusted code-base of KeY is much larger, in particular, it includes SMT solvers.
-%
-%
-%
-%     There's also a follow-up on dual pivot quicksort??
-%
-%       Proving JDK’s Dual Pivot Quicksort Correct
-%       Bernhard Beckert, Jonas Schiffl, Peter H. Schmitt, and Mattias Ulbrich
-%
-%         Bottom up. Refactored the code to make it more prover-friendly.
-%
-%
-%
-%
-%     12. de Gouw S, de Boer FS, Rot J. Proof pearl: the KeY to correct and stable sorting. J. Autom. Reason. 2014;53(2):129–139. doi: 10.1007/s10817-013-9300-y. [CrossRef] [Google Scholar]
-%       radix sort, counting sort
-%
-%
-%   Why3? Verified some simple sorting algorithm. No evidence of implementations or benchmarks.
-%
-%   Functional quicksort: Common knowledge. Few lines of Isabelle!
-%
-%   Heapsort-AFP
-%     not executable
-%
-%   Bockeneck et al.
-%     Insertion sort, quicksort partitioning (Lomutu?)
-%     Not executable
-%
-%
-%
-%
-%
-% Future work:
-%   Proof complexity (Use refinement-with-time)
-%   More optimizations (eg PD-Qsort)
-%   Specialized algorithms for integers and strings (radix sort, burstsort, etc ... -> Boost library)
-
-\bibliographystyle{ACM-Reference-Format}
+\paragraph{Acknowledgements}
+We received funding from DFG grant LA 3292/1 "Verifizierte Model Checker" and VeTSS grant "Formal Verification of Information Flow Security for Relational Databases".
+
+% Further structure
+%
+%   *** Verification.
+%   Introduce modularization, specs first.
+%
+%   show specs for slice_sort.
+%
+%   complete dep diagram for Introsort and Pdqsort
+%
+%   Pdqsort: phases of algorithm (pivot-front, partitioned, sorted-left, sorted-right)
+%
+%   Unguarded optimization: Handled by parameter
+%
+%   Deferred swap in heapsort: By refinement
+%
+%   *** Imperative Implementation
+%   quick Sepref intro
+%   easy case: pure array elements
+%     specialization of inserts
+%     give example for ug_insert
+%
+%   hard case: array elements on heaps
+%     ownership, etc.
+%     explicit ownership
+%     solved by data refinement step
+%
+%   The LLVM code generator
+%     mention header-generation extension
+%     show example
+%
+%   *** Benchmarks
+%     Try scatter plot
+%
+%   *** Conclusions
+%
+%
+%
+
+\bibliographystyle{abbrv}
 \bibliography{bib}

### 2020-05-04 papers/sorting/IJCAR_reviews_for_rebuttal.txt
IJCAR'20 paper

Dear Peter Lammich,

Thank you for your submission to IJCAR 2020. The IJCAR 2020
rebuttal period will be between March 9 and March 12 (AoE).

During this time, you will have access to the current state of
your reviews and have the opportunity to submit a response.
Please keep in mind the following during this
process:

* The response must focus on any factual errors in the reviews
and any questions posed by the reviewers. It must not provide
new research results or reformulate the presentation. Try to be
as concise and to the point as possible.

* The rebuttal period is an opportunity to react to the
reviews, but not a requirement to do so. Thus, if you feel the
reviews are accurate and the reviewers have not asked any
questions, then you do not have to respond.

* The reviews are as submitted by the PC members, without any
coordination between them. Thus, there may be inconsistencies.
Furthermore, these are not the final versions of the reviews.
The reviews can later be updated to take into account the
discussions at the program committee meeting, and we may find
it necessary to solicit other outside reviews after the
rebuttal period.

* The program committee will read your responses carefully and
take this information into account during the discussions. On
the other hand, the program committee will not directly respond
to your responses, either before the program committee meeting
or in the final versions of the reviews.

* Your response will be seen by all PC members who have access
to the discussion of your paper, so please try to be polite and
constructive.

The reviews on your paper are attached to this letter. To
submit your response you should log on the EasyChair Web page
for IJCAR 2020 and select your submission on the menu.

----------------------- REVIEW 1 ---------------------
SUBMISSION: 128
TITLE: Efficient Verified Implementation of Introsort and Pdqsort
AUTHORS: Peter Lammich

----------- Overall evaluation -----------
SCORE: 2 (accept)
----- TEXT:
The paper presents an application of the Isabelle-LLVM framework: two state-of-the-art sorting algorithms (i.e., Introsort and Pdqsort) have been formally verified in the Isabelle/HOL theorem prover. The notable feature of this verification effort is that the verified algorithms can be extracted to LLVM representations, and further compiled to machine code that shows competitive performance against highly optimised implementations in the GNU C++ Standard Library. The performance is marvellous since most of the time the extracted code from a theorem prover (which is usually in a functional language) is several orders of magnitude slower than an optimised C/C++ implementations. Therefore, I believe this work is of significant importance.

The paper is well written and easy to follow. Both the proofs and the benchmark can be easily accessed online. Overall, I don't have any complaint against this paper and wish to see more follow-ups of the Isabelle-LLVM framework.

----------------------- REVIEW 2 ---------------------
SUBMISSION: 128
TITLE: Efficient Verified Implementation of Introsort and Pdqsort
AUTHORS: Peter Lammich

----------- Overall evaluation -----------
SCORE: 2 (accept)
----- TEXT:
The paper presents a verification effort for some sorting algorithms that are used in practice in C++ libraries.  The algorithms are non-trivial, and algorithms of the same kind have had many bugs in the past.  The paper can also be seen as a case study for the Isabelle Refinement Framework (by the same author).  After verification in the Isabelle Framework, the algorithms are translated to LLVM using the Isabelle-LLVM framework (also by the same author).  The generated algorithms (which are verified down to the LLVM level) behave on par with the implementations provided in libraries.

The paper is very clearly written.  It is a case study, but the author has chosen the right level of abstraction to explain the verification process and the tools used.  The results are nice, exhaustive, and convincingly presented.  My sole concern about the paper is that it is basically a case study for tools previously published.  This is however afaik the first proper case study, so I believe the paper should be accepted to IJCAR.

Minor comments:

p2.  Isn't the pivot generally left out of the recursive call?  In line 11, Algo 1, it seems it is also sorted in the second recursive call.

p2. Notation l..<h is maybe unconventional, maybe explain in a few words?

p3 : ...which leaves the pivot-element as first element of the left partition.  Could you recheck?

p3, Algo 2.  Maybe the arguments of PDQSORT on line 1 are just a cut and paste of line 3: lm and d are unused in this function.

p4. the the threshold

p8. It seems notation for assert and spec changed.  There might be a good reason, but if not, fix.

p9. The the Sepref

p14. The explanation of Fig 2. could probably made simpler to understand

p13. person hours are interesting, but lines of code would also be of interest

----------------------- REVIEW 3 ---------------------
SUBMISSION: 128
TITLE: Efficient Verified Implementation of Introsort and Pdqsort
AUTHORS: Peter Lammich

----------- Overall evaluation -----------
SCORE: 1 (weak accept)
----- TEXT:
The verification of sorting algorithms that are close to actual
implementation is fiendishly difficult. The paper manages to verify
in Isabelle/HOL the Introsort and Pdqsort algorithms, which are both
part of C++ libraries. The idea is to implement abstract functions
in Isabelle/HOL and then refine them to LLVM code. This produces
code that is correct but can also be run against benchmarks (and
it turns out the code is rather competitive). The verification is
aided by Isabelle-LLVM framework. This work is a good datapoint
for the formal verification of implemented algorithms (as opposed to
idealised algorithms shown in textbooks). I therefore recommend the
acceptance of the paper.

A slight let-down of the paper is that it assumes quite a bit of background
knowledge of the Isabelle-LLVM framework. References are given, which
is appreciated. A longer version of the paper might make the material
more "stand-alone".

------------------------------------------------------

Best wishes,

Nicolas Peltier
Viorica Sofronie-Stokkermans

### 2020-05-04 papers/sorting/main.tex
IJCAR'20 paper

@@
 
-  As generic sorting algorithm, the GNU C++ Standard Library implements the Introsort algorithm,
-  a combination of quicksort, heapsort, and insertion sort. The Boost C++ Libraries implement Pdqsort,
-  an extension of Introsort that achieves linear runtime on inputs with certain patterns.
-
-  We verify Introsort and Pdqsort in the Isabelle LLVM verification framework,
+  As generic sorting algorithm, the GNU C++ Standard Library implements the introsort algorithm,
+  a combination of quicksort, heapsort, and insertion sort. The Boost C++ Libraries implement pdqsort,
+  an extension of introsort that achieves linear runtime on inputs with certain patterns.
+
+  We verify introsort and pdqsort in the Isabelle LLVM verification framework,
   closely following the state-of-the-art implementations from GNU and Boost.
@@
 %   In this paper, we present verified and efficient implementations of
-%   Musser's introspective sorting algorithm (Introsort)~\cite{Muss97} and its
-%   extension \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}.
+%   Musser's introspective sorting algorithm (introsort)~\cite{Muss97} and its
+%   extension \emph{pattern-defeating quicksort} (pdqsort)~\cite{pdqsort}.
 %   Using the Isabelle Refinement Framework with its LLVM backend~\cite{La19-llvm},
-%   we closely follow the implementations of Introsort and Pdqsort from
+%   we closely follow the implementations of introsort and pdqsort from
 %   the GNU C++ Library (libstdc++)~\cite{libstdc++} and the Boost C++ Libraries~\cite{boost}.
@@
   Sorting algorithms are an important part of any standard library.
-  The GNU C++ Library (libstdc++)~\cite{libstdc++} implements Musser's introspective sorting algorithm (Introsort)~\cite{Muss97}.
+  The GNU C++ Library (libstdc++)~\cite{libstdc++} implements Musser's introspective sorting algorithm (introsort)~\cite{Muss97}.
   It is a combination of quicksort, heapsort, and insertion sort,
@@
   The Boost C++ Libraries~\cite{boost} provide a state-of-the-art implementation of
-  \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}, an extension of Introsort to
+  \emph{pattern-defeating quicksort} (pdqsort)~\cite{pdqsort}, an extension of introsort to
   achieve better performance on inputs that contain certain patterns like already sorted sequences.
@@
   Sorting algorithms in standard libraries have not always been correct.
-  The Timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
+  The timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
   Also, many real-world mergesort implementations suffered from an overflow bug~\cite{bs_flaw_blogpost}.
   Finally, LLVM's libc++~\cite{libc++} implements a different quicksort based sorting algorithm. While it may be functionally correct,
-  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by January 2020.}.
-
-  In this paper, we present efficient implementations of Introsort and Pdqsort that are verified down to their
+  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by April 2020.}.
+
+  In this paper, we present efficient implementations of introsort and pdqsort that are verified down to their
   LLVM intermediate representation~\cite{LLVM-manual}. The verification uses the
@@
 \section{The Introsort and Pdqsort Algorithms}
-  The Introsort algorithm by Musser~\cite{Muss97} is a generic unstable sorting algorithm
+  The introsort algorithm by Musser~\cite{Muss97} is a generic unstable sorting algorithm
   that combines the good average-case runtime of quicksort~\cite{Hoare61} with the optimal $O(n\log(n))$ worst-case complexity of heapsort~\cite{Will64}.
@@
 
-  Algorithm~\ref{alg:introsort} shows our implementation of Introsort, which closely follows the implementation in libstdc++~\cite{libstdc++}.
-  The function \textsc{introsort} sorts the slice \is$l..<h$ of the list\footnote{%
+  Algorithm~\ref{alg:introsort} shows our implementation of introsort, which closely follows the implementation in libstdc++~\cite{libstdc++}.
+  The function \textsc{introsort} sorts the slice from index \is$l$ (inclusive) up to index \is{h} (exclusive) of the list\footnote{%
     Our formalization initially uses lists to represent the sequence of elements to be sorted, and refines them to arrays later (cf.\ Sec.\ \ref{sec:imp_impl}).
@@
   threshold positions away from its final position in the sorted list. The remaining sorting is then done by insertion sort (line~\ref{l:intrs:finalis}).
-  The function \textsc{introsort\_aux} implements a recursive quicksort scheme: recursion stops if the slice becomes smaller than the the threshold (line~\ref{l:intrs:threshold}).
+  The function \textsc{introsort\_aux} implements a recursive quicksort scheme: recursion stops if the slice becomes smaller than the threshold (line~\ref{l:intrs:threshold}).
   If the maximum recursion depth is exhausted, heapsort is used to sort the slice (line~\ref{l:intrs:heapsort}).
@@
   recursively invoked for the two partitions (line~\ref{l:intrs:rec1}--\ref{l:intrs:rec2}).
+  Here, \textsc{partition\_pivot} moves the pivot element to the first element of the left partition, and returns the start index of the right partition.
 
@@
   \begin{algorithmic}[1]
-    \Procedure{pdqsort}{$lm,xs,l,h,d$}
+    \Procedure{pdqsort}{$xs,l,h$}
       \If{$h-l>1$}~\Call{pdqsort\_aux}{$\true,xs,l,h,\log(h-l)$}\EndIf \label{l:pdq:aux}
@@
 
-  Algorithm~\ref{alg:pdqsort} shows our implementation of Pdqsort.
-  As for Introsort, the wrapper \textsc{pdqsort} just initializes a depth counter, and then calls the function \textsc{pdqsort\_aux}
+  Algorithm~\ref{alg:pdqsort} shows our implementation of pdqsort.
+  As for introsort, the wrapper \textsc{pdqsort} just initializes a depth counter, and then calls the function \textsc{pdqsort\_aux}
   (line~\ref{l:pdq:aux}),
-  which, in contrast to Introsort, completely sorts the list, such that no final insertion sort is necessary.
+  which, in contrast to introsort, completely sorts the list, such that no final insertion sort is necessary.
   Again, the \textsc{pdqsort\_aux} function implements a recursive quicksort scheme, however, with a few additional optimizations.
@@
   Next, we check for a highly unbalanced partitioning (line~\ref{l:pdq:unbal}), i.e., if one partition is less than 1/8th of the overall size.
-  After encountering a certain number of highly unbalanced partitionings, Pdqsort switches to heapsort (line~\ref{l:pdq:heapsort}).
+  After encountering a certain number of highly unbalanced partitionings, pdqsort switches to heapsort (line~\ref{l:pdq:heapsort}).
   Otherwise, it will shuffle some elements in both partitions, trying to break up patterns in the input (line~\ref{l:pdq:shuffle}).
   If the input was already partitioned wrt.\ the selected pivot (indicated by the flag $ap$),
-  Pdqsort will optimistically try to sort both partitions with insertion sort (line~\ref{l:pdq:maybe}).
+  pdqsort will optimistically try to sort both partitions with insertion sort (line~\ref{l:pdq:maybe}).
   However, these insertion sorts abort if they cannot sort the list with a small number of swaps, limiting the penalty for being too optimistic.
@@
 
-  Our implementation of Pdqsort closely follows the implementation we found in Boost~\cite{boost}.
+  Our implementation of pdqsort closely follows the implementation we found in Boost~\cite{boost}.
   Again, we omitted a manual tail call optimization that LLVM does automatically.
-  Moreover, for certain comparison functions, Boost's Pdqsort uses a special branch-aware partitioning algorithm~\cite{EdWe16}.
+  Moreover, for certain comparison functions, Boost's pdqsort uses a special branch-aware partitioning algorithm~\cite{EdWe16}.
   We leave its verification to future work, but note that it will easily integrate in our existing formalization.
 
-  While Introsort and Pdqsort are based on the same idea, this presentation focuses on the more complex Pdqsort:
+  While introsort and pdqsort are based on the same idea, this presentation focuses on the more complex pdqsort:
   apart from the more involved \textsc{pdqsort\_aux} function,
-  \textsc{pivot\_to\_front} uses Tukey's 'ninther' pivot selection~\cite{Bent93}, while Introsort uses the simpler median-of-three scheme.
+  \textsc{pivot\_to\_front} uses Tukey's 'ninther' pivot selection~\cite{Bent93}, while introsort uses the simpler median-of-three scheme.
   It has two partitioning algorithms used in different situations, and the \textsc{partition\_right} algorithm also
@@
   here \is$|xs|$ is the length of the list \is{xs} and
-  \is{xs[I]} is the slice of the list \is{xs} for indexes in the interval \is{I}.
+  \is{xs[I]} is the slice of the list \is{xs} for indexes in the interval \is{I}\footnote{%
+    An interval from index \is{l} (inclusive) to \is{h} (exclusive) is denoted as \is{l..<h}.
+    If both indexes are exclusive, we write \is{l<..<h}.
+  }.
 %   and \is{xs[l..]} is short for \is{xs[l..<|xs|${}$]}.
@@
 
-  Using the above statement, and an analogous statement for Introsort, we can prove the main correctness theorem:
+  Using the above statement, and an analogous statement for introsort, we can prove the main correctness theorem:
   \begin{theorem}\label{thm:sorting_correct}
@@
 %   the comparison $l<i$ in line~\ref{alg:insert:loophead}. In particular when element comparisons are cheap, i.e.~have cost similar to index comparisons,
-%   omitting a comparison from the very inner loop can have a significant effect. Again, we show the version for Pdqsort here, where the unguarded optimization
-%   is used for non-leftmost partitions. The final insertion sort of Introsort uses a similar optimization, where we switch to unguarded insertion once
+%   omitting a comparison from the very inner loop can have a significant effect. Again, we show the version for pdqsort here, where the unguarded optimization
+%   is used for non-leftmost partitions. The final insertion sort of introsort uses a similar optimization, where we switch to unguarded insertion once
 %   the first threshold elements have been sorted.
@@
   \begin{lstlisting}
-    ASSERT (sorted xs[l..<i] \<and> l<=i \<and> i<|xs| \<and> (G \<or> xs[l-1] <= xs[i]));
-    SPEC (%xs'. xs=$_{l,i+1}$xs' \<and> sorted xs[l..<i+1])
+    assert sorted xs[l..<i] \<and> l<=i \<and> i<|xs| \<and> (G \<or> xs[l-1] <= xs[i]);
+    spec xs'. xs=$_{l,i+1}$xs' \<and> sorted xs[l..<i+1]
   \end{lstlisting}
@@
 \section{Imperative Implementation}\label{sec:imp_impl}
-  We have presented a refinement based approach to verify the Introsort and Pdqsort algorithms,
+  We have presented a refinement based approach to verify the introsort and pdqsort algorithms,
   including most optimizations we found in their libstdc++ and Boost implementations. However, the algorithms are still expressed
@@
   \subsection{The Sepref Tool}
-  The the Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{nres}-monad,
+  The Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{nres}-monad,
   keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
@@
   \begin{lstlisting}
-    move xs i == assert (i<|xs| \and xs[i]~=None); return (the (xs[i]), xs[i:=None])
+    move xs i == assert i<|xs| \and xs[i]~=None; return (the (xs[i]), xs[i:=None])
   \end{lstlisting}
@@
   \begin{lstlisting}
-    cmp_idxs xs i j == assert (i<|xs| \<and> j<|xs|); return xs[i] < xs[j]
+    cmp_idxs xs i j == assert i<|xs| \<and> j<|xs|; return xs[i] < xs[j]
   \end{lstlisting}
@@
   can generate LLVM text.
-  For example, the Pdqsort algorithm for strings yields an LLVM function with the signature:
+  For example, the pdqsort algorithm for strings yields an LLVM function with the signature:
 %   Isabelle-LLVM comes with a code generator that exports this program to actual LLVM text, which is subsequently compiled using the LLVM toolchain.
-%   For example, the Pdqsort algorithms for strings yields an LLVM function with the signature:
+%   For example, the pdqsort algorithms for strings yields an LLVM function with the signature:
   \begin{lstlisting}[language=llvm,literate={}]
@@
 \section{Benchmarks}
-%   We have verified the libstdc++ implementation of Introsort, and the Boost implementation of Pdqsort.
+%   We have verified the libstdc++ implementation of introsort, and the Boost implementation of pdqsort.
 %   We benchmarked our verified implementations against the unverified originals.
@@
 
+  \begin{figure}
+  \plotpdqsort
+  \plotintrosort
+  \caption{Benchmarking our verified implementations against the unverified originals.
+    For each element type, machine, and distribution, the value $(t_1/t_2-1)*s$ is shown,
+    where $t_1$ is the slower time, $t_2$ is the faster time, and $s = 100$ if our implementation is slower,
+    and $s=-100$ if the original implementation is slower. That is, a positive value $p$ indicates that our implementation is slower,
+    requiring $100+p$ percent of the run time of the original. Analogously,
+    a negative value $-p$ means that the original implementation is slower, requiring $100+p$ percent of our implementation's run time.
+  }\label{fig:benchres}
+  \end{figure}
+
   We sorted integer arrays with $n=10^8$ elements, and string arrays with $n=10^7$ elements.
   For strings, all implementations use the same data structure and compare function.
-  For integers, we disable Pdqsorts branch-aware partitioning, which we have not yet verified.
+  For integers, we disable pdqsorts branch-aware partitioning, which we have not yet verified.
   For strings, it does not apply anyway.
@@
 %   The charts below compare the run-times for the verified and unverified algorithm.
-  \begin{figure}
-  \plotpdqsort
-  \plotintrosort
-  \caption{Benchmarking our verified implementations against the unverified originals.
-    For each element type, machine, and distribution, the value $(t_1/t_2-1)*s$ is shown,
-    where $t_1$ is the slower time, $t_2$ is the faster time, and $s = 100$ if the unverified algorithm is faster,
-    and $s=-100$ if the verified algorithm is faster.
-  }\label{fig:benchres}
-  \end{figure}
 
@@
 %   \end{figure}
-  They indicate that both our Pdqsort and Introsort implementations are competitive.
-  There is one outlier for Pdqsort for already sorted integer arrays on the laptop. We have not yet understood its exact reason.
+  They indicate that both our pdqsort and introsort implementations are competitive.
+  There is one outlier for pdqsort for already sorted integer arrays on the laptop. We have not yet understood its exact reason.
   The remaining cases differ by less than 20\%, and in many cases our verified algorithm is actually faster.
@@
 %   Except for already sorted integer sequences on the laptop machine, the algorithm
-%   For Pdqsort, we have
+%   For pdqsort, we have
 %   some outliers $>20\%$ for equal, rev-sorted, and sorted. We have not yet understood their exact reason, but they may hint at
@@
 \section{Conclusions}
-We have presented the first verification of the Introsort and Pdqsort algorithms.
+We have presented the first verification of the introsort and pdqsort algorithms.
 We verified state-of-the-art implementations, down to LLVM intermediate representation.
@@
 Apart from our work, the only other verified real-world implementations of sorting algorithms
-are Java implementations of Timsort and dual-pivot quicksort that have been verified with KeY~\cite{GRBBH15,BSSU17}.
+are Java implementations of timsort and dual-pivot quicksort that have been verified with KeY~\cite{GRBBH15,BSSU17}.
 
@@
 
-The verification of Introsort took us about 100 person hours.
-After we had set up most of the infrastructure for Introsort, we could verify the more
-complex Pdqsort in about 30h.
+The verification of introsort took us about 100 person hours.
+After we had set up most of the infrastructure for introsort, we could verify the more
+complex pdqsort in about 30h. The development consists of roughly 8700 lines of Isabelle text,
+of which 2400 lines are for introsort and 2200 lines for pdqsort. The rest is boilerplate and libraries
+shared between both algorithms, among them 1500 lines for the verification of heapsort.
 
@@
 However, they use a simplistic partitioning scheme, do not report on code generation or benchmarking,
-nor do they combine their separate algorithms to get Introsort.
-
-The Timsort algorithm, which was used in the Java standard library, has been verified
+nor do they combine their separate algorithms to get introsort.
+
+The timsort algorithm, which was used in the Java standard library, has been verified
 with the KeY tool~\cite{GRBBH15}. A bug was found and fixed during the verification.
@@
 %   Proof complexity (Use refinement-with-time)
-%   More optimizations (eg PD-Qsort)
 %   Specialized algorithms for integers and strings (radix sort, burstsort, etc ... -> Boost library)
@@
 %
-%   complete dep diagram for Introsort and Pdqsort
+%   complete dep diagram for introsort and pdqsort
 %

### 2020-05-04 papers/2020_Enschede_Talk/easyenum.rb

@@ -1,0 +1,138 @@
+#!/usr/bin/ruby
+
+#     #title            - frame
+#     #!title           - containsverbatim frame
+#     #[options] title  - explicit options, must not contain "]"
+#     #![options] title
+#
+#     * text      - itemize
+#     *# text     - enumerate (only first item needs #)
+#     *& text     - description
+#
+#     *<2->       - Overlay specification
+#     *#<2->, *&<2->
+#
+#     *name: text - named item
+#     *#name: text, *&name: text
+#     *<2->name:
+#     
+#     %%BEGIN-FILTER: bash-command
+#       text
+#     %%END-FILTER
+#
+#
+
+
+class EItem 
+  attr_accessor :on_depth, :ty
+
+  def initialize d,ty
+    @on_depth=d
+    @ty=ty
+  end
+
+  def open par
+    puts "".ljust(@on_depth - 1) + "\\begin{#{@ty}}#{par}"
+  end
+
+  def close
+    puts "".ljust(@on_depth - 1) + "\\end{#{@ty}}"
+  end
+end
+
+$item_stack = []
+
+def depth text
+  text.gsub(/\t/,"        ");
+  /^( *\*?)/ =~ text
+  return $1.length
+end
+
+def app_list l,t
+  if l=="" or l==nil
+    return t
+  else
+    return l+","+t
+  end
+end
+
+def process_line text
+  if /^[ \t]*%%BEGIN-FILTER:(.*)/ =~ text
+    flt=$1
+    acc=""
+    while gets
+      break if /^[ \t]*%%END-FILTER/=~ $_
+      #acc=acc+"\n" if acc!=""
+      acc=acc+$_;
+    end
+    acc.chop!
+    res=`#{flt} <<HERE
+#{acc}
+HERE`
+    puts res
+  elsif /^[ \t]*%/ =~ text
+    puts text
+  elsif /^[ \t]*$/ =~ text
+    puts text
+  else
+    d = 0
+    d = depth text unless text==nil
+
+    while (!$item_stack.empty? && $item_stack.last.on_depth > d)
+      $item_stack.last.close
+      $item_stack.pop
+    end
+
+    if (text != nil)
+  
+      last_d=0
+      last_d=$item_stack.last.on_depth unless $item_stack.empty?
+  
+      texts=text.strip
+  
+
+      #      1     2 3               45                6
[…]

### 2020-07-04 benchmarks/sorting/eval_benchmark.sh
IJCAR rename

@@ -8,5 +8,5 @@ cat log/allnew-2020-01.log | ./__eval_benchmark.awk > ~/tmp/data.tex
 gawk '$1=="@" { echo=($2=="introsort"); next }  echo {print} ' ~/tmp/data.tex  > ~/tmp/data_introsort.tex
 gawk '$1=="@" { echo=($2=="pdqsort"); next }  echo {print} ' ~/tmp/data.tex > ~/tmp/data_pdqsort.tex
 
-cp ~/tmp/data_introsort.tex ../../papers/sorting/
-cp ~/tmp/data_pdqsort.tex ../../papers/sorting/
+# cp ~/tmp/data_introsort.tex ../../papers/sorting/
+# cp ~/tmp/data_pdqsort.tex ../../papers/sorting/

### 2020-07-04 mkhtml.sh

@@ -33,12 +33,21 @@ if $REBUILD; then
   make
   cd $BASEDIR
 
+  cd papers/IJCAR2020
+  make
+  cd $BASEDIR
+
   cd papers/2019_Rennes_Talk
   make
   cd $BASEDIR
 
+  cd papers/2020_Enschede_Talk
+  make
+  cd $BASEDIR
+
   cd thys
-  isabelle build -v -d '$AFP' -D .
+  isabelle build -v -D .
+#   isabelle build -v -d '$AFP' -D .
   cd $BASEDIR
 
   ./mkdist.sh
@@ -55,9 +64,11 @@ cp index.md html/
 cp dist.tgz html/
 cp LICENSE html/
 cp etc/logo/logo_200.png html/
+cp papers/IJCAR2020/main.pdf html/paper_IJCAR2020.pdf
 cp papers/ITP2019/main.pdf html/paper_ITP2019.pdf
 cp papers/ITP2019/talk/pres.pdf html/slides_ITP2019.pdf
 cp papers/2019_Rennes_Talk/pres.pdf html/rennes2019.pdf
+cp papers/2020_Enschede_Talk/pres.pdf html/enschede2020.pdf
 
 pandoc -V pagetitle="Isabelle LLVM" -s index.md > html/index.html
 

### 2021-03-04 mkhtml.sh
IJCAR20 mods

@@ -64,9 +64,13 @@ cp index.md html/
 cp dist.tgz html/
 cp LICENSE html/
 cp etc/logo/logo_200.png html/
+
 cp papers/IJCAR2020/main.pdf html/paper_IJCAR2020.pdf
+cp papers/IJCAR2020/talk/pres.pdf html/slides_IJCAR2020.pdf
+
 cp papers/ITP2019/main.pdf html/paper_ITP2019.pdf
 cp papers/ITP2019/talk/pres.pdf html/slides_ITP2019.pdf
+
 cp papers/2019_Rennes_Talk/pres.pdf html/rennes2019.pdf
 cp papers/2020_Enschede_Talk/pres.pdf html/enschede2020.pdf
 

### 2021-11-13 benchmarks/sorting/all_benchmarks.sh
Parallel branch

@@ -49,17 +49,53 @@ function run_int_std() {
   run uint64 std::sort $1 $NIELEM $NITER
 }
 
+function run_int_std_par() {
+  run uint64 std::parsort $1 $NIELEM $NITER
+}
+
+
 function run_int_isa() {
   run uint64 isabelle::sort $1 $NIELEM $NITER
 }
+
+function run_int_naive_par() {
+  run uint64 naive::parqsort $1 $NIELEM $NITER
+}
+
+function run_int_isa_par() {
+  run uint64 isabelle::parqsort $1 $NIELEM $NITER
+}
+
+function run_int_sample() {
+  run uint64 boost::sample_sort $1 $NIELEM $NITER
+}
+
+
 
 function run_str_std() {
   run llstring std::sort $1 $NSELEM $NITER
 }
 
+function run_str_std_par() {
+  run llstring std::parsort $1 $NSELEM $NITER
+}
+
 function run_str_isa() {
   run llstring isabelle::sort $1 $NSELEM $NITER
 }
+
+function run_str_naive_par() {
+  run llstring naive::parqsort $1 $NSELEM $NITER
+}
+
+function run_str_isa_par() {
+  run llstring isabelle::parqsort $1 $NSELEM $NITER
+}
+
+function run_str_sample() {
+  run llstring boost::sample_sort $1 $NSELEM $NITER
+}
+
 
 function run_pdq_int_std() {
   run uint64 boost::pdqsort $1 $NIELEM $NITER
@@ -85,18 +121,32 @@ function run_int_gcc() {
 }
 
 for i in $INT_DATA; do
-  run_int_isa $i
+
   run_int_std $i
-  run_pdq_int_isa $i
-  run_pdq_int_std $i
+  run_int_std_par $i
+  run_int_isa_par $i
+  run_int_naive_par $i
+  run_int_sample $i
+
+
+#   run_int_isa $i
+#   run_int_std $i
+#   run_pdq_int_isa $i
+#   run_pdq_int_std $i
 #   run_int_gcc $i
 done
 
 for i in $STR_DATA; do
-  run_str_isa $i
   run_str_std $i
-  run_pdq_str_isa $i
-  run_pdq_str_std $i
+  run_str_std_par $i
+  run_str_isa_par $i
+  run_str_naive_par $i
+  run_str_sample $i
+
+#   run_str_isa $i
+#   run_str_std $i
+#   run_pdq_str_isa $i
+#   run_pdq_str_std $i
 done
 
 

### 2021-11-13 benchmarks/sorting/do_benchmark.cpp

@@ -1,5 +1,7 @@
 // Re-used parts from BOOST sorting benchmark suite
 
+#include <execution>
+
 #include <algorithm>
 #include <iostream>
 #include <iomanip>
@@ -22,6 +24,279 @@ extern "C" {
 using namespace std;
 namespace bsort = boost::sort;
 namespace bsc = boost::sort::common;
+
+
+// Disable branch-prediction optimization for all iterators, to make it comparable with our impl!
+template<class Iter, class Compare>
+inline void boost_pdqsort(Iter first, Iter last, Compare comp) {
+    if (first == last) return;
+    boost::sort::pdqsort_detail::pdqsort_loop<Iter, Compare, false >( first, last, comp, boost::sort::pdqsort_detail::log2(last - first));
+}
+
+template<class Iter>
+inline void boost_pdqsort(Iter first, Iter last) {
+    typedef typename std::iterator_traits<Iter>::value_type T;
+    boost_pdqsort(first, last, std::less<T>());
+}
+
+
+
+
+
+//////////////////////////////////
+// Parallel experiments
+//////////////////////////////////
+
+std::atomic<size_t> bad_partitions (0);
+std::atomic<size_t> threads_spawned (0);
+std::atomic<size_t> depth_limit (0);
+std::atomic<size_t> big_partitions (0);
+std::atomic<size_t> small_partitions (0);
+
+
+template<typename I, typename Compare> I find_pivot(I first, I last, Compare comp) {
+  size_t size = last-first;
+
+  size_t num_samples = std::min ((size_t)std::__lg(size) * 4, (size_t)64);
+
+  size_t samples[num_samples];
+
+  size_t incr = (size-1) / num_samples;
+  size_t extra = (size-1) % num_samples;
+  size_t idx = 0;
+
+  for (size_t i=0;i<num_samples;++i) {
+    samples[i]=idx;
+    idx += incr;
+    if (extra) {++idx; --extra;}
+  }
+
+
+
+
+//   clog<<"Samples out of "<<size<<": ";
+//   for (size_t i=0;i<num_samples;++i) clog<<samples[i]<<" ";
+//   clog<<endl;
+//
+//   clog<<"Values: ";
+//   for (size_t i=0;i<num_samples;++i) clog<< *(first+samples[i]) <<" ";
+//   clog<<endl;
+
+
[…]

### 2021-11-13 benchmarks/sorting/eval_benchmark.sh

@@ -1,12 +1,25 @@
 #!/bin/bash
 
-rm ~/tmp/data*.tex
+set -e
+
+rm -f ~/tmp/data*.tex
 
 # cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
-cat log/allnew-2020-01.log | ./__eval_benchmark.awk > ~/tmp/data.tex
+cat log/par-2021.log | ./__eval_benchmark.awk S uint64 > ~/tmp/data_server_uint64.tex
+cat log/par-2021.log | ./__eval_benchmark.awk S llstring > ~/tmp/data_server_llstring.tex
 
-gawk '$1=="@" { echo=($2=="introsort"); next }  echo {print} ' ~/tmp/data.tex  > ~/tmp/data_introsort.tex
-gawk '$1=="@" { echo=($2=="pdqsort"); next }  echo {print} ' ~/tmp/data.tex > ~/tmp/data_pdqsort.tex
+cat log/par-2021.log | ./__eval_benchmark.awk L uint64 > ~/tmp/data_laptop_uint64.tex
+cat log/par-2021.log | ./__eval_benchmark.awk L llstring > ~/tmp/data_laptop_llstring.tex
 
-# cp ~/tmp/data_introsort.tex ../../papers/sorting/
-# cp ~/tmp/data_pdqsort.tex ../../papers/sorting/
+# cat log/par-2021.log | ./__eval_benchmark.awk
+#
+# exit
+
+# grep ">L" ~/tmp/data.tex | sed -re 's/^>. //' > repeat_laptop.sh
+# grep ">S" ~/tmp/data.tex | sed -re 's/^>. //' > repeat_server.sh
+
+
+# gawk '$1=="@" { echo=($2=="server"); next }  echo {print} ' ~/tmp/data.tex  > ~/tmp/data_server.tex
+# gawk '$1=="@" { echo=($2=="laptop"); next }  echo {print} ' ~/tmp/data.tex > ~/tmp/data_laptop.tex
+#
+cp ~/tmp/data*.tex ../../papers/llvm_par/

### 2021-11-13 benchmarks/sorting/eval_speedup.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+rm -f ~/tmp/sudata*.tex
+
+# cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
+cat log/par-speedup-2021.log | ./__eval_speedup.awk > ~/tmp/sudata.tex
+
+gawk '$1=="@" { echo=($2=="server"); next }  echo {print} ' ~/tmp/sudata.tex  > ~/tmp/sudata_server.tex
+gawk '$1=="@" { echo=($2=="laptop"); next }  echo {print} ' ~/tmp/sudata.tex > ~/tmp/sudata_laptop.tex
+#
+cp ~/tmp/sudata_server.tex ../../papers/llvm_par/
+cp ~/tmp/sudata_laptop.tex ../../papers/llvm_par/

### 2021-11-13 benchmarks/sorting/log/fetch_logs.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+rsync -avz lxcisa0:sorting_benchmarks/log/ server/

### 2021-11-13 benchmarks/sorting/repeat_laptop.sh

@@ -2,42 +2,12 @@
 echo -n '# '; date 
 echo -n '# '; uname -a 
 echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring boost::pdqsort rev-sorted-end-10 10000000 10
-
+./do_benchmark llstring naive::parqsort rev-sorted-end-1 10000000 10
+echo '# Repeated b/c difference > 15\% (1433,1751)'
+./do_benchmark llstring naive::parqsort random-boolean 10000000 10
+./do_benchmark llstring isabelle::parqsort random-boolean 10000000 10
+echo '# Repeated b/c difference > 15\% (771,1026)'
+./do_benchmark llstring naive::parqsort equal 10000000 10
+./do_benchmark llstring isabelle::parqsort equal 10000000 10
 echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring isabelle::pdqsort rev-sorted-end-10 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring boost::pdqsort sorted-end-10 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring boost::pdqsort rev-sorted-middle-.1 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring boost::pdqsort rev-sorted-middle-10 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring boost::pdqsort almost-sorted-.1 10000000 10
-
-echo '# Repeated b/c difference > 15\% (154,201)'
-./do_benchmark uint64 boost::pdqsort sorted 100000000 10
-./do_benchmark uint64 isabelle::pdqsort sorted 100000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring boost::pdqsort rev-sorted-middle-1 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring isabelle::pdqsort rev-sorted-middle-1 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring boost::pdqsort sorted-middle-10 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
-./do_benchmark llstring isabelle::pdqsort sorted-middle-10 10000000 10
-
-echo '# Repeated b/c noise > 2\%'
[…]

### 2021-11-13 benchmarks/sorting/repeat_server.sh

@@ -2,13 +2,47 @@
 echo -n '# '; date 
 echo -n '# '; uname -a 
 echo '# Repeated b/c noise > 2\%'
-./do_benchmark uint64 isabelle::pdqsort sorted-end-.1 100000000 10
-
-echo '# Repeated b/c difference > 15\% (8142,6584)'
-./do_benchmark llstring boost::pdqsort rev-sorted-middle-10 10000000 10
-./do_benchmark llstring isabelle::pdqsort rev-sorted-middle-10 10000000 10
-
-echo '# Repeated b/c difference > 15\% (3899,3371)'
-./do_benchmark uint64 boost::pdqsort almost-sorted-1 100000000 10
-./do_benchmark uint64 isabelle::pdqsort almost-sorted-1 100000000 10
-
+./do_benchmark uint64 naive::parqsort random-boolean 100000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::sample_sort organ-pipe 10000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::sample_sort sorted-end-10 10000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::sample_sort equal 10000000 10
+echo '# Repeated b/c difference > 15\% (561,667)'
+./do_benchmark llstring naive::parqsort equal 10000000 10
+./do_benchmark llstring isabelle::parqsort equal 10000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark uint64 naive::parqsort equal 100000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::sample_sort rev-sorted-middle-.1 10000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark uint64 isabelle::parqsort rev-sorted-middle-.1 100000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark uint64 naive::parqsort rev-sorted 100000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark uint64 isabelle::parqsort rev-sorted 100000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::sample_sort random 10000000 10
+echo '# Repeated b/c noise > 2\%'
+./do_benchmark llstring boost::sample_sort almost-sorted-.1 10000000 10
[…]

### 2021-11-13 benchmarks/sorting/speedup_benchmarks.sh

@@ -1,0 +1,115 @@
+#!/bin/bash
+
+# make do_benchmark
+
+########################## BEGIN CONFIG SECTION ##################
+
+mtype=$1
+if [[ $mtype == "server" ]]; then
+  function tmask() {
+    printf "0x%x" $((2**($1 * 2) - 1 ))
+  }
+
+  TNUM=22
+
+elif [[ $mtype == "laptop" ]]; then
+  function tmask() {
+    printf "0x%x" $((2**$1 - 1 ))
+  }
+
+  TNUM=12
+else
+  echo "Unknown machine config: $mtype"
+  exit 1
+fi
+
+# Defaults
+
+export NITER=10   # Number of iterations
+export NIELEM=100000000 # Number of integer elements
+export NSELEM=10000000 # Number of string elements
+
+# Quick benchmark run
+# export NITER=3   # Number of iterations
+# export NIELEM=1000000 # Number of integer elements
+# export NSELEM=1000000 # Number of string elements
+
+export INT_DATA="random"
+
+export STR_DATA="random"
+
+# CUSTOM Ad-HOC Overrides
+
+# export INT_DATA="random organ-pipe"
+# export STR_DATA=""
+# export NIELEM=1000 # Number of integer elements
+# export NSELEM=1000 # Number of string elements
+
+
+########################## END CONFIG SECTION ##################
+
+
+export LOGFILE="log/sortbench-$(date -Iseconds).log"
+
+echo "Writing log to $LOGFILE"
+
+echo "# Sorting benchmark" | tee "$LOGFILE"
+
+( echo -n "# "; date ) | tee -a "$LOGFILE"
+
+( echo -n "# "; uname -a ) | tee -a "$LOGFILE"
+
+echo "@ $mtype" | tee -a "$LOGFILE"
+
+function run() {
+  TM=$(tmask $1)
+  echo "$1 cores (mask: $TM)" | tee -a "$LOGFILE"
+  ( taskset $TM ./do_benchmark $2 $3 $4 $5 $6 || echo "\n***ERROR $?" ) 2>&1 | tee -a "$LOGFILE"
+  echo | tee -a "$LOGFILE"
+}
+
+for distr in $INT_DATA; do
+  for ((j=1;j<=$TNUM;++j)); do
+    run $j uint64 isabelle::parqsort $distr $NIELEM $NITER
+  done
[…]

### 2021-11-13 mkhtml.sh

@@ -51,6 +51,10 @@ if $REBUILD; then
   make
   cd $BASEDIR
 
+  cd papers/llvm_par
+  make
+  cd $BASEDIR
+
   cd thys
   isabelle build -v -D .
 #   isabelle build -v -d '$AFP' -D .
@@ -70,6 +74,7 @@ cp index.md html/
 cp dist.tgz html/
 cp dist-2020.tgz html/
 cp dist-v1.1.tgz html/
+cp dist-v2.0.tgz html/
 cp LICENSE html/
 cp etc/logo/logo_200.png html/
 
@@ -84,7 +89,10 @@ cp papers/2020_Enschede_Talk/pres.pdf html/enschede2020.pdf
 cp papers/2021_Enschede_Talk/pres.pdf html/enschede2021.pdf
 cp papers/2021_RF_Pres/pres.pdf html/RF_pres.pdf
 
-pandoc -V pagetitle="Isabelle LLVM" -s index.md > html/index.html
+cp papers/llvm_par/main.pdf html/paper_llvm_par.pdf
+
+
+pandoc -V pagetitle="Isabelle LLVM Parallel" -s index.md > html/index.html
 
 
 if $UPLOAD; then

### 2021-11-13 regression/gencode/sorting.h

@@ -18,8 +18,10 @@ void str_free(llstring*);
 void str_init(llstring*);
 void str_append(llstring*, char);
 llstring* str_pdqsort(llstring*, int64_t, int64_t);
+llstring* str_par_sort(llstring*, int64_t);
 uint64_t* pdqsort(uint64_t*, int64_t, int64_t);
 llstring* str_introsort(llstring*, int64_t, int64_t);
+uint64_t* par_sort(uint64_t*, int64_t);
 uint64_t* introsort(uint64_t*, int64_t, int64_t);
 
 #endif

### 2021-11-13 regression/gencode/test_avx512f.h

@@ -1,0 +1,15 @@
+// Generated by Isabelle-LLVM. Do not modify.
+#ifndef _TEST_AVX512F_H
+#define _TEST_AVX512F_H 1
+
+
+
+
+
+
+double test_avx512f_to_ninf(double, double);
+double test_avx512f_to_pinf(double, double);
+double test_avx512f_to_zero(double, double);
+double test_avx512f_to_nearest(double, double);
+
+#endif

### 2021-11-13 regression/gencode/test_double.h

@@ -1,0 +1,12 @@
+// Generated by Isabelle-LLVM. Do not modify.
+#ifndef _TEST_DOUBLE_H
+#define _TEST_DOUBLE_H 1
+
+
+
+
+
+
+double test_double(double, double);
+
+#endif

### 2021-11-13 regression/src/double_lib.c

@@ -1,0 +1,57 @@
+#include <stdlib.h>
+#include <stdint.h>
+#include <assert.h>
+#include <math.h>
+
+_Static_assert(sizeof(double) == sizeof(uint64_t),"sizeof(double) == sizeof(uint64_t)");
+
+double i2d(uint64_t i) {
+  double r;
+  *(uint64_t*)(&r) = i;
+  return r;
+}
+
+uint64_t d2i(double i) {
+  uint64_t r;
+  *(double*)(&r) = i;
+  return r;
+}
+
+double nd(double x) {
+  if (isnan(x)) return 0; else return x;
+}
+
+double rd() {
+  return rand() * 42.0 / ((double)RAND_MAX);
+}
+
+
+
+uint16_t ru16() {
+  return rand() % 0xFFFF; // poor quality random numbers (sigh!)
+}
+
+uint32_t ru32() {
+  return (((uint32_t)ru16())<<16) | ru16();
+}
+
+uint64_t ru64() {
+  return (((uint64_t)ru32())<<32) | ru32();
+}
+
+double rd1() {
+  return i2d(ru64());
+}
+
+double rd2() {
+  double x;
+
+  do {
+    x = rd1();
+  } while (isnan(x));
+
+  return x;
+}
+
+
+

### 2021-11-13 regression/src/kmp.c

@@ -12,8 +12,8 @@ string str(const char *s) {
 
 void test_kmp(string a, string b) {
 
-  int64_t la=a.len;
-  int64_t lb=b.len;
+//   int64_t la=a.len;
+//   int64_t lb=b.len;
 
   char *da = a.str;
   char *db = b.str;

### 2021-11-13 regression/src/sorting.c

@@ -1,3 +1,5 @@
+#error obsolete, replaced by sorting.cpp
+
 #include <stdio.h>
 #include <stdint.h>
 #include <string.h>
@@ -58,7 +60,7 @@ void test_str() {
 
 
 int main(int argc, char **argv) {
-//test_int();
+  test_int();
   test_str();
 }
 

### 2021-11-13 regression/src/sorting.cpp

@@ -1,0 +1,142 @@
+#include <cstdlib>
+#include <cstdint>
+#include <cassert>
+#include <chrono>
+#include <iostream>
+#include <algorithm>
+#include <vector>
+#include <cmath>
+
+extern "C" {
+  #include "sorting.h"
+}
+
+
+size_t const N = 1000000;
+size_t const STRLEN = 50;
+size_t const NSTR = 500000;
+
+using namespace std;
+
+
+
+
+template <class D,class T> void check_sorted(vector<T> data, D comp) {
+  assert (is_sorted(data.begin(), data.end(),comp));
+}
+
+template<typename A> void time_op(std::string name, A op) {
+  auto start = std::chrono::system_clock::now();
+  op();
+  auto finish = std::chrono::system_clock::now();
+
+  std::chrono::duration<double> duration = finish - start;
+
+  std::cout<<"  "<<name<<"."<<"time: "<<round(duration.count()*1000)<<std::endl;
+}
+
+template<class IA, class compare, class Op>
+void sort_test(string name, std::vector<IA> &A, compare comp, Op op, bool verify=true) {
+  time_op(name, [&]{op(A);});
+
+//   for (auto i : A) clog<<i<<" ";
+//   clog<<endl;
+
+  if (verify) check_sorted(A,comp);
+}
+
+void isa_introsort_uint(vector<uint64_t> &xs) {
+  introsort(xs.data(),0,xs.size());
+}
+
+void isa_pdqsort_uint(vector<uint64_t> &xs) {
+  pdqsort(xs.data(),0,xs.size());
+}
+
+void isa_parqsort_uint(vector<uint64_t> &xs) {
+  par_sort(xs.data(),xs.size());
+}
+
+void isa_introsort_llstring(vector<llstring> &xs) {
+  str_introsort(xs.data(),0,xs.size());
+}
+
+void isa_pdqsort_llstring(vector<llstring> &xs) {
+  str_pdqsort(xs.data(),0,xs.size());
+}
+
+void isa_parqsort_llstring(vector<llstring> &xs) {
+  str_par_sort(xs.data(),xs.size());
+}
+
+
+
+template <class Op> void sort_test_uint64(string name, Op op) {
+  srand(1);
+
+  vector<uint64_t> xs;
+  for (size_t i=0;i<N;++i) xs.push_back(rand());
+  sort_test(name, xs, less<uint64_t>(),op);
+}
+
+
+char rand_char() {
[…]

### 2021-11-13 regression/src/test_avx512f.c

@@ -1,0 +1,76 @@
+#include <stdio.h>
+#include <stdint.h>
+#include <assert.h>
+#include <math.h>
+#include <immintrin.h>
+#include <avx512fintrin.h>
+#include "test_avx512f.h"
+
+#include "double_lib.c"
+
+
+double fmadd_sd(double a, double b, double c) {
+  return _mm_cvtsd_f64(_mm_fmadd_sd(_mm_set_sd(a), _mm_set_sd(b), _mm_set_sd(c)));
+}
+
+double c_test_avx_fp(double a, double b) {
+  return (sqrt(a*a + b*b) - a/b) + a;
+}
+
+double c_test_avx(double a, double b) {
+  return (sqrt(fmadd_sd(b,b,a*a)) - a/b) + a;
+}
+
+
+void test_avx(double a, double b) {
+
+//   printf("%.20f, %.20f\n",a,b);
+
+  double rv_n = c_test_avx(a,b);
+  double rv_fp = c_test_avx_fp(a,b);
+  double t_n = test_avx512f_to_nearest(a,b);
+  double t_up = test_avx512f_to_pinf(a,b);
+  double t_down = test_avx512f_to_ninf(a,b);
+  double t_z = test_avx512f_to_zero(a,b);
+
+//   printf("t_n  = %.20e\nrv_n = %.20e\n",t_n,rv_n);
+
+  assert (d2i(t_n) == d2i(rv_n));
+
+  assert (nd(t_n) == nd(rv_n));
+
+  // fix bounds, should they be nan
+  if (isnan(t_down)) t_down=-INFINITY;
+  if (isnan(t_up)) t_up=INFINITY;
+
+  assert (isnan(rv_n) || t_down <= rv_n && rv_n <= t_up);
+  assert (isnan(t_n) || t_down <= t_n && t_n <= t_up);
+  assert (isnan(t_z) || t_down <= t_z && t_z <= t_up);
+  assert (isnan(rv_fp) || t_down <= rv_fp && rv_fp <= t_up);
+
+//   printf("%.30f\n%.30f\n%.30f\n%.30f\n%.30f\n",rv_n,t_n,t_up,t_down,t_z);
+
+//   if (t_n != t_down && t_n != t_up) printf("d < n < u\n");
+//   if (t_z != t_down && t_z != t_up) printf("d < z < u\n");
+//   if (rv_n != rv_fp) printf("rv_n != rv_fp\n");
+
+
[…]

### 2021-11-13 regression/src/test_double.c

@@ -1,0 +1,42 @@
+#include <stdio.h>
+#include <stdint.h>
+#include <assert.h>
+#include <math.h>
+#include "test_double.h"
+
+#include "double_lib.c"
+
+double c_test_double(double a, double b) {
+  return fmod (sqrt(a*a + b*b) - a/b, a+b);
+}
+
+void test(double a, double b) {
+
+  assert(isnan(a) || i2d(d2i(a)) == a);
+
+  double rv = c_test_double(a,b);
+  double v = test_double(a,b);
+
+//   printf("%e,%e\n",a,b);
+
+
+  assert(d2i(v) == d2i(rv));
+
+  assert(isnan(v) && isnan(rv) || v==rv );
+}
+
+int main(int argc, char **argv) {
+
+  test(42,15);
+
+  srand(42);
+
+//   for (size_t i=0;i<100; ++i) {
+//     uint64_t x = ru64();
+//     printf("%lx %.10e\n",x,i2d(x));
+//   }
+
+  for (size_t i=0;i<1000000; ++i)
+    test(rd1(),rd1());
+}
+

### 2021-11-13 src/lib_isabelle_llvm.c

@@ -1,3 +1,7 @@
+/*
+ * Isabelle LLVM support library
+ */
+
 #include <stdlib.h>
 #include <stdio.h>
 

### 2021-11-13 src/lib_isabelle_llvm_par.cpp

@@ -1,0 +1,11 @@
+/*
+ * Isabelle LLVM Parallel support library
+ */
+#include <future>
+extern "C" {
+  void isabelle_llvm_parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2) {
+    auto s1 = std::async(std::launch::async, f1,x1);
+    f2(x2);
+    s1.get();
+  }
+}

### 2021-11-13 thys-2018/examples/code/bin_search_cli.c

@@ -1,44 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-#include <time.h>
-
-typedef uint64_t elem_t;
-
-typedef struct {
-  int64_t len;
-  elem_t *data;
-} larray_t;
-
-int64_t bin_search (larray_t xs, elem_t x);
-
-int main(int argc, char **argv) {
-
-  if (argc != 2) {printf("usage: bs <size>\n"); exit(1);}
-  
-  int64_t len = strtol(argv[1],NULL,10);
-
-//  int64_t len = 10000000;
-
-  elem_t* ptr = (elem_t*)calloc(len,sizeof(elem_t));
-  
-  for (int64_t i=0;i<len;++i) ptr[i] = ((elem_t)i) * 5;
-  
-  larray_t arr = { len, ptr };
-  
-  clock_t time = clock();
-  
-  int64_t i = 0;
-  
-  for (elem_t x=0;x<5*len;x+=2)
-    i += bin_search(arr,x);
-  
-  time = clock() - time;  
-    
-  printf("Time: %fs, (idx checksum: %ld)\n",((double)time)/CLOCKS_PER_SEC,i);
-  
-  uint64_t time_ms = ((double)time)/CLOCKS_PER_SEC * 1000;
-  printf("@ %ld %ld\n",len,time_ms);
-  
-}

### 2021-11-13 thys-2018/examples/code/kmpcli.c

@@ -1,87 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-#include <time.h>
-
-typedef struct {
-  int64_t len;
-  char *str;
-} string;
-
-int64_t kmp(string,string);
-
-
-string string_of_str(char *s) { string r = {strlen(s),s}; return r;}
-
-void error(char * msg) {
-  printf("Error: %s\n",msg);
-  exit(1);
-}
-
-
-string string_of_file(char *fname) {
-  // code snippet found on fundza.com
-  /* declare a file pointer */
-  FILE    *infile;
-  char    *buffer;
-  int64_t numbytes;
-  
-  /* open an existing file for reading */
-  infile = fopen(fname, "r");
-  
-  /* quit if the file does not exist */
-  if(infile == NULL) error("No such file");
-  
-  /* Get the number of bytes */
-  fseek(infile, 0L, SEEK_END);
-  numbytes = ftell(infile);
-  
-  /* reset the file position indicator to 
-  the beginning of the file */
-  fseek(infile, 0L, SEEK_SET);	
-  
-  /* grab sufficient memory for the 
-  buffer to hold the text */
-  buffer = (char*)calloc(numbytes, sizeof(char));	
-  
-  /* memory error */
-  if(buffer == NULL) error("Out of memory");
-  
-  /* copy all the text into the buffer */
-  fread(buffer, sizeof(char), numbytes, infile);
-  fclose(infile);
-  
-  string res = {numbytes,buffer};
-  return res;
-}
-
-void free_string(string s) { free (s.str); }
-
-
-
-int main (int argc, char** argv) {
-  
-  if (argc != 4) {printf("usage: kmp <num-iterations> <search-string> <textfile>\n"); exit(1);}
-  
-  int n = strtol(argv[1],NULL,10);
-  string s1 = string_of_str(argv[2]);
-  string s2 = string_of_file(argv[3]);
-  
-  clock_t time = clock();
-  
-  for (int i=1;i<n;++i) kmp(s1,s2);
-  int64_t res = kmp(s1,s2);
-
-  time = clock() - time;
-  double dtime = ((double)time)/CLOCKS_PER_SEC*1000000/n;
-    
-  printf("Time: %fus\n",dtime);
-  printf("result (%d iterations) = %ld\n", n, res);
-  
-  printf("@(%d) '%s': %d\n",n, s1.str, (int)(dtime));
-  
-  free_string(s2);
-  
-  return 0;
-}

### 2021-11-13 thys-2020/examples/code/bin_search.h

@@ -1,14 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _BIN_SEARCH_H
-#define _BIN_SEARCH_H 1
-
-
-typedef uint64_t elem_t;
-typedef struct {
-  int64_t len;
-  elem_t*data;
-} larray_t;
-
-int64_t bin_search(larray_t, elem_t);
-
-#endif

### 2021-11-13 thys-2020/examples/code/euclid.h

@@ -1,10 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _EUCLID_H
-#define _EUCLID_H 1
-
-
-
-
-uint64_t euclid(uint64_t, uint64_t);
-
-#endif

### 2021-11-13 thys-2020/examples/code/exp.h

@@ -1,11 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _EXP_H
-#define _EXP_H 1
-
-
-
-
-uint64_t exp64(uint64_t);
-uint32_t exp32(uint32_t);
-
-#endif

### 2021-11-13 thys-2020/examples/code/heapsort.h

@@ -1,10 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _HEAPSORT_H
-#define _HEAPSORT_H 1
-
-
-
-
-uint64_t* heapsort(uint64_t*, int64_t, int64_t);
-
-#endif

### 2021-11-13 thys-2020/examples/code/introsort.h

@@ -1,25 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _INTROSORT_H
-#define _INTROSORT_H 1
-
-
-typedef struct {
-  int64_t size;
-  struct {
-    int64_t capacity;
-    char*data;
-  };
-} llstring;
-
-char llstrcmp(llstring*, llstring*);
-void str_init(llstring*);
-void str_append(llstring*, char);
-llstring* str_pdqsort(llstring*, int64_t, int64_t);
-uint64_t* pdqsort(uint64_t*, int64_t, int64_t);
-llstring* str_introsort(llstring*, int64_t, int64_t);
-uint64_t* heapsort(uint64_t*, int64_t, int64_t);
-uint64_t* introsort(uint64_t*, int64_t, int64_t);
-uint64_t* introsort_aux(uint64_t*, int64_t, int64_t, int64_t);
-uint64_t* insertion_sort(uint64_t*, int64_t, int64_t);
-
-#endif

### 2021-11-13 thys-2020/examples/code/introsortXXX.h

@@ -1,23 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _INTROSORTXXX_H
-#define _INTROSORTXXX_H 1
-
-
-typedef struct {
-  int64_t size;
-  struct {
-    int64_t capacity;
-    char*data;
-  };
-} llstring;
-
-char llstrcmp(llstring*, llstring*);
-void str_init(llstring*);
-void str_append(llstring*, char);
-llstring* str_introsort(llstring*, int64_t, int64_t);
-uint64_t* heapsort(uint64_t*, int64_t, int64_t);
-uint64_t* introsort(uint64_t*, int64_t, int64_t);
-uint64_t* insertion_sort(uint64_t*, int64_t, int64_t);
-uint64_t* introsort_aux(uint64_t*, int64_t, int64_t, int64_t);
-
-#endif

### 2021-11-13 thys-2020/examples/code/introsort_param.h

@@ -1,24 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _INTROSORT_PARAM_H
-#define _INTROSORT_PARAM_H 1
-
-
-typedef struct {
-  int64_t size;
-  struct {
-    int64_t capacity;
-    int32_t*data;
-  };
-} array_list_32_64;
-typedef struct {
-  int64_t size;
-  struct {
-    int64_t capacity;
-    int64_t*data;
-  };
-} array_list_64_64;
-
-uint64_t* heapsort_idxs(array_list_32_64, uint64_t*, int64_t, int64_t);
-uint64_t* introsort_idxs(array_list_32_64, uint64_t*, int64_t, int64_t);
-
-#endif

### 2021-11-13 thys-2020/examples/code/kmp.h

@@ -1,13 +1,0 @@
-// Generated by Isabelle-LLVM. Do not modify.
-#ifndef _KMP_H
-#define _KMP_H 1
-
-
-typedef struct {
-  int64_t len;
-  char*str;
-} string;
-
-int64_t kmp(string, string);
-
-#endif

### 2021-11-13 thys-2020/examples/code/kmpcli.c

@@ -1,88 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-#include <time.h>
-#include "kmp.h"
-
-/*typedef struct {
-  int64_t len;
-  char *str;
-} string;
-
-int64_t kmp(string,string);
-*/
-
-string string_of_str(char *s) { string r = {strlen(s),s}; return r;}
-
-void error(char * msg) {
-  printf("Error: %s\n",msg);
-  exit(1);
-}
-
-
-string string_of_file(char *fname) {
-  // code snippet found on fundza.com
-  /* declare a file pointer */
-  FILE    *infile;
-  char    *buffer;
-  int64_t numbytes;
-  
-  /* open an existing file for reading */
-  infile = fopen(fname, "r");
-  
-  /* quit if the file does not exist */
-  if(infile == NULL) error("No such file");
-  
-  /* Get the number of bytes */
-  fseek(infile, 0L, SEEK_END);
-  numbytes = ftell(infile);
-  
-  /* reset the file position indicator to 
-  the beginning of the file */
-  fseek(infile, 0L, SEEK_SET);	
-  
-  /* grab sufficient memory for the 
-  buffer to hold the text */
-  buffer = (char*)calloc(numbytes, sizeof(char));	
-  
-  /* memory error */
-  if(buffer == NULL) error("Out of memory");
-  
-  /* copy all the text into the buffer */
-  fread(buffer, sizeof(char), numbytes, infile);
-  fclose(infile);
-  
-  string res = {numbytes,buffer};
-  return res;
-}
-
-void free_string(string s) { free (s.str); }
-
-
-
-int main (int argc, char** argv) {
-  
-  if (argc != 4) {printf("usage: kmp <num-iterations> <search-string> <textfile>\n"); exit(1);}
-  
-  int n = strtol(argv[1],NULL,10);
-  string s1 = string_of_str(argv[2]);
-  string s2 = string_of_file(argv[3]);
-  
-  clock_t time = clock();
-  
-  for (int i=1;i<n;++i) kmp(s1,s2);
-  int64_t res = kmp(s1,s2);
-
-  time = clock() - time;
-  double dtime = ((double)time)/CLOCKS_PER_SEC*1000000/n;
-    
-  printf("Time: %fus\n",dtime);
-  printf("result (%d iterations) = %ld\n", n, res);
-  
[…]

### 2021-11-13 thys-2020/sepref/IICF/Impl/Heaps/sort_cli.c

@@ -1,35 +1,0 @@
-#include <stdint.h>
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-#include <time.h>
-
-
-typedef struct {
-  int64_t l;
-  int64_t c;
-  int64_t *data;
-} array_list;
-
-extern array_list sort(array_list);
-extern array_list arl_new();
-extern array_list arl_push_back(array_list,int64_t);
-
-
-int main(int argc, char **argv) {
-  array_list a;
-  a.l=0;
-  a.c=0;
-  a.data=0;
-  
-  a = arl_new();
-  
-  printf("%ld, %ld\n",a.l,a.c);
-  
-  arl_push_back(a,42);
-  
-  // for (int i=0;i<5;i++) a=arl_push_back(a,i);
-  
-  return 0;
-}
-

### 2021-11-13 thys/basic/kernel/make_tmpls.sh

@@ -1,0 +1,13 @@
+#!/bin/bash
+
+for i in *.tmpl.ll; do
+  o="${i%.ll}.ml"
+
+  echo $i " --> " $o
+
+  echo "(* Generated code, do not modify *)" > $o
+  echo "(* Generated from $i on $(date) by $(whoami) *)" >> $o
+  sed -re 's/\$\{([A-Za-z0-9_]*)\}/" ^ \1 ^ "/g; s/\$([A-Za-z0-9_]*)/" ^ \1 ^ "/g; s/^$|^[^&].*/"\0\\n" ^/;s/^& ?//' $i >> $o
+done
+
+

### 2021-11-13 thys/examples/code/introsort.h

@@ -14,14 +14,14 @@ typedef struct llstring {
 } llstring;
 
 char llstrcmp(llstring*, llstring*);
+void str_free(llstring*);
 void str_init(llstring*);
 void str_append(llstring*, char);
 llstring* str_pdqsort(llstring*, int64_t, int64_t);
+llstring* str_par_sort(llstring*, int64_t);
 uint64_t* pdqsort(uint64_t*, int64_t, int64_t);
 llstring* str_introsort(llstring*, int64_t, int64_t);
-uint64_t* heapsort(uint64_t*, int64_t, int64_t);
+uint64_t* par_sort(uint64_t*, int64_t);
 uint64_t* introsort(uint64_t*, int64_t, int64_t);
-uint64_t* introsort_aux(uint64_t*, int64_t, int64_t, int64_t);
-uint64_t* insertion_sort(uint64_t*, int64_t, int64_t);
 
 #endif

### 2022-01-19 papers/llvm_par/ESOP22_rebuttal.txt
ESOP2022 submission

Dear reviewers. Thank you for your useful reviews. We'll try to clarify the paper accordingly,
and make it accessible to a larger audience by using less Isabelle-specific notation.
With this rebuttal, we're trying to answer your questions and clarify misunderstandings.

The paper finds a sweet spot between expressiveness and simplicity:
the semantics is simple enough to be integrated into a well-proven scalable
shallow-embedding approach, and expressive enough for interesting algorithms.
We now can parallelize some complex and highly optimized algorithms, making
them more efficient in practice. Instead of many toy examples, we decided for
one substantial case study.
If this is not enough contribution for ESOP, we accept that.

@Reviewer A: the hnr predicate and sepref tool is generic. The formalized data structures are typically re-usable.
The application-specific boilerplate is to set up synthesis of refined versions of the program, procedure by procedure,
specifying what data structures to use, and making sure the program contains enough assertions to justify
the refinement (e.g., that computations don't overflow).

@Reviewer B: the main challenge was to find the sweet spot between expressiveness and compatibility with
  existing techniques. Once found, implementing the tool became practical. As it allows re-use of sequential algorithms,
  implementing a parallel sorting algorithm based on highly optimized sequential algorithms became feasible.

@Reviewer C: unfortunately, the review is based on a bunch of misunderstandings.
We will try to clarify the paper accordingly, to make it more readable for the Coq/IRIS community.

Overclaiming:
  1) we formalize a sufficiently large part of LLVM to generate practical code.
  Adding additional features would be useless for our application, and unnecessarily increase the trusted code base.

  2) we, indeed, were not aware of the Actris example. It verifies a simple, unoptimized, abstract algorithm, while
    we verify an optimized concrete implementation. We'll relativise the claim.

  3) [39] handles total correctness of *sequential programs*, and leaves parallel programs for future work.

Comparison with IRIS:
IRIS and the Isabelle Refinement Framework (IRF) have the same ultimate goal, providing verified software.
While IRIS tries to be super-expressive, supporting every thinkable feature of concurrency,
features to the IRF are added in an incremental way, driven by the need of verifying actual implementations.
While IRIS, eventually, will be able to verify actual implementations of most real-world concurrent algorithms,
the IRF can verify actual implementations of some important parallel algorithms now.

Misunderstandings about what is verified:

The separation logic rules/Hoare triples *are* proved/defined wrt. the underlying semantics.

The translation to LLVM *is* verified, down to our concise LLVM semantics.
This goes much further than the typical code extraction to a functional language,
but not as far as tools like CompCert, that also verify the optimization and machine code generation.

Using sorted and mset is Isabelle's established way of specifying a sorting algorithm.
arrA and idxA specify the concepts of array / machine word. For correctness statements over low-level code,
separation of concerns into natural concepts is essential for the readability of the specification.

The parallel combinator, indeed, is asymmetric. However, as also reviewer C acknowledges,
this approach is sound for the features that we consider. We are aware (and discuss in the paper) that
extending the supported features will require adjustments. We note that the asymmetry could easily be avoided by
identifying NTERM and FAIL already in the semantics.

############### Submitted rebuttal end here #############################

Rebuttal strategy:

  Emphasize papers contributions again. (implicitly say:
    the simplicity and ease of integration is the STRONG point, not the weak one.
    The obtained result is decent!
  )

  Address reviewer 3:
    state difference in approach between this paper and IRIS!
      this paper: less general, but practical results
      IRIS: more general, but only much less complex verified algorithms.

    some of their claims are factually false! start with these.
      adequacy, hoare-triples wrt underlying semantics, verify specification, total correctness iris

    some claims are actually not severe (benign races, LLVM semantics, verify translation to LLVM, *first* verification of par-sort-algo, dubious semantics)

  Then address other reviewers:
    only one case study --> yes, but substantial one!
    the other concerns: already addressed in paper, but explain again anyway.

TCB:
  also argument: even if a formalization in IRIS style is considered more elegant,
  the TCB of my approach is still very small compared to other verification tools
  like deductive verifiers.

Ideas for rebuttal:

General: this paper finds a sweet spot, where a parallel semantics can be formalized as a simple extension to a sequential one.
  * the resulting semantics is simple, and thus easy to manually verify.
  * it's backwards compatible to the existing sequential formalization
  * and expressive enough for realistic examples

In summary: parallelism for cheap

  * existing tools and formalizations for sequential semantics can easily be adapted.
    If there is a substantial body of sequential verifications, this backwards compatibility is essential,
    as it allows reuse and thus high (efficiency) gain at low (formalization) cost.
  * while this approach is less general, it's still expressive enough for practical relevant examples

In the reviews, I see these arguments being turned around and used against the paper.

case study:
  decided to go for one substantial case study rather than many small toy examples.
  The case study shows how the extension seamlessly integrates into the already existing architecture,
  allowing to parallelize existing verified algorithm at little effort, but with great speed-up effect.

LLVM semantics:
  the presented framework goes down to a fragment of LLVM that is big enough for code generation.
  Formalizing anything more of LLVM in this context would
    1) not be necessary.
    2) even be disadvantageous, as it would increase the TCB
  Note that formalizing LLVM optimization passes is an entirely orthogonal endeavour (see VeLLVM project), and
  this project would immediately profit from a verified LLVM backend.
  I will tone down the statement, to clarify that only a fragment has been verified.

Read-only memory:
  the semantics already supports it. Just extending the separation logic is,
  actually, straightforward and it's well known how to do it.
  Extending the automation of Sepref accordingly would be less straightforward.

  The contribution of this would mainly focus on technical details of how Sepref works,
  and how to adjust it to fractional separation logic, and the practical gain would be
  to perhaps be able to include a toy-example matrix multiplication algorithm.
  Admittedly, I hoped that the contribution of this paper, including showing how to parallelize
  a realistic sorting algorithm rather than toy examples, would be sufficient.

#1: benchmark: indeed, industrial strength parallel sorting algorithms are more optimized than this one,
  and I have also included such an algorithm in the benchmarks.
  Going the other way round, i.e., starting at an industrial strength sorting algorithm and working upwards, was done for the sequential pdqsort algorithm (that I use as a subroutine).
  For this paper, while trying to use a realistic example, I also had to keep it simple enough such that it's still suited as
  an example.

In the description of the Sepref tool, you mention that "The synthesis is automatic, but usually requires some program-specific setup and boilerplate". I'm not sure I fully understand what this program-specific setup is. Are you referring to the refinement assertions and hnr rules for the program operations? Do these have to be provided by the user? I would have thought some of them are fairly generic. Could you clarify everything one would have to provide in order for the sorting algorithm to go through?
  hnr and the rules for the program operations are generic.
  What is required is some setup to tell Sepref which concrete data structures to use for the abstract data types in the program, and to ensure the abstract program contains enough assertions
  such that Sepref can prove the refinement (e.g., indexes are in bounds, numbers must fit into 64 bit words). These are program specific, and require some boilerplate code to set up.

  Sorting algorithm: enough assertions in the abstract algorithm to justify the refinement of lists to arrays and mathematical integers to 64 bit words,
  plus boilerplate to synthesize the concrete algorithm from the abstract one, subroutine by subroutine.

#2
"Still, I think that the contribution of this paper is too narrow for ESOP. It only considers one algorithm, and it seems that only small extensions of the existing components were actually needed: including disjoint parallelism construct in LLVM syntax and semantics, standard disjoint parallelism rule in separation logic, extensions of the Sepref tool. The latter part was the not sufficiently clear to me, and I was not convinced it will work for other examples."

That's actually the point of the contribution: only small extensions, in particular to the trusted code base,
are needed to get a great effect, that also integrates well with existing formalizations, and thus allows great speed-up effects by parallelization at low effort!

The approach will work for all algorithms that can be phrased in terms of a parallel combinator on disjoint memory.
I acknowledge that there may be other "refinement idioms" that are useful for different algorithms,
but these can be straightforwardly added when needed, without changing anything at the base of the formalization.

Questions for authors’ response
What were the key technical challenges to overcome in this verification effort?
> integrating the extension to parallelism without breaking the existing framework nor bloweing up the TCB too much.
Thus, existing formalizations can be re-used, and parallelism can be added cheaply.

In particular, it is expected that completely disjoint parallelism is captured as a sequential computation. Then, what is needed here beyond syntactic race detection?
> Semantic race detection. E.g., indexes of array accesses can, typically, not be proved disjoint syntactically.
  When syntactic race detection can be used to justify disjointness: good, probably the disjointness proof will be easy!

#3

Overclaims.
  The definition of the parallel combinator is dubious (page 5).
  The trusted computing base (TCB) is too large.

  Actually, the TCB is only slightly extended (as compared to the TCB for sequential algorithms):
    the extension to the semantics is straightforward, and explainable with a straightforward and obvious argument.
    It also maintains the shallow embedding of control flow, which results in an elegant and concise semantics.

    While a semantics that accounts for interleaving or even more sophisticated memory models would
    work for more general programs (with IO, sync, etc), it would loose the
    shallow embedding property, and be inherently more complicated, and thus harder to check.

Page 2 says this paper presents "the first verification of a parallel sorting algorithm". See Actris: Session-Type Based Reasoning in Separation Logic (POPL 2020).
   granted, I did not know this work. However: their sorting algorithm is a toy example, which they explicitly do not optimize for efficiency, nor can they extract efficient code.
   I'll reference the paper and weaken the claim.

Page 22 says "total correctness is a non-trivial endeavour that is subject of active research [39]". But it is unfair. [39] indeed provides a way to prove total correctness, and they can prove termination using almost the same idea employed in this paper, i.e., using the decreasing index.
  they have not (yet) done it for parallel programs, only for sequential programs. And at the cost of a much more complicated semantics (e.g. TCB).

The definition of the parallel combinator is dubious (page 5).
  it is OK for the proposed (simple) fragment. And as I suggest in the paper, will not work for more complicated semantics with sync/communication.
  The distinction between NTERM and FAIL is mainly for historical reasons. The Hoare-triple identifies them anyway, so no unsoundness is to be expected here.

The current definition doesn't support benign races.
  no, this was a deliberate decision, as this was not required. Adding such support is easily possible by keeping track of writes as addresses and values.
  This would slightly increase the complexity of the semantics.

The underlying separation logic's adequacy should be proved as in Iris [18].
  The separation logic is not part of the TCB. All rules are actually proved.

The meaning of Hoare triples should be defined w.r.t. the underlying semantic model as in Iris [18].
  It actually is!

The arrA, idxA, sorted, mset functions should be verified.
  Using "sorted" and "mset" is the standard Isabelle way of specifying a sorting algorithm!
  Other standard specifications can easily be derived from them.

  The relation arrA defines the concept of an array in memory, mapping it to a list of elements.
  And idxA defines the concept of a fixed-bit word, mapping it to a mathematical integer.
  While it might be possible to merge their definitions with the definition of sortedness, the resulting
  specification would be much harder to read. It was a deliberate decision to split the specification into
  natural, and easy to verify parts, rather than to rely on a monolithic and complex specification.

The translation to LLVM should be verified.
  It is verified down to the shallowly embedded semantics of LLVM.
  As any translation can only be verified against a semantics of the target, and the
  shallowly embedded semantics used in this paper is concise and elegant, and I don't get the point.
  Formalizing more complete and complicated deeply embedded semantics of LLVM, and translating from
  the shallow embedding to these deep embedding is an orthogonal issue.

Questions for authors’ response
In my eyes, many of the claimed contributions are actually already made by prior work,
for the reasons I describe above. Would you please elaborate on your contributions
so that I can better judge this paper?

  see above.

Overall merit
B.
OK paper, but I will not champion it

Confidence
Y.
I am knowledgeable in this area, but not an expert

Paper summary
The paper presents a technique for generating verified parallel algorithms in LLVM intermediate representation with total correctness guarantees. This is based in the Isabelle Refinement Framework and makes use of a refinement approach:

The user provides a sequential abstract program. This program contains hints for the subsequent refinement steps. For instance, the npar combinator, which executes two blocks sequentially, is a hint to refine to parallel execution. Conversely, nseq is the same as npar, but it is translated to sequential execution.
The Sepref tool then refines the abstract program to a concrete LLVM representation. The authors modified the semantics of LLVM to report the read and written memory locations. This allows defining a parallel combinator that fails whenever there is a data race.
The reasoning infrastructure is based on separation logic. The authors defined a separation algebra, abstraction function from concrete memory to separation algebra, weakest precondition predicate, validity of Hoare triples, and a Verification Condition Generator.
The result is an LLVM parallel program and a refinement proof.

As a case study, the authors verify total correctness for a parallel sorting algorithm. This algorithm is based on a verified sequential version of Introsort and makes use of a verified pdqsort.

Strengths
This paper makes progress towards providing total correctness guarantees for parallel algorithms in the Isabelle Refinement Framework.

Weaknesses
It only discusses one benchmark.
There is no support for read-only shared memory, meaning that only threads that run on disjoint memory can to be executed in parallel. The authors do discuss this as future work.
Explanation of your recommendation and comments to authors
I liked this paper and I think it makes a worthy contribution by taking a step towards providing total correctness guarantees for parallel algorithms. While there are some shortcomings (one single benchmark and no support for shared read-only memory), this paper still makes reasonable progress.

Comments/questions:

It would have been nice to see at least another case study.
For the definition of the parallel combinator, I think that the definition for "conflict a1 a2" is missing an outer negation?
With respect to presentation, I think that discussing a bit more about the refinement process (and Sepref) earlier in the paper could better motivate the formalisation. As it is now, I didn't fully understand the big picture and the aim of all the formalisation until quite late in the paper.
For the performance comparison to the C++ version, I guess ideally you would be using some existing optimised implementation rather than implementing your own. However, I imagine such an implementation doesn't exist. Probably the easiest way is to go the other way around and start from an existing optimised implementation for which you then generate your own LLVM one. That would make for an interesting comparison.
Typos:

page 3: "separation logic 3" --> "separation logic (Section 3)"
page 12: "how to refine this two an actual" --> "how to refine this to an actual"
Questions for authors’ response
In the description of the Sepref tool, you mention that "The synthesis is automatic, but usually requires some program-specific setup and boilerplate". I'm not sure I fully understand what this program-specific setup is. Are you referring to the refinement assertions and hnr rules for the program operations? Do these have to be provided by the user? I would have thought some of them are fairly generic. Could you clarify everything one would have to provide in order for the sorting algorithm to go through?

Review #100B
Overall merit
C.
Weak paper, though I will not fight strongly against it

Confidence
Y.
I am knowledgeable in this area, but not an expert

Paper summary
This submission reports on a formal verification effort of a parallel sorting algorithm employing disjoint parallelism (race free, and moreover, the forked threads never access the same variables). Verification is carried out in a framework for refinement of sequential programs in Isabelle (IRF) using a pre-existing LLVM model in Isabelle. Some extensions of these frameworks are needed, e.g., with a programming construct for disjoint parallelism, and a Hoare rule for it. Successful verification entails (total) correctness of the LLVM code (assuming its formalized semantics is correct). The (only) case study is parallel quick-sort with certain heuristics for preventing small partitions and for picking a good pivots. Verification is done by using a simple abstract parallel program for sorting which lies between the actual implementation and the sequential abstraction.

Strengths
correctness is fully verified in Isabelle/HOL

the verified implementation is not so far from standard real worlds ones

Weaknesses
contribution seems rather narrow given previous work on sequential programs

The main solver Sepref (based on certain heuristics) is not sufficiently clear

The paper is hard to read, composed mostly by Isabelle-style code snippets

Explanation of your recommendation and comments to authors
I was surprised to see that parallel sorting algorithms were not verified in proof assistants before, and I consider this as a useful contribution.

Still, I think that the contribution of this paper is too narrow for ESOP. It only considers one algorithm, and it seems that only small extensions of the existing components were actually needed: including disjoint parallelism construct in LLVM syntax and semantics, standard disjoint parallelism rule in separation logic, extensions of the Sepref tool. The latter part was the not sufficiently clear to me, and I was not convinced it will work for other examples.

I also found the paper very hard to follow, and not sufficiently accessible for ESOP. The technical contents is given mostly as code with sparse explanations. Even well known notations (e.g., for Hoare triples and points-to assertions) are encoded as predicates as in the Isabelle implementation. This makes the paper look like a documentation of the formalization. Standard mathematical definitions and theorems (preferably in textual environments) and standard notations can help to follow the material, and identify what are the new insights in this submission.

Questions for authors’ response
What were the key technical challenges to overcome in this verification effort?

In particular, it is expected that completely disjoint parallelism is captured as a sequential computation. Then, what is needed here beyond syntactic race detection?

Review #100C
Overall merit
D.
Reject

Confidence
X.
I am an expert in this area

Paper summary
Designs a verification condition generator (VCG) based on separation logic for parallel algorithms.
Verifies the total correctness of a parallel sort algorithm.
Translates the parallel sort algorithm to LLVM.
Strengths
Designs a decent VCG for parallel algorithms.
Weaknesses
Overclaims.
The definition of the parallel combinator is dubious (page 5).
The trusted computing base (TCB) is too large.
Explanation of your recommendation and comments to authors
Overclaims.

Page 1 says this paper presents "the first formalization of parallelism for a practically usable LLVM semantics." But the semantics presented in Section 2 is too simple to be called a formalization of LLVM semantics.

Page 2 says this paper presents "the first verification of a parallel sorting algorithm". See Actris: Session-Type Based Reasoning in Separation Logic (POPL 2020).

Page 22 says "total correctness is a non-trivial endeavour that is subject of active research [39]". But it is unfair. [39] indeed provides a way to prove total correctness, and they can prove termination using almost the same idea employed in this paper, i.e., using the decreasing index.

The definition of the parallel combinator is dubious (page 5).

If m1 is NTERM and m2 is FAIL, then the result is NTERM, which is quite wrong because m2 is FAIL. Also, it introduces asymmetry: if m1 and m2 are FAIL and NTERM, respectively, then the result is FAIL.

Identifying NTERM and FAIL is maybe okay for the supposed model of parallelism, where there is no communication, but in the presence of I/O or communication, it is not justified: an NTERM thread may communicate with another thread.

The current definition doesn't support benign races.

The trusted computing base (TCB) is too large.

I think at least:

The underlying separation logic's adequacy should be proved as in Iris [18].
The meaning of Hoare triples should be defined w.r.t. the underlying semantic model as in Iris [18].
The arrA, idxA, sorted, mset functions should be verified.
The translation to LLVM should be verified.

Questions for authors’ response
In my eyes, many of the claimed contributions are actually already made by prior work,
for the reasons I describe above. Would you please elaborate on your contributions
so that I can better judge this paper?

### 2022-03-08 papers/2022_Enschede_Talk/pres.i.tex
new stuff

\documentclass[fleqn]{beamer}

\mode<presentation>
{
  \usetheme{default}
%   \useinnertheme[shadow=true]{rounded}

  \useinnertheme{circles}

%    \useoutertheme{infolines}
  % or ...
  \setbeamersize{text margin left=1em,text margin right=1em}

%   \setbeamercovered{transparent}
  % or whatever (possibly just delete it)

  \beamertemplatenavigationsymbolsempty

% Display frame numbers in footline
  \setbeamertemplate{footline}[frame number]
}

\usepackage{etex}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath}
% \usepackage{amsthm}
% \usepackage{stmaryrd}
\usepackage{lmodern}
\usepackage{mathpartir}

\usepackage{centernot}

\usepackage{colortbl}
\usepackage{multirow}

\usepackage[purexy]{qsymbols}
\usepackage{graphicx}

\usepackage{cprotect}
\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{relsize}

\usepackage[noend]{algpseudocode}

\usepackage{ifthen}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{scopes}
\usetikzlibrary{decorations}
  \usetikzlibrary{decorations.pathmorphing}
\usetikzlibrary{arrows}
\usetikzlibrary{automata}
\usetikzlibrary{positioning}
\usetikzlibrary{chains}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{shapes.callouts}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.multipart}
\usetikzlibrary{fit}
\usetikzlibrary{calc}

\usepackage{pgfplots}

% \tikzstyle{stack}=[inner sep=0pt,minimum size=2mm]
% \tikzstyle{ssline}=[->,snake=snake,segment amplitude=.2mm,segment length=3mm,line after snake=1mm]
% \tikzstyle{fgnode}=[circle,draw,inner sep=0pt,minimum size=2mm]

\tikzset{
  -|-/.style={
    to path={
      (\tikztostart) -| ($(\tikztostart)!#1!(\tikztotarget)$) |- (\tikztotarget)
      \tikztonodes
    }
  },
  -|-/.default=0.5,
  |-|/.style={
    to path={
      (\tikztostart) |- ($(\tikztostart)!#1!(\tikztotarget)$) -| (\tikztotarget)
      \tikztonodes
    }
  },
  |-|/.default=0.5,
}

\tikzset{
  invisible/.style={opacity=0},
  visible on/.style={alt=#1{}{invisible}},
  alt/.code args={<#1>#2#3}{%
    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}} % \pgfkeysalso doesn't change the path
  },
}

% The goal is to translate
%    \overlaynode<red,blue>{hallo};
% into
%    \node[red]{hallo};
%    \node[blue]{hallo};
\makeatletter
\def\overlaynode<#1>#2;{
        \gdef\stacknodecommonpart{#2}
        \pgfkeys{/typeset node/.list={#1}}
        % we are lazy
        % pgfkeys will translate
        %    \pgfkeys{/typeset node/.list={red,blue}}
        % into
        %    \pgfkeys{/typeset node=red}
        %    \pgfkeys{/typeset node=blue}
}
\pgfkeys{
    /typeset node/.code={
        \edef\pgf@marshal{\noexpand\node[#1]\stacknodecommonpart;}
        \pgf@marshal
    }
}
% \tikz{
%     \overlaynode<red,{blue,xshift=1}>{{Hello}}; % notice the nested {{}}
% %   \overlaynode<red,{blue,xshift=1}>[]{Hello}; % workaround
%     \overlaynode<red,{blue,rotate=5}>at(1,0)[draw]{from};
%     \overlaynode<red,{blue,scale=1.1}>[circle]at(2,0)[draw]{the};
%     \overlaynode<red,{blue,opacity=.5}>[fill=yellow!50]at(3,0){other};
%     \overlaynode<{red,rectangle callout,fill},{blue,ellipse callout,draw}>at(4,0){side};
% }
% \tikz{
%     \overlaynode<
%             {fill=red,callout absolute pointer=(45:2)},
%             {fill=yellow,callout absolute pointer=(135:2),callout pointer shorten=1cm},
%             {fill=green,callout relative pointer=(-135:2),callout pointer width=.5cm},
%             {fill=blue,callout relative pointer=(-45:2),callout pointer shorten=1cm,callout pointer width=.5cm}
%         >
%         [rectangle callout,text=white]
%         at(0,0){can you}
%     ;
% }

\def\overlaynodedrawfill{\pgfutil@ifnextchar[{\overlaynodedrawfill@opt}{\overlaynodedrawfill@opt[]}}
\def\overlaynodedrawfill@opt[#1]<#2>#3;{
    \begin{scope}[transparency group,draw=black,fill=white,line cap=round,line join=round,#1]
        \pgfmathsetmacro\pgflinewidthdouble{2\pgflinewidth}
        \overlaynode<#2>[draw=pgfstrokecolor,line width=\pgflinewidthdouble]#3;
        \overlaynode<#2>[fill=pgffillcolor]#3;
    \end{scope}
}

\makeatother

% \tikz{
%     \overlaynodedrawfill[draw=magenta,fill=cyan,opacity=.5]<
%             {callout absolute pointer=(45:2)},
%             {callout absolute pointer=(135:2),callout pointer shorten=1cm},
%             {callout relative pointer=(-135:2),callout pointer width=.5cm},
%             {callout relative pointer=(-45:2),callout pointer shorten=1cm,callout pointer width=.5cm}
%         >
%         [rectangle callout]
%         at(0,0){hear me}
%     ;
% }
%
% \tikz{
%     \overlaynodedrawfill<
%             {cloud callout,callout absolute pointer=(90:2),inner sep=-20},
%             {rectangle callout,callout absolute pointer=(210:2),minimum height=30},
%             {ellipse callout,callout absolute  pointer=(-30:2)}
%         >
%         at(0,0){nice to meet you}
%     ;
% }
%
% \tikz{
%     \overlaynodedrawfill<
%             {starburst},
%             {cloud,inner sep=-10}
%         >
%         at(0,0){where you been}
%     ;
% }
%
% \usetikzlibrary{shapes.arrows}
% \tikz{
%     \overlaynodedrawfill[arrow box,text opacity=0,minimum height=50,minimum width=50,inner xsep=-10]<
%             {rotate=22.5},
%             {rotate=45},
%             {rotate=67.5},
%             {text opacity=1}
%         >
%         at(0,0){I can}
%     ;
% }
%

\usepackage{packages/isabelle}
\usepackage{packages/isabelletags}
\usepackage{packages/isabellesym}
\usepackage{packages/comment}

% \isabellestyle{it}

\def\isachardoublequote{}%
\def\isachardoublequoteopen{}%
\def\isachardoublequoteclose{}%

\newcommand{\isainnerkeyword}[1]{{\textbf{#1}}}
\newcommand{\isasymexistsA}{\isamath{\exists_{\textsc A}\,}}

\def\isadelimproof{}
\def\endisadelimproof{}
\def\isatagproof{}
\def\endisatagproof{}
\def\isafoldproof{}
\def\isadelimproof{}
\def\endisadelimproof{}

\def\isastylescript{\sl}%

\input{lstisabelle}
\newcommand{\isai}{\lstinline[language=isabelle,basicstyle=\normalsize\ttfamily\slshape]}

\newcommand\CC{C\nolinebreak[4]\hspace{-.05em}\raisebox{.3ex}{\relsize{-1}{\textbf{++}}}}

\newcommand{\eqdef}{\mathrel{{=}_{def}}}
\newcommand{\iffdef}{\mathrel{{\mathord{\iff}\!\!}_{def}}}

\makeatletter
\newcommand*{\overlaynumber}{\number\beamer@slideinframe}
\makeatother

\AtBeginSection[] % Do nothing for \section*
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

\title{Refinement of Parallel Algorithms down to LLVM}

% \subtitle
% {Subtitle} % (optional)

\author[Peter Lammich]{Peter Lammich}
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\institute[UoM] % (optional, but mostly needed)
{ University of Twente}
% - Use the \inst command only if there are several affiliations.
% - Keep it simple, no one is interested in your street address.

\date {Feb 2022}
% {2008-12-01}

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

% \pgfdeclareimage[height=0.5cm]{university-logo}{university-logo-filename}
% \logo{\pgfuseimage{university-logo}}

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

%\mathchardef\-="2D
%\renewcommand\-{\text{-}}

\newcommand{\mc}{\color{blue}}
\newcommand{\term}[1]{{\mc#1}}

\let\olddisplaystyle\displaystyle
\newcommand{\mydisplaystyle}{\olddisplaystyle\mc}
\let\displaystyle\mydisplaystyle

\newcommand{\smc}{\everymath{\mc}}
\smc

\lstset{basicstyle=\color{blue}}

% \newcommand<>{\btikzset}[2]{\alt#3{\tikzset{#1}}{\tikzset{#2}}}

\tikzset{onslide/.code args={<#1>#2}{%
  \only<#1>{\pgfkeysalso{#2}} % \pgfkeysalso doesn't change the path
}}

\tikzset{uncover/.code args={<#1>#2}{%
  \uncover<#1>{\pgfkeysalso{#2}} % \pgfkeysalso doesn't change the path
}}

\tikzset{>=latex}

\lstset{autogobble}

\newcommand{\natN}{{\text{nat}_{\mathord{<}N}}}

\newcommand{\high}[1]{{\color{blue}#1}}

\begin{document}
% \input{macros}

\begin{frame}
  \titlepage
\end{frame}

\newcommand{\insertsectitle}{}

\setbeamertemplate{frametitle}{\vspace{.7em}\insertframetitle\hfill \small\raisebox{10pt}{\insertsectitle}}

# The Isabelle Refinement Framework
  \definecolor{colA}{RGB}{11, 97,164}
  \definecolor{colB}{RGB}{204,246,  0}
  \definecolor{colC}{RGB}{166,  0,166}
  \definecolor{colD}{RGB}{255,146,  0}
  \definecolor{colTCB}{RGB}{255, 0, 0}

  \center
  \begin{tikzpicture}[node distance=.01\textwidth,outer sep=0pt,
  basicLayer/.style={
    draw=black,
    rounded corners,
    text width=.9\textwidth,
    font={\sffamily\bfseries\color{black}},
    align=center,
    minimum height=.03\textheight,
%    text height=12pt,
%    text depth=9pt
    },
  layerA/.style={basicLayer,fill=colA!30},
  layerB/.style={basicLayer,fill=colB!30},
  layerC/.style={basicLayer,fill=colC!30},
  layerD/.style={basicLayer,fill=colD!30},
  layerDD/.style={basicLayer,fill=colD!40,text width=.2\textwidth},
  layerTCB/.style={basicLayer,fill=colTCB!30,text width=.2\textwidth},
  layerTCBB/.style={basicLayer,fill=colTCB!40,text width=.2\textwidth},
  ]

  \node[layerA] (AlgDS) {Algorithms and Data Structures};

  \node[layerB, below=of AlgDS.south west, anchor=north west, text width=.437\textwidth] (ne) {NE-Monad};
  \node[layerA, right=of ne.north east, anchor=north west, text width=.433\textwidth, minimum height=.134\textheight] (llDS) {Low-Level Algorithms \\and Data Structures};

  \node[layerB, below=of ne.south west, anchor=north west, text width=.437\textwidth] (sepref) {Sepref};

  \node[layerC, below=of sepref.south west, anchor=north west] (sepl) {Program Logic and VCG};
  \node[layerD, below=of sepl.south west, anchor=north west, minimum height=.34\textheight] (sembox) {};

  \node[layerTCB, below left=of sembox.north east, minimum height=.32\textheight, text width=.537\textwidth] (tcbbox) {};

  \node[above right=of sembox.south west,font={\sffamily\bfseries\color{black}}] {Back End};

  \node[above left=of tcbbox.south east,font={\sffamily\bfseries\color{black}}] {Trusted Code Base};

%   \node[layerD, right=of sembox.west] (preproc) {Preprocessor};
%   \node[layerD, right=1cm of preproc] (instr) {Instructions};
%   \node[layerD, right=1cm of instr] (codegen) {Code Generator};

  \node[layerTCBB, below left=of tcbbox.north east] (codegen) {Code Generator};
  \node[layerTCBB, left=1cm of codegen] (instr) {Instructions};
  \node[layerDD, left=1cm of instr] (preproc) {Preprocessor};

  \node[layerTCBB, below=of instr] (mmodel) {Memory Model};
  \node[layerTCBB, below=of mmodel] (mmonad) {SE-Monad};

  \draw[->] (preproc) -- (instr);
  \draw[->] (instr) -- (codegen);

  \end{tikzpicture}

{

\usebackgroundtemplate{%
\only<3->{
\tikz\node[opacity=0.1] {\includegraphics[height=\paperheight]{isabelle-llvm.png}};
}
\only<2>{
\tikz\node[opacity=1] {\includegraphics[height=\paperheight]{isabelle-llvm.png}};
}
}

#! Isabelle LLVM Back End
  *<2-> Shallow embedding of small fragment of LLVM
    * just enough to express our programs
    * code generator translates to actual LLVM text
  *<3-> Simple memory model {\small
    \begin{lstlisting}
      datatype addr == ADDR (bidx: nat) (idx: nat)
      datatype ptr == PTR_NULL   |   PTR_ADDR (the_addr: addr)
      datatype val == LL_INT lint   |   LL_STRUCT val list   |   LL_PTR ptr

      datatype block == FRESH   |   FREED   |   is_alloc: ALLOC (vals: "val list")
      typedef memory == { \mu :: nat => block. finite {b. \mu b ~= FRESH} }
    \end{lstlisting}}

  *<4-> Using state-error monad
    \begin{lstlisting}
      \alpha llM = memory => (FAIL | SUCC (\alpha \x memory))
    \end{lstlisting}

}

#! Shallow Embedding Example
  \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
  {\small\centering
  \begin{minipage}[t]{.35\textwidth}
    \begin{lstlisting}[escapechar=!]
      fib:: 64 word !\mknd{ndparams}!=> 64 word ll!\mknd{ndrettype}!M
      fib n = do {
        t <- ll_icmp_ule n 1;
        llc_if!\mknd{stmtif}! t
          (return n)
          (do {
            n_1 <- ll_sub!\mknd{stmtsub}! n 1;
            a   <- fib!\mknd{stmtrec}! n_1;
            n_2 <- ll_sub n!\mknd{ndargs}! 2;
            b   <-!\mknd{ndbind}! fib n_2;
            c   <- ll_add a b;
            return!\mknd{ndreturn}! c
          }) }
    \end{lstlisting}
%     \begin{tikzpicture}[remember picture,overlay,ND/.style={rectangle callout, anchor=west,draw , fill=red!50, callout absolute pointer={#1}}]
%       \node[ND=(ndrettype.east),anchor=south] at ($(ndrettype.north) + (0,.5cm)$) {\footnotesize state/error monad};
%       \pause
%       \node[ND=(ndparams.east)] at ($(ndparams -| 6,0) + (0,.5cm)$) {\footnotesize parameter types: words, pointers, pairs};
%       \pause
%       \node[ND=(stmtif.east)] at ($(stmtif -| 6,0) + (0,.5cm)$) {\footnotesize control flow (if, [optional: while])};
%       \pause
%       \node[ND=(stmtsub.east)] at ($(stmtsub -| 6,0)+(0,.5cm)$) {\footnotesize standard instructions};
%       \pause
%       \node[ND=(stmtrec.east)] at ($(stmtrec -| 6,0)+(0,.5cm)$) {\footnotesize function calls};
%       \pause
%       \node[ND=(ndargs.east)] at ($(ndargs -| 6,0)+(0,.5cm)$) {\footnotesize arguments: variables and constants};
%       \pause
%       \node[ND=(ndbind.east)] at ($(ndbind -| 6,0)+(0,.5cm)$) {\footnotesize monad: bind, return};
%     \end{tikzpicture}
    \begin{tikzpicture}[remember picture,overlay,ND/.style={anchor=west}]
      \pause
      \node[ND,anchor=south] at ($(ndrettype.north) + (0,.5cm)$) (X) {\footnotesize state/error monad};
      \draw (X) -- (ndrettype);

      \pause
      \node[ND] at ($(ndparams -| 6,0) + (0,.5cm)$) (X) {\footnotesize types: word, pointer, struct};
      \draw (X.west) -- (ndparams);

      \pause
      \node[ND=(ndbind.east)] at ($(ndbind -| 6,0)+(0,.5cm)$) (X) {\footnotesize monad: bind, return};
      \draw (X.west) -- (ndbind);
      \draw (X.west) -- (ndreturn);

      \pause
      \node[ND=(stmtsub.east)] at ($(stmtsub -| 6,0)+(0,.5cm)$) (X) {\footnotesize standard instructions (\is{ll_<opcode>})};
      \draw (X.west) -- (stmtsub);

      \pause
      \node[ND=(ndargs.east)] at ($(ndargs -| 6,0)+(0,.5cm)$) (X) {\footnotesize arguments: variables and constants};
      \draw (X.west) -- (ndargs);

      \pause
      \node[ND=(stmtif.east)] at ($(stmtif -| 6,0) + (0,.5cm)$) (X) {\footnotesize control flow (if, [optional: while])};
      \draw (X.west) -- (stmtif);

      \pause
      \node[ND=(stmtrec.east)] at ($(stmtrec -| 6,0)+(0,.5cm)$) (X) {\footnotesize function calls (rec.\ via fixp in ccpo)};
      \draw (X.west) -- (stmtrec);

    \end{tikzpicture}
  \end{minipage}
  }

#! Code Generation
  \uncover<2->{compiling control flow + pretty printing}

  {\small\centering
  \begin{minipage}[t]{.35\textwidth}
    \begin{lstlisting}
      fib:: 64 word => 64 word llM
      fib n = do {
        t <- ll_icmp_ule n 1;
        llc_if t

          (return n)
          (do {
            n_1 <- ll_sub n 1;
            a   <- fib n_1;
            n_2 <- ll_sub n 2;
            b   <- fib n_2;
            c   <- ll_add a b;
            return c
          }) }
    \end{lstlisting}
  \end{minipage}
  % \hspace*{.04\textwidth}
  \hfill\pause
  \begin{minipage}[t]{.57\textwidth}
    \begin{lstlisting}[language=LLVM, literate={}]
    define i64 @fib(i64 %n) {
      start:
        %t = icmp ule i64 %n, 1
        br i1 %t, label %then, label %else
      then:
        br label %ctd_if
      else:
        %n_1 = sub i64 %n, 1
        %a    = call i64 @fib (i64 %n_1)
        %n_2 = sub i64 %n, 2
        %b    = call i64 @fib (i64 %n_2)
        %c    = add i64 %a, %b
        br label %ctd_if
      ctd_if:
        %x1a = phi i64 [%n,%then], [%c,%else]
        ret i64 %x1a }
    \end{lstlisting}
  \end{minipage}
  }

#! Preprocessor
  * Only restricted terms accepted by code generator
    * good to keep code generation simple
    * tedious to write manually
  * Preprocessor transforms terms into restricted format
    * proves equality (via Isabelle kernel)
  * Motto: Keep TCB small, preprocessor makes it usable

#! Example: Preprocessing Euclid's Algorithm
  \small
  \begin{lstlisting}
  euclid :: 64 word => 64 word => 64 word
  euclid a b = do {
    (a,b) <- llc_while
      (%(a,b) => ll_cmp (a ~= b))
      (%(a,b) => if (a<=b) then return (a,b-a) else return (a-b,b))
      (a,b);
    return a }
  \end{lstlisting}
  \pause
  preprocessor defines function \is{euclid_0} and proves
  \begin{lstlisting}
  euclid a b = do {
      ab <- ll_insert_1 init a; ab <- ll_insert_2 ab b;
      ab <- euclid_0 ab;
      ll_extract_1 ab  }
  euclid_0 s = do {
    a <- ll_extract_1 s;
    b <- ll_extract_2 s;
    ctd <- ll_icmp_ne a b;
    llc_if ctd do {...; euclid_0 ...} }
  \end{lstlisting}

#! Reasoning about LLVM Programs
  *<+-> Separation Logic
    \begin{lstlisting}
      \<alpha> :: memory -> amemory :: sep_algebra
      wp c Q s == \<exists>r s'. c s = SUCC r s' \and Q r (\<alpha> s')
      {P} c {Q} == \<forall>F s. (P * F) (\<alpha> s) --> wp c (\<lambda>r s'. (Q r * F) s') s
    \end{lstlisting}

    * defined wrt.\ shallowly embedded semantics
    * proof rules are proved theorems!

  *<+-> Automation: VCG, frame inference, heuristics to discharge VCs
    * these prove theorems!

  *<+-> Basic Data Structures: signed/unsigned integers, Booleans, arrays

# Sepref
  * Semi-automatic translation of functional to imperative program
  * Data refinement to imperative DS
    * e.g. list to array
  * Proves refinement theorem

#! Example: Binary Search
  \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
  \begin{lstlisting}
    definition "bin_search xs x = do {
      (l,h) <- while (bin_search$\mknd{invar}$_invar xs x)
        (%(l,h). l<h)
        (%(l,h). do {
          ass$\mknd{assn}$ert (l<|xs| \<and> h<=|xs| \<and> l<=h);
          let m = l +$\mknd{midc}$ (h-l) div 2;
          if xs!m < x then return (m+1,h) else return (l,m)
        })
        (0,|xs|);
      return l
    }"
  \end{lstlisting}

  \onslide<2->

  \begin{tikzpicture}[remember picture,overlay,ND/.style={anchor=west}]
    \only<2->{
    \node[ND,anchor=west] at ($(invar) + (.5cm,-.5cm)$) (X) {\footnotesize invariant annotation};
    \draw (X.west) -- (invar);
    }

    \only<3->{
    \node[ND,anchor=west] at ($(assn) + (3cm,-3cm)$) (X) {\footnotesize hint for subsequent refinement};
    \draw (X.west) -- (assn);
    }

    \only<4->{
    \node[ND,anchor=west] at ($(midc) + (2,-1cm)$) (X) {\footnotesize overflow-safe midpoint computation};
    \draw (X.west) -- (midc);
    }
  \end{tikzpicture}

  \onslide<5->

  \begin{lstlisting}
  lemma bin_search_correct:
    sorted xs ==> bin_search xs x <= spec i. i=find_index (<=y) xs
  \end{lstlisting}

#! Example: Binary Search --- Refinement
  \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
  \onslide<1->

  \begin{lstlisting}[escapechar=!]
  sepref_def bin_search$_\dagger$ is bin_search
    :: "(array$_A$!\mknd{assn1}! int$_A^{64}$)$^k$ $$* (int!\mknd{assn2}!$_A^{64}$)$^k$ -> int!\mknd{assn3}!$_A^{64}$
    unfolding bin_search_def
    apply (rule hfref_with_rdomI,!\mknd{assn4}! annot_snat_const "64")
    by sep!\mknd{auto}!ref
  \end{lstlisting}
  \onslide<2->

  \begin{tikzpicture}[remember picture,overlay,ND/.style={anchor=west}]
  \only<2-3>{
    \only<2->{
    \node[ND,anchor=south] at ($(assn1) + (4cm,-3cm)$) (X) {\footnotesize hints for data refinement};
    \draw (X) -- (assn1);
    \draw (X) -- (assn2);
    \draw (X) -- (assn3);
    \draw (X) -- (assn4);
    }

    \only<3->{
    \node[ND] at ($(auto) + (0,-2cm)$) (X) {\footnotesize automatic synthesis + proof};
    \draw (X.west) -- (auto);
    }
  }
  \end{tikzpicture}

  \onslide<4->
  proves: \is{(bin_search$_\dagger$, bin_search) \<in> (array$_A$ int$_A^{64}$)$^k$ * int${_A^{64}}^k$ -> int$_A^{64}$}
  \onslide<5->

  Combination with \is{bin_search_correct} yields:
  \begin{lstlisting}
  theorem bin_search$_\dagger$_correct:
    {(array$_A$ int$_A^{64}$ xs xs$_\dagger$ * int$_A^{64}$ x x$_\dagger$ * sorted xs)}
      (bin_search$_\dagger$ xs$_\dagger$ x$_\dagger$)
    {%i$_\dagger$. \<exists>i. array$_A$ int$_A^{64}$ xs xs$_\dagger$ * int$_A^{64}$ x x$_\dagger$ * int$_A^{64}$ i i$_\dagger$ * i=find_index (<=y) xs}
  \end{lstlisting}

#! Example: Binary Search --- Generated Code
  \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}

  \begin{lstlisting}[escapechar=!]
  export_llvm bin_se!\mknd{isaconst}!arch$_\dagger$ is int64_t bin_!\mknd{csig}!search(larray_t, elem_t)
  defines
    typedef !\mknd{tdefs}!int64_t elem_t;
    typedef struct { int64_t len; elem_t *data; } larray_t;
  file "!\mknd{file}!code/bin_search.ll"
  \end{lstlisting}

  \begin{tikzpicture}[remember picture,overlay,ND/.style={anchor=west}]
  \only<2-5>{
    \only<2->{
    \node[ND,anchor=south] at ($(isaconst) + (0,-3cm)$) (X) {\footnotesize Isabelle constant};
    \draw (X) -- (isaconst);
    }

    \only<3->{
    \node[ND] at ($(csig) + (0,-3cm)$) (X) {\footnotesize LLVM name and signature};
    \draw (X.west) -- (csig);
    }

    \only<4->{
    \node[ND] at ($(tdefs) + (0,-3cm)$) (X) {\footnotesize C-style typedefs};
    \draw (X.west) -- (tdefs);
    }

    \only<5->{
    \node[ND] at ($(file) + (0,-3cm)$) (X) {\footnotesize target file};
    \draw (X.west) -- (file);
    }
  }
  \end{tikzpicture}

  \onslide<6->

  Produces LLVM code and header file:
  \begin{lstlisting}[language=C]
    typedef int64_t elem_t;
    typedef struct {
      int64_t len;
      elem_t*data;
    } larray_t;

    int64_t bin_search(larray_t,elem_t);
  \end{lstlisting}

{
\usebackgroundtemplate{%
\only<3->{
\tikz\node[opacity=0.1] {\includegraphics[height=\paperheight]{isabelle-llvm-par.png}};
}
\only<2>{
\tikz\node[opacity=1] {\includegraphics[height=\paperheight]{isabelle-llvm-par.png}};
}
}

#![t] Isabelle-LLVM Parallel
  \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}

  *<2-> Add parallel operator: $c_1 \parallel c_2$
    *<3-> execute in parallel, fail on data race
  *<4-> Shallow embedding: make program report memory accesses

    \only<4>{
    \begin{tikzpicture}[node distance = 1cm, auto]
      \node (m) {$\mu$};
      \node[below left = of m] (m1) {$\rho_1,\mu_1$};
      \node[below right = of m] (m2) {$\rho_2,\mu_2$};
      \node[below right = of m1] (m') {$\mu'$};

      \draw[->] (m) -- node[swap] {\small $c_1$}  (m1);
      \draw[->] (m) -- node {\small $c_2$}  (m2);

      \draw[->] (m1) -- (m');
      \draw[->] (m2) -- (m');

      \node[above = .7em of m'] {\tiny combine};
      \node[above = 2em of m'] {\tiny check race};

    \end{tikzpicture}
    }

    \begin{onlyenv}<5-6>
    \begin{lstlisting}
      (c_1 || c_2) \mu ==
        (r_1,\rho$\mknd{r1}$_1,\mu_1) <- c_1 \mu --"\rightcomment{execute first strand}"
        (r_2,\rho$\mknd{r2}$_2,\mu_2) <- c_2 \mu --"\rightcomment{execute second strand}"
        assert no_race \rho_1 \rho_2 --"\rightcomment{fail on data race}"
        \mu' = combine \rho_1 \mu_1   \rho_2 \mu_2 --"\rightcomment{combine states}"
        return ((r_1,r_2), \rho_1 \union \rho_2, \mu')

    \end{lstlisting}

    \begin{tikzpicture}[remember picture,overlay,ND/.style={anchor=west}]
      \only<6->{
      \node[ND,anchor=south] at ($(r1) + (3cm,-.1cm)$) (X) {\footnotesize Access report};
      \draw (X) -- (r1);
      \draw (X) -- (r2);
      }
    \end{tikzpicture}
    \end{onlyenv}

}

#! Memory Allocation
  *<+-> Currently: deterministic semantics
    * $c_1~\mu$ and $c_2~\mu$ will allocate same memory
    * cannot be combined!

  *<+-> Use nondeterminism
    \begin{lstlisting}
      \alpha llM = memory => FAIL | SUCC ((\alpha \x report \x memory) set)
    \end{lstlisting}
    * malloc nondeterministically allocates some free address
    * on combination: exclude infeasible possibilities

  \onslide<+->
  \begin{lstlisting}
    (c_1 || c_2) \mu ==
      (r_1,\rho_1,\mu_1) <- c_1 \mu
      (r_2,\rho_2,\mu_2) <- c_2 \mu
      assume \rho_1.alloc \<inter> \rho_2.alloc = \emptyset --"\rightcomment{ignore infeasible combinations}"
      assert no_race \rho_1 \rho_2 --"\rightcomment{fail on data race}"
      \mu' = combine \rho_1 \mu_1   \rho_2 \mu_2 --"\rightcomment{combine states}"
      return ((r_1,r_2), \rho_1 \union \rho_2, \mu')

  \end{lstlisting}

#! Invariants
  * We prove for \is{llM} (enforced by subtype)
    * access reports are consistent with observed changes in memory
    * there is at least one possible result (no magic happens)

  * Sanity check for semantics

  * Allows us to prove symmetry of $\parallel$
    \begin{lstlisting}
      c_1 || c_2  =  swapres (c_2 || c_1)

      swapres m == (r_1,r_2)<-m; return (r_2,r_1)
    \end{lstlisting}

#! Code generator
  *<+-> We add \is{llc_par f_1 f_2 x_1 x_2 == f_1 x_1 || f_2 x_2}
    * \is{f_1, f_2} must be functions
  *<+-> Code generator generates
    * type casting boilerplate
    * call to external \is{parallel} function

    {\small
    \begin{lstlisting}[language=C++]
    void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2)
    \end{lstlisting}
    }
  *<+-> For example, implemented using TBB:
    {\small
    \begin{lstlisting}[language=C++]
    {
      tbb::parallel_invoke([=]{f1(x1);}, [=]{f2(x2);});
    }
    \end{lstlisting}
    }

#! Amending higher layers of IRF
  *<+-> Prove concurrency rule
    \begin{lstlisting}
      ${}$        {P_1} c_1 {Q_1}    \and    {P_2} c_2 {Q_2}
    \end{lstlisting}

    \vspace*{-.8em}
    \begin{lstlisting}
      ==> {P_1 * P_2} c_1 || c_2 {%(r_1,r_2). Q_1 r_1 * Q_2 r_2}
    \end{lstlisting}

  *<+-> Sepref refines sequential to parallel execution
    \begin{lstlisting}
      npar f_1 f_2 x_1 x_2 == r_1 <- f_1 x_1; r_2 <- f_2 x_2; return (r_1,r_2)
    \end{lstlisting}
    refined to \is{llc_par}.

  *<+-> Backwards compatible with sequential Sepref!
    * Easy porting of existing algorithms

#! Parallel Quicksort (basic)

  \begin{lstlisting}
    psort xs ==
      if |xs|<=1 then return xs    --"\rightcomment{trivially sorted}"
      else
        (xs,m) <- partition_spec xs;  --"\rightcomment{partition}"
        (_,xs) <- with_split m xs (%xs_1 xs_2.
          npar psort psort xs_1 xs_2 --"\rightcomment{recursively sort partitions}"
        );
        return xs

    with_split i xs f ==
      assert (i < |xs|);       --"\rightcomment{split point must be in list}"
      (xs_1,xs_2) <- f (take i xs) (drop i xs); --"\rightcomment{execute f with halfs}"
      assert (|xs_1| = i \and |xs_2| = |xs| - i); --"\rightcomment{length of halfs must not change}"
      return (xs_1@xs_2) --"\rightcomment{return both halfs}"

  \end{lstlisting}

#! Parallel Quicksort (refined)
  \small
  \begin{lstlisting}
    psort xs n ==
      assert n=|xs|;
      if n<=1 then return xs
      else psort_aux xs n (log2 n * 2) --"\rightcomment{recursion depth limit}"

    psort_aux xs n d ==
      assert n=|xs|  --"\rightcomment{extra parameter for length}"
      if d=0 \or n<100000 then sort_spec xs --"\rightcomment{fallback to seq-sort}"
      else
        (xs,m) <- partition_spec xs;
        let bad = m<n div 8 \or (n-m < n div 8) --"\rightcomment{check unbalanced partition}"
        (_,xs) <- with_split m xs (%xs_1 xs_2.
          if bad then --"\rightcomment{sequentially recurse for unbalanced}"
            nseq psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
          else --"\rightcomment{recurse in parallel for balanced}"
            npar psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
        );
        return xs"

  \end{lstlisting}

#![t] Parallel Quicksort (Sepref + Code Export)
  \small

  *<1-> Sepref generates imperative program
    * using existing sequential \is{pdqsort} for fallback
    * using (new) sampling partitioner (proved correct + refined separately)

  *<2-> Correctness theorem:
    {\only<3->{\tiny}
    \begin{lstlisting}
    {arr$_A$ xs xs$_\dagger$ * idx$_A$ n n$_\dagger$ * n = |xs|}
      (psort$_\dagger$ xs$_\dagger$ n$_\dagger$)
    {%r. r=xs$_\dagger$ * \<exists> xs'. arr$_A$ xs' xs$_\dagger$ * sorted xs' * mset xs' = mset xs}
    \end{lstlisting}}
  *<3-> Instantiation to concrete weak ordering + code export
    {\only<4->{\tiny}
    \begin{lstlisting}
      interpretation unat: pcmp (%_. <) (%_. ll_icmp_ult) unat$^{64}_A$ \<proof>
      interpretation str: pcmp (%_. <) (%_. strcmp) str$^{64}_A$ \<proof>

      export_llvm
        unat.psort$_\dagger$ is "uint64_t* psort(uint64_t*, int64_t)"
        str.psort$_\dagger$ is "llstring* str_psort(llstring*, int64_t)"
        defines
          typedef struct {int64_t sz; struct {int64_t cap; char *data;};} llstring;
        file psort.ll
    \end{lstlisting}}

  *<4-> Link against C++ benchmark driver
    \begin{lstlisting}
      clang++ [...] lib_isabelle_llvm.cpp psort.ll benchmark.cpp
    \end{lstlisting}

\tikzset{myaxis/.style={
      ymajorgrids,
      title style={at={(0.2,.85)}},
      ybar=0pt,
      width=\textwidth,
      height=.85\textheight,
      bar width=3pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,},
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      legend style = {
%         legend pos= north east,
      at = {(.87,.97)},
%       cells={anchor=west},
      font=\scriptsize
      }
}}

# Benchmarks

  \begin{tikzpicture}
    \begin{axis}[
      title={\large Laptop (6 core), uint64},
      ymajorgrids,
      title style={at={(0.2,.85)}},
      ybar=0pt,
      width=\textwidth,
      height=.85\textheight,
      bar width=3pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,},
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      legend style = {
%         legend pos= north east,
      at = {(.87,.97)},
%       cells={anchor=west},
      font=\scriptsize
      }
    ]
      \input{data_laptop_uint64}
    \end{axis}
  \end{tikzpicture}

# Benchmarks

  \begin{tikzpicture}
    \begin{axis}[
      title={\large Laptop (6 core), string},
      ymajorgrids,
      title style={at={(0.2,.85)}},
      ybar=0pt,
      width=\textwidth,
      height=.85\textheight,
      bar width=3pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,},
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      legend style = {
%         legend pos= north east,
      at = {(.87,.97)},
%       cells={anchor=west},
      font=\scriptsize
      }
    ]
      \input{data_laptop_llstring}
    \end{axis}
  \end{tikzpicture}

# Benchmarks

  \begin{tikzpicture}
    \begin{axis}[
      title={\large Server (22 core), uint64},
      ymajorgrids,
      title style={at={(0.2,.85)}},
      ybar=0pt,
      width=\textwidth,
      height=.85\textheight,
      bar width=3pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,},
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      legend style = {
%         legend pos= north east,
      at = {(.87,.97)},
%       cells={anchor=west},
      font=\scriptsize
      }
    ]
      \input{data_server_uint64}
    \end{axis}
  \end{tikzpicture}

# Benchmarks

  \begin{tikzpicture}
    \begin{axis}[
      title={\large Server (22 core), string},
      ymajorgrids,
      title style={at={(0.2,.85)}},
      ybar=0pt,
      width=\textwidth,
      height=.85\textheight,
      bar width=3pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,},
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      legend style = {
%         legend pos= north east,
      at = {(.87,.97)},
%       cells={anchor=west},
      font=\scriptsize
      }
    ]
      \input{data_server_llstring}
    \end{axis}
  \end{tikzpicture}

# Speedup
  \begin{tikzpicture}
    \begin{axis}[
      xlabel near ticks,
      legend style = {
        at = {(.87,.3)},
        cells={anchor=west},
        font=\scriptsize
      },
      ymajorgrids,
      title={\large Laptop},
      title style={at={(0.15,.85)}},
      width=\textwidth,
      height=\textheight
    ]
      \input{sudata_laptop}

      \draw[black, thin, sharp plot]
      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});

%       \addplot[black, thin, sharp plot, update limits=false] {\x};
    \end{axis}
  \end{tikzpicture}

# Speedup
  \begin{tikzpicture}
    \begin{axis}[
      xlabel near ticks,
      legend style = {
        at = {(.87,.3)},
        cells={anchor=west},
        font=\scriptsize
      },
      ymajorgrids,
      title={\large Server},
      title style={at={(0.15,.85)}},
      width=\textwidth,
      height=\textheight
    ]
      \input{sudata_server}

      \draw[black, thin, sharp plot]
      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});

%       \addplot[black, thin, sharp plot, update limits=false] {\x};
    \end{axis}
  \end{tikzpicture}

{
\usebackgroundtemplate{%
\tikz\node[opacity=0.1] {\includegraphics[height=\paperheight]{isabelle-llvm-par.png}};
}

# Conclusion
  * Verification of parallel programs
    * stepwise refinement to tackle complexity
    * down to LLVM, small TCB
    * \high{fast} verified programs

  * Idea: shallow embedding, using access reports
    * backwards compatible with sequential IRF

  * Future work
    * state-of-the-art parallel sorting
    * fractional separation logic
    * more concurrency
    * complexity of parallel algorithms
    * GP-GPUs

}

\end{document}

### 2022-03-08 papers/2022_Enschede_Talk/pres.tex
new stuff

--- 2022-03-08 papers/2022_Enschede_Talk/pres.i.tex
@@
 
-# The Isabelle Refinement Framework
+\begin{frame}{The Isabelle Refinement Framework}
   \definecolor{colA}{RGB}{11, 97,164}
@@
 
+\end{frame}
 {
@@
 
-#! Isabelle LLVM Back End
-  *<2-> Shallow embedding of small fragment of LLVM
-    * just enough to express our programs
-    * code generator translates to actual LLVM text
-  *<3-> Simple memory model {\small
+\begin{frame}[fragile]{Isabelle LLVM Back End}
+  \begin{itemize}
+   \item<2-> Shallow embedding of small fragment of LLVM
+    \begin{itemize}
+     \item just enough to express our programs
+     \item code generator translates to actual LLVM text
+    \end{itemize}
+   \item<3-> Simple memory model {\small
     \begin{lstlisting}
@@
 
-  *<4-> Using state-error monad
+   \item<4-> Using state-error monad
     \begin{lstlisting}
@@
 
-}
-
-#! Shallow Embedding Example
+  \end{itemize}
+\end{frame}
+}
+
+\begin{frame}[fragile]{Shallow Embedding Example}
   \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
@@
 
-#! Code Generation
+\end{frame}
+\begin{frame}[fragile]{Code Generation}
   \uncover<2->{compiling control flow + pretty printing}
@@
 
-#! Preprocessor
-  * Only restricted terms accepted by code generator
-    * good to keep code generation simple
-    * tedious to write manually
-  * Preprocessor transforms terms into restricted format
-    * proves equality (via Isabelle kernel)
-  * Motto: Keep TCB small, preprocessor makes it usable
-
-#! Example: Preprocessing Euclid's Algorithm
+\end{frame}
+\begin{frame}[fragile]{Preprocessor}
+  \begin{itemize}
+   \item Only restricted terms accepted by code generator
+    \begin{itemize}
+     \item good to keep code generation simple
+     \item tedious to write manually
+    \end{itemize}
+   \item Preprocessor transforms terms into restricted format
+    \begin{itemize}
+     \item proves equality (via Isabelle kernel)
+    \end{itemize}
+   \item Motto: Keep TCB small, preprocessor makes it usable
+
+  \end{itemize}
+\end{frame}
+\begin{frame}[fragile]{Example: Preprocessing Euclid's Algorithm}
   \small
@@
 
-#! Reasoning about LLVM Programs
-  *<+-> Separation Logic
+\end{frame}
+\begin{frame}[fragile]{Reasoning about LLVM Programs}
+  \begin{itemize}
+   \item<+-> Separation Logic
     \begin{lstlisting}
@@
 
-    * defined wrt.\ shallowly embedded semantics
-    * proof rules are proved theorems!
-
-  *<+-> Automation: VCG, frame inference, heuristics to discharge VCs
-    * these prove theorems!
-
-  *<+-> Basic Data Structures: signed/unsigned integers, Booleans, arrays
-
-# Sepref
-  * Semi-automatic translation of functional to imperative program
-  * Data refinement to imperative DS
-    * e.g. list to array
-  * Proves refinement theorem
-
-#! Example: Binary Search
+    \begin{itemize}
+     \item defined wrt.\ shallowly embedded semantics
+     \item proof rules are proved theorems!
+
+    \end{itemize}
+   \item<+-> Automation: VCG, frame inference, heuristics to discharge VCs
+    \begin{itemize}
+     \item these prove theorems!
+
+    \end{itemize}
+   \item<+-> Basic Data Structures: signed/unsigned integers, Booleans, arrays
+
+  \end{itemize}
+\end{frame}
+\begin{frame}{Sepref}
+  \begin{itemize}
+   \item Semi-automatic translation of functional to imperative program
+   \item Data refinement to imperative DS
+    \begin{itemize}
+     \item e.g. list to array
+    \end{itemize}
+   \item Proves refinement theorem
+
+  \end{itemize}
+\end{frame}
+\begin{frame}[fragile]{Example: Binary Search}
   \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
@@
 
-#! Example: Binary Search --- Refinement
+\end{frame}
+\begin{frame}[fragile]{Example: Binary Search --- Refinement}
   \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
@@
 
-#! Example: Binary Search --- Generated Code
+\end{frame}
+\begin{frame}[fragile]{Example: Binary Search --- Generated Code}
   \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
@@
 
-#![t] Isabelle-LLVM Parallel
+\begin{frame}[t,fragile]{Isabelle-LLVM Parallel}
   \newcommand{\mknd}[1]{\makebox[0pt]{\tikz[remember picture]{\node (#1) {};}}}
 
-  *<2-> Add parallel operator: $c_1 \parallel c_2$
-    *<3-> execute in parallel, fail on data race
-  *<4-> Shallow embedding: make program report memory accesses
+  \begin{itemize}
+   \item<2-> Add parallel operator: $c_1 \parallel c_2$
+    \begin{itemize}
+     \item<3-> execute in parallel, fail on data race
+    \end{itemize}
+   \item<4-> Shallow embedding: make program report memory accesses
 
@@
 
-}
-
-#! Memory Allocation
-  *<+-> Currently: deterministic semantics
-    * $c_1~\mu$ and $c_2~\mu$ will allocate same memory
-    * cannot be combined!
-
-  *<+-> Use nondeterminism
+  \end{itemize}
+\end{frame}
+}
+
+\begin{frame}[fragile]{Memory Allocation}
+  \begin{itemize}
+   \item<+-> Currently: deterministic semantics
+    \begin{itemize}
+     \item $c_1~\mu$ and $c_2~\mu$ will allocate same memory
+     \item cannot be combined!
+
+    \end{itemize}
+   \item<+-> Use nondeterminism
     \begin{lstlisting}
@@
     \end{lstlisting}
-    * malloc nondeterministically allocates some free address
-    * on combination: exclude infeasible possibilities
-
+    \begin{itemize}
+     \item malloc nondeterministically allocates some free address
+     \item on combination: exclude infeasible possibilities
+
+    \end{itemize}
+  \end{itemize}
   \onslide<+->
@@
 
-#! Invariants
-  * We prove for \is{llM} (enforced by subtype)
-    * access reports are consistent with observed changes in memory
-    * there is at least one possible result (no magic happens)
-
-  * Sanity check for semantics
-
-  * Allows us to prove symmetry of $\parallel$
+\end{frame}
+\begin{frame}[fragile]{Invariants}
+  \begin{itemize}
+   \item We prove for \is{llM} (enforced by subtype)
+    \begin{itemize}
+     \item access reports are consistent with observed changes in memory
+     \item there is at least one possible result (no magic happens)
+
+    \end{itemize}
+   \item Sanity check for semantics
+
+   \item Allows us to prove symmetry of $\parallel$
     \begin{lstlisting}
@@
 
-#! Code generator
-  *<+-> We add \is{llc_par f_1 f_2 x_1 x_2 == f_1 x_1 || f_2 x_2}
-    * \is{f_1, f_2} must be functions
-  *<+-> Code generator generates
-    * type casting boilerplate
-    * call to external \is{parallel} function
-
+  \end{itemize}
+\end{frame}
+\begin{frame}[fragile]{Code generator}
+  \begin{itemize}
+   \item<+-> We add \is{llc_par f_1 f_2 x_1 x_2 == f_1 x_1 || f_2 x_2}
+    \begin{itemize}
+     \item \is{f_1, f_2} must be functions
+    \end{itemize}
+   \item<+-> Code generator generates
+    \begin{itemize}
+     \item type casting boilerplate
+     \item call to external \is{parallel} function
+
+    \end{itemize}
     {\small
@@
     }
-  *<+-> For example, implemented using TBB:
+   \item<+-> For example, implemented using TBB:
     {\small
@@
 
-#! Amending higher layers of IRF
-  *<+-> Prove concurrency rule
+  \end{itemize}
+\end{frame}
+\begin{frame}[fragile]{Amending higher layers of IRF}
+  \begin{itemize}
+   \item<+-> Prove concurrency rule
     \begin{lstlisting}
@@
 
-  *<+-> Sepref refines sequential to parallel execution
+   \item<+-> Sepref refines sequential to parallel execution
     \begin{lstlisting}
@@
 
-  *<+-> Backwards compatible with sequential Sepref!
-    * Easy porting of existing algorithms
-
-#! Parallel Quicksort (basic)
+   \item<+-> Backwards compatible with sequential Sepref!
+    \begin{itemize}
+     \item Easy porting of existing algorithms
+
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+\begin{frame}[fragile]{Parallel Quicksort (basic)}
 
@@
 
-#! Parallel Quicksort (refined)
+\end{frame}
+\begin{frame}[fragile]{Parallel Quicksort (refined)}
   \small
@@
 
-#![t] Parallel Quicksort (Sepref + Code Export)
+\end{frame}
+\begin{frame}[t,fragile]{Parallel Quicksort (Sepref + Code Export)}
   \small
 
-  *<1-> Sepref generates imperative program
-    * using existing sequential \is{pdqsort} for fallback
-    * using (new) sampling partitioner (proved correct + refined separately)
-
-  *<2-> Correctness theorem:
+  \begin{itemize}
+   \item<1-> Sepref generates imperative program
+    \begin{itemize}
+     \item using existing sequential \is{pdqsort} for fallback
+     \item using (new) sampling partitioner (proved correct + refined separately)
+
+    \end{itemize}
+   \item<2-> Correctness theorem:
     {\only<3->{\tiny}
@@
     \end{lstlisting}}
-  *<3-> Instantiation to concrete weak ordering + code export
+   \item<3-> Instantiation to concrete weak ordering + code export
     {\only<4->{\tiny}
@@
 
-  *<4-> Link against C++ benchmark driver
+   \item<4-> Link against C++ benchmark driver
     \begin{lstlisting}
@@
 
+  \end{itemize}
+\end{frame}
 \tikzset{myaxis/.style={
@@
 
-# Benchmarks
+\begin{frame}{Benchmarks}
 
@@
 
-# Benchmarks
+\end{frame}
+\begin{frame}{Benchmarks}
 
@@
 
-# Speedup
+\end{frame}
+\begin{frame}{Speedup}
   \begin{tikzpicture}
@@
 
-# Conclusion
-  * Verification of parallel programs
-    * stepwise refinement to tackle complexity
-    * down to LLVM, small TCB
-    * \high{fast} verified programs
-
-  * Idea: shallow embedding, using access reports
-    * backwards compatible with sequential IRF
-
-  * Future work
-    * state-of-the-art parallel sorting
-    * fractional separation logic
-    * more concurrency
-    * complexity of parallel algorithms
-    * GP-GPUs
-
+\begin{frame}{Conclusion}
+  \begin{itemize}
+   \item Verification of parallel programs
+    \begin{itemize}
+     \item stepwise refinement to tackle complexity
+     \item down to LLVM, small TCB
+     \item \high{fast} verified programs
+
+    \end{itemize}
+   \item Idea: shallow embedding, using access reports
+    \begin{itemize}
+     \item backwards compatible with sequential IRF
+
+    \end{itemize}
+   \item Future work
+    \begin{itemize}
+     \item state-of-the-art parallel sorting
+     \item fractional separation logic
+     \item more concurrency
+     \item complexity of parallel algorithms
+     \item GP-GPUs
+
+    \end{itemize}
+  \end{itemize}
+\end{frame}
 }

### 2022-03-08 papers/llvm_par/main.tex
new stuff

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

%
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\bibliographystyle{plainurl}% the mandatory bibstyle

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{balance}

\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}

%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
\usepackage{mathpartir}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{cprotect}

\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

\usepackage{multirow}
\usepackage{siunitx}

\usepackage{tikz}
\usetikzlibrary{positioning,fit,shadows,decorations, arrows, shapes, decorations.markings,decorations.pathmorphing,calc}

\usepackage{pgfplots}

\tikzset{
  -|-/.style={
    to path={
      (\tikztostart) -| ($(\tikztostart)!#1!(\tikztotarget)$) |- (\tikztotarget)
      \tikztonodes
    }
  },
  -|-/.default=0.5,
  |-|/.style={
    to path={
      (\tikztostart) |- ($(\tikztostart)!#1!(\tikztotarget)$) -| (\tikztotarget)
      \tikztonodes
    }
  },
  |-|/.default=0.5,
}

\clubpenalty = 10000
\widowpenalty = 10000
\displaywidowpenalty = 10000

\input{lstisabelle}
\lstset{captionpos=b}
\lstset{numberbychapter=false}
\lstset{autogobble}

\newcolumntype{Y}{>{\RaggedRight\arraybackslash}X} % Multiline column, automatic width

  \newcommand{\false}{\textrm{false}}
  \newcommand{\true}{\textrm{true}}

% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]

% \overfullrule=8pt

\title{Refinement of Parallel Algorithms down to LLVM}

% \author{Peter Lammich}
% \institute{University of Twente}
% \author{A.~Nonymous}
% \institute{Institute}
% \email{p.lammich@utwente.nl}
%
% \authorinfo{Peter Lammich}
%            {The University of Manchester, UK}
%            {peter.lammich@machester.ac.uk}

\author{Peter Lammich}{University of Twente, Netherlands}{p.lammich@utwente.nl}{https://orcid.org/0000-0003-3576-0504}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\authorrunning{Peter Lammich} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Peter Lammich} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[500]{Software and its engineering~Formal software verification}
\ccsdesc[500]{Theory of computation~Semantics and reasoning}
\ccsdesc[300]{Computing methodologies~Parallel algorithms}

% \ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Isabelle,Concurrent Separation Logic,Parallel Sorting,LLVM} %TODO mandatory; please add comma-separated list of keywords

% \acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
  We present a stepwise refinement approach to develop
  verified parallel algorithms, down to efficient LLVM code.
  The resulting algorithms' performance is competitive with their
  counterparts implemented in C/C++.
%
  Our approach is backwards compatible with the Isabelle Refinement Framework,
  such that existing sequential formalizations can easily be adapted
  or re-used.
%
  As case study, we verify a parallel quicksort algorithm, and
  show that it performs on par with its C++ implementation,
  and is competitive to state-of-the-art parallel sorting algorithms.
\end{abstract}

\section{Introduction}
% Modern hardware has become faster by getting more parallel, while the
% sequential execution speed has not significantly increased in the last years.
% To keep up with this development, and fully utilize the capabilities of modern hardware,
% parallel algorithms are required.
%
% However, parallel algorithms come with new subtle classes of errors, like race conditions,
% which are particularly hard to find by testing. Thus, formal methods that can guarantee
% absence of all errors, including the hard-to-find ones, become even more important for
% parallel algorithms.

We present a stepwise refinement approach to develop
verified and efficient parallel algorithms. Our method can verify total correctness down to LLVM intermediate code.
The resulting verified implementations are competitive with state-of-the-art
unverified implementations. Our approach is backwards compatible to
the Isabelle Refinement Framework (IRF), a powerful tool to verify efficient sequential software,
such as model checkers~\cite{ELNN13,BrLa18,WiLa18}, SAT solvers~\cite{La17_CADE,La17_SAT,FBL18},
or graph algorithms~\cite{La14,LaSe16,LaSe19}.
This paper adds parallel execution to the IRF's toolbox, without invalidating the existing formalizations,
which can now be used as sequential building blocks for parallel algorithms, or be modified to add
parallelization.

As a case study, we verify total correctness of a parallel quicksort algorithm,
re-using an existing verification of state-of-the-art sequential sorting algorithms~\cite{La20}.
Our verified parallel sorting algorithm is competitive to state-of-the-art parallel sorting algorithms.

% The main idea is to add a parallel combinator to an existing
% sequential LLVM semantics, by extending the thread semantics to record memory accesses, and
% check these for data-race freedom upon joining the parallel threads.
% Based on this idea, we extend the existing Isabelle Refinement Framework~\cite{La19-llvm}
% to support parallel algorithms. The resulting framework is largely backwards compatible to
% the original one, such that we could easily port existing algorithms and data structures,
% and use them as building blocks for parallel algorithms.

% The contributions of this paper are as follows:
% \begin{itemize}
%   \item the (up to our knowledge) first verification method for parallel LLVM programs.
%
%   \item a stepwise refinement approach to verify total correctness of parallel programs.
%
%   \item the first verified implementation of a parallel sorting algorithm.
% \end{itemize}

\noindent Formalization available at: \url{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}

\subsection{Overview}\label{sec:intro_overview}
This paper is based on the Isabelle Refinement Framework (IRF), a continuing effort
to verify efficient implementations of complex algorithms, using stepwise refinement techniques.
Figure~\ref{fig:IRF-overview} displays the components of the Isabelle Refinement Framework.

\begin{figure}
  \definecolor{colA}{RGB}{11, 97,164}
  \definecolor{colB}{RGB}{204,246,  0}
  \definecolor{colC}{RGB}{166,  0,166}
  \definecolor{colD}{RGB}{255,146,  0}
  \definecolor{colTCB}{RGB}{255, 0, 0}

  \center
  \begin{tikzpicture}[node distance=.01\textwidth,outer sep=0pt,
  basicLayer/.style={
    draw=black,
    rounded corners,
    text width=.9\textwidth,
    font={\sffamily\bfseries\color{black}},
    align=center,
    minimum height=.03\textheight,
%    text height=12pt,
%    text depth=9pt
    },
  layerA/.style={basicLayer,fill=colA!30},
  layerB/.style={basicLayer,fill=colB!30},
  layerC/.style={basicLayer,fill=colC!30},
  layerD/.style={basicLayer,fill=colD!30},
  layerDD/.style={basicLayer,fill=colD!40,text width=.2\textwidth},
  layerTCB/.style={basicLayer,fill=colTCB!30,text width=.2\textwidth},
  layerTCBB/.style={basicLayer,fill=colTCB!40,text width=.2\textwidth},
  ]

  \node[layerA] (AlgDS) {Algorithms and Data Structures};

  \node[layerB, below=of AlgDS.south west, anchor=north west, text width=.437\textwidth] (ne) {NE-Monad};
  \node[layerA, right=of ne.north east, anchor=north west, text width=.437\textwidth, minimum height=.067\textheight] (llDS) {Low-Level Algorithms \\and Data Structures};

  \node[layerB, below=of ne.south west, anchor=north west, text width=.437\textwidth] (sepref) {Sepref};

  \node[layerC, below=of sepref.south west, anchor=north west] (sepl) {Program Logic and VCG};
  \node[layerD, below=of sepl.south west, anchor=north west, minimum height=.13\textheight] (sembox) {};

  \node[layerTCB, below left=of sembox.north east, minimum height=.12\textheight, text width=.537\textwidth] (tcbbox) {};

  \node[right=of sembox.west,text width=.2\textwidth,font={\sffamily\bfseries\color{black}}] {Back\\ End};

  \node[above left=of tcbbox.south east,font={\sffamily\bfseries\color{black}}] {Trusted Code Base};

%   \node[layerD, right=of sembox.west] (preproc) {Preprocessor};
%   \node[layerD, right=1cm of preproc] (instr) {Instructions};
%   \node[layerD, right=1cm of instr] (codegen) {Code Generator};

  \node[layerTCBB, below left=of tcbbox.north east] (codegen) {Code Generator};
  \node[layerTCBB, left=1cm of codegen] (instr) {Instructions};
  \node[layerDD, left=1cm of instr] (preproc) {Preprocessor};

  \node[layerTCBB, below=of instr] (mmodel) {Memory Model};
  \node[layerTCBB, below=of mmodel] (mmonad) {SE-Monad};

  \draw[->] (preproc) -- (instr);
  \draw[->] (instr) -- (codegen);

  \end{tikzpicture}

  \caption{Components of the Isabelle Refinement Framework, with focus on the back end.}\label{fig:IRF-overview}
\end{figure}

%   \begin{tikzpicture}[node distance=.01\textwidth,outer sep=0pt,
%   layerA/.style={
%     draw=black,
%     fill=blue!20,
%     rounded corners,
%     text width=.9\textwidth,
%     font={\sffamily\bfseries\color{black}},
%     align=center,
%     minimum height=.03\textheight,
% %    text height=12pt,
% %    text depth=9pt
%     },
%   layerB/.style={layerA,fill=green!20},
%   layerC/.style={layerA,fill=red!20},
%   layerD/.style={layerA,fill=red!30,text width=.2\textwidth},
%   ]
%
%   \node[layerA] (AlgDS) {Algorithms and Data Structures};
%
%   \node[layerB, below=of AlgDS.south west, anchor=north west, text width=.437\textwidth] (ne) {Nres-Monad};
%   \node[layerA, right=of ne.north east, anchor=north west, text width=.437\textwidth, minimum height=.067\textheight] (llDS) {Low-Level Algorithms \\and Data Structures};
%
%   \node[layerB, below=of ne.south west, anchor=north west, text width=.437\textwidth] (sepref) {Sepref};
%
%   \node[layerC, below=of sepref.south west, anchor=north west] (sepl) {Program Logic and VCG};
%   \node[layerC, below=of sepl.south west, anchor=north west, minimum height=.120\textheight] (sembox) {};
%
%   \node[below right=of sembox.north west] {Back End Semantics};
%
%
%   \node[layerD, right=of sembox.west] (isacode) {Isa Codegen};
%   \node[layerD, right=of isacode] (imphol) {Imp HOL};
%   \node[layerD, right=of imphol] (llvm) {LLVM};
%   \node[layerD, right=of llvm] (llvm-par) {LLVM par};
%
%   \node[layerD, below=of imphol] (impholtime) {Imp HOL time};
%   \node[layerD, below=of llvm] (llvmtime) {LLVM time};
%
%
%   \end{tikzpicture}

The back end layer handles the translation from Isabelle/HOL to the actual target language.
The instructions of the target language are shallowly embedded into Isabelle/HOL, using
a state-error (SE) monad. An instruction with undefined behaviour,
or behaviour outside our supported fragment, raises an error.
The state of the monad is the memory, represented via a memory model.
The code generator translates the instructions to actual code.
These components form the trusted code base, while all the remaining components of the
Isabelle Refinement Framework generate proofs. In the back-end, the preprocessor transforms
expressions to the syntactically restricted format required by the code generator,
proving semantic equality of the original and transformed expression.
While there exist back ends for purely functional code~\cite{LaTu12,La13},
and sequential imperative code~\cite{La15,La19-llvm}, this paper describes a back end
for parallel imperative LLVM code (Section~\ref{sec:parsem}).

On top of the back-end, a program logic is used to prove programs correct.
It uses separation logic, and provides automation like a verification condition generator (VCG).
In Section~\ref{sec:seplogic}, we describe our formalization of concurrent separation logic~\cite{OH04}, and our VCG.

This basic program logic and VCG can already be used to verify simple low-level algorithms and data structures,
like dynamic arrays and linked lists. More complex developments typically use a stepwise refinement approach,
starting at purely functional programs modelled in a nondeterminism-error (NE) monad~\cite{LaTu12}.
A semi-automatic refinement procedure (Sepref~\cite{La15,La19-llvm}) translates from the purely functional
code to imperative code, refining abstract functional data types to concrete imperative ones.
In Section~\ref{sec:refine_parallel}, we describe our extensions to support refinement to parallel executions,
and a fine-grained tracking of pointer equalities, required to parallelize computations that work on
disjoint parts of the same array.

Using our approach, complex algorithms and data structures can be developed and
refined to optimized efficient code. The stepwise refinement ensures a separation of concerns
between high-level algorithmic ideas and low-level optimizations. We have used this approach to
verify a wide range of practically efficient algorithms~\cite{ELNN13,BrLa18,WiLa18,La17_CADE,La17_SAT,FBL18,La14,LaSe16,LaSe19,La20}.
In Section~\ref{sec:parsort}, we use our techniques to verify a parallel sorting algorithm,
with competitive performance wrt.\ unverified state-of-the-art algorithms.

Section~\ref{sec:concl} concludes the paper and discusses related and future work.

%
% While the ne-monad and associated tooling has largely stayed the same since it's first version~\cite{LaTu12},
% the Sepref tool has been continuously adapted and improved with each new back end that was added.
% Again, our work is based on the Sepref tool for LLVM without time~\cite{La19-llvm}.
%
%
% While the Imperative HOL versions of the IRF are based on a separation logic formalization
% by Lammich and Meis~\cite{LaMe12}, the LLVM versions are based on a more general framework by Klein et al.~\cite{KKB12,KKB12_afp}.
%
% While this reasoning infrastructure can already be used to verify simple low-level algorithms and data structures,
% more complex developments are typically refined from a purely functional,
% monadic representation of nondeterministic programs (ne-monad), originally introduced in~\cite{LaTu12}.
% A typical development will do multiple refinement steps inside the ne-monad, and then use a
% semi-automatic tool (Sepref) to refine from the ne-monad into the back end semantics.
% While the ne-monad and associated tooling has largely stayed the same since it's first version~\cite{LaTu12},
% the Sepref tool has been continuously adapted and improved with each new back end that was added.
% Again, our work is based on the Sepref tool for LLVM without time~\cite{La19-llvm}.
%
% We have also implemented parallelism for the Imperative HOL back end.
% This was done as a feasibility study, and uses the exactly same ideas and
% techniques as described in this paper for the more efficient LLVM back end.
% While a description of this initial implementation would not add any new
% ideas to this paper, it might still be useful in the context of older developments,
% that have not yet been ported to the LLVM back end.
%
% The rest of this paper is roughly organized along the structure of the Isabelle Refinement Framework:
% We first describe our parallel LLVM semantics (Section~\ref{sec:parsem}) and separation logic~\ref{sec:seplogic}.
% Then, we describe our changes to the refinement infrastructure (Section~\ref{sec:refine_parallel}).
% Section~\ref{sec:parsort} describes our parallel sorting algorithm case study and benchmark results,
% and Section~\ref{sec:concl} concludes the paper and discusses related and future work.
%

%
%
%
% Contributions
%   shallow embedding of parallel semantics
%
%   technical: formalization of parallel separation logic. TOTAL CORRECTNESS.
%   integration of parallel call into refinement framework. backwards compatible (re-use of seq).
%
%   technical: automatic refinement with concrete-identity tracking
%
%   formalization of simple parallel quicksort, using state-of-the-art sequential sort as subroutine
%
%   sample partitioning scheme, template instantiation by preprocessor

\section{A Back End for LLVM with Parallel Execution}\label{sec:parsem}
  We formalize a semantics for parallel execution, shallowly embedded into Isabelle/HOL.
  As for the existing sequential back ends~\cite{La15,La19-llvm}, the shallow embedding is key
  to the flexibility and feasibility of the approach.
  The main idea is to make an execution report the memory that it accesses,
  and use this information to raise an error when joining executions that
  would have exhibited a data race. We use this to model an instruction that
  calls two functions in parallel, and waits until both have returned.

  \subsection{State-Nondeterminism-Error Monad with Access Reports}\label{sec:monad}
  We define the underlying monad in two steps. We start with a nondeterminism-error monad,
  and then lift it to a state monad and add access reports.
  Defining a nondeterminism-error monad is standard:
  \begin{lstlisting}
    'a neM == spec ("'a => bool") | fail
    return x == spec (%r. r=x)
    bind fail f == fail
    bind (spec P) f == if ?x. P x & f x = fail then fail
                      else spec (%r. ?x Q. P x & f x = spec Q & Q r)
  \end{lstlisting}
  A program either fails, or yields a possible set of results (\is{spec P}), described by its characteristic function \is{P}.
  The \is{return} operation yields exactly one result, and \is{bind} combines all possible results,
  failing if there is a possibility to fail.

  Now assume that we have a state (memory) type \is{'\mu}, and an access report type \is{'\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == '\mu => ('x \x '\rho \x '\mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, and adds up the access reports.

  Typically, the access report will contain read and written addresses, such that we can detect a data race.
  Moreover, if parallel executions can allocate memory, we must detect those
  executions where the memory manager allocated the same block in both parallel strands.
  As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
  Let \is{norace :: '\rho => '\rho => bool} and \is{feasible :: '\rho => '\rho => bool} be symmetric predicates,
  and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
  compose two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
  \begin{lstlisting}
    (m_1 || m_2) \mu ==
      (x_1,(r_1,\mu_1)) <- m_1 \mu; (x_2,(r_2,\mu_2)) <- m_2 \mu;  --"\rightcomment{execute both strands}"
      assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
      assert norace \rho_1 \rho_2; --"\rightcomment{fail on data race}"
      return$_{ne}$ ((x_1,x_2), combine (\rho_1,\mu_1) (\rho_2,\mu_2))  --"\rightcomment{combine results}"

    assume P == if P then return () else spec (%_. False)
    assert P == if P then return () else fail
  \end{lstlisting}
  Here, we use \is{assume} to ignore infeasible executions, and \is{assert} to fail on data races.
  Note that, if one parallel strand fails, and the other parallel strand has no possible
  results \is{spec (\<lambda>_. False)}, the behaviour of the parallel composition is not clear.
  For this reason, we fix an invariant \is{invar$_M$ :: ('\mu => ('x \x '\rho \x '\mu) neM) => bool},
  which implies that every non-failing execution has at least one possible result.
  We define the actual type \is{M} as the subtype satisfying \is{invar$_M$}.
  Thus, we have to prove that every combinator and instruction of our semantics
  preserves the invariant, which is an important sanity check.
  As additional sanity check, we prove symmetry of parallel composition:
  \begin{lstlisting}
    m_1 || m_2 = mswap (m_2 || m_1)     $\text{where}$    mswap m == (x_1,x_2)<-m; return (x_2,x_1)
  \end{lstlisting}
%   where \is|mswap m = (x_1,x_2)<-m; return (x_2,x_1)| swaps the result tuple.

  \subsection{Memory Model}
  Our memory model supports blocks of values, where values can be integers, structures, or pointers into a block:
  \begin{lstlisting}
    datatype addr == ADDR (bidx: nat) (idx: nat)
    datatype ptr == PTR_NULL   |   PTR_ADDR (the_addr: addr)
    datatype val == LL_INT lint   |   LL_STRUCT val list   |   LL_PTR ptr

    datatype block == FRESH   |   FREED   |   is_alloc: ALLOC (vals: "val list")
    typedef memory == { \mu :: nat => block. finite {b. \mu b ~= FRESH} }
  \end{lstlisting}
  A block is either fresh, freed, or allocated, and a memory is a mapping from block indexes to blocks,
  such that only finitely many blocks are not fresh. Every block's state transitions from fresh to
  allocated to freed. This avoids ever reusing the same block, and thus allows us to semantically detect use after free errors.
  Every program execution can only allocate finitely many blocks,
  such that we will never run out of fresh blocks\footnote{If the actual system does run out of memory, we will terminate the program in a defined way.}.
  An allocated block contains an array of values, modelled as a list. Thus, an address consists of a block number, and an index into the array.

  To access and modify memory, we define the functions \is{valid}, \is{get}, and \is{put}:
  \begin{lstlisting}
    valid \mu (ADDR b i) == is_alloc (\mu b) & i<|vals (\mu b)|
    get \mu (ADDR b i) == vals (\mu b) ! i
    put \mu (ADDR b i) x == \mu(b := BLOCK ((vals (\mu b))[i:=x]))
  \end{lstlisting}
  where \is{|xs|} is the length of list \is{xs}, \is{xs!i} returns the \is$i$th element of list \is$xs$,
  and \is{xs[i:=x]} replaces the \is$i$th element of \is{xs} by \is{x}.

  Note that our LLVM semantics does not support conversion of pointers to integers, nor comparison
  or difference of pointers to different blocks. This way, a program cannot see the internal representation of a pointer,
  and we can choose a simple abstract representation, while being
  faithful wrt.\ any actual memory manager implementation.

  \subsection{Access Reports}
  We now fix the state of the M-monad to be memory, and the access reports to be sets
  of read and written addresses, as well as sets of allocated and freed blocks:
  \begin{lstlisting}
    acc == ( r :: addr set; w :: addr set; a :: nat set; f :: nat set )
    0 == ( {},{},{},{} )
    (r_1,w_1,a_1,f_1) + (r_2,w_2,a_2,f_2) == ( r_1\union r_2, w_1\union w_2, a_1\union a_2, f_1\union f_2 )
  \end{lstlisting}

  Two parallel executions are feasible if they did not allocate the same block, and they have a data race
  if one strand accesses addresses or blocks modified by the other strand:
  \begin{lstlisting}
    feasible (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) == a_1 \inter a_2 = {}

    norace (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) ==
      let m_1 = w_1 \<union> { ADDR b i. b \<in> a_1 \<union> f_1 } in
      let m_2 = w_2 \<union> { ADDR b i. b \<in> a_2 \<union> f_2 } in
        (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
  \end{lstlisting}
  The invariant for \is{M} states that blocks transition only from fresh to allocated to free,
  allocated blocks never change their size, and the access report matches the observable state change (\is{consistent}).
  It also states, that for each finite set of blocks $B$, there is an execution that does not allocate blocks from $B$.
  The latter is required to show that we always find feasible parallel executions:
  \begin{lstlisting}
    invar$_M$ c == \<forall>\mu P. c \mu = spec P ==>
      (\<forall>x \rho \mu'. P (x,\rho,\mu') ==> consistent \mu \rho \mu')
    & (\<forall>B. finite B ==> (\<exists>x \rho \mu'. P (x,\rho,\mu') & \rho.a \inter B = {} ))
  \end{lstlisting}
  The combine function joins the access reports and memories,
  preferring allocated over fresh, and freed over allocated memory. When joining two allocated blocks, the written addresses
  from the access report are used to join the blocks. We skip the rather technical definition of combine, and just state the
  relevant properties: Let \is{\rho_1=(r_1,w_1,a_1,f_1)} and \is{\rho_2=(r_2,w_2,a_2,f_2)} be feasible and race free access reports,
  and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}, and \is{addr} a valid address in \is{\mu'}. Then
  \begin{lstlisting}
    \mu' b = FRESH <--> if b \in a_2 \union f_2 then \mu_2 b = FRESH else \mu_1 b = FRESH
    \mu' b = FREED <--> if b \in a_2 \union f_2 then \mu_2 b = FREED else \mu_1 b = FREED
    is_alloc (\mu' b) <--> if b \in b \in a_2 then is_alloc (\mu_2 b) else is_alloc (\mu_1 b)
    get \mu' addr = if addr \in w_2 \or addr.bidx \in a_2 \union f_2 then get \mu_2 addr else get \mu_1 addr
  \end{lstlisting}

  \subsection{LLVM Instructions}
  Based on the M-monad, we define shallowly embedded LLVM instructions. For most instructions,
  this is analogous to the sequential case~\cite{La19-llvm}.
  The exceptions are memory allocation, which nondeterministically allocates some available
  block (the original formalization deterministically counted up the block indexes), and an instruction for parallel
  function call:
  \begin{lstlisting}
    llc_par f g a b == f a || g b
  \end{lstlisting}
  The code generator only accepts this, if \is{f} and \is{g} are constants (i.e., function names).
  It then generates some type-casting boilerplate, and a call to an external
  \is{parallel} function, which we implement using the Threading Building Blocks~\cite{TBB} library:
  \begin{lstlisting}[language=C++]
  void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2) {
    tbb::parallel_invoke([=]{f1(x1);}, [=]{f2(x2);}); }
  \end{lstlisting}
  I.e., the two functions \is{f1(x1)} and \is{f2(x2)} are called in parallel. The generated boilerplate code sets up \is{x1} and \is{x2} to point to both, the actual arguments and space for the results.

%
%
%
%
%
%
%
%
%
%
%   \subsection{Executions}\label{sec:parsem:execs}
%   To model executions, we use a state-nondeterminism-error monad, where each execution reports the memory that it has accessed.
%   This information is used to make a parallel execution fail in case of data races.
%   We start with a standard nondeterminism-error (ne) monad, where an execution either fails, or returns a set of possible results, specified by its characteristic function.
%   \begin{lstlisting}
%     'a neM == spec ("'a => bool") | fail
%     return x == spec (%r. r=x)
%     bind fail f == fail
%     bind (spec P) f == if ?x. P x & f x = fail then fail
%                       else spec (%r. ?x Q. P x & f x = spec Q & Q r)
%   \end{lstlisting}
%   We use the shortcut notation \is{x<-m; f x} for \is|bind m (\<lambda>x. f x)|.
%
%   In the \is{ne}-monad, we can define the functions for memory management. We define functions for memory allocation,
%   freeing, loading, and storing. In addition to returning a result and a new memory, these functions also return
%   information which memory was accessed, i.e., the read and written addresses, as well as the allocated and freed blocks:
%   \begin{lstlisting}
%     macc == { r: addr set; w: addr set; a: nat set; f: nat set }
%   \end{lstlisting}
%   Memory accesses form a commutative monoid where \is{0} means no accesses, and \is{(+)} joins the accesses.
%   \begin{lstlisting}
%     0 == { r={},w={},a={},f={} }
%     {r_1,w_1,a_1,f_1} + {r_2,w_2,a_2,f_2} == { r=r_1\union r_2, w=w_1\union w_2, a=a_1\union a_2, f=f_1\union f_2 }
%   \end{lstlisting}
%
%   We display the definitions of memory allocation and storing. Freeing and loading are defined analogously:
%   \begin{lstlisting}
%     alloc :: val list => memory => (nat \x macc \x memory)
%     alloc vs \mu ==
%       b <- spec b. \mu b = FRESH;  --"\rightcomment{Select some fresh block}"
%       return (b, macc (r={},w={},a={b},f={}), \mu(b:=ALLOC vs))
%
%     store :: addr => val => memory => (unit \x macc \x memory)
%     store addr v \mu ==
%       assert valid \mu addr;
%       return ((), macc (r={},w={addr},a={},f={}), put \mu addr v)
%
%     assert True == return ()     assert False == fail
%   \end{lstlisting}
%   The \is{alloc} function nondeterministically selects a fresh block of memory, and initializes it
%   with the given list \is{vs} of values. Additionally, it reports an allocation access of the allocated block.
%   The \is{store} function checks if the given address is valid. Then, it updates the memory and reports a write access.
%   Here, \is{assert} fails if the argument is false.
%
%   Next, we lift these definitions to a monad of type \is{'a M = memory => ('a \x macc \x memory)}.
%   The return function does not modify the state and reports no memory accesses, and bind sequentially composes
%   the state change, and joins the reported memory accesses:
%   \begin{lstlisting}
%     return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
%
%     bind$_M$ m f \mu == (x,acc_1,\mu) <- m \mu; (r,acc_2,\mu) <- f x \mu; return$_{ne}$ (r,acc_1+acc_2,\mu)
%   \end{lstlisting}
%   note that we use subscripts ${}_{M}$ and ${}_{ne}$ to disambiguate the monad operations when not clear from context.
%
%   Moreover, we define parallel execution:
%   \begin{lstlisting}
%     (m_1 || m_2) \mu ==
%       (x_1,acc_1,\mu_1) <- m_1 \mu; (x_2,acc_2,\mu_2) <- m_2 \mu;  --"\rightcomment{execute both strands}"
%       assume acc_1.a \inter acc_2.a = {};  --"\rightcomment{Filter out impossible combinations}"
%       assert norace acc_1 acc_2; --"\rightcomment{fail on data race}"
%       return$_{ne}$ ((x_1,x_2),acc_1+acc_2, combine \mu_1 acc_2 \mu_2 )  --"\rightcomment{Combine results}"
%
%     assume P == if P then return () else spec (%_. False)
%   \end{lstlisting}
%   That is, we execute both programs from the initial memory \is{\mu}.
%   Then, we filter out impossible executions, i.e., where the memory manager allocated the
%   same block in both branches. For the remaining executions, we fail if there is a data race, i.e.,
%   if the accessed addresses and blocks of one strand interfere with the modified addresses and blocks
%   of the other strand:
%   \begin{lstlisting}
%     intf_norace acc_1 acc_2 ==
%       let m_1 = acc_1.w \<union> { ADDR b i. b \<in> acc_1.a \<union> acc_1.f } in
%       let m_2 = acc_2.w \<union> { ADDR b i. b \<in> acc_2.a \<union> acc_2.f } in
%       (acc_1.r \<union> m_1) \<inter> m_2 = {}  &   m_1 \<inter> (acc_2.r \<union> m_2) = {}
%   \end{lstlisting}
%   Finally, we combine the memories from the two executions, using the reported memory accesses.
%   The \is{combine \mu_1 acc_2 \mu_2} function combines the memories, amending \is{mu_1}
%   with values from \is{mu_2} at blocks and addresses contained in \is{acc_2}.
%   We skip its technical definition, but show the relevant properties instead.
%   Let \is{\mu' = combine \mu_1 acc_2 \mu_2}, \is{af_2=acc_2.a \union acc_2.f}, and \is{a = ADDR b i}, then
%   \begin{lstlisting}
%     \mu' b = FRESH <--> if b \in af_2 then \mu_2 b = FRESH else \mu_1 b = FRESH
%     \mu' b = FREED <--> if b \in af_2 then \mu_2 b = FREED else \mu_1 b = FREED
%     is_alloc (\mu' b) <--> if b \in af_2 then is_alloc (\mu_2 b) else is_alloc (\mu_1 b)
%     valid \mu' a ==> get \mu' a = if a \in acc_2.w \or b \in af_2 then get \mu_2 a else get \mu_1 a
%   \end{lstlisting}
%
%   \subsection{Invariants}
%   Our execution semantics only make sense if the reported memory accesses actually match the actual memory accesses,
%   and all programs fail or have at least on possible result. Otherwise, we might miss data-races, or get an unclear meaning
%   of a failing execution in parallel to one without any results. Note that our code generator will only accept functions that are
%   constructed from the primitive building blocks of our semantics. Thus, we only have to check these in order to believe in the correctness of any verified program.
%   To further reduce this trusted code base, we actually define our monad \is{M} as a subtype. Thus, for every primitive function that we define, we have to actually
%   prove that it satisfies these assumptions:
%   \begin{lstlisting}
%     typedef 'a M = "{\mu:: memory => ('a \x macc \x memory) neM. invarM \mu}"
%   \end{lstlisting}
%   Here, the invariant ensures that the reported memory accesses are consistent with the observed changes in the memory,
%   and that there are no empty result sets.
%
%   The invariant preservation proof is easy for all primitives but parallel execution.
%   Here, we have to show that not all executions are filtered out as impossible, which would yield an empty set of results.
%   For this, we generalize the invariant to include the property that,
%   for every finite set \is{B} of blocks, there is an execution that does not allocate blocks from \is{B}.
%   Thus, we will always find compatible parallel executions.
%
%   As an additional sanity check of our semantics, we prove that the parallel operator is symmetric:
%   \begin{lstlisting}
%     m_1 || m_2 = mswap (m_2 || m_1)
%   \end{lstlisting}
%   where \is|mswap m = (x_1,x_2)<-m; return (x_2,x_1)| swaps the result tuple.
%
%
%   \subsection{Extending the LLVM Backend}
%   We have defined the monad \is{M}, that supports basic primitives for memory access and parallel execution.
%   Using this monad, we define functions for each supported LLVM instruction, as well as functions for memory management,
%   that are provided by a small support library. The code generator ensures that everything is composed of only these functions,
%   and translates them to the corresponding LLVM instructions. For details we refer the reader to~\cite{La19-llvm},
%   and only describe our new primitive for parallel function call:
%   \begin{lstlisting}
%     llc_par f g a b == f a || g b
%   \end{lstlisting}
%   The code generator only accepts this, if \is{f} and \is{g} are constants, rather than more complex terms.
%   It then generates some type-casting boilerplate, and a call to an external
%   \is{parallel} function, which we implement using the Threading Building Blocks~\cite{TBB} library:
%   \begin{lstlisting}[language=C++]
%   void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2) {
%     tbb::parallel_invoke([=]{f1(x1);}, [=]{f2(x2);}); }
%   \end{lstlisting}
%   I.e., the two functions \is{f1(x1)} and \is{f2(x2)} are called in parallel. The generated boilerplate code sets up the
%   arguments \is{x1} and \is{x2} to point to both, the actual arguments and space for the results.
%
%   %TODO Explain how syntactic restrictions work.
%
%
% %   Prepropessor!

  \section{Parallel Separation Logic}\label{sec:seplogic}
  In the previous section, we have defined a shallow embedding of LLVM programs into Isabelle/HOL.
  We now describe how to reason about these programs, using separation logic.

%   We use separation logic in order to reason about programs. As the program itself, also the separation logic is shallowly embedded,
%   i.e., the notion of Hoare-triple is directly defined wrt.\ the semantics, and a Hoare-rule is proved as a lemma.
%   Thus, our Hoare-logic is not part of the trusted code base. In this section, we briefly sketch the separation logic of Isabelle-LLVM~\cite{La19-llvm},
%   and describe how we extend it to prove the standard Hoare-rule for parallel execution~\cite{OH04}, i.e., that two threads that
%   run on disjoint memory can be executed in parallel:
%   \[
%     \frac{\{P_1\} ~c_1~ \{Q_1\}~~~\{P_2\} ~c_2~ \{Q_2\}}{ \{P_1 * P_2\} ~c_1 \parallel c_2~ \{Q_1 * Q_2\} }
%   \]

  \subsection{Separation Algebra}
  In order to reason about memory with separation logic, we define an abstraction function from the memory into a separation algebra~\cite{CHY07}.
  Separation algebras formalize the intuition of combining disjoint parts of memory.
  They come with a \emph{zero} (\is{0}) that describes the empty part, a \emph{disjointness predicate} $a\#b$
  describing that the parts $a$ and $b$ do not overlap, and a \emph{disjoint union} $a+b$ that combines two disjoint parts.
  For the exact definition of a separation algebra, we refer to~\cite{CHY07,KKB12}. We note that
  separation algebras naturally extend over functions and pairs, in a pointwise manner.

%   The separation algebra abstracts away from details of the memory model
%   (like our distinction between fresh and freed blocks), and introduces the necessary
%   formalism to support partial memories. This abstraction layer allows to use two different
%   formalizations of the memory model, a concise and simple \emph{concrete memory model} for the semantics,
%   and an \emph{abstract memory model} that forms a separation algebra.
%   Note that the abstract memory model is only used for reasoning, and does not contribute to the
%   trusted code base of our LLVM semantics.
%
%   In the following, we briefly sketch the concepts of separation algebra and our abstract memory model.
%   For a more detailed explanation, we refer to \cite{CHY07,La19-llvm}.
%
%   A separation algebra~\cite{CHY07} is a structure with a zero, a disjointness predicate $a\#b$,
%   and a disjoint union $a+b$.
%   Intuitively, elements describe parts of the memory.
%   Zero describes the empty memory, $a\#b$ means that $a$ and $b$ describe disjoint parts of the memory,
%   and $a+b$ describes the memory described by the union of $a$ and $b$.
%   For the exact definition of a separation algebra, we refer to~\cite{CHY07,KKB12}. We note that
%   separation algebras naturally extend over functions and pairs, in a pointwise manner.

  \begin{example}(Trivial Separation Algebra)
    The type \is{\alpha option = None | Some \alpha} forms a separation algebra with:
    \begin{lstlisting}
      0 == None     a # b == a=0 \or b=0      a + 0 == a     0 + b == b
    \end{lstlisting}
    Intuitively, this separation algebra does not allow for combination of contents, except if one side is zero.
    While it is not very useful on its own, the trivial separation algebra is a useful building block for
    more complex separation algebras.
  \end{example}

  For our memory model, we define the following abstraction function:
  \begin{lstlisting}
    \alpha :: memory -> (addr -> val option) \x (nat -> nat option)
    \alpha \mu == (\alpha$_m$ \mu, \alpha$_b$ \mu)

    \alpha$_m$ \mu addr == if valid \mu addr then Some (get \mu addr) else 0
    \alpha$_b$ \mu b == if is_alloc (\mu b) then TRIV (|vals (\mu b)|) else 0
  \end{lstlisting}
  An abstract memory \is{\alpha \mu} consists of two parts:
  \is{\alpha$_m$ \mu} is a map from addresses to the values stored there. It is used to reason about load and store operations.
  \is{\alpha$_b$ \mu} is a map from block indexes to the sizes of the corresponding blocks.
    It is used to ensure that one owns all addresses of a block when freeing it.

  We continue to define a separation logic: assertions are predicates over separation algebra elements.
  The basic connectives are defined as follows:
  \begin{lstlisting}
    false a == False     true a == True     \<box> a == a=0
    (P*Q) a == \<exists> a_1 a_2. a_1 # a_2 \and a = a_1 + a_2 \and P a_1 \and Q a_2
  \end{lstlisting}
  That is, the assertion \is{false} never holds and the assertion \is{true} holds for all abstract memories.
  The empty assertion \is{\<box>} holds for the zero memory, and the separating conjunction \is{P*Q}
  holds if the memory can be split into two disjoint parts, such that \is{P} holds for one,
  and \is{Q} holds for the other part. The lifting assertion \is{\<up>\<phi>} holds iff the Boolean value \is{\<phi>} is true:
  \begin{lstlisting}
    \<up>\<phi> == if \<phi> then \<box> else false
  \end{lstlisting}
  It does not depend on the memory, and is used to lift plain logical statements into separation logic.
  When clear from the context, we omit the \is{\<up>}-symbol, and just mix plain statements with separation logic assertions.
%   , e.g., \is{array xs p * i<|xs|}

  \subsection{Weakest Preconditions and Hoare Triples}\label{sec:wp_hoare}
  We define a \emph{weakest precondition} predicate directly via the semantics:
  \begin{lstlisting}
    wp m Q \mu == case m \mu of spec Q' => \<forall>x \rho \mu'. Q' (x,\rho,\mu') ==> Q x \rho \mu' | fail => False
  \end{lstlisting}
  That is, \is{wp m Q \mu} holds, iff program \is$m$ run on memory \is$\mu$ does not fail, and all possible results (return value \is{x}, access report \is{\rho}, new memory \is{\mu'})
  satisfy the \emph{postcondition} \is{Q}.

  To set up a verification condition generator based on separation logic,
  we standardize the postcondition: the reported memory accesses must be disjoint
  from some abstract memory \is{amf}, called the \emph{frame}. We define the
  \emph{weakest precondition with frame}:
  \begin{lstlisting}
    wpf amf c Q \mu == wp c (%x \rho \mu'. Q x \mu' & disjoint \rho amf) \mu

    disjoint (r,w,a,f) (m,b) == (\<forall>addr. m addr ~= 0 ==> addr ~: r \union w & addr.bidx ~: f)
                                 & (\<forall>i. b i ~= 0 ==> i ~: f)
  \end{lstlisting}
  that is, when executed on memory \is{\mu}, the program \is{c} does not fail, every return value \is{x} and new memory \is{\mu'} satisfies \is{Q},
  and no memory described by the frame \is{amf} is accessed.

  Equipped with a weakest precondition with access restrictions, we define a Hoare-triple:
  \begin{lstlisting}
    ABS amf P \mu == \<exists>am. am ## amf \and \<alpha> \mu = am+amf \and P am

    ht P c Q == \<forall>\mu amf. ABS amf P \mu ==> wpf amf c (%x \mu'. ABS amf (Q x) \mu') \mu
  \end{lstlisting}
  The predicate \is{ABS amf P \mu} specifies that the abstract memory \is$\<alpha> \mu$ can be
  split into a part \is$am$ and the given frame \is$amf$, such that \is{am} satisfies the precondition $P$.
  A Hoare-triple \is{ht P c Q} specifies that for all memories and frames
  for which the precondition holds (\is{ABS amf P \mu}), the program will succeed,
  not using any memory of the frame, and every result will satisfy the postcondition
  wrt.\ the original frame (\is{ABS amf (Q x) \mu'}).

  \subsection{Verification Condition Generator}
  The verification condition generator is implemented as a proof tactic that works on subgoals of the form:
  \begin{lstlisting}
    ABS amf P \mu \and ... ==> wpf amf c Q \mu
  \end{lstlisting}
  The tactic is guided by the syntax of the command \is{c}.
  Basic monad combinators are broken down using the following rules:
  \begin{lstlisting}
    Q r \mu ==> wpf amf (return r) \mu
    wpf amf m (%x. wpf amf (f x) Q) \mu ==> wpf amf ({x <- m; f x}) Q \mu
  \end{lstlisting}
  For other instructions and user defined functions, the VCG expects a Hoare-triple to be already proved.
  It then uses the following rule:
  \begin{lstlisting}
    ht P c Q \and ABS amf P' \mu   --"\rightcomment{match Hoare triple and current state}"
    \and P' |- P*F --"\rightcomment{infer frame}"
    \and (!!r \mu. ABS amf (Q r ** F) \mu ==> Q' r \mu) --"\rightcomment{continue with postcondition}"
    ==> wpf amf c Q' \mu
  \end{lstlisting}
  To process a command \is{c}, the first assumption is instantiated with the Hoare-triple for \is{c},
  and the second assumption with the assertion \is{P'} for the current state.
  Then, a simple syntactic heuristics infers a frame \is{F} and proves that the
  current assertion \is{P'} entails the required precondition \is{P} and the frame.
  Finally, verification condition generation continues with the postcondition \is{Q} and the frame as current assertion.

  \subsection{Hoare-Triples for Instructions}
  To use the VCG to verify LLVM programs, we have to prove Hoare triples for the LLVM instructions.
  For parallel calls, we prove the well-known disjoint concurrency rule~\cite{OH04}:
  \begin{lstlisting}
    ht P_1 c_1 Q_1 \and ht P_2 c_2 Q_2 ==> ht (P_1 * P_2) (par c_1 c_2) (%(r_1,r_2). Q_1 r_1 * Q_2 r_2)
  \end{lstlisting}
  That is, commands with disjoint preconditions can be executed in parallel.

  For memory operations, we prove:
  \begin{lstlisting}
  |= {n~=0} ll_malloc TYPE(\<alpha>) n {%p. range {0..<n} (%_. init) p * b_tag n p}
  |= {range {0..<n} xs p * b_tag n p} ll_free p {%_. \<box>}
  |= {pto x p} ll_load p {%r. r=x * pto x p}
  |= {pto xx p} ll_store x p {%_. pto x p}
  \end{lstlisting}
  Here \is{b_tag n p} asserts that \is{p} points to the beginning of a block of size \is$n$, and
  \is{range I f p} describes that for all $i\in I$, $p+i$ points to value $f~i$.
  Intuitively, \is{ll_malloc} creates a block of size $n$, initialized with the default \is{init} value,
  and a tag. If one possesses both, the whole block and the tag, it can be deallocated by free.
  The rules for load and store are straightforward, where \is{pto x p} describes that $p$ points to value $x$.

  \section{Refinement for Parallel Programs}\label{sec:refine_parallel}
  At this point, we have described a separation logic framework for parallel programs in LLVM.
  It is largely backwards compatible with the framework for sequential programs described in~\cite{La19-llvm},
  such that we could easily port the algorithms formalized there to our new framework.
  The next step towards verifying complex programs is to set up a stepwise refinement framework.
  In this section we describe the refinement infrastructure of the
  Isabelle Refinement Framework, focusing on our changes to support parallel algorithms.

  \subsection{Abstract Programs}
  Abstract programs are shallowly embedded into the nondeterminism error monad \is{'a ne} (cf.\ Section~\ref{sec:monad}).
  They are purely functional, not modifying memory, or differentiating between sequential and parallel execution.
  We define a \emph{refinement ordering} on \is{ne}:
  \begin{lstlisting}
    spec P <= spec Q == \<forall>x. P x ==> Q x         fail $\not\le$ spec Q         m <= fail
  \end{lstlisting}
  Intuitively, \is{m_1 <= m_2} means
  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
  Note that \is$<=$ is a complete lattice, with top element \is$fail$.

%   Arbitrary recursive programs can be defined via a fixed-point construction~\cite{Kr10}.
  We use refinement and assertions to
  specify that a program \is{m} satisfies a specification with precondition \is$P$ and postcondition \is$Q$:
  \begin{lstlisting}
    m <= assert P; spec x. Q x
  \end{lstlisting}
  If the precondition is false, the right hand side is \is{fail}, and the statement trivially holds.
  Otherwise, $m$ cannot fail, and every possible result $x$ of $m$ must satisfy $Q$.

  For a detailed description on using the \is{ne} monad for stepwise refinement based program verification, we refer the reader to~\cite{LaTu12}.

  \subsection{The Sepref Tool}
  The Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{ne}-monad,
  keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
  The synthesis is automatic, but requires use annotations to the abstract program.

  The main concept of the Sepref tool is refinement between an abstract program \is{c} in the \is{ne} monad,
  and a concrete program \is{c$_\dagger$} in the \is{M} monad, as expressed by the \is{hnr}-predicate:
  \begin{lstlisting}
    "hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ==
        c ~= fail ==> ht \<Gamma> c$_\dagger$ (%x. \<exists>x$_\dagger$. \<Gamma>' * R x x$_\dagger$ * \<up>(return x <= c \and CP x$_\dagger$))"
  \end{lstlisting}
  That is, either the abstract program \is{c} fails,
  or for a memory described by assertion \is{\<Gamma>},
  the LLVM program \is{c$_\dagger$} succeeds with \is{r$_\dagger$},
  such that the new memory is described by \is{\<Gamma>' * R x x$_\dagger$},
  for a possible result \is{x} of the abstract program \is{c}. Moreover, the predicate \is{CP} holds for the concrete result.
  Note that \is{hnr} trivially holds for a failing abstract program. This makes sense, as we prove that
  the abstract program does not fail anyway. Moreover it allows us to assume that assertions actually hold during the refinement proof:
  \begin{lstlisting}
    ( \<phi> ==> hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ) ==> hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP (assert \<phi>; c)
  \end{lstlisting}

  \begin{example}(Refinement of lists to arrays)\label{ex:ref_array}
    We define abstract programs for indexing and updating a list:
    \begin{lstlisting}
      lget xs i == assert (i<|xs|); return xs!i     lset xs i x == assert (i<|xs|); return xs[i:=x]
    \end{lstlisting}
    These programs assert that the index is in bounds, and then return the accessed element (\is{xs!i})
    or the updated list (\is{xs[i:=x]}) respectively. The following assertion links a pointer to a list of
    elements stored at the pointed-to location:
    \begin{lstlisting}
      arr$_A$ xs p = range {0..<|xs|} (%i. xs!i) p
    \end{lstlisting}
    That is, for every $i<|xs|$, $p+i$ points to the $i$th element of $xs$.
    On arrays, indexing and updating of arrays is implemented by:
    \begin{lstlisting}
      aget p i == ll_ofs_ptr p i; ll_load p         aset p i x == ll_ofs_ptr p i; ll_store x p; return p
    \end{lstlisting}
    And the abstract and concrete programs are linked by the following refinement theorems:
    \begin{lstlisting}
      hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ i i$_\dagger$)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs xs$_\dagger$$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i i$_\dagger$)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ (%_. True)$\hspace*{-1pt}$ (lget xs i)
      hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ i i$_\dagger$) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i i$_\dagger$) arr$_A$ (%r. r=xs$_\dagger$) (lset xs i x)
    \end{lstlisting}
    That is, if the list \is$xs$ is refined by array \is{xs$_\dagger$}, and the natural number \is$i$ is refined
    by the fixed-width\footnote{We use Isabelle's word library here, which encodes the actual width as a type variable,
    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word \is{i$_\dagger$} (\is{idx$_A$ i i$_\dagger$}),
    the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
    The resulting memory will still contain the original array. Note that there is no explicit precondition that the array access
    is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.
    The \is{aset} operation will return a pointer to an array that refines the updated list returned by \is{lset}.
    As the array is updated in place, the original refinement of the array is no longer valid.
    Moreover, the returned pointer \is{r} will be the same as the argument pointer \is{xs$_\dagger$}.
    This information is important for refining to parallel programs on disjoint parts of an
    array (cf.\ Section~\ref{sec::array_split}).

%     xxx: include pointer-preservation for list_set

  \end{example}

  Given refinement assertions for the parameters, and \is{hnr}-rules for all operations in a program,
  the Sepref tool automatically synthesizes an LLVM program from an abstract \is{ne} program.
  The tool tries to automatically discharge additional proof obligations, typically arising from
  translating arithmetic operations from unbounded numbers to fixed width numbers. Where automatic proof fails,
  the user has to add assertions to the abstract program to help the proof.
  The main difference of our tool wrt.\ the existing Sepref tool~\cite{La19-llvm} is the additional
  condition (\is{CP}) on the concrete result, which is used to track pointer equalities.
  We have added a heuristics to automatically synthesize and discharge these equalities.

  \subsection{Array Splitting}\label{sec::array_split}
  An important concept for parallel programs is to concurrently operate on disjoint parts of the memory,
  e.g., different slices of the same array. However, abstractly, arrays are just lists. They are updated
  by returning a new list, and there is no way to express that the new list is stored at the same address as the old list.
  Nevertheless, in order to refine a program that updates two disjoint slices of a list to one that updates disjoint
  parts of the array in place, we need to know that the result is stored in the same array as the input.
  This is handled by the \is{CP} argument to \is{hnr}. To indicate that operations shall be refined to disjoint parts of the same array, we introduce the
  combinator \is{with_split} for abstract programs:
  \begin{lstlisting}
    with_split i xs f ==
      assert (i < |xs|);
      (xs_1,xs_2) <- f (take i xs) (drop i xs);
      assert (|xs_1| = i \and |xs_2| = |xs| - i);
      return (xs_1@xs_2)
  \end{lstlisting}
  Abstractly, this is an annotation that is inlined when proving the abstract program correct.
  However, Sepref will translate it to the concrete combinator \is{awith_split}:
  \begin{lstlisting}
    awith_split i xs$_\dagger$ f$_\dagger$ == xs$_\dagger$_2 <- ll_ofs_ptr xs$_\dagger$ i; f$_\dagger$ xs$_\dagger$ xs$_\dagger$_2; return xs$_\dagger$

    hnr (arr$_A$ xs_1 xs$_\dagger$_1 * arr$_A$ xs_2 xs$_\dagger$_2) (f$_\dagger$ xs$_\dagger$_1 xs$_\dagger$_2) \<box>
        (arr$_A$ \x arr$_A$) (%(xs$_\dagger$_1',xs$_\dagger$_2'). xs$_\dagger$_1'=xs$_\dagger$_1 \and xs$_\dagger$_2' = xs$_\dagger$_2)
        (f xs_1 xs_2)"
    ==>
    hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ i i$_\dagger$) (awith_split i$_\dagger$ xs$_\dagger$ f$_\dagger$)
        (idx$_A$ i i$_\dagger$) (%xs xs$_\dagger$. arr$_A$ xs xs$_\dagger$) (%xs$_\dagger$'. xs$_\dagger$'=xs$_\dagger$)
        (with_split i xs f)
  \end{lstlisting}
  The refinement of the function \is{f} to \is{f$_\dagger$} requires an additional proof that the returned pointers
  are equal to the argument pointers (\is{xs$_\dagger$_1'=xs$_\dagger$_1 \and xs$_\dagger$_2' = xs$_\dagger$_2}).
  Sepref tries to prove that automatically, using a simple heuristics.

  \subsection{Refinement to Parallel Execution}
  The purely functional abstract programs have no notion of parallel execution.
  To indicate that refinement to parallel execution is desired, we define an abstract annotation
  \is{npar}:
  \begin{lstlisting}
    npar f g a b == x <- f a; y <- g b; return (x,y)

    ${}$  hnr Ax (f$_\dagger$ x$_\dagger$) Ax' Rx CP_1 (f x) \and hnr Ay (g$_\dagger$ y$_\dagger$) Ay' Ry CP_2 (g y)"
    ==>
      hnr (Ax * Ay) (llc_par f$_\dagger$ g$_\dagger$ x$_\dagger$ y$_\dagger$) (Ax' * Ay') (Rx \x Ry)
        (%(x$_\dagger'$,y$_\dagger$'). CP_1 x$_\dagger'$ \and CP_2 y$_\dagger'$) (npar f g x y)"
  \end{lstlisting}
  This rule can be used to automatically parallelize any (independent) abstract computations.
  For convenience, we also define \is{nseq}. Abstractly, it's the same as \is{npar},
  but Sepref translates it to sequential execution.

  \section{A Parallel Sorting Algorithm}\label{sec:parsort}
  To test the usability of our framework, we verify a parallel sorting algorithm.
  We start with the abstract specification of an algorithm that sorts a list:
  \begin{lstlisting}
  sort_spec xs = spec xs'. mset xs'=mset xs & sorted xs
  \end{lstlisting}
  That is, we return a sorted permutation of the original list.
  Note that this is a standard specification of sorting in Isabelle.
  Reusing the existing development of an abstract introsort algorithm~\cite{La20},
  we easily prove with a few refinement steps that the following abstract algorithm implements \is{sort_spec}:
  \begin{lstlisting}[numbers=left,xleftmargin=1.9em,frame=single,framexleftmargin=.5em]
    psort xs n == assert n=|xs|; if n<=1 then return xs else psort_aux$\hspace*{-1pt}$ xs$\hspace*{-1pt}$ n$\hspace*{-1pt}$ (log2$\hspace*{-1pt}$ n$\hspace*{-1pt}$ *$\hspace*{-1pt}$ 2)

    psort_aux xs n d ==
      assert n=|xs|
      if d=0 \or n<100000 then sort_spec xs $\label{l:psort:fallback}$
      else
        (xs,m) <- partition_spec xs; $\label{l:psort:partition}$
        let bad = m<n div 8 \or (n-m < n div 8)  $\label{l:psort:bad}$
        (_,xs) <- with_split m xs (%xs_1 xs_2.
          if bad then nseq psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1) $\label{l:psort:rec_seq}$
          else npar psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)  $\label{l:psort:rec_par}$
        );
        return xs"

    lemma psort xs |xs| <= sort_spec xs
  \end{lstlisting}
  This algorithm is derived from the well-known quicksort and introsort algorithms~\cite{Muss97}:
  like quicksort, it partitions the list (line~\ref{l:psort:partition}), and then recursively sorts the partitions in parallel (l.~\ref{l:psort:rec_par}).
  Like introsort, when the recursion gets too deep, or the list too short, we fall back to
  some (not yet specified) sequential sorting algorithm (l.~\ref{l:psort:fallback}). Similarly, when the partitioning is very unbalanced (l.~\ref{l:psort:bad}),
  we sort the partitions sequentially (l.~\ref{l:psort:rec_seq}). These optimizations aim at not spawning threads for small sorting tasks, where
  the overhead of thread creation outweighs the advantages of parallel execution.
  A more technical aspect is the extra parameter \is{n} that we introduced for the list length.
  Thus, we can refine the list to just a pointer to an array, and still access its length\footnote{Alternatively, we could refine a list to a pair of array pointer and length.}.

  \subsection{Implementation and Correctness Theorem}\label{sec:impl_corr}
  Next, we have to provide implementations for the fallback \is{sort_spec}, and for \is{partition_spec}.
  These implementations must be proved to be in-place, i.e., return a pointer to the same array.
  It was straightforward to amend our existing formalization of \is{pdqsort}~\cite{La20} with the in-place proofs:
  once we had amended the refinement statements, and bug-fixed the pointer equality proving heuristics that we added to Sepref,
  the proofs were automatic.

  Given the implementations of \is{sort_spec} and \is{partition_spec}, the Sepref tool
  generates an LLVM program \is{psort$_\dagger$} from the abstract \is{psort}, and proves a corresponding refinement lemma:
  \begin{lstlisting}
    hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ n n$_\dagger$) (psort$_\dagger$ xs$_\dagger$ n$_\dagger$) (idx$_A$ n n$_\dagger$) arr$_A$ (%r. r = xs$_\dagger$) (psort xs n)
  \end{lstlisting}
  Combining this with the correctness lemma of the abstract \is{psort} algorithm, and unfolding the definition of \is{hnr},
  we prove the following Hoare-triple for our final implementation:
  \begin{lstlisting}
  "ht (arr$_A$ xs xs$_\dagger$ * idx$_A$ n n$_\dagger$ * n = |xs|)
    (psort$_\dagger$ xs$_\dagger$ n$_\dagger$)
    (%r. r=xs$_\dagger$ * \<exists> xs'. arr$_A$ xs' xs$_\dagger$ * sorted xs' * mset xs' = mset xs)"
  \end{lstlisting}
  That is, for a pointer \is{xs$_\dagger$} to an array, whose contents are described by list \is{xs} (\is{arr$_A$}), and a fixed-size word \is{n$_\dagger$} representing the natural number \is{n} (\is{idx$_A$}),
  which must be the number of elements in the list \is{xs}, our sorting algorithm returns the original pointer \is{xs$_\dagger$}, and the array contents are now \is{xs'}, which is sorted and a permutation of \is{xs}.
  Note that this statement uses our semantically defined Hoare triples (cf.\ Section~\ref{sec:wp_hoare}).
  In particular, its correctness does not depend on the refinement steps, the Sepref tool, or the VCG.

%
%   For partitioning, we could also use the existing ninther partitioning algorithm that comes with \is{pdqsort}.
%   However, choosing good partitions is crucial to the degree of parallelization achieved by our algorithm, such that we decided to
%   implement a more precise sampling partitioner: to find a good pivot, we take a certain number (currently 64) of equidistant samples,
%   and use their median as pivot. To find the median, we simply use quicksort\footnote{
%     Using the more efficient quickselect algorithm does not bring any performance advantage here, as the 64 elements are negligible
%     compared to the size of the whole partition. As we had already formalized sorting algorithms, it was easy to reuse them,
%     instead of investing the effort of developing a quickselect algorithm from scratch, for no practical performance gain.}.
%   The efficient partitioning algorithms used by introsort and pdqsort rely on the pivot element to be moved
%   in front of the elements to be partitioned. Thus, we need an algorithm that determines the index of a pivot element,
%   rather than just its value. Abstractly, we define the following algorithm:
%   \begin{lstlisting} sample_pivot xs
%     let ns = min |xs| 64;
%     idxs <- equidist |xs| ns;
%     idxs <- sort_wrt_spec (%i j. xs!i < xs!j) idxs;
%     RETURN (idxs!(ns div 2))
%   \end{lstlisting}
%   here, \is{equidist n ns} computes a list of $ns$ (roughly) equidistant integers in the range ${0..<n}$,
%   and \is{sort_wrt_spec lt} sorts a list wrt.\ the specified comparison operator.
%   To refine this algorithm to an imperative version, we need to support a comparison
%   function that, itself, can access memory (the array where )
%
%
%
%
%   , such that
%   we can then swap it with the first element of the array to be partitioned.
%
%
%   move the pivot element to the front of the array to be sorted,
%   to have a guard value there, which saves some indexing comparison

  \subsection{A Sampling Partitioner}
  While we could simply re-use the existing partitioning algorithm from the pdqsort formalization,
  which uses a pseudomedian of nine pivot selection, we observe that the quality of the pivot is
  particularly important for a balanced parallelization. Moreover, the partitioning in the \is{psort_aux} procedure
  is only done for arrays above a quite big size threshold. Thus, we can invest a little more work to find
  a good pivot, which is still negligible compared to the cost of sorting the resulting partitions.
  We choose a sampling approach, using the median of $64$ equidistant samples as pivot.
  The highly optimized partitioning algorithms that we use swap the pivot to the front of the partition,
  such that we need to determine its index, rather than just its value. We simply use quicksort to find the median\footnote{
  We leave verification of efficient median algorithms, e.g., quickselect, to future work. Note that the overhead of sorting $64$ elements is negligible compared to the large partition that has to be sorted.}:
  \begin{lstlisting}
    sample xs == $is$ <- equidist |xs| 64; $is$ <- sort_wrt (%i j. xs!i < xs!j) $is$; return ($is$!32)
  \end{lstlisting}
  Proving that this algorithm finds a valid pivot index is straightforward. More challenging is to refine it to
  purely imperative LLVM code, which does not support closures like \is{\<lambda>i j. xs!i < xs!j}.

  We resolve such closures over the comparison function manually: using Isabelle's locale mechanism~\cite{KWP99},
  we parametrize over the comparison function. Moreover, we thread through an extra parameter for the data
  captured by the closure:
  \begin{lstlisting}
    locale pcmp =
      fixes lt :: 'p => 'e => 'e => bool and lt$_\dagger$ :: 'p$_\dagger$ => 'e$_\dagger$ => 'e$_\dagger$ => bool
        and par$_A$ :: 'p => 'p$_\dagger$ => assn and elem$_A$ :: 'e => 'e$_\dagger$ => assn
      assumes \<forall>p. weak_ordering (lt p)
      assumes hnr (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi) (lt$_\dagger$ pi ai bi)
        (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True) (lt p a b)
  \end{lstlisting}
  This defines a context in which we have an abstract compare function \is{lt} for the abstract elements of type \is{'e}.
  It takes an extra parameter of type \is{'p} (e.g.\ the list $xs$), and forms a weak ordering\footnote{
    A weak ordering is induced by a mapping of the elements into a total ordering. It is the standard prerequisite for sorting algorithms in C++~\cite{Josu12}.
  }. Moreover, we have a concrete implementation \is{lt$_\dagger$} or the compare function, wrt.\ the refinement assertions
  \is{par$_A$} for the parameter and \is{elem$_A$} for the elements.

  Our sorting algorithm is developed and verified in the context of this locale (to avoid confusion, our presentation has, up to now, just used \is{<} and \is{sorted} instead of \is{lt p} and \is{sorted_wrt (lt p)}).
  To get a sorting algorithm for an actual compare function, we have to instantiate the locale,
  providing an abstract and concrete compare function, along with a proof that the abstract function is a weak ordering,
  and the concrete function refines the abstract one. For our example of sorting indexes into an array,
  where the array elements are, themselves, compared by a parametrized function \is{lt}, we get:
  \begin{lstlisting}
    interpretation idx: pcmp lt_idx lt_idx$_\dagger$ (par$_A$ \<times> arr$_A$) idx$_A$   \<proof>

    lt_idx (p,xs) i j == lt p (xs!i) (xs!j)
    lt_idx$_\dagger$ (p$_\dagger$,xs$_\dagger$) i$_\dagger$ j$_\dagger$ == x$_\dagger$<-aget xs$_\dagger$ i$_\dagger$; y$_\dagger$<-aget xs$_\dagger$ j$_\dagger$; lt$_\dagger$ p$_\dagger$ x$_\dagger$ y$_\dagger$
  \end{lstlisting}
  this yields sorting algorithms for sorting indexes, taking an extra parameter for the array to index into.
  For our sampling application, we use \is{idx.introsort xs}.

  \subsection{Code Generation}
  Finally, we instantiate the sorting algorithms to sort unsigned integers and strings:
  \begin{lstlisting}
    interpretation unat: pcmp (%_. <) (%_. ll_icmp_ult) unat$^{64}_A$ \<proof>
    interpretation str: pcmp (%_. <) (%_. strcmp) str$^{64}_A$ \<proof>
  \end{lstlisting}
  This yields implementations \is{unat.psort$_\dagger$} and \is{str.psort$_\dagger$}, and automatically
  proves instantiated versions of the correctness theorems.

  In a last step, we use our code generator to generate actual LLVM text, as well as a C header file with the
  signatures of the generated functions\footnote{
    For technical reasons, we represent the array size as non-negative signed integer, thus the C signature uses \lstinline[language=C]{int64_t}.
    Moreover, we use a string implementation based on dynamic arrays, rather than C's zero terminated strings.
  }:
  \begin{lstlisting}
    export_llvm
      unat.psort$_\dagger$ is "uint64_t* psort(uint64_t*, int64_t)"
      str.psort$_\dagger$ is "llstring* str_psort(llstring*, int64_t)"
      defines typedef struct {int64_t size; struct {int64_t capacity; char *data;};} llstring;
      file psort.ll
  \end{lstlisting}
  This checks that the specified C signatures are compatible with the actual types, and then
  generates \is{psort.ll} and \is{psort.h}, which can be used in a standard C/C++ toolchain.

%   This can now be compiled, and linked against other C programs using the generated header.
%
%
%   , parametrizing the sorting algorithm over
%   the comparison function
%
%
%
%
%
%   : we take $64$ equidistant samples,
%   and use the median of these samples as pivot.
%
%   Formalization of this sampling algorithm in the Refinement Framework is straightforward.
%   To easily integrate with the existing (highly optimized) implementation of the partitioning algorithm,
%   we have to determine the \emph{index} of the pivot element, swap it with the first element of the array,
%   and then use the existing partitioning algorithm.
%
%   This suggests an algorithm to find the median of a list of sample indexes, comparing each index by looking it up in
%   the list to be sorted, and comparing the corresponding elements.
%   We briefly sketch our approach to formalize such parametrized algorithms using locales, and
%   an extension to Isabelle-LLVM's preprocessor to allow for painless code generation.
%   Note that, for the sake of readability, we have slightly idealized the presentation.
% %   \footnote{
% %     We have omitted some of the fixed parameters, and only describe a parametrized locale.
% %     In practice, we have two versions of the locale, one with a parameter, and one without. While they are still based on the same abstract algorithms, they duplicate some boilerplate code.
% %   }.
%
%
%   A \emph{locale}~\cite{KWP99} is a named context that fixes some parameters and makes assumptions about these parameters.
%   Inside a locale, definitions and theorems can depend on the parameters and assumptions.
%   Later, a locale can be instantiated with actual terms for the parameters, proving that they satisfy the assumptions.
%   This makes available instantiated versions of the definitions and theorems proved inside the locale.
%   Instantiations can also be nested, that is, a locale can be instantiated in the context of another locale.
%
%   We define a locale to set up a parametrized compare function as follows:
%   \begin{lstlisting}
%     locale pcmp =
%       fixes lt lti par$_A$ elem$_A$
%       assumes weak_ordering (lt p)
%       assumes hnr
%         (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)
%         (lti pi ai bi)
%         (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True)
%         (lt p a b)
%
%   \end{lstlisting}
%   that is, we fix an abstract compare function \is{lt}. It takes one extra parameter \is{p},
%   and is a weak ordering\footnote{
%     A weak ordering can be defined by mapping the elements into a total ordering, and is the standard prerequisite for sorting algorithms in C++~\cite{Josu12}.
%   } for each such \is{p}. We also fix a concrete compare function \is{lti}, that refines the abstract compare
%   function, with refinement assertion \is{par$_A$} for the parameter, and \is{elem$_A$} for the elements to be sorted.
%
%   Inside the \is{pcmp} locale, we define our sorting and median functions, threading through the explicit parameter \is{p}.
%   Among others, this will yield a function \is{median p xs}, which returns the median index of the elements in list \is{xs},
%   comparing wrt.\ \is{lt p}. We also get a corresponding implementation \is{median_impl}\footnote{We simply reused a sequential sorting algorithm to completely sort the list and then obtain the middle element. While this is not the most efficient way of finding a median, the overhead when sorting $64$ element arrays is negligible compared to the time spent for sorting the partitions that are at least $10^5$ elements large. Thus, a more efficient algorithm like quickselect would not have any measurable effect on the overall performance, and we leave its verification to future work.}.
%
%   To implement the sampling algorithm, \is{inside} the \is{pcmp} locale, we define
%   a comparison function for indexes, parametrized with the original parameter and the array to be sorted.
%   For this, we instantiate the \is{pcmp} locale again\footnote{In the actual formalization, we use a slightly more complex locale hierarchy to carefully avoid infinite recursive instantiations.}:
%   \begin{lstlisting}
%     context pcmp begin
%
%       definition "lt_idx (p,xs) i j == lt p (xs!i) (xs!j)"
%       definition "lt_idx_impl == ..."
%
%       sublocale IDX: pcmp lt_idx lt_idx_impl (par$_A$ \<times> arr$_A$) idx$_A$   \<proof>
%   \end{lstlisting}
%   This makes available a function \is{IDX.median} and its implementation, which compares indexes into arrays as desired.
%   This is then used in the partitioning function.
%
%   Next, the \is{pcmp} locale is instantiated globally, for example to sort unsigned integers:
%   \begin{lstlisting}
%     global_interpretation unat: pcmp (<) ll_icmp_ult unat$^{64}_A$ \<proof>
%   \end{lstlisting}
%   The \is{global_interpretation} command instantiates the locale for comparing natural
%   numbers implemented by 64 bit words (\is{unat$^{64}_A$}). It yields a proof obligation that
%   the LLVM instruction \is{ll_icmp_ult} actually implements the \is{(<)} operation on natural numbers,
%   which is easily discharged.
%

%
%
%
%
%   For the sake of clarity, we have omitted the fact that our whole development so far has been parametrized
%   over a weak ordering, and a refinement assertion for the array elements.
%   A weak ordering is any ordering induced by a mapping into a total ordering.
%   It is the standard requirement for compare operations used for sorting in C++~\cite{Josu12}.
%   Before we can generate LLVM code, we have to instantiate the compare operator accordingly.
%   For example, to sort natural numbers implemented by unsigned integers (\is{unat$_A$}), we instantiate:
%   \begin{lstlisting}
%     global_interpretation unat: sort_impl_context "(<)" ll_icmp_ult unat$_A$
%   \end{lstlisting}
%   This yields a proof obligation that the LLVM instruction \is{ll_icmp_ult} on LLVM integers
%   actually implements the \is{(<)} operation on natural numbers, wrt.\ the refinement assertion \is{unat$_A$},
%   which is easily discharged. We then get instantiated versions of \is{psort$_\dagger$}, and also of the correctness theorem.
%
%   In a last step, we use our code generator to generate actual LLVM text, as well as a C header file with the
%   signatures of the generated functions\footnote{
%     For technical reasons, we represent the array size as non-negative signed integer, thus the second argument of the C signature has
%     the signed type \lstinline[language=C]{int64_t}.
%   }:
%   \begin{lstlisting}
%     export_llvm unat.psort$_\dagger$ is "uint64_t* psort(uint64_t*, int64_t)" file psort.ll
%   \end{lstlisting}
%   This can now be compiled, and linked against other C programs using the generated header.
%   A similar instantiation is done for the lexicographic ordering on strings, represented by dynamic arrays of bytes.
%

%
%
%
%
%
%
%
%
%   the partition gets too small, or the recursion too deep
%
%
%
%
%   That is, for a valid range of indexes \is{l..<h}, the new list is a permutation of the old list,
%   equal to the old list outside the range, and sorted inside the range.
%
%
%   Our version of the Sepref tool is mostly backwards compatible with the original Sepref tool for Isabelle-LLVM~\cite{La19-llvm}.
%   The only difference is the handling of concrete pointer equalities (cf.\ Section~\ref{sec::array_split}).
%   Thus, it was a mainly straightforward task to port the efficient verified Introsort and pdqsort algorithms~\cite{La20} to the new tool,
%   additionally proving that the concrete algorithm is in-place, i.e., the returned array pointer equals the argument pointer.
%   Equipped with these refinements, the Sepref tool can synthesize a parallel sorting algorithm from \is{sort_aux} (Section~\ref{sec:sortspec}), using the already existing partitioning algorithm and the
%   pdqsort algorithm for the abstract \is{partition_spec} and \is{sort_spec} specifications.
%
%   Actually, we used refinement steps on the abstract level to add two optimizations:
%   first, we add an explicit parameter for the length of the list. This allows us to refine the list
%   to just an array pointer\footnote{Alternatively, we could refine a list to a pair of array pointer and explicit length.}.
%   Second, to prevent creation of parallel threads for small partitions, we proceed sequentially if the created partitions
%   are too unbalanced. The following is the algorithm that we input into Sepref\footnote{We have omitted some
%   of the assertions required to prove that the arithmetic operations do not overflow.}:
%   \begin{lstlisting}
%     psort_aux xs n d ==
%       assert n=|xs|
%       if d=0 \or n<100000 then {
%         slice_sort_spec xs 0 n
%       } else {
%         (xs,m) <- slice_partition_spec xs 0 n;
%         let bad = m<n div 8 \or (n-m < n div 8)
%         (_,xs) <- with_split m xs (%xs_1 xs_2. {
%           if bad then nseq psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
%           else npar psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
%         });
%         return xs
%       }"
%
%     nseq f_1 f_2 x_1 x_2 == r_1 <- f_1 x_1; r_2 <- f_2 x_2; return (x_1,x_2)
%
%     psort xs n ==
%       assert n=|xs|
%       if n<=1 then return xs
%       else psort_aux xs n (log2 n * 2)
%
%   \end{lstlisting}
%   where \is{nseq} is abstractly the same as \is{npar}, but is translated to sequential execution by the Sepref tool.
%   Moreover, \is{slice_sort_spec} and \is{slice_partition_spec} are the specifications for sorting and partitioning
%   only a slice of the array, leaving the rest unchanged. These match the interface of the original sequential algorithms
%   proved correct in~\cite{La20}. Finally, \is{psort} wraps the algorithm to run with an initial depth limit of $2\lfloor\log_2 n\rfloor$.

%   \subsection{A Sampling Partitioner}
%   While we could simply re-use the existing partitioning algorithm from the Introsort formalization,
%   which uses a median-of-three pivot selection, we observe that the quality of the pivot is
%   particularly important for a balanced parallelization. Moreover, the partitioning in the \is{psort_aux} procedure
%   is only done for arrays above a quite big size threshold. Thus, we can invest a little more work to find
%   a good pivot, which is still negligible compared to the cost of sorting the resulting partitions.
%
%   We choose a rather simple sampling approach: for an array of size $n$, we take $\min(n,64)$ equidistant samples,
%   and use the median of these samples as pivot.
%
%   Formalization of this sampling algorithm in the Refinement Framework is straightforward.
%   To easily integrate with the existing (highly optimized) implementation of the partitioning algorithm,
%   we have to determine the \emph{index} of the pivot element, swap it with the first element of the array,
%   and then use the existing partitioning algorithm.
%
%   This suggests an algorithm to find the median of a list of sample indexes, comparing each index by looking it up in
%   the list to be sorted, and comparing the corresponding elements.
%   We briefly sketch our approach to formalize such parametrized algorithms using locales, and
%   an extension to Isabelle-LLVM's preprocessor to allow for painless code generation.
%   Note that, for the sake of readability, we have slightly idealized the presentation\footnote{
%     We have omitted some of the fixed parameters, and only describe a parametrized locale.
%     In practice, we have two versions of the locale, one with a parameter, and one without. While they are still based on the same abstract algorithms, they duplicate some boilerplate code.
%   }.
%
%   A \emph{locale}~\cite{KWP99} is a named context that fixes some parameters and makes assumptions about these parameters.
%   Inside a locale, definitions and theorems can depend on the parameters and assumptions.
%   Later, a locale can be instantiated with actual terms for the parameters, proving that they satisfy the assumptions.
%   This makes available instantiated versions of the definitions and theorems proved inside the locale.
%   Instantiations can also be nested, that is, a locale can be instantiated in the context of another locale.
%
%   We define a locale to set up a parametrized compare function as follows:
%   \begin{lstlisting}
%     locale pcmp =
%       fixes lt lti par$_A$ elem$_A$
%       assumes weak_ordering (lt p)
%       assumes hnr
%         (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)
%         (lti pi ai bi)
%         (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True)
%         (lt p a b)
%
%   \end{lstlisting}
%   that is, we fix an abstract compare function \is{lt}. It takes one extra parameter \is{p},
%   and is a weak ordering\footnote{
%     A weak ordering can be defined by mapping the elements into a total ordering, and is the standard prerequisite for sorting algorithms.
%   } for each such \is{p}. We also fix a concrete compare function \is{lti}, that refines the abstract compare
%   function, with refinement assertion \is{par$_A$} for the parameter, and \is{elem$_A$} for the elements to be sorted.
%
%   Inside the \is{pcmp} locale, we define our sorting and median functions, threading through the explicit parameter \is{p}.
%   Among others, this will yield a function \is{median p xs}, which returns the median index of the elements in list \is{xs},
%   comparing wrt.\ \is{lt p}. We also get a corresponding implementation \is{median_impl}\footnote{We simply reused a sequential sorting algorithm to completely sort the list and then obtain the middle element. While this is not the most efficient way of finding a median, the overhead when sorting $64$ element arrays is negligible compared to the time spent for sorting the partitions that are at least $10^5$ elements large. Thus, a more efficient algorithm like quickselect would not have any measurable effect on the overall performance, and we leave its verification to future work.}.
%
%   To implement the sampling algorithm, \is{inside} the \is{pcmp} locale, we define
%   a comparison function for indexes, parametrized with the original parameter and the array to be sorted.
%   For this, we instantiate the \is{pcmp} locale again\footnote{In the actual formalization, we use a slightly more complex locale hierarchy to carefully avoid infinite recursive instantiations.}:
%
%   \begin{lstlisting}
%     context pcmp begin
%
%       definition "lt_idx (p,xs) i j == lt p (xs!i) (xs!j)"
%       definition "lt_idx_impl == ..."
%
%       sublocale IDX: pcmp lt_idx lt_idx_impl (par$_A$ \<times> arr$_A$) idx$_A$   \<proof>
%
%   \end{lstlisting}
%   This makes available a function \is{IDX.median} and its implementation, which compares indexes into arrays as desired.
%   This is then used in the partitioning function.
%
%   Finally, the \is{pcmp} locale is instantiated globally, for example to sort unsigned integers, and the resulting implementation is
%   exported to LLVM text:
%   \begin{lstlisting}
%     global_interpretation unat: pcmp (<) ll_icmp_ult unat$^{64}_A$   \<proof>
%
%     export_llvm unat.psort$_\dagger$
%   \end{lstlisting}
%   The \is{global_interpretation} command instantiates the locale for comparing natural
%   numbers implemented by 64 bit words (\is{unat$^{64}_A$}). Note that the compare function is not
%   parametrized. Finally, the \is{export_llvm} command generates LLVM text.
%   It first uses a preprocessor to generate and prove correct a set of code equations, which are then pretty printed to actual LLVM text.
%   The preprocessor does not contribute to the trusted code base, as it proves correct its transformations.
%   At this point, we have added a transformation to automatically instantiate the higher-order parameters generated by Isabelle's locale mechanism.
%   This saves a lot of boilerplate code that was necessary in the original sorting algorithm formalization~\cite{La20}, and became unmanageable
%   once we added the sublocale for comparing indexes into an array.

%
%   xxx: array splitting
%
%   then short explanation that proofs are semi-automatic.
%
%
%   also mention optimizations of algorithm (bad-partition)
%
%
%
%   Sepref comes with standard setup to refine lists to arrays. List updates are refined to destructive array updates,
%   as long as the old version of the list is not used after the update.
%   It also provides setup to refine unbounded natural numbers to bounded integers.
%   It tries to discharge the resulting in-bounds proof obligations automatically.
%   If this is not possible, it relies on hints from the user.
%
%
%     - intro to sepref
%     - parallel hnr rule
%     - identity tracking
%     - array splitting

%   \section{Parallel Quicksort}
%   \subsection{The Algorithm}
%
%   \subsection{Array Data Structure}

%   \subsection{Final Implementation and Correctness Theorem}
%   From this algorithm, the Sepref tool generates an imperative implementation \is{psort$_\dagger$} and proves the corresponding \is{hnr} refinement lemma.
%   Finally, our code generator translates \is{psort$_\dagger$} to an LLVM function.
%   Combining the refinement lemma with the correctness lemma of the abstract \is{psort} algorithm, and unfolding the definition of \is{hnr},
%   we get the following Hoare-triple, that states correctness of our final algorithm:
%   \begin{lstlisting}
%   "ht (arr$_A$ xs xsi * idx$_A$ n ni * n = |xs|)
%     (psort$_\dagger$ xsi ni)
%     (%r. r=xsi * \<exists> xs'. arr$_A$ xs' xsi * sorted xs' * mset xs' = mset xs)"
%   \end{lstlisting}
%   That is, for a pointer \is{xsi} to an array, whose contents are described by list \is{xs} (\is{arr$_A$}), and a fixed-size word \is{ni} representing the natural number \is{n} (\is{idx$_A$}),
%   which must be the number of elements in the list \is{xs}, our sorting algorithm returns the original pointer \is{xsi}, and the array contents are now \is{xs'},
%   which is sorted and a permutation of \is{xs}.
%   Note that, while the trusted code base of this statement includes our LLVM semantics and code generator, as well as Isabelle's inference kernel,
%   it does not include the refinement steps, verification condition generation, Hoare-rules, or automated theorem provers invoked by sledgehammer~\cite{BBP13},
%   which we used to prove this statement.
%

%
%
%   \subsection{Specification of Sorting Algorithms}\label{sec:sortspec}
%   The first step to verify a sorting algorithm is to specify the desired result.
%   We specify an algorithm that sorts a list \is{xs} as follows:
%   \begin{lstlisting}
%     sort_spec xs == spec xs'. sorted xs' \and mset xs' = mset xs
%   \end{lstlisting}
%   where \is{mset xs} is the multiset of the elements in the list \is{xs}.
%   That is, the resulting list \is{xs'} is sorted, and a permutation of the original list \is{xs}.
%
%   For quicksort-like sorting algorithms, the main function is partitioning, which reshuffles a
%   given list into two partitions, such that any element in the first partition is less than or equal
%   to any element in the second partition. We specify a partitioning function that assumes at least four elements,
%   and is guaranteed to return non-empty partitions. The result is returned as a reshuffled list \is{xs'},
%   and the index \is{m} of the first element of the second partition:
%   \begin{lstlisting}
%     partition_spec xs = assert (4 <= length xs);
%       spec (xs',m). mset xs' = mset xs \and 0<m \and m<|xs|
%       \and (\<forall>i\<in>{0..<m}. \<forall>j\<in>{m..<|xs|}. xs'!i <= xs'!j)
%   \end{lstlisting}
%
%   Using the above specifications as building blocks, we can define a simple abstract sorting algorithm:
%   \begin{lstlisting}
%     sort_aux xs d ==
%       if d=0 \or |xs|<100000 then { --"\rightcomment{check depth bound or size threshold}"
%         sort_spec xs --"\rightcomment{sort with some (sequential) algorithm}"
%       } else {
%         (xs,m) <- partition_spec xs;  --"\rightcomment{partition}"
%         with_split m xs (%xs_1 xs_2. { --"\rightcomment{work on partitions $xs_1$ and $xs_2$}"
%           npar sort_aux sort_aux (xs_1,d-1) (xs_2,d-1) --"\rightcomment{recursively sort in parallel}"
%         })
%       }"
%   \end{lstlisting}
%   where
%   \begin{lstlisting}
%     with_split m xs f ==
%       assert (m < |xs|);
%       (xs_1,xs_2) <- f (take m xs) (drop m xs);
%       assert (|xs_1| = m \and |xs_2| = |xs| - m);
%       return (xs_1@xs_2)
%
%     npar f g x y == r1 <- f x; r2 <- g y; return (r1,r2)
%   \end{lstlisting}
%   where \is{@} appends two lists.
%   This algorithm is derived from Introsort~\cite{Muss97}, a quicksort with a depth bound to ensure $O(n\log n)$ worst-case complexity.
%   If the recursion depth bound $d$ is reached, or the list size is below a certain threshold,
%   we sort the list using some yet unspecified sorting algorithm. Otherwise, we partition the list
%   and recursively sort the two partitions.
%
%   Here, we have introduced two new combinators,
%   which are hints for the following refinement steps: (1) the \is{with_split} combinator splits a given list into two parts,
%   and invokes a function that can change the content of these two parts, but not their length. It then returns
%   the concatenation of the two changed parts. It is a hint for the further refinement steps to implement the list as
%   an array that is modified in-place. (2) the \is{npar} combinator simply executes two blocks sequentially.
%   Note that, for our stateless abstract programs, there is no difference between parallel and (independent) sequential execution.
%   However, \is{npar} is a hint to the subsequent refinement steps to refine to parallel execution.
%
%   Using the tools of the Isabelle Refinement Framework~\cite{LaTu12}, and some existing theories copied from~\cite{La19-llvm},
%   it is straightforward to prove that this sorting algorithm is correct, i.e.:
%   \begin{lstlisting}
%     sort_aux xs d <= sort_spec xs
%   \end{lstlisting}
%
%   In the next sections, we will explain how to refine this two an actual parallel sorting algorithm in LLVM,
%   re-using the existing sequential pdqsort algorithm~\cite{La19-llvm} for small or too deep partitions, and a sampling algorithm for finding good pivots for partitioning.
%
%
%
%   \section{A Parallel Sorting Algorithm}\label{sec:parsort}
%   Our version of the Sepref tool is mostly backwards compatible with the original Sepref tool for Isabelle-LLVM~\cite{La19-llvm}.
%   The only difference is the handling of concrete pointer equalities (cf.\ Section~\ref{sec::array_split}).
%   Thus, it was a mainly straightforward task to port the efficient verified Introsort and pdqsort algorithms~\cite{La20} to the new tool,
%   additionally proving that the concrete algorithm is in-place, i.e., the returned array pointer equals the argument pointer.
%   Equipped with these refinements, the Sepref tool can synthesize a parallel sorting algorithm from \is{sort_aux} (Section~\ref{sec:sortspec}), using the already existing partitioning algorithm and the
%   pdqsort algorithm for the abstract \is{partition_spec} and \is{sort_spec} specifications.
%
%   Actually, we used refinement steps on the abstract level to add two optimizations:
%   first, we add an explicit parameter for the length of the list. This allows us to refine the list
%   to just an array pointer\footnote{Alternatively, we could refine a list to a pair of array pointer and explicit length.}.
%   Second, to prevent creation of parallel threads for small partitions, we proceed sequentially if the created partitions
%   are too unbalanced. The following is the algorithm that we input into Sepref\footnote{We have omitted some
%   of the assertions required to prove that the arithmetic operations do not overflow.}:
%   \begin{lstlisting}
%     psort_aux xs n d ==
%       assert n=|xs|
%       if d=0 \or n<100000 then {
%         slice_sort_spec xs 0 n
%       } else {
%         (xs,m) <- slice_partition_spec xs 0 n;
%         let bad = m<n div 8 \or (n-m < n div 8)
%         (_,xs) <- with_split m xs (%xs_1 xs_2. {
%           if bad then nseq psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
%           else npar psort_aux psort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
%         });
%         return xs
%       }"
%
%     nseq f_1 f_2 x_1 x_2 == r_1 <- f_1 x_1; r_2 <- f_2 x_2; return (x_1,x_2)
%
%     psort xs n ==
%       assert n=|xs|
%       if n<=1 then return xs
%       else psort_aux xs n (log2 n * 2)
%
%   \end{lstlisting}
%   where \is{nseq} is abstractly the same as \is{npar}, but is translated to sequential execution by the Sepref tool.
%   Moreover, \is{slice_sort_spec} and \is{slice_partition_spec} are the specifications for sorting and partitioning
%   only a slice of the array, leaving the rest unchanged. These match the interface of the original sequential algorithms
%   proved correct in~\cite{La20}. Finally, \is{psort} wraps the algorithm to run with an initial depth limit of $2\lfloor\log_2 n\rfloor$.
%
% %   \subsection{A Sampling Partitioner}
% %   While we could simply re-use the existing partitioning algorithm from the Introsort formalization,
% %   which uses a median-of-three pivot selection, we observe that the quality of the pivot is
% %   particularly important for a balanced parallelization. Moreover, the partitioning in the \is{psort_aux} procedure
% %   is only done for arrays above a quite big size threshold. Thus, we can invest a little more work to find
% %   a good pivot, which is still negligible compared to the cost of sorting the resulting partitions.
% %
% %   We choose a rather simple sampling approach: for an array of size $n$, we take $\min(n,64)$ equidistant samples,
% %   and use the median of these samples as pivot.
% %
% %   Formalization of this sampling algorithm in the Refinement Framework is straightforward.
% %   To easily integrate with the existing (highly optimized) implementation of the partitioning algorithm,
% %   we have to determine the \emph{index} of the pivot element, swap it with the first element of the array,
% %   and then use the existing partitioning algorithm.
% %
% %   This suggests an algorithm to find the median of a list of sample indexes, comparing each index by looking it up in
% %   the list to be sorted, and comparing the corresponding elements.
% %   We briefly sketch our approach to formalize such parametrized algorithms using locales, and
% %   an extension to Isabelle-LLVM's preprocessor to allow for painless code generation.
% %   Note that, for the sake of readability, we have slightly idealized the presentation\footnote{
% %     We have omitted some of the fixed parameters, and only describe a parametrized locale.
% %     In practice, we have two versions of the locale, one with a parameter, and one without. While they are still based on the same abstract algorithms, they duplicate some boilerplate code.
% %   }.
% %
% %   A \emph{locale}~\cite{KWP99} is a named context that fixes some parameters and makes assumptions about these parameters.
% %   Inside a locale, definitions and theorems can depend on the parameters and assumptions.
% %   Later, a locale can be instantiated with actual terms for the parameters, proving that they satisfy the assumptions.
% %   This makes available instantiated versions of the definitions and theorems proved inside the locale.
% %   Instantiations can also be nested, that is, a locale can be instantiated in the context of another locale.
% %
% %   We define a locale to set up a parametrized compare function as follows:
% %   \begin{lstlisting}
% %     locale pcmp =
% %       fixes lt lti par$_A$ elem$_A$
% %       assumes weak_ordering (lt p)
% %       assumes hnr
% %         (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)
% %         (lti pi ai bi)
% %         (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True)
% %         (lt p a b)
% %
% %   \end{lstlisting}
% %   that is, we fix an abstract compare function \is{lt}. It takes one extra parameter \is{p},
% %   and is a weak ordering\footnote{
% %     A weak ordering can be defined by mapping the elements into a total ordering, and is the standard prerequisite for sorting algorithms.
% %   } for each such \is{p}. We also fix a concrete compare function \is{lti}, that refines the abstract compare
% %   function, with refinement assertion \is{par$_A$} for the parameter, and \is{elem$_A$} for the elements to be sorted.
% %
% %   Inside the \is{pcmp} locale, we define our sorting and median functions, threading through the explicit parameter \is{p}.
% %   Among others, this will yield a function \is{median p xs}, which returns the median index of the elements in list \is{xs},
% %   comparing wrt.\ \is{lt p}. We also get a corresponding implementation \is{median_impl}\footnote{We simply reused a sequential sorting algorithm to completely sort the list and then obtain the middle element. While this is not the most efficient way of finding a median, the overhead when sorting $64$ element arrays is negligible compared to the time spent for sorting the partitions that are at least $10^5$ elements large. Thus, a more efficient algorithm like quickselect would not have any measurable effect on the overall performance, and we leave its verification to future work.}.
% %
% %   To implement the sampling algorithm, \is{inside} the \is{pcmp} locale, we define
% %   a comparison function for indexes, parametrized with the original parameter and the array to be sorted.
% %   For this, we instantiate the \is{pcmp} locale again\footnote{In the actual formalization, we use a slightly more complex locale hierarchy to carefully avoid infinite recursive instantiations.}:
% %
% %   \begin{lstlisting}
% %     context pcmp begin
% %
% %       definition "lt_idx (p,xs) i j == lt p (xs!i) (xs!j)"
% %       definition "lt_idx_impl == ..."
% %
% %       sublocale IDX: pcmp lt_idx lt_idx_impl (par$_A$ \<times> arr$_A$) idx$_A$   \<proof>
% %
% %   \end{lstlisting}
% %   This makes available a function \is{IDX.median} and its implementation, which compares indexes into arrays as desired.
% %   This is then used in the partitioning function.
% %
% %   Finally, the \is{pcmp} locale is instantiated globally, for example to sort unsigned integers, and the resulting implementation is
% %   exported to LLVM text:
% %   \begin{lstlisting}
% %     global_interpretation unat: pcmp (<) ll_icmp_ult unat$^{64}_A$   \<proof>
% %
% %     export_llvm unat.psort$_\dagger$
% %   \end{lstlisting}
% %   The \is{global_interpretation} command instantiates the locale for comparing natural
% %   numbers implemented by 64 bit words (\is{unat$^{64}_A$}). Note that the compare function is not
% %   parametrized. Finally, the \is{export_llvm} command generates LLVM text.
% %   It first uses a preprocessor to generate and prove correct a set of code equations, which are then pretty printed to actual LLVM text.
% %   The preprocessor does not contribute to the trusted code base, as it proves correct its transformations.
% %   At this point, we have added a transformation to automatically instantiate the higher-order parameters generated by Isabelle's locale mechanism.
% %   This saves a lot of boilerplate code that was necessary in the original sorting algorithm formalization~\cite{La20}, and became unmanageable
% %   once we added the sublocale for comparing indexes into an array.
%
%
% %
% %   xxx: array splitting
% %
% %   then short explanation that proofs are semi-automatic.
% %
% %
% %   also mention optimizations of algorithm (bad-partition)
% %
% %
% %
% %   Sepref comes with standard setup to refine lists to arrays. List updates are refined to destructive array updates,
% %   as long as the old version of the list is not used after the update.
% %   It also provides setup to refine unbounded natural numbers to bounded integers.
% %   It tries to discharge the resulting in-bounds proof obligations automatically.
% %   If this is not possible, it relies on hints from the user.
% %
% %
% %     - intro to sepref
% %     - parallel hnr rule
% %     - identity tracking
% %     - array splitting
%
% %   \section{Parallel Quicksort}
% %   \subsection{The Algorithm}
% %
% %   \subsection{Array Data Structure}
%
%
%
%
%   \subsection{Final Implementation and Correctness Theorem}
%   From this algorithm, the Sepref tool generates an imperative implementation \is{psort$_\dagger$} and proves the corresponding \is{hnr} refinement lemma.
%   Finally, our code generator translates \is{psort$_\dagger$} to an LLVM function.
%   Combining the refinement lemma with the correctness lemma of the abstract \is{psort} algorithm, and unfolding the definition of \is{hnr},
%   we get the following Hoare-triple, that states correctness of our final algorithm:
%   \begin{lstlisting}
%   "ht (arr$_A$ xs xsi * idx$_A$ n ni * n = |xs|)
%     (psort$_\dagger$ xsi ni)
%     (%r. r=xsi * \<exists> xs'. arr$_A$ xs' xsi * sorted xs' * mset xs' = mset xs)"
%   \end{lstlisting}
%   That is, for a pointer \is{xsi} to an array, whose contents are described by list \is{xs} (\is{arr$_A$}), and a fixed-size word \is{ni} representing the natural number \is{n} (\is{idx$_A$}),
%   which must be the number of elements in the list \is{xs}, our sorting algorithm returns the original pointer \is{xsi}, and the array contents are now \is{xs'},
%   which is sorted and a permutation of \is{xs}.
%   Note that, while the trusted code base of this statement includes our LLVM semantics and code generator, as well as Isabelle's inference kernel,
%   it does not include the refinement steps, verification condition generation, Hoare-rules, or automated theorem provers invoked by sledgehammer~\cite{BBP13},
%   which we used to prove this statement.
%

  \subsection{Benchmarks}\label{sec:benchmark}
  \newcommand{\plottim}[4]{
  \begin{tikzpicture}
    \begin{axis}[
      ymajorgrids,
      title={\large #1},
      title style={at={(0.1,.75)}},
      ybar=0pt,
      width=\textwidth,
      height=.25\textheight,
      bar width=3pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,}
      ,
      #3
    ]

      \input{#2}

      \addplot[black, thin, sharp plot,update limits=false]
	coordinates {([normalized]-10,1) ([normalized]100,1)};
% 	node[above] at (axis cs:1950,4.3e7) {Houses};
      #4
    \end{axis}
  \end{tikzpicture}
  }

  \begin{figure}
    \plottim{Laptop, uint64}{data_laptop_uint64}{xticklabels={}}{\legend{}}
    \plottim{Server, uint64}{data_server_uint64}{xticklabels={}}{\legend{}}
    \plottim{Laptop, string}{data_laptop_llstring}{xticklabels={}}{\legend{}}
    \plottim{Server, string}{data_server_llstring}{
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      ymax=1000,
      legend style = {
%         legend pos= north east,
        at = {(.87,.97)},
        cells={anchor=west},
        font=\scriptsize
      },
    }{}

  \caption{Runtimes in milliseconds for sorting various distributions of unsigned 64 bit integers and strings with
    our verified parallel sorting algorithm, C++'s standard parallel sorting algorithm, and Boost's parallel sample sort algorithm.
    The experiments were performed on a server machine with 22 AMD Opteron 6176 cores and 128GiB of RAM, and a laptop with a
    6 core (12 threads) i7-10750H CPU and 32GiB of RAM.
  }\label{fig:benchres}
  \end{figure}

  \begin{figure}

  \begin{tikzpicture}
    \begin{axis}[
      xlabel near ticks,
      legend style = {
        at = {(.87,.3)},
        cells={anchor=west},
        font=\scriptsize
      },
      ymajorgrids,
      title={\large Laptop},
      title style={at={(0.15,.85)}},
      width=.54\textwidth,
      height=.3\textheight
    ]
      \input{sudata_laptop}

      \draw[black, thin, sharp plot]
      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});

%       \addplot[black, thin, sharp plot, update limits=false] {\x};
    \end{axis}
  \end{tikzpicture}\hfill
  \begin{tikzpicture}
    \begin{axis}[
      xlabel near ticks,
      legend style = {
        at = {(.87,.3)},
        cells={anchor=west},
        font=\scriptsize
      },
      ymajorgrids,
      title={\large Server},
      title style={at={(0.15,.85)}},
      width=.54\textwidth,
      height=.3\textheight
    ]
      \input{sudata_server}

      \draw[black, thin, sharp plot]
      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});

%       \addplot[black, thin, sharp plot, update limits=false] {\x};
    \end{axis}
  \end{tikzpicture}
  \caption{Speedup of the various implementations, for sorting unsigned 64 bit integers with a random distribution, on
    a server with 22 AMD Opteron 6176 cores and 128GiB of RAM, and a laptop with a
    6 core (12 threads) i7-10750H CPU and 32GiB of RAM.
    The x axis ranges over the number of cores, and the y-axis gives the speedup wrt.\ the same implementation run on only one core.
    The thin black lines indicate linear speedup.
  }\label{fig:speedup}

  \end{figure}

  We have benchmarked our verified sorting algorithm against a direct implementation of the same algorithm in C++.
  The result was that both implementations have the same runtime, up to some minor noise.
  This indicates that there is no systemic slowdown: algorithms verified with our framework run as fast as their unverified counterparts implemented in C++.

  We also benchmarked against the state-of-the-art implementations \is{std::sort} with execution policy \is{par_unseq} from the
  GNU C++ standard library~\cite{libstdc++},
  and \is{sample_sort} from the Boost C++ libraries~\cite{boost,boost-sort}.
  We have benchmarked the algorithm on two different machines, and various input distributions. The results are shown in Figure~\ref{fig:benchres}.
  While our verified algorithm is clearly competitive for integer sorting on the less parallel laptop
  machine, it's slightly less efficient for sorting strings on the highly parallel server machine.
  Nevertheless, we believe that our verified implementation is already useful in practice,
  and leave further optimizations to future work.

  Finally, we measured the speedup that the implementations achieve for a certain number of cores.
  The results are displayed in Figure~\ref{fig:speedup}. While the speedup on the moderately parallel
  laptop is comparable to the one of the C++ standard library, our implementation
  achieves lower speedups than the state-of-the-art on the highly parallel server.
  Again, we leave further optimizations to future work.

%   Again, in particular for higher numbers of cores,
%   the speedup achieved by the state-of-the-art algorithms is better than for our simple algorithm,
%   highlighting some optimization potential.

%
%   Nevertheless, our verified algorithm is still competitive when compared to state-of-the-art implementations.
%
%   Nevertheless, while our simple algorithm cannot compete with the best practical implementations,
%   it is still significantly faster than sequential sorting in most cases.
%   Moreover, our verified implementation has the same performance as an unverified C++ implementation
%   of the same algorithm.
%
%   We have benchmarked our algorithm on standard laptop and server hardware,
%   for sorting 64 bit integers, with various input distributions.
%   The runtimes of our verified algorithm ($t_1$) and its unverified C++ implementation ($t_2$) is within 6\%
%   of each other ($.94 <= t_1/t_2 <= 1.06$).
%   Figure~\ref{fig:benchres} shows the results of comparing our verified parallel sorting algorithm against
%   C++ standard sequential sorting algorithm, and the state-of-the-art sample sort implementation from the Boost Libraries.
%   Sample sort s always faster than sequential sort, and faster than our verified algorithm in most cases.
%   However, except for one outlier, even our simple verified algorithm is significantly faster than the sequential algorithm.
%   Thus, we expect our verified algorithm to be useful in practice, and leave further optimizations to future work.

  \section{Conclusions}\label{sec:concl}
    We have presented a stepwise refinement approach to verify total correctness of efficient parallel algorithms.
    Our approach targets LLVM as back end, and there is no systemic efficiency loss in our approach
    when compared to unverified algorithms implemented in C++.

    The trusted code base of our approach is relatively small:
    apart from Isabelle's inference kernel, it contains our shallow embedding of a small fragment of
    the LLVM semantics, and the code generator.
    All other tools that we used, e.g., our Hoare logic, Sepref tool, and Refinement Framework for abstract programs,
    ultimately prove a correctness theorem that only depends on our shallowly embedded semantics.

    As a case study, we have implemented a parallel sorting algorithm.
    It uses an existing verified sequential pdqsort algorithm as a building block,
    and is competitive with state-of-the-art parallel sorting algorithms,
    at least on moderately parallel hardware.

    The main idea of our parallel extension is to shallowly embed the semantics of a
    parallel combinator into a sequential semantics, by making the
    semantics report the accessed memory locations, and fail if there is a potential data race.
    We only needed to changed the lower levels of our existing framework for sequential LLVM~\cite{La19-llvm}.
    Higher-level tools like the VCG and Sepref remained largely unchanged and backwards compatible.
    This greatly simplified reusing of existing verification projects, like the sequential pdqsort algorithm~\cite{La20}.

%     In our approach, the high-level algorithms are phrased as purely functional and sequential algorithms,
%     that, however, already contain hints how they should be refined to imperative parallel algorithms.
%     Only during the last refinement step, performed by Sepref, actual imperative and parallel code is generated.

    \subsection{Related Work}
    While there is extensive work on parallel sorting algorithm (e.g.~\cite{CNLM08,AMI21}),
    there seems to be almost no work on their formal verification. The only work we are aware of is
    a distributed merge sort algorithm~\cite{HKBK20}, for which ''no effort has been made to make it efficient''\cite[Sec.~2]{HKBK20},
    nor any executable code has been generated or benchmarked. Another verification~\cite{SaHu20} uses
    the VerCors deductive verifier to prove the permutation property (\is{mset xs' = mset xs})
    of odd-even transposition sort~\cite{Ha72}, but neither the sortedness property nor termination.

    Concurrent separation logic is used by many verification tools such
    as VerCors~\cite{Vercors}, and also formalized in proof assistants, for example in the VST~\cite{VST}
    and IRIS~\cite{JKJA18} projects for Coq~\cite{BeCa10}. These formalizations contain elaborate concepts to
    reason about communication between threads via shared memory, and are typically used to
    verify partial correctness of subtle concurrent algorithms (e.g.~\cite{MeJo21}).
    Reasoning about total correctness is more complicated in the step-indexed separation logic provided by IRIS,
    and currently only supported for sequential programs~\cite{SGGT21}.
    Our approach is less expressive, but naturally supports total correctness, and is already sufficient for
    many practically relevant parallel algorithms like sorting, matrix-multiplication, or parallel algorithms from
    the C++ STL.

%     parallel sorting?
%
%     IRIS and VST: more expressive concurrent operators and sep-logic. Only partial correctness.
%       Focus on subtle concurrency, rather than simple parallelism but with subtly optimized sequential parts.

    \subsection{Future Work}
      An obvious next step is to implement a fractional separation logic~\cite{BCOP05},
      to reason about parallel threads that share read-only memory. While our semantics
      already supports shared read-only memory, our separation logic does not.
      We believe that implementing a fractional separation logic will be straightforward,
      and mainly pose technical issues for automatic frame inference.

      Another obvious next step is to verify a state-of-the-art parallel sorting algorithm,
      like Boost's sample sort. Like our current algorithm, sample sort does not require
      advanced synchronization concepts, and can be implemented only with a parallel combinator.

      Finally, the Sepref framework has recently been extended to reason about complexity of (sequential)
      LLVM programs~\cite{HaLa21-toplas,HaLa21}. This line of work could be combined with our parallel extension, to verify the
      complexity (e.g. work and span) of parallel algorithms.

      Extending our approach towards more advanced
      synchronization like locks or atomic operations may be possible: instead of accessed memory addresses,
      a thread could report a set of possible traces, which are checked for race-freedom and then combined.

      Finally, our framework currently targets multicore CPUs. Another important architecture are general purpose GPUs.
      As LLVM is also available for GPUs, porting our framework to this architecture should be possible.
      We even expect that barrier synchronization, which is important in the GPU context, can be integrated into our approach.

%
%       for a parallel combinator
%       that forks and joins threads in a syntactic block: instead of accessed memory addresses,
%       a thread could return a trace of memory accesses and synchronization points, and this trace
%       could be checked for race-freedom at the (syntactic) join point. Extending our idea to more
%       general thread creation operations like spawning arbitrary threads is expected to be more complicated.
%
%
%
%       will probably require a deeper
%       embedding of the LLVM semantics, loosing some of the advantages of the current shallow embedding.
%       Nevertheless,
%
%
%       fractional sep-logic,
%       state-of-the-art parallel sorting
%       time
%       extension to invariants, weak memory, etc.,
%

\clearpage

\bibliography{bib}

\end{document}

### 2022-03-08 papers/llvm_par/main_long.tex
new stuff

\documentclass[runningheads]{llncs}
%
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{balance}

\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}

%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
\usepackage{mathpartir}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{cprotect}

\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

\usepackage{multirow}
\usepackage{siunitx}

\usepackage{tikz}
\usetikzlibrary{positioning,fit,shadows,decorations, arrows, shapes, decorations.markings,decorations.pathmorphing,calc}

\usepackage{pgfplots}

\tikzset{
  -|-/.style={
    to path={
      (\tikztostart) -| ($(\tikztostart)!#1!(\tikztotarget)$) |- (\tikztotarget)
      \tikztonodes
    }
  },
  -|-/.default=0.5,
  |-|/.style={
    to path={
      (\tikztostart) |- ($(\tikztostart)!#1!(\tikztotarget)$) -| (\tikztotarget)
      \tikztonodes
    }
  },
  |-|/.default=0.5,
}

\clubpenalty = 10000
\widowpenalty = 10000
\displaywidowpenalty = 10000

\input{lstisabelle}
\lstset{captionpos=b}
\lstset{numberbychapter=false}
\lstset{autogobble}

\newcolumntype{Y}{>{\RaggedRight\arraybackslash}X} % Multiline column, automatic width

  \newcommand{\false}{\textrm{false}}
  \newcommand{\true}{\textrm{true}}

% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]

% \overfullrule=8pt

\begin{document}

\title{Refinement of Parallel Algorithms down to LLVM}

% \author{Peter Lammich}
% \institute{University of Twente}
\author{A.~Nonymous}
% \institute{Institute}
% \email{p.lammich@utwente.nl}
%
% \authorinfo{Peter Lammich}
%            {The University of Manchester, UK}
%            {peter.lammich@machester.ac.uk}

\maketitle

\begin{abstract}
  We present a stepwise refinement approach to develop
  verified parallel algorithms, down to efficient LLVM code.
  The resulting algorithms' performance is competitive with their
  counterparts implemented in C/C++.

  Our approach is backwards compatible with the Isabelle Refinement Framework,
  such that existing sequential formalizations can easily be adapted
  or re-used.

  As case study, we verify a simple parallel sorting algorithm, and
  show that it performs on par with its C++ implementation,
  and is competitive to state-of-the-art parallel sorting algorithms.
\end{abstract}

\section{Introduction}
Modern hardware has become faster by getting more parallel, while the
sequential execution speed has not significantly increased in the last years.
To keep up with this development, and fully utilize the capabilities of modern hardware,
parallel algorithms are required.

However, parallel algorithms come with new subtle classes of errors, like race conditions,
which are particularly hard to find by testing. Thus, formal methods that can guarantee
absence of all errors, including the hard-to-find ones, become even more important for
parallel algorithms.

In this paper, we present a stepwise refinement approach to develop
verified and efficient parallel algorithms. Our method can verify total correctness down to LLVM intermediate code.
The resulting verified implementations are competitive with state-of-the-art
unverified implementations. Our approach is backwards compatible to
the Isabelle Refinement Framework (IRF), a powerful tool to verify efficient sequential software,
such as model checkers~\cite{ELNN13,BrLa18,WiLa18}, SAT solvers~\cite{La17_CADE,La17_SAT,FBL18},
or graph algorithms~\cite{La14,LaSe16,LaSe19}.
This paper adds parallel execution to the IRF's toolbox, without invalidating the existing formalizations,
which can now be used as sequential building blocks for parallel algorithms, or be modified to add
parallelization.

As a case study, we verify total correctness of a parallel sorting algorithm,
re-using an existing verification of state-of-the-art sequential sorting algorithms~\cite{La20}.
Our verified parallel sorting algorithm is competitive to state-of-the-art parallel sorting algorithms,
at least on moderately parallel hardware.

% The main idea is to add a parallel combinator to an existing
% sequential LLVM semantics, by extending the thread semantics to record memory accesses, and
% check these for data-race freedom upon joining the parallel threads.
% Based on this idea, we extend the existing Isabelle Refinement Framework~\cite{La19-llvm}
% to support parallel algorithms. The resulting framework is largely backwards compatible to
% the original one, such that we could easily port existing algorithms and data structures,
% and use them as building blocks for parallel algorithms.

The contributions of this paper are as follows:
\begin{itemize}
  \item a semantics for LLVM that supports parallel execution.
    Up to our knowledge, this is the first formalization of parallelism for a practically usable LLVM semantics.

  \item a framework that supports a stepwise refinement approach to verify total correctness of parallel programs.

  \item a verified parallel sorting algorithm. Up to our knowledge, this is the
    first verification of a parallel sorting algorithm, let alone a competitive implementation.

\end{itemize}

The formalization and benchmark scripts have been submitted as (anonymized) supplementary material.

\subsection{Overview}\label{sec:intro_overview}
This paper is based on the Isabelle Refinement Framework (IRF), a continuing effort
to verify efficient implementations of complex algorithms, using stepwise refinement techniques.
In this paragraph, we give a brief overview of the components and versions of the IRF,
and explain how our work relates to them.
Figure~\ref{fig:IRF-overview} displays the components of the Isabelle Refinement Framework.

\begin{figure}
  \center
  \begin{tikzpicture}[node distance=.01\textwidth,outer sep=0pt,
  layerA/.style={
    draw=black,
    fill=blue!20,
    rounded corners,
    text width=.9\textwidth,
    font={\sffamily\bfseries\color{black}},
    align=center,
    minimum height=.03\textheight,
%    text height=12pt,
%    text depth=9pt
    },
  layerB/.style={layerA,fill=green!20},
  layerC/.style={layerA,fill=red!20},
  layerD/.style={layerA,fill=red!30,text width=.2\textwidth},
  ]

  \node[layerA] (AlgDS) {Algorithms and Data Structures};

  \node[layerB, below=of AlgDS.south west, anchor=north west, text width=.437\textwidth] (nres) {Nres-Monad};
  \node[layerA, right=of nres.north east, anchor=north west, text width=.437\textwidth, minimum height=.067\textheight] (llDS) {Low-Level Algorithms \\and Data Structures};

  \node[layerB, below=of nres.south west, anchor=north west, text width=.437\textwidth] (sepref) {Sepref};

  \node[layerC, below=of sepref.south west, anchor=north west] (sepl) {Program Logic and VCG};
  \node[layerC, below=of sepl.south west, anchor=north west, minimum height=.120\textheight] (sembox) {};

  \node[below right=of sembox.north west] {Back End Semantics};

  \node[layerD, right=of sembox.west] (isacode) {Isa Codegen};
  \node[layerD, right=of isacode] (imphol) {Imp HOL};
  \node[layerD, right=of imphol] (llvm) {LLVM};
  \node[layerD, right=of llvm] (llvm-par) {LLVM par};

  \node[layerD, below=of imphol] (impholtime) {Imp HOL time};
  \node[layerD, below=of llvm] (llvmtime) {LLVM time};

  \end{tikzpicture}

  \caption{Overview of the different components of the Isabelle Refinement Framework}\label{fig:IRF-overview}
\end{figure}

At the basis, there is a semantics of a back end language.
While the first version of the IRF used Isabelle's code generator to generate purely functional code~\cite{LaTu12,La13},
this has subsequently been extended to imperative functional code~\cite{La15} using Imperative HOL~\cite{BKHEM08}, and purely imperative LLVM code~\cite{La19-llvm}.
For the imperative back ends, also versions that can reason about run-time complexity are available~\cite{HaLa19,HaLa21,HaLa21-toplas}.
The work in this paper is based on the LLVM semantics without time~\cite{La19-llvm}, adding support for parallel programs.

On the next level, a program logic and verification condition generator is provided.
While this is trivial when reasoning about purely functional code in Isabelle/HOL,
the imperative versions use separation logic, and provide automation like frame-inference heuristics.
While the Imperative HOL versions of the IRF are based on a separation logic formalization
by Lammich and Meis~\cite{LaMe12}, the LLVM versions are based on a more general framework by Klein et al.~\cite{KKB12,KKB12_afp}.

While this reasoning infrastructure can already be used to verify simple low-level algorithms and data structures,
more complex developments are typically refined from a purely functional,
monadic representation of nondeterministic programs (nres-monad), originally introduced in~\cite{LaTu12}.
A typical development will do multiple refinement steps inside the nres-monad, and then use a
semi-automatic tool (Sepref) to refine from the nres-monad into the back end semantics.
While the nres-monad and associated tooling has largely stayed the same since it's first version~\cite{LaTu12},
the Sepref tool has been continuously adapted and improved with each new back end that was added.
Again, our work is based on the Sepref tool for LLVM without time~\cite{La19-llvm}.

We have also implemented parallelism for the Imperative HOL back end.
This was done as a feasibility study, and uses the exactly same ideas and
techniques as described in this paper for the more efficient LLVM back end.
While a description of this initial implementation would not add any new
ideas to this paper, it might still be useful in the context of older developments,
that have not yet been ported to the LLVM back end.

The rest of this paper is roughly organized along the structure of the Isabelle Refinement Framework:
We first describe our parallel LLVM semantics (Section~\ref{sec:parsem}) and separation logic~\ref{sec:seplogic}.
Then, we describe our changes to the refinement infrastructure (Section~\ref{sec:refine_parallel}).
Section~\ref{sec:parsort} describes our parallel sorting algorithm case study and benchmark results,
and Section~\ref{sec:concl} concludes the paper and discusses related and future work.

%
%
%
% Contributions
%   shallow embedding of parallel semantics
%
%   technical: formalization of parallel separation logic. TOTAL CORRECTNESS.
%   integration of parallel call into refinement framework. backwards compatible (re-use of seq).
%
%   technical: automatic refinement with concrete-identity tracking
%
%   formalization of simple parallel quicksort, using state-of-the-art sequential sort as subroutine
%
%   sample partitioning scheme, template instantiation by preprocessor

\section{Shallow Embedding of Parallel Semantics}\label{sec:parsem}
  Isabelle LLVM~\cite{La19-llvm} is a shallow embedding of a fragment of the LLVM semantics into the interactive theorem prover Isabelle/HOL~\cite{NPW02}.
  It comes with a set of tools to facilitate algorithm development, and has been successfully used to verify efficient \emph{sequential} algorithms~\cite{La20}.
  The shallow embedding of Isabelle LLVM is key to its lightweightness and flexibility. In particular, the control flow is embedded as a state-error monad,
  and registers are mapped to bound variables. Only the memory model is deeply embedded.

  We now describe an extension of Isabelle LLVM to support parallel algorithms,
  while keeping the shallow embedding. The resulting framework is backwards compatible to the
  original sequential Isabelle LLVM, such that porting existing verifications and tools is easy.
  The basic idea of our extension, which we detail in the following, is to support a parallel
  combinator \is{f || g}, that executes \is$f$ and \is$g$ in parallel, and fails if there is a possible data race.

  \subsection{Memory Model}
  We define a simple memory model\footnote{The original formalization~\cite{La19-llvm} also supports pointers into tuples.
  While complicated to implement, this feature has not been used in practice. To avoid additional complications
  with conflicting memory accesses from different but overlapping addresses, we dropped this feature.
  Now, each memory location is uniquely identified by an address, greatly simplifying the definition of data race.},
  that supports arrays of values, where values can be pointers into arrays, integers, or tuples:
  \begin{lstlisting}
    datatype addr = ADDR (block: nat) (idx: int)
    datatype ptr = PTR_NULL | PTR_ADDR (the_addr: addr)
    datatype val = LL_STRUCT "val list"  |  LL_INT lint  |  LL_PTR ptr
    typedef memory = val list option list
  \end{lstlisting}
  where \is{\<alpha> option = None | Some \<alpha>} is Isabelle's option type.

  Here, a memory is a list of blocks, where each block is either freed (\is{None}), or a list of values.
  Newly allocated blocks are simply appended to the end of the block list.
  An address consists of a block index and a value index, and a pointer is either null, or an address.
  Note that our LLVM semantics does not support conversion of pointers to integers, nor comparison
  or difference of pointers to different blocks. This way, a program cannot see the internal representation of a pointer,
  and we can choose a simple abstract representation and deterministic allocation function, while being
  faithful wrt.\ any actual memory manager implementation.

  The control flow is embedded into a state-error monad, which additionally records the memory addresses accessed by a computation\footnote{For technical reasons, we distinguish nontermination (\is{NTERM}) from errors (\is{FAIL}).}:
  \begin{lstlisting}
    datatype acc = MACC (reads: "addr set") (writes: "addr set")
    datatype \<alpha> mres = NTERM | FAIL | SUCC \<alpha> acc memory
    datatype \<alpha> M = M (run: "memory => \<alpha> mres")
  \end{lstlisting}
  Here, a memory access (\is{acc}) records the set of read and written addresses.
  A monadic result (\is{\<alpha> mres}) describes either a nonterminating (\is{NTERM}) or failed (\is{FAIL})
  computation, or a successful computation (\is{SUCC x a m}), with result \is{x}, accessed addresses \is{a}, and new memory content \is{m}.
  A program (\is{\<alpha> M}) is a function from an initial memory to a monadic result.

  For memory accesses, we define the operations \is$0$ and \is$(+)$, and show that they form a commutative monoid:
  \begin{lstlisting}
    0 == MACC {} {}"
    MACC r_1 w_1 + MACC r_2 w_2 == MACC (r_1 \<union> r_2) (w_1 \<union> w_2)"
  \end{lstlisting}
  where \is{==} indicates defining equations.
  Moreover, we extend the $(+)$ operation to monadic results:
  \begin{lstlisting}
    (SUCC x a s) + a' == SUCC x (a+a') s
    FAIL + a == FAIL
    NTERM + a == NTERM
  \end{lstlisting}

  The basic monad combinators are \is{return x}, which returns a result without accessing any memory,
  and bind (\is{x<-m; f x}), which first executes \is{m}, binds its result to variable \is{x}, and then executes \is{f x}.
  They are defined as follows:
  \begin{lstlisting}
    return x == M (SUCC x 0)
  \end{lstlisting}%
  \vspace*{-1em}
  \begin{lstlisting}
    x<-m; f x == M (%s. case run m s of
      SUCC x a s => (run (f x) s) + a
    | NTERM => NTERM
    | FAIL => FAIL)
  \end{lstlisting}
  Here, the return operation makes no memory accesses ($0$),
  and the bind operation adds the memory accesses of both parts.
  If the second part does not depend on the result of the first part,
  we write \is{m_1; m_2} as shortcut notation for \is{x<-m_1; m_2}.

  Using the accessed memory, we can define a parallel combinator \is{m_1 || m_2},
  which executes the programs \is{m_1} and \is{m_2} in parallel, and returns a pair of the results.
  When there is a data race, the combinator returns \is{FAIL}.
  \begin{lstlisting}
    m_1 || m_2 == M (%s. case run m_1 s of --"\rightcomment{run $m_1$}"
      SUCC x_1 a_1 s => (case run m_2 s of --"\rightcomment{$m_1$ succeeds, run $m_2$}"
        SUCC x_2 a_2 s => if conflict a_1 a_2 then FAIL --"\rightcomment{$m_2$ succeeds, check race}"
                          else SUCC (x_1,x_2) (a_1+a_2) s --"\rightcomment{no race, add mem acc}"
      | FAIL => FAIL | NTERM => NTERM) --"\rightcomment{$m_2$ fails or diverges}"
    | FAIL => FAIL | NTERM => NTERM) --"\rightcomment{$m_1$ fails or diverges}"
  \end{lstlisting}
  where
  \begin{lstlisting}
  conflict a_1 a_2 == writes a_1 \<inter> accs a_2 = {}  \<and>  writes a_2 \<inter> accs a_1 = {}
  accs a == reads a \<union> writes a
  \end{lstlisting}

  Note that we simply execute the programs sequentially on the given memory.
  As the programs only access disjoint addresses, the order of execution is irrelevant for the result
  --- as long as the programs do not allocate new memory:
  in this case, the allocated addresses are different, depending on the order of execution.
  In particular, we cannot prove \is{par m_1 m_2 = swap(par m_2 m_1)}, where \is{swap} swaps the elements of the result tuple.
  However, we argue that our formalization is still correct, as the details of pointers,
  like block addresses, are not visible to the program.

  Formally, this could be proved as follows:
  we define two states to be similar
  (\is{s_1 $\approx$ s_2}), iff there exists a bijection \is$f$ on block indexes, such that \is{f s_1 = s_2}.
  We lift $\approx$ to monadic results (\is{mres}), identifying \is{FAIL} and \is{NTERM}.
  We call a program \emph{well-formed}, if it behaves similar on similar states,
  only depends on the addresses reported as read, and only modifies the addresses reported as written.
%   \begin{lstlisting}
%     well_formed m ==
%          !s_1 s_2. s_1 \approx s_2 ==> run m s_1 \approx run m s_2
%     \and !s x a s'. run m s = SUCC x a s' \and s_2 = s on a
%   \end{lstlisting}
  Well-formedness syntactically extends over the basic monad operations. For well-formed programs,
  we can actually prove \is{par m_1 m_2 $\approx$ swap(par m_2 m_1)}.

  We leave a mechanised formalization of the above sketch to future work.

%   While it is possible to rigorously formalize the above sketch in Isabelle, and it would give a nice sanity check of our semantics,
%   we believe that the property is intuitive enough to leave such an endeavour to future work.

  \subsection{Extending the LLVM Backend}
  While the parallel operator gives us a semantics for parallel execution,
  it is difficult to map it to LLVM in its full generality, as this would mean to parallelly
  execute two parts of the same procedure, including all captured registers.
  Thus, our actual LLVM semantics only allows for a slightly more restricted parallel operator,
  which calls two single-argument functions in parallel, and returns the pair of their results:
  \begin{lstlisting}
    llc_par f g a b == par (f a) (g b)
  \end{lstlisting}
  where \is{f} and \is{g} must be constants referring to LLVM functions.

  We have extended the code generator to map \is{llc_par} to an external \cs{parallel()} function with the following signature:
  \begin{lstlisting}[language=C++]
    void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2)
  \end{lstlisting}
  Its semantics is to perform the calls \cs{f1(x1)} and \cs{f2(x2)} in parallel.
  Here, the arguments \cs{x1} and \cs{x2} point to some stack-allocated memory
  containing the actual function arguments, and space for the results.
  The actual implementation of the \cs{parallel} function has to be provided by a (trusted) support library.
  For our experiments, we use a very concise C++ implementation, based on the Threading Building Blocks~\cite{TBB} library:
  \begin{lstlisting}[language=C++]
  void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2) {
    tbb::parallel_invoke([=]{f1(x1);}, [=]{f2(x2);});
  }
  \end{lstlisting}

  %TODO Explain how syntactic restrictions work.

%   Prepropessor!

  \section{Parallel Separation Logic}\label{sec:seplogic}
  In order to reason about programs with a parallel operator, we define a parallel separation logic~\cite{OH04}.
  For our simple parallel combinator that does not allow for communication between parallel threads,
  a simple parallel rule is sufficient:
  \[
    \frac{\{P_1\} ~c_1~ \{Q_1\}~~~\{P_2\} ~c_2~ \{Q_2\}}{ \{P_1 * P_2\} ~c_1 \parallel c_2~ \{Q_1 * Q_2\} }
  \]
  That is, two threads $c_1$ and $c_2$ that run on disjoint memory can be executed in parallel. In this section, we briefly sketch the separation logic framework for Isabelle LLVM~\cite{La19-llvm}, and describe its extension to parallel separation logic.

  \subsection{Separation Algebra}
  In order to reason about memory with separation logic, a standard approach is
  to define an abstraction function from the memory into a separation algebra.
  The separation algebra can abstract away from details of the memory model
  (like our choice of using a list of blocks), and introduce the necessary
  formalism to support partial memories. This abstraction layer allows to use two different
  formalizations of the memory model, a concise and simple \emph{concrete memory model} for the semantics,
  and an \emph{abstract memory model} that forms a separation algebra.
  Note that the abstract memory model is only used for reasoning, and does not contribute to the
  trusted code base of our LLVM semantics.

  In the following, we briefly sketch the concepts of separation algebra and our abstract memory model.
  For a more detailed explanation, we refer to \cite{CHY07,La19-llvm}.

  A separation algebra~\cite{CHY07} is a structure with a zero, a disjointness predicate $a\#b$,
  and a disjoint union $a+b$.
  Intuitively, elements describe parts of the memory.
  Zero describes the empty memory, $a\#b$ means that $a$ and $b$ describe disjoint parts of the memory,
  and $a+b$ describes the memory described by the union of $a$ and $b$.
  For the exact definition of a separation algebra, we refer to~\cite{CHY07,KKB12}. We note that
  separation algebras naturally extend over functions and pairs, in a pointwise manner.

  \begin{example}(Trivial Separation Algebra)
    We define the type \is{\<alpha> tsa = 0 | TRIV \<alpha>}. This forms a separation algebra where the operators are defined as follows:
    \begin{lstlisting}
      a # b == a=0 \or b=0      a + 0 == a     0 + b == b
    \end{lstlisting}
    Intuitively, this separation algebra does not allow for combination of contents, except if one side is zero.
    While it is not very useful on its own, the trivial separation algebra is a useful building block for
    more complex separation algebras.
  \end{example}

  For our memory model, we define the following abstraction function:
  \begin{lstlisting}
    \alpha :: memory -> (addr -> val tsa) \x (nat -> nat tsa)
    \alpha s == (\alpha$_m$ s, \alpha$_b$ s)
  \end{lstlisting}
  where
  \begin{lstlisting}
    \alpha$_m$ s a == if valid_addr a s then TRIV get a s else 0
    \alpha$_b$ s b == if valid_block b s then TRIV (block_size b s) else 0
    valid_addr (ADDR b i) s == b < |s| \and s!b ~= None \and i<|the (s!b)|
    get (ADDR b i) s == the (s!b) ! i
    valid_block b s == b < |s| \and s!b ~= None
    block_size b s == |the (s!b)|
  \end{lstlisting}
  where \is{|xs|} is the length of list \is{xs}, \is{xs!i} is the $i$th element, and \is{the (Some x) == x} is the selector function of the option type.
  That is, an abstract memory \is{\alpha s} consists of two parts:
  \begin{enumerate}
  \item \is{\alpha$_m$ s} is a map from addresses to the values stored there. It is used to reason about load and store operations.
  \item \is{\alpha$_b$ s} is a map from block indexes to the sizes of the corresponding blocks.
    It is used to ensure that one owns all addresses of a block when freeing it.
  \end{enumerate}

  We continue to define a separation logic: assertions are predicates over separation algebra elements.
  The basic connectives are defined as follows:
  \begin{lstlisting}
    false a == False     true a == True     \<box> a == a=0
    (P*Q) a == \<exists> a_1 a_2. a_1 # a_2 \and a = a_1 + a_2 \and P a_1 \and Q a_2
  \end{lstlisting}
  That is, the assertion \is{false} never holds and the assertion \is{true} holds for all abstract memories.
  The empty assertion \is{\<box>} holds for the zero memory, and the separating conjunction \is{P*Q}
  holds if the memory can be split into two disjoint parts, such that \is{P} holds for one,
  and \is{Q} holds for the other part. The lifting assertion \is{\<up>\<phi>} holds iff the Boolean value \is{\<phi>} is true:
  \begin{lstlisting}
    \<up>\<phi> == if \<phi> then \<box> else false
  \end{lstlisting}
  It does not depend on the memory, and is used to lift plain logical statements into separation logic.
  When clear from the context, we omit the \is{\<up>}-symbol, and just mix plain statements with separation logic assertions.
%   , e.g., \is{array xs p * i<|xs|}

  \subsection{Weakest Preconditions}
  We define a \emph{weakest precondition} predicate as usual:
  \begin{lstlisting}
    wp m Q s == case run m s of
      SUCC x a s' => Q x a s' | _ => False
  \end{lstlisting}
  That is, \is{wp m Q s} holds, iff program \is$m$ run on state \is$s$ succeeds with a result \is{x}, memory accesses \is{a},
  and new state \is{s'} that satisfy \emph{postcondition} \is$Q x a s'$.

  To set up a verification condition generator based on separation logic,
  the postcondition should be a separation logic assertion over the corresponding abstract state, i.e.,
  we want \is$Q x a s$ to have the form \is{(Q' x) (\alpha s)}, where \is{Q'} is a separation logic assertion that depends on the result.
  However, it is not clear how to incorporate the accessed addresses \is{a} into this form, as they have no meaning on the level of assertions.

  We observe that, for reasoning about parallel programs, it suffices to ensure that one thread does not access memory that is used by another, parallel thread.
  Moreover, we require the program to be
  well-formed in the sense that it only reports accessed memory that was allocated at the start
  of the program, or got newly allocated during the program's execution, and that an address can
  only transition from unallocated to allocated to freed. In particular, once freed, an address
  will never be re-used\footnote{Note that this property is already required for a (deterministic)
  sequential semantics. Otherwise, internals of the memory model implementation would be visible
  to the program via pointer comparisons. Giving newly allocated memory a unique block
  and restricting pointer comparison to pointers to the same block ensures that our memory model is
  sound wrt.\ any actual implementation.}.

  For our memory model, we define the set of allocated and freed addresses, and the newly
  allocated memory between two states:
  \begin{lstlisting}
    valid_addrs s == { a | valid_addr a s }
    freed_addrs s == { ADDR b i | b<|s| \and s!b = None }
    used_mem s == valid_addrs s \<union> freed_addrs s
    new_mem s s' == used_mem s' - used_mem s
  \end{lstlisting}
  Note that, for \is{freed_addrs}, we do not keep track of the block size of a freed block, and simply report all addresses with this block index as freed.
  We also define a predicate to state that memory only transitions from fresh to allocated to freed:
  \begin{lstlisting}
    valid_trans s s' == valid_addrs s \<subseteq> valid_addrs s' \<union> freed_addrs s'
                    \and freed_addrs s \<subseteq> freed_addrs s'
  \end{lstlisting}
  Finally, we define the \emph{weakest precondition with access restrictions} as follows:
  \begin{lstlisting}
    wpa A c Q s == wp c (%r a s'.
      Q r s' --"\rightcomment{postcondition holds}"
      \and accs a \<subseteq> allocated s \<union> new_mem s s' \and valid_trans s s' --"\rightcomment{well formed}"
      \and accs a \<inter> A = {}) s --"\rightcomment{no addresses in A accessed}"
  \end{lstlisting}
  That is, \is{wpa A c Q s} states that execution of program $c$ in state $s$ yields a result \is{r} and a new
  state \is{s'} that satisfies the postcondition \is$Q r s'$. Moreover, the execution is well-formed,
  and does not access any addresses in $A$.

  \subsection{Hoare-Triples}
  Equipped with a weakest precondition with access restrictions, we define a Hoare-triple as follows:
  \begin{lstlisting}
    STATE asf P s == \<exists>as. as ## asf \and \<alpha> s = as+asf \and P as

    ht P c Q == \<forall>s asf. STATE asf P s
                      ==> wpa (addrs asf) c (%r s'. STATE asf (Q r) s') s)
  \end{lstlisting}
  Intuitively, the predicate \is{STATE asf P s} specifies that the abstract memory \is$\<alpha> s$ can be
  split into a part \is$as$ and the given frame \is$asf$, such that \is{as} satisfies assertion $P$.
  A Hoare-triple specifies that for all states \is$s$ and frames \is$asf$,
  such that \is$P$ describes \is$s$ wrt.\ the frame (\is{STATE asf P s}), the program \is$c$ will succeed
  on state \is$s$, not using any addresses of the frame, and the result and new state will satisfy
  assertion \is$Q$ wrt.\ the original frame (\is{STATE asf (Q r) s'}).

  Using this notion of Hoare-triple, we can prove the standard Hoare rules, in particular
  the frame rule and parallel rule:
  \begin{lstlisting}
    ht P c Q ==> ht (P * F) c (% r. Q r * F)

    ht P_1 c_1 Q_1 \and ht P_2 c_2 Q_2
    ==> ht (P_1 * P_2) (par c_1 c_2) (%(r_1,r_2). Q_1 r_1 * Q_2 r_2)
  \end{lstlisting}

  \subsection{Verification Condition Generator}
  Based on our notion of weakest precondition and Hoare-triples, we implement a verification condition generator (VCG).
  Like the semantics and separation logic, also the VCG is shallowly embedded into HOL.
  For technical reasons\footnote{
    This form simplifies reasoning using Isabelle's standard tactics. In particular, for the Hoare-triple form,
    an application of a consequence rule may introduce a unification variable for the
    postcondition. This cannot depend on bound variables introduced later, rendering the goal unprovable.
  }, our VCG does not work on Hoare-triples, but on subgoals of the form:
  \begin{lstlisting}
    STATE asf P s \and ... ==> wpa A c Q s
  \end{lstlisting}
  It repeatedly normalizes the state, and then applies rules that decompose the command $c$.
  Normalization of the state includes moving lifted assertions to HOL assumptions, and eliminating existential quantifiers.
  For example, the goal
  \begin{lstlisting}
    STATE asf (\<exists>x. P_1 a x * P_2 x b * \<up>( i < n)) s ==> wpa A c Q s
  \end{lstlisting}
  becomes
  \begin{lstlisting}
    !!x. i<n \and STATE asf (P_1 a x * P_2 x b) s ==> wpa A c Q s
  \end{lstlisting}
  There are rules for decomposing the basic monad combinators:
  \begin{lstlisting}
    Q r s ==> wpa A (return r) s

    wpa A m (%x. wpa A (f x) Q) s ==> wpa A ({x <- m; f x}) Q s
  \end{lstlisting}
  Moreover, any Hoare-triple can be turned into a rule for the VCG.
  Using the frame rule, it also generates a frame inference proof obligation:
  \begin{lstlisting}
    ht P c Q
    \and STATE asf P' s \and P' |- P*F
    \and (!!r s. STATE asf (Q r ** F) s ==> Q' r s)
    ==> wpa (addrs asf) c Q' s
  \end{lstlisting}
  That is, if the VCG encounters a proof obligation for a command \is{c}, for which there is a Hoare-triple \is{ht P c Q},
  it will first generate a frame inference proof obligation from the current state \is{P'} to the precondition \is{P},
  use a simple heuristics to automatically infer the frame,
  and then continue to derive the original postcondition \is{Q'} from a state described by the postcondition \is{Q}
  of the Hoare-triple and the inferred frame \is{F}.

  \subsection{Hoare-Triples for Memory Operations}
  Finally, we prove Hoare-rules for all basic LLVM instructions, and register them with our VCG infrastructure.
  We display the rules for memory operations:
  \begin{lstlisting}
  |= {n~=0} ll_malloc TYPE(\<alpha>) n {%p. range {0..<n} (%_. init) p * b_tag n p}
  |= {range {0..<n} xs p * b_tag n p} ll_free p {%_. \<box>}
  |= {pto x p} ll_load p {%r. r=x * pto x p}
  |= {pto xx p} ll_store x p {%_. pto x p}
  \end{lstlisting}
  Here \is{b_tag n p} asserts that \is{p} points to the beginning of a block of size \is$n$, and
  \is{range I f p} describes that for all $i\in I$, $p+i$ points to value $f~i$.
  Intuitively, \is{ll_malloc} creates a block of size $n$, initialized with the default \is{init} value,
  and a tag. If one possesses both, the whole block and the tag, it can be deallocated by free.
  The rules for load and store are straightforward, where \is{pto x p} describes that $p$ points to value $x$.

  \section{Refinement for Parallel Programs}\label{sec:refine_parallel}
  At this point, we have described a separation logic framework for parallel programs in LLVM.
  It is largely backwards compatible with the framework for sequential programs described in~\cite{La19-llvm},
  such that we could easily port the data structures formalized there to our new framework.
  We also can verify simple parallel programs, using our VCG.

  The next step towards verifying complex programs is to set up a stepwise refinement framework.
  In this section we describe the refinement infrastructure of the
  Isabelle Refinement Framework (cf.~Section~\ref{sec:intro_overview}), focusing on our changes to support parallel algorithms.

  \subsection{Abstract Programs}
  Abstract programs are shallowly embedded. They are terms of type \is{\<alpha> nres}, representing the result of a nondeterministic stateless computation:
  \begin{lstlisting}
    \<alpha> nres == fail | spec (\<alpha> => bool)
  \end{lstlisting}
  Here, \is{fail} represents possible non-termination or assertion violation, and \is{spec P} represents
  a result nondeterministically chosen to satisfy predicate $P$.
  The \emph{refinement ordering} on \is{nres} is defined by:
  \begin{lstlisting}
    spec P <= spec Q == \<forall>x. P x ==> Q x         fail $\not\le$ spec Q         m <= fail
  \end{lstlisting}
  Intuitively, \is{m_1 <= m_2} means
  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
  Note that \is$<=$ is a complete lattice, with top element \is$fail$.
  The \emph{monad combinators} are defined as:
  \begin{lstlisting}
    return x  ==  spec y. y=x
    x <- spec P; f x  ==  |_|{f x | P x}         x <- fail; f  ==  fail
  \end{lstlisting}
  Here, \is{return x} deterministically returns \is{x}, and \is{x<-m; f x} chooses a result of \is$m$, binds it to \is{x},
  and then executes \is$f x$. If \is$m$ may fail, then the bind may also fail.
%   We write \is{ x<-m; f x} for \is{bind m (\<lambda>x. f x)}, and \is{m_1; m_2} for \is{bind m_1 (\<lambda>_. m_2)}.

  Arbitrary recursive programs can be defined via a fixed-point construction~\cite{Kr10}.
  An \emph{assertion} fails if its condition is not met, otherwise it returns the unit value:
  \begin{lstlisting}
    assert \<phi> == if \<phi> then return () else fail;
  \end{lstlisting}
  Assertions are used to express that a program \is{m} satisfies a specification with
  precondition \is$P$ and postcondition \is$Q$:
  \begin{lstlisting}
    m <= assert P; spec x. Q x
  \end{lstlisting}
  If the precondition is false, the right hand side is \is{fail}, and the statement trivially holds.
  Otherwise, $m$ cannot fail, and every possible result $x$ of $m$ must satisfy $Q$.

  \subsection{Specification of Sorting Algorithms}\label{sec:sortspec}
  The first step to verify a sorting algorithm is to specify the desired result.
  We specify an algorithm that sorts a list \is{xs} as follows:
  \begin{lstlisting}
    sort_spec xs == spec xs'. sorted xs' \and mset xs' = mset xs
  \end{lstlisting}
  where \is{mset xs} is the multiset of the elements in the list \is{xs}.
  That is, the resulting list \is{xs'} is sorted, and a permutation of the original list \is{xs}.

  For quicksort-like sorting algorithms, the main function is partitioning, which reshuffles a
  given list into two partitions, such that any element in the first partition is less than or equal
  to any element in the second partition. We specify a partitioning function that assumes at least four elements,
  and is guaranteed to return non-empty partitions. The result is returned as a reshuffled list \is{xs'},
  and the index \is{m} of the first element of the second partition:
  \begin{lstlisting}
    partition_spec xs = assert (4 <= length xs);
      spec (xs',m). mset xs' = mset xs \and 0<m \and m<|xs|
      \and (\<forall>i\<in>{0..<m}. \<forall>j\<in>{m..<|xs|}. xs'!i <= xs'!j)
  \end{lstlisting}

  Using the above specifications as building blocks, we can define a simple abstract sorting algorithm:
  \begin{lstlisting}
    sort_aux xs d ==
      if d=0 \or |xs|<100000 then { --"\rightcomment{check depth bound or size threshold}"
        sort_spec xs --"\rightcomment{sort with some (sequential) algorithm}"
      } else {
        (xs,m) <- partition_spec xs;  --"\rightcomment{partition}"
        with_split m xs (%xs_1 xs_2. { --"\rightcomment{work on partitions $xs_1$ and $xs_2$}"
          npar sort_aux sort_aux (xs_1,d-1) (xs_2,d-1) --"\rightcomment{recursively sort in parallel}"
        })
      }"
  \end{lstlisting}
  where
  \begin{lstlisting}
    with_split m xs f ==
      assert (m < |xs|);
      (xs_1,xs_2) <- f (take m xs) (drop m xs);
      assert (|xs_1| = m \and |xs_2| = |xs| - m);
      return (xs_1@xs_2)

    npar f g x y == r1 <- f x; r2 <- g y; return (r1,r2)
  \end{lstlisting}
  where \is{@} appends two lists.
  This algorithm is derived from Introsort~\cite{Muss97}, a quicksort with a depth bound to ensure $O(n\log n)$ worst-case complexity.
  If the recursion depth bound $d$ is reached, or the list size is below a certain threshold,
  we sort the list using some yet unspecified sorting algorithm. Otherwise, we partition the list
  and recursively sort the two partitions.

  Here, we have introduced two new combinators,
  which are hints for the following refinement steps: (1) the \is{with_split} combinator splits a given list into two parts,
  and invokes a function that can change the content of these two parts, but not their length. It then returns
  the concatenation of the two changed parts. It is a hint for the further refinement steps to implement the list as
  an array that is modified in-place. (2) the \is{npar} combinator simply executes two blocks sequentially.
  However, it is a hint to the subsequent refinement steps to refine to parallel execution.

  Using the tools of the Isabelle Refinement Framework, and some existing theories copied from~\cite{La19-llvm},
  it is straightforward to prove that this sorting algorithm is correct, i.e.:
  \begin{lstlisting}
    sort_aux xs d <= sort_spec xs
  \end{lstlisting}

  In the next sections, we will explain how to refine this two an actual parallel sorting algorithm in LLVM,
  re-using the existing sequential pdqsort algorithm~\cite{La19-llvm} for small or too deep partitions, and a sampling algorithm for finding good pivots for partitioning.

  \subsection{The Sepref Tool}
  The Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{nres}-monad,
  keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
  The synthesis is automatic, but usually requires some program-specific setup and boilerplate.
  In this section we briefly introduce the Sepref tool, focusing on our extensions.
  For a more detailed discussion, we refer the reader to~\cite{La15,La19-llvm}.

  The main concept of the Sepref tool is refinement between a functional program \is{m} in the nres monad, and an LLVM program \is{c},
  as expressed by the \is{hnr}-predicate\footnote{This definition is the same as in \cite{La19-llvm}, except that we added the \is{CP} parameter.}:
  \begin{lstlisting}
    "hnr \<Gamma> c \<Gamma>' R CP m ==
    m ~= fail ==> ht \<Gamma> c (%r. \<exists>x. \<Gamma>' * R x r * \<up>(return x <= m \and CP r))"
  \end{lstlisting}
  That is, either \is{m} fails, or for a memory described by \is{\<Gamma>}, the LLVM program \is{c} succeeds with concrete result \is{r},
  such that there is an abstract result \is{x}, the new memory is described by \is{\<Gamma>' * R x r}, and
  \is{x} is a possible result of \is$m$. Moreover, \is{CP r} holds for the concrete result \is{r}.
  Note that the refinement trivially holds for a failing abstract program. This makes sense, as we prove that
  the abstract program does not fail anyway. Moreover, during the refinement proof, we can assume that
  assertions actually hold, as stated by the following rule:
  \begin{lstlisting}
    ( \<phi> ==> hnr \<Gamma> c \<Gamma>' R CP m ) ==> hnr \<Gamma> c \<Gamma>' R CP (assert \<phi>; m)
  \end{lstlisting}

  \begin{example}(Refinement of lists to arrays)\label{ex:ref_array}
    We define abstract programs for indexing and updating a list:
    \begin{lstlisting}
      lget xs i == assert (i<|xs|); return xs!i
      lset xs i x == assert (i<|xs|); return xs[i:=x]
    \end{lstlisting}
    That is, these programs assert that the index is in bounds, and then return the accessed element (\is{xs!i})
    or the updated list (\is{xs[i:=x]}) respectively.

    The following assertion links a pointer to a list of elements stored at the pointed-to location:
    \begin{lstlisting}
      arr$_A$ xs p = range {0..<|xs|} (%i. xs!i) p
    \end{lstlisting}
    That is, for every $i<|xs|$, $p+i$ points to the $i$th element of $xs$.

    The following Isabelle-LLVM programs index and update an array:
    \begin{lstlisting}
      aget p i == ll_ofs_ptr p i; ll_load p
      aset p i x == ll_ofs_ptr p i; ll_store x p; return p
    \end{lstlisting}

    We can link the abstract and concrete programs by the following theorems:
    \begin{lstlisting}
      hnr (arr$_A$ xs p * idx$_A$ i ii) (aget p ii)
          (arr$_A$ xs p * idx$_A$ i ii) id$_A$ (%_. True) (lget xs i)
      hnr (arr$_A$ xs p * idx$_A$ i ii) (aset p ii x)
          (idx$_A$ i ii) arr$_A$ (%r. r=p) (lset xs i x)
    \end{lstlisting}
    That is, if the list $xs$ is refined by array $p$, and the natural number $i$ is refined
    by the fixed-width\footnote{We use Isabelle's word library here, which encodes the actual width as a type variable,
    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word $ii$ (\is{idx$_A$ i ii}),
    the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
    The resulting memory will still contain the original array. Note that there is no explicit precondition that the array access
    is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.

    The \is{aset} operation will return a pointer to an array that refines the updated list returned by \is{lset}.
    As the array is updated in place, the original refinement of the array is no longer valid.
    Moreover, the returned pointer \is{r} will be the same as the argument pointer \is{p}.
    This information is important for refining to parallel programs on disjoint parts of an
    array (cf.\ Section~\ref{sec::array_split}).

%     xxx: include pointer-preservation for list_set

  \end{example}

  Given refinement assertions for the parameters, and \is{hnr}-rules for all operations in a program,
  the Sepref tool automatically synthesizes an LLVM program from an abstract \is{nres} program.
  The tool tries to automatically discharge additional proof obligations, typically arising from
  translating arithmetic operations from unbounded numbers to fixed width numbers. Where automatic proof fails,
  the user has to add assertions to the abstract program to help the proof. We have only slightly changed the existing Sepref tool,
  mainly adding a heuristics to keep track of concrete pointer equalities.

  \subsection{Array Splitting}\label{sec::array_split}
  An important concept for parallel programs is to concurrently operate on disjoint parts of the memory,
  e.g., different slices of the same array. However, abstractly, arrays are just lists. They are updated
  by returning a new list, and there is no way to express that the new list is stored at the same address as the old list.
  Nevertheless, in order to refine a program that updates two disjoint slices of a list to one that updates disjoint
  parts of the array in place, we need to know that the result is stored in the same array as the input.
  This is handled by the \is{CP} argument to \is{hnr}, as illustrated in Example~\ref{ex:ref_array}.

  We now refine the abstract \is{with_split} combinator from Section~\ref{sec:sortspec} to arrays.
  The corresponding concrete LLVM operation is\footnote{LLVM does not support higher-order functions, such as \is{awith_split}. However, the preprocessor of our code generator can inline such functions, to obtain valid LLVM code for non-recursive functions.}:
  \begin{lstlisting}
    awith_split i a m ==
      a_2 <- ll_ofs_ptr a i;
      m a a_2;
      return a
  \end{lstlisting}
  The abstract and concrete functions are connected by the following rule:
  \begin{lstlisting}
    hnr (arr$_A$ xs_1 xsi_1 * arr$_A$ xs_2 xsi_2) (mi xsi_1 xsi_2) \<box>
        (arr$_A$ \x arr$_A$) (%(xsi_1',xsi_2'). xsi_1'=xsi_1 \and xsi_2' = xsi_2)
        (m xs_1 xs_2)"
    ==>
    hnr (arr$_A$ xs xsi * idx$_A$ n ni) (awith_split ni xsi mi)
        (idx$_A$ n ni) (%xs xsi. arr$_A$ xs xsi) (%xsi'. xsi'=xsi)
        (with_split n xs m)
  \end{lstlisting}
  That is, in order to refine an abstract \is{with_split n xs m} combinator,
  we have to refine the inner program \is{m} that operates on the split lists,
  in a way that the concrete pointers for the result (\is{xsi_1',xsi_2'}) are the
  same as the arguments (\is{xsi_1, xsi_2}). Thus, when the Sepref tool encounters a
  \is{with_split} combinator for a list that is refined by an array, it will apply the above rule,
  i.e., synthesize a refinement for the inner program, and try to prove that the return values are the
  same as the arguments.

  \subsection{Refinement to Parallel Execution}
  In Section~\ref{sec:sortspec}, we have used the nres-combinator \is{npar f g x y} as a
  mere abbreviation for executing \is{f x} and \is{g y} sequentially. In fact, for purely functional
  programs, there is no (semantic) difference between sequential or parallel execution,
  as long as one execution does not depend on the result of the other. For concrete programs that share memory,
  we have to additionally ensure that the two programs operate on disjoint parts of the memory.
  As disjointness is given by the separation conjunction, we get the following rule to refine the abstract \is{npar}
  to actual parallel execution:
  \begin{lstlisting}
    ${}$  hnr Ax (fi xi) Ax' Rx CP_1 (f x) \and hnr Ay (gi yi) Ay' Ry CP_2 (g y)"
    ==>
      hnr (Ax * Ay) (llc_par fi gi xi yi) (Ax' * Ay') (Rx \x Ry)
        (%(rxi,ryi). CP_1 rxi \and CP_2 ryi) (npar f g x y)"
  \end{lstlisting}
  That is, the Sepref tool will try to refine the argument functions first, and then synthesize a program
  using \is{llc_par} to execute them in parallel.

  \section{A Parallel Sorting Algorithm}\label{sec:parsort}
  Our version of the Sepref tool is mostly backwards compatible with the original Sepref tool for Isabelle-LLVM~\cite{La19-llvm}.
  The only difference is the handling of concrete pointer equalities (cf.\ Section~\ref{sec::array_split}).
  Thus, it was a mainly straightforward task to port the efficient verified Introsort and pdqsort algorithms~\cite{La20} to the new tool,
  additionally proving that the concrete algorithm is in-place, i.e., the returned array pointer equals the argument pointer.
  Equipped with these refinements, the Sepref tool can synthesize a parallel sorting algorithm from \is{sort_aux} (Section~\ref{sec:sortspec}), using the already existing partitioning algorithm and the
  pdqsort algorithm for the abstract \is{partition_spec} and \is{sort_spec} specifications.

  Actually, we used refinement steps on the abstract level to add two optimizations:
  first, we add an explicit parameter for the length of the list. This allows us to refine the list
  to just an array pointer\footnote{Alternatively, we could refine a list to a pair of array pointer and explicit length.}.
  Second, to prevent creation of parallel threads for small partitions, we proceed sequentially if the created partitions
  are too unbalanced. The following is the algorithm that we input into Sepref\footnote{We have omitted some
  of the assertions required to prove that the arithmetic operations do not overflow.}:
  \begin{lstlisting}
    par_sort_aux xs n d ==
      assert n=|xs|
      if d=0 \or n<100000 then {
        slice_sort_spec xs 0 n
      } else {
        (xs,m) <- slice_partition_spec xs 0 n;
        let bad = m<n div 8 \or (n-m < n div 8)
        (_,xs) <- with_split m xs (%xs_1 xs_2. {
          if bad then nseq par_sort_aux par_sort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
          else npar par_sort_aux par_sort_aux (xs_1,m,d-1) (xs_2,n-m,d-1)
        });
        return xs
      }"

    nseq f_1 f_2 x_1 x_2 == r_1 <- f_1 x_1; r_2 <- f_2 x_2; return (x_1,x_2)

    par_sort xs n ==
      assert n=|xs|
      if n<=1 then return xs
      else par_sort_aux xs n (log2 n * 2)

  \end{lstlisting}
  where \is{nseq} is abstractly the same as \is{npar}, but is translated to sequential execution by the Sepref tool.
  Moreover, \is{slice_sort_spec} and \is{slice_partition_spec} are the specifications for sorting and partitioning
  only a slice of the array, leaving the rest unchanged. These match the interface of the original sequential algorithms
  proved correct in~\cite{La20}. Finally, \is{par_sort} wraps the algorithm to run with an initial depth limit of $2\lfloor\log_2 n\rfloor$.

  \subsection{Final Implementation and Correctness Theorem}
  From this algorithm, the Sepref tool generates an imperative implementation \is{par_sort_impl} and proves the corresponding \is{hnr} refinement lemma.
  Combining this with the correctness proof of the abstract \is{par_sort} algorithm, and unfolding the definition of \is{hnr},
  we get the following Hoare-triple, that states correctness of our final algorithm:
  \begin{lstlisting}
  "ht (arr$_A$ xs xsi * idx$_A$ n ni * n = |xs|)
    (par_sort_impl xsi ni)
    (%r. r=xsi * \<exists> xs'. arr$_A$ xs' xsi * sorted xs' * mset xs' = mset xs)"
  \end{lstlisting}
  That is, for a pointer \is{xsi} to an array, whose contents are described by list \is{xs} (\is{arr$_A$}), and a fixed-size word \is{ni} representing the natural number \is{n} (\is{idx$_A$}),
  which must be the number of elements in the list \is{xs}, our sorting algorithm returns the original pointer \is{xsi}, and the array contents are now \is{xs'},
  which is sorted and a permutation of \is{xs}. The trusted code base (TCB) of this statement is our
  formalization of separation logic, Hoare-triples, \is{arr$_A$}, \is{idx$_A$}, the \is{sorted} and \is{mset} functions,
  as well as the LLVM semantics, the code generator, Isabelle's inference kernel, and, finally, the LLVM compiler.
  While this may seem a lot, we emphasize that it, in particular, does not include the nres-monad, the Sepref tool, the VCG,
  the Hoare-logic rules, or any high-level proof tactics or automated theorem provers. While all these components are actually used to create the proof,
  Isabelle's LCF style kernel ensures that they are not part of the TCB.

  \subsection{A Sampling Partitioner}
  While we could simply re-use the existing partitioning algorithm from the Introsort formalization,
  which uses a median-of-three pivot selection, we observe that the quality of the pivot is
  particularly important for a balanced parallelization. Moreover, the partitioning in the \is{par_sort_aux} procedure
  is only done for arrays above a quite big size threshold. Thus, we can invest a little more work to find
  a good pivot, which is still negligible compared to the cost of sorting the resulting partitions.

  We choose a rather simple sampling approach: for an array of size $n$, we take $\min(n,64)$ equidistant samples,
  and use the median of these samples as pivot.

  Formalization of this sampling algorithm in the Refinement Framework is straightforward.
  To easily integrate with the existing (highly optimized) implementation of the partitioning algorithm,
  we have to determine the \emph{index} of the pivot element, swap it with the first element of the array,
  and then use the existing partitioning algorithm.

  This suggests an algorithm to find the median of a list of sample indexes, comparing each index by looking it up in
  the list to be sorted, and comparing the corresponding elements.
  We briefly sketch our approach to formalize such parametrized algorithms using locales, and
  an extension to Isabelle-LLVM's preprocessor to allow for painless code generation.
  Note that, for the sake of readability, we have slightly idealized the presentation\footnote{
    We have omitted some of the fixed parameters, and only describe a parametrized locale.
    In practice, we have two versions of the locale, one with a parameter, and one without. While they are still based on the same abstract algorithms, they duplicate some boilerplate code.
  }.

  A \emph{locale}~\cite{KWP99} is a named context that fixes some parameters and makes assumptions about these parameters.
  Inside a locale, definitions and theorems can depend on the parameters and assumptions.
  Later, a locale can be instantiated with actual terms for the parameters, proving that they satisfy the assumptions.
  This makes available instantiated versions of the definitions and theorems proved inside the locale.
  Instantiations can also be nested, that is, a locale can be instantiated in the context of another locale.

  We define a locale to set up a parametrized compare function as follows:
  \begin{lstlisting}
    locale pcmp =
      fixes lt lti par$_A$ elem$_A$
      assumes weak_ordering (lt p)
      assumes hnr
        (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)
        (lti pi ai bi)
        (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True)
        (lt p a b)

  \end{lstlisting}
  that is, we fix an abstract compare function \is{lt}. It takes one extra parameter \is{p},
  and is a weak ordering\footnote{
    A weak ordering can be defined by mapping the elements into a total ordering, and is the standard prerequisite for sorting algorithms.
  } for each such \is{p}. We also fix a concrete compare function \is{lti}, that refines the abstract compare
  function, with refinement assertion \is{par$_A$} for the parameter, and \is{elem$_A$} for the elements to be sorted.

  Inside the \is{pcmp} locale, we define our sorting and median functions, threading through the explicit parameter \is{p}.
  Among others, this will yield a function \is{median p xs}, which returns the median index of the elements in list \is{xs},
  comparing wrt.\ \is{lt p}. We also get a corresponding implementation \is{median_impl}\footnote{We simply reused a sequential sorting algorithm to completely sort the list and then obtain the middle element. While this is not the most efficient way of finding a median, the overhead when sorting $64$ element arrays is negligible compared to the time spent for sorting the partitions that are at least $10^5$ elements large. Thus, a more efficient algorithm like quickselect would not have any measurable effect on the overall performance, and we leave its verification to future work.}.

  To implement the sampling algorithm, \is{inside} the \is{pcmp} locale, we define
  a comparison function for indexes, parametrized with the original parameter and the array to be sorted.
  For this, we instantiate the \is{pcmp} locale again\footnote{In the actual formalization, we use a slightly more complex locale hierarchy to carefully avoid infinite recursive instantiations.}:

  \begin{lstlisting}
    context pcmp begin

      definition "lt_idx (p,xs) i j == lt p (xs!i) (xs!j)"
      definition "lt_idx_impl == ..."

      sublocale IDX: pcmp lt_idx lt_idx_impl (par$_A$ \<times> arr$_A$) idx$_A$   \<proof>

  \end{lstlisting}
  This makes available a function \is{IDX.median} and its implementation, which compares indexes into arrays as desired.
  This is then used in the partitioning function.

  Finally, the \is{pcmp} locale is instantiated globally, for example to sort unsigned integers, and the resulting implementation is
  exported to LLVM text:
  \begin{lstlisting}
    global_interpretation unat: pcmp (<) ll_icmp_ult unat$^{64}_A$   \<proof>

    export_llvm unat.par_sort_impl
  \end{lstlisting}
  The \is{global_interpretation} command instantiates the locale for comparing natural
  numbers implemented by 64 bit words (\is{unat$^{64}_A$}). Note that the compare function is not
  parametrized. Finally, the \is{export_llvm} command generates LLVM text.
  It first uses a preprocessor to generate and prove correct a set of code equations, which are then pretty printed to actual LLVM text.
  The preprocessor does not contribute to the trusted code base, as it proves correct its transformations.
  At this point, we have added a transformation to automatically instantiate the higher-order parameters generated by Isabelle's locale mechanism.
  This saves a lot of boilerplate code that was necessary in the original sorting algorithm formalization~\cite{La20}, and became unmanageable
  once we added the sublocale for comparing indexes into an array.

%
%   xxx: array splitting
%
%   then short explanation that proofs are semi-automatic.
%
%
%   also mention optimizations of algorithm (bad-partition)
%
%
%
%   Sepref comes with standard setup to refine lists to arrays. List updates are refined to destructive array updates,
%   as long as the old version of the list is not used after the update.
%   It also provides setup to refine unbounded natural numbers to bounded integers.
%   It tries to discharge the resulting in-bounds proof obligations automatically.
%   If this is not possible, it relies on hints from the user.
%
%
%     - intro to sepref
%     - parallel hnr rule
%     - identity tracking
%     - array splitting

%   \section{Parallel Quicksort}
%   \subsection{The Algorithm}
%
%   \subsection{Array Data Structure}

  \subsection{Benchmarks}
  \newcommand{\plottim}[4]{
  \begin{tikzpicture}
    \begin{axis}[
      ymajorgrids,
      title={\large #1},
      title style={at={(0.15,.75)}},
      ybar=0pt,
      width=\textwidth,
      height=.25\textheight,
      bar width=3pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,}
      ,
      #3
    ]

      \input{#2}

      \addplot[black, thin, sharp plot,update limits=false]
	coordinates {([normalized]-10,1) ([normalized]100,1)};
% 	node[above] at (axis cs:1950,4.3e7) {Houses};
      #4
    \end{axis}
  \end{tikzpicture}
  }

  \begin{figure}
    \plottim{Laptop, uint64}{data_laptop_uint64}{xticklabels={}}{\legend{}}
    \plottim{Server, uint64}{data_server_uint64}{xticklabels={}}{\legend{}}
    \plottim{Laptop, string}{data_laptop_llstring}{xticklabels={}}{\legend{}}
    \plottim{Server, string}{data_server_llstring}{
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      ymax=1000,
      legend style = {
%         legend pos= north east,
        at = {(.87,.97)},
        cells={anchor=west},
        font=\scriptsize
      },
    }{}

  \caption{Runtimes in milliseconds for sorting various distributions of unsigned 64 bit integers and strings with
    C++ standard sorting algorithm, our verified parallel sorting algorithm and Boosts sample sort algorithm.
    The experiments are performed on a server machine with 22 AMD Opteron 6176 cores and 128GiB of RAM, and a laptop with a
    6 core (12 threads) i7-10750H CPU and 32GiB of RAM.
  }\label{fig:benchres}
  \end{figure}

  \begin{figure}

  \begin{tikzpicture}
    \begin{axis}[
      xlabel near ticks,
      legend style = {
        at = {(.87,.3)},
        cells={anchor=west},
        font=\scriptsize
      },
      ymajorgrids,
      title={\large Laptop},
      title style={at={(0.15,.85)}},
      width=\columnwidth,
      height=.4\textheight
    ]
      \input{sudata_laptop}

      \draw[black, thin, sharp plot]
      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});

%       \addplot[black, thin, sharp plot, update limits=false] {\x};
    \end{axis}
  \end{tikzpicture}

  \begin{tikzpicture}
    \begin{axis}[
      xlabel near ticks,
      legend style = {
        at = {(.87,.3)},
        cells={anchor=west},
        font=\scriptsize
      },
      ymajorgrids,
      title={\large Server},
      title style={at={(0.15,.85)}},
      width=\columnwidth,
      height=.4\textheight
    ]
      \input{sudata_server}

      \draw[black, thin, sharp plot]
      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});

%       \addplot[black, thin, sharp plot, update limits=false] {\x};
    \end{axis}
  \end{tikzpicture}
  \caption{Speedup of the various implementations, for sorting unsigned 64 bit integers with a random distribution, on
    a server with 22 AMD Opteron 6176 cores and 128GiB of RAM, and a laptop with a
    6 core (12 threads) i7-10750H CPU and 32GiB of RAM.
    The x axis ranges over the number of cores, and the y-axis gives the speedup wrt.\ the same implementation run on only one core.
    The thin black lines indicate linear speedup.
  }\label{fig:speedup}

  \end{figure}

  The resulting parallel sorting algorithm is quite simplistic,
  ignoring a lot of research on efficient parallel sorting algorithms~\cite{CNLM08},
  as well as practically well-tested implementations~\cite{boost-sort}.
  In this section, we report on benchmarking our verified algorithm against unverified algorithms.

  In a first experiment, we implemented the same simplistic algorithm directly in C++, and benchmarked it
  against its verified counterpart on various sets of distributions.
  The result was that both implementations have the same runtime, up to some minor noise.
  This indicates that there is no systemic slowdown: algorithms verified with our framework run as fast as their unverified counterparts implemented in C++.

  In a second experiment, we benchmark our verified algorithm against state-of-the-art implementations:
  we use the parallel \is{std::sort(execution_policy::par_unseq)} from the GNU C++ standard library~\cite{libstdc++}, and the
  \is{sample_sort} implementation from the Boost C++ libraries~\cite{boost}.
  We have benchmarked the algorithm on two different machines, and various input distributions. The results are shown in Figure~\ref{fig:benchres}.
  While our simple algorithm is clearly competitive for integer sorting on the less parallel laptop
  machine, it's slightly less efficient for sorting strings on the highly parallel server machine.
  Nevertheless, we believe that our simple verified algorithm is already useful in practice,
  and leave further optimizations to future work.

  In a third experiment, we measure the speedup that the algorithm achieves for a certain number of cores.
  The results are displayed in Figure~\ref{fig:speedup}. While the speedup on the moderately parallel
  laptop is comparable to the one of the C++ standard library, our simple algorithm
  achieves lower speedups than the state-of-the-art implementations on the highly parallel server.
  Again, we leave further optimization of our algorithm to future work.

%   Again, in particular for higher numbers of cores,
%   the speedup achieved by the state-of-the-art algorithms is better than for our simple algorithm,
%   highlighting some optimization potential.

%
%   Nevertheless, our verified algorithm is still competitive when compared to state-of-the-art implementations.
%
%   Nevertheless, while our simple algorithm cannot compete with the best practical implementations,
%   it is still significantly faster than sequential sorting in most cases.
%   Moreover, our verified implementation has the same performance as an unverified C++ implementation
%   of the same algorithm.
%
%   We have benchmarked our algorithm on standard laptop and server hardware,
%   for sorting 64 bit integers, with various input distributions.
%   The runtimes of our verified algorithm ($t_1$) and its unverified C++ implementation ($t_2$) is within 6\%
%   of each other ($.94 <= t_1/t_2 <= 1.06$).
%   Figure~\ref{fig:benchres} shows the results of comparing our verified parallel sorting algorithm against
%   C++ standard sequential sorting algorithm, and the state-of-the-art sample sort implementation from the Boost Libraries.
%   Sample sort s always faster than sequential sort, and faster than our verified algorithm in most cases.
%   However, except for one outlier, even our simple verified algorithm is significantly faster than the sequential algorithm.
%   Thus, we expect our verified algorithm to be useful in practice, and leave further optimizations to future work.

  \section{Conclusions}\label{sec:concl}
    We have presented a stepwise refinement approach to verify total correctness of efficient parallel algorithms.
    Our approach targets LLVM as back end, and there is no systemic efficiency loss in our approach
    when compared to unverified algorithms implemented in C++.
    As a case study, we have implemented a simple parallel sorting algorithm.
    It uses an existing verified pdqsort algorithm as a building block,
    and is competitive with state-of-the-art parallel sorting algorithms,
    at least on moderately parallel hardware.

    The main idea of our parallel extension is to shallowly embed the semantics of a
    parallel combinator into an existing sequential semantics, by making the existing
    semantics report the accessed memory locations, and fail if there is a potential data race.
    This only changed the lower levels of the existing tooling, while higher-level tools
    like VCG and Sepref remained largely unchanged and mostly backwards compatible.
    This allowed us to easily port the verification of already existing sequential algorithms
    to our parallel framework, where they could be used as building blocks for parallel algorithms.

    In our approach, the high-level algorithms are phrased as purely functional and sequential algorithms,
    that, however, already contain hints how they should be refined to imperative parallel algorithms.
    Only during the last refinement step, performed by Sepref, actual imperative and parallel code is generated.

    The trusted code base of our approach is relatively small.
    To the trusted code base of the original Isabelle-LLVM tool, we only added the semantics and code generation for the parallel combinator.

    \subsection{Related Work}
    While there is extensive work on parallel sorting algorithm (e.g.~\cite{CNLM08,AMI21}),
    there seems to be almost no work on their formal verification. The only work we are
    aware of~\cite{SaHu20} uses the VerCors deductive verifier to prove the permutation property (\is{mset xs' = mset xs}),
    but neither the sortedness property nor termination, of odd-even transposition sort~\cite{Ha72}.

    Concurrent separation logic is used by many verification tools such
    as VerCors~\cite{Vercors}, and also formalized in proof assistants, for example in the VST~\cite{VST}
    and IRIS~\cite{JKJA18} projects for Coq~\cite{BeCa10}. These formalizations contain elaborate concepts to
    reason about communication between threads via shared memory, and are typically used to
    verify subtle concurrent algorithms (e.g.~\cite{MeJo21}).
    However, they can only reason about partial correctness, and extending them to total correctness
    is a non-trivial endeavour that is subject of active research~\cite{SGGT21}.
    On the other hand, our (less expressive) separation logic naturally supports total correctness,
    and is sufficient for many parallel algorithms.

%     parallel sorting?
%
%     IRIS and VST: more expressive concurrent operators and sep-logic. Only partial correctness.
%       Focus on subtle concurrency, rather than simple parallelism but with subtly optimized sequential parts.

    \subsection{Future Work}
      An obvious next step for our work is to implement a fractional separation logic~\cite{BCOP05},
      to reason about parallel threads that share read-only memory. While our semantics
      already supports shared read-only memory, our separation logic does not.
      We believe that implementing a fractional separation logic will be straightforward,
      and mainly pose technical issues for automatic frame inference in our VCG and Sepref tool.

      Another obvious next step is to verify a state-of-the-art parallel sorting algorithm,
      like Boost's sample sort. As our simple algorithm, sample sort does not require
      advanced synchronization concepts, and can be implemented only with a parallel combinator.

      Finally, the Sepref framework has recently been extended to reason about complexity of (sequential)
      LLVM programs~\cite{HaLa21-toplas,HaLa21}. This line of work could be combined with our parallel extension, to verify the
      complexity (e.g. work and span) of parallel algorithms.

      Extending our approach towards more advanced
      synchronization like locks or atomic operations may be possible: instead of accessed memory addresses,
      a thread could return a set of possible traces, which are checked for race-freedom and then combined.
      However, this would require a nondeterministic LLVM semantics instead of the current deterministic one.

      Finally, our framework currently targets multicore CPUs. Another emerging architecture are general purpose GPUs.
      As LLVM is also available for GPUs, porting our framework to this architecture should be possible.
      We even expect that barrier synchronization, which is important in the GPU context, can be easily integrated into our approach.

%
%       for a parallel combinator
%       that forks and joins threads in a syntactic block: instead of accessed memory addresses,
%       a thread could return a trace of memory accesses and synchronization points, and this trace
%       could be checked for race-freedom at the (syntactic) join point. Extending our idea to more
%       general thread creation operations like spawning arbitrary threads is expected to be more complicated.
%
%
%
%       will probably require a deeper
%       embedding of the LLVM semantics, loosing some of the advantages of the current shallow embedding.
%       Nevertheless,
%
%
%       fractional sep-logic,
%       state-of-the-art parallel sorting
%       time
%       extension to invariants, weak memory, etc.,
%

\clearpage

\bibliographystyle{abbrv}
\bibliography{bib}

\end{document}

### 2022-03-18 thys/lib/IEEE_Floating_Point/document/root.tex
Fixed fmul_add semantics

\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{isabelle,isabellesym}

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

\parindent 0pt\parskip 0.5ex

\begin{document}

\title{A Formal Model of IEEE Floating Point Arithmetic}
\author{Lei Yu}
\maketitle

\begin{abstract}
  This development provides a formal model of IEEE-754 floating-point
  arithmetic. This formalization, including formal specification of the
  standard and proofs of important properties of floating-point arithmetic,
  forms the foundation for verifying programs with floating-point computation.
  There is also a code generation setup for floats so that we can execute
  programs using this formalization in functional programming languages. The
  definitions of the IEEE standard in Isabelle is ported from HOL Light
  \cite{harrison1997floating}.
\end{abstract}

\tableofcontents

% include generated text of all theories
\input{session}

\bibliographystyle{abbrv}
\bibliography{root}

\end{document}

### 2022-05-12 papers/ITP2022/main.tex
renaming files

--- 2022-03-08 papers/llvm_par/main.tex
@@
 \usepackage{float}
-\usepackage{wrapfig}
+% \usepackage{wrapfig}
 
@@
 
+\supplementdetails[subcategory={Isabelle Formalization}]{Software}{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}
+
 % \acknowledgements{I want to thank \dots}%optional
@@
 
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
+% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \EventEditors{John Q. Open and Joan R. Access}
+% \EventNoEds{2}
+% \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+% \EventShortTitle{CVIT 2016}
+% \EventAcronym{CVIT}
+% \EventYear{2016}
+% \EventDate{December 24--27, 2016}
+% \EventLocation{Little Whinging, United Kingdom}
+% \EventLogo{}
+% \SeriesVolume{42}
+% \ArticleNo{23}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\EventEditors{June Andronick and Leonardo de Moura}
 \EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventLongTitle{13th International Conference on Interactive Theorem Proving (ITP 2022)}
+\EventShortTitle{ITP 2022}
+\EventAcronym{ITP}
+\EventYear{2022}
+\EventDate{August 7--10, 2022}
+\EventLocation{Haifa, Israel}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\SeriesVolume{237}
+\ArticleNo{12}
 
@@
 
-\noindent Formalization available at: \url{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}
+% \noindent The formalization is available at \url{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}.
 
 \subsection{Overview}\label{sec:intro_overview}
-This paper is based on the Isabelle Refinement Framework (IRF), a continuing effort
+This paper is based on the Isabelle Refinement Framework, a continuing effort
 to verify efficient implementations of complex algorithms, using stepwise refinement techniques.
@@
 
-This basic program logic and VCG can already be used to verify simple low-level algorithms and data structures,
+At the level of the program logic and VCG, our framework can already be used to verify simple low-level algorithms and data structures,
 like dynamic arrays and linked lists. More complex developments typically use a stepwise refinement approach,
@@
   and then lift it to a state monad and add access reports.
-  Defining a nondeterminism-error monad is standard:
+  Defining a nondeterminism-error monad is straightforward in Isabelle/HOL:
   \begin{lstlisting}
@@
   Here, \is{return} does not change the state, and reports no accesses (\is$0$),
-  and \is{bind} sequentially composes the executions, and adds up the access reports.
-
-  Typically, the access report will contain read and written addresses, such that we can detect a data race.
+  and \is{bind} sequentially composes the executions, threading through the state $\mu$, and adding up the access reports \is$r_1$ and \is$r_2$.
+
+  Typically, the access report will contain read and written addresses, such that data races can be detected.
   Moreover, if parallel executions can allocate memory, we must detect those
@@
     (m_1 || m_2) \mu ==
-      (x_1,(r_1,\mu_1)) <- m_1 \mu; (x_2,(r_2,\mu_2)) <- m_2 \mu;  --"\rightcomment{execute both strands}"
+      (x_1,r_1,\mu_1) <- m_1 \mu; (x_2,r_2,\mu_2) <- m_2 \mu;  --"\rightcomment{execute both strands}"
       assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
@@
     get \mu (ADDR b i) == vals (\mu b) ! i
-    put \mu (ADDR b i) x == \mu(b := BLOCK ((vals (\mu b))[i:=x]))
+    put \mu (ADDR b i) x == \mu(b := ALLOC ((vals (\mu b))[i:=x]))
   \end{lstlisting}
@@
   \begin{lstlisting}
-    \mu' b = FRESH <--> if b \in a_2 \union f_2 then \mu_2 b = FRESH else \mu_1 b = FRESH
-    \mu' b = FREED <--> if b \in a_2 \union f_2 then \mu_2 b = FREED else \mu_1 b = FREED
-    is_alloc (\mu' b) <--> if b \in b \in a_2 then is_alloc (\mu_2 b) else is_alloc (\mu_1 b)
-    get \mu' addr = if addr \in w_2 \or addr.bidx \in a_2 \union f_2 then get \mu_2 addr else get \mu_1 addr
-  \end{lstlisting}
+    (1)  \mu' b = FRESH <--> \mu b = FRESH \and b ~: a_1 \union a_2
+    (2)  is_alloc (\mu' b) <--> (is_alloc (\mu b) \or b \<in> a_1 \union a_2) \and b ~: f_1 \union f_2
+    (3)  \mu' b = FREED <--> \mu b = FREED \or b \in f_1 \union f_2
+
+    (4)  a \in w_1 \or b \in a_1 ==> get_addr \mu' a = get_addr \mu_1 a
+    (5)  a \in w_2 \or b \in a_2 ==> get_addr \mu' a = get_addr \mu_2 a
+    (6)  a ~: w_1\<union>w_2 \or b ~: a_1\<union>a_2 ==> get_addr \mu' a = get_addr \mu a
+  \end{lstlisting}
+  The properties (1)--(3) define the state of blocks in the combined memory:
+  a fresh block in \is{\mu'} was fresh already in \is{\mu}, and has not been allocated (1);
+  an allocated block was already allocated or has been allocated, but has not been freed (2); and a freed block was already freed, or has been freed (3).
+  The properties (4)--(6) define the content: addresses written or allocated in the first or second execution get their content from \is{\mu_1} (4) or \is{\mu_2} (5) respectively.
+  Addresses not written or allocated at all keep their original content (6).
 
@@
     \alpha$_m$ \mu addr == if valid \mu addr then Some (get \mu addr) else 0
-    \alpha$_b$ \mu b == if is_alloc (\mu b) then TRIV (|vals (\mu b)|) else 0
+    \alpha$_b$ \mu b == if is_alloc (\mu b) then Some (|vals (\mu b)|) else 0
   \end{lstlisting}
@@
   \end{lstlisting}
-  It does not depend on the memory, and is used to lift plain logical statements into separation logic.
+  It is used to lift plain logical statements into separation logic assertions owning no memory.
   When clear from the context, we omit the \is{\<up>}-symbol, and just mix plain statements with separation logic assertions.
@@
   \begin{lstlisting}
-    ABS amf P \mu == \<exists>am. am ## amf \and \<alpha> \mu = am+amf \and P am
+    ABS amf P \mu == \<exists>am. am # amf \and \<alpha> \mu = am+amf \and P am
 
@@
   \begin{lstlisting}
-    Q r \mu ==> wpf amf (return r) \mu
+    Q r \mu ==> wpf amf (return r) Q \mu
     wpf amf m (%x. wpf amf (f x) Q) \mu ==> wpf amf ({x <- m; f x}) Q \mu
@@
     \and P' |- P*F --"\rightcomment{infer frame}"
-    \and (!!r \mu. ABS amf (Q r ** F) \mu ==> Q' r \mu) --"\rightcomment{continue with postcondition}"
+    \and (!!r \mu. ABS amf (Q r * F) \mu ==> Q' r \mu) --"\rightcomment{continue with postcondition}"
     ==> wpf amf c Q' \mu
@@
   |= {pto x p} ll_load p {%r. r=x * pto x p}
-  |= {pto xx p} ll_store x p {%_. pto x p}
+  |= {pto y p} ll_store x p {%_. pto x p}
   \end{lstlisting}
@@
   \subsection{Abstract Programs}
-  Abstract programs are shallowly embedded into the nondeterminism error monad \is{'a ne} (cf.\ Section~\ref{sec:monad}).
+  Abstract programs are shallowly embedded into the nondeterminism error monad \is{'a neM} (cf.\ Section~\ref{sec:monad}).
   They are purely functional, not modifying memory, or differentiating between sequential and parallel execution.
-  We define a \emph{refinement ordering} on \is{ne}:
+  We define a \emph{refinement ordering} on \is{neM}:
   \begin{lstlisting}
@@
 
-  For a detailed description on using the \is{ne} monad for stepwise refinement based program verification, we refer the reader to~\cite{LaTu12}.
+  For a detailed description on using the \is{ne}-monad for stepwise refinement based program verification, we refer the reader to~\cite{LaTu12}.
 
@@
   During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
-  The synthesis is automatic, but requires use annotations to the abstract program.
-
-  The main concept of the Sepref tool is refinement between an abstract program \is{c} in the \is{ne} monad,
+  The synthesis is automatic, but requires annotations to the abstract program.
+
+  The main concept of the Sepref tool is refinement between an abstract program \is{c} in the \is{ne}-monad,
   and a concrete program \is{c$_\dagger$} in the \is{M} monad, as expressed by the \is{hnr}-predicate:
@@
     "hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ==
-        c ~= fail ==> ht \<Gamma> c$_\dagger$ (%x. \<exists>x$_\dagger$. \<Gamma>' * R x x$_\dagger$ * \<up>(return x <= c \and CP x$_\dagger$))"
+        c ~= fail ==> ht \<Gamma> c$_\dagger$ (%x$_\dagger$. \<exists>x. \<Gamma>' * R x x$_\dagger$ * \<up>(return x <= c \and CP x$_\dagger$))"
   \end{lstlisting}
@@
   or for a memory described by assertion \is{\<Gamma>},
-  the LLVM program \is{c$_\dagger$} succeeds with \is{r$_\dagger$},
+  the LLVM program \is{c$_\dagger$} succeeds with \is{x$_\dagger$},
   such that the new memory is described by \is{\<Gamma>' * R x x$_\dagger$},
@@
   Given refinement assertions for the parameters, and \is{hnr}-rules for all operations in a program,
-  the Sepref tool automatically synthesizes an LLVM program from an abstract \is{ne} program.
+  the Sepref tool automatically synthesizes an LLVM program from an abstract \is{neM} program.
   The tool tries to automatically discharge additional proof obligations, typically arising from
@@
     A weak ordering is induced by a mapping of the elements into a total ordering. It is the standard prerequisite for sorting algorithms in C++~\cite{Josu12}.
-  }. Moreover, we have a concrete implementation \is{lt$_\dagger$} or the compare function, wrt.\ the refinement assertions
+  }. Note that the strict compare function \is{lt} also induces a non-strict version \is{le p a b == \<not>lt p b a}.
+  Moreover, we have a concrete implementation \is{lt$_\dagger$} of the compare function, wrt.\ the refinement assertions
   \is{par$_A$} for the parameter and \is{elem$_A$} for the elements.
 
-  Our sorting algorithm is developed and verified in the context of this locale (to avoid confusion, our presentation has, up to now, just used \is{<} and \is{sorted} instead of \is{lt p} and \is{sorted_wrt (lt p)}).
+  Our sorting algorithm is developed and verified in the context of this locale (to avoid confusion, our presentation has, up to now, just used \is{<}, \is{<=}, and \is{sorted} instead of \is{lt p}, \is{le p}, and \is{sorted_wrt (le p)}).
   To get a sorting algorithm for an actual compare function, we have to instantiate the locale,
@@
       width=\textwidth,
-      height=.25\textheight,
+      height=.249\textheight,
       bar width=3pt,
@@
     semantics report the accessed memory locations, and fail if there is a potential data race.
-    We only needed to changed the lower levels of our existing framework for sequential LLVM~\cite{La19-llvm}.
+    We only needed to change the lower levels of our existing framework for sequential LLVM~\cite{La19-llvm}.
     Higher-level tools like the VCG and Sepref remained largely unchanged and backwards compatible.

### 2022-05-12 papers/ITP2022/sudata_laptop.tex
renaming files

\addplot+[color=red, mark color=black, mark options={fill=red}] coordinates {
  (1, 1.00)
  (2, 1.80)
  (3, 2.45)
  (4, 2.96)
  (5, 3.25)
  (6, 3.48)
  (7, 3.61)
  (8, 3.69)
  (9, 3.77)
  (10, 3.92)
  (11, 4.00)
  (12, 4.10)
};
\addlegendentry{verified};

\addplot+[color=black, mark color=black, mark options={fill=black}] coordinates {
  (1, 1.00)
  (2, 1.91)
  (3, 2.69)
  (4, 3.43)
  (5, 4.03)
  (6, 4.29)
  (7, 4.42)
  (8, 4.53)
  (9, 4.62)
  (10, 4.72)
  (11, 4.74)
  (12, 4.80)
};
\addlegendentry{sample sort};

\addplot+[color=blue, mark color=black, mark options={fill=blue}] coordinates {
  (1, 1.00)
  (2, 1.78)
  (3, 2.42)
  (4, 2.98)
  (5, 3.26)
  (6, 3.48)
  (7, 3.59)
  (8, 3.75)
  (9, 3.93)
  (10, 3.97)
  (11, 4.16)
  (12, 4.23)
};
\addlegendentry{std::sort(par-unseq)};

### 2023-03-23 papers/JAR_SI_ITP2022/ESOP22_reviews_before_rebuttal.txt
JAR SI ITP paper

--- 2022-01-19 papers/llvm_par/ESOP22_rebuttal.txt
@@
-Dear reviewers. Thank you for your useful reviews. We'll try to clarify the paper accordingly,
-and make it accessible to a larger audience by using less Isabelle-specific notation.
-With this rebuttal, we're trying to answer your questions and clarify misunderstandings.
-
-The paper finds a sweet spot between expressiveness and simplicity:
-the semantics is simple enough to be integrated into a well-proven scalable
-shallow-embedding approach, and expressive enough for interesting algorithms.
-We now can parallelize some complex and highly optimized algorithms, making
-them more efficient in practice. Instead of many toy examples, we decided for
-one substantial case study.
-If this is not enough contribution for ESOP, we accept that.
-
-@Reviewer A: the hnr predicate and sepref tool is generic. The formalized data structures are typically re-usable.
-The application-specific boilerplate is to set up synthesis of refined versions of the program, procedure by procedure,
-specifying what data structures to use, and making sure the program contains enough assertions to justify
-the refinement (e.g., that computations don't overflow).
-
-@Reviewer B: the main challenge was to find the sweet spot between expressiveness and compatibility with
-  existing techniques. Once found, implementing the tool became practical. As it allows re-use of sequential algorithms,
-  implementing a parallel sorting algorithm based on highly optimized sequential algorithms became feasible.
-
-@Reviewer C: unfortunately, the review is based on a bunch of misunderstandings.
-We will try to clarify the paper accordingly, to make it more readable for the Coq/IRIS community.
-
-Overclaiming:
-  1) we formalize a sufficiently large part of LLVM to generate practical code.
-  Adding additional features would be useless for our application, and unnecessarily increase the trusted code base.
-
-  2) we, indeed, were not aware of the Actris example. It verifies a simple, unoptimized, abstract algorithm, while
-    we verify an optimized concrete implementation. We'll relativise the claim.
-
-  3) [39] handles total correctness of *sequential programs*, and leaves parallel programs for future work.
-
-Comparison with IRIS:
-IRIS and the Isabelle Refinement Framework (IRF) have the same ultimate goal, providing verified software.
-While IRIS tries to be super-expressive, supporting every thinkable feature of concurrency,
-features to the IRF are added in an incremental way, driven by the need of verifying actual implementations.
-While IRIS, eventually, will be able to verify actual implementations of most real-world concurrent algorithms,
-the IRF can verify actual implementations of some important parallel algorithms now.
-
-Misunderstandings about what is verified:
-
-The separation logic rules/Hoare triples *are* proved/defined wrt. the underlying semantics.
-
-The translation to LLVM *is* verified, down to our concise LLVM semantics.
-This goes much further than the typical code extraction to a functional language,
-but not as far as tools like CompCert, that also verify the optimization and machine code generation.
-
-Using sorted and mset is Isabelle's established way of specifying a sorting algorithm.
-arrA and idxA specify the concepts of array / machine word. For correctness statements over low-level code,
-separation of concerns into natural concepts is essential for the readability of the specification.
-
-The parallel combinator, indeed, is asymmetric. However, as also reviewer C acknowledges,
-this approach is sound for the features that we consider. We are aware (and discuss in the paper) that
-extending the supported features will require adjustments. We note that the asymmetry could easily be avoided by
-identifying NTERM and FAIL already in the semantics.
-
-############### Submitted rebuttal end here #############################
-
-Rebuttal strategy:
-
-  Emphasize papers contributions again. (implicitly say:
-    the simplicity and ease of integration is the STRONG point, not the weak one.
-    The obtained result is decent!
-  )
-
-  Address reviewer 3:
-    state difference in approach between this paper and IRIS!
-      this paper: less general, but practical results
-      IRIS: more general, but only much less complex verified algorithms.
-
-    some of their claims are factually false! start with these.
-      adequacy, hoare-triples wrt underlying semantics, verify specification, total correctness iris
-
-    some claims are actually not severe (benign races, LLVM semantics, verify translation to LLVM, *first* verification of par-sort-algo, dubious semantics)
-
-  Then address other reviewers:
-    only one case study --> yes, but substantial one!
-    the other concerns: already addressed in paper, but explain again anyway.
-
-TCB:
-  also argument: even if a formalization in IRIS style is considered more elegant,
-  the TCB of my approach is still very small compared to other verification tools
-  like deductive verifiers.
-
-Ideas for rebuttal:
-
-General: this paper finds a sweet spot, where a parallel semantics can be formalized as a simple extension to a sequential one.
-  * the resulting semantics is simple, and thus easy to manually verify.
-  * it's backwards compatible to the existing sequential formalization
-  * and expressive enough for realistic examples
-
-In summary: parallelism for cheap
-
-  * existing tools and formalizations for sequential semantics can easily be adapted.
-    If there is a substantial body of sequential verifications, this backwards compatibility is essential,
-    as it allows reuse and thus high (efficiency) gain at low (formalization) cost.
-  * while this approach is less general, it's still expressive enough for practical relevant examples
-
-In the reviews, I see these arguments being turned around and used against the paper.
-
-case study:
-  decided to go for one substantial case study rather than many small toy examples.
-  The case study shows how the extension seamlessly integrates into the already existing architecture,
-  allowing to parallelize existing verified algorithm at little effort, but with great speed-up effect.
-
-LLVM semantics:
-  the presented framework goes down to a fragment of LLVM that is big enough for code generation.
-  Formalizing anything more of LLVM in this context would
-    1) not be necessary.
-    2) even be disadvantageous, as it would increase the TCB
-  Note that formalizing LLVM optimization passes is an entirely orthogonal endeavour (see VeLLVM project), and
-  this project would immediately profit from a verified LLVM backend.
-  I will tone down the statement, to clarify that only a fragment has been verified.
-
-Read-only memory:
-  the semantics already supports it. Just extending the separation logic is,
-  actually, straightforward and it's well known how to do it.
-  Extending the automation of Sepref accordingly would be less straightforward.
-
-  The contribution of this would mainly focus on technical details of how Sepref works,
-  and how to adjust it to fractional separation logic, and the practical gain would be
-  to perhaps be able to include a toy-example matrix multiplication algorithm.
-  Admittedly, I hoped that the contribution of this paper, including showing how to parallelize
-  a realistic sorting algorithm rather than toy examples, would be sufficient.
-
-#1: benchmark: indeed, industrial strength parallel sorting algorithms are more optimized than this one,
-  and I have also included such an algorithm in the benchmarks.
-  Going the other way round, i.e., starting at an industrial strength sorting algorithm and working upwards, was done for the sequential pdqsort algorithm (that I use as a subroutine).
-  For this paper, while trying to use a realistic example, I also had to keep it simple enough such that it's still suited as
-  an example.
-
-In the description of the Sepref tool, you mention that "The synthesis is automatic, but usually requires some program-specific setup and boilerplate".
-I'm not sure I fully understand what this program-specific setup is.
-Are you referring to the refinement assertions and hnr rules for the program operations?
-Do these have to be provided by the user?
-I would have thought some of them are fairly generic.
-Could you clarify everything one would have to provide in order for the sorting algorithm to go through?
-  hnr and the rules for the program operations are generic.
-  What is required is some setup to tell Sepref which concrete data structures to use for the abstract data types in the program, and to ensure the abstract program contains enough assertions
-  such that Sepref can prove the refinement (e.g., indexes are in bounds, numbers must fit into 64 bit words). These are program specific, and require some boilerplate code to set up.
-
-  Sorting algorithm: enough assertions in the abstract algorithm to justify the refinement of lists to arrays and mathematical integers to 64 bit words,
-  plus boilerplate to synthesize the concrete algorithm from the abstract one, subroutine by subroutine.
-
-#2
-"Still, I think that the contribution of this paper is too narrow for ESOP.
-It only considers one algorithm, and it seems that only small extensions of the existing components were actually needed: including disjoint parallelism construct in LLVM syntax and semantics, standard disjoint parallelism rule in separation logic, extensions of the Sepref tool.
-The latter part was the not sufficiently clear to me, and I was not convinced it will work for other examples."
-
-That's actually the point of the contribution: only small extensions, in particular to the trusted code base,
-are needed to get a great effect, that also integrates well with existing formalizations, and thus allows great speed-up effects by parallelization at low effort!
-
-The approach will work for all algorithms that can be phrased in terms of a parallel combinator on disjoint memory.
-I acknowledge that there may be other "refinement idioms" that are useful for different algorithms,
-but these can be straightforwardly added when needed, without changing anything at the base of the formalization.
-
-Questions for authors’ response
-What were the key technical challenges to overcome in this verification effort?
-> integrating the extension to parallelism without breaking the existing framework nor bloweing up the TCB too much.
-Thus, existing formalizations can be re-used, and parallelism can be added cheaply.
-
-In particular, it is expected that completely disjoint parallelism is captured as a sequential computation. Then, what is needed here beyond syntactic race detection?
-> Semantic race detection. E.g., indexes of array accesses can, typically, not be proved disjoint syntactically.
-  When syntactic race detection can be used to justify disjointness: good, probably the disjointness proof will be easy!
-
-#3
-
-Overclaims.
-  The definition of the parallel combinator is dubious (page 5).
-  The trusted computing base (TCB) is too large.
-
-  Actually, the TCB is only slightly extended (as compared to the TCB for sequential algorithms):
-    the extension to the semantics is straightforward, and explainable with a straightforward and obvious argument.
-    It also maintains the shallow embedding of control flow, which results in an elegant and concise semantics.
-
-    While a semantics that accounts for interleaving or even more sophisticated memory models would
-    work for more general programs (with IO, sync, etc), it would loose the
-    shallow embedding property, and be inherently more complicated, and thus harder to check.
-
-Page 2 says this paper presents "the first verification of a parallel sorting algorithm". See Actris: Session-Type Based Reasoning in Separation Logic (POPL 2020).
-   granted, I did not know this work. However: their sorting algorithm is a toy example, which they explicitly do not optimize for efficiency, nor can they extract efficient code.
-   I'll reference the paper and weaken the claim.
-
-Page 22 says "total correctness is a non-trivial endeavour that is subject of active research [39]".
-But it is unfair.
-[39] indeed provides a way to prove total correctness, and they can prove termination using almost the same idea employed in this paper, i.e., using the decreasing index.
-  they have not (yet) done it for parallel programs, only for sequential programs. And at the cost of a much more complicated semantics (e.g. TCB).
-
-The definition of the parallel combinator is dubious (page 5).
-  it is OK for the proposed (simple) fragment. And as I suggest in the paper, will not work for more complicated semantics with sync/communication.
-  The distinction between NTERM and FAIL is mainly for historical reasons. The Hoare-triple identifies them anyway, so no unsoundness is to be expected here.
-
-The current definition doesn't support benign races.
-  no, this was a deliberate decision, as this was not required. Adding such support is easily possible by keeping track of writes as addresses and values.
-  This would slightly increase the complexity of the semantics.
-
-The underlying separation logic's adequacy should be proved as in Iris [18].
-  The separation logic is not part of the TCB. All rules are actually proved.
-
-The meaning of Hoare triples should be defined w.r.t. the underlying semantic model as in Iris [18].
-  It actually is!
-
-The arrA, idxA, sorted, mset functions should be verified.
-  Using "sorted" and "mset" is the standard Isabelle way of specifying a sorting algorithm!
-  Other standard specifications can easily be derived from them.
-
-  The relation arrA defines the concept of an array in memory, mapping it to a list of elements.
-  And idxA defines the concept of a fixed-bit word, mapping it to a mathematical integer.
-  While it might be possible to merge their definitions with the definition of sortedness, the resulting
-  specification would be much harder to read. It was a deliberate decision to split the specification into
-  natural, and easy to verify parts, rather than to rely on a monolithic and complex specification.
-
-The translation to LLVM should be verified.
-  It is verified down to the shallowly embedded semantics of LLVM.
-  As any translation can only be verified against a semantics of the target, and the
-  shallowly embedded semantics used in this paper is concise and elegant, and I don't get the point.
-  Formalizing more complete and complicated deeply embedded semantics of LLVM, and translating from
-  the shallow embedding to these deep embedding is an orthogonal issue.
-
-Questions for authors’ response
-In my eyes, many of the claimed contributions are actually already made by prior work,
-for the reasons I describe above. Would you please elaborate on your contributions
-so that I can better judge this paper?
-
-  see above.
-
 Overall merit
@@
 The translation to LLVM should be verified.
-
 Questions for authors’ response
-In my eyes, many of the claimed contributions are actually already made by prior work,
-for the reasons I describe above. Would you please elaborate on your contributions
-so that I can better judge this paper?
+In my eyes, many of the claimed contributions are actually already made by prior work, for the reasons I describe above.
+Would you please elaborate on your contributions so that I can better judge this paper?

### 2023-03-23 papers/JAR_SI_ITP2022/main.tex
JAR SI ITP paper

--- 2022-05-12 papers/ITP2022/main.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles.
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-%
-% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
+% TODO:
+%   try to prevent page breaks for shorter listings
+%   update website, prominently link to html-thys covered in this paper
+
+% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[sn-mathphys,Numbered]{sn-jnl}
 
@@
 \usepackage[utf8]{inputenc}
-\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
-\usepackage{balance}
-
-\usepackage{amsmath}
-\usepackage{amssymb}
-% \usepackage{amsthm}
+
+\usepackage{graphicx}%
+\usepackage{multirow}%
+\usepackage{amsmath,amssymb,amsfonts}%
+\usepackage{amsthm}%
+\usepackage{mathrsfs}%
+\usepackage[title]{appendix}%
+\usepackage{xcolor}%
+\usepackage{textcomp}%
+\usepackage{manyfoot}%
+\usepackage{booktabs}%
+\usepackage[boxed]{algorithm}%
+\usepackage{algorithmicx}%
+\usepackage[noend]{algpseudocode}
+\usepackage{listings}%
+
+% \usepackage[T1]{fontenc}
+% \usepackage[utf8]{inputenc}
+% \usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
+% \usepackage{balance}
 
 %\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
-\usepackage{mathpartir}
-\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
-\usepackage[boxed]{algorithm}
-\usepackage{enumerate}
+% \usepackage{mathpartir}
+% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
+% \usepackage[boxed]{algorithm}
+% \usepackage{enumerate}
 \usepackage{cprotect}
 
-\usepackage{listings}
 \usepackage{lstautogobble}
-\usepackage{graphicx}
 \usepackage{tabularx}
-\usepackage{booktabs}
 \usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
-\usepackage{color}
-\usepackage[noend]{algpseudocode}
 \usepackage{caption}
 \usepackage[font=scriptsize]{subcaption}
-\usepackage{hyperref}
-\usepackage{float}
+% \usepackage{hyperref}
+% \usepackage{float}
 % \usepackage{wrapfig}
 
-\usepackage{multirow}
-\usepackage{siunitx}
+% \usepackage{multirow}
+% \usepackage{siunitx}
 
@@
 
-% \theoremstyle{definition}
-% \newtheorem{example}{Example}[section]
-
-% \overfullrule=8pt
-
-\title{Refinement of Parallel Algorithms down to LLVM}
-
-% \author{Peter Lammich}
-% \institute{University of Twente}
-% \author{A.~Nonymous}
-% \institute{Institute}
-% \email{p.lammich@utwente.nl}
-%
-% \authorinfo{Peter Lammich}
-%            {The University of Manchester, UK}
-%            {peter.lammich@machester.ac.uk}
-
-\author{Peter Lammich}{University of Twente, Netherlands}{p.lammich@utwente.nl}{https://orcid.org/0000-0003-3576-0504}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional.
-Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
-
-\authorrunning{Peter Lammich} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Peter Lammich} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\ccsdesc[500]{Software and its engineering~Formal software verification}
-\ccsdesc[500]{Theory of computation~Semantics and reasoning}
-\ccsdesc[300]{Computing methodologies~Parallel algorithms}
-
-% \ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-\keywords{Isabelle,Concurrent Separation Logic,Parallel Sorting,LLVM} %TODO mandatory; please add comma-separated list of keywords
-
-\supplementdetails[subcategory={Isabelle Formalization}]{Software}{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}
-
-% \acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \EventEditors{John Q. Open and Joan R. Access}
-% \EventNoEds{2}
-% \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-% \EventShortTitle{CVIT 2016}
-% \EventAcronym{CVIT}
-% \EventYear{2016}
-% \EventDate{December 24--27, 2016}
-% \EventLocation{Little Whinging, United Kingdom}
-% \EventLogo{}
-% \SeriesVolume{42}
-% \ArticleNo{23}
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\EventEditors{June Andronick and Leonardo de Moura}
-\EventNoEds{2}
-\EventLongTitle{13th International Conference on Interactive Theorem Proving (ITP 2022)}
-\EventShortTitle{ITP 2022}
-\EventAcronym{ITP}
-\EventYear{2022}
-\EventDate{August 7--10, 2022}
-\EventLocation{Haifa, Israel}
-\EventLogo{}
-\SeriesVolume{237}
-\ArticleNo{12}
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}%
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
+
+\theoremstyle{definition}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
+
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}%
+
+\raggedbottom
 
@@
 
-\maketitle
-
-\begin{abstract}
+\title[Refinement of Parallel Algorithms down to LLVM]{Refinement of Parallel Algorithms down to LLVM}
+\subtitle{applied to practically efficient parallel sorting}
+
+\author*[1]{\fnm{Peter} \sur{Lammich}}\email{p.lammich@utwente.nl}
+\affil*[1]{\orgdiv{Faculty of Electrical Engineering, Mathematics and Computer Science} \orgname{University of Twente} \orgaddress{\city{Enschede} \country{Netherlands}}}
+
+\abstract{
   We present a stepwise refinement approach to develop
@@
 %
-  As case study, we verify a parallel quicksort algorithm, and
-  show that it performs on par with its C++ implementation,
-  and is competitive to state-of-the-art parallel sorting algorithms.
-\end{abstract}
+  As case study, we verify a parallel quicksort algorithm that is
+  competitive to unverified state-of-the-art algorithms.
+}
+
+\keywords{Isabelle,Concurrent Separation Logic,Parallel Sorting,LLVM}
+
+\maketitle
 
@@
 unverified implementations. Our approach is backwards compatible to
-the Isabelle Refinement Framework (IRF), a powerful tool to verify efficient sequential software,
+the Isabelle Refinement Framework (IRF)~\cite{La19-llvm}, a powerful tool to verify efficient sequential software,
 such as model checkers~\cite{ELNN13,BrLa18,WiLa18}, SAT solvers~\cite{La17_CADE,La17_SAT,FBL18},
@@
 re-using an existing verification of state-of-the-art sequential sorting algorithms~\cite{La20}.
-Our verified parallel sorting algorithm is competitive to state-of-the-art parallel sorting algorithms.
+Our verified parallel sorting algorithm is competitive to state-of-the-art parallel sorting algorithms
+from GNU's C++ standard library and the Boost C++ Libraries.
+
+This paper is an extended version of our ITP 2023 paper~\cite{La22}.
+The main new contribution is a verified parallel partitioning algorithm,
+which significantly improves the efficiency and scalability of our sorting algorithm from~\cite{La22}.
+To this end, we added a description of the interval list data structure (Section~\ref{sec:mod_ds})
+and the \is{with_idxs} combinator (Section~\ref{sec:array_split}), which are required by the parallel partitioner.
+The parallel partitioning algorithm itself is described in Section~\ref{sec:ppar}, and Section~\ref{sec:benchmark}
+contains updated and more extensive benchmarks.
 
@@
 
-% \noindent The formalization is available at \url{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}.
+\noindent The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.
 
@@
 This paper is based on the Isabelle Refinement Framework, a continuing effort
-to verify efficient implementations of complex algorithms, using stepwise refinement techniques.
+to verify efficient implementations of complex algorithms, using stepwise refinement techniques~\cite{LaLo10,LaTu12,La13,La15,La19-llvm,HaLa21}.
 Figure~\ref{fig:IRF-overview} displays the components of the Isabelle Refinement Framework.
-
+%
 \begin{figure}
@@
 
-  \caption{Components of the Isabelle Refinement Framework, with focus on the back end.}\label{fig:IRF-overview}
+  \caption{Components of the IRF, with focus on the back end.}\label{fig:IRF-overview}
 \end{figure}
-
-%   \begin{tikzpicture}[node distance=.01\textwidth,outer sep=0pt,
-%   layerA/.style={
-%     draw=black,
-%     fill=blue!20,
-%     rounded corners,
-%     text width=.9\textwidth,
-%     font={\sffamily\bfseries\color{black}},
-%     align=center,
-%     minimum height=.03\textheight,
-% %    text height=12pt,
-% %    text depth=9pt
-%     },
-%   layerB/.style={layerA,fill=green!20},
-%   layerC/.style={layerA,fill=red!20},
-%   layerD/.style={layerA,fill=red!30,text width=.2\textwidth},
-%   ]
-%
-%   \node[layerA] (AlgDS) {Algorithms and Data Structures};
-%
-%   \node[layerB, below=of AlgDS.south west, anchor=north west, text width=.437\textwidth] (ne) {Nres-Monad};
-%   \node[layerA, right=of ne.north east, anchor=north west, text width=.437\textwidth, minimum height=.067\textheight] (llDS) {Low-Level Algorithms \\and Data Structures};
-%
-%   \node[layerB, below=of ne.south west, anchor=north west, text width=.437\textwidth] (sepref) {Sepref};
-%
-%   \node[layerC, below=of sepref.south west, anchor=north west] (sepl) {Program Logic and VCG};
-%   \node[layerC, below=of sepl.south west, anchor=north west, minimum height=.120\textheight] (sembox) {};
-%
-%   \node[below right=of sembox.north west] {Back End Semantics};
-%
-%
-%   \node[layerD, right=of sembox.west] (isacode) {Isa Codegen};
-%   \node[layerD, right=of isacode] (imphol) {Imp HOL};
-%   \node[layerD, right=of imphol] (llvm) {LLVM};
-%   \node[layerD, right=of llvm] (llvm-par) {LLVM par};
-%
-%   \node[layerD, below=of imphol] (impholtime) {Imp HOL time};
-%   \node[layerD, below=of llvm] (llvmtime) {LLVM time};
-%
-%
-%   \end{tikzpicture}
-
+%
 The back end layer handles the translation from Isabelle/HOL to the actual target language.
@@
 
-At the level of the program logic and VCG, our framework can already be used to verify simple low-level algorithms and data structures,
+At the level of the program logic and VCG, our framework can be used to verify simple low-level algorithms and data structures,
 like dynamic arrays and linked lists. More complex developments typically use a stepwise refinement approach,
@@
 
+\subsection{Notation}
+We mainly use Isabelle/HOL notation, with some shortcuts and adaptations for presentation in a paper.
+In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
+e.g., \is{\<forall>x. x<4 ==> x<=3} (in Isabelle, one must use \is{-->} here).
+
+Type variables are \is{'a, 'b, ...} and a function types are written curried as \is{'a => 'b => 'c}.
+Types can be annotated to any symbol or term, e.g., \is{(x::nat) < 4 ==> x<=3}.
+Function application is written as \is{f x}, and function update is \is{f(x:=y)}.
+For definitions, we use \is{==}, e.g., \is{f x == x+1}.
+
+Algebraic datatypes are written as \is{'a option == Some (the: 'a) | None}. This also defines the selector function \is{the :: 'a option => 'a}.
+Values of the tuple type \is{'a_1 \x ... \x 'a_n} are written as \is{(a_1,...,a_n)}. Lists have type \is{'a list},
+and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@xs_2} for list concatenation.
+We also use the list notation \is{[1,2,3]}. Then length of list \is{xs} is \is{|xs|}.
+
+The empty set is \is${} :: 'a set$. The set \is${l..<h}$ contains all elements between $l$ inclusive and $h$ exclusive, and the set \is${l..}$ contains all elements greater than or equal to $l$.
+Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
+Otherwise, we will explicitly write, e.g. \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.
+
+Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x>42}.
+We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
+Also, we can use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.
+
 \section{A Back End for LLVM with Parallel Execution}\label{sec:parsem}
@@
   to the flexibility and feasibility of the approach.
-  The main idea is to make an execution report the memory that it accesses,
+  The main idea is to make an execution report its accessed memory,
   and use this information to raise an error when joining executions that
@@
   The \is{return} operation yields exactly one result, and \is{bind} combines all possible results,
-  failing if there is a possibility to fail.
+  failing if there is a possibility to fail. We use the notation \is{x<-m; f x} for \is{bind m (\<lambda>x. f x)},
+  and \is{m_1; m_2} for \is{bind m_1 (\<lambda>_. m_2)}.
 
@@
   Here, \is{return} does not change the state, and reports no accesses (\is$0$),
-  and \is{bind} sequentially composes the executions, threading through the state $\mu$, and adding up the access reports \is$r_1$ and \is$r_2$.
+  and \is{bind} sequentially composes the executions, threading through the state $\mu$ and adding up the access reports \is$r_1$ and \is$r_2$.
 
@@
   As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
-  Let \is{norace :: '\rho => '\rho => bool} and \is{feasible :: '\rho => '\rho => bool} be symmetric predicates,
+  Let \is{feasible :: '\rho => '\rho => bool} and \is{norace :: '\rho => '\rho => bool} be symmetric predicates,
   and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
-  compose two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
+  combine two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
   \begin{lstlisting}
@@
   Note that, if one parallel strand fails, and the other parallel strand has no possible
-  results \is{spec (\<lambda>_. False)}, the behaviour of the parallel composition is not clear.
+  results (\is{spec (\<lambda>_. False)}), the behaviour of the parallel composition is not clear.
   For this reason, we fix an invariant \is{invar$_M$ :: ('\mu => ('x \x '\rho \x '\mu) neM) => bool},
@@
   \subsection{Access Reports}
-  We now fix the state of the M-monad to be memory, and the access reports to be sets
-  of read and written addresses, as well as sets of allocated and freed blocks:
-  \begin{lstlisting}
-    acc == ( r :: addr set; w :: addr set; a :: nat set; f :: nat set )
+  We now fix the state of the M-monad to be memory, and the access reports to be tuples \is{(r,w,a,f)} of
+  read and written addresses, as well as sets of allocated and freed blocks:
+  \begin{lstlisting}
+    acc == addr set \x addr set \x nat set \x nat set
     0 == ( {},{},{},{} )
@@
 
-  Two parallel executions are feasible if they did not allocate the same block, and they have a data race
-  if one strand accesses addresses or blocks modified by the other strand:
+  Two parallel executions are feasible if they did not allocate the same block. They have a data race
+  if one execution accesses addresses or blocks modified by the other:
   \begin{lstlisting}
@@
         (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
-  \end{lstlisting}
-  The invariant for \is{M} states that blocks transition only from fresh to allocated to free,
-  allocated blocks never change their size, and the access report matches the observable state change (\is{consistent}).
-  It also states, that for each finite set of blocks $B$, there is an execution that does not allocate blocks from $B$.
-  The latter is required to show that we always find feasible parallel executions:
-  \begin{lstlisting}
-    invar$_M$ c == \<forall>\mu P. c \mu = spec P ==>
-      (\<forall>x \rho \mu'. P (x,\rho,\mu') ==> consistent \mu \rho \mu')
-    & (\<forall>B. finite B ==> (\<exists>x \rho \mu'. P (x,\rho,\mu') & \rho.a \inter B = {} ))
   \end{lstlisting}
@@
   and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
-  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}, and \is{addr} a valid address in \is{\mu'}. Then
+  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}. Then
   \begin{lstlisting}
@@
     (3)  \mu' b = FREED <--> \mu b = FREED \or b \in f_1 \union f_2
-
-    (4)  a \in w_1 \or b \in a_1 ==> get_addr \mu' a = get_addr \mu_1 a
-    (5)  a \in w_2 \or b \in a_2 ==> get_addr \mu' a = get_addr \mu_2 a
-    (6)  a ~: w_1\<union>w_2 \or b ~: a_1\<union>a_2 ==> get_addr \mu' a = get_addr \mu a
+  \end{lstlisting}
+  Moreover, for all addresses \is{addr = ADDR b i} with \is{valid \mu' addr}:
+  \begin{lstlisting}
+    (4)  addr \in w_1 \or b \in a_1 ==> get \mu' addr = get \mu_1 addr
+    (5)  addr \in w_2 \or b \in a_2 ==> get \mu' addr = get \mu_2 addr
+    (6)  addr ~: w_1\<union>w_2 & b ~: a_1\<union>a_2 ==> get \mu' addr = get \mu addr
   \end{lstlisting}
@@
   The properties (4)--(6) define the content: addresses written or allocated in the first or second execution get their content from \is{\mu_1} (4) or \is{\mu_2} (5) respectively.
-  Addresses not written or allocated at all keep their original content (6).
+  Addresses not written nor allocated at all keep their original content (6).
+
+  \subsection{The Interface of the M-Monad}
+  The invariant for \is{M} states that blocks transition only from fresh to allocated to free,
+  allocated blocks never change their size, and the access report matches the observable state change (\is{consistent}).
+  It also states, that for each finite set of blocks $B$, there is an execution that does not allocate blocks from $B$.
+  The latter is required to show that we always find feasible parallel executions:
+  \begin{lstlisting}
+    invar$_M$ c == \<forall>\mu P. c \mu = spec P ==>
+      (\<forall>x \rho \mu'. P (x,\rho,\mu') ==> consistent \mu \rho \mu')
+    & (\<forall>B. finite B ==> (\<exists>x \rho \mu'. P (x,\rho,\mu') & \rho.a \inter B = {} ))
+  \end{lstlisting}
+
+  To define functions in the M-monad, we have to show that they satisfy this invariant.
+  For \is{return} and \is{bind}, this is straightforward. The proof for the parallel operator is slightly more involved,
+  using the properties of \is{combine}, and the invariant for the operands to obtain a feasible parallel execution.
+
+  Moreover, the M monad provides the memory management functions \is{Mmalloc}, \is{Mfree}, \is{Mload}, \is{Mstore}, and \is{Mvalid_addr}.
+  The latter function checks if a given address is valid, and is used to check if pointer arithmetic can be performed
+  on that address. Currently, it behaves like loading from that address, in particular it does not support pointers
+  one past the end of an allocated block. We leave integration of such pointers to future work.
+
+  \begin{example}[{Memory Allocation}]
+    To define memory allocation in \is{M}, we first define the allocation function in the underlying nondeterminism-error monad:
+    \begin{lstlisting}
+    malloc vs \mu ==
+      b <- spec b. is_FRESH \mu b;
+      return (b, ({},{},{b},{}), \mu(b:=ALLOC vs))
+    \end{lstlisting}
+    This function selects an arbitrary fresh block \is$b$, and initializes it with the given list \is$vs$ of values.
+    It returns the allocated block, an access report for the allocation, and the updated memory.
+
+    We then show that \is{malloc} satisfies the invariant of M: we correctly report the allocated block,
+    and we can select any fresh block. As our memory model guarantees an infinite supply of fresh blocks,
+    any finite set of blocks can be avoided:
+    \begin{lstlisting}
+      invarM_malloc: invarM (malloc vs)
+    \end{lstlisting}
+
+    Finally, we define the corresponding function in the M monad, using Isabelle's lifting and transfer package~\cite{HuKu13}:
+    \begin{lstlisting}
+      lift_definition Mmalloc :: val list => M is malloc by (rule invarM_malloc)
+    \end{lstlisting}
+    The other memory management functions are defined analogously.
+  \end{example}
 
@@
   this is analogous to the sequential case~\cite{La19-llvm}.
-  The exceptions are memory allocation, which nondeterministically allocates some available
-  block (the original formalization deterministically counted up the block indexes), and an instruction for parallel
-  function call:
+  Additionally, we define an instruction for parallel function call:
   \begin{lstlisting}
@@
   In the previous section, we have defined a shallow embedding of LLVM programs into Isabelle/HOL.
-  We now describe how to reason about these programs, using separation logic.
+  We now reason about these programs, using separation logic.
 
@@
   For the exact definition of a separation algebra, we refer to~\cite{CHY07,KKB12}. We note that
-  separation algebras naturally extend over functions and pairs, in a pointwise manner.
+  separation algebras naturally extend over functions and pairs in a pointwise manner.
 
@@
   \begin{example}(Trivial Separation Algebra)
-    The type \is{\alpha option = None | Some \alpha} forms a separation algebra with:
+    The type \is{'a option = None | Some 'a} forms a separation algebra with:
     \begin{lstlisting}
@@
   \begin{lstlisting}
-    \alpha :: memory -> (addr -> val option) \x (nat -> nat option)
+    \alpha :: memory => (addr => val option) \x (nat => nat option)
     \alpha \mu == (\alpha$_m$ \mu, \alpha$_b$ \mu)
@@
     false a == False     true a == True     \<box> a == a=0
-    (P*Q) a == \<exists> a_1 a_2. a_1 # a_2 \and a = a_1 + a_2 \and P a_1 \and Q a_2
+    (P*Q) a == \<exists>a_1 a_2. a_1 # a_2 \and a = a_1 + a_2 \and P a_1 \and Q a_2
   \end{lstlisting}
@@
   \begin{lstlisting}
-    wp m Q \mu == case m \mu of spec Q' => \<forall>x \rho \mu'. Q' (x,\rho,\mu') ==> Q x \rho \mu' | fail => False
+    wp m Q \mu == case m \mu of spec Q' => \<forall>x \rho \mu'. Q' (x,\rho,\mu') ==> Q x \rho \mu'
+                                   | fail => False
   \end{lstlisting}
@@
 
-  Equipped with a weakest precondition with access restrictions, we define a Hoare-triple:
+  Equipped with \is{wpf}, we define a Hoare-triple:
   \begin{lstlisting}
@@
   \begin{lstlisting}
-  |= {n~=0} ll_malloc TYPE(\<alpha>) n {%p. range {0..<n} (%_. init) p * b_tag n p}
-  |= {range {0..<n} xs p * b_tag n p} ll_free p {%_. \<box>}
-  |= {pto x p} ll_load p {%r. r=x * pto x p}
-  |= {pto y p} ll_store x p {%_. pto x p}
-  \end{lstlisting}
-  Here \is{b_tag n p} asserts that \is{p} points to the beginning of a block of size \is$n$, and
-  \is{range I f p} describes that for all $i\in I$, $p+i$ points to value $f~i$.
+  |= {n~=0} ll_malloc TYPE('a) n {%p. range {0..<n} p (%_. init) * b_tag p n}
+  |= {range {0..<n} p f * b_tag p n} ll_free p {%_. \<box>}
+  |= {pto p x} ll_load p {%r. r=x * pto p x}
+  |= {pto p y} ll_store x p {%_. pto p x}
+  \end{lstlisting}
+  Here \is{b_tag p n} asserts that \is{p} points to the beginning of a block of size \is$n$, and
+  \is{range I p f} describes that for all $i\in I$, $p+i$ points to value $f~i$.
   Intuitively, \is{ll_malloc} creates a block of size $n$, initialized with the default \is{init} value,
   and a tag. If one possesses both, the whole block and the tag, it can be deallocated by free.
-  The rules for load and store are straightforward, where \is{pto x p} describes that $p$ points to value $x$.
-
-  \section{Refinement for Parallel Programs}\label{sec:refine_parallel}
+  The rules for load and store are straightforward, where \is{pto p x} describes that $p$ points to value $x$.
+
+  \section{Refinement for (Parallel) Programs}\label{sec:refine_parallel}
   At this point, we have described a separation logic framework for parallel programs in LLVM.
@@
   In this section we describe the refinement infrastructure of the
-  Isabelle Refinement Framework, focusing on our changes to support parallel algorithms.
+  Isabelle Refinement Framework.
 
@@
   Abstract programs are shallowly embedded into the nondeterminism error monad \is{'a neM} (cf.\ Section~\ref{sec:monad}).
-  They are purely functional, not modifying memory, or differentiating between sequential and parallel execution.
+  They are purely functional and have no notion of parallel execution.
   We define a \emph{refinement ordering} on \is{neM}:
@@
 
+  \begin{example}[Swapping multiple elements]\label{ex:swap_spec}
+    We specify an operation to perform multiple swaps.
+It takes two disjoint sets of indices \is{s_1} and \is{s_2}, and a list \is{xs}.
+It then swaps each index in \is{s_1} with some index in \is{s_2}.
+    The precondition of this operation assumes that the index sets are in range, disjoint, and have the same cardinality:
+    \begin{lstlisting}
+      swap_spec_pre s_1 s_2 xs == s_1 \inter s_2 = {} & s_1 \union s_2 \<subseteq> {0..<|xs|} & |s_1| = |s_2|
+    \end{lstlisting}
+    The postcondition ensures that the resulting list is a permutation of the original list,
+    the elements at indexes outside \is{s_1 \union s_2} are unchanged, and that each element in \is{s_1} is swapped with one in \is{s_2}:
+    \begin{lstlisting}
+      swap_spec_post s_1 s_2 xs xs' ==
+        mset xs' = mset xs & (\<forall>i\<notin>s_1 \union s_2. i < |xs| ==> xs'!i=xs!i)
+      & (\<forall>i\<in>s_1. \<exists>j\<in>s_2. xs'!i=xs!j) & (\<forall>j\<in>s_2. \<exists>i\<in>s_1. xs'!j=xs!i)
+    \end{lstlisting}
+
+    As a sanity check, we prove that our specification is not vacuous, i.e., that for every input that satisfies the precondition, there
+    exists an output that satisfies the postcondition:
+    \begin{lstlisting}
+      swap_spec_pre s_1 s_2 xs ==> \<exists>xs'. swap_spec_post s_1 s_2 xs xs'
+    \end{lstlisting}
+    Note that this is only a sanity check lemma to detect problems early.
+    Should we accidentally insert a vacuous specification here, we won't be able to prove refinement to an M-monad program later, which cannot be vacuous due to \is{invar$_M$}.
+
+    In the \is{ne} monad, we then specify:
+    \begin{lstlisting}
+      swap_spec s_1 s_2 xs ==
+        assert (swap_spec_pre s_1 s_2 xs); spec xs'. swap_spec_post s_1 s_2 xs xs'
+    \end{lstlisting}
+%
+    In Section~\ref{sec:ppar} we will refine this specification to a parallel implementation in LLVM.
+  \end{example}
+
   \subsection{The Sepref Tool}
@@
   keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
-  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
+  During the symbolic execution, the tool synthesizes an Isabelle-LLVM program, together with a refinement proof.
   The synthesis is automatic, but requires annotations to the abstract program.
@@
     "hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ==
-        c ~= fail ==> ht \<Gamma> c$_\dagger$ (%x$_\dagger$. \<exists>x. \<Gamma>' * R x x$_\dagger$ * \<up>(return x <= c \and CP x$_\dagger$))"
+        c ~= fail ==> ht \<Gamma> c$_\dagger$ (%r$_\dagger$. \<exists>r. \<Gamma>' * R r$_\dagger$ r * \<up>(return r <= c \and CP r$_\dagger$))"
   \end{lstlisting}
@@
   or for a memory described by assertion \is{\<Gamma>},
-  the LLVM program \is{c$_\dagger$} succeeds with \is{x$_\dagger$},
-  such that the new memory is described by \is{\<Gamma>' * R x x$_\dagger$},
-  for a possible result \is{x} of the abstract program \is{c}. Moreover, the predicate \is{CP} holds for the concrete result.
+  the LLVM program \is{c$_\dagger$} succeeds with result \is{r$_\dagger$},
+  such that the new memory is described by \is{\<Gamma>' * R r$_\dagger$ r},
+  for a possible result \is{r} of the abstract program \is{c}. Moreover, the predicate \is{CP} holds for the concrete result.
   Note that \is{hnr} trivially holds for a failing abstract program. This makes sense, as we prove that
-  the abstract program does not fail anyway. Moreover it allows us to assume that assertions actually hold during the refinement proof:
+  the abstract program does not fail anyway. It allows us to assume abstract assertions during the refinement proof:
   \begin{lstlisting}
@@
     \begin{lstlisting}
-      lget xs i == assert (i<|xs|); return xs!i     lset xs i x == assert (i<|xs|); return xs[i:=x]
+      lget xs i == assert (i<|xs|); return xs!i
+      lset xs i x == assert (i<|xs|); return xs[i:=x]
     \end{lstlisting}
@@
     \begin{lstlisting}
-      arr$_A$ xs p = range {0..<|xs|} (%i. xs!i) p
+      arr$_A$ p xs = range {0..<|xs|} p (%i. xs!i)
     \end{lstlisting}
     That is, for every $i<|xs|$, $p+i$ points to the $i$th element of $xs$.
-    On arrays, indexing and updating of arrays is implemented by:
-    \begin{lstlisting}
-      aget p i == ll_ofs_ptr p i; ll_load p         aset p i x == ll_ofs_ptr p i; ll_store x p; return p
-    \end{lstlisting}
-    And the abstract and concrete programs are linked by the following refinement theorems:
-    \begin{lstlisting}
-      hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ i i$_\dagger$)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs xs$_\dagger$$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i i$_\dagger$)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ (%_.
-True)$\hspace*{-1pt}$ (lget xs i)
-      hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ i i$_\dagger$) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i i$_\dagger$) arr$_A$ (%r. r=xs$_\dagger$) (lset xs i x)
+    Assertions like \is{arr$_A$}, that relate concrete to abstract values, are called \emph{refinement relations}.
+    If we want to emphasize that they depend on the heap, we also call them refinement \emph{assertions}.
+
+    Indexing and updating of arrays is implemented by:
+    \begin{lstlisting}
+      aget p i == ll_ofs_ptr p i; ll_load p
+      aset p i x == ll_ofs_ptr p i; ll_store x p; return p
+    \end{lstlisting}
+    The abstract and concrete programs are linked by the following refinement theorems:
+    \begin{lstlisting}
+      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs$_\dagger$ xs$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ $\top$$\hspace*{-1pt}$ (lget xs i)
+      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r=xs$_\dagger$) (lset xs i x)
     \end{lstlisting}
@@
     by the fixed-width\footnote{We use Isabelle's word library here, which encodes the actual width as a type variable,
-    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word \is{i$_\dagger$} (\is{idx$_A$ i i$_\dagger$}),
+    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word \is{i$_\dagger$} (\is{idx$_A$ i$_\dagger$ i}),
     the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
-    The resulting memory will still contain the original array. Note that there is no explicit precondition that the array access
+    The resulting memory will still contain the original array. Note that there is no explicit precondition ($\top$) that the array access
     is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.
@@
     This information is important for refining to parallel programs on disjoint parts of an
-    array (cf.\ Section~\ref{sec::array_split}).
+    array (cf.\ Section~\ref{sec:array_split}).
+
+    To increase readability, we introduce an (almost) point-free notation for refinement theorems.
+    The theorems for the array operations above can also be written as:
+    \begin{lstlisting}
+      aget, lget : arr$_A$ -> idx$_A$ -> id$_A$
+      aset, lset : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> id$_A$ -> arr$_A$:r [r=xs$_\dagger$]
+    \end{lstlisting}
+    The first theorem simply states that the first argument is refined by \is{arr$_A$},
+    the second argument by \is{idx$_A$}, and the result by \is{id$_A$}.
+    The second theorem adds the annotation $\cdot^d$ to the refinement for the array argument,
+    indicating that this argument will be destroyed, i.e., the refinement is no longer valid when the function returns.
+    Moreover, it binds the array argument to the name \is{xs$_\dagger$} and the result to $r$. These names are used in the pointer equality predicate \is{[r=xs$_\dagger$]}
+    at the end, indicating that the result will be the same pointer as the array argument.
 
@@
 
-  Given refinement assertions for the parameters, and \is{hnr}-rules for all operations in a program,
+  Given refinement relations for the parameters, and refinement theorems for all operations in a program,
   the Sepref tool automatically synthesizes an LLVM program from an abstract \is{neM} program.
@@
 
-  \subsection{Array Splitting}\label{sec::array_split}
+  \subsection{Modular Data Structure Development}\label{sec:mod_ds}
+  The Refinement Framework allows us to build more complex data structures, using already existing ones as building blocks,
+  and chaining together several refinements.
+  We describe the development of an interval list data structure, which we need for our parallel partitioning algorithm (cf.~\ref{sec:ppar}).
+
+  A pair of natural numbers \is$(l,h)$ can be used to represent the set \is${l..<h}$. We define \is{iv$_R$} to be
+  the refinement relation between intervals (pairs) and sets.
+  Moreover, we define operations for constructing an interval, testing if an interval is empty, intersection, and cardinality.
+  We show that these operations refine the corresponding operations on sets:
+  \begin{lstlisting}
+    iv$_R$ :: (nat \x nat) => nat set => bool                    iv$_R$ (l,h) s == s = {l..<h}
+
+    iv l h == (l,h)                                                 iv, {_..<_} : id$_R$ -> id$_R$ -> iv$_R$
+    iv_is_empty (l,h) == h<=l                                    iv_is_empty, (={}) : iv$_R$ -> id$_R$
+    iv_inter (l_1,h_1) (l_2,h_2) == ( max l_1 l_2, min h_1 h_2 )     iv_inter, (\inter) : iv$_R$ -> iv$_R$ -> iv$_R$
+    iv_card (l,h) == if h<l then 0 else h-l                iv_card, |_| : iv$_R$ -> id$_R$
+  \end{lstlisting}
+  Note that \is$h<l ==> {l..<h} = {}$, and we do not enforce $l\le h$ for our representation. Thus, no checks are needed for construction and intersection.
+  However, we use a check to avoid underflow when computing the cardinality.
+
+  Analogously, we implement open intervals with a single number, and define operations to construct an open interval,
+  and to intersect a closed and an open interval:
+  \begin{lstlisting}
+    ivo$_R$ :: nat => nat set => bool        ivo$_R$ l s == s = {l..}
+
+    ivo l == l                                         ivo, {_..} : id$_R$ -> iv$_R$
+    iv_inter_ivo (l_1,h_1) l_2 == (max l_1 l_2) h_1    iv_inter_ivo, (\inter) : iv$_R$ -> ivo$_R$ -> iv$_R$
+  \end{lstlisting}
+
+  Next, we use Sepref to implement the natural numbers by fixed-sized words (\is{idx$_A$}).
+  For example, given the definition of \is{iv_inter}, and an annotation to implement natural numbers by 64 bit words,
+  Sepref synthesizes the Isabelle LLVM program \is{iv_inter$_\dagger$} and proves the refinement theorem:
+  \begin{lstlisting}
+    iv_inter$_\dagger$, iv_inter :  idx$_A$ \x idx$_A$ -> idx$_A$ \x idx$_A$ -> idx$_A$ \x idx$_A$
+  \end{lstlisting}
+  We then define \is{iv$_A$ == (idx$_A$ \x idx$_A$) O iv$_R$} as the composition of the two refinements (word to nat to set).
+  With the help of Sepref's FCOMP tool, we can automatically compose the refinement lemmas.
+  For example, composing the refinement lemmas for \is{iv_inter$_\dagger$, iv_inter} and \is{iv_inter, (\inter)} yields:
+  \begin{lstlisting}
+    iv_inter$_\dagger$, (\inter) :  iv$_A$ -> iv$_A$ -> iv$_A$
+  \end{lstlisting}
+  Thus, we obtain imperative implementations of the set operations. We proceed analogously for open intervals.
+
+  Next, we implement a set as the union of a list of non-empty, pairwise disjoint, and finite sets.
+  While that seems to make little sense at first glance, we will later implement the sets in the list by intervals,
+  and the list itself by dynamic arrays, to obtain an imperative interval list data structure.
+  We define operations for constructing an empty set, emptiness test,
+  disjoint union with a single set, cardinality, and a more specialized operation \is{split}, which splits off
+  a non-empty set from the list:
+  \begin{lstlisting}
+    ivl$_R$ :: nat set list => nat set => bool
+    ivl$_R$ ls s == s = \<Union> set s & pw_disjoint ls & \<forall>x:set ls. finite x & x ~= {}
+
+    ivl_empty == []                                 ivl_empty, {} : ivl$_R$
+    ivl_is_empty ls == ls=[]                       ivl_is_empty, (={}) : ivl$_R$ -> id$_R$
+    ivl_dj_un s ls == s#ls                         ivl_dj_un, ($\dot\cup$) : id$_R$ -> ivl$_R$ -> ivl$_R$
+    ivl_card ls == fold (%s c. c=c+|s|) ls 0     ivl_card, card : ivl$_R$ -> id$_R$
+    ivl_split (s#ls) == (s,ls)                      ivl_split, split : ivl$_R$ -> id$_R$ \x ivl$_R$
+      split s == assert (s~={}); spec (s_1,s_2). s = s_1 $\dot\cup$ s_2 & s_1~={}
+  \end{lstlisting}
+
+  We then refine the lists of sets to array lists (dynamic arrays) of intervals: \is{al$_A$ iv$_A$}.
+  Here, \is{al$_A$} is the refinement assertion from lists to the array list data structure from the IRF collections library.
+  As argument, it takes a refinement assertion for the list elements.
+  Again, Sepref automatically generates imperative implementations of the \is{ivl} functions and proves the corresponding
+  refinement lemmas. Combining them with the refinements to sets yields the desired imperative interval list data structure,
+  with the refinement relation \is{ivl$_A$ == al$_A$ iv$_A$ O ivl$_R$}.
+  For example, for joining a single interval to the list, and for splitting off an interval, we get:
+  \begin{lstlisting}
+    ivl_dj_un$_\dagger$, ($\dot\cup$) : iv$_A$ -> ivl$_A$^d -> ivl$_A$
+    ivl_split$_\dagger$, split : ivl$_A$^d -> iv$_A$ \x ivl$_A$
+  \end{lstlisting}
+  Note that we update the underlying dynamic array destructively, hence the $\cdot^d$ annotation to the argument refinements.
+
+  In a last step, we define some operations on (finite) sets, and use Sepref to directly refine them to arrays, without any explicit intermediate steps.
+  For example, intersecting two finite sets can be expressed as:
+  \begin{lstlisting}
+    set_inter_it s_1 s_2 ==
+      assert(finite s_2);
+      r={}; while (s_2 ~= {}) ( (ss,s_2) <- split s_2; r = (s_1 \inter ss) $\dot\cup$ r );
+      return r
+  \end{lstlisting}
+  It is straightforward to prove that this algorithm returns \is$s_1 \inter s_2$.
+  Also, Sepref can implement $s_1$ with a closed or open interval, and $s_2$ with an interval array, yielding:
+  \begin{lstlisting}
+    iv_inter_ivl$_\dagger$, (\inter) : iv$_A$ -> ivl$_A$^d -> ivl$_A$
+    ivo_inter_ivl$_\dagger$, (\inter) : ivo$_A$ -> ivl$_A$^d -> ivl$_A$
+  \end{lstlisting}
+
+  We have demonstrated one way of modularly developing an interval list data structure based on a dynamic array.
+  By separating the actual intervals from the list data structure, the proofs about the interval
+  list where independent of the interval implementation.
+  This is a design choice, and a more direct design, using \is{(nat \x nat) list} as intermediate data structure, is certainly possible.
+
+  \subsection{Array Splitting}\label{sec:array_split}
   An important concept for parallel programs is to concurrently operate on disjoint parts of the memory,
@@
   \begin{lstlisting}
-    awith_split i xs$_\dagger$ f$_\dagger$ == xs$_\dagger$_2 <- ll_ofs_ptr xs$_\dagger$ i; f$_\dagger$ xs$_\dagger$ xs$_\dagger$_2; return xs$_\dagger$
-
-    hnr (arr$_A$ xs_1 xs$_\dagger$_1 * arr$_A$ xs_2 xs$_\dagger$_2) (f$_\dagger$ xs$_\dagger$_1 xs$_\dagger$_2) \<box>
-        (arr$_A$ \x arr$_A$) (%(xs$_\dagger$_1',xs$_\dagger$_2'). xs$_\dagger$_1'=xs$_\dagger$_1 \and xs$_\dagger$_2' = xs$_\dagger$_2)
+    awith_split i p f$_\dagger$ == p_2 <- ll_ofs_ptr p i; f$_\dagger$ p p_2; return p
+  \end{lstlisting}
+  The corresponding refinement theorem is:
+  \begin{lstlisting}
+    awith_split, with_split :
+      arr$_A$^d:p -> idx$_A$ -> (arr$_A$^d:p_1 -> arr$_A$^d:p_2 -> arr$_A$:r_1 \x arr$_A$:r_2 [r_1=p_1 & r_2=p_2])
+      -> arr$_A$:r [r=p]
+  \end{lstlisting}
+  or, equivalently, in pointwise notation:
+  \begin{lstlisting}
+    hnr (arr$_A$ p_1 xs_1 * arr$_A$ p_2 xs_2) (f$_\dagger$ p_1 p_2) \<box>
+        (arr$_A$ \x arr$_A$) (%(r_1,r_2). r_1=p_1 \and r_2 = p_2)
         (f xs_1 xs_2)"
     ==>
-    hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ i i$_\dagger$) (awith_split i$_\dagger$ xs$_\dagger$ f$_\dagger$)
-        (idx$_A$ i i$_\dagger$) (%xs xs$_\dagger$. arr$_A$ xs xs$_\dagger$) (%xs$_\dagger$'. xs$_\dagger$'=xs$_\dagger$)
+    hnr (arr$_A$ p xs * idx$_A$ i$_\dagger$ i) (awith_split i$_\dagger$ p f$_\dagger$)
+        (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r=p)
         (with_split i xs f)
   \end{lstlisting}
-  The refinement of the function \is{f} to \is{f$_\dagger$} requires an additional proof that the returned pointers
-  are equal to the argument pointers (\is{xs$_\dagger$_1'=xs$_\dagger$_1 \and xs$_\dagger$_2' = xs$_\dagger$_2}).
-  Sepref tries to prove that automatically, using a simple heuristics.
+  The refinement of the function argument (\is{f} to \is{f$_\dagger$}) requires an additional proof that the returned pointers
+  are equal to the argument pointers (\is{r_1 = p_1 \and r_2 = p_2}).
+  Sepref tries to prove that automatically, using its pointer equality heuristics.
+
+  Splitting an array into two halves allows us to abstractly treat the array and its two halves just as lists,
+  which simplifies the abstract proofs: the fact that the two halves come from the same array is only visible
+  at a later refinement stage. However, while splitting an array in halves is adequate for many operations,
+  it is not a workable abstraction for swapping multiple elements in parallel (cf.~Sec.~\ref{ex:swap_spec}):
+  while, in theory, we could split the array element-wise, this would incur a considerable proof burden.
+
+  A more elegant solution is to keep track of which elements of a list can be accessed already on the abstract level. To this end,
+  we model a list of option values, \is{None} meaning that we cannot access this element.
+  We start by defining functions to abstractly handle lists of option values.
+  These functions work on the actual list, and the \emph{structure} of the list,
+  which is a list of Booleans indicating which elements we do \emph{not} own.
+  Using the structure of the list as an explicit concept simplifies abstract proofs, as, typically,
+  the values in the list will change, while the structure is preserved.
+  The following functions obtain the structure of a list, and determine if two
+  structures are compatible, i.e., have the same lengths and own disjoint indexes:
+  \begin{lstlisting}
+    sl_struct :: 'a option list => bool list        sl_struct xs == map (=None) xs
+    sl_compat :: bool list => bool list => bool   sl_compat s_1 s_2 == list_all2 (\<or>) s_1 s_2
+  \end{lstlisting}
+  We also define functions to split and join lists:
+  \begin{lstlisting}
+    sl_split s xs == map (%i. if i\<in>s & i < |xs| then xs!i else None) [0..<|xs|${}$]
+    join_option None y == y     join_option x None == x      join_option _ _ == None
+    sl_join xs_1 xs_2 == map2 join_option xs_1 xs_2
+  \end{lstlisting}
+  Here, \is{sl_split s xs} returns a list that owns the indexes that are in \is{s} and owned by \is{xs},
+  and \is{sl_join xs_1 xs_2} joins the elements of two (compatible) lists.
+
+  Analogously to \is{with_split}, we define a combinator \is{with_idxs s xs f}, that splits
+  the list \is{xs} into the lists with the indexes \is{s}, and without the indexes \is{s}, executes \is{f} on these lists,
+  and joins the resulting lists:
+  \begin{lstlisting}
+  with_idxs s xs m ==
+    assert (\<forall>i\<in>s. i<|xs| & xs!i ~= None);
+    let (xs_1, xs_2) = (sl_split s xs, sl_split (-s) xs);
+    (xs$_1'$,xs$_2'$) <- f xs_1 xs_2;
+    assert (sl_struct xs$_1'$ = sl_struct xs_1 & sl_struct xs$_2'$ = sl_struct xs_2);
+    return (sl_join xs$_1'$ xs$_2'$)
+  \end{lstlisting}
+
+  On the concrete side, we define the refinement assertion \is{oarr$_A$} between arrays and lists of options.
+  It only owns the indexes of the array that are not None:
+  \begin{lstlisting}
+  oarr$_A$ p xs == range {i | i. i<length xs & xs!i~=None}) p (%i. the (xs!i))
+  \end{lstlisting}
+
+  We implement abstract operations for accessing the list, and show the corresponding refinement lemmas:
+  \begin{lstlisting}
+    olget xs i == assert (i<|xs| & xs!i ~= None); return the (xs!i)
+    olset xs i x == assert (i<|xs| & xs!i ~= None); return the (xs[i:=x])
+
+    aget, olget : oarr$_A$ -> idx$_A$ -> id$_A$
+    aset, olset : oarr$_A$^d:p -> idx$_A$ -> id$_A$ -> oarr$_A$:r [r=p]
+  \end{lstlisting}
+  We also define conversion operations between plain lists and lists of option values:
+  \begin{lstlisting}
+    l2o xs = map Some xs                                    return, l2o : arr$_A$ -> oarr$_A$
+    o2l xs = assert (None ~: set xs); map the xs        return, o2l : oarr$_A$ -> arr$_A$
+  \end{lstlisting}
+  These conversion operations are important to limit the proof overhead when using lists of option
+  values: only where fine-grained ownership control is needed, we use option values.
+  When we are done, and have reassembled all parts of the list, we convert it back to a plain list.
+
+  Finally, we define \is{awith_idxs} and prove its refinement theorem:
+  \begin{lstlisting}
+    awith_idxs p f == f p p; return p
+
+    awith_idxs, with_idxs s :
+      oarr$_A$^d:p -> (oarr$_A$^d:p_1 -> oarr$_A$^d:p_2 -> oarr$_A$:r_1 \x oarr$_A$:r_2 [r_1=p_1 & r_2=p_2])
+      -> oarr$_A$:r [r=p]
+  \end{lstlisting}
+  Note that the set $s$ of indexes does not have a concrete counterpart.
+  It is a ghost variable that controls the split on the abstract level.
 
   \subsection{Refinement to Parallel Execution}
-  The purely functional abstract programs have no notion of parallel execution.
+  Our abstract programs have no notion of parallel execution.
   To indicate that refinement to parallel execution is desired, we define an abstract annotation
@@
     npar f g a b == x <- f a; y <- g b; return (x,y)
-
+  \end{lstlisting}
+  Its refinement rule is:
+  \begin{lstlisting}
     ${}$  hnr Ax (f$_\dagger$ x$_\dagger$) Ax' Rx CP_1 (f x) \and hnr Ay (g$_\dagger$ y$_\dagger$) Ay' Ry CP_2 (g y)"
@@
   \begin{lstlisting}
-  sort_spec xs = spec xs'. mset xs'=mset xs & sorted xs
-  \end{lstlisting}
-  That is, we return a sorted permutation of the original list.
-  Note that this is a standard specification of sorting in Isabelle.
-  Reusing the existing development of an abstract introsort algorithm~\cite{La20},
-  we easily prove with a few refinement steps that the following abstract algorithm implements \is{sort_spec}:
+  sort_spec xs == spec xs'. mset xs'=mset xs & sorted xs
+  \end{lstlisting}
+  I.e., we return a sorted permutation of the original list.
+  This is a standard specification of sorting in Isabelle, and easily proved equivalent to other, more explicit specifications:
+  \begin{lstlisting}
+    sort_spec xs = spec xs'.
+      (\<forall>x. count_list xs' x = count_list xs x) & (\<forall>i j. i<j & j<length xs' ==> xs'!j $\not<$ xs'!i)
+  \end{lstlisting}
+
+  \begin{figure}
   \begin{lstlisting}[numbers=left,xleftmargin=1.9em,frame=single,framexleftmargin=.5em]
-    psort xs n == assert n=|xs|; if n<=1 then return xs else psort_aux$\hspace*{-1pt}$ xs$\hspace*{-1pt}$ n$\hspace*{-1pt}$ (log2$\hspace*{-1pt}$ n$\hspace*{-1pt}$ *$\hspace*{-1pt}$ 2)
+    psort xs n ==
+      assert n=|xs|; if n<=1 then return xs else psort_aux$\hspace*{-1pt}$ xs$\hspace*{-1pt}$ n$\hspace*{-1pt}$ (log2$\hspace*{-1pt}$ n$\hspace*{-1pt}$ *$\hspace*{-1pt}$ 2)
 
@@
         return xs"
-
-    lemma psort xs |xs| <= sort_spec xs
-  \end{lstlisting}
+  \end{lstlisting}
+  \caption{Abstract version of our parallel quicksort algorithm.}\label{fig:psort}
+  \end{figure}
+
+  Figure~\ref{fig:psort} show our abstract parallel sorting algorithm \is{psort}.
   This algorithm is derived from the well-known quicksort and introsort algorithms~\cite{Muss97}:
@@
 
+  Reusing our existing development of an abstract introsort algorithm~\cite{La20},
+  we prove with a few refinement steps that \is{psort} implements \is{sort_spec}:
+  \begin{lstlisting}
+    psort xs |xs| <= sort_spec xs
+  \end{lstlisting}
+
   \subsection{Implementation and Correctness Theorem}\label{sec:impl_corr}
@@
   It was straightforward to amend our existing formalization of \is{pdqsort}~\cite{La20} with the in-place proofs:
-  once we had amended the refinement statements, and bug-fixed the pointer equality proving heuristics that we added to Sepref,
+  once we had amended the refinement statements and bug-fixed the pointer equality proving heuristics,
   the proofs were automatic.
 
-  Given the implementations of \is{sort_spec} and \is{partition_spec}, the Sepref tool
+  Given implementations of \is{sort_spec} and \is{partition_spec}, Sepref
   generates an LLVM program \is{psort$_\dagger$} from the abstract \is{psort}, and proves a corresponding refinement lemma:
   \begin{lstlisting}
-    hnr (arr$_A$ xs xs$_\dagger$ * idx$_A$ n n$_\dagger$) (psort$_\dagger$ xs$_\dagger$ n$_\dagger$) (idx$_A$ n n$_\dagger$) arr$_A$ (%r. r = xs$_\dagger$) (psort xs n)
+    psort$_\dagger$, psort : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> arr$_A$:r [r=xs$_\dagger$]
   \end{lstlisting}
@@
   \begin{lstlisting}
-  "ht (arr$_A$ xs xs$_\dagger$ * idx$_A$ n n$_\dagger$ * n = |xs|)
+  "ht (arr$_A$ xs$_\dagger$ xs  *  idx$_A$ n$_\dagger$ n  *  n = |xs|)
     (psort$_\dagger$ xs$_\dagger$ n$_\dagger$)
-    (%r. r=xs$_\dagger$ * \<exists> xs'. arr$_A$ xs' xs$_\dagger$ * sorted xs' * mset xs' = mset xs)"
-  \end{lstlisting}
-  That is, for a pointer \is{xs$_\dagger$} to an array, whose contents are described by list \is{xs} (\is{arr$_A$}), and a fixed-size word \is{n$_\dagger$} representing the natural number \is{n} (\is{idx$_A$}),
-  which must be the number of elements in the list \is{xs}, our sorting algorithm returns the original pointer \is{xs$_\dagger$}, and the array contents are now \is{xs'}, which is sorted and a permutation of \is{xs}.
+    (%r. r=xs$_\dagger$  *  \<exists> xs'. arr$_A$ xs$_\dagger$ xs'  *  sorted xs'  *  mset xs' = mset xs)"
+  \end{lstlisting}
+  That is, for a pointer \is{xs$_\dagger$} to an array, whose content is described by list \is{xs} (\is{arr$_A$}), and a fixed-size word \is{n$_\dagger$} representing the natural number \is{n} (\is{idx$_A$}),
+  which must be the number of elements in the list \is{xs}, our sorting algorithm returns the original pointer \is{xs$_\dagger$}, and the array content now is \is{xs'}, which is sorted and a permutation of \is{xs}.
   Note that this statement uses our semantically defined Hoare triples (cf.\ Section~\ref{sec:wp_hoare}).
-  In particular, its correctness does not depend on the refinement steps, the Sepref tool, or the VCG.
+  In particular, it does not depend on the refinement steps, the Sepref tool, or the VCG.
 
@@
 
-  \subsection{A Sampling Partitioner}
+  \subsection{Sampling Pivot Selection}
   While we could simply re-use the existing partitioning algorithm from the pdqsort formalization,
@@
   We choose a sampling approach, using the median of $64$ equidistant samples as pivot.
-  The highly optimized partitioning algorithms that we use swap the pivot to the front of the partition,
-  such that we need to determine its index, rather than just its value. We simply use quicksort to find the median\footnote{
+  We simply use quicksort to find the index of the pivot\footnote{
   We leave verification of efficient median algorithms, e.g., quickselect, to future work.
@@
       fixes lt :: 'p => 'e => 'e => bool and lt$_\dagger$ :: 'p$_\dagger$ => 'e$_\dagger$ => 'e$_\dagger$ => bool
-        and par$_A$ :: 'p => 'p$_\dagger$ => assn and elem$_A$ :: 'e => 'e$_\dagger$ => assn
+        and par$_A$ :: 'p$_\dagger$ => 'p => assn and elem$_A$ :: 'e$_\dagger$ => 'e => assn
       assumes \<forall>p. weak_ordering (lt p)
-      assumes hnr (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi) (lt$_\dagger$ pi ai bi)
-        (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True) (lt p a b)
+      assumes lt$_\dagger$, lt : par$_A$ -> elem$_A$ -> elem$_A$ -> bool$_A$
   \end{lstlisting}
@@
 
-  Our sorting algorithm is developed and verified in the context of this locale (to avoid confusion, our presentation has, up to now, just used \is{<}, \is{<=}, and \is{sorted} instead of \is{lt p}, \is{le p}, and \is{sorted_wrt (le p)}).
-  To get a sorting algorithm for an actual compare function, we have to instantiate the locale,
-  providing an abstract and concrete compare function, along with a proof that the abstract function is a weak ordering,
-  and the concrete function refines the abstract one. For our example of sorting indexes into an array,
-  where the array elements are, themselves, compared by a parametrized function \is{lt}, we get:
-  \begin{lstlisting}
-    interpretation idx: pcmp lt_idx lt_idx$_\dagger$ (par$_A$ \<times> arr$_A$) idx$_A$   \<proof>
-
-    lt_idx (p,xs) i j == lt p (xs!i) (xs!j)
-    lt_idx$_\dagger$ (p$_\dagger$,xs$_\dagger$) i$_\dagger$ j$_\dagger$ == x$_\dagger$<-aget xs$_\dagger$ i$_\dagger$; y$_\dagger$<-aget xs$_\dagger$ j$_\dagger$; lt$_\dagger$ p$_\dagger$ x$_\dagger$ y$_\dagger$
-  \end{lstlisting}
-  this yields sorting algorithms for sorting indexes, taking an extra parameter for the array to index into.
+  Our sorting algorithms are developed and verified in the context of this locale (to avoid confusion, our presentation has, up to now, just used \is{<}, \is{<=}, and \is{sorted} instead of \is{lt p}, \is{le p}, and \is{sorted_wrt (le p)}).
+  To get an actual sorting algorithm, we instantiate the locale with an abstract and concrete compare function, proving
+  that the abstract function is a weak ordering, and that the concrete function refines the abstract one.
+  For our example of sorting indexes into an array, where the array elements themselves are compared by a function \is{lt}, we get:
+  \begin{lstlisting}
+    interpretation idx: pcmp lt_idx lt_idx$_\dagger$ arr$_A$ idx$_A$   \<proof>
+
+    lt_idx xs i j == lt (xs!i) (xs!j)
+    lt_idx$_\dagger$ xs$_\dagger$ i$_\dagger$ j$_\dagger$ == x$_\dagger$<-aget xs$_\dagger$ i$_\dagger$; y$_\dagger$<-aget xs$_\dagger$ j$_\dagger$; lt$_\dagger$ x$_\dagger$ y$_\dagger$
+  \end{lstlisting}
+  This yields sorting algorithms for sorting indexes, taking an extra parameter for the array to index into.
   For our sampling application, we use \is{idx.introsort xs}.
+
+  \subsection{Parallel Partitioning}\label{sec:ppar}
+  While our parallel quicksort scheme parallelizes the sorting, partitioning is still a bottleneck:
+  before the first thread is even spawned, the whole array needs to be partitioned. On the next recursion level,
+  only two partitionings can run in parallel, and so on. That is, initially,
+  most processors will be idle. To this end, the partitioning itself can be parallelized.
+  The parallel partitioning algorithms used in the latest research on practically efficient sorting
+  algorithms~\cite{AWFS22} are branchless $k$-way algorithms, which use atomic operations to orchestrate the parallel threads.
+  In contrast, we \emph{only} verify a 2-way partitioning algorithm that uses parallel calls as its only synchronization mechanism.
+  This is a compromise between verification effort and efficiency, taking into account the features currently supported by Isabelle-LLVM.
+  The idea of our parallel partitioning algorithm is sketched in Figure~\ref{fig:ppar}.
+
+  \begin{figure}
+    \begin{tikzpicture}[
+      node distance = 3mm,
+      base/.style={rectangle, draw, minimum height = .7cm},
+      small/.style={base,fill=blue!30},
+      big/.style={base,fill=red!30},
+    ]
+      % 1)
+      \node (1xl) {1)};
+      \node[base, right =     of 1xl, minimum width = 4cm] (1x1) {};
+      \node[base, right = 0cm of 1x1, minimum width = 4cm] (1x2) {};
+      \node[base, right = 0cm of 1x2, minimum width = 4cm] (1x3) {};
+
+      % 2)
+      \node[below = 1cm of 1xl] (2xl) {2)};
+      \node[small, right =     of 2xl , minimum width = 2cm] (2x1s) {$\le p$};
+      \node[big  , right = 0cm of 2x1s, minimum width = 2cm] (2x1b) {$\ge p$};
+
+      \node[small, right = 0cm of 2x1b, minimum width = 2.5cm] (2x2s) {$\le p$};
+      \node[big  , right = 0cm of 2x2s, minimum width = 1.5cm] (2x2b) {$\ge p$};
+
+      \node[small, right = 0cm of 2x2b, minimum width = 1.5cm] (2x3s) {$\le p$};
+      \node[big  , right = 0cm of 2x3s, minimum width = 2.5cm] (2x3b) {$\ge p$};
+
+      % 3)
+      \node[below = 1cm of 2xl] (3xl) {3)};
+      \node[base , right =     of 3xl , minimum width = 2cm] (3x1s) {};
+      \node[big  , right = 0cm of 3x1s, minimum width = 2cm] (3x1b) {$bs$};
+
+      \node[base , right = 0cm of 3x1b, minimum width = 2cm]  (3x2s) {};
+        % m
+      \node[small, right = 0cm of 3x2s, minimum width = .5cm] (3x2sx) {};
+      \node[at = (3x2sx.center), anchor=center] {$ss$};
+
+      \node[base , right = 0cm of 3x2sx, minimum width = 1.5cm] (3x2b) {};
+
+      \node[small, right = 0cm of 3x2b, minimum width = 1.5cm] (3x3s) {$ss$};
+      \node[base , right = 0cm of 3x3s, minimum width = 2.5cm] (3x3b) {};
+
+      \draw ($ (3x2sx.north west) + (0,.3cm) $) -- ($ (3x2sx.south west) - (0,.3cm) $);
+      \node [ above = 2mm of 3x2sx ] {$m$};
+
+      % 4)
+      \node[below = 1cm of 3xl] (4xl) {4)};
+      \node[base , right =     of 4xl , minimum width = 2cm] (4x1s) {};
+      \node[small, right = 0cm of 4x1s, minimum width = 2cm] (4x1b) {};
+
+      \node[base , right = 0cm of 4x1b, minimum width = 2cm]  (4x2s) {};
+        % m
+      \node[big  , right = 0cm of 4x2s, minimum width = .5cm] (4x2sx) {};
+      \node[base , right = 0cm of 4x2sx, minimum width = 1.5cm] (4x2b) {};
+
+      \node[big  , right = 0cm of 4x2b, minimum width = 1.5cm] (4x3s) {};
+      \node[base , right = 0cm of 4x3s, minimum width = 2.5cm] (4x3b) {};
+
+      \draw ($ (4x1b.north west) + (.5cm,0) $) -- ($ (4x1b.south west) + (.5cm,0) $);
+
+      \coordinate (swap1left) at ($ (4x1b.north west) + (.25,0) $);
+      \draw[<->,thick]
+        (swap1left)
+        -- ($ (swap1left) + (0,.2) $)
+        -- ($ (4x2sx.north) + (0,.2) $)
+        -- (4x2sx.north)
+        ;
+
+      \coordinate (swap2left) at ($ (4x1b.north west) + (1.25,0) $);
+      \draw[<->,thick]
+        (swap2left)
+        -- ($ (swap2left) + (0,.5) $)
+        -- ($ (4x3s.north) + (0,.5) $)
+        -- (4x3s.north)
+        ;
+
+    \end{tikzpicture}
+
+    \caption{Illustration of the phases of our parallel partitioning algorithm, after a pivot $p$ has been picked.
+      Step~1 splits the array into slices. Step~2 partitions the slices in parallel. Step~3 computes the start
+      index $m$ of the right partition as the sum of the sizes of all left partitions. It then determines the indexes
+      of misplaced elements: the set $bs$ contains right-partition elements that are left of $m$,
+      and the set $ss$ contains left partition elements that are right of $m$. Step~4 then swaps the misplaced elements in parallel.
+      While Steps~1 and~3 are computationally cheap, Steps~2 and~4 do the main part of the work in parallel.
+%
+%       1) the array is split into slices,
+%       2) the slices are partitioned in parallel. This results into the array consisting of intervals
+%       of \emph{small} ($\le p$) and \emph{big} ($\ge p$) elements.
+%       3) the size $m$ of the left partition is computed as the sum of all small intervals.
+%         This is also the index where the right partition will start in the final result.
+%         Then, two sets $ss$ and $bs$ of index intervals are computed, such that $ss$ contains the indexes
+%         of small elements that are right of $m$, and $bs$ contains the indexes of big elements left of $m$.
+%       4) The elements in $ss$ and $bs$ are swapped in parallel.
+    }\label{fig:ppar}
+  \end{figure}
+
+  We specify our algorithm on sets of indexes, and then refine it to intervals and interval arrays (cf. Section~\ref{sec:mod_ds}).
+  First, we specify Steps~1 and~2, i.e., returning a permutation of the list, along with the sets \is$ls$ of
+  indexes that belong to a left partition, and \is$rs$ of indexes that belong to a right partition. In Figure~\ref{fig:ppar},
+  \is{ls} corresponds to the set of blue ($\le p$) intervals, and \is$rs$ to the set of red ($\ge p$) intervals:
+  \begin{lstlisting}
+    ppart_spec p xs == spec (xs',ls,rs).
+      mset xs' = mset xs
+    & ls \union rs = {0..<|xs'|} & ls \inter rs = {}
+    & (\<forall>i\<in>ls. xs'!i <= p) & (\<forall>i\<in>rs. xs'!i <= p)
+  \end{lstlisting}
+
+  The whole partitioning algorithm is specified as follows:
+  \begin{lstlisting}
+    ppart1 p xs ==
+      (xs,ls,rs) <- ppart_spec p xs;   --"\rightcomment{Step 1 and 2}"
+      let m = |ls|; ss = {i\<in>ls. m<=i}; bs = {i\<in>rs. i<m};   --"\rightcomment{Step 3}"
+      xs <- swap_spec ss bs xs;        --"\rightcomment{Step 4}"
+      return (m,xs)
+  \end{lstlisting}
+  A straightforward proof, only using arguments on lists and sets of indexes, shows that this algorithm partitions the list:
+  \begin{lstlisting}
+  ppart1 p xs <= spec (m, xs').
+      mset xs' = mset xs & m <= |xs'| & (\<forall>i<m. xs'!i <= p) & (\<forall>i\<in>{m..<|xs'|}. p <= xs'!i)"
+  \end{lstlisting}
+
+  The set operations in Step~3 are implemented by the operations \is{ivo_inter_ivl$_\dagger$} and \is{iv_inter_ivl$_\dagger$} of our interval arrays (cf.~Section~\ref{sec:mod_ds}),
+  using the equalities: \is$ {i\<in>ls. m<=i} = {m..} \inter ls$ and \is${i\<in>rs. i<m} = {0..<m} \inter rs$.
+  Thus, we are only missing implementations for \is{ppart_spec} and \is{swap_spec}.
+
+  The refinement of the parallel partitioning \is{ppart_spec} is similar to that of the parallel sorting algorithm \is{psort}:
+  using \is{with_split} and \is{npar}, we split the list into slices that we then partition with a sequential algorithm.
+%   Another technical difficulty we had to solve is the representation of the pivot element: in our original sorting algorithm formalizations, we have always swapped the pivot to the first element of the array, and only passed the index of the pivot element to the partitioning procedure.
+Now, we actually propagate copies of the pivot element itself to the different parallel partitioning threads.
+This means that the element type must provide a (deep) copy function.
+While this is trivial for integers, we had to implement and prove correct such a function for strings.
+
+  The parallel swapping algorithm is refined as follows:
+  \begin{lstlisting}
+  par_swap ss bs xs ==
+    if (ss={}) then return xs
+    else xs <- l2o xs; xs <- par_swap_aux ss bs xs; xs <- o2l xs; return xs
+
+  par_swap_aux ss bs xs ==
+    ((ss_1,ss_2),(bs_1,bs_2))
+      <- spec ((ss_1,ss_2),(bs_1,bs_2)). ss=ss_1$\dot\cup$ss_2 & bs=bs_1$\dot\cup$bs_2 & |ss_1|=|bs_1| & ss_1~={}
+
+    if (ss_2={}) then swap_opt_spec ss_1 bs_1 xs
+    else
+      with_idxs (ss_1 \union bs_1) xs (%xs_1 xs_2.
+        (xs_1,xs_2) <- npar swap_opt_spec par_swap_aux (ss_1,bs_1,xs_1) (ss_2,bs_2,xs_2);
+        return (xs_1,xs_2)
+  \end{lstlisting}
+  The main procedure \is{par_swap} first checks if there are any indexes to swap.
+  Then, it converts the plain list to a list of option values (\is{l2o}), invokes the actual parallel swapping procedure \is{par_swap_aux},
+  and converts the result back to a plain list (\is{o2l}).
+
+  The \is{par_swap_aux} procedure first splits off equally sized, non-empty sets \is{ss_1} and \is{bs_1} from the index sets,
+  and then swaps these in parallel with the rest. Here, \is{swap_opt_spec} is the lifting of \is{swap_spec} to lists of option values,
+  and \is{with_idxs} ensures that the swap operation will own the necessary elements.
+
+  We prove that our algorithm is correct (\is{par_swap ss bs xs <= swap_spec ss bs xs}),
+  and use Sepref, a straightforward implementation of sequential swapping, and our interval array implementation
+  to refine it to efficient imperative Isabelle-LLVM code.
+
+  Combining all refinements in this section gives us a parallel partitioning algorithm. When we wanted to show that it
+  satisfies the specification \is{partition_spec} as required by our parallel sorting algorithm, we discovered that we
+  also need to prove that neither partition can be empty.
+While this is certainly possible along the same lines as it is proved for sequential partitioning, we chose a pragmatic solution here: we dynamically check for the extreme case of
+  one partition being empty, and fix that with an additional swap. The runtime impact of this check is negligible, but it greatly simplifies the correctness proof.
 
@@
   \begin{lstlisting}
-    interpretation unat: pcmp (%_. <) (%_. ll_icmp_ult) unat$^{64}_A$ \<proof>
-    interpretation str: pcmp (%_. <) (%_. strcmp) str$^{64}_A$ \<proof>
-  \end{lstlisting}
-  This yields implementations \is{unat.psort$_\dagger$} and \is{str.psort$_\dagger$}, and automatically
-  proves instantiated versions of the correctness theorems.
-
-  In a last step, we use our code generator to generate actual LLVM text, as well as a C header file with the
+    interpretation unat: cmp (%_. <) (%_. ll_icmp_ult) unat$^{64}_A$ \<proof>
+    interpretation str: cmp (%_. <) (%_. strcmp) str$^{64}_A$ \<proof>
+  \end{lstlisting}
+  Here, the locale \is{cmp} is the version of \is{pcmp} without an extra parameter to the compare function\footnote{
+    Parameters to the compare function are currently not supported for parallel sorting algorithms, as we cannot
+    efficiently share the parameter between multiple threads. Integrating fractional separation logic into Sepref,
+    which would enable such a sharing, is left to future work. }.
+  This yields implementations \is{unat.psort$_\dagger$} and \is{str.psort$_\dagger$}, and instantiated versions of the correctness theorem.
+
+  We then use our code generator to generate actual LLVM text, as well as a C header file with the
   signatures of the generated functions\footnote{
@@
       str.psort$_\dagger$ is "llstring* str_psort(llstring*, int64_t)"
-      defines typedef struct {int64_t size; struct {int64_t capacity; char *data;};} llstring;
+      defines
+        $\textrm{\tt typedef}$ $\textrm{\tt struct}$ {int64_t size; $\textrm{\tt struct}$ {int64_t capacity; char *data;};} llstring;
       file psort.ll
@@
       title={\large #1},
-      title style={at={(0.1,.75)}},
+      title style={at={(0.12,.75)}},
       ybar=0pt,
@@
     \plottim{Laptop, uint64}{data_laptop_uint64}{xticklabels={}}{\legend{}}
-    \plottim{Server, uint64}{data_server_uint64}{xticklabels={}}{\legend{}}
+    \plottim{Server, uint64}{data_server_uint64}{xticklabels={},ymax=1000}{\legend{}}
     \plottim{Laptop, string}{data_laptop_llstring}{xticklabels={}}{\legend{}}
@@
         cells={anchor=west},
-        font=\scriptsize
+        font=\tiny
       },
@@
 
-  \caption{Runtimes in milliseconds for sorting various distributions of unsigned 64 bit integers and strings with
+  \caption{Runtimes in milliseconds for sorting various distributions of $10^8$ unsigned 64 bit integers and $10^7$ strings with
     our verified parallel sorting algorithm, C++'s standard parallel sorting algorithm, and Boost's parallel sample sort algorithm.
@@
       legend style = {
-        at = {(.87,.3)},
+        at = {(.97,.4)},
         cells={anchor=west},
-        font=\scriptsize
+        font=\footnotesize
       },
       ymajorgrids,
-      title={\large Laptop},
-      title style={at={(0.15,.85)}},
+      xmin=1, ymin=1, xmax=12, ymax=8,
+      title={\large Laptop (u64)},
+      title style={at={(0.25,.75)}},
       width=.54\textwidth,
@@
     ]
-      \input{sudata_laptop}
+      \input{sudata_uint64_laptop}
 
@@
       ymajorgrids,
-      title={\large Server},
-      title style={at={(0.15,.85)}},
+      xmin=1, ymin=1, xmax=12, ymax=8,
+      title={\large Laptop (str)},
+      title style={at={(0.25,.75)}},
       width=.54\textwidth,
@@
     ]
-      \input{sudata_server}
+      \input{sudata_llstring_laptop}
 
@@
 
+      \legend{}
+
+%       \addplot[black, thin, sharp plot, update limits=false] {\x};
+    \end{axis}
+  \end{tikzpicture}\\
+  \begin{tikzpicture}
+    \begin{axis}[
+      xlabel near ticks,
+      legend style = {
+        at = {(.87,.3)},
+        cells={anchor=west},
+        font=\scriptsize
+      },
+      ymajorgrids,
+      xmin=1, ymin=1, xmax=22, ymax=15,
+      title={\large Server (u64)},
+      title style={at={(0.25,.75)}},
+      width=.54\textwidth,
+      height=.3\textheight
+    ]
+      \input{sudata_uint64_server}
+
+      \draw[black, thin, sharp plot]
+      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
+      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});
+
+      \legend{}
+
+%       \addplot[black, thin, sharp plot, update limits=false] {\x};
+    \end{axis}
+  \end{tikzpicture}\hfill
+  \begin{tikzpicture}
+    \begin{axis}[
+      xlabel near ticks,
+      legend style = {
+        at = {(.97,.4)},
+        cells={anchor=west},
+        font=\footnotesize
+      },
+      ymajorgrids,
+      xmin=1, ymin=1, xmax=22, ymax=15,
+      title={\large Server (str)},
+      title style={at={(0.25,.75)}},
+      width=.54\textwidth,
+      height=.3\textheight
+    ]
+      \input{sudata_llstring_server}
+
+      \draw[black, thin, sharp plot]
+      (axis cs:\pgfkeysvalueof{/pgfplots/xmin},\pgfkeysvalueof{/pgfplots/xmin}) --
+      (axis cs:\pgfkeysvalueof{/pgfplots/xmax},\pgfkeysvalueof{/pgfplots/xmax});
+
+      \legend{}
+
 %       \addplot[black, thin, sharp plot, update limits=false] {\x};
@@
   \end{tikzpicture}
-  \caption{Speedup of the various implementations, for sorting unsigned 64 bit integers with a random distribution, on
-    a server with 22 AMD Opteron 6176 cores and 128GiB of RAM, and a laptop with a
-    6 core (12 threads) i7-10750H CPU and 32GiB of RAM.
+  \caption{Speedup of the various implementations for sorting $10^8$ integers and $10^7$ strings with a random distribution.
     The x axis ranges over the number of cores, and the y-axis gives the speedup wrt.\ the same implementation run on only one core.
@@
 
+  \begin{figure}
+
+  \begin{tikzpicture}
+    \begin{loglogaxis}[
+      xlabel near ticks,
+      legend style = {
+        at = {(.97,.4)},
+        cells={anchor=west},
+        font=\footnotesize
+      },
+      ymajorgrids,
+      xmin=10^4,xmax=10^8,
+      title={\large Server (u64)},
+      title style={at={(0.25,.75)}},
+      width=.54\textwidth,
+      height=.3\textheight
+    ]
+      \input{sadata_uint64_server}
+    \end{loglogaxis}
+  \end{tikzpicture}\hfill
+  \begin{tikzpicture}
+    \begin{loglogaxis}[
+      xlabel near ticks,
+      legend style = {
+        at = {(.97,.4)},
+        cells={anchor=west},
+        font=\footnotesize
+      },
+      ymajorgrids,
+      xmin=10^3,xmax=10^7,
+      title={\large Server (str)},
+      title style={at={(0.25,.75)}},
+      width=.54\textwidth,
+      height=.3\textheight
+    ]
+      \input{sadata_llstring_server}
+      \legend{}
+    \end{loglogaxis}
+  \end{tikzpicture}
+  \caption{Performance for sorting small arrays of randomly distributed integers and strings.
+    The $y$-axis shows the runtime in milliseconds, and the $x$-axis the array size. Note that both axis are logarithmic.
+  }\label{fig:smallarrays}
+
+  \end{figure}
+
   We have benchmarked our verified sorting algorithm against a direct implementation of the same algorithm in C++.
@@
   and \is{sample_sort} from the Boost C++ libraries~\cite{boost,boost-sort}.
-  We have benchmarked the algorithm on two different machines, and various input distributions. The results are shown in Figure~\ref{fig:benchres}.
-  While our verified algorithm is clearly competitive for integer sorting on the less parallel laptop
-  machine, it's slightly less efficient for sorting strings on the highly parallel server machine.
-  Nevertheless, we believe that our verified implementation is already useful in practice,
-  and leave further optimizations to future work.
-
-  Finally, we measured the speedup that the implementations achieve for a certain number of cores.
-  The results are displayed in Figure~\ref{fig:speedup}. While the speedup on the moderately parallel
-  laptop is comparable to the one of the C++ standard library, our implementation
-  achieves lower speedups than the state-of-the-art on the highly parallel server.
-  Again, we leave further optimizations to future work.
+  We have benchmarked the algorithm on two different machines, and various input distributions. The results are shown in Figure~\ref{fig:benchres}:
+  our verified algorithm is clearly competitive with the unverified state-of-the-art implementations.
+  Only for a few string-sorting benchmarks, it is slightly slower. We leave improving on this for future work.
+%
+%
+%   While our verified algorithm is clearly competitive for integer sorting on the less parallel laptop
+%   machine, it's slightly less efficient for sorting strings on the highly parallel server machine.
+%   Nevertheless, we believe that our verified implementation is already useful in practice,
+%   and leave further optimizations to future work.
+
+  We also measured the speedup that the implementations achieve for a certain number of cores.
+  The results are displayed in Figure~\ref{fig:speedup}: again, our verified implementation is clearly competitive.
+
+  The previous two benchmarks used relatively large input sizes.
+  Figure~\ref{fig:smallarrays} displays a benchmark for smaller input sizes. While we are still competitive with \is{sample_sort},
+  \is{std::sort} is clearly faster for small arrays. This result is expected: our parallelization uses hard-coded thresholds
+  to switch to sequential algorithms, which are independent of the number of available processors or the input size.
+  We leave fine-tuning of the parallelization scheme to future work.
 
@@
     the LLVM semantics, and the code generator.
-    All other tools that we used, e.g., our Hoare logic, Sepref tool, and Refinement Framework for abstract programs,
+    All other tools that we used, e.g., our Hoare logic, the Sepref tool, and the Refinement Framework for abstract programs,
     ultimately prove a correctness theorem that only depends on our shallowly embedded semantics.
@@
     It uses an existing verified sequential pdqsort algorithm as a building block,
-    and is competitive with state-of-the-art parallel sorting algorithms,
-    at least on moderately parallel hardware.
+    and is competitive with state-of-the-art parallel sorting algorithms.
 
@@
 
+    While the verification of our sorting algorithms uses a top-down approach, we actually started with
+    implementing, benchmarking, and fine-tuning the algorithms in C++. This gave us a quick option to
+    find an algorithm that is efficient, and, at the same time not too complex for verification,
+    without having to verify each intermediate step towards this algorithm.
+    Only then we used our top-down approach to first formalize the abstract ideas behind the algorithm,
+    and then refine it to an efficient implementation close to what we had written in C++.
+    At this point, one may ask why not directly verify the C++ implementation: while this might be
+    possible, the required work and steps would be similar: to manage the complexity of such a verification,
+    several bottom-up refinement steps would be necessary, ultimately arriving at something similarly abstract
+    as our initial abstract algorithm.
+
+%     The effort for our formalization is hard to estimate. We have no precise records how long it
+%     took us to formalize the parallel extension to Isabelle LLVM: the initial idea was born when trying to solve
+%     Challenge~3 from the VerifyThis 2021 competition~\cite{VerifyThis21}.
+%     After abandoning an initial formalization for the Imperative/HOL backend, because the achievable performance
+%     gain was too low, we transferred the idea to the LLVM backend, using our existing sequential sorting algorithms.
+%     The extension to parallel partitioning
+%
+%
+%     The implementation of
+%
+%     formalizing the idea for the Imperative/HOL
+%
+%
+%     The initial formalization of the sequential sorting algorithms
+%     took roughly 130 person hours, and 8700 lines of Isabelle code.
+%
+%
+
 %     In our approach, the high-level algorithms are phrased as purely functional and sequential algorithms,
@@
     \subsection{Related Work}
-    While there is extensive work on parallel sorting algorithm (e.g.~\cite{CNLM08,AMI21}),
+    While there is extensive work on parallel sorting algorithm (e.g.~\cite{AWFS22,CNLM08,AMI21}),
     there seems to be almost no work on their formal verification. The only work we are aware of is
@@
 
-      Another obvious next step is to verify a state-of-the-art parallel sorting algorithm,
-      like Boost's sample sort. Like our current algorithm, sample sort does not require
-      advanced synchronization concepts, and can be implemented only with a parallel combinator.
-
-      Finally, the Sepref framework has recently been extended to reason about complexity of (sequential)
-      LLVM programs~\cite{HaLa21-toplas,HaLa21}. This line of work could be combined with our parallel extension, to verify the
-      complexity (e.g. work and span) of parallel algorithms.
-
       Extending our approach towards more advanced
@@
       a thread could report a set of possible traces, which are checked for race-freedom and then combined.
-
-      Finally, our framework currently targets multicore CPUs. Another important architecture are general purpose GPUs.
+      Moreover, our framework currently targets multicore CPUs. Another important architecture are general purpose GPUs.
       As LLVM is also available for GPUs, porting our framework to this architecture should be possible.
-      We even expect that barrier synchronization, which is important in the GPU context, can be integrated into our approach.
+      We even expect that we can model barrier synchronization, which is important in the GPU context.
+
+      Finally, the Sepref framework has recently been extended to reason about complexity of (sequential)
+      LLVM programs~\cite{HaLa21-toplas,HaLa21}. This could be combined with our parallel extension, to verify the
+      complexity (e.g. work and span) of parallel algorithms.
+
+      Another direction for future work is to further optimize our verified sorting algorithm.
+      We expect that tuning our parallelization scheme will improve the speedup for smaller input sizes.
+      Also, while we are competitive with standard library implementations, recent research indicates that there is still
+      some room for improvement, for example with the IPS$^4$o algorithm \cite{AWFS22}. While this algorithm uses atomic operations
+      in one place, many other of its optimizations, for example branchless decision trees for multi-way partitioning,
+      only require features already supported by our framework.
 
@@
 
-\clearpage
+% \clearpage
 

### 2023-03-23 papers/JAR_SI_ITP2022/sadata_llstring_laptop.tex
JAR SI ITP paper

\addplot+[color=red, mark color=black, mark options={fill=red}] coordinates {
  (1000, 0.07)
  (2783, 0.26)
  (7743, 0.85)
  (21544, 2.71)
  (59948, 8.49)
  (166810, 16.94)
  (464159, 33.00)
  (1291550, 96.47)
  (3593814, 295.55)
  (10000000, 898.21)
};
\addlegendentry{verified};

\addplot+[color=black, mark color=black, mark options={fill=black}] coordinates {
  (1000, 0.08)
  (2783, 0.26)
  (7743, 0.81)
  (21544, 2.51)
  (59948, 8.00)
  (166810, 6.38)
  (464159, 25.89)
  (1291550, 86.31)
  (3593814, 294.83)
  (10000000, 962.54)
};
\addlegendentry{sample sort};

\addplot+[color=blue, mark color=black, mark options={fill=blue}] coordinates {
  (1000, 0.06)
  (2783, 0.10)
  (7743, 0.20)
  (21544, 0.49)
  (59948, 1.23)
  (166810, 5.23)
  (464159, 23.66)
  (1291550, 82.55)
  (3593814, 300.77)
  (10000000, 1015.65)
};
\addlegendentry{std::sort};

### 2023-03-23 papers/JAR_SI_ITP2022/sadata_llstring_server.tex
JAR SI ITP paper

\addplot+[color=red, mark color=black, mark options={fill=red}] coordinates {
  (1000, 0.18)
  (2783, 0.25)
  (7743, 0.84)
  (21544, 2.71)
  (59948, 8.61)
  (166810, 16.24)
  (464159, 27.09)
  (1291550, 71.66)
  (3593814, 163.82)
  (10000000, 416.31)
};
\addlegendentry{verified};

\addplot+[color=black, mark color=black, mark options={fill=black}] coordinates {
  (1000, 0.17)
  (2783, 0.56)
  (7743, 0.85)
  (21544, 2.79)
  (59948, 8.62)
  (166810, 31.70)
  (464159, 39.46)
  (1291550, 86.36)
  (3593814, 189.04)
  (10000000, 470.72)
};
\addlegendentry{sample sort};

\addplot+[color=blue, mark color=black, mark options={fill=blue}] coordinates {
  (1000, 0.13)
  (2783, 0.19)
  (7743, 0.38)
  (21544, 0.63)
  (59948, 0.90)
  (166810, 2.14)
  (464159, 7.00)
  (1291550, 34.22)
  (3593814, 128.24)
  (10000000, 461.74)
};
\addlegendentry{std::sort};

### 2023-03-23 papers/JAR_SI_ITP2022/sadata_uint64_laptop.tex
JAR SI ITP paper

\addplot+[color=red, mark color=black, mark options={fill=red}] coordinates {
  (10000, 0.48)
  (27826, 1.46)
  (77426, 4.05)
  (215443, 4.71)
  (599484, 9.13)
  (1668101, 19.85)
  (4641589, 45.21)
  (12915497, 138.15)
  (35938137, 391.04)
  (100000000, 1206.60)
};
\addlegendentry{verified};

\addplot+[color=black, mark color=black, mark options={fill=black}] coordinates {
  (10000, 0.34)
  (27826, 0.96)
  (77426, 0.95)
  (215443, 1.81)
  (599484, 4.12)
  (1668101, 12.44)
  (4641589, 40.17)
  (12915497, 122.38)
  (35938137, 352.97)
  (100000000, 1040.94)
};
\addlegendentry{sample sort};

\addplot+[color=blue, mark color=black, mark options={fill=blue}] coordinates {
  (10000, 0.09)
  (27826, 0.23)
  (77426, 0.61)
  (215443, 1.78)
  (599484, 5.47)
  (1668101, 16.12)
  (4641589, 51.32)
  (12915497, 159.44)
  (35938137, 491.69)
  (100000000, 1426.85)
};
\addlegendentry{std::sort};

### 2023-03-23 papers/JAR_SI_ITP2022/sadata_uint64_server.tex
JAR SI ITP paper

\addplot+[color=red, mark color=black, mark options={fill=red}] coordinates {
  (10000, 0.39)
  (27826, 1.21)
  (77426, 3.69)
  (215443, 11.12)
  (599484, 15.01)
  (1668101, 19.28)
  (4641589, 34.43)
  (12915497, 84.48)
  (35938137, 232.61)
  (100000000, 733.26)
};
\addlegendentry{verified};

\addplot+[color=black, mark color=black, mark options={fill=black}] coordinates {
  (10000, 0.58)
  (27826, 0.94)
  (77426, 13.99)
  (215443, 17.99)
  (599484, 21.07)
  (1668101, 29.71)
  (4641589, 69.88)
  (12915497, 176.65)
  (35938137, 377.31)
  (100000000, 765.52)
};
\addlegendentry{sample sort};

\addplot+[color=blue, mark color=black, mark options={fill=blue}] coordinates {
  (10000, 0.14)
  (27826, 0.31)
  (77426, 0.74)
  (215443, 1.08)
  (599484, 2.84)
  (1668101, 7.83)
  (4641589, 25.13)
  (12915497, 95.56)
  (35938137, 277.38)
  (100000000, 821.05)
};
\addlegendentry{std::sort};

### 2023-03-23 papers/JAR_SI_ITP2022/sn-article.tex
JAR SI ITP paper

% Version 1.2 of SN LaTeX, November 2022
%
% See section 11 of the User Manual for version history 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing

%%=======================================================%%
%% to print line numbers in the margin use lineno option %%
%%=======================================================%%

%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%======================================================%%
%% to compile with pdflatex/xelatex use pdflatex option %%
%%======================================================%%

%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%Note: the following reference styles support Namedate and Numbered referencing. By default the style follows the most common style. To switch between the options you can add or remove “Numbered” in the optional parenthesis. 
%%The option is available for: sn-basic.bst, sn-vancouver.bst, sn-chicago.bst, sn-mathphys.bst. %

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
\documentclass[sn-mathphys,Numbered]{sn-jnl}% Math and Physical Sciences Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver,Numbered]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
%%\documentclass[default]{sn-jnl}% Default
%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published 
%%%%  by Springer Nature. The guidance has been prepared in partnership with 
%%%%  production teams to conform to Springer Nature technical requirements. 
%%%%  Editorial and presentation requirements differ among journal portfolios and 
%%%%  research disciplines. You may find sections in this template are irrelevant 
%%%%  to your work and are empowered to omit any such section if allowed by the 
%%%%  journal you intend to submit to. The submission guidelines and policies 
%%%%  of the journal take precedence. A detailed User Manual is available in the 
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%\jyear{2021}%

%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}% 
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

\begin{document}

\title[Article Title]{Article Title}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1,2]{\fnm{First} \sur{Author}}\email{iauthor@gmail.com}

\author[2,3]{\fnm{Second} \sur{Author}}\email{iiauthor@gmail.com}
\equalcont{These authors contributed equally to this work.}

\author[1,2]{\fnm{Third} \sur{Author}}\email{iiiauthor@gmail.com}
\equalcont{These authors contributed equally to this work.}

\affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{100190}, \state{State}, \country{Country}}}

\affil[2]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{10587}, \state{State}, \country{Country}}}

\affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}

%%==================================%%
%% sample for unstructured abstract %%
%%==================================%%

\abstract{The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.}

%%================================%%
%% Sample for structured abstract %%
%%================================%%

% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}

\keywords{keyword1, Keyword2, Keyword3, Keyword4}

%%\pacs[JEL Classification]{D8, H51}

%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}

\maketitle

\section{Introduction}\label{sec1}

The Introduction section, of referenced text \cite{bib1} expands on the background of the work (some overlap with the Abstract is acceptable). The introduction should not include subheadings.

Springer Nature does not impose a strict layout as standard however authors are advised to check the individual requirements for the journal they are planning to submit to as there may be journal-level preferences. When preparing your text please also be aware that some stylistic choices are not supported in full text XML (publication version), including coloured font. These will not be replicated in the typeset article if it is accepted.

\section{Results}\label{sec2}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\section{This is an example for first level head---section head}\label{sec3}

\subsection{This is an example for second level head---subsection head}\label{subsec2}

\subsubsection{This is an example for third level head---subsubsection head}\label{subsubsec2}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\section{Equations}\label{sec4}

Equations in \LaTeX\ can either be inline or on-a-line by itself (``display equations''). For
inline equations use the \verb+$...$+ commands. E.g.: The equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

For display equations (with auto generated equation numbers)
one can use the equation or align environments:
\begin{equation}
\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q}.\label{eq1}
\end{equation}
where,
\begin{align}
D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu \label{eq2}
\end{align}
Notice the use of \verb+\nonumber+ in the align environment at the end
of each line, except the last, so as not to produce equation numbers on
lines where no equation numbers are required. The \verb+\label{}+ command
should only be used at the last line of an align environment where
\verb+\nonumber+ is not used.
\begin{equation}
Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
    + \frac{\ln(c_2/5)}{15} \right]
\end{equation}
The class file also supports the use of \verb+\mathbb{}+, \verb+\mathscr{}+ and
\verb+\mathcal{}+ commands. As such \verb+\mathbb{R}+, \verb+\mathscr{R}+
and \verb+\mathcal{R}+ produces $\mathbb{R}$, $\mathscr{R}$ and $\mathcal{R}$
respectively (refer Subsubsection~\ref{subsubsec2}).

\section{Tables}\label{sec5}

Tables can be inserted via the normal table and tabular environment. To put
footnotes inside tables you should use \verb+\footnotetext[]{...}+ tag.
The footnote appears just below the table itself (refer Tables~\ref{tab1} and \ref{tab2}). 
For the corresponding footnotemark use \verb+\footnotemark[...]+

\begin{table}[h]
\caption{Caption text}\label{tab1}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2  & Column 3 & Column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
\end{table}

\noindent
The input format for the above table is as follows:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{verbatim}
\begin{table}[<placement-specifier>]
\caption{<table-caption>}\label{<table-label>}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2 & Column 3 & Column 4\\
\midrule
row 1 & data 1 & data 2	 & data 3 \\
row 2 & data 4 & data 5\footnotemark[1] & data 6 \\
row 3 & data 7 & data 8	 & data 9\footnotemark[2]\\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. 
This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.
This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. 
This is an example of table footnote.}
\end{table}
\end{verbatim}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

\begin{table}[h]
\caption{Example of a lengthy table which is set to full textwidth}\label{tab2}
\begin{tabular*}{\textwidth}{@{\extracolsep\fill}lcccccc}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]} & \multicolumn{3}{@{}c@{}}{Element 2\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote. This is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.}
\footnotetext[2]{Example for a second table footnote.}
\end{table}

\vfill\eject

In case of double column layout, tables which do not fit in single column width should be set to full text width. For this, you need to use \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ instead of \verb+\begin{table}+ \verb+...+ \verb+\end{table}+ environment. Lengthy tables which do not fit in textwidth should be set as rotated table. For this, you need to use \verb+\begin{sidewaystable}+ \verb+...+ \verb+\end{sidewaystable}+ instead of \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ environment. This environment puts tables rotated to single column width. For tables rotated to double column width, use \verb+\begin{sidewaystable*}+ \verb+...+ \verb+\end{sidewaystable*}+.

\begin{sidewaystable}
\caption{Tables which are too long to fit, should be written using the ``sidewaystable'' environment as shown here}\label{tab3}
\begin{tabular*}{\textheight}{@{\extracolsep\fill}lcccccc}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]}& \multicolumn{3}{@{}c@{}}{Element\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Projectile & Energy	& $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
Element 5 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 6 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{This is an example of table footnote.}
\end{sidewaystable}

\section{Figures}\label{sec6}

As per the \LaTeX\ standards you need to use eps images for \LaTeX\ compilation and \verb+pdf/jpg/png+ images for \verb+PDFLaTeX+ compilation. This is one of the major difference between \LaTeX\ and \verb+PDFLaTeX+. Each image should be from a single input .eps/vector image file. Avoid using subfigures. The command for inserting images for \LaTeX\ and \verb+PDFLaTeX+ can be generalized. The package used to insert images in \verb+LaTeX/PDFLaTeX+ is the graphicx package. Figures can be inserted via the normal figure environment as shown in the below example:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{verbatim}
\begin{figure}[<placement-specifier>]
\centering
\includegraphics{<eps-file>}
\caption{<figure-caption>}\label{<figure-label>}
\end{figure}
\end{verbatim}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

\begin{figure}[h]%
\centering
\includegraphics[width=0.9\textwidth]{fig.eps}
\caption{This is a widefig. This is an example of long caption this is an example of long caption  this is an example of long caption this is an example of long caption}\label{fig1}
\end{figure}

In case of double column layout, the above format puts figure captions/images to single column width. To get spanned images, we need to provide \verb+\begin{figure*}+ \verb+...+ \verb+\end{figure*}+.

For sample purpose, we have included the width of images in the optional argument of \verb+\includegraphics+ tag. Please ignore this.

\section{Algorithms, Program codes and Listings}\label{sec7}

Packages \verb+algorithm+, \verb+algorithmicx+ and \verb+algpseudocode+ are used for setting algorithms in \LaTeX\ using the format:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{verbatim}
\begin{algorithm}
\caption{<alg-caption>}\label{<alg-label>}
\begin{algorithmic}[1]
. . .
\end{algorithmic}
\end{algorithm}
\end{verbatim}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

You may refer above listed package documentations for more details before setting \verb+algorithm+ environment. For program codes, the ``verbatim'' package is required and the command to be used is \verb+\begin{verbatim}+ \verb+...+ \verb+\end{verbatim}+.

Similarly, for \verb+listings+, use the \verb+listings+ package. \verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ is used to set environments similar to \verb+verbatim+ environment. Refer to the \verb+lstlisting+ package documentation for more details.

A fast exponentiation procedure:

\lstset{texcl=true,basicstyle=\small\sf,commentstyle=\small\rm,mathescape=true,escapeinside={(*}{*)}}
\begin{lstlisting}
begin
  for $i:=1$ to $10$ step $1$ do
      expt($2,i$);  
      newline() od                (*\textrm{Comments will be set flush to the right margin}*)
where
proc expt($x,n$) $\equiv$
  $z:=1$;
  do if $n=0$ then exit fi;
     do if odd($n$) then exit fi;                 
        comment: (*\textrm{This is a comment statement;}*)
        $n:=n/2$; $x:=x*x$ od;
     { $n>0$ };
     $n:=n-1$; $z:=z*x$ od;
  print($z$). 
end
\end{lstlisting}

\begin{algorithm}
\caption{Calculate $y = x^n$}\label{algo1}
\begin{algorithmic}[1]
\Require $n \geq 0 \vee x \neq 0$
\Ensure $y = x^n$ 
\State $y \Leftarrow 1$
\If{$n < 0$}\label{algln2}
        \State $X \Leftarrow 1 / x$
        \State $N \Leftarrow -n$
\Else
        \State $X \Leftarrow x$
        \State $N \Leftarrow n$
\EndIf
\While{$N \neq 0$}
        \If{$N$ is even}
            \State $X \Leftarrow X \times X$
            \State $N \Leftarrow N / 2$
        \Else[$N$ is odd]
            \State $y \Leftarrow y \times X$
            \State $N \Leftarrow N - 1$
        \EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{minipage}{\hsize}%
\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}% Set your language (you can change the language for each code-block optionally)
%%% Start your code-block
\begin{lstlisting}
for i:=maxint to 0 do
begin
{ do nothing }
end;
Write('Case insensitive ');
Write('Pascal keywords.');
\end{lstlisting}
\end{minipage}

\section{Cross referencing}\label{sec8}

Environments such as figure, table, equation and align can have a label
declared via the \verb+\label{#label}+ command. For figures and table
environments use the \verb+\label{}+ command inside or just
below the \verb+\caption{}+ command. You can then use the
\verb+\ref{#label}+ command to cross-reference them. As an example, consider
the label declared for Figure~\ref{fig1} which is
\verb+\label{fig1}+. To cross-reference it, use the command 
\verb+Figure \ref{fig1}+, for which it comes up as
``Figure~\ref{fig1}''.

To reference line numbers in an algorithm, consider the label declared for the line number 2 of Algorithm~\ref{algo1} is \verb+\label{algln2}+. To cross-reference it, use the command \verb+\ref{algln2}+ for which it comes up as line~\ref{algln2} of Algorithm~\ref{algo1}.

\subsection{Details on reference citations}\label{subsec7}

Standard \LaTeX\ permits only numerical citations. To support both numerical and author-year citations this template uses \verb+natbib+ \LaTeX\ package. For style guidance please refer to the template user manual.

Here is an example for \verb+\cite{...}+: \cite{bib1}. Another example for \verb+\citep{...}+: \citep{bib2}. For author-year citation mode, \verb+\cite{...}+ prints Jones et al. (1990) and \verb+\citep{...}+ prints (Jones et al., 1990).

All cited bib entries are printed at the end of this article: \cite{bib3}, \cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, \cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.

\section{Examples for theorem like environments}\label{sec10}

For theorem like environments, we require \verb+amsthm+ package. There are three types of predefined theorem styles exists---\verb+thmstyleone+, \verb+thmstyletwo+ and \verb+thmstylethree+

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{tabular}{|l|p{19pc}|}
\hline
\verb+thmstyleone+ & Numbered, theorem head in bold font and theorem text in italic style \\\hline
\verb+thmstyletwo+ & Numbered, theorem head in roman font and theorem text in italic style \\\hline
\verb+thmstylethree+ & Numbered, theorem head in bold font and theorem text in roman style \\\hline
\end{tabular}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

For mathematics journals, theorem styles can be included as shown in the following examples:

\begin{theorem}[Theorem subhead]\label{thm1}
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text. 
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text. 
Example theorem text. 
\end{theorem}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{proposition}
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text. 
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text. 
\end{proposition}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{example}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. 
\end{example}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{remark}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. 
\end{remark}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{definition}[Definition sub head]
Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. 
\end{definition}

Additionally a predefined ``proof'' environment is available: \verb+\begin{proof}+ \verb+...+ \verb+\end{proof}+. This prints a ``Proof'' head in italic font style and the ``body text'' in roman font style with an open square at the end of each proof environment.

\begin{proof}
Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. 
\end{proof}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{proof}[Proof of Theorem~{\upshape\ref{thm1}}]
Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. 
\end{proof}

\noindent
For a quote environment, use \verb+\begin{quote}...\end{quote}+
\begin{quote}
Quoted text example. Aliquam porttitor quam a lacus. Praesent vel arcu ut tortor cursus volutpat. In vitae pede quis diam bibendum placerat. Fusce elementum
convallis neque. Sed dolor orci, scelerisque ac, dapibus nec, ultricies ut, mi. Duis nec dui quis leo sagittis commodo.
\end{quote}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text (refer Figure~\ref{fig1}). Sample body text. Sample body text. Sample body text (refer Table~\ref{tab3}).

\section{Methods}\label{sec11}

Topical subheadings are allowed. Authors must ensure that their Methods section includes adequate experimental and characterization data necessary for others in the field to reproduce their work. Authors are encouraged to include RIIDs where appropriate.

\textbf{Ethical approval declarations} (only required where applicable) Any article reporting experiment/s carried out on (i)~live vertebrate (or higher invertebrates), (ii)~humans or (iii)~human samples must include an unambiguous statement within the methods section that meets the following requirements:

\begin{enumerate}[1.]
\item Approval: a statement which confirms that all experimental protocols were approved by a named institutional and/or licensing committee. Please identify the approving body in the methods section

\item Accordance: a statement explicitly saying that the methods were carried out in accordance with the relevant guidelines and regulations

\item Informed consent (for experiments involving humans or human tissue samples): include a statement confirming that informed consent was obtained from all participants and/or their legal guardian/s
\end{enumerate}

If your manuscript includes potentially identifying patient/participant information, or if it describes human transplantation research, or if it reports results of a clinical trial then  additional information will be required. Please visit (\url{https://www.nature.com/nature-research/editorial-policies}) for Nature Portfolio journals, (\url{https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214}) for Springer Nature journals, or (\url{https://www.biomedcentral.com/getpublished/editorial-policies\#ethics+and+consent}) for BMC.

\section{Discussion}\label{sec12}

Discussions should be brief and focused. In some disciplines use of Discussion or `Conclusion' is interchangeable. It is not mandatory to use both. Some journals prefer a section `Results and Discussion' followed by a section `Conclusion'. Please refer to Journal-level guidance for any specific requirements.

\section{Conclusion}\label{sec13}

Conclusions may be used to restate your hypothesis or research question, restate your major findings, explain the relevance and the added value of your work, highlight any limitations of your study, describe future directions for research and recommendations.

In some disciplines use of Discussion or 'Conclusion' is interchangeable. It is not mandatory to use both. Please refer to Journal-level guidance for any specific requirements.

\backmatter

\bmhead{Supplementary information}

If your article has accompanying supplementary file/s please state so here.

Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information. This may be requested by the editorial team/s if it is missing.

Please refer to Journal-level guidance for any specific requirements.

\bmhead{Acknowledgments}

Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

\section*{Declarations}

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading `Declarations':

\begin{itemize}
\item Funding
\item Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
\item Ethics approval 
\item Consent to participate
\item Consent for publication
\item Availability of data and materials
\item Code availability 
\item Authors' contributions
\end{itemize}

\noindent
If any of the sections are not relevant to your manuscript, please include the heading and write `Not applicable' for that section.

%%===================================================%%
%% For presentation purpose, we have included        %%
%% \bigskip command. please ignore this.             %%
%%===================================================%%
\bigskip
\begin{flushleft}%
Editorial Policies for:

\bigskip\noindent
Springer journals and proceedings: \url{https://www.springer.com/gp/editorial-policies}

\bigskip\noindent
Nature Portfolio journals: \url{https://www.nature.com/nature-research/editorial-policies}

\bigskip\noindent
\textit{Scientific Reports}: \url{https://www.nature.com/srep/journal-policies/editorial-policies}

\bigskip\noindent
BMC journals: \url{https://www.biomedcentral.com/getpublished/editorial-policies}
\end{flushleft}

\begin{appendices}

\section{Section title of first appendix}\label{secA1}

An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.

%%=============================================%%
%% For submissions to Nature Portfolio Journals %%
%% please use the heading ``Extended Data''.   %%
%%=============================================%%

%%=============================================================%%
%% Sample for another appendix section			       %%
%%=============================================================%%

%% \section{Example of another appendix section}\label{secA2}%
%% Appendices may be used for helpful, supporting or essential material that would otherwise 
%% clutter, break up or be distracting to the text. Appendices can consist of sections, figures, 
%% tables and equations etc.

\end{appendices}

%%===========================================================================================%%
%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
%% system, please include the references within the manuscript file itself. You may do this  %%
%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
%% file, and delete the associated \verb+\bibliography+ commands.                            %%
%%===========================================================================================%%

\bibliography{sn-bibliography}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl

\end{document}

### 2023-03-23 papers/JAR_SI_ITP2022/sudata_llstring_laptop.tex
JAR SI ITP paper

\addplot+[color=red, mark color=black, mark options={fill=red}] coordinates {
  (1, 1.00)
  (2, 1.83)
  (3, 2.52)
  (4, 3.03)
  (5, 3.42)
  (6, 3.67)
  (7, 3.92)
  (8, 3.95)
  (9, 4.20)
  (10, 4.19)
  (11, 4.54)
  (12, 4.65)
};
\addlegendentry{verified};

\addplot+[color=black, mark color=black, mark options={fill=black}] coordinates {
  (1, 1.00)
  (2, 1.92)
  (3, 2.72)
  (4, 3.37)
  (5, 3.90)
  (6, 4.17)
  (7, 4.30)
  (8, 4.43)
  (9, 4.58)
  (10, 4.69)
  (11, 4.71)
  (12, 4.79)
};
\addlegendentry{sample sort};

\addplot+[color=blue, mark color=black, mark options={fill=blue}] coordinates {
  (1, 1.00)
  (2, 1.75)
  (3, 2.37)
  (4, 2.84)
  (5, 3.17)
  (6, 3.18)
  (7, 3.48)
  (8, 3.52)
  (9, 3.74)
  (10, 3.87)
  (11, 4.08)
  (12, 4.14)
};
\addlegendentry{std::sort};

### 2023-03-23 papers/JAR_SI_ITP2022/sudata_uint64_laptop.tex
JAR SI ITP paper

\addplot+[color=red, mark color=black, mark options={fill=red}] coordinates {
  (1, 1.00)
  (2, 1.81)
  (3, 2.68)
  (4, 3.52)
  (5, 4.35)
  (6, 5.15)
  (7, 5.51)
  (8, 5.95)
  (9, 6.14)
  (10, 6.34)
  (11, 6.87)
  (12, 6.72)
};
\addlegendentry{verified};

\addplot+[color=black, mark color=black, mark options={fill=black}] coordinates {
  (1, 1.00)
  (2, 1.92)
  (3, 2.84)
  (4, 3.73)
  (5, 4.55)
  (6, 5.31)
  (7, 5.79)
  (8, 6.20)
  (9, 6.29)
  (10, 6.42)
  (11, 6.60)
  (12, 6.85)
};
\addlegendentry{sample sort};

\addplot+[color=blue, mark color=black, mark options={fill=blue}] coordinates {
  (1, 1.00)
  (2, 1.93)
  (3, 2.86)
  (4, 3.79)
  (5, 4.70)
  (6, 5.55)
  (7, 5.99)
  (8, 6.38)
  (9, 6.74)
  (10, 7.11)
  (11, 7.35)
  (12, 7.61)
};
\addlegendentry{std::sort};

### 2023-04-11 papers/JAR_SI_ITP2022/main.tex
upd paper

@@
 
-On top of the back-end, a program logic is used to prove programs correct.
-It uses separation logic, and provides automation like a verification condition generator (VCG).
-In Section~\ref{sec:seplogic}, we describe our formalization of concurrent separation logic~\cite{OH04}, and our VCG.
+On top of the back-end, we formalize a concurrent separation logic~\cite{OH04} and implement a verification condition generator (VCG). (Section~\ref{sec:seplogic}).
 
@@
 refined to optimized efficient code. The stepwise refinement ensures a separation of concerns
-between high-level algorithmic ideas and low-level optimizations. We have used this approach to
-verify a wide range of practically efficient algorithms~\cite{ELNN13,BrLa18,WiLa18,La17_CADE,La17_SAT,FBL18,La14,LaSe16,LaSe19,La20}.
-In Section~\ref{sec:parsort}, we use our techniques to verify a parallel sorting algorithm,
+between high-level algorithmic ideas and low-level optimizations. We have already used this approach to
+verify a wide range of practically efficient sequential algorithms~\cite{ELNN13,BrLa18,WiLa18,La17_CADE,La17_SAT,FBL18,La14,LaSe16,LaSe19,La20}.
+In Section~\ref{sec:parsort}, we use our extended techniques to verify a parallel sorting algorithm,
 with competitive performance wrt.\ unverified state-of-the-art algorithms.
@@
 \subsection{Notation}
+Formal statements in this paper correspond to theorems proved in our Isabelle/HOL formalization,
+though sometimes simplified to improve the clarity of the presentation.
+
 We mainly use Isabelle/HOL notation, with some shortcuts and adaptations for presentation in a paper.
 In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
-e.g., \is{\<forall>x. x<4 ==> x<=3} (in Isabelle, one must use \is{-->} here).
-
-Type variables are \is{'a, 'b, ...} and a function types are written curried as \is{'a => 'b => 'c}.
-Types can be annotated to any symbol or term, e.g., \is{(x::nat) < 4 ==> x<=3}.
-Function application is written as \is{f x}, and function update is \is{f(x:=y)}.
+e.g., \is{\<forall>x. x<4 ==> x<=4} (in Isabelle, one must use \is{-->} here). Free variables are
+universally quantified at the top-level.
+
+Type variables are \is{'a, 'b, ...} and function types are written curried as \is{'a => 'b => 'c}.
+Types can be annotated to any term, e.g., \is{(x::nat) < 4 ==> x<=3}.
+Function application is written as \is{f x y}, and function update is \is{f(x:=y)}.
 For definitions, we use \is{==}, e.g., \is{f x == x+1}.
@@
 Algebraic datatypes are written as \is{'a option == Some (the: 'a) | None}. This also defines the selector function \is{the :: 'a option => 'a}.
-Values of the tuple type \is{'a_1 \x ... \x 'a_n} are written as \is{(a_1,...,a_n)}. Lists have type \is{'a list},
+Values of the tuple type \is{'a_1 \x ... \x 'a$_n$} are written as \is{(a_1,...,a$_n$)}. Lists have type \is{'a list},
 and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@xs_2} for list concatenation.
-We also use the list notation \is{[1,2,3]}. Then length of list \is{xs} is \is{|xs|}.
+We also use the list notation \is{[1,2,3]}. The length of list \is{xs} is \is{|xs|}.
 
@@
 Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
-Otherwise, we will explicitly write, e.g. \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.
+Otherwise, we will explicitly write, e.g., \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.
 
@@
 We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
-Also, we can use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.
+We also use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.
 
@@
 
-  Now assume that we have a state (memory) type \is{'\mu}, and an access report type \is{'\rho}, which forms a monoid (\is{0,+}).
+  Now assume that we have a state (memory) type \is{\mu}, and an access report type \is{\rho}, which forms a monoid (\is{0,+}).
   With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
   \begin{lstlisting}
-    'x M == '\mu => ('x \x '\rho \x '\mu) neM
+    'x M == \mu => ('x \x \rho \x \mu) neM
     return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
@@
   As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
-  Let \is{feasible :: '\rho => '\rho => bool} and \is{norace :: '\rho => '\rho => bool} be symmetric predicates,
-  and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
+  Let \is{feasible :: \rho => \rho => bool} and \is{norace :: \rho => \rho => bool} be symmetric predicates,
+  and let \is{combine :: (\rho \x \mu) => (\rho \x \mu) => (\rho \x \mu) } be a commutative operator to
   combine two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
@@
   results (\is{spec (\<lambda>_. False)}), the behaviour of the parallel composition is not clear.
-  For this reason, we fix an invariant \is{invar$_M$ :: ('\mu => ('x \x '\rho \x '\mu) neM) => bool},
+  For this reason, we fix an invariant \is{invar$_M$ :: (\mu => ('x \x \rho \x \mu) neM) => bool},
   which implies that every non-failing execution has at least one possible result.
@@
 
-    We then show that \is{malloc} satisfies the invariant of M: we correctly report the allocated block,
-    and we can select any fresh block. As our memory model guarantees an infinite supply of fresh blocks,
-    any finite set of blocks can be avoided:
+    We then show that \is{malloc} satisfies the invariant of M: we correctly report the allocated block.
+    Moreover, we can select any fresh block. As our memory model guarantees an infinite supply of fresh blocks,
+    any finite set of blocks can be avoided.
     \begin{lstlisting}
@@
     the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
-    The resulting memory will still contain the original array. Note that there is no explicit precondition ($\top$) that the array access
+    The resulting memory will still contain the original array. Note that there is no explicit precondition that the array access
     is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.
@@
     ivl$_R$ :: nat set list => nat set => bool
-    ivl$_R$ ls s == s = \<Union> set s & pw_disjoint ls & \<forall>x:set ls. finite x & x ~= {}
+    ivl$_R$ ls s == s = \<Union>set s  &  pw_disjoint ls  &  (\<forall>x:set ls. finite x & x ~= {})
 
@@
   Here, \is{al$_A$} is the refinement assertion from lists to the array list data structure from the IRF collections library.
-  As argument, it takes a refinement assertion for the list elements.
+  As argument, it takes the refinement relation for the list elements.
   Again, Sepref automatically generates imperative implementations of the \is{ivl} functions and proves the corresponding
@@
   \begin{lstlisting}
-  with_idxs s xs m ==
+  with_idxs s xs f ==
     assert (\<forall>i\<in>s. i<|xs| & xs!i ~= None);
@@
 
-  Figure~\ref{fig:psort} show our abstract parallel sorting algorithm \is{psort}.
+  Figure~\ref{fig:psort} shows our abstract parallel sorting algorithm \is{psort}.
   This algorithm is derived from the well-known quicksort and introsort algorithms~\cite{Muss97}:
@@
   \begin{lstlisting}
-    interpretation idx: pcmp lt_idx lt_idx$_\dagger$ arr$_A$ idx$_A$   \<proof>
-
     lt_idx xs i j == lt (xs!i) (xs!j)
     lt_idx$_\dagger$ xs$_\dagger$ i$_\dagger$ j$_\dagger$ == x$_\dagger$<-aget xs$_\dagger$ i$_\dagger$; y$_\dagger$<-aget xs$_\dagger$ j$_\dagger$; lt$_\dagger$ x$_\dagger$ y$_\dagger$
+
+    interpretation idx: pcmp lt_idx lt_idx$_\dagger$ arr$_A$ idx$_A$   \<proof>
   \end{lstlisting}
@@
   \end{tikzpicture}
-  \caption{Performance for sorting small arrays of randomly distributed integers and strings.
+  \caption{Runtimes for sorting small arrays of randomly distributed integers and strings.
     The $y$-axis shows the runtime in milliseconds, and the $x$-axis the array size. Note that both axis are logarithmic.
@@
     While the verification of our sorting algorithms uses a top-down approach, we actually started with
-    implementing, benchmarking, and fine-tuning the algorithms in C++. This gave us a quick option to
+    implementing, benchmarking, and fine-tuning the algorithms in C++. This gave us a quick way to
     find an algorithm that is efficient, and, at the same time not too complex for verification,
@@
     \subsection{Related Work}
-    While there is extensive work on parallel sorting algorithm (e.g.~\cite{AWFS22,CNLM08,AMI21}),
+    While there is extensive work on parallel sorting algorithms (e.g.~\cite{AWFS22,CNLM08,AMI21}),
     there seems to be almost no work on their formal verification. The only work we are aware of is

### 2024-03-22 papers/JAR_SI_ITP2022/main.tex
Handled reviewer's comments

@@
   The resulting algorithms' performance is competitive with their
-  counterparts implemented in C/C++.
+  counterparts implemented in C++.
 %
@@
 
-This paper is an extended version of our ITP 2023 paper~\cite{La22}.
+This paper is an extended version of our ITP 2022 paper~\cite{La22}.
 The main new contribution is a verified parallel partitioning algorithm,
@@
 
-\noindent The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.
+The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.
 
@@
 
-On top of the back-end, we formalize a concurrent separation logic~\cite{OH04} and implement a verification condition generator (VCG). (Section~\ref{sec:seplogic}).
+On top of the back-end, we formalize a concurrent separation logic~\cite{OH04} and implement a verification condition generator (VCG), cf.\ Section~\ref{sec:seplogic}.
 
@@
 In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
-e.g., \is{\<forall>x. x<4 ==> x<=4} (in Isabelle, one must use \is{-->} here). Free variables are
+e.g., \is{\<forall>x. x\:<\:4 ==> x\:<=\:4} (in Isabelle, one must use \is{-->} here). Free variables are
 universally quantified at the top-level.
@@
 Type variables are \is{'a, 'b, ...} and function types are written curried as \is{'a => 'b => 'c}.
-Types can be annotated to any term, e.g., \is{(x::nat) < 4 ==> x<=3}.
+Types can be annotated to any term, e.g., \is{(x::nat)\:<\:4 ==> x\:<=\:3}.
 Function application is written as \is{f x y}, and function update is \is{f(x:=y)}.
@@
 Values of the tuple type \is{'a_1 \x ... \x 'a$_n$} are written as \is{(a_1,...,a$_n$)}. Lists have type \is{'a list},
-and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@xs_2} for list concatenation.
+and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@\:xs_2} for list concatenation.
 We also use the list notation \is{[1,2,3]}. The length of list \is{xs} is \is{|xs|}.
@@
 Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
-Otherwise, we will explicitly write, e.g., \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.
-
-Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x>42}.
+Otherwise, we will explicitly write, e.g., \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of a finite set \is{s} is \is{|s|}.
+
+Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x\:>\:42}.
 We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
@@
   and \is{bind} sequentially composes the executions, threading through the state $\mu$ and adding up the access reports \is$r_1$ and \is$r_2$.
+Note that we use the names \is$r :: \rho$ for reports, \is$\mu :: \mu$ for memory, and \is$m :: _ M$ for monads.
 
@@
   \begin{lstlisting}
-    datatype addr == ADDR (bidx: nat) (idx: nat)
-    datatype ptr == PTR_NULL   |   PTR_ADDR (the_addr: addr)
-    datatype val == LL_INT lint   |   LL_STRUCT val list   |   LL_PTR ptr
-
-    datatype block == FRESH   |   FREED   |   is_alloc: ALLOC (vals: "val list")
-    typedef memory == { \mu :: nat => block. finite {b. \mu b ~= FRESH} }
-  \end{lstlisting}
-  A block is either fresh, freed, or allocated, and a memory is a mapping from block indexes to blocks,
+    bidx == nat
+    addr == ADDR bidx nat
+    ptr == PTR_NULL   |   PTR_ADDR (the_addr: addr)
+    val == LL_INT lint   |   LL_STRUCT val list   |   LL_PTR ptr
+
+    block == FRESH   |   FREED   |   is_alloc: ALLOC (vals: "val list")
+    typedef memory == { \mu :: bidx => block. finite {b. \mu b ~= FRESH} }
+  \end{lstlisting}
+  A block is either fresh, freed, or allocated, and a memory is a mapping from block indexes (\is{bidx}) to blocks,
   such that only finitely many blocks are not fresh. Every block's state transitions from fresh to
@@
   such that we will never run out of fresh blocks\footnote{If the actual system does run out of memory, we will terminate the program in a defined way.}.
-  An allocated block contains an array of values, modelled as a list. Thus, an address consists of a block number, and an index into the array.
+  An allocated block contains an array of values, modelled as a list. Thus, an address \is$ADDR bi i$ consists of a block index \is$bi$, and an index \is{i} into the array.
 
@@
   \begin{lstlisting}
-    valid \mu (ADDR b i) == is_alloc (\mu b) & i<|vals (\mu b)|
+    valid \mu (ADDR b i) == is_alloc (\mu b) & i < |vals (\mu b)|
     get \mu (ADDR b i) == vals (\mu b) ! i
@@
   \begin{lstlisting}
-    acc == addr set \x addr set \x nat set \x nat set
+    acc == addr set \x addr set \x bidx set \x bidx set
     0 == ( {},{},{},{} )
-    (r_1,w_1,a_1,f_1) + (r_2,w_2,a_2,f_2) == ( r_1\union r_2, w_1\union w_2, a_1\union a_2, f_1\union f_2 )
+    (r_1,w_1,a_1,f_1) + (r_2,w_2,a_2,f_2) == ( r_1 \union r_2, w_1 \union w_2, a_1 \union a_2, f_1 \union f_2 )
   \end{lstlisting}
@@
     (5)  addr \in w_2 \or b \in a_2 ==> get \mu' addr = get \mu_2 addr
-    (6)  addr ~: w_1\<union>w_2 & b ~: a_1\<union>a_2 ==> get \mu' addr = get \mu addr
+    (6)  addr ~: w_1\:\<union>\:w_2 & b ~: a_1\:\<union>\:a_2 ==> get \mu' addr = get \mu addr
   \end{lstlisting}
@@
   this is analogous to the sequential case~\cite{La19-llvm}.
-  Additionally, we define an instruction for parallel function call:
+  Additionally, we define an instruction for a parallel function call:
   \begin{lstlisting}
@@
   \is{parallel} function, which we implement using the Threading Building Blocks~\cite{TBB} library:
-  \begin{lstlisting}[language=C++]
+  \begin{lstlisting}[language=C++, basicstyle=\ttfamily]
   void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2) {
@@
   \begin{lstlisting}
-    \alpha :: memory => (addr => val option) \x (nat => nat option)
+    \alpha :: memory => (addr => val option) \x (bidx => nat option)
     \alpha \mu == (\alpha$_m$ \mu, \alpha$_b$ \mu)
@@
   that is, when executed on memory \is{\mu}, the program \is{c} does not fail, every return value \is{x} and new memory \is{\mu'} satisfies \is{Q},
-  and no memory described by the frame \is{amf} is accessed.
+  and no memory described by the frame \is{amf}~~is accessed.
 
@@
     We specify an operation to perform multiple swaps.
-It takes two disjoint sets of indices \is{s_1} and \is{s_2}, and a list \is{xs}.
+It takes two disjoint sets of indexes \is{s_1} and \is{s_2}, and a list \is{xs}.
 It then swaps each index in \is{s_1} with some index in \is{s_2}.
@@
       swap_spec_post s_1 s_2 xs xs' ==
-        mset xs' = mset xs & (\<forall>i\<notin>s_1 \union s_2. i < |xs| ==> xs'!i=xs!i)
-      & (\<forall>i\<in>s_1. \<exists>j\<in>s_2. xs'!i=xs!j) & (\<forall>j\<in>s_2. \<exists>i\<in>s_1. xs'!j=xs!i)
-    \end{lstlisting}
+        mset xs' = mset xs & (\<forall>i\<notin>s_1 \union s_2. i < |xs| ==> xs'!i = xs!i)
+      & (\<forall>i\<in>s_1. \<exists>j\<in>s_2. xs'!i = xs!j) & (\<forall>j\<in>s_2. \<exists>i\<in>s_1. xs'!j = xs!i)
+    \end{lstlisting}
+    Here, \is{mset xs} is the multiset of elements of the list \is{xs}, and \is{mset xs = mset xs'} is the standard way to express permutation in Isabelle.
 
@@
     \begin{lstlisting}
-      lget xs i == assert (i<|xs|); return xs!i
-      lset xs i x == assert (i<|xs|); return xs[i:=x]
+      lget xs i == assert (i < |xs|); return xs!i
+      lset xs i x == assert (i < |xs|); return xs[i:=x]
     \end{lstlisting}
@@
     \end{lstlisting}
+    where \is{ll_ofs_ptr} is the Isabelle-LLVM instruction for offsetting a pointer by an index.
     The abstract and concrete programs are linked by the following refinement theorems:
@@
       hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs$_\dagger$ xs$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ $\top$$\hspace*{-1pt}$ (lget xs i)
-      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r=xs$_\dagger$) (lset xs i x)
+      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r\:=\:xs$_\dagger$) (lset xs i x)
     \end{lstlisting}
@@
       aget, lget : arr$_A$ -> idx$_A$ -> id$_A$
-      aset, lset : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> id$_A$ -> arr$_A$:r [r=xs$_\dagger$]
+      aset, lset : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> id$_A$ -> arr$_A$:r [r\:=\:xs$_\dagger$]
     \end{lstlisting}
@@
     indicating that this argument will be destroyed, i.e., the refinement is no longer valid when the function returns.
-    Moreover, it binds the array argument to the name \is{xs$_\dagger$} and the result to $r$. These names are used in the pointer equality predicate \is{[r=xs$_\dagger$]}
+    Moreover, it binds the array argument to the name \is{xs$_\dagger$} and the result to $r$. These names are used in the pointer equality predicate \is{[r\:=\:xs$_\dagger$]}
     at the end, indicating that the result will be the same pointer as the array argument.
@@
     iv l h == (l,h)                                                 iv, {_..<_} : id$_R$ -> id$_R$ -> iv$_R$
-    iv_is_empty (l,h) == h<=l                                    iv_is_empty, (={}) : iv$_R$ -> id$_R$
+    iv_is_empty (l,h) == h\:<=\:l                                   iv_is_empty, (={}) : iv$_R$ -> id$_R$
     iv_inter (l_1,h_1) (l_2,h_2) == ( max l_1 l_2, min h_1 h_2 )     iv_inter, (\inter) : iv$_R$ -> iv$_R$ -> iv$_R$
-    iv_card (l,h) == if h<l then 0 else h-l                iv_card, |_| : iv$_R$ -> id$_R$
+    iv_card (l,h) == if h\:<\:l then 0 else h-l               iv_card, |_| : iv$_R$ -> id$_R$
   \end{lstlisting}
@@
     ivl_empty == []                                 ivl_empty, {} : ivl$_R$
-    ivl_is_empty ls == ls=[]                       ivl_is_empty, (={}) : ivl$_R$ -> id$_R$
+    ivl_is_empty ls == ls = []                     ivl_is_empty, (={}) : ivl$_R$ -> id$_R$
     ivl_dj_un s ls == s#ls                         ivl_dj_un, ($\dot\cup$) : id$_R$ -> ivl$_R$ -> ivl$_R$
@@
     ivl_split (s#ls) == (s,ls)                      ivl_split, split : ivl$_R$ -> id$_R$ \x ivl$_R$
-      split s == assert (s~={}); spec (s_1,s_2). s = s_1 $\dot\cup$ s_2 & s_1~={}
+      split s == assert (s ~= {}); spec (s_1,s_2). s = s_1 $\dot\cup$ s_2 & s_1 ~= {}
   \end{lstlisting}
@@
   list where independent of the interval implementation.
-  This is a design choice, and a more direct design, using \is{(nat \x nat) list} as intermediate data structure, is certainly possible.
+  This is a design choice, and a more direct design, e.g., using \is{(nat \x nat) list} as intermediate data structure, is certainly possible.
 
@@
       assert (|xs_1| = i \and |xs_2| = |xs| - i);
-      return (xs_1@xs_2)
+      return (xs_1@\:xs_2)
   \end{lstlisting}
@@
     awith_split, with_split :
-      arr$_A$^d:p -> idx$_A$ -> (arr$_A$^d:p_1 -> arr$_A$^d:p_2 -> arr$_A$:r_1 \x arr$_A$:r_2 [r_1=p_1 & r_2=p_2])
+      arr$_A$^d:p -> idx$_A$ -> (arr$_A$^d:p_1 -> arr$_A$^d:p_2 -> arr$_A$:r_1 \x arr$_A$:r_2 [r_1=\:p_1 & r_2=\:p_2])
       -> arr$_A$:r [r=p]
@@
     hnr (arr$_A$ p_1 xs_1 * arr$_A$ p_2 xs_2) (f$_\dagger$ p_1 p_2) \<box>
-        (arr$_A$ \x arr$_A$) (%(r_1,r_2). r_1=p_1 \and r_2 = p_2)
+        (arr$_A$ \x arr$_A$) (%(r_1,r_2). r_1\:=\:p_1 \and r_2\:=\:p_2)
         (f xs_1 xs_2)"
@@
     hnr (arr$_A$ p xs * idx$_A$ i$_\dagger$ i) (awith_split i$_\dagger$ p f$_\dagger$)
-        (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r=p)
+        (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r\:=\:p)
         (with_split i xs f)
@@
 
-  Splitting an array into two halves allows us to abstractly treat the array and its two halves just as lists,
-  which simplifies the abstract proofs: the fact that the two halves come from the same array is only visible
-  at a later refinement stage. However, while splitting an array in halves is adequate for many operations,
+  Splitting an array into two parts allows us to abstractly treat the array and its two parts just as lists,
+  which simplifies the abstract proofs: the fact that the two parts come from the same array is only visible
+  at a later refinement stage. However, while splitting an array in parts is adequate for many operations,
   it is not a workable abstraction for swapping multiple elements in parallel (cf.~Sec.~\ref{ex:swap_spec}):
@@
   \end{lstlisting}
+  Here, \is{list_all2} is the natural relator for lists, i.e., \is{list_all2 (\<or>) s_1 s_2} means that the lists
+  \is{s_1} and \is{s_2} have the same length, and for each index, the element in at least one of the lists is true.
+
   We also define functions to split and join lists:
@@
   and \is{sl_join xs_1 xs_2} joins the elements of two (compatible) lists.
+The function \is{map2 f xs ys == map (\<lambda>(x,y).
+f x y) (zip xs ys)} combines two lists element-wise, using the binary function \is$f$.
 
@@
   with_idxs s xs f ==
-    assert (\<forall>i\<in>s. i<|xs| & xs!i ~= None);
+    assert (\<forall>i\<in>s. i\:<\:|xs| & xs!i ~= None);
     let (xs_1, xs_2) = (sl_split s xs, sl_split (-s) xs);
@@
   \begin{lstlisting}
-  oarr$_A$ p xs == range {i | i. i<length xs & xs!i~=None}) p (%i. the (xs!i))
+  oarr$_A$ p xs == range {i | i. i\:<\:length xs & xs!i\:~=\:None}) p (%i. the (xs!i))
   \end{lstlisting}
@@
   \begin{lstlisting}
-    olget xs i == assert (i<|xs| & xs!i ~= None); return the (xs!i)
-    olset xs i x == assert (i<|xs| & xs!i ~= None); return the (xs[i:=x])
+    olget xs i == assert (i < |xs| & xs!i ~= None); return the (xs!i)
+    olset xs i x == assert (i < |xs| & xs!i ~= None); return the (xs[i:=x])
 
@@
   \begin{lstlisting}
-  sort_spec xs == spec xs'. mset xs'=mset xs & sorted xs
+  sort_spec xs == spec xs'. mset xs' = mset xs & sorted xs
   \end{lstlisting}
@@
     psort xs n ==
-      assert n=|xs|; if n<=1 then return xs else psort_aux$\hspace*{-1pt}$ xs$\hspace*{-1pt}$ n$\hspace*{-1pt}$ (log2$\hspace*{-1pt}$ n$\hspace*{-1pt}$ *$\hspace*{-1pt}$ 2)
+      assert n\:=\:|xs|; if n\:<=\:1 then return xs else psort_aux$\hspace*{-1pt}$ xs$\hspace*{-1pt}$ n$\hspace*{-1pt}$ (log2$\hspace*{-1pt}$ n$\hspace*{-1pt}$ *$\hspace*{-1pt}$ 2)
 
     psort_aux xs n d ==
-      assert n=|xs|
-      if d=0 \or n<100000 then sort_spec xs $\label{l:psort:fallback}$
+      assert n = |xs|
+      if d = 0 \or n < 100000 then sort_spec xs $\label{l:psort:fallback}$
       else
         (xs,m) <- partition_spec xs; $\label{l:psort:partition}$
-        let bad = m<n div 8 \or (n-m < n div 8)  $\label{l:psort:bad}$
+        let bad = m < n div 8 \or (n-m < n div 8)  $\label{l:psort:bad}$
         (_,xs) <- with_split m xs (%xs_1 xs_2.
@@
   \begin{lstlisting}
-    psort$_\dagger$, psort : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> arr$_A$:r [r=xs$_\dagger$]
+    psort$_\dagger$, psort : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> arr$_A$:r [r = xs$_\dagger$]
   \end{lstlisting}
@@
     (psort$_\dagger$ xs$_\dagger$ n$_\dagger$)
-    (%r. r=xs$_\dagger$  *  \<exists> xs'. arr$_A$ xs$_\dagger$ xs'  *  sorted xs'  *  mset xs' = mset xs)"
+    (%r. r = xs$_\dagger$  *  \<exists> xs'. arr$_A$ xs$_\dagger$ xs'  *  sorted xs'  *  mset xs' = mset xs)"
   \end{lstlisting}
@@
       (xs,ls,rs) <- ppart_spec p xs;   --"\rightcomment{Step 1 and 2}"
-      let m = |ls|; ss = {i\<in>ls. m<=i}; bs = {i\<in>rs. i<m};   --"\rightcomment{Step 3}"
+      let m = |ls|; ss = {i\<in>ls. m\:<=\:i}; bs = {i\<in>rs. i\:<\:m};   --"\rightcomment{Step 3}"
       xs <- swap_spec ss bs xs;        --"\rightcomment{Step 4}"
@@
     ((ss_1,ss_2),(bs_1,bs_2))
-      <- spec ((ss_1,ss_2),(bs_1,bs_2)). ss=ss_1$\dot\cup$ss_2 & bs=bs_1$\dot\cup$bs_2 & |ss_1|=|bs_1| & ss_1~={}
-
-    if (ss_2={}) then swap_opt_spec ss_1 bs_1 xs
+      <- spec ((ss_1,ss_2),(bs_1,bs_2)). ss\:=\:ss_1$\dot\cup$ss_2 & bs\:=\:bs_1$\dot\cup$bs_2 & |ss_1|\:=\:|bs_1| & ss_1~={}
+
+    if (ss_2 = {}) then swap_opt_spec ss_1 bs_1 xs
     else
@@
       height=.249\textheight,
-      bar width=3pt,
+      bar width=2.5pt,
       xtick=data,
@@
   \begin{figure}
-    \plottim{Laptop, uint64}{data_laptop_uint64}{xticklabels={}}{\legend{}}
-    \plottim{Server, uint64}{data_server_uint64}{xticklabels={},ymax=1000}{\legend{}}
-    \plottim{Laptop, string}{data_laptop_llstring}{xticklabels={}}{\legend{}}
+    \plottim{Laptop, uint64}{data_laptop_uint64}{xticklabels={},ytick distance={200}}{\legend{}}
+    \plottim{Server, uint64}{data_server_uint64}{xticklabels={},ytick distance={200},ymax=1100}{\legend{}}
+    \plottim{Laptop, string}{data_laptop_llstring}{xticklabels={},ytick distance={200},ymax=1100}{\legend{}}
     \plottim{Server, string}{data_server_llstring}{
       xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
-      ymax=1000,
+      ytick distance={200},ymax=1000,
       legend style = {
@@
 
-  We also benchmarked against the state-of-the-art implementations \is{std::sort} with execution policy \is{par_unseq} from the
-  GNU C++ standard library~\cite{libstdc++},
+  We also benchmarked against the old verified algorithm with sequential partitioning from our ITP 2022 paper~\cite{La22}, as well as against the state-of-the-art implementations \is{std::sort} with execution policy \is{par_unseq} from the GNU C++ standard library~\cite{libstdc++},
   and \is{sample_sort} from the Boost C++ libraries~\cite{boost,boost-sort}.
@@
   Only for a few string-sorting benchmarks, it is slightly slower. We leave improving on this for future work.
+  Compared to the old verified algorithm, the parallel partitioning algorithm is more efficient in many cases, and there are only a few cases where it is slightly less efficient.
+
 %
@@
   We also measured the speedup that the implementations achieve for a certain number of cores.
-  The results are displayed in Figure~\ref{fig:speedup}: again, our verified implementation is clearly competitive.
+  The results are displayed in Figure~\ref{fig:speedup}: again, our verified implementation is clearly competitive, and it scales better than the old verified algorithm.
 
@@
     without having to verify each intermediate step towards this algorithm.
-    Only then we used our top-down approach to first formalize the abstract ideas behind the algorithm,
+    Only then did we use our top-down approach to first formalize the abstract ideas behind the algorithm,
     and then refine it to an efficient implementation close to what we had written in C++.
@@
     as our initial abstract algorithm.
+
+    \subsection{Statistics}
+    The Isabelle Refinement Framework for LLVM consists of roughly 45k lines of theory text and Isabelle-ML code (referred to as kLOC from now on). The sorting algorithms comprise another 14kLOC.
+
+    Integrating the initial version of the parallel semantics into the framework, and verifying the parallel sorting algorithm with sequential partitioning, as reported in our ITP 2022 paper~\cite{La22}, took us roughly 6 months\footnote{As we do not have precise work time logs, we report calendar months during which we worked almost full-time on the project.}.
+During this time, we abandoned an initial formalization for the Imperative/HOL back end, as the achievable performance was unsatisfactory.
+We also abandoned a naive formalization of the parallel operator for the LLVM back end, to finally arrive at what is described in our ITP 2022 paper.
+Effectively, we added about 6kLOC to the framework and adapted most of the remaining code.
+The parallel sorting algorithm adds another kLOC.
+    The parallel partitioning algorithm and its support data structures add 4kLOC, and took us one month to develop.
+
+%
+%     Developing the parallel partitioning algorithm took us another month, adding 4000kLOC.
+%
+%
+%     May 2021--October 2021  first versions (Imp-HOL, par as seq semantics, parallel sorting) 5month
+%       framework: +3kLOC, adapted most of 40kLOC
+%       sorting: +1000kLOC
+%
+%     Jan 2022 second version (ndet par semantics) 1month
+%       framework: heavily changed 3000kLOC, adapted rest (40 kLOC)
+%
+%     Mar 2023 parallel partitioning 1 month
+%       +4000kLOC
+%
+%     We have started developing the parallel extension to Isabelle LLVM in May 2021: we first worked on a prototype for the Imperative HOL back end, which we then ported to the LLVM back end, to obtain the first verified parallel sorting algorithms by October.
+We then decided to change our initial semantics of parallel calls again to what is described in the paper
+%
+%
+%     Code: the parallel extension: around 3000LOC added to the Backend and Program Logic, and adapted a substantial amount of the 40kLOC that made up the higher levels of the Refinement Framework.
+%
+%
+%
+%    par-sort (seq partitioning):
+%
+%    431 Sorting_Parsort.thy
+%    320 Sorting_Sample_Partition.thy
+%
+%
+%    parallel partioning:
+%
+%    705 IICF_DS_Array_Idxs.thy
+%    884 IICF_DS_Interval_List.thy
+%    512 IICF_Shared_Lists.thy
+%    536 Sorting_Guarded_Partition.thy
+%   1612 Sorting_Par_Partition.thy
+%
+%   5000 total
+%
+%
+%
+% Back-End basic: 6429
+% Low-Level Alg + ds: 3544
+% lib: 12145
+% sepref: 19104
+% Program logic and vcg: 3243
+%
+% NE-Monad (from AFP) 10kLOC
+%
+% examples/sorting: 14001
 
@@
     there seems to be almost no work on their formal verification. The only work we are aware of is
-    a distributed merge sort algorithm~\cite{HKBK20}, for which ''no effort has been made to make it efficient''\cite[Sec.~2]{HKBK20},
+    a distributed merge sort algorithm~\cite{HKBK20}, for which ''no effort has been made to make it efficient''~\cite[Sec.~2]{HKBK20},
     nor any executable code has been generated or benchmarked. Another verification~\cite{SaHu20} uses
@@
 
+      An important aspect is the scalability of our tools.
+The current implementation scales to small software, like the verified IsaSAT-solver~\cite{FlLa23}, which actually uses our sequential sorting algorithms.
+However, for projects of this size, the times needed by Sepref and the LLVM code exporter easily get into the order of dozens of minutes, and some manual performance tweaks are required (cf.~\cite{FlLa23}[Sec.~5]).
+      Further improvement of the scalability is left to future work.
+
       Another direction for future work is to further optimize our verified sorting algorithm.

### 2024-03-22 statistics.sh

@@ -1,0 +1,36 @@
+#!/bin/bash
+
+DDIR="statistics"
+
+mkdir -p "$DDIR"
+
+
+FFILE="$DDIR/files.txt"
+
+if ! [ -f "$FFILE" ]; then
+  echo -n "Collecting active files ..."
+
+  isabelle build -D thys -l -n > "$FFILE"
+
+  echo "Done"
+fi
+
+DIRS="thys/basic thys/ds thys/lib thys/sepref thys/vcg thys/examples/sorting"
+
+for i in $DIRS; do
+
+  j=${i%/}
+  j=${j#thys/}
+
+  FILES="$DDIR/files_${j//\//-}.txt"
+
+
+  if grep "$i" $FFILE > "$FILES"; then
+
+    echo "$j: $(cat "$FILES" | xargs cat | wc -l)"
+
+  fi
+
+
+done
+

### 2024-03-23 benchmarks/sorting/distr_report.tex
better reproducible benchmark scripts

\documentclass[a4]{article}

\usepackage[T1]{fontenc}
\usepackage{color}

\usepackage{pgfplots}
\usepgfplotslibrary{groupplots}

\begin{document}

  \newcommand{\plottim}[3]{
  \begin{tikzpicture}
    \begin{axis}[
      ymajorgrids,
      title={\large #1},
      title style={at={(0.12,.75)}},
      ybar=0pt,
      width=\textwidth,
      height=.249\textheight,
      bar width=2.5pt,
      xtick=data,
      symbolic x coords={rev-sorted-end-10,rev-sorted-end-1,sorted-end-.1,almost-sorted-50,random-boolean,organ-pipe,sorted-end-10,equal,rev-sorted-middle-.1,rev-sorted,sorted-middle-1,rev-sorted-middle-10,random,almost-sorted-.1,sorted,rev-sorted-middle-1,sorted-middle-.1,almost-sorted-10,almost-sorted-1,sorted-middle-10,rev-sorted-end-.1,sorted-end-1,random-dup-10,}
      ,
      xlabel near ticks, x tick label style={rotate=45,anchor=east,font=\tiny},
      ytick distance={200},
      legend style = {
%         legend pos= north east,
        at = {(.87,.97)},
        cells={anchor=west},
        font=\tiny
      },
    ]

      \input{#2}

      \addplot[black, thin, sharp plot,update limits=false]
	coordinates {([normalized]-10,1) ([normalized]100,1)};
% 	node[above] at (axis cs:1950,4.3e7) {Houses};
    \end{axis}
  \end{tikzpicture}
  }

  \plottim{uint64}{d_uint64.tex}

  \clearpage

  \plottim{llstring}{d_llstring.tex}

\end{document}

### 2024-03-23 benchmarks/sorting/all_benchmarks.sh

@@ -1,168 +1,0 @@
-#!/bin/bash
-
-# make do_benchmark
-
-########################## BEGIN CONFIG SECTION ##################
-
-# Defaults
-
-export NITER=10   # Number of iterations
-export NIELEM=100000000 # Number of integer elements
-export NSELEM=10000000 # Number of string elements
-
-# Quick benchmark run
-# export NITER=3   # Number of iterations
-# export NIELEM=1000 # Number of integer elements
-# export NSELEM=1000 # Number of string elements
-
-export INT_DATA="random random-dup-10 random-boolean organ-pipe sorted equal almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
-
-export STR_DATA="random random-dup-10 random-boolean organ-pipe sorted equal   almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
-
-
-export ALL_INT_ALGOS="isabelle::sort isabelle::parqsort isabelle::pparqsort isabelle::pdqsort std::sort std::parsort boost::pdqsort boost::pdqsort_bp boost::sample_sort boost::bi_sort naive::parqsort naive::psort naive::parqsort_vs naive::parqsort_vs_pp"
-
-export ALL_STR_ALGOS="isabelle::sort isabelle::parqsort isabelle::pparqsort isabelle::pdqsort boost::pdqsort std::sort std::parsort boost::sample_sort boost::bi_sort naive::parqsort naive::parqsort_vs naive::parqsort_vs_pp"
-
-export INT_ALGOS="isabelle::parqsort isabelle::pparqsort std::parsort boost::sample_sort"
[…]

### 2024-03-23 benchmarks/sorting/distr_benchmarks.sh

@@ -1,0 +1,174 @@
+#!/bin/bash
+
+# make do_benchmark
+
+########################## BEGIN CONFIG SECTION ##################
+
+set -e
+
+# Defaults
+
+export NITER=10   # Number of iterations
+export NIELEM=100000000 # Number of integer elements
+export NSELEM=10000000 # Number of string elements
+
+# Quick benchmark run
+# export NITER=3   # Number of iterations
+# export NIELEM=10000000 # Number of integer elements
+# export NSELEM=1000000 # Number of string elements
+
+export INT_DATA="random random-dup-10 random-boolean organ-pipe sorted equal almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
+
+export STR_DATA="random random-dup-10 random-boolean organ-pipe sorted equal   almost-sorted-.1 almost-sorted-1 almost-sorted-10 almost-sorted-50 sorted-end-.1 sorted-end-1 sorted-end-10 sorted-middle-.1 sorted-middle-1 sorted-middle-10 rev-sorted rev-sorted-end-.1 rev-sorted-end-1 rev-sorted-end-10 rev-sorted-middle-.1 rev-sorted-middle-1 rev-sorted-middle-10"
+
+
+export ALL_INT_ALGOS="isabelle::sort isabelle::parqsort isabelle::pparqsort isabelle::pdqsort std::sort std::parsort boost::pdqsort boost::pdqsort_bp boost::sample_sort boost::bi_sort naive::parqsort naive::psort naive::parqsort_vs naive::parqsort_vs_pp"
+
+export ALL_STR_ALGOS="isabelle::sort isabelle::parqsort isabelle::pparqsort isabelle::pdqsort boost::pdqsort std::sort std::parsort boost::sample_sort boost::bi_sort naive::parqsort naive::parqsort_vs naive::parqsort_vs_pp"
+
[…]

### 2024-03-23 benchmarks/sorting/eval_benchmark.sh

@@ -2,24 +2,54 @@
 
 set -e
 
-rm -f ~/tmp/data*.tex
+function error() {
+  echo "Error: $@"
+  exit 1
+}
 
-# cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
-cat log/server/par-2023.log | ./__eval_benchmark.awk S uint64 > ~/tmp/data_server_uint64.tex
-cat log/server/par-2023.log | ./__eval_benchmark.awk S llstring > ~/tmp/data_server_llstring.tex
+rm -rf generate
+mkdir generate
 
-cat log/par-2023.log | ./__eval_benchmark.awk L uint64 > ~/tmp/data_laptop_uint64.tex
-cat log/par-2023.log | ./__eval_benchmark.awk L llstring > ~/tmp/data_laptop_llstring.tex
+LOGFILE="log/sortbench-latest.log"
 
-# cat log/par-2021.log | ./__eval_benchmark.awk
+if [ "$1" != "" ]; then
+  LOGFILE="$1"
+fi
+
+OUTFILE="${LOGFILE%.log}.pdf"
+
+
+./__eval_benchmark.awk uint64 < "$LOGFILE" > generate/d_uint64.tex
+./__eval_benchmark.awk llstring < "$LOGFILE" > generate/d_llstring.tex
+
+cp distr_report.tex generate/
+
+cd generate
+pdflatex distr_report
+cd ..
+
+cp "generate/distr_report.pdf" "$OUTFILE"
+echo "Generated report at $OUTFILE"
+
+
+# rm -f ~/tmp/data*.tex
 #
-# exit
-
-# grep ">L" ~/tmp/data.tex | sed -re 's/^>. //' > repeat_laptop.sh
-# grep ">S" ~/tmp/data.tex | sed -re 's/^>. //' > repeat_server.sh
-
-
-# gawk '$1=="@" { echo=($2=="server"); next }  echo {print} ' ~/tmp/data.tex  > ~/tmp/data_server.tex
-# gawk '$1=="@" { echo=($2=="laptop"); next }  echo {print} ' ~/tmp/data.tex > ~/tmp/data_laptop.tex
+# # cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
+# cat log/server/par-2023.log | ./__eval_benchmark.awk S uint64 > ~/tmp/data_server_uint64.tex
[…]

### 2024-03-23 benchmarks/sorting/eval_smallarrays.sh

@@ -1,17 +1,43 @@
 #!/bin/bash
 
-rm -f ~/tmp/sadata*.tex
+LOGFILE="log/sortbench-smallarrays-latest.log"
 
-# cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
+if [ "$1" != "" ]; then
+  LOGFILE="$1"
+fi
 
-cat log/par-smallarrays-2023.log | ./__eval_smallarrays.awk uint64 > ~/tmp/sadata_uint64_laptop.tex
-cat log/server/par-smallarrays-2023.log | ./__eval_smallarrays.awk uint64 > ~/tmp/sadata_uint64_server.tex
-
-cat log/par-smallarrays-2023.log | ./__eval_smallarrays.awk llstring > ~/tmp/sadata_llstring_laptop.tex
-cat log/server/par-smallarrays-2023.log | ./__eval_smallarrays.awk llstring > ~/tmp/sadata_llstring_server.tex
+OUTFILE="${LOGFILE%.log}.pdf"
 
 
-# gawk '$1=="@" { echo=($2=="server"); next }  echo {print} ' ~/tmp/sudata.tex  > ~/tmp/sudata_server.tex
-# gawk '$1=="@" { echo=($2=="laptop"); next }  echo {print} ' ~/tmp/sudata.tex > ~/tmp/sudata_laptop.tex
+rm -rf generate
+mkdir -p generate
+
+./__eval_smallarrays.awk uint64 < "$LOGFILE" > generate/d_uint64.tex
+./__eval_smallarrays.awk llstring < "$LOGFILE" > generate/d_llstring.tex
+
+cp smallarrays_report.tex generate/
+
+cd generate
+pdflatex smallarrays_report
+cd ..
+
+cp "generate/smallarrays_report.pdf" "$OUTFILE"
+echo "Generated report at $OUTFILE"
+
+
+
+# rm -f ~/tmp/sadata*.tex
 #
-cp ~/tmp/sadata_{uint64,llstring}_{laptop,server}.tex ../../papers/JAR_SI_ITP2022/
+# # cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
+#
+# cat log/par-smallarrays-2023.log | ./__eval_smallarrays.awk uint64 > ~/tmp/sadata_uint64_laptop.tex
[…]

### 2024-03-23 benchmarks/sorting/eval_speedup.sh

@@ -1,17 +1,45 @@
 #!/bin/bash
 
-rm -f ~/tmp/sudata*.tex
+LOGFILE="log/sortbench-speedup-latest.log"
 
-# cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
+if [ "$1" != "" ]; then
+  LOGFILE="$1"
+fi
 
-cat log/par-speedup-2023.log | ./__eval_speedup.awk uint64 > ~/tmp/sudata_uint64_laptop.tex
-cat log/server/par-speedup-2023.log | ./__eval_speedup.awk uint64 > ~/tmp/sudata_uint64_server.tex
-
-cat log/par-speedup-2023.log | ./__eval_speedup.awk llstring > ~/tmp/sudata_llstring_laptop.tex
-cat log/server/par-speedup-2023.log | ./__eval_speedup.awk llstring > ~/tmp/sudata_llstring_server.tex
+OUTFILE="${LOGFILE%.log}.pdf"
 
 
-# gawk '$1=="@" { echo=($2=="server"); next }  echo {print} ' ~/tmp/sudata.tex  > ~/tmp/sudata_server.tex
-# gawk '$1=="@" { echo=($2=="laptop"); next }  echo {print} ' ~/tmp/sudata.tex > ~/tmp/sudata_laptop.tex
+rm -rf generate
+mkdir -p generate
+
+./__eval_speedup.awk uint64 < "$LOGFILE" > generate/d_uint64.tex
+./__eval_speedup.awk llstring < "$LOGFILE" > generate/d_llstring.tex
+
+cp speedup_report.tex generate/
+
+cd generate
+pdflatex speedup_report
+cd ..
+
+cp "generate/speedup_report.pdf" "$OUTFILE"
+echo "Generated report at $OUTFILE"
+
+
 #
-cp ~/tmp/sudata_{uint64,llstring}_{laptop,server}.tex ../../papers/JAR_SI_ITP2022/
+#
+#
+# rm -f ~/tmp/sudata*.tex
+#
+# # cat log/sortbench-2020-01-all.log | ./__eval_benchmark.awk > /tmp/data.tex
+#
+# cat log/par-speedup-2023.log | ./__eval_speedup.awk uint64 > ~/tmp/sudata_uint64_laptop.tex
[…]

### 2024-03-23 benchmarks/sorting/smallarrays_benchmarks.sh

@@ -16,8 +16,8 @@ export S_SIZES="1000 2783 7743 21544 59948 166810 464159 1291550 3593814 1000000
 export INT_DATA="random"
 export STR_DATA="random"
 
-export INT_ALGOS="isabelle::pparqsort std::parsort boost::sample_sort"
-export STR_ALGOS="isabelle::pparqsort std::parsort boost::sample_sort"
+export INT_ALGOS="isabelle::pparqsort isabelle::parqsort std::parsort boost::sample_sort"
+export STR_ALGOS="isabelle::pparqsort isabelle::parqsort std::parsort boost::sample_sort"
 
 # export INT_ALGOS="isabelle::pparqsort naive::parqsort_vs_pp"
 # export STR_ALGOS="isabelle::pparqsort naive::parqsort_vs_pp"
@@ -33,8 +33,13 @@ export STR_ALGOS="isabelle::pparqsort std::parsort boost::sample_sort"
 
 ########################## END CONFIG SECTION ##################
 
+LOGDATE="$(date -Iseconds)"
+export LOGFILE="log/sortbench-smallarrays-$LOGDATE.log"
 
-export LOGFILE="log/sortbench-smallarrays-$(date -Iseconds).log"
+rm -f "log/sortbench-smallarrays-latest.log"
+ln -s "sortbench-smallarrays-$LOGDATE.log" "log/sortbench-smallarrays-latest.log"
+
+
 
 echo "Writing log to $LOGFILE"
 

### 2024-03-23 benchmarks/sorting/speedup_benchmarks.sh

@@ -4,6 +4,7 @@
 
 ########################## BEGIN CONFIG SECTION ##################
 
+
 mtype=$1
 if [[ $mtype == "server" ]]; then
   function tmask() {
@@ -20,7 +21,16 @@ elif [[ $mtype == "laptop" ]]; then
   TNUM=12
 else
   echo "Unknown machine config: $mtype"
-  exit 1
+
+  TNUM=$(cat /proc/cpuinfo | grep '^processor' | wc -l)
+
+  echo "Guessing processor allocation: $TNUM cores's at Ids: 0..<$TNUM"
+  echo "Guess may be wrong depending on you machine!"
+
+  function tmask() {
+    printf "0x%x" $((2**$1 - 1 ))
+  }
+
 fi
 
 # Defaults
@@ -38,14 +48,15 @@ export INT_DATA="random"
 
 export STR_DATA="random"
 
-# export INT_ALGOS="isabelle::parqsort isabelle::pparqsort std::parsort boost::sample_sort"
-# export STR_ALGOS="isabelle::parqsort isabelle::pparqsort std::parsort boost::sample_sort"
-
-export INT_ALGOS="isabelle::pparqsort naive::parqsort_vs_pp"
-export STR_ALGOS="isabelle::pparqsort naive::parqsort_vs_pp"
+export INT_ALGOS="isabelle::parqsort isabelle::pparqsort std::parsort boost::sample_sort"
+export STR_ALGOS="isabelle::parqsort isabelle::pparqsort std::parsort boost::sample_sort"
 
 
 # CUSTOM Ad-HOC Overrides
+
+# export INT_ALGOS="isabelle::pparqsort naive::parqsort_vs_pp"
+# export STR_ALGOS="isabelle::pparqsort naive::parqsort_vs_pp"
+
 
 # export INT_DATA="random organ-pipe"
 # export STR_DATA=""
@@ -56,7 +67,11 @@ export STR_ALGOS="isabelle::pparqsort naive::parqsort_vs_pp"
 ########################## END CONFIG SECTION ##################
 
 
-export LOGFILE="log/sortbench-speedup-$(date -Iseconds).log"
+export LOGDATE="$(date -Iseconds)"
+
+export LOGFILE="log/sortbench-speedup-$LOGDATE.log"
+rm -f "log/sortbench-speedup-latest.log"
[…]

### 2024-03-23 papers/JAR_SI_ITP2022/main.tex
Diagrams uod

@@
 %         legend pos= north east,
-        at = {(.87,.97)},
+        at = {(.77,.97)},
         cells={anchor=west},
@@
       legend style = {
-        at = {(.97,.4)},
+        at = {(.97,.35)},
         cells={anchor=west},
-        font=\footnotesize
+        font=\tiny
       },

### 2024-03-25 papers/JAR_SI_ITP2022/main.tex
last changes before revision submission

@@
 
-The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.
+Isabelle LLVM is hosted at \href{https://lammich.github.io/isabelle_llvm/index.html}{lammich.github.io/isabelle\_llvm}.
+The version described in this paper has been archived~\cite{Isabelle_LLVM_artefact}.
+%
+% The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.
 
@@
   \end{lstlisting}
-  This yields sorting algorithms for sorting indexes, taking an extra parameter for the array to index into.
-  For our sampling application, we use \is{idx.introsort xs}.
+  This instantiates the generic sorting algorithms defined in the \is{pcmp} locale to use \is{lt_idx xs} as comparison function, taking \is{xs} as an extra parameter.
+To sort our list \is{$is$} of sample indexes into \is{xs}, we use the instantiation of the introsort algorithm: \is{idx.introsort xs $is$}.
 
@@
 
-    \subsection{Statistics}
+    \subsection{Development Effort}
     The Isabelle Refinement Framework for LLVM consists of roughly 45k lines of theory text and Isabelle-ML code (referred to as kLOC from now on). The sorting algorithms comprise another 14kLOC.
@@
       An important aspect is the scalability of our tools.
-The current implementation scales to small software, like the verified IsaSAT-solver~\cite{FlLa23}, which actually uses our sequential sorting algorithms.
+The current implementation scales to small software systems, like the verified IsaSAT-solver~\cite{FlLa23}, which actually uses our sequential sorting algorithms.
 However, for projects of this size, the times needed by Sepref and the LLVM code exporter easily get into the order of dozens of minutes, and some manual performance tweaks are required (cf.~\cite{FlLa23}[Sec.~5]).
